### 2019-12-17 interim/interim.tex
Add LaTeX build directory to .gitignore

\documentclass[12pt,twoside]{report}

\newcommand{\reporttitle}{Type-safe Webservices Generation: Interim Report}
\newcommand{\reportauthor}{Anson Miu}
\newcommand{\reporttype}{MEng Individual Project}
\newcommand{\supervisor}{Prof. Nobuko Yoshida}
\newcommand{\sndmarker}{TBC}
\newcommand{\cid}{your college-id number}

% include files that load packages and define macros
\input{includes} % various packages needed for maths etc.
\input{notation} % short-hand notation and macros

%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
% front page
\input{titlepage}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\doublespacing
\tableofcontents
\singlespacing

%%%%%%%%%%%%%%%%%%%%%%%%%%%% Main document

\chapter{Introduction}

\chapter{Background}

\chapter{Project Plan}
\section{Theory}
\section{Implementation}
\section{Timetable}

\chapter{Evaluation Plan}
\section{Theory}
\section{Implementation}

\clearpage

\end{document}
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2019-12-17 interim/sections/misc.tex
Add LaTeX build directory to .gitignore

\section{Introduction}
This is a template for coursework submission. Many macros and definitions can be found in \texttt{notation.tex}. This document is not an introduction to LaTeX. General advice if get stuck: Use your favorite search engine. A great source is also \mbox{\url{https://en.wikibooks.org/wiki/LaTeX}}.

\section{Basics}

\subsection{Figures}
A figure can be included as follows:
\begin{figure}[tb]
\centering % this centers the figure
\includegraphics[width = 0.7\hsize]{./figures/imperial} % this includes the figure and specifies that it should span 0.7 times the horizontal size of the page
\caption{This is a figure.} % caption of the figure
\label{fig:imperial figure} % a label. When we refer to this label from the text, the figure number is included automatically
\end{figure}
Fig.~\ref{fig:imperial figure} shows the Imperial College logo.

Some guidelines:
\begin{itemize}
\item Always use vector graphics (scale free)
\item In graphs, label the axes
\item Make sure the font size (labels, axes) is sufficiently large
\item When using colors, avoid red and green together (color blindness)
\item Use different line styles (solid, dashed, dotted etc.) and different markers to make it easier to distinguish between lines
\end{itemize}

\subsection{Notation}
\begin{table}[tb]
\caption{Notation}
\label{tab:notation}
\centering
\begin{tabular}{ll}
Scalars & $x$\\
Vectors & $\vec x$\\
Matrices & $\mat X$\\
Transpose & $\T$\\
Inverse & $\inv$\\
Real numbers & $\R$\\
Expected values & $\E$\\
\end{tabular}
\end{table}
Table~\ref{tab:notation} lists some notation with some useful shortcuts (see latex source code).

\subsubsection{Equations}
Here are a few guidelines regarding equations
\begin{itemize}
\item Please use the \texttt{align} environment for equations (\texttt{eqnarray} is buggy)
\item Please number all equations: It will make things easier when we need to refer to equation numbers. If you always use the \texttt{align} environment, equations are numbered by default.
\item Vectors are by default column vectors, and we write 
\begin{align}
\vec x &= \colvec{1,2}
\end{align}
\item Note that the same macro (\texttt{$\backslash$colvec}) can produce vectors of variable lengths, as
\begin{align}
\vec y &= \colvec{1,2,3,4}
\end{align}
\item Matrices can be created with the same command. The \& switches to the next column:
\begin{align}
\mat A = \begin{bmatrix}
1 & 2 & 3\\
3 & 4 & 5
\end{bmatrix}
\end{align}
\item Determinants. We provide a simple macro (\texttt{$\backslash$matdet}) whose argument is just a matrix array:
\begin{align}
\matdet{
1 & 2 & 3\\
3 & 4 & 5\\
2 & 2 & 2
}
\end{align}
\item If you do longer manipulations, please explain what you are doing: Try to avoid sequences of equations without text breaking up. Here is an example:
We consider
\begin{align}
U_1 = [\colvec{1,1,0,0},\, \colvec{0,1,1,0},\, \colvec{0,0,1,1}]
\subset\R^4, \quad 
U_2 = [\colvec{-1,1,2,0},\, \colvec{0,1,0,0}]
\subset\R^4\,.
\end{align}
To find a basis of $U_1\cap U_2$, we need to find all $\vec x \in V$ that can be represented as linear combinations of the basis vectors of $U_1$ and $U_2$, i.e., 
\begin{align}
\sum_{i=1}^3 \lambda_i \vec b_i = \vec x = \sum_{j=1}^2 \psi_j \vec c_j\,,
\end{align}
where $\vec b_i$ and $\vec c_j$ are the basis vectors of $U_1$ and $U_2$, respectively.
%
The matrix $\mat A = [\vec b_1|\vec b_2|\vec b_3| -\vec c_1|-\vec
c_2]$ is given as
\begin{align}
\mat A = 
\begin{bmatrix}
1 & 0 & 0 & 1 & 0\\
1 & 1 & 0 & -1 & -1\\
0 & 1 & 1 & -2 & 0\\
0 & 0 & 1 & 0 & 0
\end{bmatrix}\,.
\end{align}
By using Gaussian elimination, we determine the corresponding reduced row echelon form 
\begin{align}
\begin{bmatrix}
1 & 0 & 0 & 1& 0\\
0 & 1 & 0 & -2 & 0\\
0 & 0 & 1 & 0 & 0\\
0 & 0 & 0 & 0 & 1
\end{bmatrix}
\,.
\end{align}
We keep in mind that we are interested in finding $\lambda_1,\lambda_2,\lambda_3\in\R$ and/or $\psi_1,\psi_2\in\R$ with 
\begin{align}
\begin{bmatrix}
1 & 0 & 0 & 1& 0\\
0 & 1 & 0 & -2 & 0\\
0 & 0 & 1 & 0 & 0\\
0 & 0 & 0 & 0 & 1
\end{bmatrix}
\colvec{\lambda_1, \lambda_2, \lambda_3, \psi_1, \psi_2}
=\vec 0\,.
\end{align}
From here, we can immediately see that $\psi_2=0$ and $\psi_1\in\R$ is
a free variable since it corresponds to a non-pivot column, and our solution is 
\begin{align}
U_1\cap U_2 = \psi_1\vec c_1 =  [ \colvec{-1,1,2,0} ]
\,, \quad \psi_1\in\R\,.
\end{align}
\end{itemize}

\subsection{Gaussian elimination}
We provide a template for Gaussian elimination. It is not perfect, but it may be useful:

\begin{elimination}[6]{5}{8mm}{1}
    \eliminationstep
    {
1 & - 2 & 1 & -1 & 1 &  0\\     
0 & 0 & -1 & 1 & -3 & 2\\
0 & 0 & 0 & -3 & 6 & -3\\
0 & 0 & -1 & -2 & 3 & a
    }
    {
\\
\\
\\
-R_2
    }
    \\
\eliminationstep
 {
1 & - 2 & 1 & -1 & 1 &  0\\     
0 & 0 & -1 & 1 & -3 & 2\\
0 & 0 & 0 & -3 & 6 & -3\\
0 & 0 & 0 & -3 & 6 & a-2
    }
    {
\\
\\
\\
-R_3
    }\\
\eliminationstep{
1 & - 2 & 1 & -1 & 1 &  0\\     
0 & 0 & -1 & 1 & -3 & 2\\
0 & 0 & 0 & -3 & 6 & -3\\
0 & 0 & 0 & 0 & 0 & a+1
}
{
\\
\cdot (-1)\\
\cdot (-\tfrac{1}{3})\\
\\}
\\
\eliminationstep{
1 & - 2 & 1 & -1 & 1 &  0\\     
0 & 0 & 1 & -1 & 3 & -2\\
0 & 0 & 0 & 1 & -2 & 1\\
0 & 0 & 0 & 0 & 0 & a+1
}{}
\end{elimination}

The arguments of this environment are:
\begin{enumerate}
\item Number of columns (in the augmented matrix)
\item Number of free variables (equals the number of columns after which the vertical line is drawn)
\item Column width
\item Stretch factor, which can stretch the rows further apart.
\end{enumerate}

\newpage
\section{Answer Template}
\begin{enumerate}[1)]
\item Discrete models

\begin{enumerate}[a)]
\addtocounter{enumii}{2} % change to enumi if you use sections rather than enumerate for question numbers
\item 
\item
\item 
\end{enumerate}

\item Differentiation

\begin{enumerate}[a)]
\item 
\item
\addtocounter{enumii}{1} 
\item 
\item
\end{enumerate}

\item Continuous Models

\begin{enumerate}[a)]
\item 
\item
\item 
\item 
\item
\item 
\item 
\end{enumerate}

\item Linear Regression

\begin{enumerate}[a)]
\item 
\item
\item 
\item 
\end{enumerate}

\item Ridge Regression

\begin{enumerate}[a)]
\item 
\item
\item

\begin{enumerate}[i)]
\item 
\item
\end{enumerate}

\end{enumerate}

\item Bayesian Linear Regression

\begin{enumerate}[a)]
\addtocounter{enumii}{1} 
\item 
\item
\item 
\item (bonus)
\end{enumerate}

\end{enumerate}

### 2019-12-20 interim/sections/background/codegen.tex
Start setting up interim scaffolding

\section{Code Generation}

\subsection{Overview}
\begin{itemize}
\item Motivate the practical implementation of session types in programming
\item What needs to be verified - I/O behaviour (introduce the term behaviour typing) and linear channel usage
\end{itemize}

\subsection{Static Session Typing}
\begin{itemize}
\item PureScript implementation that addresses the 2 criteria statically
\item \begin{itemize}
\item Relevance for web development
\item Static behavioural typing
\item Static channel linearity by construction
\item Pros - type dependencies (expressing dependent types)
\item Cons
\end{itemize}
\item Other initiatives -- creating language with first-class channel primitives (e.g. SILL)
\end{itemize}

PureScript - \cite{PS2019}

\subsection{Runtime Monitors}

\subsection{Hybrid Session Verification}
\begin{itemize}
\item How the aforementioned 2 approaches motivated hybrid session verification
\item Workflow using Scribble toolchain 
\item Static behavioural typing, dynamic channel usage runtime checks
\item Pros - Abstract I/O interfaces
\item Pros - Input futures for `complex communication patterns' motivated by SMTP example
\item Cons - ``practical compromise'' as outlined in paper
\end{itemize}

Java - \cite{HY2016}

\subsection{Comparison}
\begin{itemize}
\item Comparison table between the 3 approaches for how they deal with behavioural typing 
\end{itemize}

### 2019-12-20 interim/sections/background/scribble.tex
Start setting up interim scaffolding

\section{The Scribble Language}

\subsection{Overview}
\begin{itemize}
\item Introduce Scribble language as implementation of MPST
\end{itemize}

\subsection{Protocol Specification}
\begin{itemize}
\item Show parallels between Scribble protocol and MPST global type
\end{itemize}

\subsection{Endpoint Finite State Machines}
\begin{itemize}
\item Show parallels between Scribble protocol-to-EFSM and MPST global-to-local projections
\item Formalise syntax and properties of EFSM derived from well-formed protocols
\end{itemize}

\begin{figure}[!hb]
\doublespacing
\[
\begin{array}{rlr}

\text{EFSM} ::= & \mathbb{R} \times \mathbb{L} \times \mathbb{T} \times \Sigma \times \mathbb{S} \times \delta & \text{Endpoint FSM} \\

\mathbb{R} ::= & r,~r',~\dots & \text{Role Identifiers} \\

\mathbb{L} ::= & l,~l',~\dots & \text{Message Label Identifiers} \\

\mathbb{T} ::= & \tt{int},~\tt{bool},~T,~T',~\dots & \text{Payload Format Types} \\

\Sigma ::= & & \text{Actions} \\
     & r!l(\tilde{T}) \quad \text{where } \tilde{T} \subseteq \mathbb{T} & \text{Output} \\
\mid & r?l(\tilde{T}) \quad \text{where } \tilde{T} \subseteq \mathbb{T} & \text{Input} \\

\mathbb{S} ::= & S,~S',\dots & \text{State Identifiers} \\

\delta ::= & \mathbb{S} \times \Sigma \rightharpoonup \mathbb{S} & \text{State Transition Function} \\

\end{array}
\]
\singlespacing
\caption{Syntax for Endpoint FSM}
\end{figure}

\begin{figure}[!hb]
\doublespacing
\[
\begin{array}{rl}

\texttt{initial}_{\mathbb{R}, \mathbb{L}, \mathbb{T}, \Sigma, \mathbb{S}, \delta}(S) \iff & \nexists S' \in \mathbb{S},\alpha \in \Sigma.~\delta(S',\alpha) = S \\
\texttt{terminal}_{\mathbb{R}, \mathbb{L}, \mathbb{T}, \Sigma, \mathbb{S}, \delta}(S) \iff & \delta(S) = \emptyset \\
\texttt{output}_{\mathbb{R}, \mathbb{L}, \mathbb{T}, \Sigma, \mathbb{S}, \delta}(S) \iff & \delta(S) = \{\,\alpha \in \Sigma \mid \exists r \in \mathbb{R}, l \in \mathbb{L}, \tilde{T} \subseteq{\mathbb{T}}.~\alpha = r!l(\tilde{T})\,\} \\
\texttt{input}_{\mathbb{R}, \mathbb{L}, \mathbb{T}, \Sigma, \mathbb{S}, \delta}(S) \iff & \delta(S) = \{\,\alpha \in \Sigma \mid \exists r \in \mathbb{R}, l \in \mathbb{L}, \tilde{T} \subseteq{\mathbb{T}}.~\alpha = r?l(\tilde{T})\,\} \\
\end{array}
\]
\singlespacing
\caption{Types of EFSM States}
\end{figure}

### 2019-12-20 interim/sections/background/session_types.tex
Start setting up interim scaffolding

\section{Session Types}

\subsection{Overview}
\begin{itemize}
\item Motivate different ways of modelling concurrency (shared memory, message passing), narrow focus to message passing
\item Outline relevance of process algebra and session types as the relevant typing discipline
\item Outline existing implementations of session types in practice
\end{itemize}

\subsection{Process Calculus}
\begin{itemize}
\item Introduce asynchronous pi calculus
\item Motivate how this can model interaction between concurrent processes through reduction rule in operational semantics
\end{itemize}

\subsection{Binary Session Types}
\begin{itemize}
\item Introduce synchronous pi calculus with branching and selection
\item Motivate with travel agency example
\item Introduce binary session types, notion of duality and subtyping for well-typed binary proesses
\item Guarantees for well-typed sessions?
\end{itemize}

\subsection{Multiparty session types}
\begin{itemize}
\item Motivate the extension from BST to MPST (example of battleship game, existing distributed protocols that involve multiple participants?)
\item Introduce global type, projection
\item How does the notion of compatibility (which used to be duality) adapt to MPST?
\end{itemize}

### 2019-12-26 interim/sections/background/session_types.tex
Update .bib for papers read
Interim write-up: session type overview

@@
 \subsection{Overview}
-\begin{itemize}
-\item Motivate different ways of modelling concurrency (shared memory, message passing), narrow focus to message passing
-\item Outline relevance of process algebra and session types as the relevant typing discipline
-\item Outline existing implementations of session types in practice
-\end{itemize}
+% Motivate different ways of modelling concurrency, narrow the focus to message passing
+Web applications are one of many examples of distributed systems in practice.
+Distributed systems are built upon the interaction between concurrent processes, which can be implemented using the two main communication abstractions in \term{shared memory} and \term{message passing}.
 
-\subsection{Process Calculus}
+Shared memory provides processes with the impression of a logical single large monolithic memory but requires programmers to understand consistency models in order to correctly reason about the consistency of shared state.
+
+Message passing interprets the interaction between processes as the exchange of messages, and best describes the communication transports found in web applications, ranging from the stateless request-response client-server interactions via HTTP to full-duplex communication channels via the WebSocket protocol \cite{WebSocketRFC}.
+
+% Outline relevance of process algebra and session types as the relevant typing discipline
+The process algebra $\pi$-calculus introduced by Milner \cite{Milner1989} provides a formalism of the message passing abstraction in terms of the basic building blocks of sending and receiving processes, along with inductively defined continuation processes.
+The composition of these primitives allow us to describe more complex communication sessions.
+Session types define the typing discipline for the $\pi$-calculus and provide reliability guarantees for communication sessions; the latter addresses a key challenge when reasoning about the correctness of distributed systems.
+
+% Outline existing implementations of session types in practice
+Many studies are done on the practical applications of session types, from developing languages providing native session type support \cite{ATS2016} to implementing session types in existing programming languages across different paradigms.
+Implementations of the latter approach differ by how they leverage the design philosophy and features provided by the programming language.
+For example, King et al.
+\cite{PureScript2019} leveraged the expressive type system of PureScript to perform static session type checking, whilst Neykova and Yoshida \cite{Python2017} opted for dynamic approaches to check the conformance of Python programs with respect to session types.
+
+\subsection{Asynchronous $\pi$-calculus}
 \begin{itemize}
@@
 
-\subsection{Multiparty session types}
+\subsection{Multiparty Session Types}
 \begin{itemize}
@@
 \end{itemize}
+
+Gentle introduction to MPST - \cite{MPST}

### 2019-12-26 interim/sections/background/session_types.tex
Interim write-up: session type pi calculus

@@
 % Outline relevance of process algebra and session types as the relevant typing discipline
-The process algebra $\pi$-calculus introduced by Milner \cite{Milner1989} provides a formalism of the message passing abstraction in terms of the basic building blocks of sending and receiving processes, along with inductively defined continuation processes.
+The process algebra $\pi$-calculus introduced by Milner in \cite{Milner1999} provides a formalism of the message passing abstraction in terms of the basic building blocks of sending and receiving processes, along with inductively defined continuation processes.
 The composition of these primitives allow us to describe more complex communication sessions.
@@
 For example, King et al.
-\cite{PureScript2019} leveraged the expressive type system of PureScript to perform static session type checking, whilst Neykova and Yoshida \cite{Python2017} opted for dynamic approaches to check the conformance of Python programs with respect to session types.
+leveraged the expressive type system of PureScript to perform static session type checking in \cite{PureScript2019}, whilst Neykova and Yoshida introduced dynamic approaches to check the conformance of Python programs with respect to session types in \cite{Python2017}.
 
-\subsection{Asynchronous $\pi$-calculus}
+\subsection{Asynchronous $\pi$-calculus}\label{section_async}
+
+% Introduce pi-calculus, monadic asynchronous pi-calculus.
+The $\pi$-calculus models concurrent computation, where processes can execute in parallel and communicate via shared names.
+We first consider the asynchronous $\pi$-calculus introduced by Honda and Tokoro in \cite{AsyncHonda}.
+Among the many flavours of the calculus which vary depending on the application domain, we outline the variant as presented in \cite{C406Lecture}.
+
+Figure \ref{fig:async} defines the syntax of processes in asynchronous $\pi$-calculus; the asynchrony comes from the lack of continuation in the output process.
+
 \begin{itemize}
-\item Introduce asynchronous pi calculus
-\item Motivate how this can model interaction between concurrent processes through reduction rule in operational semantics
+\item $\mathbf{0}$ is the nil process and represents inactivity.
+\item $\piout{u}{v}$ is the output process that will send value $v$ on $u$.
+\item $\piin{u}{x}.P$ is the input process that, upon receiving a message on $u$, will bind the message to $x$ and carry on executing $P$ under this binding.
+\item $P\mid Q$ represents the parallel composition of processes executing simultaneously.
+\item $!P$ represents the parallel composition of infinite instances of $P$; more specifically, $!P \equiv P \mid {!P}$.
+\item $(\nu a)~P$ represents a name restriction where any occurrence of $a$ in $P$ is local and will not interfere with other names outside the scope of $P$.
 \end{itemize}
 
+\begin{figure}[!hb]
+\doublespacing
+\[
+\begin{array}{rlr}
+
+P,Q ::= & & \text{Processes} \\
+     & \mathbf{0} & \text{Nil Process} \\
+\mid & \piout{u}{v} & \text{Output} \\
+\mid & \piin{u}{x}.P & \text{Input} \\     
+\mid & P \mid Q & \text{Parallel Composition} \\
+\mid & !P & \text{Replication} \\
+\mid & (\nu a)~P & \text{Restriction} \\
+
+u,v ::= & & \text{Identifiers} \\
+     & a, b, c,~\dots & \text{Names} \\
+\mid & x, y, z,~\dots & \text{Variables} \\
+
+\end{array}
+\]
+\singlespacing
+\caption{Syntax of Asynchronous $\pi$-calculus}
+\label{fig:async}
+\end{figure}
+
+% Motivate how this models interactions between concurrent processes through the reduction rule in operational semantics.
+The operational semantics model the interaction between parallel processes.
+Whilst \cite{C406Lecture} presents the full operational semantics, we highlight the \rulename{Comm} reduction rule which specifically models message passing:
+if the parallel composition of an input process and output process share the same name, the composition reduces to the continuation of the input process, substituting the variable $x$ with the message received $v$.
+We omit the definitions of substitution, free variables and free names, $\alpha$-equivalence and structural congruence; the interested reader may refer to \cite{C406Lecture}.
+
+$$
+\infer[\mbox{\rulename{Comm}}]{\piout{a}{v} \mid \piin{a}{x}.P ~\longrightarrow ~ P[v/x] }{}
+$$
+
+We additionally define a process $P$ to be \term{stuck} if $P$ is not the nil process and $P$ cannot be reduced any further.
+For example, the process $P = \piin{a}{x}.\mathbf{0} \mid \piout{b}{v}$ is stuck as the parallel composition of an input process and an output process that do not share the same name cannot be reduced using \rulename{Comm}.
+In practice, a stuck process contains communications that will never be executed.
+
 \subsection{Binary Session Types}
+% Introduce synchronous pi calculus with branching and selection
+A \term{binary session} is a parallel composition of two processes, each representing a distinct participant.
+In the context of web applications, a binary session would describe the interactions between client and server.
+Without loss of generality, a \term{session} represents the sequence of send and receive actions of a single participant.
+
+We introduce a \term{synchronous} session calculus built upon \mathref{\ref{section_async}} motivated by \cite{C406Lecture}.
+Figure \ref{fig:sync} defines the syntax formally; we briefly discuss the main additions below:
+
 \begin{itemize}
-\item Introduce synchronous pi calculus with branching and selection
-\item Motivate with travel agency example
-\item Introduce binary session types, notion of duality and subtyping for well-typed binary proesses
-\item Guarantees for well-typed sessions?
+\item \dots
 \end{itemize}
+
+\begin{figure}[!hb]
+\doublespacing
+\[
+\begin{array}{rlr}
+
+e, e' ::= & & \text{Expressions} \\
+	 & v & \text{Values} \\
+\mid	 & x & \text{Variables} \\
+\mid & e + e'~\mid~e - e' & \text{Arithmetic Operators} \\
+\mid & e = e'~\mid~e < e' ~\mid~e > e' & \text{Relational Operators} \\
+\mid & e \wedge e'~\mid~e \vee e' ~\mid~\neg e & \text{Logical Operators} \\
+
+\parti{p} ::= & \parti{Client},~\parti{Server}& \text{Participant} \\
+
+P,Q ::= & & \text{Processes} \\
+     & \mathbf{0} & \text{Nil Process} \\
+\mid & \send{\parti{p}}{e}.P & \text{Send} \\
+\mid & \recv{\parti{p}}{x}.P & \text{Receive} \\
+\mid & \bra{\parti{p}}{l_i:P_i}_{i\in I} & \text{Branching} \\
+\mid & \sel{\parti{p}}{l_i}.~P & \text{Selection} \\
+\mid & {\tt{if}}~e~{\tt{then}}~P~{\tt{else}}~Q & \text{Conditional} \\
+\mid & \mu X.~P & \text{Recursive Process} \\
+\mid & X & \text{Process Variable} \\
+\mathcal{M} ::= & \parti{p} :: P~\mid~\parti{q} :: Q & \text{Binary Session} \\
+\end{array}
+\]
+
+\singlespacing
+\caption{Syntax of Session Calculus with Branching, Selection and Recursion}
+\label{fig:sync}
+\end{figure}
+
+$$
+\infer[\mbox{\rulename{Comm}}]{
+	\parti{p} :: \piout{\parti{q}}{e}.P \mid 
+	\parti{q} ::\piin{\parti{p}}{x}.Q ~\longrightarrow~ \parti{p} :: P \mid \parti{q} ::Q[v/x] }{
+	\begin{array}{ccc}
+	e \downarrow v 
+	&
+	& \parti{p} \neq \parti{q}
+	\end{array}
+}
+$$
+
+$$
+\infer[\mbox{\rulename{Label}}]{
+	\parti{p} :: \sel{\parti{q}}{l}.~P \mid 
+	\parti{q} :: \bra{\parti{p}}{l_i:Q_i}_{i\in I} ~\longrightarrow~ \parti{p} :: P \mid \parti{q} ::Q_j
+}{
+	\begin{array}{ccc}
+	\exists	j \in I. l_j = l
+	&
+	& \parti{p} \neq \parti{q}
+	\end{array}
+}
+$$
+
+% Introduce binary session types, notion of duality and subtyping for well-typed binary processes
+\term{Session types} represent the type theory for our session calculus.
+
+% Guarantees for well-typed sessions
+The definition of stuck processes from \mathref{\ref{section_async}} motivate the discussion of communication errors that may occur during interactions among participants.
+We outline two of the main classes of errors:
+
+\begin{itemize}
+\item \textbf{Deadlock}: 
+\item \textbf{Communication mismatch}: 
+\end{itemize}
+
+The type theory for binary sessions ensure that well-typed binary sessions are guaranteed to be free from the aforementioned communication errors through the concept of \term{duality}.
 

### 2019-12-27 interim/sections/background/session_types.tex
Interim write-up:  binary session types

@@
 
-We introduce a \term{synchronous} session calculus built upon \mathref{\ref{section_async}} motivated by \cite{C406Lecture}.
-Figure \ref{fig:sync} defines the syntax formally; we briefly discuss the main additions below:
+We introduce a \term{synchronous} session calculus inspired by \cite{MPST}.
+Figure \ref{fig:sync} defines the syntax formally; we briefly discuss the main components and how it differs from the variant introduced in \cite{C406Lecture}:
 
+% polyadic synchronous
+% communication must come with label
+% equi-recursive recursion, assume guarded
 \begin{itemize}
-\item \dots
+\item \textbf{Synchronous communication}: Processes that send a message will have a continuation which will be executed upon a successful send. 
+\item \textbf{Polyadic communication}: A vector of values can be communicated at once.
+\item \textbf{Branching and selection}: A branching process can offer a set of branches, each defined by its own label identifier and continuation process.
+A selection process can select a branch by sending the corresponding label identifier alongside the payload to the branching process.
+
+\item \textbf{Labelled messages}: a label identifier is attached to all messages; the input process in \mathref{\ref{section_async}} is generalised as a branching process that offers one branch.
 \end{itemize}
+
+The \rulename{Comm} rule in the operational semantics for this calculus exemplifies these new additions: given a binary session between distinct participants $\parti{p}$ and $\parti{q}$ where $\parti{q}$ offers a set of labelled branches, if $\parti{p}$ selects a label offered by $\parti{q}$ and sends a vector of expressions $e_1, \dots, e_n$ that evaluate\footnote{We adopt the operational semantics for expression evaluation $e \downarrow v$ as defined in \cite{C406Lecture}.} to the corresponding vector of values $v_1, \dots, v_n$, the session reduces to a session with the continuation from the selection process composed with the continuation from the selected branch of the branching process, the latter having the variables $x_1, \dots, x_n$ substituted with the vector of values $v_1, \dots, v_n$ received.
+
+$$
+\infer[\mbox{\rulename{Comm}}]{
+	\parti{p} :: \sel{\parti{q}}{l}{e_1 \dots e_n}.~P \mid 
+	\parti{q} :: \bra{\parti{p}}{l_i(x_1 \dots x_n):Q_i}_{i\in I} ~\longrightarrow~ \parti{p} :: P \mid \parti{q} ::Q_j[v_k/x_k]^n_{k=1}
+}{
+	\begin{array}{cccccc}
+	\exists	j \in I. l_j = l
+	&
+	e_1 \downarrow v_1 
+	& \dots 
+	& e_n \downarrow v_n
+	&
+	& \parti{p} \neq \parti{q}
+	\end{array}
+}
+$$
 
@@
 
+v ::= & \underline{n}~\mid~\texttt{true}~\mid~\texttt{false} & \text{Values} \\
 e, e' ::= & & \text{Expressions} \\
@@
 \mid & e \wedge e'~\mid~e \vee e' ~\mid~\neg e & \text{Logical Operators} \\
+\mid & e \oplus e' & \text{Non-Determinism} \\
 
@@
      & \mathbf{0} & \text{Nil Process} \\
-\mid & \send{\parti{p}}{e}.P & \text{Send} \\
-\mid & \recv{\parti{p}}{x}.P & \text{Receive} \\
-\mid & \bra{\parti{p}}{l_i:P_i}_{i\in I} & \text{Branching} \\
-\mid & \sel{\parti{p}}{l_i}.~P & \text{Selection} \\
-\mid & {\tt{if}}~e~{\tt{then}}~P~{\tt{else}}~Q & \text{Conditional} \\
+\mid & \bra{\parti{p}}{l_i(x_1 \dots x_n):P_i}_{i\in I} & \text{Branching} \\
+\mid & \sel{\parti{p}}{l}{e_1 \dots e_n}.~P & \text{Selection} \\
+\mid & \texttt{if}~e~\texttt{then}~P~\texttt{else}~Q & \text{Conditional} \\
 \mid & \mu X.~P & \text{Recursive Process} \\
 \mid & X & \text{Process Variable} \\
+
+l, l' ::= & {\tt{``str''}} & \text{Label Identifiers} \\
+
 \mathcal{M} ::= & \parti{p} :: P~\mid~\parti{q} :: Q & \text{Binary Session} \\
@@
 
+Additionally, the calculus introduces:
+
+\begin{itemize}
+\item \textbf{Conditionals}: If $e \downarrow \texttt{true}$, the process $\texttt{if}~e~\texttt{then}~P~\texttt{else}~Q$ reduces to $P$; if $e \downarrow \texttt{false}$, the process $\texttt{if}~e~\texttt{then}~P~\texttt{else}~Q$ reduces to $Q$.
+\item \textbf{Recursion}: Following the equirecursive approach, the occurrence of the process variable $X$ in the recursive process can be expanded into the process transparently; more specifically, $\mu X.
+P \equiv P[(\mu X.
+P)/X]$.
+\end{itemize}
+
+% Introduce binary session types
+\term{Session types} represent the type theory for our session calculus. We define the syntax of session types for binary sessions in figure \ref{fig:bst}.
+
+\begin{figure}[!hb]
+\doublespacing
+\[
+\begin{array}{rlr}
+
+U ::= & \text{\tt{int}}~\mid~\text{\tt{bool}} & \text{Sorts} \\
+
+S ::= & & \text{Session Types} \\
+     & \mathbf{end} & \text{Termination} \\
+\mid & \tybra{\parti{p}}{l_i(U_1 \dots U_n):S_i}_{i\in I} & \text{Branching} \\
+\mid & \tysel{\parti{p}}{l_i(U_1 \dots U_n):S_i}_{i\in I} & \text{Selection} \\
+\mid & \mu \mathbf{t}.~S & \text{Recursive Type} \\
+\mid & \mathbf{t} & \text{Type Variable} \\
+\end{array}
+\]
+
+\singlespacing
+\caption{Syntax of Session Types}
+\label{fig:bst}
+\end{figure}
+
+We derive the type of a process with a \term{typing judgement} of the form $\Gamma \vdash P: S$, which reads, \textit{under the typing context $\Gamma$, process $P$ has session type $S$}.
+
+The \term{typing context} records typing assumptions used as part of the derivation: in the case of binary session types, the context maps expressions to sorts, and process variables to session types.
+A typing judgement is constructed in terms of inference rules defined inductively on the structure of processes and expressions.
+
+We present the rules for \rulename{Ty-Sel} and \rulename{Ty-Bra}, the remaining rules follow from \cite{C406Lecture} and can be trivially defined as they leverage the syntactic similarities between session types and our session calculus.
+
 $$
-\infer[\mbox{\rulename{Comm}}]{
-	\parti{p} :: \piout{\parti{q}}{e}.P \mid 
-	\parti{q} ::\piin{\parti{p}}{x}.Q ~\longrightarrow~ \parti{p} :: P \mid \parti{q} ::Q[v/x] }{
+\infer[\mbox{\rulename{Ty-Bra}}]{
+	\Gamma \vdash \bra{\parti{p}}{l_i(x_1 \dots x_n):P_i}_{i\in I} : \tybra{\parti{p}}{l_i(U_1 \dots U_n):S_i}_{i\in I}
+	}{
 	\begin{array}{ccc}
-	e \downarrow v 
+	\forall i \in I.
 	&
-	& \parti{p} \neq \parti{q}
+	&
+    \Gamma, x_1: U_1, \dots, x_n: U_n \vdash P_i : S_i		
 	\end{array}
@@
 $$
-\infer[\mbox{\rulename{Label}}]{
-	\parti{p} :: \sel{\parti{q}}{l}.~P \mid 
-	\parti{q} :: \bra{\parti{p}}{l_i:Q_i}_{i\in I} ~\longrightarrow~ \parti{p} :: P \mid \parti{q} ::Q_j
+\infer[\mbox{\rulename{Ty-Sel}}]{
+	\Gamma \vdash \sel{\parti{p}}{l}{e_1 \dots e_n}.P : \tysel{\parti{p}}{l(U_1 \dots U_n): S}
 	}{
-	\begin{array}{ccc}
-	\exists	j \in I. l_j = l
+	\begin{array}{cccc}
+	\Gamma \vdash e_1 : U_1
 	&
-	& \parti{p} \neq \parti{q}
+	\dots
+	&
+	\Gamma \vdash e_n : U_n
+	&
+    \Gamma \vdash P : S
 	\end{array}
@@
 
-% Introduce binary session types, notion of duality and subtyping for well-typed binary processes
-\term{Session types} represent the type theory for our session calculus.
-
-% Guarantees for well-typed sessions
+% Guarantees for well-typed sessions expressed through duality, also notion of subtyping.
 The definition of stuck processes from \mathref{\ref{section_async}} motivate the discussion of communication errors that may occur during interactions among participants.
@@
 \begin{itemize}
-\item \textbf{Deadlock}: 
-\item \textbf{Communication mismatch}: 
+\item \textbf{Deadlock}: Progress cannot be made when the two participants expect to be receiving a message from each other at the same time.
+\item \textbf{Communication mismatch}: Progress cannot be made when the selection process sends a message with a label identifier not offered by the branching process; likewise, the payload sent must be compatible with the sort expected by the branching process for the selected branch.
 \end{itemize}
 
-The type theory for binary sessions ensure that well-typed binary sessions are guaranteed to be free from the aforementioned communication errors through the concept of \term{duality}.
+Session types ensure that well-typed binary sessions are guaranteed to be free from these communication errors through the concept of \term{duality}.
+Duality defines a notion of \term{compatibility} between processes: two session types are dual with respect to each other if the communication between them (i.e.
+pairs of sending and receiving actions) always match (i.e.
+with respect to the selected label and message payload type).
+We define $\overline{S}$ as the dual type of $S$ in Table \ref{table:dual}.
+
+% Row height in tables
+\renewcommand{\arraystretch}{1.2}
+\begin{center}
+\begin{tabular}{rcl}
+$\overline{\mathbf{end}}$ & = & $\mathbf{end}$ \\
+$\overline{\tybra{\parti{p}}{l_i(U_1 \dots U_n):S_i}_{i\in I}}$ & = & $\tysel{\parti{q}}{l_i(U_1 \dots U_n):\overline{S_i}}_{i\in I}$ \\
+$\overline{\tysel{\parti{p}}{l_i(U_1 \dots U_n):S_i}_{i\in I}}$ & = & $\tybra{\parti{q}}{l_i(U_1 \dots U_n):\overline{S_i}}_{i\in I}$ \\
+$\overline{\mu \mathbf{t}. S}$ & = & $\mu t. \overline{S}$ \\
+$\overline{\mathbf{t}}$ & = & $\mathbf{t}$ \\ 
+\end{tabular}
+\captionof{table}{Duality of Binary Session Types involving participants $\parti{p}$ and $\parti{q}$}
+\label{table:dual}
+\end{center}
+\renewcommand{\arraystretch}{1}
+
+Consequently, a binary session is well-typed if the participating processes are typed to be dual with respect to each other: we illustrate this in \rulename{MTy}.
+
+$$
+\infer[\mbox{\rulename{MTy}}]{
+	\vdash \parti{p} :: P \mid 
+	\parti{q} :: Q
+}{
+	\begin{array}{lr}
+	\cdot \vdash P: S
+	&
+	\cdot \vdash Q: \bar{S}
+	\end{array}
+}
+$$
+
+The definition of duality alone restricts the definition of well-typed binary sessions to those where the two processes are derived to be \textit{exactly} dual types of one another.
+Consider the pair of session types below:
+
+$$
+\begin{array}{rl}
+S_{\text{Client}} &= \tysel{\parti{Server}}{Succ(\texttt{int}): \tybra{\parti{Server}}{Res(\texttt{int}): \mathbf{end}}} \\
+S_{\text{Server}} &= \tybra{\parti{Client}}{Succ(\texttt{int}): \tysel{\parti{Client}}{Res(\texttt{int}): \mathbf{end}},~Quit(): \mathbf{end}}
+\end{array}
+$$
+
+Whilst $\overline{S_{\text{Client}}} \neq S_{\text{Server}}$, this pair of session types is intuitively compatible as the client is selecting a branch offered by the server, where the session types for the continuations of this branch for both participants are indeed dual.
+
+This motivates the concept of \term{subtypes}, which allows a process to be typed by its ``supertype'' when required.
+$\leqslant$ defines the subtyping relation: $S \leqslant S'$ reads \textit{$S$ is a subtype of $S'$}, and is defined coinductively.
+
+We present the inference rules for \rulename{Sub-Bra} and \rulename{Sub-Sel} inspired by \cite{MPST} but adapted for polyadic communication; the intuition behind subtyping and subsorting is outlined below:
+
+\begin{itemize}
+\item \textbf{Branching}: A process can offer more branches 
+\item \textbf{Selection}:
+\end{itemize}
+
+\begin{prooftree}
+\AxiomC{$\forall i \in I.$}
+\AxiomC{$U'_1 \leqslant U_1 \dots U'_n \leqslant U_n$}
+\AxiomC{$S_i \leqslant S'_i$}
+\RightLabel{\mbox{\rulename{Sub-Bra}}}
+\doubleLine
+\TrinaryInfC{$
+	\tybra{\parti{p}}{l_i(U_1 \dots U_n):S_i}_{i\in I \cup J}
+	\leqslant
+	\tybra{\parti{p}}{l_i(U'_1 \dots U'_n):S'_i}_{i\in I}
+$}
+\end{prooftree}
+
+\begin{prooftree}
+\AxiomC{$\forall i \in I.$}
+\AxiomC{$U_1 \leqslant U'_1 \dots U_n \leqslant U'_n$}
+\AxiomC{$S_i \leqslant S'_i$}
+\RightLabel{\mbox{\rulename{Sub-Sel}}}
+\doubleLine
+\TrinaryInfC{$
+	\tysel{\parti{p}}{l_i(U_1 \dots U_n):S_i}_{i\in I}
+	\leqslant
+	\tysel{\parti{p}}{l_i(U'_1 \dots U'_n):S'_i}_{i\in I \cup J}
+$}
+\end{prooftree}
+
+We also introduce the \rulename{Ty-Sub} to incorporate the subtyping relation into the typing judgement.
+
+$$
+\infer[\mbox{\rulename{Ty-Sub}}]{
+	\Gamma \vdash P : S'
+}{
+	\begin{array}{lr}
+	\Gamma \vdash P: S
+	&
+	S \leqslant S'
+	\end{array}
+}
+$$
+
+This allows us to construct a derivation to show that the binary session
+
+\[
+\mathcal{M} = \parti{Client} :: P_\text{Client}~\mid~\parti{Server} :: P_\text{Server}
+\]
+
+is well-typed, assuming the $P_\text{Client}$ and $P_\text{Server}$ are typed $S_\text{Client}$ and $S_\text{Server}$ respectively.
+
+\begin{prooftree}
+\AxiomC{\vdots}
+\UnaryInfC{$\cdot \vdash P_\text{Client}: S_\text{Client}$}
+\AxiomC{\vdots}
+\UnaryInfC{$\cdot \vdash P_\text{Server}: S_\text{Server}$}
+\AxiomC{\vdots}
+\doubleLine
+\UnaryInfC{$S_\text{Server} \leqslant \overline{S_\text{Client}}$}
+
+%\AxiomC{$\cdot \vdash P_\text{Client}: S_\text{Client}$}
+%\AxiomC{$\cdot \vdash P_\text{Server}: S_\text{Server}$}
+%\AxiomC{$S_\text{Server} \leqslant \overline{S_\text{Client}}$}
+\RightLabel{\rulename{Ty-Sub}}
+\BinaryInfC{$\cdot \vdash P_\text{Server}: \overline{S_\text{Client}}$}
+\RightLabel{\rulename{MTy}}
+\BinaryInfC{$
+	\vdash \parti{Client} :: P_\text{Client} \mid 
+	\parti{Server} :: P_\text{Server}
+$}
+\end{prooftree}
 
@@
 Gentle introduction to MPST - \cite{MPST}
+
+\begin{figure}[!hb]
+\doublespacing
+\[
+\begin{array}{rlr}
+
+G ::= & & \text{Global Types} \\
+     & \mathbf{end} & \text{Termination} \\
+\mid & \parti{p} \to \parti{q}:[U]; G & \text{Termination} \\
+\mid & \tysend{\parti{p}}{U};~S & \text{Value Send} \\
+\mid & \tyrecv{\parti{p}}{U};~S & \text{Value Receive} \\
+\mid & \tybra{\parti{p}}{l_i:S_i}_{i\in I} & \text{Branching} \\
+\mid & \tysel{\parti{p}}{l_i:S_i}_{i\in I} & \text{Selection} \\
+\mid & \mu \mathbf{t}.~S & \text{Recursive Type} \\
+\mid & \mathbf{t} & \text{Type Variable} \\
+\end{array}
+\]
+
+\singlespacing
+\caption{Syntax of Global Types}
+\label{fig:mpst}
+\end{figure}
+
+$$
+\infer[\mbox{\rulename{MTy}}]{
+	\vdash \prod_{i \in I} \parti{p_i} :: P_i ~: G
+}{
+	\begin{array}{llll}
+	\forall i \in I. 
+	&
+	\cdot \vdash P_i: G \upharpoonright \parti{p_i}
+	&
+	&
+	\text{pt}(G) \subseteq \left\{\parti{p_i} \mid i \in I \right\}
+	\end{array}
+}
+$$

### 2019-12-28 interim/sections/background/session_types.tex
Interim write-up:  binary session types

@@
 	&
-	\cdot \vdash Q: \bar{S}
+	\cdot \vdash Q: \overline{S}
 	\end{array}
@@
 This motivates the concept of \term{subtypes}, which allows a process to be typed by its ``supertype'' when required.
-$\leqslant$ defines the subtyping relation: $S \leqslant S'$ reads \textit{$S$ is a subtype of $S'$}, and is defined coinductively.
+$\leqslant$\footnote{The $\leqslant$ operator is also an overloaded relation on sorts to express subsorting.} defines the subtyping relation: $S \leqslant S'$ reads \textit{$S$ is a subtype of $S'$}, and is defined coinductively on the structure of session types.
 
@@
 \begin{itemize}
-\item \textbf{Branching}: A process can offer more branches 
-\item \textbf{Selection}:
+\item \textbf{Branching}: The supertype of a branching process offers a subset of the branches and expects more specific types of payload; intuitively, if a process expects to receive an \texttt{int}, it can handle a \texttt{nat} payload.
+\item \textbf{Selection}: The supertype of a selection process offers a superset of the internal choices and can send more generic types of payload; intuitively, if a process sends a \texttt{nat}, the payload is compatible with receivers expecting a more generic \texttt{int} payload.
 \end{itemize}
@@
 
-We also introduce the \rulename{Ty-Sub} to incorporate the subtyping relation into the typing judgement.
+We also introduce subsumption in \rulename{Ty-Sub} to incorporate the subtyping relation into the typing judgement.
 
@@
 
-is well-typed, assuming the $P_\text{Client}$ and $P_\text{Server}$ are typed $S_\text{Client}$ and $S_\text{Server}$ respectively.
+is well-typed, assuming $P_\text{Client}$ and $P_\text{Server}$ are typed $S_\text{Client}$ and $S_\text{Server}$ respectively.
 
@@
 \end{itemize}
-
-Gentle introduction to MPST - \cite{MPST}
 

### 2019-12-28 interim/sections/background/session_types.tex
Interim write-up: multiparty session types

@@
 
-$$
+\[
 \infer[\mbox{\rulename{Comm}}]{\piout{a}{v} \mid \piin{a}{x}.P ~\longrightarrow ~ P[v/x] }{}
-$$
+\]
 
@@
 
-\subsection{Binary Session Types}
+\subsection{Binary Session Types}\label{section_bst}
 % Introduce synchronous pi calculus with branching and selection
@@
 
-$$
+\[
 \infer[\mbox{\rulename{Comm}}]{
@@
 }
-$$
+\]
 
@@
 
-$$
+\[
 \infer[\mbox{\rulename{Ty-Bra}}]{
@@
 }
-$$
+\]
 
-$$
+\[
 \infer[\mbox{\rulename{Ty-Sel}}]{
@@
 
-$$
+\[
 \infer[\mbox{\rulename{MTy}}]{
@@
 
-$$
+\[
 \begin{array}{rl}
-S_{\text{Client}} &= \tysel{\parti{Server}}{Succ(\texttt{int}): \tybra{\parti{Server}}{Res(\texttt{int}): \mathbf{end}}} \\
-S_{\text{Server}} &= \tybra{\parti{Client}}{Succ(\texttt{int}): \tysel{\parti{Client}}{Res(\texttt{int}): \mathbf{end}},~Quit(): \mathbf{end}}
+S_{\text{Client}} &= \tysel{\parti{Server}}{\text{Succ}(\texttt{int}): \tybra{\parti{Server}}{\text{Res}(\texttt{int}): \mathbf{end}}} \\
+S_{\text{Server}} &= \tybra{\parti{Client}}{\text{Succ}(\texttt{int}): \tysel{\parti{Client}}{\text{Res}(\texttt{int}): \mathbf{end}},~\text{Quit}(): \mathbf{end}}
 \end{array}
-$$
+\]
 
@@
 
-$$
+\[
 \infer[\mbox{\rulename{Ty-Sub}}]{
@@
 \subsection{Multiparty Session Types}
-\begin{itemize}
-\item Motivate the extension from BST to MPST (example of battleship game, existing distributed protocols that involve multiple participants?)
-\item Introduce global type, projection
-\item How does the notion of compatibility (which used to be duality) adapt to MPST?
-\end{itemize}
 
+% Motivate the extension from BST to MPST (example of battleship game, existing distributed protocols that involve multiple participants?)
+Whilst binary session types provide communication guarantees between exactly 2 participants, distributed systems generally involve more parties in practice.
+This is equally relevant in interactive web applications, as motivated by the \textit{Battleships} game example in \cite{PureScript2019} where the server coordinates interactions between two players.
+
+Whilst there is a natural syntactical extension to our session calculus for describing multiparty sessions\footnote{We also adopt the shorthand $\mathcal{M} ::= \prod^n_{i=1} \parti{p_i} :: P_i$ as used in the literature.} as
+
+\[
+\begin{array}{rl}
+\mathcal{M} ::= & \parti{p_1} :: P_1~\mid~\parti{p_2} :: P_2~\mid~\dots~\mid \parti{p_n} :: P_n
+\end{array}
+\]
+
+the same cannot be said for the binary session typing discipline, particularly with respect to duality.
+The same notion of duality does not extend to the decomposition of multiparty interactions into multiple binary sessions:  \cite{C406Lecture} and \cite{MPST} both present counterexamples of well-typed binary sessions that, when composed to represent a multiparty session, results in communication errors thus violating guarantees of well-typed sessions.
+
+Honda et al.
+presents \term{multiparty session types} in \cite{MPAST} to extend the binary session typing discipline for sessions involving more than 2 participants, whilst redefining the notion of compatibility in this multiparty context.
+Multiparty session types are defined in terms a \term{global type}, which provides a bird's eye view of the communication protocol describing the interactions between pairs of participants.
+Figure \ref{fig:mpst} defines the syntax of global types inspired by \cite{MPST} and adapted to be compatible with our session calculus presented in \mathref{\ref{section_bst}}.
+
+% Introduce global type, projection
 \begin{figure}[!hb]
@@
      & \mathbf{end} & \text{Termination} \\
-\mid & \parti{p} \to \parti{q}:[U]; G & \text{Termination} \\
-\mid & \tysend{\parti{p}}{U};~S & \text{Value Send} \\
-\mid & \tyrecv{\parti{p}}{U};~S & \text{Value Receive} \\
-\mid & \tybra{\parti{p}}{l_i:S_i}_{i\in I} & \text{Branching} \\
-\mid & \tysel{\parti{p}}{l_i:S_i}_{i\in I} & \text{Selection} \\
-\mid & \mu \mathbf{t}.~S & \text{Recursive Type} \\
+\mid & \parti{p} \to \parti{q}:\left\{l_i(U_1 \dots U_n).~G_i\right\}_{i \in I} & \text{Message Exchange} \\
+\mid & \mu \mathbf{t}.~G & \text{Recursive Type} \\
 \mid & \mathbf{t} & \text{Type Variable} \\
@@
 
-$$
+To check the conformance of a participant's process against the protocol specification described by the global type, we \term{project} the global type onto the participant to obtain a session type that only preserves the interactions described by the global type that pertain to the participant.
+Projection is defined by the $\upharpoonright$ operator, more commonly seen in literature in its infix form as $G \upharpoonright \parti{p}$ describing the projection of global type $G$ for participant $\parti{p}$.
+Intuitively, the projected local type of a participant describes the protocol from the viewpoint of the participant.
+
+More formally, projection can be interpreted as a \textit{partial function} ${\upharpoonright} :: G \times \parti{p} \rightharpoonup S$, as the projection for a participant may be undefined for an ill-formed global type; \cite{C406Lecture} presents examples of where this is the case, and \cite{MPST} presents the formal definition of projection.
+
+% How does the notion of compatibility (which used to be duality) adapt to MPST?
+The notion of compatibility in multiparty session types is still captured by \rulename{MTy}, but adapted to consider the local projections for all participants as supposed to dual types in the binary case.
+
+\[
 \infer[\mbox{\rulename{MTy}}]{
-	\vdash \prod_{i \in I} \parti{p_i} :: P_i ~: G
+	{\displaystyle \vdash \prod^{}_{i \in I} \parti{p_i} :: P_i ~: G} 
 }{
@@
 }
-$$
+\]
+
+For a multiparty session $\mathcal{M} = \prod_{i \in I}\parti{p_i} :: P_i$ to be well-typed by a global type $G$, we require:
+
+\begin{enumerate}
+\item All participant processes $\parti{p_i} :: P_i$ to be well-typed with respect to their corresponding well-defined projection $G \upharpoonright \parti{p_i}$, and 
+\item $G$ does not describe interactions with participants not defined in $\mathcal{M}$ 
+\end{enumerate}
+
+Well-typed multiparty session enjoys the following communication guarantees as outlined in \cite{GentleMPST}:
+
+\begin{itemize}
+\item \textbf{Communication safety}: The types of sent and expected messages will always match.
+\item \textbf{Protocol fidelity}: The exchange of messages between processes will abide by the global protocol.
+\item \textbf{Progress}: Messages sent by a process will be eventually received, and a process waiting for a message will eventually receive one; this also means there will not be any sent but unreceived messages.
+\end{itemize}
+
+This motivates an elegant, decentralised solution for checking protocol conformance in practice: once the global type for the protocol is defined, local processes can verify their implementation against their corresponding projection in isolation, independent of each other.
+Figure \ref{fig:mpst_workflow}\todo{Create figure} illustrates this diagrammatically.
+
+\begin{figure}[!h]
+\caption{Type checking with Multiparty Session Types}
+\label{fig:mpst_workflow}
+\end{figure}

### 2019-12-28 interim/sections/background/scribble.tex
Update .bib for papers read
Interim write-up:  binary session types
Interim write-up:  start on Scribble

@@
 \subsection{Overview}
-\begin{itemize}
-\item Introduce Scribble language as implementation of MPST
-\end{itemize}
+% Introduce Scribble as an implementation of MPST
 
-\subsection{Protocol Specification}
-\begin{itemize}
-\item Show parallels between Scribble protocol and MPST global type
-\end{itemize}
+% Protocol specification: show parallels between Scribble protocol and MPST global type
+
+\begin{lstlisting}[language=Java,numbers=left,frame=single]
+// TODO: Adder Protocol
+\end{lstlisting}
+
+\[
+\begin{array}{rl}
+G = \mu\mathbf{t}.\parti{Client}\to\parti{Server}:\{
+& \text{Add}(\texttt{int}, \texttt{int}): \parti{Server}\to\parti{Client}: \{\text{Res}(\texttt{int}): \mathbf{t}\}, \\
+& \text{Quit}(): \parti{Server}\to\parti{Client}: \{\text{Terminate}(): \mathbf{end}\}\\
+\} &
+\end{array}
+\]
 
@@
 
-\mathbb{T} ::= & \tt{int},~\tt{bool},~T,~T',~\dots & \text{Payload Format Types} \\
+\mathbb{T} ::= & \texttt{int},~\texttt{bool},~T,~T',~\dots & \text{Payload Format Types} \\
 

### 2019-12-31 interim/sections/background/scribble.tex
Interim write-up: Scribble overview

@@
-\section{The Scribble Language}
+\section{The Scribble Protocol Language}
 
@@
 % Introduce Scribble as an implementation of MPST
+Whilst session type theory represents the type language for concurrent processes, it also forms the theoretical basis of proposals introduced to implement session types for real-world application development: the Scribble language is one such proposal.
+
+Scribble \cite{Scribble} is a platform-independent description language for the specification of message-passing protocols.
+The language describes the behaviour of communicating processes at a high level of abstraction: more importantly, the description is independent from implementation details in the same way that the type signature of a function declaration is decoupled from the corresponding function definition.
 
 % Protocol specification: show parallels between Scribble protocol and MPST global type
+A Scribble protocol specification describes an agreement of how participating systems, referred to as \term{roles}, interact.
+The protocol stipulates the sequence of structured messages exchanged between roles; each message is labelled with a name and the type of payload carried by the message.
 
-\begin{lstlisting}[language=Java,numbers=left,frame=single]
-// TODO: Adder Protocol
+We present an example of a Scribble protocol in Figure \ref{fig:adder_scr} adapted from \cite{Hybrid2016}.
+The protocol specifies an arithmetic web service offered by a server to a client, represented by roles $S$ and $C$ respectively.
+The client is permitted to either:
+
+\begin{itemize}
+\item Send two \texttt{int}s attached to an \texttt{Add} message, where the server will respond with an \texttt{int} in a message labelled \texttt{Res}, and the protocol recurses; or,
+\item Send a \texttt{Quit} message, where the server will respond with a \texttt{Terminate} message and the protocol ends.
+\end{itemize}
+
+The platform-independent nature of Scribble can be observed from the \texttt{type} declaration on Line 1: the developer has the freedom to specify message payload formats and data types from the target language of the implementation - in this case, aliasing the built-in Java integer as \texttt{int} throughout the protocol.
+
+\begin{figure}[!h]
+\begin{lstlisting}
+type <java> "java.lang.Integer" from "rt.jar" as int;
+
+global protocol Adder(role C, role S) {
+	choice at C {
+		Add(int, int)	from C to S;
+		Res(int)		from S to C;
+		do Adder(C, S);
+	} or {
+		Quit()		from C to S;
+		Terminate()	from S to C;	
+	}
+}
 \end{lstlisting}
+\caption{Adder Protocol in Scribble}
+\label{fig:adder_scr}
+\end{figure}
+
+The simplicity of the protocol specification language reflects the design goals for Scribble, as outlined in \cite{Scribble}, to be easy to read, write and maintain, even for developers who are not accustomed to the formalities in protocol specification.
+Moreover, we clearly observe the parallels between the Scribble language and multiparty session type theory (MPST), from the homomorphic mapping between Scribble roles and MPST participants to the syntactic similarities between the specification in Figure \ref{fig:adder_scr} and the global type below written in the calculus.
 
@@
 \begin{array}{rl}
-G = \mu\mathbf{t}.\parti{Client}\to\parti{Server}:\{
-& \text{Add}(\texttt{int}, \texttt{int}): \parti{Server}\to\parti{Client}: \{\text{Res}(\texttt{int}): \mathbf{t}\}, \\
-& \text{Quit}(): \parti{Server}\to\parti{Client}: \{\text{Terminate}(): \mathbf{end}\}\\
+G = \mu\mathbf{t}.\parti{C}\to\parti{S}:\{
+& \text{Add}(\texttt{int}, \texttt{int}): \parti{S}\to\parti{C}: \{\text{Res}(\texttt{int}): \mathbf{t}\}, \\
+& \text{Quit}(): \parti{S}\to\parti{C}: \{\text{Terminate}(): \mathbf{end}\}\\
 \} &

### 2019-12-31 interim/sections/background/scribble.tex
Interim write-up: begin Scribble EFSM

@@
 The simplicity of the protocol specification language reflects the design goals for Scribble, as outlined in \cite{Scribble}, to be easy to read, write and maintain, even for developers who are not accustomed to the formalities in protocol specification.
-Moreover, we clearly observe the parallels between the Scribble language and multiparty session type theory (MPST), from the homomorphic mapping between Scribble roles and MPST participants to the syntactic similarities between the specification in Figure \ref{fig:adder_scr} and the global type below written in the calculus.
+Moreover, we clearly observe the parallels between the Scribble language and multiparty session type (MPST) theory, from the homomorphic mapping between Scribble roles and MPST participants to the syntactic similarities between the specification in Figure \ref{fig:adder_scr} and the global type below written in the calculus.
 
@@
 \subsection{Endpoint Finite State Machines}
-\begin{itemize}
-\item Show parallels between Scribble protocol-to-EFSM and MPST global-to-local projections
-\item Formalise syntax and properties of EFSM derived from well-formed protocols
-\end{itemize}
+% Show parallels between Scribble protocol-to-EFSM and MPST global-to-local projections
+The protocol specification language is a component of the broader Scribble project in \cite{Scribble}; through the project, Honda et al.
+also aims to facilitate the development of endpoint applications that conform to user-specified protocols.
+
+A Scribble global protocol can be projected to a role to obtain the local specification of said protocol from the role's viewpoint.
+This is analogous to the standard algorithmic projections in multiparty session type theory: the projected local specification, or local protocol, only preserves the interactions in which the target role is involved.
+Any user-defined type declarations in the global protocol will also be preserved.
+This allows local roles, also referred to as \term{endpoints}, to verify their implementation against their local protocol for conformance, independent of other endpoints.
+The communication safety guarantees from MPST theory also apply here: if the implementation for each endpoint is verified against its local protocol, the multiparty system as a whole will conform to the global protocol.
+
+% Formalise syntax and properties of EFSM derived from well-formed protocols
+To facilitate the verification process, a local protocol is converted into 
+an \term{endpoint finite state machine} (EFSM).
+An EFSM encodes the control flow of the local protocol, where an initial state is defined and each transition from some state to a successor state corresponds to a valid IO action (i.e.
+sending to or receiving from another role) permitted at the endpoint at that state.
 

### 2020-01-14 PLACES_2020/paper.tex
PLACES: load EPTCS template

\documentclass[submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{PLACES 2020} % Name of the event you are submitting to
%\usepackage{breakurl}             % Not needed if you use pdflatex only.

\title{Multiparty Session Type-safe Web Development in TypeScript}
\author{Rob van Glabbeek
\institute{NICTA\\ Sydney, Australia}
\email{rvg@cs.stanford.edu}
\and
Co Author
\institute{Stanford Univeristy\\
California, USA}
\email{is@gmail.com}
\and
Co Author
\institute{Stanford Univeristy\\
California, USA}
\email{is@gmail.com}
\and
Co Author
\institute{Stanford Univeristy\\
California, USA}
\email{is@gmail.com}
}
\def\titlerunning{A Longtitled Paper}
\def\authorrunning{R.J. van Glabbeek, C. Author \& Y.S. Else}
\begin{document}
\maketitle

\begin{abstract}
This is a sentence in the abstract.
This is another sentence in the abstract.
This is yet another sentence in the abstract.
This is the final sentence in the abstract.
\end{abstract}

\section{Introduction}

The optional arguments of {\tt $\backslash$documentclass$\{$eptcs$\}$} are
\begin{itemize}
\item at most one of
{\tt adraft},
{\tt submission} or
{\tt preliminary},
\item at most one of {\tt publicdomain} or {\tt copyright},
\item and optionally {\tt creativecommons},
  \begin{itemize}
  \item possibly augmented with
    \begin{itemize}
    \item {\tt noderivs}
    \item or {\tt sharealike},
    \end{itemize}
  \item and possibly augmented with {\tt noncommercial}.
  \end{itemize}
\end{itemize}
We use {\tt adraft} rather than {\tt draft} so as not to confuse hyperref.
The style-file option {\tt submission} is for papers that are
submitted to {\tt $\backslash$event}, where the value of the latter is
to be filled in in line 2 of the tex-file. Use {\tt preliminary} only
for papers that are accepted but not yet published. The final version
of your paper that is to be uploaded at the EPTCS website should have
none of these style-file options.

By means of the style-file option
\href{http://creativecommons.org/about/license/}{creativecommons}
authors equip their paper with a Creative Commons license that allows
everyone to copy, distribute, display, and perform their copyrighted
work and derivative works based upon it, but only if they give credit
the way you request. By invoking the additional style-file option {\tt
noderivs} you let others copy, distribute, display, and perform only
verbatim copies of your work, but not derivative works based upon
it. Alternatively, the {\tt sharealike} option allows others to
distribute derivative works only under a license identical to the
license that governs your work. Finally, you can invoke the option
{\tt noncommercial} that let others copy, distribute, display, and
perform your work and derivative works based upon it for
noncommercial purposes only.

Authors' (multiple) affiliations and emails use the commands
{\tt $\backslash$institute} and {\tt $\backslash$email}.
Both are optional.
Authors should moreover supply
{\tt $\backslash$titlerunning} and {\tt $\backslash$authorrunning},
and in case the copyrightholders are not the authors also
{\tt $\backslash$copyrightholders}.
As illustrated above, heuristic solutions may be called for to share
affiliations. Authors may apply their own creativity here.

Exactly 46 lines fit on a page.
The rest is like any normal {\LaTeX} article.
We will spare you the details.
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\hfill6\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\hfill11\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.

Here starts a new paragraph. The rest is like any normal {\LaTeX} article.
We will spare you the details.
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\hfill16\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\hfill21\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\hfill26\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\hfill31\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\hfill36\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\hfill41\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.\hfill46\\
The rest is like any normal {\LaTeX} article.
We will spare you the details.
The rest is like any normal {\LaTeX} article.
We will spare you the details.
The rest is like any normal {\LaTeX} article.
We will spare you the details.
The rest is like any normal {\LaTeX} article.
We will spare you the details.
The rest is like any normal {\LaTeX} article.
We will spare you the details.
The rest is like any normal {\LaTeX} article.
We will spare you the details.
The rest is like any normal {\LaTeX} article.
We will spare you the details.
The rest is like any normal {\LaTeX} article.
We will spare you the details.
The rest is like any normal {\LaTeX} article.
We will spare you the details.
The rest is like any normal {\LaTeX} article.
We will spare you the details.
The rest is like any normal {\LaTeX} article.
We will spare you the details.
The rest is like any normal {\LaTeX} article.
We will spare you the details.
The rest is like any normal {\LaTeX} article.
We will spare you the details.
The rest is like any normal {\LaTeX} article.
We will spare you the details.
The rest is like any normal {\LaTeX} article.
We will spare you the details.
The rest is like any normal {\LaTeX} article.
We will spare you the details.
The rest is like any normal {\LaTeX} article.
We will spare you the details.

\section{Prefaces}

Volume editors may create prefaces using this very template,
with {\tt $\backslash$title$\{$Preface$\}$} and {\tt $\backslash$author$\{\}$}.

\section{Bibliography}

We request that you use
\href{http://www.cse.unsw.edu.au/~rvg/EPTCS/eptcs.bst}
{\tt $\backslash$bibliographystyle$\{$eptcs$\}$}
\cite{bibliographystylewebpage}. Compared to the original {\LaTeX}
{\tt $\backslash$biblio\-graphystyle$\{$plain$\}$},
it ignores the field {\tt month}, and uses the extra
bibtex fields {\tt eid}, {\tt doi}, {\tt ee} and {\tt url}.
The first is for electronic identifiers (typically the number $n$
indicating the $n^{\rm th}$ paper in an issue) of papers in electronic
journals that do not use page numbers. The other three are to refer,
with life links, to electronic incarnations of the paper.

Almost all publishers use digital object identifiers (DOIs) as a
persistent way to locate electronic publications. Prefixing the DOI of
any paper with {\tt http://dx.doi.org/} yields a URI that resolves to the
current location (URL) of the response page\footnote{Nowadays, papers
  that are published electronically tend
  to have a \emph{response page} that lists the title, authors and
  abstract of the paper, and links to the actual manifestations of
  the paper (e.g.\ as {\tt dvi}- or {\tt pdf}-file). Sometimes
  publishers charge money to access the paper itself, but the response
  page is always freely available.}
of that paper. When the location of the response page changes (for
instance through a merge of publishers), the DOI of the paper remains
the same and (through an update by the publisher) the corresponding
URI will then resolve to the new location. For that reason a reference
ought to contain the DOI of a paper, with a life link to corresponding
URI, rather than a direct reference or link to the current URL of
publisher's response page. This is the r\^ole of the bibtex field {\tt doi}.
DOIs of papers can often be found through
\url{http://www.crossref.org/guestquery};\footnote{For papers that will appear
  in EPTCS and use \href{http://www.cse.unsw.edu.au/~rvg/EPTCS/eptcs.bst}
  {\tt $\backslash$bibliographystyle$\{$eptcs$\}$} there is no need to
  find DOIs on this website, as EPTCS will look them up for you
  automatically upon submission of a first version of your paper;
  these DOIs can then be incorporated in the final version, together
  with the remaining DOIs that need to found at DBLP or publisher's webpages.}
the second method {\it Search on article title}, only using the {\bf
surname} of the first-listed author, works best.  
Other places to find DOIs are DBLP and the response pages for cited
papers (maintained by their publishers).
{\bf EPTCS requires the inclusion of a DOI in each cited paper, when available.}

Often an official publication is only available against payment, but
as a courtesy to readers that do not wish to pay, the authors also
make the paper available free of charge at a repository such as
\url{arXiv.org}. In such a case it is recommended to also refer and
link to the URL of the response page of the paper in such a
repository.  This can be done using the bibtex fields {\tt ee} or {\tt
url}, which are treated as synonyms.  These fields should not be used
to duplicate information that is already provided through the DOI of
the paper.
You can find archival-quality URL's for most recently published papers
in DBLP---they are in the bibtex-field {\tt ee}. In fact, it is often
useful to check your references against DBLP records anyway, or just find
them there in the first place.

When using {\LaTeX} rather than {\tt pdflatex} to typeset your paper, by
default no linebreaking within long URLs is allowed. This leads often
to very ugly output, that moreover is different from the output
generated when using {\tt pdflatex}. This problem is repaired when
invoking \href{http://www.cse.unsw.edu.au/~rvg/EPTCS/breakurl.sty}
{\tt $\backslash$usepackage$\{$breakurl$\}$}: it allows linebreaking
within links and yield the same output as obtained by default with
{\tt pdflatex}. 
When invoking {\tt pdflatex}, the package {\tt breakurl} is ignored.

\nocite{*}
\bibliographystyle{eptcs}
\bibliography{generic}
\end{document}

### 2020-01-19 PLACES_2020/paper.tex
PLACES 2020: start structure
PLACES 2020: introduction and Scribble sections

@@
 
+\usepackage[strings]{underscore}
+\usepackage{color}
+\definecolor{purple}{rgb}{0.65, 0.12, 0.82}
+\definecolor{darkgray}{rgb}{.4,.4,.4}
+
+\usepackage{listings}
+\lstdefinelanguage{Scribble}{
+  keywords={module, type, from, as, global,protocol, role, to, choice, at, or, do},
+  keywordstyle=\color{blue}\bfseries,
+  identifierstyle=\color{black},
+  sensitive=false,
+  comment=[l]{//},
+  morecomment=[s]{/*}{*/},
+  commentstyle=\color{darkgray}\ttfamily,
+  stringstyle=\color{purple}\ttfamily,
+  morestring=[b]',
+  morestring=[b]"
+}
+\lstset{
+	basicstyle=\footnotesize\ttfamily,
+	numbers=left,
+	frame=single,
+}
+\usepackage{caption}
+
+\newcommand{\sectionref}[1]{\textsection~\ref{#1}}
+
 \title{Multiparty Session Type-safe Web Development in TypeScript}
-\author{Rob van Glabbeek
-\institute{NICTA\\ Sydney, Australia}
-\email{rvg@cs.stanford.edu}
+\author{Anson Miu
+\institute{Imperial College London}
 \and
-Co Author
-\institute{Stanford Univeristy\\
-California, USA}
-\email{is@gmail.com}
+Fangyi Zhou
+\institute{Imperial College London}
 \and
-Co Author
-\institute{Stanford Univeristy\\
-California, USA}
-\email{is@gmail.com}
+Francisco Ferreira
+\institute{Imperial College London}
 \and
-Co Author
-\institute{Stanford Univeristy\\
-California, USA}
-\email{is@gmail.com}
+Nobuko Yoshida
+\institute{Imperial College London}
 }
-\def\titlerunning{A Longtitled Paper}
-\def\authorrunning{R.J. van Glabbeek, C. Author \& Y.S. Else}
+\def\titlerunning{MPST-safe Web Development in TypeScript}
+\def\authorrunning{A. Miu, F. Zhou, F. Ferreira \& N. Yoshida}
 \begin{document}
@@
 \begin{abstract}
+% Situation: webapps, interactive, JavaScript more powerful but remains a poor language for maintaing large applications
+
+% Problem: hard to maintain/test
+% New languages (like Dart and TypeScript)
+
+% Solution: MPST for web development
+
 This is a sentence in the abstract.
@@
 This is the final sentence in the abstract.
+
+% Problem: similar to PureScript paper
+
+% In this paper, we present...
+
+% TypeScript
+% - gradual type system, superset of JavaScript
+% - we define how our approach achieves "gradual linearity" (can we define this?)
+%Our work targets TypeScript - a superset of \dots
+
+%We demonstrate the effectiveness of our approach through a web-based game of \textit{Noughts and Crosses} where communication is performed through WebSocket connections.
+
 \end{abstract}
@@
 \section{Introduction}
+% Situation: stateful communication over WebSocket and interactive web applications
 
-The optional arguments of {\tt $\backslash$documentclass$\{$eptcs$\}$} are
-\begin{itemize}
-\item at most one of
-{\tt adraft},
-{\tt submission} or
-{\tt preliminary},
-\item at most one of {\tt publicdomain} or {\tt copyright},
-\item and optionally {\tt creativecommons},
-  \begin{itemize}
-  \item possibly augmented with
-    \begin{itemize}
-    \item {\tt noderivs}
-    \item or {\tt sharealike},
-    \end{itemize}
-  \item and possibly augmented with {\tt noncommercial}.
-  \end{itemize}
-\end{itemize}
-We use {\tt adraft} rather than {\tt draft} so as not to confuse hyperref.
-The style-file option {\tt submission} is for papers that are
-submitted to {\tt $\backslash$event}, where the value of the latter is
-to be filled in in line 2 of the tex-file. Use {\tt preliminary} only
-for papers that are accepted but not yet published. The final version
-of your paper that is to be uploaded at the EPTCS website should have
-none of these style-file options.
+Modern interactive web applications aim to provide a highly responsive user experience by minimising the communication latency between clients and servers. 
+Whilst the HTTP request-response model is sufficient for retrieving static assets, applying the same stateless communication approach for interactive use cases (such as a multiplayer game) introduces undesirable performance overhead from having to frequently set up new connections for client-server interactions.
+Developers have since adopted other communication transport abstractions over HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy low-latency full-duplex client-server communication in their applications over a single persistent connection.
+Enabling more complex communication patterns caters for more interactive use cases (such as real-time multiplayer games), but introduces additional concerns to the developer with respect to implementation correctness.
 
-By means of the style-file option
-\href{http://creativecommons.org/about/license/}{creativecommons}
-authors equip their paper with a Creative Commons license that allows
-everyone to copy, distribute, display, and perform their copyrighted
-work and derivative works based upon it, but only if they give credit
-the way you request. By invoking the additional style-file option {\tt
-noderivs} you let others copy, distribute, display, and perform only
-verbatim copies of your work, but not derivative works based upon
-it. Alternatively, the {\tt sharealike} option allows others to
-distribute derivative works only under a license identical to the
-license that governs your work. Finally, you can invoke the option
-{\tt noncommercial} that let others copy, distribute, display, and
-perform your work and derivative works based upon it for
-noncommercial purposes only.
+% Example: naughts and crosses
+Consider a classic turn-based board game of \textit{Noughts and Crosses} between two players.
+Both players are identified by either noughts or crosses respectively, and take turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player wins (when their markers form one straight line on the board) or a stalemate is reached (when all cells are occupied).
 
-Authors' (multiple) affiliations and emails use the commands
-{\tt $\backslash$institute} and {\tt $\backslash$email}.
-Both are optional.
-Authors should moreover supply
-{\tt $\backslash$titlerunning} and {\tt $\backslash$authorrunning},
-and in case the copyrightholders are not the authors also
-{\tt $\backslash$copyrightholders}.
-As illustrated above, heuristic solutions may be called for to share
-affiliations. Authors may apply their own creativity here.
+Figure \ref{fig:tictactoe} illustrates a possible implementation of the game as an interactive real-time multiplayer web application.
+Both players are connected to the game server by bidirectional WebSocket connections and interact with the game from their web browser, which serve a \textit{single-page application} (SPA) of the game client.
+SPAs feature a single HTML page and dynamically renders content via JavaScript in the browser.
+Players take turns to make a move on the game board and the server implements the game logic to progress the game forward until a result (either a win/loss or draw) can be declared.
 
-Exactly 46 lines fit on a page.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\hfill6\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\hfill11\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
+Whilst WebSockets make this web-based implementation possible, it introduces the developer to a new family of communication errors, even for this simple game.
+In addition to the usual testing for game logic correctness, the developer needs to test against \textit{deadlocks} (e.g.
+both players waiting for each other to make a move at the same time) and \textit{communication mismatches} (e.g.
+player 1 sending a boolean to the game server instead of the board coordinates).
+The complexity of these errors, which correlate to the complexity of tests required against these errors, scale with the complexity of the communication patterns involved.
 
-Here starts a new paragraph. The rest is like any normal {\LaTeX} article.
-We will spare you the details.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\hfill16\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\hfill21\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\hfill26\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\hfill31\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\hfill36\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\hfill41\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.\hfill46\\
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
-The rest is like any normal {\LaTeX} article.
-We will spare you the details.
+\textit{Multiparty Session Types} (MPST) \cite{MPST} provide a framework for formally specifying a structured communication pattern between concurrent processes and verifying implementations for correctness with respect to the communications aspect.
+By specifying the client-server interactions of our game as a protocol and verifying the implementations against the protocol for well-formedness, MPST theory guarantees well-formed implementations to be free from communication errors.
 
-\section{Prefaces}
+This MPST methodology has been applied to target applications written in both statically and dynamically typed languages in \cite{Hybrid2016, Python2017}, but do not apply to the event-driven paradigm of interactive web applications.
+A proposal that caters for web applications in \cite{PureScript2019} targets PureScript to maximise the static guarantees it can provide, but the strong type system make it difficult to extract common code and the functional paradigm is arguably unidiomatic relative to mainstream web development in JavaScript, which limits developer productivity.
+JavaScript began as the language of the web browser, but a shallow learning curve, an ever-growing ecosystem of libraries and frameworks (such as React.js for writing SPAs) and a successful server-side single-threaded event loop implementation in \textit{Node.js} has propelled it as the de-facto choice for full-stack web application development in a single language.
 
-Volume editors may create prefaces using this very template,
-with {\tt $\backslash$title$\{$Preface$\}$} and {\tt $\backslash$author$\{\}$}.
+Whilst the proposal in \cite{Python2017} could be adapted to verify interactive web applications written in JavaScript, the emergence of Microsoft's TypeScript \cite{TypeScript} with a gradual type system motivates the need for a MPST-based workflow for verifying full-stack TypeScript web applications, such that it is flexible with respect to idiomatic practices and frameworks to be useful in industry whilst also leveraging the language features to maximise the static guarantees it is able to provide.
+Such a workflow would ultimately decrease the overhead for incorporating MPST into mainstream web development, which reduces development time by programmatically verifying implementations for communication correctness.
 
-\section{Bibliography}
+\paragraph{Contributions}
+This paper presents a workflow for developing type-safe interactive SPAs motivated by the MPST framework: \textbf{(1)} An endpoint API code generation workflow targeting TypeScript-based web applications for multiparty sessions; \textbf{(2)} An encoding of session types in server-side TypeScript that enforces static linearity; and \textbf{(3)} An encoding of session types in browser-side TypeScript using the React framework that guarantees affine usage of communication channels.
 
-We request that you use
-\href{http://www.cse.unsw.edu.au/~rvg/EPTCS/eptcs.bst}
-{\tt $\backslash$bibliographystyle$\{$eptcs$\}$}
-\cite{bibliographystylewebpage}. Compared to the original {\LaTeX}
-{\tt $\backslash$biblio\-graphystyle$\{$plain$\}$},
-it ignores the field {\tt month}, and uses the extra
-bibtex fields {\tt eid}, {\tt doi}, {\tt ee} and {\tt url}.
-The first is for electronic identifiers (typically the number $n$
-indicating the $n^{\rm th}$ paper in an issue) of papers in electronic
-journals that do not use page numbers. The other three are to refer,
-with life links, to electronic incarnations of the paper.
+Figure \ref{fig:workflow} illustrates our proposed development workflow: we extend the Scribble framework (\sectionref{section:scribble}) and generate TypeScript endpoints for server-side (\sectionref{section:server}) and browser-side (\sectionref{section:browser}) targets.
+We will use the \textit{Noughts and Crosses} game as our running example in the rest of the paper and show how our approach is a practical compromise that combines benefits from static session typing and being compatible with common libraries and JavaScript idiomatic practices (\sectionref{section:example}).
 
-Almost all publishers use digital object identifiers (DOIs) as a
-persistent way to locate electronic publications. Prefixing the DOI of
-any paper with {\tt http://dx.doi.org/} yields a URI that resolves to the
-current location (URL) of the response page\footnote{Nowadays, papers
-  that are published electronically tend
-  to have a \emph{response page} that lists the title, authors and
-  abstract of the paper, and links to the actual manifestations of
-  the paper (e.g.\ as {\tt dvi}- or {\tt pdf}-file). Sometimes
-  publishers charge money to access the paper itself, but the response
-  page is always freely available.}
-of that paper. When the location of the response page changes (for
-instance through a merge of publishers), the DOI of the paper remains
-the same and (through an update by the publisher) the corresponding
-URI will then resolve to the new location. For that reason a reference
-ought to contain the DOI of a paper, with a life link to corresponding
-URI, rather than a direct reference or link to the current URL of
-publisher's response page. This is the r\^ole of the bibtex field {\tt doi}.
-DOIs of papers can often be found through
-\url{http://www.crossref.org/guestquery};\footnote{For papers that will appear
-  in EPTCS and use \href{http://www.cse.unsw.edu.au/~rvg/EPTCS/eptcs.bst}
-  {\tt $\backslash$bibliographystyle$\{$eptcs$\}$} there is no need to
-  find DOIs on this website, as EPTCS will look them up for you
-  automatically upon submission of a first version of your paper;
-  these DOIs can then be incorporated in the final version, together
-  with the remaining DOIs that need to found at DBLP or publisher's webpages.}
-the second method {\it Search on article title}, only using the {\bf
-surname} of the first-listed author, works best.  
-Other places to find DOIs are DBLP and the response pages for cited
-papers (maintained by their publishers).
-{\bf EPTCS requires the inclusion of a DOI in each cited paper, when available.}
+\section{The Scribble Framework}
+\label{section:scribble}
+Development begins by specifying the permitted communications between participants as a \textit{global protocol} in Scribble, a MPST-based framework introduced in \cite{Scribble} which features a protocol specification language and code generation toolchain.
+This means that a protocol specified in Scribble maps to some \textit{global type} in multiparty session type theory \cite{MPST}.
 
-Often an official publication is only available against payment, but
-as a courtesy to readers that do not wish to pay, the authors also
-make the paper available free of charge at a repository such as
-\url{arXiv.org}. In such a case it is recommended to also refer and
-link to the URL of the response page of the paper in such a
-repository.  This can be done using the bibtex fields {\tt ee} or {\tt
-url}, which are treated as synonyms.  These fields should not be used
-to duplicate information that is already provided through the DOI of
-the paper.
-You can find archival-quality URL's for most recently published papers
-in DBLP---they are in the bibtex-field {\tt ee}. In fact, it is often
-useful to check your references against DBLP records anyway, or just find
-them there in the first place.
+We specify the \textit{Noughts and Crosses} game as a Scribble protocol in Listing \ref{lst:game} and explain the syntax of the language through this example. 
+Scribble protocols are organised into \textit{modules}: line 1 declares the name of the module.
+Line 2 declares a message type to be used as payload types in the global protocol: this would correspond to an \texttt{import \{ Coordinate as Point \} from './Types'} statement in TypeScript, where the developer provides the definition of the \texttt{Coordinate} interface under a \texttt{Types.ts} file.
+Line 4 declares the name of the global protocol (\texttt{Game}) and specifies the participants involved in the protocol as \textit{roles}; the protocol body is defined by lines 5-16.
+A message passing statement from a sender (e.g.
+\texttt{P1}) to a receiver (e.g.
+\texttt{Svr}) require a \textit{label identifier} followed by a list of \textit{payload types} in parentheses (e.g.
+\texttt{Pos(Point) from P1 to Svr} on line 5).
+Semicolons denote continuations for message passing statement in the same way that they express sequential statement execution in programming languages such as Java.
 
-When using {\LaTeX} rather than {\tt pdflatex} to typeset your paper, by
-default no linebreaking within long URLs is allowed. This leads often
-to very ugly output, that moreover is different from the output
-generated when using {\tt pdflatex}. This problem is repaired when
-invoking \href{http://www.cse.unsw.edu.au/~rvg/EPTCS/breakurl.sty}
-{\tt $\backslash$usepackage$\{$breakurl$\}$}: it allows linebreaking
-within links and yield the same output as obtained by default with
-{\tt pdflatex}. 
-When invoking {\tt pdflatex}, the package {\tt breakurl} is ignored.
+\begin{figure}[!h]
+\begin{lstlisting}[
+	language=Scribble
+]
+module NoughtsAndCrosses;
+type <typescript> "Coordinate" from "./Types" as Point;	// Position on board
 
+global protocol Game(role Svr, role P1, role P2) {
+  Pos(Point) from P1 to Svr;
+  choice at Svr {
+    Lose(Point) from Svr to P2;
+    Win(Point) from Svr to P1;
+  } or {
+    Draw(Point) from Svr to P2;
+    Draw(Point) from Svr to P1;
+  } or {
+    Update(Point) from Svr to P2;
+    Update(Point) from Svr to P1;
+    do Game(Svr, P2, P1);
+  }
+}
+\end{lstlisting}
+\captionof{lstlisting}{Main body of the \textit{Noughts and Crosses} protocol.}
+\label{lst:game}
+\end{figure}
+
+The \texttt{Game} protocol involves three roles: the server (\texttt{Svr}) and two players (\texttt{P1} and \texttt{P2}).
+First, \texttt{P1} makes a move by sending a coordinate to place the marker on the game board to \texttt{Svr}.
+Upon receiving the \texttt{Pos} message, \texttt{Svr} will implement the game logic to decide the outcome of \texttt{P1}'s move and communicate with \texttt{P2} accordingly.
+We express this as a \texttt{choice} statement (lines 6-15), where \texttt{Svr} decides whether \texttt{P1}'s move resulted in a loss for \texttt{P2} (lines 7-8), stalemate (lines 10-11) or an unfinished game (lines 13-15).
+
+If \texttt{P1} performed a winning move (i.e.
+the latest coordinate resulted in three-in-a-row for \texttt{P1}), then \texttt{Svr} sends a \texttt{Lose} message to \texttt{P2} and a \texttt{Win} message to \texttt{P1} with the coordinates of the latest move.
+These coordinates are sent to allow the applications running in the browser for both players to update their version of the game state respectively; how they choose to render this on the DOM is up to their implementation, independent of the protocol.
+Similar message passing actions are sent in the case of a draw, just with updated labels for \texttt{Draw}.
+
+Otherwise, the game has not ended, so \texttt{Svr} updates both players with the coordinates of the latest move by sending an \texttt{Update} message, and the game proceeds to allow \texttt{P2} to make a move.
+We express this as a recursive call to the \texttt{Game} protocol using a \texttt{do} statement: note that we swap the role parameters \texttt{P2} and \texttt{P1} to express the turn-taking semantics of the game.
+In short, the global protocol describes \texttt{P1} and \texttt{P2} taking turns until \texttt{Svr} declares a winner/loser or draw.
+
+We leverage the Scribble toolchain to check for protocol well-formedness.
+This directly corresponds to the multiparty session type theory of verifying valid local type \textit{projections} for all participants of a global type.
+The implementation of this validation by the toolchain follow from the algorithmic projection procedures defined in \cite{MPST}.
+We obtain \textit{endpoint protocols} for each role in a well-formed global protocol.
+An endpoint protocol only preserves the interactions defined by the global protocol in which the target role is involved.
+For example, the endpoint protocol for \texttt{P1} will not include the exchange of messages between \texttt{Svr} and \texttt{P2}.
+An endpoint protocol corresponds to an equivalent \textit{Endpoint Finite State Machine} (EFSM) in the Scribble toolchain.
+We adopt the formalism of EFSMs in \cite{Hybrid2016}, and use the EFSMs as a basis for API generation.
+
+\section{Encoding Session Types in TypeScript}
+% Design choices
+
+\subsection{Server-side API generation}
+\label{section:server}
+% How we interpret EFSM for
+
+\subsubsection{Linear channel usage}
+% Callback approach does not expose any channels
+
+\subsection{Browser-side API generation}
+\label{section:browser}
+
+\subsubsection{The React Framework}
+
+\subsubsection{Affine channel usage}
+
+\section{Case Study}
+\label{section:example}
+% TicTacToe example - backend
+
+% TicTacToe example - frontend
+
+\section{Related Works}
+% New languages written  (e.g. ATS)
+%	- addresses the challenge of resource linearity with a liner type system
+%	- problem with webdev: not compatible with libraries, technologies, flexibility of JavaScript (ultimately the language of the browser)
+
+% API generation
+% - original Python implementation, generating monitors (Rumi's paper)
+%	- Gradual type system lets us do better than just dynamic runtime checks
+%	- could do the same approach with generating JavaScript monitors, but we have TS and Dart and can leverage more powerful type systems to do more static/compiler checks
+
+% - Java (hybrid approach)
+%	- leverage strongly typed system for static behavioural typing
+%	- minimse dupliction by defining states in terms of IO interfaces (same IO state with different successor can still share code)
+%	- runtime linearity checks
+
+% Typestates (Mungo)
+% - the React components we generate is essentially behaving like typestates of the role, where 
+% each role has its set of permitted actions.
+
+% Session types in web development
+% - purescript looking at the whole workflow
+	% - problem: PureScript isn't as prevalent
+% - simon fowler paper on MVU targeting linearity for frontend GUI
+
+\section{Conclusion and Future Works}
+
+% Future work
+% - generalise to encode protocols as typestates (e.g. Mungo example)
+% - WebRTC for full MPST (since current limitation is w.r.t. MPST protocols with 1-server many-client roles
+
+%\section{Introduction}
+%% Emergence of interactive JavaScript-intensive applications that maintain continuous stateful
+%
+%The optional arguments of {\tt $\backslash$documentclass$\{$eptcs$\}$} are
+%\begin{itemize}
+%\item at most one of
+%{\tt adraft},
+%{\tt submission} or
+%{\tt preliminary},
+%\item at most one of {\tt publicdomain} or {\tt copyright},
+%\item and optionally {\tt creativecommons},
+%  \begin{itemize}
+%  \item possibly augmented with
+%    \begin{itemize}
+%    \item {\tt noderivs}
+%    \item or {\tt sharealike},
+%    \end{itemize}
+%  \item and possibly augmented with {\tt noncommercial}.
+%  \end{itemize}
+%\end{itemize}
+%We use {\tt adraft} rather than {\tt draft} so as not to confuse hyperref.
+%The style-file option {\tt submission} is for papers that are
+%submitted to {\tt $\backslash$event}, where the value of the latter is
+%to be filled in in line 2 of the tex-file. Use {\tt preliminary} only
+%for papers that are accepted but not yet published. The final version
+%of your paper that is to be uploaded at the EPTCS website should have
+%none of these style-file options.
+%
+%By means of the style-file option
+%\href{http://creativecommons.org/about/license/}{creativecommons}
+%authors equip their paper with a Creative Commons license that allows
+%everyone to copy, distribute, display, and perform their copyrighted
+%work and derivative works based upon it, but only if they give credit
+%the way you request. By invoking the additional style-file option {\tt
+%noderivs} you let others copy, distribute, display, and perform only
+%verbatim copies of your work, but not derivative works based upon
+%it. Alternatively, the {\tt sharealike} option allows others to
+%distribute derivative works only under a license identical to the
+%license that governs your work. Finally, you can invoke the option
+%{\tt noncommercial} that let others copy, distribute, display, and
+%perform your work and derivative works based upon it for
+%noncommercial purposes only.
+%
+%Authors' (multiple) affiliations and emails use the commands
+%{\tt $\backslash$institute} and {\tt $\backslash$email}.
+%Both are optional.
+%Authors should moreover supply
+%{\tt $\backslash$titlerunning} and {\tt $\backslash$authorrunning},
+%and in case the copyrightholders are not the authors also
+%{\tt $\backslash$copyrightholders}.
+%As illustrated above, heuristic solutions may be called for to share
+%affiliations. Authors may apply their own creativity here.
+%
+%Exactly 46 lines fit on a page.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\hfill6\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\hfill11\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%
+%Here starts a new paragraph. The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\hfill16\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\hfill21\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\hfill26\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\hfill31\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\hfill36\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\hfill41\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.\hfill46\\
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%The rest is like any normal {\LaTeX} article.
+%We will spare you the details.
+%
+%\section{Prefaces}
+%
+%Volume editors may create prefaces using this very template,
+%with {\tt $\backslash$title$\{$Preface$\}$} and {\tt $\backslash$author$\{\}$}.
+%
+%\section{Bibliography}
+%
+%We request that you use
+%\href{http://www.cse.unsw.edu.au/~rvg/EPTCS/eptcs.bst}
+%{\tt $\backslash$bibliographystyle$\{$eptcs$\}$}
+%\cite{bibliographystylewebpage}. Compared to the original {\LaTeX}
+%{\tt $\backslash$biblio\-graphystyle$\{$plain$\}$},
+%it ignores the field {\tt month}, and uses the extra
+%bibtex fields {\tt eid}, {\tt doi}, {\tt ee} and {\tt url}.
+%The first is for electronic identifiers (typically the number $n$
+%indicating the $n^{\rm th}$ paper in an issue) of papers in electronic
+%journals that do not use page numbers. The other three are to refer,
+%with life links, to electronic incarnations of the paper.
+%
+%Almost all publishers use digital object identifiers (DOIs) as a
+%persistent way to locate electronic publications. Prefixing the DOI of
+%any paper with {\tt http://dx.doi.org/} yields a URI that resolves to the
+%current location (URL) of the response page\footnote{Nowadays, papers
+%  that are published electronically tend
+%  to have a \emph{response page} that lists the title, authors and
+%  abstract of the paper, and links to the actual manifestations of
+%  the paper (e.g.\ as {\tt dvi}- or {\tt pdf}-file). Sometimes
+%  publishers charge money to access the paper itself, but the response
+%  page is always freely available.}
+%of that paper. When the location of the response page changes (for
+%instance through a merge of publishers), the DOI of the paper remains
+%the same and (through an update by the publisher) the corresponding
+%URI will then resolve to the new location. For that reason a reference
+%ought to contain the DOI of a paper, with a life link to corresponding
+%URI, rather than a direct reference or link to the current URL of
+%publisher's response page. This is the r\^ole of the bibtex field {\tt doi}.
+%DOIs of papers can often be found through
+%\url{http://www.crossref.org/guestquery};\footnote{For papers that will appear
+%  in EPTCS and use \href{http://www.cse.unsw.edu.au/~rvg/EPTCS/eptcs.bst}
+%  {\tt $\backslash$bibliographystyle$\{$eptcs$\}$} there is no need to
+%  find DOIs on this website, as EPTCS will look them up for you
+%  automatically upon submission of a first version of your paper;
+%  these DOIs can then be incorporated in the final version, together
+%  with the remaining DOIs that need to found at DBLP or publisher's webpages.}
+%the second method {\it Search on article title}, only using the {\bf
+%surname} of the first-listed author, works best.  
+%Other places to find DOIs are DBLP and the response pages for cited
+%papers (maintained by their publishers).
+%{\bf EPTCS requires the inclusion of a DOI in each cited paper, when available.}
+%
+%Often an official publication is only available against payment, but
+%as a courtesy to readers that do not wish to pay, the authors also
+%make the paper available free of charge at a repository such as
+%\url{arXiv.org}. In such a case it is recommended to also refer and
+%link to the URL of the response page of the paper in such a
+%repository.  This can be done using the bibtex fields {\tt ee} or {\tt
+%url}, which are treated as synonyms.  These fields should not be used
+%to duplicate information that is already provided through the DOI of
+%the paper.
+%You can find archival-quality URL's for most recently published papers
+%in DBLP---they are in the bibtex-field {\tt ee}. In fact, it is often
+%useful to check your references against DBLP records anyway, or just find
+%them there in the first place.
+%
+%When using {\LaTeX} rather than {\tt pdflatex} to typeset your paper, by
+%default no linebreaking within long URLs is allowed. This leads often
+%to very ugly output, that moreover is different from the output
+%generated when using {\tt pdflatex}. This problem is repaired when
+%invoking \href{http://www.cse.unsw.edu.au/~rvg/EPTCS/breakurl.sty}
+%{\tt $\backslash$usepackage$\{$breakurl$\}$}: it allows linebreaking
+%within links and yield the same output as obtained by default with
+%{\tt pdflatex}. 
+%When invoking {\tt pdflatex}, the package {\tt breakurl} is ignored.
+%
 \nocite{*}

### 2020-01-20 PLACES_2020/paper.tex
PLACES 2020: begin server-side implementation

@@
 
+\usepackage{stmaryrd}
 \usepackage[strings]{underscore}
@@
 
+Developers can validate the communication aspects of their implementation against their EFSM to verify protocol conformance.
+Our approach integrates the EFSM into the development workflow by encoding session types as TypeScript types.
+We appreciate the inherent differences in development considerations between front- and back-end web development: business logic tends to initiate communication in the back-end, whilst communication in the front-end is mostly driven by user interactions with the browser.
+This motivates our design choice of generating different session types encodings for the target application on the server (\sectionref{section:server}) and in the browser (\sectionref{section:browser}).
+
 \subsection{Server-side API generation}
 \label{section:server}
-% How we interpret EFSM for
+For server-side targets, we encode EFSM states as TypeScript types depending on the type of state.
+We adopt the EFSM definition presented in \cite{Hybrid2016} to consider receive and send states separately.
+We assign each TypeScript encoding its state identifier from the EFSM as its type alias, thus providing syntactic sugar when referring to the successor state in the TypeScript encoding of the current state.
+For any state $S$ in the EFSM, we refer to the TypeScript type alias of its encoding as $\llbracket S \rrbracket$.
+
+We make a key design decision \textit{not} to expose communication channels in the TypeScript session type encodings to provide linearity guarantees (\sectionref{section:serverlinear}).
+Our encodings sufficiently exposes seams for the developer to inject their business logic, whilst the generated session API (\sectionref{section:serversessionapi}) handles the sending and receiving of messages; as a result, the encodings do not concern with the role involved in the send/receive action.
+We outline the encodings below and provide examples for the \textit{Noughts and Crosses} game server (\sectionref{section:exampleserver}).
+
+\paragraph{Branching state} We consider a receive state as a unary branching state for conciseness.
+A branching state is encoded as an \textit{object literal}, with each branch corresponding to a member field.
+A branch expecting to receive a message labelled $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as an \textit{member field} named $\texttt{label}_i$ of function type \texttt{(payload:$\texttt{T}_i$) => $\llbracket S_i \rrbracket$}.
+The developer implements a branching operation by passing callbacks for each branch, parameterised by the expected message payload type for that branch.
+
+\paragraph{Selection state} We consider a send state as a unary selection state for conciseness.
+A selection state is encoded as a \textit{union type} of internal choice encodings: each internal choice sending a message labelled $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as a \textit{tuple type} of \texttt{[Label.label$_i$, T$_i$, $\llbracket S_i \rrbracket$]}.
+The developer implements a selection operation by passing the selected label and payload to send in the message.
+We generate a \textit{string enum} (named \texttt{Label}) wrapping the labels in the protocol, hence the enum member access in the first element of the tuple type.
+
+\subsubsection{Session API}
+\label{section:serversessionapi}
 
 \subsubsection{Linear channel usage}
+\label{section:serverlinear}
+Our callback-oriented design for server-side API generation provides eguarantees on state channel linearity by enforcing the following two properties.
+
+\paragraph{Repeat use} By adopting a callback-oriented design for server-side API generation, channels are not directly accessed by the programmer which makes \textit{reuse} impossible.
+Consider the \textit{Noughts and Crosses} server endpoint: the programmer must pass a callback that handles the receipt of a \texttt{Pos(Point)} message and returns either a \texttt{Lose(Point)}, \texttt{Draw(Point)} or \texttt{Update(Point)} message with the appropriate continuation.
+Upon receiving the \texttt{Pos(Point)} message from a player, our lightweight runtime invokes the callback to get \texttt{Svr}'s choice from the return value, and perform the appropriate send.
+This is transparent to the programmer and it is impossible for the programmer to send or receive messages more than once, by design.
+
+\paragraph{Unused} Callbacks for all states of the EFSM must be supplied to the constructor in order to create the session.
+This will cover the terminal state (if it exists in the EFSM, as \cite{Hybrid2016} notes that an EFSM contains at most one terminal state).
+
+Thus, \dots
+
 % Callback approach does not expose any channels
@@
 
+\subsubsection{Model transitions}
+
 \subsubsection{Affine channel usage}
@@
 \label{section:example}
+We apply our framework to implement a web-based implementation of the \textit{Noughts and Crosses} running example in TypeScript; the interested reader can find the full implementation in \cite{NoughtsAndCrosses}.
+In addition to showing the multiparty session type safety achieved by our generated APIs, we also show that our library design welcomes idiomatic JavaScript practices in the user implementation and is interoperable with common front- and back-end frameworks.
+
+\subsection{Game server}
+\label{section:exampleserver}
 % TicTacToe example - backend
 
+% Note: no state space explosion, allows for reusability
+
+\subsection{Game clients}
 % TicTacToe example - frontend
@@
 \section{Related Works}
+\paragraph{Languages with native session type support} \dots
 % New languages written  (e.g. ATS)
@@
 
+\paragraph{Endpoint API generation} \dots
 % API generation
@@
 
-% Typestates (Mungo)
-% - the React components we generate is essentially behaving like typestates of the role, where 
-% each role has its set of permitted actions.
-
+\paragraph{Session types in web development} \dots
 % Session types in web development
@@
 % - simon fowler paper on MVU targeting linearity for frontend GUI
+
+\paragraph{Typestate programming} \dots
+% Typestates (Mungo)
+% - the React components we generate is essentially behaving like typestates of the role, where 
+% each role has its set of permitted actions.
 

### 2020-01-21 PLACES_2020/paper.tex
PLACES 2020: begin front-end API generation

@@
 
+\usepackage{wrapfig}
+\usepackage{graphicx}
+\usepackage{subcaption}
 \usepackage{stmaryrd}
@@
   keywordstyle=\color{blue}\bfseries,
+  identifierstyle=\color{black},
+  sensitive=false,
+  comment=[l]{//},
+  morecomment=[s]{/*}{*/},
+  commentstyle=\color{darkgray}\ttfamily,
+  stringstyle=\color{purple}\ttfamily,
+  morestring=[b]',
+  morestring=[b]"
+}
+\lstdefinelanguage{JavaScript}{
+  keywords={typeof, new, true, false, catch, function, return, null, catch, switch, var, if, in, while, do, else, case, break, class, export, boolean, throw, implements, import, this, const, let},
+  keywordstyle=\color{blue}\bfseries,
+%  ndkeywords={class, export, boolean, throw, implements, import, this},
+%  ndkeywordstyle=\color{darkgray}\bfseries,
   identifierstyle=\color{black},
@@
 Figure \ref{fig:workflow} illustrates our proposed development workflow: we extend the Scribble framework (\sectionref{section:scribble}) and generate TypeScript endpoints for server-side (\sectionref{section:server}) and browser-side (\sectionref{section:browser}) targets.
-We will use the \textit{Noughts and Crosses} game as our running example in the rest of the paper and show how our approach is a practical compromise that combines benefits from static session typing and being compatible with common libraries and JavaScript idiomatic practices (\sectionref{section:example}).
+We will use the \textit{Noughts and Crosses} game as our running example in the rest of the paper and show how our approach is a practical compromise that combines benefits from static session typing and being compatible with common libraries and idiomatic JavaScript practices (\sectionref{section:example}).
 
@@
 \label{section:server}
+
+\begin{wrapfigure}{r}{0.5\textwidth}
+  \begin{center}
+    \includegraphics[width=0.48\textwidth]{figures/efsm_svr.png}
+  \end{center}
+
+  \captionof{figure}{EFSM for \texttt{Svr} in \textit{Noughts and Crosses}.}
+  \label{fig:efsmsvr}
+\end{wrapfigure}
+
+We will refer to the \texttt{Svr} EFSM (fig.~\ref{fig:efsmsvr}) as a running example in this section.
 For server-side targets, we encode EFSM states as TypeScript types depending on the type of state.
@@
 Our encodings sufficiently exposes seams for the developer to inject their business logic, whilst the generated session API (\sectionref{section:serversessionapi}) handles the sending and receiving of messages; as a result, the encodings do not concern with the role involved in the send/receive action.
-We outline the encodings below and provide examples for the \textit{Noughts and Crosses} game server (\sectionref{section:exampleserver}).
+We outline the encodings below using examples from the \textit{Noughts and Crosses} game server (listing \ref{lst:svr}).
 
 \paragraph{Branching state} We consider a receive state as a unary branching state for conciseness.
-A branching state is encoded as an \textit{object literal}, with each branch corresponding to a member field.
+A branching state is encoded as an \textit{object literal} \cite{TypeScriptSpec}, with each branch corresponding to a member field.
 A branch expecting to receive a message labelled $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as an \textit{member field} named $\texttt{label}_i$ of function type \texttt{(payload:$\texttt{T}_i$) => $\llbracket S_i \rrbracket$}.
@@
 \paragraph{Selection state} We consider a send state as a unary selection state for conciseness.
-A selection state is encoded as a \textit{union type} of internal choice encodings: each internal choice sending a message labelled $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as a \textit{tuple type} of \texttt{[Label.label$_i$, T$_i$, $\llbracket S_i \rrbracket$]}.
+A selection state is encoded as a \textit{union type} \cite{TypeScriptSpec} of internal choice encodings: each internal choice sending a message labelled $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as a \textit{tuple type} of \texttt{[Labels.label$_i$, T$_i$, $\llbracket S_i \rrbracket$]}.
 The developer implements a selection operation by passing the selected label and payload to send in the message.
-We generate a \textit{string enum} (named \texttt{Label}) wrapping the labels in the protocol, hence the enum member access in the first element of the tuple type.
+We generate a \textit{string enum} (named \texttt{Labels}) wrapping the labels in the protocol, hence the enum member access in the first element of the tuple type.
 
-\subsubsection{Session API}
+\begin{figure}[!h]
+\begin{lstlisting}[language=JavaScript, title=\texttt{./NoughtsAndCrosses/Game.ts}]
+import { Coordinate as Point } from "./Types";
+import { Labels } from "./Constants";
+
+export type S13 = { Pos: (payload: Point) => S15 }
+export type S15 = [Labels.Lose, Point, S16]
+		| [Labels.Draw, Point, S17]
+		| [Labels.Update, Point, S18]
+\end{lstlisting}  
+\captionof{lstlisting}{Example encodings from \textit{Noughts and Crosses} \texttt{Svr} EFSM.}
+\label{lst:svr}
+\end{figure}
+
+\subsubsection{Session Runtime and Protocol API}
 \label{section:serversessionapi}
+
+Our session runtime performs communication in a way that conforms to the protocol specification.
+The runtime listens to message (receive) events on the communication channel, invokes the corresponding callback to obtain the value to send next, and performs the send.
+
+The \textit{Protocol API} is a class that interfaces between the session runtime and the user implementation of the EFSM state encodings.
+Upon construction, the Protocol class waits for all client roles to connect via WebSocket, then proceeds to the initial state of the EFSM and performs state transitions depending on the permitted communication patterns from the EFSM.
+The Protocol class constructor requires the WebSocket server and the following handlers:
+
+\paragraph{Initial state} When all roles have joined the session, the Protocol class needs to know what to do in the initial state to start communication, be it sending a message to a client or offering branches to a client.
+For the \textit{Noughts and Crosses} game server, this would be a callback to handle the receipt of the \texttt{Pos(Point)} message from \texttt{P1}.
+
+\paragraph{Receive handlers for each role} The runtime needs to bind a message event listener to the WebSocket between the server and each (client) role, and needs to know how to handle the incoming messages.
+For each role, the user is required to pass callbacks that encode the receive states for that role.
+The type alias for such a handler is the \textit{intersection type} \cite{TypeScriptSpec} of all receive states concerning that role.
+Considering a hypothetical web-service providing addition and square root functionality to a client role \texttt{C}, if \texttt{S1} is a state expecting to receive \texttt{Add(int, int)} from \texttt{C} and \texttt{S4} is a state expecting to receive \texttt{Sqrt(int)} from \texttt{C}, we define \texttt{type Receive_C = S1 \& S4} as the receive handler type alias and the user needs to pass some value of \texttt{Receive_C} when instantiating the session through the Protocol API.
 
@@
 \label{section:serverlinear}
-Our callback-oriented design for server-side API generation provides eguarantees on state channel linearity by enforcing the following two properties.
+Our callback-oriented design for server-side API generation provides guarantees on state channel linearity by preventing the two properties detailed below.
+Here, channel linearity is guaranteed based on the correctness of our library design and session runtime implementation, which means a faulty implementation could violate this, but is up to the library author to verify once rather than the end user.
 
-\paragraph{Repeat use} By adopting a callback-oriented design for server-side API generation, channels are not directly accessed by the programmer which makes \textit{reuse} impossible.
+\paragraph{Repeat use} By adopting a callback-oriented design for server-side API generation, channels are not directly accessed by the programmer which makes \textit{reuse} impossible; only the Protocol API is exposed but channel operations are concealed in a private class.
 Consider the \textit{Noughts and Crosses} server endpoint: the programmer must pass a callback that handles the receipt of a \texttt{Pos(Point)} message and returns either a \texttt{Lose(Point)}, \texttt{Draw(Point)} or \texttt{Update(Point)} message with the appropriate continuation.
@@
 
-\paragraph{Unused} Callbacks for all states of the EFSM must be supplied to the constructor in order to create the session.
-This will cover the terminal state (if it exists in the EFSM, as \cite{Hybrid2016} notes that an EFSM contains at most one terminal state).
-
-Thus, \dots
-
-% Callback approach does not expose any channels
+\paragraph{Unused} The initial state must be supplied to the Protocol API constructor in order to instantiate the session; this initial state is defined in terms of the successor states, which in turn has references to its successors and so forth.
+This encoding approach will cover the terminal state (if it exists in the EFSM, as \cite{Hybrid2016} notes that an EFSM contains at most one terminal state), and the session runtime guarantees this terminal state, if it exists, will be reached by construction.
 
@@
 
+\begin{wrapfigure}{r}{0.5\textwidth}
+  \begin{center}
+    \includegraphics[width=0.48\textwidth]{figures/efsm_p1.png}
+  \end{center}
+
+  \captionof{figure}{EFSM for \texttt{P1} in \textit{Noughts and Crosses}.}
+  \label{fig:efsmp1}
+\end{wrapfigure}
+
+We will refer to the \texttt{P1} EFSM (figure \ref{fig:efsmp1}) as a running example in this section.
+Preserving behavioural typing and channel linearity are known to be challenging for browser-side applications due to the event-driven nature of user interaction: in the case of \textit{Noughts and Crosses}, once the user makes a move by clicking on a cell on the game board, this click event must be deactivated until the user's next turn, otherwise the user can click again and violate channel linearity.
+Our design goal is to enforce this statically through the APIs we generate.
+For browser-side targets, we extend the approach presented in \cite{MVU2019} on \textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU) architecture.
+Each state in the EFSM is a model type and uniquely defines a \textit{view function}, set of \textit{messages} and \textit{update function}: the view function defines what to render to the DOM, the set of messages define the possible (IO) actions available at that state, and the update function defines which successor state to transition to, given some supported IO action at this state.
+
 \subsubsection{The React Framework}
+Rather than using the LINKS web programming language from \cite{MVU2019}, we apply the multiple model types approach to the React framework \cite{React} developed by Facebook.
+React is widely used in industry to create scalable single-page TypeScript applications, and we intend for our proposed workflow to be beneficial in an industrial context.
+We introduce the key features of the framework below.
 
-\subsubsection{Model transitions}
+\paragraph{Components} \dots
+% Component = basic building blocks 
+% Life cycle
+% JSX
+
+\paragraph{Virtual DOM} \dots
+% virtual-dom
+
+\paragraph{Uni-directional data flow} \dots
+% Parent provides props to child (uni-directional workflow)
+
+\subsubsection{Model types in React}
+Although React was originally intended for the \textit{Model-View-Controller} architecture, we use the framework to express the model types approach.
+
+% State = React component
+% Send transitions = model transition factory
+% Receive transitions = handled by 
+% Protocol component = handle joining and rendering of correct EFSM component
 
 \subsubsection{Affine channel usage}
+
+\paragraph{Repeat use}
+
+\paragraph{Unused} Our approach \textit{does not} statically detect whether all transitions in a certain state are bound to some UI event.
+This means that it is possible for an implementation to \textit{not} handle transitions to a terminal state but still type-check, so we cannot prevent unused states.
+Consider a hypothetical protocol where, in the current state, sending \texttt{Quit()} to some server role \texttt{S} transitions to the terminal state: \dots
 
@@
 
+We set up the WebSocket server on top of an Express.js \cite{ExpressJS}application running on Node.js \cite{NodeJS}.
+We define our own game logic in a \texttt{Board} class to keep track of the game state and expose methods to query the result.
+This custom logic is integrated into our
+\texttt{Receive\_P1} and \texttt{Receive\_P2} handlers (listing are defined using 
+encapsulate our custom game logic in a class
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=JavaScript, tabsize=4, title=\texttt{./app.ts}]
+const handleP1Move: Receive_P1 = (move: Point) => {
+	// User-implemented logic
+	board.P1(move)
+	if (board.won()) { return [Labels.Lose, move, move] }
+	else if (board.draw()) { return [Labels.Draw, move, move] }
+	else { return [Labels.Draw, move, handleP2Move] } 
+}
+
+const handleP2Move: Receive_P2 = ...
+
+// Instantiate session
+const webSocketServer: WebSocket.Server = ...
+new GameServer(webSocketServer, handleP1Move, handleP2Move)
+\end{lstlisting}  
+\captionof{lstlisting}{Example fragment of \textit{Noughts and Crosses} \texttt{Svr} implementation.}
+\label{lst:svrprotocol}
+\end{figure}
+
 % Note: no state space explosion, allows for reusability
@@
 
+\paragraph{Implement states} \dots
+
+\paragraph{Register states} \dots
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=JavaScript, tabsize=4]
+export class MakeMove extends S31 {
+	render() {
+		...
+		{board.map((row, x) => (
+			board.map((col, y) => {
+				const SelectPoint = this.props.Pos('click', event => {
+					event.preventDefault()
+					return { x, y }
+				}
+				return <SelectPoint><td>.</td></SelectPoint>
+			}
+		)}
+		...
+	}
+}
+\end{lstlisting}  
+\captionof{lstlisting}{Example fragment of \textit{Noughts and Crosses} \texttt{P1} implementation.}
+\label{lst:svrprotocol}
+\end{figure}
+
 \section{Related Works}
-\paragraph{Languages with native session type support} \dots
+%\paragraph{Languages with native session type support} \dots
 % New languages written  (e.g. ATS)

### 2020-01-21 PLACES_2020/paper.tex
PLACES 2020: update author name ordering

@@
 \lstdefinelanguage{JavaScript}{
-  keywords={typeof, new, true, false, catch, function, return, null, catch, switch, var, if, in, while, do, else, case, break, class, export, boolean, throw, implements, import, this, const, let},
+  keywords={typeof, new, true, false, catch, function, return, null, catch, switch, var, if, in, while, do, else, case, break, class, export, boolean, throw, implements, import, this, const, let, extends},
   keywordstyle=\color{blue}\bfseries,
@@
 \and
-Fangyi Zhou
-\institute{Imperial College London}
-\and
 Francisco Ferreira
@@
 \institute{Imperial College London}
+\and
+Fangyi Zhou
+\institute{Imperial College London}
 }
 \def\titlerunning{MPST-safe Web Development in TypeScript}
-\def\authorrunning{A. Miu, F. Zhou, F. Ferreira \& N. Yoshida}
+\def\authorrunning{A. Miu, F. Ferreira, N. Yoshida \& F. Zhou}
 \begin{document}
@@
 
-\paragraph{Components} \dots
-% Component = basic building blocks 
-% Life cycle
-% JSX
+\paragraph{Components} A component is a reusable UI element which contains its own markup and logic.
+Components implement a \texttt{render()} function which returns a React element, the smallest building blocks of a React application.
+Components can keep \textit{state} and the \texttt{render()} function is invoked upon a change of state.
+A simple counter can be implemented as a component, with the count stored as state, a button which increments the count when clicked and a \texttt{div} that renders the current count.
+Components can also render other components, which gives rise
 
-\paragraph{Virtual DOM} \dots
-% virtual-dom
+\paragraph{Virtual DOM} React components are rendered on a logical abstraction of the DOM, which in turn performs a \texttt{diff} on the current browser DOM and patches the delta accordingly.
+This allows the programmer to directly specify what should be rendered without having to worry about which elements to append or remove from the browser DOM.
 
-\paragraph{Uni-directional data flow} \dots
+\paragraph{Uni-directional data flow} Parent components pass data to child components as \textit{props}, \dots
+
 % Parent provides props to child (uni-directional workflow)
@@
 \subsubsection{Model types in React}
-Although React was originally intended for the \textit{Model-View-Controller} architecture, we use the framework to express the model types approach.
+We leverage React to express the model types approach in \cite{MVU2019} rather than the \textit{Model-View-Controller} (MVC) architecture it was intended for.
 
@@
 \subsubsection{Affine channel usage}
+A limitation of our browser-side session type encoding is only being able to guarantee that channel resources are used \textit{at most once} as supposed to \textit{exactly once}.
 
-\paragraph{Repeat use}
+\paragraph{Repeat use} \dots
 
@@
 \label{section:example}
-We apply our framework to implement a web-based implementation of the \textit{Noughts and Crosses} running example in TypeScript; the interested reader can find the full implementation in \cite{NoughtsAndCrosses}.
-In addition to showing the multiparty session type safety achieved by our generated APIs, we also show that our library design welcomes idiomatic JavaScript practices in the user implementation and is interoperable with common front- and back-end frameworks.
 
-\subsection{Game server}
-\label{section:exampleserver}
-% TicTacToe example - backend
-
-We set up the WebSocket server on top of an Express.js \cite{ExpressJS}application running on Node.js \cite{NodeJS}.
-We define our own game logic in a \texttt{Board} class to keep track of the game state and expose methods to query the result.
-This custom logic is integrated into our
-\texttt{Receive\_P1} and \texttt{Receive\_P2} handlers (listing are defined using 
-encapsulate our custom game logic in a class
-
-\begin{figure}[!h]
+\begin{figure}
 \begin{lstlisting}[language=JavaScript, tabsize=4, title=\texttt{./app.ts}]
@@
 // Instantiate session
-const webSocketServer: WebSocket.Server = ...
-new GameServer(webSocketServer, handleP1Move, handleP2Move)
+new NoughtsAndCrosses.Svr(webSocketServer, handleP1Move, handleP2Move)
 \end{lstlisting}  
@@
 \end{figure}
+
+We apply our framework to implement a web-based implementation of the \textit{Noughts and Crosses} running example in TypeScript; the interested reader can find the full implementation in \cite{NoughtsAndCrosses}.
+In addition to showing the multiparty session type safety achieved by our generated APIs, we also show that our library design welcomes idiomatic JavaScript practices in the user implementation and is interoperable with common front- and back-end frameworks.
+
+\subsection{Game server}
+\label{section:exampleserver}
+% TicTacToe example - backend
+
+We set up the WebSocket server inside an Express.js \cite{ExpressJS} application on top of a Node.js \cite{NodeJS} runtime.
+We define our own game logic in a \texttt{Board} class to keep track of the game state and expose methods to query the result.
+This custom logic is integrated into our
+\texttt{Receive\_P1} and \texttt{Receive\_P2} handlers (listing \ref{lst:svrprotocol}) are defined using
+
+%\dots encapsulate our custom game logic in a class
 
@@
 \captionof{lstlisting}{Example fragment of \textit{Noughts and Crosses} \texttt{P1} implementation.}
-\label{lst:svrprotocol}
+\label{lst:clientapp}
 \end{figure}
+
+For the sake of sharing the same game client implementation for both players, \cite{NoughtsAndCrosses} uses \textit{higher-order components} (HOC) to build the correct state implementations depending on which player the user chooses to be.
 
@@
 \section{Conclusion and Future Works}
+We have presented a MPST-based framework for developing full-stack interactive TypeScript applications with WebSocket communications that conform to a protocol.
+
+Future works include \dots
 

### 2020-01-21 PLACES_2020/paper.tex
PLACES 2020: progress on front-end API encoding

@@
 A simple counter can be implemented as a component, with the count stored as state, a button which increments the count when clicked and a \texttt{div} that renders the current count.
-Components can also render other components, which gives rise
+Components can also render other components, which gives rise to a parent/child relationship between components.
+Parents can pass data to children as \textit{props}, so the counter component could render a  child component \texttt{<StyledDiv count=\{this.state.count\} />} and propagate the count in the state to the child to reuse an existing UI element.
 
@@
 
-\paragraph{Uni-directional data flow} Parent components pass data to child components as \textit{props}, \dots
-
-% Parent provides props to child (uni-directional workflow)
-
 \subsubsection{Model types in React}
 We leverage React to express the model types approach in \cite{MVU2019} rather than the \textit{Model-View-Controller} (MVC) architecture it was intended for.
+
+\paragraph{State} \dots
+
+\paragraph{Model transitions - message receive} \dots
+
+\paragraph{Model transitions - message send} \dots
+
+\subsubsection{Session instantiation}
+
+\dots
 
@@
 
-\paragraph{Repeat use} \dots
+\paragraph{Repeat use} Communication channels are not exposed to the programmer so multiple sends are impossible.
+This does not restrict the programmer from binding the send action to exactly one UI event: for the \textit{Noughts and Crosses} game, we bind the \texttt{Pos(Point)} send action to each unoccupied cell on the game board, but the generated runtime ensures that, once an event bound to the send action is triggered, the send is only performed once and the successor state is rendered on the DOM to prevent multiple sends initiated by the user.
 

### 2020-01-21 PLACES_2020/paper.tex
PLACES 2020: progress on front-end API encoding

@@
 Our design goal is to enforce this statically through the APIs we generate.
+
 For browser-side targets, we extend the approach presented in \cite{MVU2019} on \textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU) architecture.
@@
 Components can also render other components, which gives rise to a parent/child relationship between components.
-Parents can pass data to children as \textit{props}, so the counter component could render a  child component \texttt{<StyledDiv count=\{this.state.count\} />} and propagate the count in the state to the child to reuse an existing UI element.
+Parents can pass data to children as \textit{props}: going back to the aforementioned example, the the counter component could render a child component \texttt{<StyledDiv count=\{this.state.count\} />} in its \texttt{render()} function, propagating the \texttt{count} from its state to the child.
+This enables reusability, and for our use case, gives control to the parent on what data to pass to its children (e.g.
+pass the payload of a received message to a child to render).
 
@@
 
-\paragraph{State} \dots
+\paragraph{State} An EFSM state is encoded as an \textit{abstract} React component.
+This is an abstract class to require the programmer to provide their own view function, which translates conveniently to the \texttt{render()} function of React components.
+Our session runtime (\sectionref{section:clientruntime}) ``executes'' the EFSM and renders the current state, so upon transitioning to a successor state, the successor's view function will be invoked, as per the semantics expressed in \cite{MVU2019}.
 
-\paragraph{Model transitions - message receive} \dots
+\paragraph{Model transitions} Transitions are encoded as React component props onto the encoded states by the session runtime (\sectionref{section:clientruntime}).
+We motivate the design choice of not exposing channel resources to provide guarantees on channel usage.
+React components in TypeScript are \textit{generics} \cite{TypeScriptSpec} which are parameterised by the permitted types of prop and state, which allows us to leverage the TypeScript compiler to verify that the props for model transitions stay local to the state they are defined for.
+The model transitions for EFSMs are message send and receive.
 
-\paragraph{Model transitions - message send} \dots
+\subparagraph{Send} We make the assumption that message sending is triggered by some user-driven UI event (e.g.
+clicking a button, pressing a key on the keyboard) which interacts with some DOM element.
+We could pass a \texttt{send()} function as a prop to the sending state, but the programmer would be free to call the function multiple times which makes channel reuse possible.
+Instead, we pass a \textit{factory function} as a prop, which will, given an HTML event and a event handler function, return a fresh React component that binds the sending action on construction, so once the bound event is triggered, our session runtime executes the event handler function to obtain the payload to send, perform the send \textit{exactly once} and transition to (which, in practice, means render) the successor state.
 
-\subsubsection{Session instantiation}
+\begin{figure}[!h]
+\begin{lstlisting}[language=JavaScript, tabsize=4]
+render() { return (
+	...
+	{board.map((row, x) => (
+		board.map((col, y) => {
+			const SelectPoint = this.props.Pos('click', event => {
+				event.preventDefault()
+				return { x, y }
+			}
+			return <SelectPoint><td>.</td></SelectPoint>
+		}
+	)}
+	...	
+)}
+\end{lstlisting}  
+\captionof{lstlisting}{Example fragment of \textit{Noughts and Crosses} \texttt{P1} implementation.}
+\label{lst:clientapp}
+\end{figure}
+
+We clarify the semantics using the \textit{Noughts and Crosses} example in listing \ref{lst:clientapp}.
+\texttt{this.props.Pos} is the factory function prop passed from the session runtime.
+For each x-y coordinate on the game board, we create a \texttt{SelectPoint} React component from the factory function (which reads ``build a React component that sends the \texttt{Pos} message with x-y coordinates as payload when the user clicks on it'') and we wrap a table cell (the game board is rendered as a HTML table) inside the \texttt{SelectPoint} component to bind the click event on the table cell.
+
+\subparagraph{Receive} The payload of a message receive transition is passed as a prop to the successor state and can be handled in the successor's view function.
+Upon receiving the message from the channel, the session runtime renders the successor of the receive state and propagates the payload as the prop.
+
+\subsubsection{Session runtime}
+\label{section:clientruntime}
 
 \dots
+
+The session runtime component is rendered by the programmer and requires the following properties to instantiate the session and execute the EFSM.
+
+\paragraph{Endpoint URL}
+
+\paragraph{Concrete state components} The programmer writes their own implementation of each state (mainly to customise how the state is rendered and inject business logic into state transitions) by extending the abstract React class components.
 
@@
 
-\begin{figure}[!h]
-\begin{lstlisting}[language=JavaScript, tabsize=4]
-export class MakeMove extends S31 {
-	render() {
-		...
-		{board.map((row, x) => (
-			board.map((col, y) => {
-				const SelectPoint = this.props.Pos('click', event => {
-					event.preventDefault()
-					return { x, y }
-				}
-				return <SelectPoint><td>.</td></SelectPoint>
-			}
-		)}
-		...
-	}
-}
-\end{lstlisting}  
-\captionof{lstlisting}{Example fragment of \textit{Noughts and Crosses} \texttt{P1} implementation.}
-\label{lst:clientapp}
-\end{figure}
+The beauty of the MPST framework is that implementation is decoupled from protocol conformance, so it would be possible to have different front-end implementations for both players; but for the sake of sharing the same game client implementation between both players for demonstration purposes, \cite{NoughtsAndCrosses} uses \textit{higher-order components} (HOC) to build the correct state implementations depending on which player the user chooses to be.
 
-For the sake of sharing the same game client implementation for both players, \cite{NoughtsAndCrosses} uses \textit{higher-order components} (HOC) to build the correct state implementations depending on which player the user chooses to be.
+We also illustrate the flexibility of our approach in being interoperable with other libraries and idiomatic JavaScript practices.
+Our generated APIs provide an implementation of the communication state machine which conforms to the protocol, but \cite{NoughtsAndCrosses} also integrates with the \textit{Redux} state management library to uniformly keep track of the game state across the game client - for simplicity, we save game state into browser local storage.
 

### 2020-01-21 PLACES_2020/paper.tex
PLACES 2020: add related works and conclusion

@@
 Both players are identified by either noughts or crosses respectively, and take turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player wins (when their markers form one straight line on the board) or a stalemate is reached (when all cells are occupied).
-
-Figure \ref{fig:tictactoe} illustrates a possible implementation of the game as an interactive real-time multiplayer web application.
-Both players are connected to the game server by bidirectional WebSocket connections and interact with the game from their web browser, which serve a \textit{single-page application} (SPA) of the game client.
+A web-based implementation may involve players connected to a game server via WebSocket connections and interacting with the game from their web browser, which serve a \textit{single-page application} (SPA) of the game client written in a popular framework like \textit{React.js} \cite{React}.
 SPAs feature a single HTML page and dynamically renders content via JavaScript in the browser.
@@
 
-This MPST methodology has been applied to target applications written in both statically and dynamically typed languages in \cite{Hybrid2016, Python2017}, but do not apply to the event-driven paradigm of interactive web applications.
-A proposal that caters for web applications in \cite{PureScript2019} targets PureScript to maximise the static guarantees it can provide, but the strong type system make it difficult to extract common code and the functional paradigm is arguably unidiomatic relative to mainstream web development in JavaScript, which limits developer productivity.
-JavaScript began as the language of the web browser, but a shallow learning curve, an ever-growing ecosystem of libraries and frameworks (such as React.js for writing SPAs) and a successful server-side single-threaded event loop implementation in \textit{Node.js} has propelled it as the de-facto choice for full-stack web application development in a single language.
-
-Whilst the proposal in \cite{Python2017} could be adapted to verify interactive web applications written in JavaScript, the emergence of Microsoft's TypeScript \cite{TypeScript} with a gradual type system motivates the need for a MPST-based workflow for verifying full-stack TypeScript web applications, such that it is flexible with respect to idiomatic practices and frameworks to be useful in industry whilst also leveraging the language features to maximise the static guarantees it is able to provide.
+We see the application of the MPST methodology to verifying interactive web applications written in TypeScript to be an interesting design space - to what extent can the MPST methodology be applied such that we leverage TypeScript language features to maximise static safety guarantees whilst preserving a flexible, natural and idiomatic workflow for TypeScript developers?
 Such a workflow would ultimately decrease the overhead for incorporating MPST into mainstream web development, which reduces development time by programmatically verifying implementations for communication correctness.
@@
 
-Figure \ref{fig:workflow} illustrates our proposed development workflow: we extend the Scribble framework (\sectionref{section:scribble}) and generate TypeScript endpoints for server-side (\sectionref{section:server}) and browser-side (\sectionref{section:browser}) targets.
+%Figure \ref{fig:workflow} illustrates our proposed development workflow: we extend the Scribble framework (\sectionref{section:scribble}) and generate TypeScript endpoints for server-side (\sectionref{section:server}) and browser-side (\sectionref{section:browser}) targets.
 We will use the \textit{Noughts and Crosses} game as our running example in the rest of the paper and show how our approach is a practical compromise that combines benefits from static session typing and being compatible with common libraries and idiomatic JavaScript practices (\sectionref{section:example}).
@@
 \label{section:scribble}
+
 Development begins by specifying the permitted communications between participants as a \textit{global protocol} in Scribble, a MPST-based framework introduced in \cite{Scribble} which features a protocol specification language and code generation toolchain.
 This means that a protocol specified in Scribble maps to some \textit{global type} in multiparty session type theory \cite{MPST}.
-
-We specify the \textit{Noughts and Crosses} game as a Scribble protocol in Listing \ref{lst:game} and explain the syntax of the language through this example. 
-Scribble protocols are organised into \textit{modules}: line 1 declares the name of the module.
-Line 2 declares a message type to be used as payload types in the global protocol: this would correspond to an \texttt{import \{ Coordinate as Point \} from './Types'} statement in TypeScript, where the developer provides the definition of the \texttt{Coordinate} interface under a \texttt{Types.ts} file.
-Line 4 declares the name of the global protocol (\texttt{Game}) and specifies the participants involved in the protocol as \textit{roles}; the protocol body is defined by lines 5-16.
-A message passing statement from a sender (e.g.
-\texttt{P1}) to a receiver (e.g.
-\texttt{Svr}) require a \textit{label identifier} followed by a list of \textit{payload types} in parentheses (e.g.
-\texttt{Pos(Point) from P1 to Svr} on line 5).
-Semicolons denote continuations for message passing statement in the same way that they express sequential statement execution in programming languages such as Java.
+We specify the \textit{Noughts and Crosses} game as a Scribble protocol in Listing \ref{lst:game}.
 
@@
 
-The \texttt{Game} protocol involves three roles: the server (\texttt{Svr}) and two players (\texttt{P1} and \texttt{P2}).
-First, \texttt{P1} makes a move by sending a coordinate to place the marker on the game board to \texttt{Svr}.
-Upon receiving the \texttt{Pos} message, \texttt{Svr} will implement the game logic to decide the outcome of \texttt{P1}'s move and communicate with \texttt{P2} accordingly.
-We express this as a \texttt{choice} statement (lines 6-15), where \texttt{Svr} decides whether \texttt{P1}'s move resulted in a loss for \texttt{P2} (lines 7-8), stalemate (lines 10-11) or an unfinished game (lines 13-15).
-
-If \texttt{P1} performed a winning move (i.e.
-the latest coordinate resulted in three-in-a-row for \texttt{P1}), then \texttt{Svr} sends a \texttt{Lose} message to \texttt{P2} and a \texttt{Win} message to \texttt{P1} with the coordinates of the latest move.
-These coordinates are sent to allow the applications running in the browser for both players to update their version of the game state respectively; how they choose to render this on the DOM is up to their implementation, independent of the protocol.
-Similar message passing actions are sent in the case of a draw, just with updated labels for \texttt{Draw}.
-
-Otherwise, the game has not ended, so \texttt{Svr} updates both players with the coordinates of the latest move by sending an \texttt{Update} message, and the game proceeds to allow \texttt{P2} to make a move.
-We express this as a recursive call to the \texttt{Game} protocol using a \texttt{do} statement: note that we swap the role parameters \texttt{P2} and \texttt{P1} to express the turn-taking semantics of the game.
-In short, the global protocol describes \texttt{P1} and \texttt{P2} taking turns until \texttt{Svr} declares a winner/loser or draw.
-
 We leverage the Scribble toolchain to check for protocol well-formedness.
@@
 We obtain \textit{endpoint protocols} for each role in a well-formed global protocol.
-An endpoint protocol only preserves the interactions defined by the global protocol in which the target role is involved.
-For example, the endpoint protocol for \texttt{P1} will not include the exchange of messages between \texttt{Svr} and \texttt{P2}.
-An endpoint protocol corresponds to an equivalent \textit{Endpoint Finite State Machine} (EFSM) in the Scribble toolchain.
-We adopt the formalism of EFSMs in \cite{Hybrid2016}, and use the EFSMs as a basis for API generation.
+An endpoint protocol only preserves the interactions defined by the global protocol in which the target role is involved, and corresponds to an equivalent \textit{Endpoint Finite State Machine} (EFSM) in the Scribble toolchain.
+We use the EFSMs as a basis for API generation and adopt the formalisms in \cite{Hybrid2016}.
 
@@
 Our encodings sufficiently exposes seams for the developer to inject their business logic, whilst the generated session API (\sectionref{section:serversessionapi}) handles the sending and receiving of messages; as a result, the encodings do not concern with the role involved in the send/receive action.
-We outline the encodings below using examples from the \textit{Noughts and Crosses} game server (listing \ref{lst:svr}).
+We outline the encodings below using examples from the \textit{Noughts and Crosses} game server (Listing \ref{lst:svr}).
 
@@
 \begin{figure}[!h]
-\begin{lstlisting}[language=JavaScript, title=\texttt{./NoughtsAndCrosses/Game.ts}]
+\begin{lstlisting}[language=JavaScript]
 import { Coordinate as Point } from "./Types";
@@
 Upon construction, the Protocol class waits for all client roles to connect via WebSocket, then proceeds to the initial state of the EFSM and performs state transitions depending on the permitted communication patterns from the EFSM.
-The Protocol class constructor requires the WebSocket server and the following handlers:
-
-\paragraph{Initial state} When all roles have joined the session, the Protocol class needs to know what to do in the initial state to start communication, be it sending a message to a client or offering branches to a client.
-For the \textit{Noughts and Crosses} game server, this would be a callback to handle the receipt of the \texttt{Pos(Point)} message from \texttt{P1}.
-
-\paragraph{Receive handlers for each role} The runtime needs to bind a message event listener to the WebSocket between the server and each (client) role, and needs to know how to handle the incoming messages.
-For each role, the user is required to pass callbacks that encode the receive states for that role.
-The type alias for such a handler is the \textit{intersection type} \cite{TypeScriptSpec} of all receive states concerning that role.
-Considering a hypothetical web-service providing addition and square root functionality to a client role \texttt{C}, if \texttt{S1} is a state expecting to receive \texttt{Add(int, int)} from \texttt{C} and \texttt{S4} is a state expecting to receive \texttt{Sqrt(int)} from \texttt{C}, we define \texttt{type Receive_C = S1 \& S4} as the receive handler type alias and the user needs to pass some value of \texttt{Receive_C} when instantiating the session through the Protocol API.
 
@@
 
-\paragraph{Repeat use} By adopting a callback-oriented design for server-side API generation, channels are not directly accessed by the programmer which makes \textit{reuse} impossible; only the Protocol API is exposed but channel operations are concealed in a private class.
-Consider the \textit{Noughts and Crosses} server endpoint: the programmer must pass a callback that handles the receipt of a \texttt{Pos(Point)} message and returns either a \texttt{Lose(Point)}, \texttt{Draw(Point)} or \texttt{Update(Point)} message with the appropriate continuation.
-Upon receiving the \texttt{Pos(Point)} message from a player, our lightweight runtime invokes the callback to get \texttt{Svr}'s choice from the return value, and perform the appropriate send.
-This is transparent to the programmer and it is impossible for the programmer to send or receive messages more than once, by design.
+\paragraph{Repeat use} We do not expose channels to the programmer, which makes \textit{reuse} impossible.
+For example, to send a message, the generated API only requires the payload that needs to be sent, and the session runtime performs the send internally, guaranteeing this action is done \textit{exactly once} by construction.
 
@@
 \label{section:browser}
+
+We will refer to the \texttt{P1} EFSM (figure \ref{fig:efsmp1}) as a running example in this section.
+Preserving behavioural typing and channel linearity are known to be challenging for browser-side applications due to the event-driven nature of user interaction: in the case of \textit{Noughts and Crosses}, once the user makes a move by clicking on a cell on the game board, this click event must be deactivated until the user's next turn, otherwise the user can click again and violate channel linearity.
+Our design goal is to enforce this statically through the APIs we generate.
+
+For browser-side targets, we extend the approach presented in \cite{MVU2019} on \textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU) architecture.
+Each state in the EFSM is a model type and uniquely defines a \textit{view function}, set of \textit{messages} and \textit{update function}: the view function defines what to render to the DOM, the set of messages define the possible (IO) actions available at that state, and the update function defines which successor state to transition to, given some supported IO action at this state.
 
@@
 \end{wrapfigure}
-
-We will refer to the \texttt{P1} EFSM (figure \ref{fig:efsmp1}) as a running example in this section.
-Preserving behavioural typing and channel linearity are known to be challenging for browser-side applications due to the event-driven nature of user interaction: in the case of \textit{Noughts and Crosses}, once the user makes a move by clicking on a cell on the game board, this click event must be deactivated until the user's next turn, otherwise the user can click again and violate channel linearity.
-Our design goal is to enforce this statically through the APIs we generate.
-
-For browser-side targets, we extend the approach presented in \cite{MVU2019} on \textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU) architecture.
-Each state in the EFSM is a model type and uniquely defines a \textit{view function}, set of \textit{messages} and \textit{update function}: the view function defines what to render to the DOM, the set of messages define the possible (IO) actions available at that state, and the update function defines which successor state to transition to, given some supported IO action at this state.
 
@@
 
-\paragraph{Virtual DOM} React components are rendered on a logical abstraction of the DOM, which in turn performs a \texttt{diff} on the current browser DOM and patches the delta accordingly.
+\paragraph{Virtual DOM (VDOM)} React components are rendered on a logical abstraction of the DOM, which in turn performs a \texttt{diff} on the current browser DOM and patches the delta accordingly.
 This allows the programmer to directly specify what should be rendered without having to worry about which elements to append or remove from the browser DOM.
@@
 \begin{lstlisting}[language=JavaScript, tabsize=4]
-render() { return (
+render() {
 	...
@@
 		board.map((col, y) => {
-			const SelectPoint = this.props.Pos('click', event => {
+			const SelectPoint = this.props.Pos('click', (event: UIEvent) => {
 				event.preventDefault()
@@
 	...	
-)}
+}
 \end{lstlisting}  
-\captionof{lstlisting}{Example fragment of \textit{Noughts and Crosses} \texttt{P1} implementation.}
+\captionof{lstlisting}{Model transition for message sending in 
+\textit{Noughts and Crosses} \texttt{P1} implementation.}
 \label{lst:clientapp}
@@
 
-We clarify the semantics using the \textit{Noughts and Crosses} example in listing \ref{lst:clientapp}.
+We clarify the semantics using the \textit{Noughts and Crosses} example in Listing \ref{lst:clientapp}.
 \texttt{this.props.Pos} is the factory function prop passed from the session runtime.
@@
 \subparagraph{Receive} The payload of a message receive transition is passed as a prop to the successor state and can be handled in the successor's view function.
-Upon receiving the message from the channel, the session runtime renders the successor of the receive state and propagates the payload as the prop.
+Upon receiving the message from the channel, the session runtime renders the successor of the receive state and propagates the payload as the prop accordingly.
 
@@
 
-\dots
+Our session runtime can be interpreted as an abstraction on top of the React VDOM that implements the EFSM by construction.
+The session runtime itself is a React component too, named after the endpoint role identifier: it opens the WebSocket connection to the server, keeps track of the current EFSM state as part of its React component state, and most importantly, renders the React component encoding of the active EFSM state.
+Channel communications are managed by the runtime, which allows it to render the successor of a receive state upon receiving a message from the channel.
+Similarly, the session runtime is responsible for passing the required props for model transitions to EFSM state React components.
+The session runtime component is rendered by the programmer and requires the following props:
 
-The session runtime component is rendered by the programmer and requires the following properties to instantiate the session and execute the EFSM.
-
-\paragraph{Endpoint URL}
+\paragraph{Endpoint URL} The session runtime needs this in order to open the WebSocket connection.
 
 \paragraph{Concrete state components} The programmer writes their own implementation of each state (mainly to customise how the state is rendered and inject business logic into state transitions) by extending the abstract React class components.
-
-% State = React component
-% Send transitions = model transition factory
-% Receive transitions = handled by 
-% Protocol component = handle joining and rendering of correct EFSM component
+The session runtime requires references to these concrete components order to render the user implementation accordingly.
 
@@
 This means that it is possible for an implementation to \textit{not} handle transitions to a terminal state but still type-check, so we cannot prevent unused states.
-Consider a hypothetical protocol where, in the current state, sending \texttt{Quit()} to some server role \texttt{S} transitions to the terminal state: \dots
 
@@
 \begin{figure}
-\begin{lstlisting}[language=JavaScript, tabsize=4, title=\texttt{./app.ts}]
+\begin{lstlisting}[language=JavaScript, tabsize=4]
 const handleP1Move: Receive_P1 = (move: Point) => {
@@
 We apply our framework to implement a web-based implementation of the \textit{Noughts and Crosses} running example in TypeScript; the interested reader can find the full implementation in \cite{NoughtsAndCrosses}.
-In addition to showing the multiparty session type safety achieved by our generated APIs, we also show that our library design welcomes idiomatic JavaScript practices in the user implementation and is interoperable with common front- and back-end frameworks.
+In addition to MPST-safety, we show that our library design welcomes idiomatic JavaScript practices in the user implementation and is interoperable with common front- and back-end frameworks.
 
@@
 We define our own game logic in a \texttt{Board} class to keep track of the game state and expose methods to query the result.
-This custom logic is integrated into our
-\texttt{Receive\_P1} and \texttt{Receive\_P2} handlers (listing \ref{lst:svrprotocol}) are defined using
-
-%\dots encapsulate our custom game logic in a class
-
-% Note: no state space explosion, allows for reusability
+This custom logic is integrated into our \texttt{Receive\_P1} and \texttt{Receive\_P2} handlers (Listing \ref{lst:svrprotocol}), so the session runtime can handle \texttt{Pos(Point)} messages from players and transition according to the injected game logic.
+Note that, by the structural typing nature of TypeScript, replacing \texttt{handleP2Move} on line 6 with \texttt{handleP1Move} would be type-correct - this allows for better code reuse as supposed to defining additionl abstractions to work around the limitations of nominal typing in \cite{Hybrid2016}.
+There is also full type erasure when transpiling to JavaScript to run the server code, so state space explosions is never a runtime consideration.
 
@@
 
-\paragraph{Implement states} \dots
-
-\paragraph{Register states} \dots
-
-The beauty of the MPST framework is that implementation is decoupled from protocol conformance, so it would be possible to have different front-end implementations for both players; but for the sake of sharing the same game client implementation between both players for demonstration purposes, \cite{NoughtsAndCrosses} uses \textit{higher-order components} (HOC) to build the correct state implementations depending on which player the user chooses to be.
-
-We also illustrate the flexibility of our approach in being interoperable with other libraries and idiomatic JavaScript practices.
-Our generated APIs provide an implementation of the communication state machine which conforms to the protocol, but \cite{NoughtsAndCrosses} also integrates with the \textit{Redux} state management library to uniformly keep track of the game state across the game client - for simplicity, we save game state into browser local storage.
+We implement the game client for players as a React application - extending from the generated abstract React components, registering our custom implementation as props for the session runtime component.
+For the sake of code reuse for demonstration purposes, \cite{NoughtsAndCrosses} uses \textit{higher-order components} (HOC) to build the correct state implementations depending on which player the user chooses to be.
+We leverage the \textit{Redux} state management library to keep track of game state, thus showing the flexibility of our library design in being interoperable with other libraries and idiomatic JavaScript practices.
 
@@
 
-\paragraph{Endpoint API generation} \dots
-% API generation
-% - original Python implementation, generating monitors (Rumi's paper)
-%	- Gradual type system lets us do better than just dynamic runtime checks
-%	- could do the same approach with generating JavaScript monitors, but we have TS and Dart and can leverage more powerful type systems to do more static/compiler checks
+\paragraph{Endpoint API generation} \cite{Python2017} targeted Python applications and the generation of runtime monitors to dynamically verify communication patterns.
+Whilst the same approach could be applied to JavaScript, we can provide more static guarantees with TypeScript's gradual typing system and compiler.
+\cite{Hybrid2016} targeted Java applications and proposed to encode the EFSM states and transitions as classes and instance methods respectively, with behavioural typing achieved statically by the type system and channel linearity guarantees achieved dynamically since channel resources are exposed and aliasing is not monitored.
 
-% - Java (hybrid approach)
-%	- leverage strongly typed system for static behavioural typing
-%	- minimse dupliction by defining states in terms of IO interfaces (same IO state with different successor can still share code)
-%	- runtime linearity checks
+\paragraph{Session types in web development} \cite{PureScript2019} targeted web development in PureScript using the \textit{Concur UI} framework and proposed a type-level encoding of EFSMs as multi-parameter type classes.
+However, it presents a trade-off between achieving static linearity guarantees from the type-level EFSM encoding under the expressive type system and providing an intuitive development experience to programmers, especially given the prevalence of JavaScript and TypeScript applications in industry.
+\cite{MVU2019} focused on applying binary session types in front-end web development and presented approaches that tackle the challenge of guaranteeing linearity in the event-driven environment.
 
-\paragraph{Session types in web development} \dots
-% Session types in web development
-% - purescript looking at the whole workflow
-	% - problem: PureScript isn't as prevalent
-% - simon fowler paper on MVU targeting linearity for frontend GUI
-
-\paragraph{Typestate programming} \dots
-% Typestates (Mungo)
-% - the React components we generate is essentially behaving like typestates of the role, where 
-% each role has its set of permitted actions.
+Our work applies the aforementioned approaches in a \textit{multiparty} context in a manner that is compatible with industrial tools and practices to ultimately invite the notion of MPST-safety into interactive web application development in industry.
 
 \section{Conclusion and Future Works}
-We have presented a MPST-based framework for developing full-stack interactive TypeScript applications with WebSocket communications that conform to a protocol.
+We have presented a MPST-based framework for developing full-stack interactive TypeScript applications with WebSocket communications that conform to a protocol, statically providing linear channel usage guarantees and affine channel usage guarantees for back-end and front-end targets respectively.
 
-Future works include \dots
+Future works include incorporating \textit{explicit connection actions} introduced in \cite{ExplicitConnections} in our API generation to better model real-world communication protocols that may feature in interactive web applications.
+Whilst our approach does support multiparty sessions, protocols are required to have exactly one role running on the server and client roles must interact with each other via the server to respect the properties of WebSocket connections, which is the transport abstraction we support.
+Extending support to WebRTC would cater for peer-to-peer communication between browsers, which further opens up possibilities for communication protocols supported by our approach.
 
@@
 
-%\section{Introduction}
-%% Emergence of interactive JavaScript-intensive applications that maintain continuous stateful
-%
-%The optional arguments of {\tt $\backslash$documentclass$\{$eptcs$\}$} are
-%\begin{itemize}
-%\item at most one of
-%{\tt adraft},
-%{\tt submission} or
-%{\tt preliminary},
-%\item at most one of {\tt publicdomain} or {\tt copyright},
-%\item and optionally {\tt creativecommons},
-%  \begin{itemize}
-%  \item possibly augmented with
-%    \begin{itemize}
-%    \item {\tt noderivs}
-%    \item or {\tt sharealike},
-%    \end{itemize}
-%  \item and possibly augmented with {\tt noncommercial}.
-%  \end{itemize}
-%\end{itemize}
-%We use {\tt adraft} rather than {\tt draft} so as not to confuse hyperref.
-%The style-file option {\tt submission} is for papers that are
-%submitted to {\tt $\backslash$event}, where the value of the latter is
-%to be filled in in line 2 of the tex-file. Use {\tt preliminary} only
-%for papers that are accepted but not yet published. The final version
-%of your paper that is to be uploaded at the EPTCS website should have
-%none of these style-file options.
-%
-%By means of the style-file option
-%\href{http://creativecommons.org/about/license/}{creativecommons}
-%authors equip their paper with a Creative Commons license that allows
-%everyone to copy, distribute, display, and perform their copyrighted
-%work and derivative works based upon it, but only if they give credit
-%the way you request. By invoking the additional style-file option {\tt
-%noderivs} you let others copy, distribute, display, and perform only
-%verbatim copies of your work, but not derivative works based upon
-%it. Alternatively, the {\tt sharealike} option allows others to
-%distribute derivative works only under a license identical to the
-%license that governs your work. Finally, you can invoke the option
-%{\tt noncommercial} that let others copy, distribute, display, and
-%perform your work and derivative works based upon it for
-%noncommercial purposes only.
-%
-%Authors' (multiple) affiliations and emails use the commands
-%{\tt $\backslash$institute} and {\tt $\backslash$email}.
-%Both are optional.
-%Authors should moreover supply
-%{\tt $\backslash$titlerunning} and {\tt $\backslash$authorrunning},
-%and in case the copyrightholders are not the authors also
-%{\tt $\backslash$copyrightholders}.
-%As illustrated above, heuristic solutions may be called for to share
-%affiliations. Authors may apply their own creativity here.
-%
-%Exactly 46 lines fit on a page.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\hfill6\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\hfill11\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%
-%Here starts a new paragraph. The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\hfill16\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\hfill21\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\hfill26\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\hfill31\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\hfill36\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\hfill41\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.\hfill46\\
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%The rest is like any normal {\LaTeX} article.
-%We will spare you the details.
-%
-%\section{Prefaces}
-%
-%Volume editors may create prefaces using this very template,
-%with {\tt $\backslash$title$\{$Preface$\}$} and {\tt $\backslash$author$\{\}$}.
-%
-%\section{Bibliography}
-%
-%We request that you use
-%\href{http://www.cse.unsw.edu.au/~rvg/EPTCS/eptcs.bst}
-%{\tt $\backslash$bibliographystyle$\{$eptcs$\}$}
-%\cite{bibliographystylewebpage}. Compared to the original {\LaTeX}
-%{\tt $\backslash$biblio\-graphystyle$\{$plain$\}$},
-%it ignores the field {\tt month}, and uses the extra
-%bibtex fields {\tt eid}, {\tt doi}, {\tt ee} and {\tt url}.
-%The first is for electronic identifiers (typically the number $n$
-%indicating the $n^{\rm th}$ paper in an issue) of papers in electronic
-%journals that do not use page numbers. The other three are to refer,
-%with life links, to electronic incarnations of the paper.
-%
-%Almost all publishers use digital object identifiers (DOIs) as a
-%persistent way to locate electronic publications. Prefixing the DOI of
-%any paper with {\tt http://dx.doi.org/} yields a URI that resolves to the
-%current location (URL) of the response page\footnote{Nowadays, papers
-%  that are published electronically tend
-%  to have a \emph{response page} that lists the title, authors and
-%  abstract of the paper, and links to the actual manifestations of
-%  the paper (e.g.\ as {\tt dvi}- or {\tt pdf}-file). Sometimes
-%  publishers charge money to access the paper itself, but the response
-%  page is always freely available.}
-%of that paper. When the location of the response page changes (for
-%instance through a merge of publishers), the DOI of the paper remains
-%the same and (through an update by the publisher) the corresponding
-%URI will then resolve to the new location. For that reason a reference
-%ought to contain the DOI of a paper, with a life link to corresponding
-%URI, rather than a direct reference or link to the current URL of
-%publisher's response page. This is the r\^ole of the bibtex field {\tt doi}.
-%DOIs of papers can often be found through
-%\url{http://www.crossref.org/guestquery};\footnote{For papers that will appear
-%  in EPTCS and use \href{http://www.cse.unsw.edu.au/~rvg/EPTCS/eptcs.bst}
-%  {\tt $\backslash$bibliographystyle$\{$eptcs$\}$} there is no need to
-%  find DOIs on this website, as EPTCS will look them up for you
-%  automatically upon submission of a first version of your paper;
-%  these DOIs can then be incorporated in the final version, together
-%  with the remaining DOIs that need to found at DBLP or publisher's webpages.}
-%the second method {\it Search on article title}, only using the {\bf
-%surname} of the first-listed author, works best.  
-%Other places to find DOIs are DBLP and the response pages for cited
-%papers (maintained by their publishers).
-%{\bf EPTCS requires the inclusion of a DOI in each cited paper, when available.}
-%
-%Often an official publication is only available against payment, but
-%as a courtesy to readers that do not wish to pay, the authors also
-%make the paper available free of charge at a repository such as
-%\url{arXiv.org}. In such a case it is recommended to also refer and
-%link to the URL of the response page of the paper in such a
-%repository.  This can be done using the bibtex fields {\tt ee} or {\tt
-%url}, which are treated as synonyms.  These fields should not be used
-%to duplicate information that is already provided through the DOI of
-%the paper.
-%You can find archival-quality URL's for most recently published papers
-%in DBLP---they are in the bibtex-field {\tt ee}. In fact, it is often
-%useful to check your references against DBLP records anyway, or just find
-%them there in the first place.
-%
-%When using {\LaTeX} rather than {\tt pdflatex} to typeset your paper, by
-%default no linebreaking within long URLs is allowed. This leads often
-%to very ugly output, that moreover is different from the output
-%generated when using {\tt pdflatex}. This problem is repaired when
-%invoking \href{http://www.cse.unsw.edu.au/~rvg/EPTCS/breakurl.sty}
-%{\tt $\backslash$usepackage$\{$breakurl$\}$}: it allows linebreaking
-%within links and yield the same output as obtained by default with
-%{\tt pdflatex}. 
-%When invoking {\tt pdflatex}, the package {\tt breakurl} is ignored.
-%
 \nocite{*}

### 2020-01-21 PLACES_2020/paper.tex
PLACES 2020: remove extra subsections

@@
 
-\paragraph{Repeat use} Communication channels are not exposed to the programmer so multiple sends are impossible.
+Communication channels are not exposed to the programmer so multiple sends are impossible.
 This does not restrict the programmer from binding the send action to exactly one UI event: for the \textit{Noughts and Crosses} game, we bind the \texttt{Pos(Point)} send action to each unoccupied cell on the game board, but the generated runtime ensures that, once an event bound to the send action is triggered, the send is only performed once and the successor state is rendered on the DOM to prevent multiple sends initiated by the user.
 
-\paragraph{Unused} Our approach \textit{does not} statically detect whether all transitions in a certain state are bound to some UI event.
+However, our approach \textit{does not} statically detect whether all transitions in a certain state are bound to some UI event.
 This means that it is possible for an implementation to \textit{not} handle transitions to a terminal state but still type-check, so we cannot prevent unused states.
@@
 
-\subsection{Game server}
-\label{section:exampleserver}
-% TicTacToe example - backend
-
-We set up the WebSocket server inside an Express.js \cite{ExpressJS} application on top of a Node.js \cite{NodeJS} runtime.
+\paragraph{Game server} We set up the WebSocket server inside an Express.js \cite{ExpressJS} application on top of a Node.js \cite{NodeJS} runtime.
 We define our own game logic in a \texttt{Board} class to keep track of the game state and expose methods to query the result.
 This custom logic is integrated into our \texttt{Receive\_P1} and \texttt{Receive\_P2} handlers (Listing \ref{lst:svrprotocol}), so the session runtime can handle \texttt{Pos(Point)} messages from players and transition according to the injected game logic.
+
 Note that, by the structural typing nature of TypeScript, replacing \texttt{handleP2Move} on line 6 with \texttt{handleP1Move} would be type-correct - this allows for better code reuse as supposed to defining additionl abstractions to work around the limitations of nominal typing in \cite{Hybrid2016}.
@@
 
-\subsection{Game clients}
-% TicTacToe example - frontend
+\paragraph{Game clients} We implement the game client for players as a React application - extending from the generated abstract React components, registering our custom implementation as props for the session runtime component.
 
-We implement the game client for players as a React application - extending from the generated abstract React components, registering our custom implementation as props for the session runtime component.
 For the sake of code reuse for demonstration purposes, \cite{NoughtsAndCrosses} uses \textit{higher-order components} (HOC) to build the correct state implementations depending on which player the user chooses to be.

### 2020-01-21 PLACES_2020/paper.tex
PLACES 2020: check TexMaker pdflatex functionality
PLACES 2020: complete first draft

@@
 \begin{abstract}
-% Situation: webapps, interactive, JavaScript more powerful but remains a poor language for maintaing large applications
+Advancements in mobile device compute power has made interactive web applications possible, allowing the web browser to render content dynamically and support low-latency communication with the server.
+This comes at a cost to the developer, who now need to reason more about the communication correctness of their application as they support more complex communication patterns.
 
-% Problem: hard to maintain/test
-% New languages (like Dart and TypeScript)
+Multiparty session types (MPST) provide a framework for verifying implementations to conform to a prescribed communication protocol.
+Existing proposals for applying the MPST framework in application development either neglect the event-driven nature of web applications, or lack compatibility with industry tools and practices which discourage mainstream adoption by web developers.
 
-% Solution: MPST for web development
-
-This is a sentence in the abstract.
-This is another sentence in the abstract.
-This is yet another sentence in the abstract.
-This is the final sentence in the abstract.
-
-% Problem: similar to PureScript paper
-
-% In this paper, we present...
-
-% TypeScript
-% - gradual type system, superset of JavaScript
-% - we define how our approach achieves "gradual linearity" (can we define this?)
-%Our work targets TypeScript - a superset of \dots
-
-%We demonstrate the effectiveness of our approach through a web-based game of \textit{Noughts and Crosses} where communication is performed through WebSocket connections.
+In this paper, we present an implementation of the MPST framework for developing interactive web applications using familiar industry tools in TypeScript and the \textit{React.js} framework.
+The global communication protocol is specified in the Scribble protocol language and checked for well-formedness by the Scribble toolchain.
+We obtain the local protocol for each role, which describes the interactions of the global communication protocols observed by the role.
+We encode the local protocol into TypeScript types, catering for server-side and client-side targets separately, and show that our encoding guarantees that only implementations which conform to the protocol will type-check.
+We demonstrate the effectiveness of our approach through a web-based implementation of the classic \textit{Noughts and Crosses} game from a MPST formalism of the game logic.
 
@@
 
-% Example: naughts and crosses
+% Example: noughts and crosses
 Consider a classic turn-based board game of \textit{Noughts and Crosses} between two players.
@@
 
-\subsubsection{Session Runtime and Protocol API}
+\subsubsection{Session runtime}
 \label{section:serversessionapi}
 
-Our session runtime performs communication in a way that conforms to the protocol specification.
+Our session runtime performs communication in a way that conforms to the protocol specification, but does not expose these IO actions to the developer by delegating the aforementioned responsibilities to an inner class.
 The runtime listens to message (receive) events on the communication channel, invokes the corresponding callback to obtain the value to send next, and performs the send.
-
-The \textit{Protocol API} is a class that interfaces between the session runtime and the user implementation of the EFSM state encodings.
-Upon construction, the Protocol class waits for all client roles to connect via WebSocket, then proceeds to the initial state of the EFSM and performs state transitions depending on the permitted communication patterns from the EFSM.
+The developer instantiates the session by constructing an instance of the session runtime class.
 
@@
 
-\paragraph{Unused} The initial state must be supplied to the Protocol API constructor in order to instantiate the session; this initial state is defined in terms of the successor states, which in turn has references to its successors and so forth.
+\paragraph{Unused} The initial state must be supplied to the session runtime constructor in order to instantiate the session; this initial state is defined in terms of the successor states, which in turn has references to its successors and so forth.
 This encoding approach will cover the terminal state (if it exists in the EFSM, as \cite{Hybrid2016} notes that an EFSM contains at most one terminal state), and the session runtime guarantees this terminal state, if it exists, will be reached by construction.

### 2020-01-22 interim/interim.tex
Start setting up interim scaffolding
Interim write-up: session type overview
Interim write-up: session type pi calculus
Interim write-up:  start on Scribble
Interim: complete project plan

@@
 
+\usepackage[nottoc]{tocbibind}
+\setcitestyle{numbers}
+
+\input{custom_includes}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\doublespacing
+%\doublespacing
 \tableofcontents
-\singlespacing
+%\singlespacing
 
@@
 
+% Rise of distributed programs -> main challenges
+
+% How to ensure correctness in general: type system, data types
+
+% Type discipline for concurrent programs - behavioural types (--> session types)
+
+% Mainstream example of distirbuted programs: web services (microservice architecture), interactive web application;
+
+% Objective:
+
 \chapter{Background}
 
+\input{sections/background/session_types.tex}
+
+\input{sections/background/scribble.tex}
+\clearpage
+
+\input{sections/background/codegen.tex}
+\clearpage
+
+\input{sections/background/typescript.tex}
+
 \chapter{Project Plan}
-\section{Theory}
-\section{Implementation}
+We aim to develop a multiparty session type-safe development workflow for building interactive full-stack TypeScript applications that conform to a communication protocol.
+This involves encoding session types using the TypeScript language and implementing a code generation workflow to generate the encodings.
+
+%As we target full-stack applications, we tackle server-side and browser-side targets separately:
+%
+%\paragraph{Server-side targets} We will analyse the approaches presented in \cite{Hybrid2016, PureScript2019, Python2017} and assess their applicability with respect to the language features and type system offered by TypeScript and general back-end web development patterns.
+At the point of writing, we have prototyped several approaches for server-side session type TypeScript encodings and will compare their effectiveness through implementing example protocols, such as the \textit{Adder} example in \cite{Hybrid2016} and the \textit{Battleships} example in \cite{PureScript2019}.
+%
+%\paragraph{Browser-side targets} We will analyse how \cite{PureScript2019} handles API generation for front-end implementations, and in particular, study their approach to behavioural typing and guaranteeing channel linearity in an event-driven environment.
+\cite{MVU2019} also presents an architecture for verifying binary session type conformance of fornt-end implementations through formalisms of the \textit{Model-View-Update} (MVU) architecture and \textit{model types}.
+We will investigate how to extend their approach for multiparty session types using TypeScript.
+At the point of writing, we have explored methods to express the methodology in \cite{MVU2019} using React.js for a multiplayer game.
+
+\section{Delivery}
+At the point of writing, we have analysed the existing code generation approaches presented in \cite{Hybrid2016, Scribble, Python2017} and assessed their applicability with respect to verifying the communication aspects of interactive web applications.
+We have also explored approaches for encoding session types into TypeScript, motivated by existing works along with similar proposals specific to web development in \cite{PureScript2019, MVU2019}.
+
+We plan to deliver our code generation implementation incrementally, aiming to get a working version of the end-to-end workflow first, then iteratively add support for more complex session type primitives.
+This minimises the risks involved in the project by ensuring we have a functional deliverable at the early stages of the project.
+We plan to deliver this basic end-to-end workflow by the end of February.
+
+The subsequent months will involve adding support for other primitives (i.e.
+selection, choice and recursion) for multiparty sessions.
+We plan to measure our progress by writing example protocols for these milestones and checking that our implementation supports those protocols by the end of each month.
+
+We will also try to connect with web developers in the community to experiment with our implementation to get feedback on ways to make it more applicable and compatible with industry practices, such that we can apply their feedback on the next iterations of our deliverables.
+
+If time permits, we may explore possible extensions of applying supporting \textit{Explicit Connection Actions} presented in \cite{FASE2017} in our workflow to support more protocols.
+
 \section{Timetable}
+We attach a preliminary timetable (Table \ref{table:timetable}) for guidance.
+
+\begin{figure}
+\centering
+\begin{tabular}{l|p{0.4\textwidth}|p{0.4\textwidth}}
+Month & Milestones & Deadlines \\
+\hline\hline
+Jan & \begin{itemize}
+\item Explore approaches for TypeScript encoding
+\item Investigate methods for code generation from EFSM
+\end{itemize} & \begin{itemize}
+\item 24th: Interim Report
+\item 25th: PLACES 2020
+\end{itemize} \\
+\hline
+Feb & \begin{itemize}
+\item Develop API generation end-to-end toolchain
+\item Support binary sessions with simple send/receive
+\end{itemize} & \begin{itemize}
+\item 14th: Project Review
+\end{itemize} \\
+\hline
+Mar & \begin{itemize}
+\item Support multiparty sessions with simple send/receive
+\end{itemize} & \begin{itemize}
+\item 16th-20th: Examinations
+\end{itemize}\\
+\hline
+Apr & \begin{itemize}
+\item Support binary sessions with selection, choice and recursion
+\item Example: calculator service
+\end{itemize} & \begin{itemize}
+\item 6th-17th: Easter break
+\end{itemize}\\
+\hline
+May & \begin{itemize}
+\item Support multiparty sessions with selection, choice and recursion
+\item Example: Tic Tac Toe game
+\end{itemize} & \begin{itemize}
+\item 15th: Health check-up
+\end{itemize}\\
+\hline
+Jun & \begin{itemize}
+\item Complete report write-up
+\end{itemize} & \begin{itemize}
+\item 15th: Final Report
+\item 26th: Final Archive
+\end{itemize}
+\end{tabular}
+\captionof{table}{Project Timetable.}
+\label{table:timetable}
+\end{figure}
 
@@
 
+\bibliographystyle{acm}
+\bibliography{interim}
+
 \end{document}

### 2020-01-22 interim/sections/background/codegen.tex
Update .bib for papers read
Interim: complete project plan

@@
 \item PureScript implementation that addresses the 2 criteria statically
-\item \begin{itemize}
+\begin{itemize}
 \item Relevance for web development
@@
 
-PureScript - \cite{PS2019}
+PureScript - \cite{PureScript2019}
 
 \subsection{Runtime Monitors}
+\begin{itemize}
+\item Python dynamic runtime monitors that addresses the 2 criteria dynamically
+\begin{itemize}
+\item Channel abstraction
+\item How invitation/conversation messages are handled
+\end{itemize}
+\end{itemize}
+
+Python - \cite{Python2017}
 
@@
 
-Java - \cite{HY2016}
+Java - \cite{Hybrid2016}
 
@@
 \end{itemize}
+
+\begin{figure}[!h]
+\centering
+\begin{tabular}{l p{0.4\textwidth} p{0.4\textwidth}}
+Language & Behavioural typing & Channel linearity \\
+\hline\hline
+Java \cite{Hybrid2016} &  & \\
+\hline
+Python \cite{Python2017} &  & \\
+\hline
+TypePureScript \cite{PureScript2019} &  & \\
+\end{tabular}
+\captionof{table}{Comparison between existing MPST code generation approaches.}
+\label{table:comparison}
+\end{figure}

### 2020-01-22 interim/interim.tex
Interim: begin skeleton of evaluation plan
Interim: complete evaluation plan

@@
 \input{sections/background/scribble.tex}
-\clearpage
 
 \input{sections/background/codegen.tex}
-\clearpage
 
@@
 \chapter{Evaluation Plan}
-\section{Theory}
-\section{Implementation}
+
+\section{Correctness}
+% Qualitatively argue how the approach achieves behavioural typing and preserves channel linearity
+We will qualitatively argue how our TypeScript encodings of session types achieve behavioural typing and preserves channel linearity.
+
+% May explore formalisms if time permits
+If time permits, we will explore possible formalisms of our implementation as a calculus upon which to prove correctness.
+This may be related to the formal language presented in \cite{UnderstandingTypeScript}, but we will need to extend this to be specific to our TypeScript encodings of session types.
+
+\section{Productivity} 
+% Compare with PureScript implementation to talk about developer productivity when using PureScript APIs with ConcurUI compared to TypeScript APIs when using React
+By adopting tools and practices used in industry, we wish to assess the extent to which our implementation boosts developer productivity by providing APIs that guarantee communication protocol conformance in a way that is compatible with existing workflows.
+Comparing the development workflow of implementing the \textit{Battleships} game using the PureScript workflow in \cite{PureScript2019} and our TypeScript-based proposal would yield interesting findings.
+
+% Survey - qualitative and quantitative feedback
+This can be carried out as a survey to be completed by web developers familiar with both languages.
+We would include questions that allow us to collect both qualitative and quantitative feedback to make effective comparisons between our approach and existing proposals.
+
+\section{Performance} Examples of performance metrics include: lines of code to be written by the developer (as a result of using the generated APIs), the size of the transpiled JavaScript assets to be loaded on the server and client browsers.
+We aim to extract these metrics from our approach and compare them against metrics extracted from a baseline implementation (i.e.
+without using the MPST methodology) and evaluate the performance overhead (if any) of our MPST-based approach.
 

### 2020-01-22 interim/interim.tex
Interim: add introduction and lstlisting styles from PLACES

@@
 \chapter{Introduction}
+% Rise of distributed programs -> main challenges
+% How to ensure correctness in general: type system, data types
+% Type discipline for concurrent programs - behavioural types (--> session types)
+% Mainstream example of distirbuted programs: web services (microservice architecture), interactive web application; 
+% Objective:
 
-% Rise of distributed programs -> main challenges
+Modern interactive web applications aim to provide a highly responsive user experience by minimising the communication latency between clients and servers. 
+Whilst the HTTP request-response model is sufficient for retrieving static assets, applying the same stateless communication approach for interactive use cases (such as a multiplayer game) introduces undesirable performance overhead from having to frequently set up new connections for client-server interactions.
+Developers have since adopted other communication transport abstractions over HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy low-latency full-duplex client-server communication in their applications over a single persistent connection.
+Enabling more complex communication patterns caters for more interactive use cases (such as real-time multiplayer games), but introduces additional concerns to the developer with respect to implementation correctness.
 
-% How to ensure correctness in general: type system, data types
+% Example: noughts and crosses
+Consider a classic turn-based board game of \textit{Noughts and Crosses} between two players.
+Both players are identified by either noughts or crosses respectively, and take turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player wins (when their markers form one straight line on the board) or a stalemate is reached (when all cells are occupied).
+A web-based implementation may involve players connected to a game server via WebSocket connections and interacting with the game from their web browser, which serve a \textit{single-page application} (SPA) of the game client written in a popular framework like \textit{React.js} \cite{React}.
+SPAs feature a single HTML page and dynamically renders content via JavaScript in the browser.
+Players take turns to make a move on the game board and the server implements the game logic to progress the game forward until a result (either a win/loss or draw) can be declared.
 
-% Type discipline for concurrent programs - behavioural types (--> session types)
+Whilst WebSockets make this web-based implementation possible, it introduces the developer to a new family of communication errors, even for this simple game.
+In addition to the usual testing for game logic correctness, the developer needs to test against \textit{deadlocks} (e.g.
+both players waiting for each other to make a move at the same time) and \textit{communication mismatches} (e.g.
+player 1 sending a boolean to the game server instead of the board coordinates).
+The complexity of these errors, which correlate to the complexity of tests required against these errors, scale with the complexity of the communication patterns involved.
 
-% Mainstream example of distirbuted programs: web services (microservice architecture), interactive web application;
+\textit{Multiparty Session Types} (MPST) \cite{MPST} provide a framework for formally specifying a structured communication pattern between concurrent processes and verifying implementations for correctness with respect to the communications aspect.
+By specifying the client-server interactions of our game as a protocol and verifying the implementations against the protocol for well-formedness, MPST theory guarantees well-formed implementations to be free from communication errors.
 
-% Objective:
+We see the application of the MPST methodology to verifying interactive web applications written in TypeScript to be an interesting design space - to what extent can the MPST methodology be applied such that we leverage TypeScript language features to maximise static safety guarantees whilst preserving a flexible, natural and idiomatic workflow for TypeScript developers?
+Such a workflow would ultimately decrease the overhead for incorporating MPST into mainstream web development, which reduces development time by programmatically verifying implementations for communication correctness.
+
+We present a workflow for developing type-safe interactive SPAs motivated by the MPST framework: \textbf{(1)} An endpoint API code generation workflow targeting TypeScript-based web applications for multiparty sessions; \textbf{(2)} An encoding of session types in server-side TypeScript that enforces static linearity; and \textbf{(3)} An encoding of session types in browser-side TypeScript using the React framework that guarantees affine usage of communication channels.
 

### 2020-01-22 interim/sections/background/codegen.tex
Interim: complete codegen Python subsection and comparison table

--- 2019-12-20 interim/sections/background/codegen.tex
@@
 \section{Code Generation}
+The EFSM describes the local session type and provides guidance to developers for verifying that their endpoint implementation conforms to the communication protocol.
+However, a direct encoding of the local session type into the target language of the implementation is usually not feasible as the EFSM assumes IO objects as first-class citizens and communication channels are linear resources, features that are left to be desired in mainstream programming languages (such as Java and Python) used by the developers' implementations.
+We note that languages with native session type support \cite{ATS2016} do exist, but their usage largely remains for research purposes as supposed to real-world application development.
 
-\subsection{Overview}
+Code generation is a common approach for verifying implementations written in the aforementioned mainstream languages against the EFSM.
+Approaches in the literature differ by how they leverage features in the target language (such as the type system), but generally define some interpretation of the EFSM in the target language and generate APIs which the developer can use to implement a target application that guarantees the following two properties:
+
 \begin{itemize}
-\item Motivate the practical implementation of session types in programming
-\item What needs to be verified - I/O behaviour (introduce the term behaviour typing) and linear channel usage
+\item \textbf{Behavioural typing}: The execution trace of messages sent and received by the application is accepted by the EFSM.
+\item \textbf{Channel linearity}: Each transition in the EFSM represents a channel resource.
+When the application transitions from some state $S$ to some successor state $S'$, it must no longer be able to access a reference (e.g.
+have an alias) to $S$.
 \end{itemize}
 
-\subsection{Static Session Typing}
+We outline the different existing approaches and summarise how they verify the aforementioned properties in \mathref{\ref{section:codegencompare}}.
+
+\subsection{Runtime Monitors}
+% What
+Neykova and Yoshida targeted the MPST methodology for Python programs in \cite{Python2017} and proposed to generate \term{runtime monitors} from the EFSM.
+These monitors expose APIs for sending and receiving messages, which is used by the developer in their implementation.
+The runtime monitor is an abstraction between the developer's implementation and the actual communication channel, and ``executes'' the EFSM internally to ensure protocol conformance.
+When the developer sends a message (with some label and payload) using the API, the runtime monitor checks whether this send action conforms to the current EFSM state, and if so, performs the send and advances to the successor state.
+Likewise, when the developer invokes a receive, the runtime monitor verifies that this is permitted at the current EFSM state before returning the received payload.
+
+We observe that this approach complements the dynamic typing nature of the Python language, which makes it sensible to perform behavioural typing at runtime.
+As the send and receive IO primitives are made available to the developer, there are no ``instances'' of channel resources created, so the developer cannot explicitly hold a reference to some state in the EFSM (let alone keep aliases), so channel linearity is trivially guaranteed here.
+
+\subsection{Type-Level Encoding}
 \begin{itemize}
 \item PureScript implementation that addresses the 2 criteria statically
-\item \begin{itemize}
+\begin{itemize}
 \item Relevance for web development
@@
 
-PureScript - \cite{PS2019}
-
-\subsection{Runtime Monitors}
+PureScript - \cite{PureScript2019}
 
@@
 
-Java - \cite{HY2016}
+Java - \cite{Hybrid2016}
 
 \subsection{Comparison}
-\begin{itemize}
-\item Comparison table between the 3 approaches for how they deal with behavioural typing 
-\end{itemize}
+\label{section:codegencompare}
+
+We compare how these existing code generation approaches provide communication safety guarantees in Table \ref{table:comparison}.
+
+\begin{figure}[!h]
+\centering
+\begin{tabular}{l || p{0.35\textwidth} | p{0.35\textwidth}}
+%\begin{tabular}{l||l|l}
+Language & Behavioural typing & Channel linearity \\
+\hline\hline
+Python \cite{Python2017} & Dynamically enforced - runtime monitor represent the EFSM and only execute supported transitions.
+& Trivially guaranteed - channel resources not exposed, developer uses \texttt{send()} and \texttt{receive()} primitives.
+\\
+\hline
+PureScript \cite{PureScript2019} & Type-level encoding - EFSM states are encoded as types, IO actions are encoded as multi-parameter type classes (to express the semantics of the state transition function) and EFSM transitions are encoded as instances of said type classes.
+& Statically guaranteed - channels are not exposed in the provided combinators to prevent reuse, and session constructor requires a \texttt{Session} continuation parameterised by initial and terminal state to prevent incomplete sessions.
+\\
+\hline
+Java \cite{Hybrid2016} & Statically guaranteed - states are encoded as classes; transitions are encoded as instance methods on the state class and return a new instance of the successor state class.
+& Checked at runtime - states keep \texttt{used} boolean flag to detect and prevent reuse, Session API implements \texttt{AutoCloseable} interface to be used in resource try-catch block to prevent unused.
+\end{tabular}
+\captionof{table}{Comparison between existing MPST code generation approaches.}
+\label{table:comparison}
+\end{figure}

### 2020-01-23 interim/sections/background/codegen.tex

@@
 \subsection{Type-Level Encoding}
-\begin{itemize}
-\item PureScript implementation that addresses the 2 criteria statically
-\begin{itemize}
-\item Relevance for web development
-\item Static behavioural typing
-\item Static channel linearity by construction
-\item Pros - type dependencies (expressing dependent types)
-\item Cons
-\end{itemize}
-\item Other initiatives -- creating language with first-class channel primitives (e.g. SILL)
-\end{itemize}
+King et al.
+presented an approach in \cite{PureScript2019} for integrating session types into web development using the PureScript language, which takes advantage of its expressive type system to provide static guarantees on implementation conformance with respect to the protocol.
+We outline the main components of their EFSM encoding:
 
-PureScript - \cite{PureScript2019}
+\paragraph{Actions as type classes} The semantics of the state transition function in the EFSM formalism express that a tuple of state-action \textit{uniquely defines} a successor state.
+These semantics can be expressed by \textit{multi-parameter type classes (MPTC) with functional dependencies}: \texttt{class Send r s t a | s -> t r a} defines \texttt{Send} as a MPTC parameterised by recipient \texttt{r}, current state \texttt{s}, successor state \texttt{t} and payload type \texttt{a}, and \texttt{s -> t r a} expresses the functional dependency that, for an instance of this type class, the current state uniquely determines the successor state, the recipient and payload type.
+These type classes are independent of the EFSM.
+
+\paragraph{Transitions as instances of type classes} By encoding states as data types, valid EFSM transitions are encoded as \textit{instances} of the type classes.
+If \texttt{S1} is an output state, sending an \texttt{Int} to \texttt{Svr} with successor \texttt{S2}, we would encode \texttt{instance SendS1 :: Send Svr S1 S2 Int}.
+Because of the functional dependency, the developer cannot instantiate an invalid transition (e.g.
+\texttt{Send Svr S1 S3 Bool}, since \texttt{S1} uniquely determines the other type parameters.
+
+This proposal is relevant to the problem we are tackling, and we appreciate that the intricacy of the library design in the communication combinators to conceal the channel resources is something we can build upon in our solution.
+We also observe the challenges of applying session types to the front-end environment, as shown by the careful choice made in \cite{PureScript2019} to use the \textit{Concur UI} framework which builds UI elements sequentially to model sequential sessions; not doing so would require binding channel resources into event listeners on UI elements, which makes linearity violation possible (e.g.
+by binding a send transition to a button click event, the user might click the button twice, thus reusing the channel).
 
 \subsection{Hybrid Session Verification}
-\begin{itemize}
-\item How the aforementioned 2 approaches motivated hybrid session verification
-\item Workflow using Scribble toolchain 
-\item Static behavioural typing, dynamic channel usage runtime checks
-\item Pros - Abstract I/O interfaces
-\item Pros - Input futures for `complex communication patterns' motivated by SMTP example
-\item Cons - ``practical compromise'' as outlined in paper
-\end{itemize}
+Whilst runtime monitors and type-level encoding achieve communication safety guarantees via dynamic and static approaches respectively, Hu and Yoshida implemented a workflow in \cite{Hybrid2016} that performs hybrid verification of communication protocol conformance for Java applications.
+We highlight the two main components below:
 
-Java - \cite{Hybrid2016}
+\paragraph{Static session typing} States are represented as classes; supported transitions on each state are represented as instance methods, parameterised by the role, label and payload involved in the message exchange.
+A send method takes the payload to send as a parameter, whilst a receive method is a blocking call that requires the caller to allocate a \texttt{Buf<T>} wrapper on the stack (where \texttt{T} is the expected payload type), then the receive method populates the payload into the wrapper and returns upon receiving from the channel.
+These instance methods return a new instance of the successor state class.
+
+\paragraph{Runtime linear channel usage checks} Each state keeps track of its usage in a private boolean flag and throws an exception when the instance method is called twice, which signifies channel reuse and a linearity violation.
+Similarly, the \texttt{SessionEndpoint} class keeps track of whether the connection is open or close, and throws an exception when program execution exits the scope of the session endpoint and a terminal state has yet to be reached, signifying a protocol violation as the session hasn't completed yet but is now out of scope.
+
+We find that this proposal strikes a good balance between maximising static communication safety guarantees whilst providing an intuitive set of APIs for developers to efficiently write their applications.
+For example, the encoding of transitions as instance methods that return new instances of the successor state exposes the channel resource, and since Java does not build support for linear resources and does not monitor aliasing of variables, linear channel usage is monitored dynamically.
+However, this complements the imperative style of application code written in Java, and takes advantage of the type system to statically enforce valid transitions in the code.
 

### 2020-01-23 interim/sections/background/scribble.tex
Interim: add introduction and lstlisting styles from PLACES

@@
 
-\subsection{Overview}
+%\subsection{Overview}
 % Introduce Scribble as an implementation of MPST
@@
 \begin{figure}[!h]
-\begin{lstlisting}
+\begin{lstlisting}[language=Scribble]
 type <java> "java.lang.Integer" from "rt.jar" as int;
@@
 sending to or receiving from another role) permitted at the endpoint at that state.
+Figure \ref{fig:efsmsyntax} presents the EFSM formalism adopted from \cite{Hybrid2016}.
 
@@
 \singlespacing
-\caption{Syntax for Endpoint FSM}
+\captionof{figure}{Syntax for Endpoint FSM}
+\label{fig:efsmsyntax}
 \end{figure}
 
-\begin{figure}[!hb]
+We highlight the semantics of the \term{state transition function}.
+Its domain defines the set of permitted actions for each state, and the range defines the successor state.
+Crucially, this is a \term{partial function} because the EFSM restricts what can be done (in terms of send and receive actions, and what label/payload types can be sent and received).
+
+\begin{figure}[!h]
 \doublespacing
@@
 
-\texttt{initial}_{\mathbb{R}, \mathbb{L}, \mathbb{T}, \Sigma, \mathbb{S}, \delta}(S) \iff & \nexists S' \in \mathbb{S},\alpha \in \Sigma.~\delta(S',\alpha) = S \\
-\texttt{terminal}_{\mathbb{R}, \mathbb{L}, \mathbb{T}, \Sigma, \mathbb{S}, \delta}(S) \iff & \delta(S) = \emptyset \\
-\texttt{output}_{\mathbb{R}, \mathbb{L}, \mathbb{T}, \Sigma, \mathbb{S}, \delta}(S) \iff & \delta(S) = \{\,\alpha \in \Sigma \mid \exists r \in \mathbb{R}, l \in \mathbb{L}, \tilde{T} \subseteq{\mathbb{T}}.~\alpha = r!l(\tilde{T})\,\} \\
-\texttt{input}_{\mathbb{R}, \mathbb{L}, \mathbb{T}, \Sigma, \mathbb{S}, \delta}(S) \iff & \delta(S) = \{\,\alpha \in \Sigma \mid \exists r \in \mathbb{R}, l \in \mathbb{L}, \tilde{T} \subseteq{\mathbb{T}}.~\alpha = r?l(\tilde{T})\,\} \\
+\texttt{initial}_{EFSM}(S) \iff & \nexists S' \in \mathbb{S},\alpha \in \Sigma.~\delta(S',\alpha) = S \\
+\texttt{terminal}_{EFSM}(S) \iff & \delta(S) = \emptyset \\
+\texttt{output}_{EFSM}(S) \iff & \delta(S) = \{\,\alpha \in \Sigma \mid \exists l \in \mathbb{L}, \tilde{T} \subseteq{\mathbb{T}}.~\alpha = r!l(\tilde{T})\,\}~\text{for some}~r \in \mathbb{R} \\
+\texttt{input}_{EFSM}(S) \iff & \delta(S) = \{\,\alpha \in \Sigma \mid \exists l \in \mathbb{L}, \tilde{T} \subseteq{\mathbb{T}}.~\alpha = r?l(\tilde{T})\,\}~\text{for some}~r \in \mathbb{R}
 \end{array}
@@
 \singlespacing
-\caption{Types of EFSM States}
+\captionof{figure}{Types of EFSM States}
+\label{fig:efsmstates}
 \end{figure}
+
+\cite{Hybrid2016} also identifies properties of EFSMs for well-formed protocol specifications: \textbf{(1)} there is exactly one initial state, \textbf{(2)} there is at most one terminal state, and \textbf{(3)} every non-terminal state is either an input state or output state.
+There are further restrictions on input and output states, namely these states must only send to or receive from the same role.
+We formalise these properties in Figure \ref{fig:efsmstates}.

### 2020-01-23 PLACES_2020/main.tex
change

--- 2020-01-21 PLACES_2020/paper.tex
@@
 
-\title{Multiparty Session Type-safe Web Development in TypeScript}
+\title{Verifying Interactive WebSocket Applications in TypeScript}
 \author{Anson Miu
@@
 }
-\def\titlerunning{MPST-safe Web Development in TypeScript}
+\def\titlerunning{Verifying Interactive WebSocket Applications in TypeScript}
 \def\authorrunning{A. Miu, F. Ferreira, N. Yoshida \& F. Zhou}
@@
 Advancements in mobile device compute power has made interactive web applications possible, allowing the web browser to render content dynamically and support low-latency communication with the server.
-This comes at a cost to the developer, who now need to reason more about the communication correctness of their application as they support more complex communication patterns.
+This comes at a cost to the developer, who now needs to reason more about the communication correctness of their application as they support more complex communication patterns.
 
@@
 
-We see the application of the MPST methodology to verifying interactive web applications written in TypeScript to be an interesting design space - to what extent can the MPST methodology be applied such that we leverage TypeScript language features to maximise static safety guarantees whilst preserving a flexible, natural and idiomatic workflow for TypeScript developers?
-Such a workflow would ultimately decrease the overhead for incorporating MPST into mainstream web development, which reduces development time by programmatically verifying implementations for communication correctness.
+We see the application of the MPST methodology to verifying
+interactive web applications written in TypeScript to be an
+interesting design space -- to what extent can the MPST methodology be
+applied such that we leverage TypeScript language features to maximise
+static safety guarantees whilst preserving a flexible, natural and
+idiomatic workflow for TypeScript developers? Such a workflow would
+ultimately decrease the overhead for incorporating MPST into
+mainstream web system, which reduces development time by
+programmatically
+verifying communication correctness of the implementation.
 
@@
 
-We leverage the Scribble toolchain to check for protocol well-formedness.
-This directly corresponds to the multiparty session type theory of verifying valid local type \textit{projections} for all participants of a global type.
-The implementation of this validation by the toolchain follow from the algorithmic projection procedures defined in \cite{MPST}.
-We obtain \textit{endpoint protocols} for each role in a well-formed global protocol.
+We leverage the Scribble toolchain to check for protocol
+well-formedness. This directly corresponds to the multiparty session
+type theory of verifying valid local type \textit{projections} for all
+participants of a global type. The implementation of this validation
+by the toolchain follows from the algorithmic projection procedures
+defined in \cite{MPST}. We obtain a set of \textit{endpoint protocols}
+-- one for each role from a well-formed global protocol.
 An endpoint protocol only preserves the interactions defined by the global protocol in which the target role is involved, and corresponds to an equivalent \textit{Endpoint Finite State Machine} (EFSM) in the Scribble toolchain.
@@
 
-Developers can validate the communication aspects of their implementation against their EFSM to verify protocol conformance.
-Our approach integrates the EFSM into the development workflow by encoding session types as TypeScript types.
-We appreciate the inherent differences in development considerations between front- and back-end web development: business logic tends to initiate communication in the back-end, whilst communication in the front-end is mostly driven by user interactions with the browser.
+Developers can validate the communication aspects of their
+implementation against their EFSM to verify protocol conformance. Our
+approach integrates the EFSM into the development workflow by encoding
+session types as TypeScript types.
+There are inherent differences between front- and back-end web development: business logic tends to initiate communication in the back-end, whilst communication in the front-end is mostly driven by user interactions with the browser.
 This motivates our design choice of generating different session types encodings for the target application on the server (\sectionref{section:server}) and in the browser (\sectionref{section:browser}).
@@
 
-We will refer to the \texttt{Svr} EFSM (fig.~\ref{fig:efsmsvr}) as a running example in this section.
-For server-side targets, we encode EFSM states as TypeScript types depending on the type of state.
-We adopt the EFSM definition presented in \cite{Hybrid2016} to consider receive and send states separately.
-We assign each TypeScript encoding its state identifier from the EFSM as its type alias, thus providing syntactic sugar when referring to the successor state in the TypeScript encoding of the current state.
+We will refer to the \texttt{Svr} EFSM (fig.~\ref{fig:efsmsvr}) as a
+running example in this section. For server-side targets, we encode
+EFSM states as TypeScript types depending on the type of state. We
+adopt the EFSM definition presented in \cite{Hybrid2016} to consider
+receive and send states separately. Each TypeScript encoding is
+assigned to its state identifier from the EFSM as its type alias, thus providing syntactic sugar when referring to the successor state in the TypeScript encoding of the current state.
 For any state $S$ in the EFSM, we refer to the TypeScript type alias of its encoding as $\llbracket S \rrbracket$.
 
-We make a key design decision \textit{not} to expose communication channels in the TypeScript session type encodings to provide linearity guarantees (\sectionref{section:serverlinear}).
-Our encodings sufficiently exposes seams for the developer to inject their business logic, whilst the generated session API (\sectionref{section:serversessionapi}) handles the sending and receiving of messages; as a result, the encodings do not concern with the role involved in the send/receive action.
-We outline the encodings below using examples from the \textit{Noughts and Crosses} game server (Listing \ref{lst:svr}).
+We make a key design decision \textit{not} to expose communication channels in the TypeScript session type encoding to provide linearity guarantees (\sectionref{section:serverlinear}).
+Our encoding sufficiently exposes seams for the developer to inject their business logic, whilst the generated session API (\sectionref{section:serversessionapi}) handles the sending and receiving of messages; as a result, the encoding does not concern with the role involved in the send/receive action.
+We outline the encoding below using examples from the \textit{Noughts and Crosses} game server (Listing \ref{lst:svr}).
 
@@
 
-We will refer to the \texttt{P1} EFSM (figure \ref{fig:efsmp1}) as a running example in this section.
-Preserving behavioural typing and channel linearity are known to be challenging for browser-side applications due to the event-driven nature of user interaction: in the case of \textit{Noughts and Crosses}, once the user makes a move by clicking on a cell on the game board, this click event must be deactivated until the user's next turn, otherwise the user can click again and violate channel linearity.
-Our design goal is to enforce this statically through the APIs we generate.
+We will refer to the \texttt{P1} EFSM (figure \ref{fig:efsmp1}) as a
+running example in this section. Preserving behavioural typing and
+channel linearity are known to be challenging for browser-side
+applications due to the event-driven nature of user interaction: in
+the case of \textit{Noughts and Crosses}, once the user makes a move
+by clicking on a cell on the game board, this click event must be
+deactivated until the user's next turn, otherwise the user can click
+again and violate channel linearity. Our design goal is to enforce
+this statically through the generated APIs.
 
@@
 
-\paragraph{Components} A component is a reusable UI element which contains its own markup and logic.
-Components implement a \texttt{render()} function which returns a React element, the smallest building blocks of a React application.
-Components can keep \textit{state} and the \texttt{render()} function is invoked upon a change of state.
-A simple counter can be implemented as a component, with the count stored as state, a button which increments the count when clicked and a \texttt{div} that renders the current count.
-Components can also render other components, which gives rise to a parent/child relationship between components.
-Parents can pass data to children as \textit{props}: going back to the aforementioned example, the the counter component could render a child component \texttt{<StyledDiv count=\{this.state.count\} />} in its \texttt{render()} function, propagating the \texttt{count} from its state to the child.
-This enables reusability, and for our use case, gives control to the parent on what data to pass to its children (e.g.
-pass the payload of a received message to a child to render).
+\paragraph{Components} A component is a reusable UI element which
+contains its own markup and logic. Components implement a
+\texttt{render()} function which returns a React element, the smallest
+building blocks of a React application. Components can keep
+\textit{state} and the \texttt{render()} function is invoked upon a
+change of state. A simple counter can be implemented as a component,
+with the count stored as state, a button which increments the count
+when clicked and a \texttt{div} that renders the current
+count. Components can also render other components, which gives rise
+to a parent/child relationship between components. Parents can pass
+data to children as \textit{props}: going back to the aforementioned
+example, the the counter component could render a child component
+\texttt{<StyledDiv count=\{this.state.count\} />} in its
+\texttt{render()} function, propagating the \texttt{count} from its
+state to the child. This enables reusability, and for our use case,
+gives a control to the parent on what data to pass to its children (e.g. pass the payload of a received message to a child to render).
 

### 2020-01-23 PLACES_2020/main.tex
PLACES: shorten code listing whitespace

@@
   choice at Svr {
-    Lose(Point) from Svr to P2;
-    Win(Point) from Svr to P1;
+    Lose(Point) from Svr to P2; Win(Point) from Svr to P1;
   } or {
-    Draw(Point) from Svr to P2;
-    Draw(Point) from Svr to P1;
+    Draw(Point) from Svr to P2; Draw(Point) from Svr to P1;
   } or {
-    Update(Point) from Svr to P2;
-    Update(Point) from Svr to P1;
+    Update(Point) from Svr to P2; Update(Point) from Svr to P1;
     do Game(Svr, P2, P1);
@@
 \begin{lstlisting}[language=JavaScript]
-import { Coordinate as Point } from "./Types";
-import { Labels } from "./Constants";
-
 export type S13 = { Pos: (payload: Point) => S15 }
@@
 \begin{lstlisting}[language=JavaScript, tabsize=4]
-render() {
-	...
+// Inside some render() function..
 {board.map((row, x) => (
-		board.map((col, y) => {
+	row.map((col, y) => {
 		const SelectPoint = this.props.Pos('click', (event: UIEvent) => {
 			event.preventDefault()
-				return { x, y }
-			}
+			return { x: x, y: y }}
 		return <SelectPoint><td>.</td></SelectPoint>
-		}
-	)}
-	...	
-}
+})}
 \end{lstlisting}  

### 2020-01-23 PLACES_2020/main.tex
PLACES: update related works and shorten paper

@@
   morestring=[b]',
-  morestring=[b]"
+  morestring=[b]",
 }
@@
 	frame=single,
+	tabsize=4
 }
@@
 Developers have since adopted other communication transport abstractions over HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy low-latency full-duplex client-server communication in their applications over a single persistent connection.
-Enabling more complex communication patterns caters for more interactive use cases (such as real-time multiplayer games), but introduces additional concerns to the developer with respect to implementation correctness.
+Enabling more complex communication patterns caters for more interactive use cases (such as real-time multiplayer games), but introduces additional correctness concerns to the developer.
 
@@
 player 1 sending a boolean to the game server instead of the board coordinates).
-The complexity of these errors, which correlate to the complexity of tests required against these errors, scale with the complexity of the communication patterns involved.
+The complexity of these errors correlate to the complexity of related tests required, scale with the complexity of communication patterns involved.
 
@@
 
-Development begins by specifying the permitted communications between participants as a \textit{global protocol} in Scribble, a MPST-based framework introduced in \cite{Scribble} which features a protocol specification language and code generation toolchain.
-This means that a protocol specified in Scribble maps to some \textit{global type} in multiparty session type theory \cite{MPST}.
+Development begins by specifying the permitted communications between participants as a \textit{global protocol} in Scribble \cite{Scribble}, a MPST-based protocol specification language and code generation toolchain.
 We specify the \textit{Noughts and Crosses} game as a Scribble protocol in Listing \ref{lst:game}.
@@
 We leverage the Scribble toolchain to check for protocol
-well-formedness. This directly corresponds to the multiparty session
-type theory of verifying valid local type \textit{projections} for all
-participants of a global type. The implementation of this validation
-by the toolchain follows from the algorithmic projection procedures
-defined in \cite{MPST}. We obtain a set of \textit{endpoint protocols}
+well-formedness.
+This directly corresponds to multiparty session
+type theory \cite{MPST}: 
+a Scribble protocol maps to some \textit{global type}, and protocol verification implements the algorithmic projection
+defined in \cite{MPST} to derive valid local type \textit{projections} for all participants.  We obtain a set of \textit{endpoint protocols}
 -- one for each role from a well-formed global protocol.
@@
 There are inherent differences between front- and back-end web development: business logic tends to initiate communication in the back-end, whilst communication in the front-end is mostly driven by user interactions with the browser.
-This motivates our design choice of generating different session types encodings for the target application on the server (\sectionref{section:server}) and in the browser (\sectionref{section:browser}).
+This motivates our design choice of generating different session types encodings for server (\sectionref{section:server}) and browser targets (\sectionref{section:browser}).
 
@@
 
-  \captionof{figure}{EFSM for \texttt{Svr} in \textit{Noughts and Crosses}.}
+  \captionof{figure}{EFSM for \texttt{Svr}.}
   \label{fig:efsmsvr}
@@
 
-We make a key design decision \textit{not} to expose communication channels in the TypeScript session type encoding to provide linearity guarantees (\sectionref{section:serverlinear}).
+We make a key design decision \textit{not} to expose communication channels in the TypeScript session type encoding to provide \textit{static} linearity guarantees (\sectionref{section:serverlinear}).
 Our encoding sufficiently exposes seams for the developer to inject their business logic, whilst the generated session API (\sectionref{section:serversessionapi}) handles the sending and receiving of messages; as a result, the encoding does not concern with the role involved in the send/receive action.
-We outline the encoding below using examples from the \textit{Noughts and Crosses} game server (Listing \ref{lst:svr}).
+We outline the encoding below using examples from the \textit{Noughts and Crosses} game (Listing \ref{lst:svr}).
 
 \paragraph{Branching state} We consider a receive state as a unary branching state for conciseness.
-A branching state is encoded as an \textit{object literal} \cite{TypeScriptSpec}, with each branch corresponding to a member field.
+A branching state is encoded as an \textit{object literal} \cite{TypeScriptSpec}, with each branch $i \in I$, $I$ denoting set of all branches, corresponding to a member field.
 A branch expecting to receive a message labelled $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as an \textit{member field} named $\texttt{label}_i$ of function type \texttt{(payload:$\texttt{T}_i$) => $\llbracket S_i \rrbracket$}.
@@
 \paragraph{Selection state} We consider a send state as a unary selection state for conciseness.
-A selection state is encoded as a \textit{union type} \cite{TypeScriptSpec} of internal choice encodings: each internal choice sending a message labelled $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as a \textit{tuple type} of \texttt{[Labels.label$_i$, T$_i$, $\llbracket S_i \rrbracket$]}.
+A selection state is encoded as a \textit{union type} \cite{TypeScriptSpec} of internal choice encodings: each internal choice $i \in I$, $I$ denoting set of all choices, sending a message labelled $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as a \textit{tuple type} of \texttt{[Labels.label$_i$, T$_i$, $\llbracket S_i \rrbracket$]}.
 The developer implements a selection operation by passing the selected label and payload to send in the message.
-We generate a \textit{string enum} (named \texttt{Labels}) wrapping the labels in the protocol, hence the enum member access in the first element of the tuple type.
+We generate a \textit{string enum} (named \texttt{Labels}) wrapping the labels in the protocol.
 
@@
 \begin{lstlisting}[language=JavaScript]
-export type S13 = { Pos: (payload: Point) => S15 }
-export type S15 = [Labels.Lose, Point, S16]
-		| [Labels.Draw, Point, S17]
-		| [Labels.Update, Point, S18]
+export type S13 = { Pos: (payload: Point) => S15 };
+export type S15 = [Labels.Lose, Point, S16] | [Labels.Draw, Point, S17]
+											| [Labels.Update, Point, S18];
 \end{lstlisting}  
@@
 
-Our session runtime performs communication in a way that conforms to the protocol specification, but does not expose these IO actions to the developer by delegating the aforementioned responsibilities to an inner class.
+Our session runtime performs communication in a protocol-conformant manner, but does not expose these IO actions to the developer by delegating the aforementioned responsibilities to an inner class.
 The runtime listens to message (receive) events on the communication channel, invokes the corresponding callback to obtain the value to send next, and performs the send.
-The developer instantiates the session by constructing an instance of the session runtime class.
+The developer instantiates the session by constructing an instance of the session runtime class, providing the WebSocket endpoint URL and initial state.
 
@@
 \label{section:serverlinear}
-Our callback-oriented design for server-side API generation provides guarantees on state channel linearity by preventing the two properties detailed below.
-Here, channel linearity is guaranteed based on the correctness of our library design and session runtime implementation, which means a faulty implementation could violate this, but is up to the library author to verify once rather than the end user.
+Channel linearity is guaranteed based on the correctness of our library design and session runtime implementation, which means a faulty implementation could violate this, but is up to the library author to verify once rather than the end user.
+Our library design prevent the two properties below:
 
@@
 \paragraph{Unused} The initial state must be supplied to the session runtime constructor in order to instantiate the session; this initial state is defined in terms of the successor states, which in turn has references to its successors and so forth.
-This encoding approach will cover the terminal state (if it exists in the EFSM, as \cite{Hybrid2016} notes that an EFSM contains at most one terminal state), and the session runtime guarantees this terminal state, if it exists, will be reached by construction.
+This encoding approach will cover the terminal state (if it exists), and the session runtime guarantees this terminal state, if it exists, will be reached by construction.
 
@@
 channel linearity are known to be challenging for browser-side
-applications due to the event-driven nature of user interaction: in
-the case of \textit{Noughts and Crosses}, once the user makes a move
+applications due to EFSM transitions being triggered by user events: 
+in the case of \textit{Noughts and Crosses}, once the user makes a move
 by clicking on a cell on the game board, this click event must be
@@
 
-\begin{wrapfigure}{r}{0.5\textwidth}
+\begin{wrapfigure}{r}{0.4\textwidth}
   \begin{center}
-    \includegraphics[width=0.48\textwidth]{figures/efsm_p1.png}
+    \includegraphics[width=0.4\textwidth]{figures/efsm_p1.png}
   \end{center}
 
-  \captionof{figure}{EFSM for \texttt{P1} in \textit{Noughts and Crosses}.}
+  \captionof{figure}{EFSM for \texttt{P1}.}
   \label{fig:efsmp1}
@@
 \subsubsection{The React Framework}
-Rather than using the LINKS web programming language from \cite{MVU2019}, we apply the multiple model types approach to the React framework \cite{React} developed by Facebook.
+Rather than using the Links web programming language from \cite{MVU2019}, we apply the multiple model types approach to the React framework \cite{React} developed by Facebook, albeit originally intended for the \textit{Model-View-Controller} (MVC) architecture.
 React is widely used in industry to create scalable single-page TypeScript applications, and we intend for our proposed workflow to be beneficial in an industrial context.
-We introduce the key features of the framework below.
+We introduce the key features of the framework.
 
@@
 change of state. A simple counter can be implemented as a component,
-with the count stored as state, a button which increments the count
-when clicked and a \texttt{div} that renders the current
-count. Components can also render other components, which gives rise
+with \texttt{count} stored as state, a button which increments \texttt{count}when clicked and a \texttt{div} that renders the current
+\texttt{count}. Components can also render other components, which gives rise
 to a parent/child relationship between components. Parents can pass
@@
 \subsubsection{Model types in React}
-We leverage React to express the model types approach in \cite{MVU2019} rather than the \textit{Model-View-Controller} (MVC) architecture it was intended for.
 
@@
 		const SelectPoint = this.props.Pos('click', (event: UIEvent) => {
-			event.preventDefault()
-			return { x: x, y: y }}
-		return <SelectPoint><td>.</td></SelectPoint>
-})}
+			event.preventDefault();
+			return { x: x, y: y };}
+		return <SelectPoint><td>.</td></SelectPoint>;
+});}
 \end{lstlisting}  
@@
 \subparagraph{Receive} The payload of a message receive transition is passed as a prop to the successor state and can be handled in the successor's view function.
-Upon receiving the message from the channel, the session runtime renders the successor of the receive state and propagates the payload as the prop accordingly.
+Upon a message receive event, the session runtime renders the successor of the receive state and propagates the payload to the successor accordingly.
 
@@
 Similarly, the session runtime is responsible for passing the required props for model transitions to EFSM state React components.
-The session runtime component is rendered by the programmer and requires the following props:
+The session runtime component is rendered by the programmer and requires, as props, the \textit{endpoint URL} (so it can open the connection) and a list of \textit{concrete state components}.
 
-\paragraph{Endpoint URL} The session runtime needs this in order to open the WebSocket connection.
-
-\paragraph{Concrete state components} The programmer writes their own implementation of each state (mainly to customise how the state is rendered and inject business logic into state transitions) by extending the abstract React class components.
+The programmer writes their own implementation of each state (mainly to customise how the state is rendered and inject business logic into state transitions) by extending the abstract React class components.
 The session runtime requires references to these concrete components order to render the user implementation accordingly.
@@
 Communication channels are not exposed to the programmer so multiple sends are impossible.
-This does not restrict the programmer from binding the send action to exactly one UI event: for the \textit{Noughts and Crosses} game, we bind the \texttt{Pos(Point)} send action to each unoccupied cell on the game board, but the generated runtime ensures that, once an event bound to the send action is triggered, the send is only performed once and the successor state is rendered on the DOM to prevent multiple sends initiated by the user.
+This does not restrict the programmer from binding the send action to exactly one UI event: for \textit{Noughts and Crosses}, we bind the \texttt{Pos(Point)} send action to each unoccupied cell on the game board, but the generated runtime ensures that, once the cell is clicked, the send is only performed once and the successor state is rendered on the DOM, so the channel resource used to send becomes unavailable.
 
@@
 const handleP1Move: Receive_P1 = (move: Point) => {
-	// User-implemented logic
-	board.P1(move)
-	if (board.won()) { return [Labels.Lose, move, move] }
-	else if (board.draw()) { return [Labels.Draw, move, move] }
-	else { return [Labels.Draw, move, handleP2Move] } 
-}
+	board.P1(move);			// User logic
+	if (board.won()) { return [Labels.Lose, move, move]; }
+	else if (board.draw()) { return [Labels.Draw, move, move]; }
+	else { return [Labels.Update, move, handleP2Move]; }}
 
-const handleP2Move: Receive_P2 = ...
-
-// Instantiate session
-new NoughtsAndCrosses.Svr(webSocketServer, handleP1Move, handleP2Move)
+// Instantiate session - `handleP2Move` defined similarly
+new NoughtsAndCrosses.Svr(webSocketServer, handleP1Move, handleP2Move);
 \end{lstlisting}  
-\captionof{lstlisting}{Example fragment of \textit{Noughts and Crosses} \texttt{Svr} implementation.}
+\captionof{lstlisting}{Session runtime instantiation for \textit{Noughts and Crosses} \texttt{Svr}.}
 \label{lst:svrprotocol}
@@
 
-\paragraph{Game clients} We implement the game client for players as a React application - extending from the generated abstract React components, registering our custom implementation as props for the session runtime component.
+\paragraph{Game clients} We implement the game client for \texttt{P1} and \texttt{P2} by extending from the generated abstract React (EFSM state) components and registering those to the session runtime component.
 
@@
 \section{Related Works}
-%\paragraph{Languages with native session type support} \dots
-% New languages written  (e.g. ATS)
-%	- addresses the challenge of resource linearity with a liner type system
-%	- problem with webdev: not compatible with libraries, technologies, flexibility of JavaScript (ultimately the language of the browser)
+Native language support for first-class linear channel resources \cite{ATS} and code generation are two main approaches for incorporating MPST into application development.
+The latter closely relates to our proposal; we highlight two areas of existing works under this approach that motivate our design choice.
 
@@
 
-Our work applies the aforementioned approaches in a \textit{multiparty} context in a manner that is compatible with industrial tools and practices to ultimately invite the notion of MPST-safety into interactive web application development in industry.
+Our work applies the aforementioned approaches in a \textit{multiparty} context using industrial tools and practices to ultimately encourage MPST-safe web application development workflows in industry.
+
+%\paragraph{Typestate programming} \dots
 
@@
 Future works include incorporating \textit{explicit connection actions} introduced in \cite{ExplicitConnections} in our API generation to better model real-world communication protocols that may feature in interactive web applications.
-Whilst our approach does support multiparty sessions, protocols are required to have exactly one role running on the server and client roles must interact with each other via the server to respect the properties of WebSocket connections, which is the transport abstraction we support.
+Whilst our approach supports multiparty sessions, the nature of WebSockets require some server-based role in the communication protocol and clients to interact via the server.
 Extending support to WebRTC would cater for peer-to-peer communication between browsers, which further opens up possibilities for communication protocols supported by our approach.

### 2020-01-23 PLACES_2020/main.tex
PLACES: minor change

@@
 The runtime listens to message (receive) events on the communication channel, invokes the corresponding callback to obtain the value to send next, and performs the send.
-The developer instantiates the session by constructing an instance of the session runtime class, providing the WebSocket endpoint URL and initial state.
+The developer instantiates the session by constructing an instance of the session runtime class, providing the WebSocket endpoint URL (to open the connection) and initial state (to execute the EFSM).
 

### 2020-01-23 interim/interim.tex
Interim: complete codegen Python subsection and comparison table
Interim: update title page and extract sections as separate .tex files

--- 2020-01-22 interim/interim.tex
@@
 \newcommand{\supervisor}{Prof. Nobuko Yoshida}
-\newcommand{\sndmarker}{TBC}
+\newcommand{\sndmarker}{Dr. Iain Phillips}
 \newcommand{\cid}{your college-id number}
@@
 
-\chapter{Introduction}
-
-% Rise of distributed programs -> main challenges
-
-% How to ensure correctness in general: type system, data types
-
-% Type discipline for concurrent programs - behavioural types (--> session types)
-
-% Mainstream example of distirbuted programs: web services (microservice architecture), interactive web application;
-
-% Objective:
+\input{sections/introduction.tex}
 
@@
 \input{sections/background/scribble.tex}
-\clearpage
 
 \input{sections/background/codegen.tex}
-\clearpage
 
-\input{sections/background/typescript.tex}
+%\input{sections/background/typescript.tex}
 
-\chapter{Project Plan}
-We aim to develop a multiparty session type-safe development workflow for building interactive full-stack TypeScript applications that conform to a communication protocol.
-This involves encoding session types using the TypeScript language and implementing a code generation workflow to generate the encodings.
+\input{sections/projectplan.tex}
 
-%As we target full-stack applications, we tackle server-side and browser-side targets separately:
-%
-%\paragraph{Server-side targets} We will analyse the approaches presented in \cite{Hybrid2016, PureScript2019, Python2017} and assess their applicability with respect to the language features and type system offered by TypeScript and general back-end web development patterns.
-At the point of writing, we have prototyped several approaches for server-side session type TypeScript encodings and will compare their effectiveness through implementing example protocols, such as the \textit{Adder} example in \cite{Hybrid2016} and the \textit{Battleships} example in \cite{PureScript2019}.
-%
-%\paragraph{Browser-side targets} We will analyse how \cite{PureScript2019} handles API generation for front-end implementations, and in particular, study their approach to behavioural typing and guaranteeing channel linearity in an event-driven environment.
-\cite{MVU2019} also presents an architecture for verifying binary session type conformance of fornt-end implementations through formalisms of the \textit{Model-View-Update} (MVU) architecture and \textit{model types}.
-We will investigate how to extend their approach for multiparty session types using TypeScript.
-At the point of writing, we have explored methods to express the methodology in \cite{MVU2019} using React.js for a multiplayer game.
-
-\section{Delivery}
-At the point of writing, we have analysed the existing code generation approaches presented in \cite{Hybrid2016, Scribble, Python2017} and assessed their applicability with respect to verifying the communication aspects of interactive web applications.
-We have also explored approaches for encoding session types into TypeScript, motivated by existing works along with similar proposals specific to web development in \cite{PureScript2019, MVU2019}.
-
-We plan to deliver our code generation implementation incrementally, aiming to get a working version of the end-to-end workflow first, then iteratively add support for more complex session type primitives.
-This minimises the risks involved in the project by ensuring we have a functional deliverable at the early stages of the project.
-We plan to deliver this basic end-to-end workflow by the end of February.
-
-The subsequent months will involve adding support for other primitives (i.e.
-selection, choice and recursion) for multiparty sessions.
-We plan to measure our progress by writing example protocols for these milestones and checking that our implementation supports those protocols by the end of each month.
-
-We will also try to connect with web developers in the community to experiment with our implementation to get feedback on ways to make it more applicable and compatible with industry practices, such that we can apply their feedback on the next iterations of our deliverables.
-
-If time permits, we may explore possible extensions of applying supporting \textit{Explicit Connection Actions} presented in \cite{FASE2017} in our workflow to support more protocols.
-
-\section{Timetable}
-We attach a preliminary timetable (Table \ref{table:timetable}) for guidance.
-
-\begin{figure}
-\centering
-\begin{tabular}{l|p{0.4\textwidth}|p{0.4\textwidth}}
-Month & Milestones & Deadlines \\
-\hline\hline
-Jan & \begin{itemize}
-\item Explore approaches for TypeScript encoding
-\item Investigate methods for code generation from EFSM
-\end{itemize} & \begin{itemize}
-\item 24th: Interim Report
-\item 25th: PLACES 2020
-\end{itemize} \\
-\hline
-Feb & \begin{itemize}
-\item Develop API generation end-to-end toolchain
-\item Support binary sessions with simple send/receive
-\end{itemize} & \begin{itemize}
-\item 14th: Project Review
-\end{itemize} \\
-\hline
-Mar & \begin{itemize}
-\item Support multiparty sessions with simple send/receive
-\end{itemize} & \begin{itemize}
-\item 16th-20th: Examinations
-\end{itemize}\\
-\hline
-Apr & \begin{itemize}
-\item Support binary sessions with selection, choice and recursion
-\item Example: calculator service
-\end{itemize} & \begin{itemize}
-\item 6th-17th: Easter break
-\end{itemize}\\
-\hline
-May & \begin{itemize}
-\item Support multiparty sessions with selection, choice and recursion
-\item Example: Tic Tac Toe game
-\end{itemize} & \begin{itemize}
-\item 15th: Health check-up
-\end{itemize}\\
-\hline
-Jun & \begin{itemize}
-\item Complete report write-up
-\end{itemize} & \begin{itemize}
-\item 15th: Final Report
-\item 26th: Final Archive
-\end{itemize}
-\end{tabular}
-\captionof{table}{Project Timetable.}
-\label{table:timetable}
-\end{figure}
-
-\chapter{Evaluation Plan}
-\section{Theory}
-\section{Implementation}
+\input{sections/evaluationplan.tex}
 

### 2020-01-23 interim/sections/introduction.tex
Interim: update title page and extract sections as separate .tex files

--- 2020-01-21 PLACES_2020/paper.tex
@@
-\documentclass[submission,copyright,creativecommons]{eptcs}
-\providecommand{\event}{PLACES 2020} % Name of the event you are submitting to
-%\usepackage{breakurl}             % Not needed if you use pdflatex only.
-
-\usepackage{wrapfig}
-\usepackage{graphicx}
-\usepackage{subcaption}
-\usepackage{stmaryrd}
-\usepackage[strings]{underscore}
-\usepackage{color}
-\definecolor{purple}{rgb}{0.65, 0.12, 0.82}
-\definecolor{darkgray}{rgb}{.4,.4,.4}
-
-\usepackage{listings}
-\lstdefinelanguage{Scribble}{
-  keywords={module, type, from, as, global,protocol, role, to, choice, at, or, do},
-  keywordstyle=\color{blue}\bfseries,
-  identifierstyle=\color{black},
-  sensitive=false,
-  comment=[l]{//},
-  morecomment=[s]{/*}{*/},
-  commentstyle=\color{darkgray}\ttfamily,
-  stringstyle=\color{purple}\ttfamily,
-  morestring=[b]',
-  morestring=[b]"
-}
-\lstdefinelanguage{JavaScript}{
-  keywords={typeof, new, true, false, catch, function, return, null, catch, switch, var, if, in, while, do, else, case, break, class, export, boolean, throw, implements, import, this, const, let, extends},
-  keywordstyle=\color{blue}\bfseries,
-%  ndkeywords={class, export, boolean, throw, implements, import, this},
-%  ndkeywordstyle=\color{darkgray}\bfseries,
-  identifierstyle=\color{black},
-  sensitive=false,
-  comment=[l]{//},
-  morecomment=[s]{/*}{*/},
-  commentstyle=\color{darkgray}\ttfamily,
-  stringstyle=\color{purple}\ttfamily,
-  morestring=[b]',
-  morestring=[b]"
-}
-\lstset{
-	basicstyle=\footnotesize\ttfamily,
-	numbers=left,
-	frame=single,
-}
-\usepackage{caption}
-
-\newcommand{\sectionref}[1]{\textsection~\ref{#1}}
-
-\title{Multiparty Session Type-safe Web Development in TypeScript}
-\author{Anson Miu
-\institute{Imperial College London}
-\and
-Francisco Ferreira
-\institute{Imperial College London}
-\and
-Nobuko Yoshida
-\institute{Imperial College London}
-\and
-Fangyi Zhou
-\institute{Imperial College London}
-}
-\def\titlerunning{MPST-safe Web Development in TypeScript}
-\def\authorrunning{A. Miu, F. Ferreira, N. Yoshida \& F. Zhou}
-\begin{document}
-\maketitle
-
-\begin{abstract}
-% Situation: webapps, interactive, JavaScript more powerful but remains a poor language for maintaing large applications
-
-% Problem: hard to maintain/test
-% New languages (like Dart and TypeScript)
-
-% Solution: MPST for web development
-
-This is a sentence in the abstract.
-This is another sentence in the abstract.
-This is yet another sentence in the abstract.
-This is the final sentence in the abstract.
-
-% Problem: similar to PureScript paper
-
-% In this paper, we present...
-
-% TypeScript
-% - gradual type system, superset of JavaScript
-% - we define how our approach achieves "gradual linearity" (can we define this?)
-%Our work targets TypeScript - a superset of \dots
-
-%We demonstrate the effectiveness of our approach through a web-based game of \textit{Noughts and Crosses} where communication is performed through WebSocket connections.
-
-\end{abstract}
-
-\section{Introduction}
-% Situation: stateful communication over WebSocket and interactive web applications
+\chapter{Introduction}
+% Rise of distributed programs -> main challenges
+% How to ensure correctness in general: type system, data types
+% Type discipline for concurrent programs - behavioural types (--> session types)
+% Mainstream example of distirbuted programs: web services (microservice architecture), interactive web application; 
+% Objective:
 
@@
 
-\paragraph{Contributions}
-This paper presents a workflow for developing type-safe interactive SPAs motivated by the MPST framework: \textbf{(1)} An endpoint API code generation workflow targeting TypeScript-based web applications for multiparty sessions; \textbf{(2)} An encoding of session types in server-side TypeScript that enforces static linearity; and \textbf{(3)} An encoding of session types in browser-side TypeScript using the React framework that guarantees affine usage of communication channels.
-
-%Figure \ref{fig:workflow} illustrates our proposed development workflow: we extend the Scribble framework (\sectionref{section:scribble}) and generate TypeScript endpoints for server-side (\sectionref{section:server}) and browser-side (\sectionref{section:browser}) targets.
-We will use the \textit{Noughts and Crosses} game as our running example in the rest of the paper and show how our approach is a practical compromise that combines benefits from static session typing and being compatible with common libraries and idiomatic JavaScript practices (\sectionref{section:example}).
-
-\section{The Scribble Framework}
-\label{section:scribble}
-
-Development begins by specifying the permitted communications between participants as a \textit{global protocol} in Scribble, a MPST-based framework introduced in \cite{Scribble} which features a protocol specification language and code generation toolchain.
-This means that a protocol specified in Scribble maps to some \textit{global type} in multiparty session type theory \cite{MPST}.
-We specify the \textit{Noughts and Crosses} game as a Scribble protocol in Listing \ref{lst:game}.
-
-\begin{figure}[!h]
-\begin{lstlisting}[
-	language=Scribble
-]
-module NoughtsAndCrosses;
-type <typescript> "Coordinate" from "./Types" as Point;	// Position on board
-
-global protocol Game(role Svr, role P1, role P2) {
-  Pos(Point) from P1 to Svr;
-  choice at Svr {
-    Lose(Point) from Svr to P2;
-    Win(Point) from Svr to P1;
-  } or {
-    Draw(Point) from Svr to P2;
-    Draw(Point) from Svr to P1;
-  } or {
-    Update(Point) from Svr to P2;
-    Update(Point) from Svr to P1;
-    do Game(Svr, P2, P1);
-  }
-}
-\end{lstlisting}
-\captionof{lstlisting}{Main body of the \textit{Noughts and Crosses} protocol.}
-\label{lst:game}
-\end{figure}
-
-We leverage the Scribble toolchain to check for protocol well-formedness.
-This directly corresponds to the multiparty session type theory of verifying valid local type \textit{projections} for all participants of a global type.
-The implementation of this validation by the toolchain follow from the algorithmic projection procedures defined in \cite{MPST}.
-We obtain \textit{endpoint protocols} for each role in a well-formed global protocol.
-An endpoint protocol only preserves the interactions defined by the global protocol in which the target role is involved, and corresponds to an equivalent \textit{Endpoint Finite State Machine} (EFSM) in the Scribble toolchain.
-We use the EFSMs as a basis for API generation and adopt the formalisms in \cite{Hybrid2016}.
-
-\section{Encoding Session Types in TypeScript}
-% Design choices
-
-Developers can validate the communication aspects of their implementation against their EFSM to verify protocol conformance.
-Our approach integrates the EFSM into the development workflow by encoding session types as TypeScript types.
-We appreciate the inherent differences in development considerations between front- and back-end web development: business logic tends to initiate communication in the back-end, whilst communication in the front-end is mostly driven by user interactions with the browser.
-This motivates our design choice of generating different session types encodings for the target application on the server (\sectionref{section:server}) and in the browser (\sectionref{section:browser}).
-
-\subsection{Server-side API generation}
-\label{section:server}
-
-\begin{wrapfigure}{r}{0.5\textwidth}
-  \begin{center}
-    \includegraphics[width=0.48\textwidth]{figures/efsm_svr.png}
-  \end{center}
-
-  \captionof{figure}{EFSM for \texttt{Svr} in \textit{Noughts and Crosses}.}
-  \label{fig:efsmsvr}
-\end{wrapfigure}
-
-We will refer to the \texttt{Svr} EFSM (fig.~\ref{fig:efsmsvr}) as a running example in this section.
-For server-side targets, we encode EFSM states as TypeScript types depending on the type of state.
-We adopt the EFSM definition presented in \cite{Hybrid2016} to consider receive and send states separately.
-We assign each TypeScript encoding its state identifier from the EFSM as its type alias, thus providing syntactic sugar when referring to the successor state in the TypeScript encoding of the current state.
-For any state $S$ in the EFSM, we refer to the TypeScript type alias of its encoding as $\llbracket S \rrbracket$.
-
-We make a key design decision \textit{not} to expose communication channels in the TypeScript session type encodings to provide linearity guarantees (\sectionref{section:serverlinear}).
-Our encodings sufficiently exposes seams for the developer to inject their business logic, whilst the generated session API (\sectionref{section:serversessionapi}) handles the sending and receiving of messages; as a result, the encodings do not concern with the role involved in the send/receive action.
-We outline the encodings below using examples from the \textit{Noughts and Crosses} game server (Listing \ref{lst:svr}).
-
-\paragraph{Branching state} We consider a receive state as a unary branching state for conciseness.
-A branching state is encoded as an \textit{object literal} \cite{TypeScriptSpec}, with each branch corresponding to a member field.
-A branch expecting to receive a message labelled $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as an \textit{member field} named $\texttt{label}_i$ of function type \texttt{(payload:$\texttt{T}_i$) => $\llbracket S_i \rrbracket$}.
-The developer implements a branching operation by passing callbacks for each branch, parameterised by the expected message payload type for that branch.
-
-\paragraph{Selection state} We consider a send state as a unary selection state for conciseness.
-A selection state is encoded as a \textit{union type} \cite{TypeScriptSpec} of internal choice encodings: each internal choice sending a message labelled $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as a \textit{tuple type} of \texttt{[Labels.label$_i$, T$_i$, $\llbracket S_i \rrbracket$]}.
-The developer implements a selection operation by passing the selected label and payload to send in the message.
-We generate a \textit{string enum} (named \texttt{Labels}) wrapping the labels in the protocol, hence the enum member access in the first element of the tuple type.
-
-\begin{figure}[!h]
-\begin{lstlisting}[language=JavaScript]
-import { Coordinate as Point } from "./Types";
-import { Labels } from "./Constants";
-
-export type S13 = { Pos: (payload: Point) => S15 }
-export type S15 = [Labels.Lose, Point, S16]
-		| [Labels.Draw, Point, S17]
-		| [Labels.Update, Point, S18]
-\end{lstlisting}  
-\captionof{lstlisting}{Example encodings from \textit{Noughts and Crosses} \texttt{Svr} EFSM.}
-\label{lst:svr}
-\end{figure}
-
-\subsubsection{Session Runtime and Protocol API}
-\label{section:serversessionapi}
-
-Our session runtime performs communication in a way that conforms to the protocol specification.
-The runtime listens to message (receive) events on the communication channel, invokes the corresponding callback to obtain the value to send next, and performs the send.
-
-The \textit{Protocol API} is a class that interfaces between the session runtime and the user implementation of the EFSM state encodings.
-Upon construction, the Protocol class waits for all client roles to connect via WebSocket, then proceeds to the initial state of the EFSM and performs state transitions depending on the permitted communication patterns from the EFSM.
-
-\subsubsection{Linear channel usage}
-\label{section:serverlinear}
-Our callback-oriented design for server-side API generation provides guarantees on state channel linearity by preventing the two properties detailed below.
-Here, channel linearity is guaranteed based on the correctness of our library design and session runtime implementation, which means a faulty implementation could violate this, but is up to the library author to verify once rather than the end user.
-
-\paragraph{Repeat use} We do not expose channels to the programmer, which makes \textit{reuse} impossible.
-For example, to send a message, the generated API only requires the payload that needs to be sent, and the session runtime performs the send internally, guaranteeing this action is done \textit{exactly once} by construction.
-
-\paragraph{Unused} The initial state must be supplied to the Protocol API constructor in order to instantiate the session; this initial state is defined in terms of the successor states, which in turn has references to its successors and so forth.
-This encoding approach will cover the terminal state (if it exists in the EFSM, as \cite{Hybrid2016} notes that an EFSM contains at most one terminal state), and the session runtime guarantees this terminal state, if it exists, will be reached by construction.
-
-\subsection{Browser-side API generation}
-\label{section:browser}
-
-We will refer to the \texttt{P1} EFSM (figure \ref{fig:efsmp1}) as a running example in this section.
-Preserving behavioural typing and channel linearity are known to be challenging for browser-side applications due to the event-driven nature of user interaction: in the case of \textit{Noughts and Crosses}, once the user makes a move by clicking on a cell on the game board, this click event must be deactivated until the user's next turn, otherwise the user can click again and violate channel linearity.
-Our design goal is to enforce this statically through the APIs we generate.
-
-For browser-side targets, we extend the approach presented in \cite{MVU2019} on \textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU) architecture.
-Each state in the EFSM is a model type and uniquely defines a \textit{view function}, set of \textit{messages} and \textit{update function}: the view function defines what to render to the DOM, the set of messages define the possible (IO) actions available at that state, and the update function defines which successor state to transition to, given some supported IO action at this state.
-
-\begin{wrapfigure}{r}{0.5\textwidth}
-  \begin{center}
-    \includegraphics[width=0.48\textwidth]{figures/efsm_p1.png}
-  \end{center}
-
-  \captionof{figure}{EFSM for \texttt{P1} in \textit{Noughts and Crosses}.}
-  \label{fig:efsmp1}
-\end{wrapfigure}
-
-\subsubsection{The React Framework}
-Rather than using the LINKS web programming language from \cite{MVU2019}, we apply the multiple model types approach to the React framework \cite{React} developed by Facebook.
-React is widely used in industry to create scalable single-page TypeScript applications, and we intend for our proposed workflow to be beneficial in an industrial context.
-We introduce the key features of the framework below.
-
-\paragraph{Components} A component is a reusable UI element which contains its own markup and logic.
-Components implement a \texttt{render()} function which returns a React element, the smallest building blocks of a React application.
-Components can keep \textit{state} and the \texttt{render()} function is invoked upon a change of state.
-A simple counter can be implemented as a component, with the count stored as state, a button which increments the count when clicked and a \texttt{div} that renders the current count.
-Components can also render other components, which gives rise to a parent/child relationship between components.
-Parents can pass data to children as \textit{props}: going back to the aforementioned example, the the counter component could render a child component \texttt{<StyledDiv count=\{this.state.count\} />} in its \texttt{render()} function, propagating the \texttt{count} from its state to the child.
-This enables reusability, and for our use case, gives control to the parent on what data to pass to its children (e.g.
-pass the payload of a received message to a child to render).
-
-\paragraph{Virtual DOM (VDOM)} React components are rendered on a logical abstraction of the DOM, which in turn performs a \texttt{diff} on the current browser DOM and patches the delta accordingly.
-This allows the programmer to directly specify what should be rendered without having to worry about which elements to append or remove from the browser DOM.
-
-\subsubsection{Model types in React}
-We leverage React to express the model types approach in \cite{MVU2019} rather than the \textit{Model-View-Controller} (MVC) architecture it was intended for.
-
-\paragraph{State} An EFSM state is encoded as an \textit{abstract} React component.
-This is an abstract class to require the programmer to provide their own view function, which translates conveniently to the \texttt{render()} function of React components.
-Our session runtime (\sectionref{section:clientruntime}) ``executes'' the EFSM and renders the current state, so upon transitioning to a successor state, the successor's view function will be invoked, as per the semantics expressed in \cite{MVU2019}.
-
-\paragraph{Model transitions} Transitions are encoded as React component props onto the encoded states by the session runtime (\sectionref{section:clientruntime}).
-We motivate the design choice of not exposing channel resources to provide guarantees on channel usage.
-React components in TypeScript are \textit{generics} \cite{TypeScriptSpec} which are parameterised by the permitted types of prop and state, which allows us to leverage the TypeScript compiler to verify that the props for model transitions stay local to the state they are defined for.
-The model transitions for EFSMs are message send and receive.
-
-\subparagraph{Send} We make the assumption that message sending is triggered by some user-driven UI event (e.g.
-clicking a button, pressing a key on the keyboard) which interacts with some DOM element.
-We could pass a \texttt{send()} function as a prop to the sending state, but the programmer would be free to call the function multiple times which makes channel reuse possible.
-Instead, we pass a \textit{factory function} as a prop, which will, given an HTML event and a event handler function, return a fresh React component that binds the sending action on construction, so once the bound event is triggered, our session runtime executes the event handler function to obtain the payload to send, perform the send \textit{exactly once} and transition to (which, in practice, means render) the successor state.
-
-\begin{figure}[!h]
-\begin{lstlisting}[language=JavaScript, tabsize=4]
-render() {
-	...
-	{board.map((row, x) => (
-		board.map((col, y) => {
-			const SelectPoint = this.props.Pos('click', (event: UIEvent) => {
-				event.preventDefault()
-				return { x, y }
-			}
-			return <SelectPoint><td>.</td></SelectPoint>
-		}
-	)}
-	...	
-}
-\end{lstlisting}  
-\captionof{lstlisting}{Model transition for message sending in 
-\textit{Noughts and Crosses} \texttt{P1} implementation.}
-\label{lst:clientapp}
-\end{figure}
-
-We clarify the semantics using the \textit{Noughts and Crosses} example in Listing \ref{lst:clientapp}.
-\texttt{this.props.Pos} is the factory function prop passed from the session runtime.
-For each x-y coordinate on the game board, we create a \texttt{SelectPoint} React component from the factory function (which reads ``build a React component that sends the \texttt{Pos} message with x-y coordinates as payload when the user clicks on it'') and we wrap a table cell (the game board is rendered as a HTML table) inside the \texttt{SelectPoint} component to bind the click event on the table cell.
-
-\subparagraph{Receive} The payload of a message receive transition is passed as a prop to the successor state and can be handled in the successor's view function.
-Upon receiving the message from the channel, the session runtime renders the successor of the receive state and propagates the payload as the prop accordingly.
-
-\subsubsection{Session runtime}
-\label{section:clientruntime}
-
-Our session runtime can be interpreted as an abstraction on top of the React VDOM that implements the EFSM by construction.
-The session runtime itself is a React component too, named after the endpoint role identifier: it opens the WebSocket connection to the server, keeps track of the current EFSM state as part of its React component state, and most importantly, renders the React component encoding of the active EFSM state.
-Channel communications are managed by the runtime, which allows it to render the successor of a receive state upon receiving a message from the channel.
-Similarly, the session runtime is responsible for passing the required props for model transitions to EFSM state React components.
-The session runtime component is rendered by the programmer and requires the following props:
-
-\paragraph{Endpoint URL} The session runtime needs this in order to open the WebSocket connection.
-
-\paragraph{Concrete state components} The programmer writes their own implementation of each state (mainly to customise how the state is rendered and inject business logic into state transitions) by extending the abstract React class components.
-The session runtime requires references to these concrete components order to render the user implementation accordingly.
-
-\subsubsection{Affine channel usage}
-A limitation of our browser-side session type encoding is only being able to guarantee that channel resources are used \textit{at most once} as supposed to \textit{exactly once}.
-
-Communication channels are not exposed to the programmer so multiple sends are impossible.
-This does not restrict the programmer from binding the send action to exactly one UI event: for the \textit{Noughts and Crosses} game, we bind the \texttt{Pos(Point)} send action to each unoccupied cell on the game board, but the generated runtime ensures that, once an event bound to the send action is triggered, the send is only performed once and the successor state is rendered on the DOM to prevent multiple sends initiated by the user.
-
-However, our approach \textit{does not} statically detect whether all transitions in a certain state are bound to some UI event.
-This means that it is possible for an implementation to \textit{not} handle transitions to a terminal state but still type-check, so we cannot prevent unused states.
-
-\section{Case Study}
-\label{section:example}
-
-\begin{figure}
-\begin{lstlisting}[language=JavaScript, tabsize=4]
-const handleP1Move: Receive_P1 = (move: Point) => {
-	// User-implemented logic
-	board.P1(move)
-	if (board.won()) { return [Labels.Lose, move, move] }
-	else if (board.draw()) { return [Labels.Draw, move, move] }
-	else { return [Labels.Draw, move, handleP2Move] } 
-}
-
-const handleP2Move: Receive_P2 = ...
-
-// Instantiate session
-new NoughtsAndCrosses.Svr(webSocketServer, handleP1Move, handleP2Move)
-\end{lstlisting}  
-\captionof{lstlisting}{Example fragment of \textit{Noughts and Crosses} \texttt{Svr} implementation.}
-\label{lst:svrprotocol}
-\end{figure}
-
-We apply our framework to implement a web-based implementation of the \textit{Noughts and Crosses} running example in TypeScript; the interested reader can find the full implementation in \cite{NoughtsAndCrosses}.
-In addition to MPST-safety, we show that our library design welcomes idiomatic JavaScript practices in the user implementation and is interoperable with common front- and back-end frameworks.
-
-\paragraph{Game server} We set up the WebSocket server inside an Express.js \cite{ExpressJS} application on top of a Node.js \cite{NodeJS} runtime.
-We define our own game logic in a \texttt{Board} class to keep track of the game state and expose methods to query the result.
-This custom logic is integrated into our \texttt{Receive\_P1} and \texttt{Receive\_P2} handlers (Listing \ref{lst:svrprotocol}), so the session runtime can handle \texttt{Pos(Point)} messages from players and transition according to the injected game logic.
-
-Note that, by the structural typing nature of TypeScript, replacing \texttt{handleP2Move} on line 6 with \texttt{handleP1Move} would be type-correct - this allows for better code reuse as supposed to defining additionl abstractions to work around the limitations of nominal typing in \cite{Hybrid2016}.
-There is also full type erasure when transpiling to JavaScript to run the server code, so state space explosions is never a runtime consideration.
-
-\paragraph{Game clients} We implement the game client for players as a React application - extending from the generated abstract React components, registering our custom implementation as props for the session runtime component.
-
-For the sake of code reuse for demonstration purposes, \cite{NoughtsAndCrosses} uses \textit{higher-order components} (HOC) to build the correct state implementations depending on which player the user chooses to be.
-We leverage the \textit{Redux} state management library to keep track of game state, thus showing the flexibility of our library design in being interoperable with other libraries and idiomatic JavaScript practices.
-
-\section{Related Works}
-%\paragraph{Languages with native session type support} \dots
-% New languages written  (e.g. ATS)
-%	- addresses the challenge of resource linearity with a liner type system
-%	- problem with webdev: not compatible with libraries, technologies, flexibility of JavaScript (ultimately the language of the browser)
-
-\paragraph{Endpoint API generation} \cite{Python2017} targeted Python applications and the generation of runtime monitors to dynamically verify communication patterns.
-Whilst the same approach could be applied to JavaScript, we can provide more static guarantees with TypeScript's gradual typing system and compiler.
-\cite{Hybrid2016} targeted Java applications and proposed to encode the EFSM states and transitions as classes and instance methods respectively, with behavioural typing achieved statically by the type system and channel linearity guarantees achieved dynamically since channel resources are exposed and aliasing is not monitored.
-
-\paragraph{Session types in web development} \cite{PureScript2019} targeted web development in PureScript using the \textit{Concur UI} framework and proposed a type-level encoding of EFSMs as multi-parameter type classes.
-However, it presents a trade-off between achieving static linearity guarantees from the type-level EFSM encoding under the expressive type system and providing an intuitive development experience to programmers, especially given the prevalence of JavaScript and TypeScript applications in industry.
-\cite{MVU2019} focused on applying binary session types in front-end web development and presented approaches that tackle the challenge of guaranteeing linearity in the event-driven environment.
-
-Our work applies the aforementioned approaches in a \textit{multiparty} context in a manner that is compatible with industrial tools and practices to ultimately invite the notion of MPST-safety into interactive web application development in industry.
-
-\section{Conclusion and Future Works}
-We have presented a MPST-based framework for developing full-stack interactive TypeScript applications with WebSocket communications that conform to a protocol, statically providing linear channel usage guarantees and affine channel usage guarantees for back-end and front-end targets respectively.
-
-Future works include incorporating \textit{explicit connection actions} introduced in \cite{ExplicitConnections} in our API generation to better model real-world communication protocols that may feature in interactive web applications.
-Whilst our approach does support multiparty sessions, protocols are required to have exactly one role running on the server and client roles must interact with each other via the server to respect the properties of WebSocket connections, which is the transport abstraction we support.
-Extending support to WebRTC would cater for peer-to-peer communication between browsers, which further opens up possibilities for communication protocols supported by our approach.
-
-% Future work
-% - generalise to encode protocols as typestates (e.g. Mungo example)
-% - WebRTC for full MPST (since current limitation is w.r.t. MPST protocols with 1-server many-client roles
-
-\nocite{*}
-\bibliographystyle{eptcs}
-\bibliography{generic}
-\end{document}
+We present a workflow for developing type-safe interactive SPAs motivated by the MPST framework: \textbf{(1)} An endpoint API code generation workflow targeting TypeScript-based web applications for multiparty sessions; \textbf{(2)} An encoding of session types in server-side TypeScript that enforces static linearity; and \textbf{(3)} An encoding of session types in browser-side TypeScript using the React framework that guarantees affine usage of communication channels.

### 2020-01-23 interim/sections/projectplan.tex
Interim: update title page and extract sections as separate .tex files

\chapter{Project Plan}
We aim to develop a multiparty session type-safe development workflow for building interactive full-stack TypeScript applications that conform to a communication protocol. This involves encoding session types using the TypeScript language and implementing a code generation workflow to generate the encodings.

\section{Delivery}
At the point of writing, we have analysed the existing code generation approaches presented in \cite{Hybrid2016, Scribble, Python2017} and assessed their applicability with respect to verifying the communication aspects of interactive web applications. We have also explored approaches for encoding session types into TypeScript, motivated by existing works along with similar proposals specific to web development in \cite{PureScript2019, MVU2019}.

We plan to deliver our code generation implementation incrementally, aiming to get a working version of the end-to-end workflow first, then iteratively add support for more complex session type primitives. This minimises the risks involved in the project by ensuring we have a functional deliverable at the early stages of the project. We plan to deliver this basic end-to-end workflow by the end of February.

The subsequent months will involve adding support for other primitives (i.e. selection, choice and recursion) for multiparty sessions. We plan to measure our progress by writing example protocols for these milestones and checking that our implementation supports those protocols by the end of each month.

We will also try to connect with web developers in the community to experiment with our implementation to get feedback on ways to make it more applicable and compatible with industry practices, such that we can apply their feedback on the next iterations of our deliverables.

If time permits, we may explore possible extensions of applying supporting \textit{Explicit Connection Actions} presented in \cite{FASE2017} in our workflow to support more protocols.

\section{Timetable}
We attach a preliminary timetable (Table \ref{table:timetable}) for guidance.

\begin{figure}
\centering
\begin{tabular}{l|p{0.4\textwidth}|p{0.4\textwidth}}
Month & Milestones & Deadlines \\
\hline\hline
Jan & \begin{itemize}
\item Explore approaches for TypeScript encoding
\item Investigate methods for code generation from EFSM
\end{itemize} & \begin{itemize}
\item 24th: Interim Report
\item 25th: PLACES 2020
\end{itemize} \\
\hline
Feb & \begin{itemize}
\item Develop API generation end-to-end toolchain
\item Support binary sessions with simple send/receive
\end{itemize} & \begin{itemize}
\item 14th: Project Review
\end{itemize} \\
\hline
Mar & \begin{itemize}
\item Support multiparty sessions with simple send/receive
\end{itemize} & \begin{itemize}
\item 16th-20th: Examinations
\end{itemize}\\
\hline
Apr & \begin{itemize}
\item Support binary sessions with selection, choice and recursion
\item Example: calculator service
\end{itemize} & \begin{itemize}
\item 6th-17th: Easter break
\end{itemize}\\
\hline
May & \begin{itemize}
\item Support multiparty sessions with selection, choice and recursion
\item Example: Tic Tac Toe game
\end{itemize} & \begin{itemize}
\item 15th: Health check-up
\end{itemize}\\
\hline
Jun & \begin{itemize}
\item Complete report write-up
\end{itemize} & \begin{itemize}
\item 15th: Final Report
\item 26th: Final Archive
\end{itemize}
\end{tabular}
\captionof{table}{Project Timetable.}
\label{table:timetable}
\end{figure}

### 2020-01-23 PLACES_2020/main.tex
FZ comments

@@
 \usepackage{color}
+\usepackage{etoolbox}
+\newtoggle{draft}
 \definecolor{purple}{rgb}{0.65, 0.12, 0.82}
@@
 
+\toggletrue{draft}
+\newcommand{\fz}[1]{\iftoggle{draft}{{\color{red}FZ: #1}}{}}
+
 \title{Verifying Interactive WebSocket Applications in TypeScript}
@@
 \begin{abstract}
-Advancements in mobile device compute power has made interactive web applications possible, allowing the web browser to render content dynamically and support low-latency communication with the server.
+Advancements in mobile device compute power have made interactive web
+applications possible, allowing the web browser to render contents dynamically and support low-latency communication with the server.
 This comes at a cost to the developer, who now needs to reason more about the communication correctness of their application as they support more complex communication patterns.
 
-Multiparty session types (MPST) provide a framework for verifying implementations to conform to a prescribed communication protocol.
-Existing proposals for applying the MPST framework in application development either neglect the event-driven nature of web applications, or lack compatibility with industry tools and practices which discourage mainstream adoption by web developers.
+Multiparty session types (MPST) provide a framework for verifying conformance
+of implementations to their prescribed communication protocol. Existing
+proposals for applying the MPST framework in application development either
+neglect the event-driven nature of web applications, or lack compatibility with
+industry tools and practices which discourages mainstream adoption by web developers.
 
@@
 type theory \cite{MPST}:
+\fz{Cite the featherweight scribble paper for the correspondence?}
 a Scribble protocol maps to some \textit{global type}, and protocol verification implements the algorithmic projection
@@
 
+\fz{Mention correctness by construction instead of verifying conformance?}
 Developers can validate the communication aspects of their
@@
 
+\fz{I think there is an important thing to note here, ``Server'' side is
+  basically the side which accepts the connection and ``Client'' side is the
+side which makes the connection. But this is orthogonal to whether it runs on
+Node or on browser? (Correct me if I'm wrong)}
+
 \subsection{Server-side API generation}
+\fz{Could you remove the state names in figures?}
 \label{section:server}
@@
 
-We will refer to the \texttt{Svr} EFSM (fig.~\ref{fig:efsmsvr}) as a
-running example in this section. For server-side targets, we encode
+We refer to the \texttt{Svr} EFSM (fig.~\ref{fig:efsmsvr}) as a
+running example in this section. \fz{What do you try to say in the next
+  sentence?} For server-side targets, we encode
 EFSM states as TypeScript types depending on the type of state. We
@@
 export type S13 = { Pos: (payload: Point) => S15 };
-export type S15 = [Labels.Lose, Point, S16] | [Labels.Draw, Point, S17]
+export type S15 = [Labels.Lose, Point, S16]
+                | [Labels.Draw, Point, S17]
                 | [Labels.Update, Point, S18];
@@
 
-\paragraph{Unused} The initial state must be supplied to the session runtime constructor in order to instantiate the session; this initial state is defined in terms of the successor states, which in turn has references to its successors and so forth.
-This encoding approach will cover the terminal state (if it exists), and the session runtime guarantees this terminal state, if it exists, will be reached by construction.
+\paragraph{Unused} The initial state must be supplied to the session runtime
+constructor in order to instantiate the session; this initial state is defined
+in terms of the successor states, which in turn has references to its
+successors and so forth. This encoding approach covers the terminal state (if it exists), and the session runtime guarantees this terminal state, if it exists, will be reached by construction.
 
@@
 
-We will refer to the \texttt{P1} EFSM (figure \ref{fig:efsmp1}) as a
+We refer to the \texttt{P1} EFSM (figure \ref{fig:efsmp1}) as a
 running example in this section. Preserving behavioural typing and
@@
 \subsubsection{The React Framework}
-Rather than using the Links web programming language from \cite{MVU2019}, we apply the multiple model types approach to the React framework \cite{React} developed by Facebook, albeit originally intended for the \textit{Model-View-Controller} (MVC) architecture.
+
+\fz{This subsubsection is a bit misplaced, probably move it up}
+
+We apply the multiple model types approach in \cite{MVU2019} to the React framework \cite{React} developed by Facebook, albeit originally intended for the \textit{Model-View-Controller} (MVC) architecture.
 React is widely used in industry to create scalable single-page TypeScript applications, and we intend for our proposed workflow to be beneficial in an industrial context.
@@
 change of state. A simple counter can be implemented as a component,
-with \texttt{count} stored as state, a button which increments \texttt{count}when clicked and a \texttt{div} that renders the current
+with \texttt{count} stored as state, a button which increments \texttt{count}
+when clicked and a \texttt{div} that renders the current
 \texttt{count}. Components can also render other components, which gives rise
@@
 
+\fz{Is VDOM needed to be covered here?}
 \paragraph{Virtual DOM (VDOM)} React components are rendered on a logical abstraction of the DOM, which in turn performs a \texttt{diff} on the current browser DOM and patches the delta accordingly.
@@
 We motivate the design choice of not exposing channel resources to provide guarantees on channel usage.
-React components in TypeScript are \textit{generics} \cite{TypeScriptSpec} which are parameterised by the permitted types of prop and state, which allows us to leverage the TypeScript compiler to verify that the props for model transitions stay local to the state they are defined for.
+React components in TypeScript are \textit{generic} \cite{TypeScriptSpec} which are parameterised by the permitted types of prop and state, which allows us to leverage the TypeScript compiler to verify that the props for model transitions stay local to the state they are defined for.
 The model transitions for EFSMs are message send and receive.
@@
 This means that it is possible for an implementation to \textit{not} handle transitions to a terminal state but still type-check, so we cannot prevent unused states.
+\fz{A client may also close the browser, hence dropping the connection?}
 
@@
 \section{Related Works}
+\fz{Try to avoid using {[X]} at the beginning of a sentence.}
+
 Native language support for first-class linear channel resources \cite{ATS} and code generation are two main approaches for incorporating MPST into application development.

### 2020-01-23 PLACES_2020/main.tex
PLACES: begin addressing FZ comments

@@
 \newcommand{\fz}[1]{\iftoggle{draft}{{\color{red}FZ: #1}}{}}
+\newcommand{\am}[1]{\iftoggle{draft}{{\color{blue}AM: #1}}{}}
 
@@
 
-\fz{Mention correctness by construction instead of verifying conformance?}
-Developers can validate the communication aspects of their
-implementation against their EFSM to verify protocol conformance. Our
-approach integrates the EFSM into the development workflow by encoding
+Developers can implement their application using APIs generated from the EFSM to guarantee correctness by construction.
+Our approach integrates the EFSM into the development workflow by encoding
 session types as TypeScript types.
@@
 Node or on browser? (Correct me if I'm wrong)}
+
+\am{Not for the case of WebSockets, because only WebSocket clients on browser can initiate connections and only WebSocket Server on Node can accept; will make this distinction clear.}
 
@@
 We refer to the \texttt{Svr} EFSM (fig.~\ref{fig:efsmsvr}) as a
-running example in this section. \fz{What do you try to say in the next
-  sentence?} For server-side targets, we encode
-EFSM states as TypeScript types depending on the type of state. We
-adopt the EFSM definition presented in \cite{Hybrid2016} to consider
+running example in this section. For server-side targets, we encode
+EFSM states as types, considering
 receive and send states separately. Each TypeScript encoding is

### 2020-01-23 PLACES_2020/main.tex
PLACES: update inrtoduction

@@
 Modern interactive web applications aim to provide a highly responsive user experience by minimising the communication latency between clients and servers.
-Whilst the HTTP request-response model is sufficient for retrieving static assets, applying the same stateless communication approach for interactive use cases (such as a multiplayer game) introduces undesirable performance overhead from having to frequently set up new connections for client-server interactions.
+Whilst the HTTP request-response model is sufficient for retrieving static assets, applying the same stateless communication approach for interactive use cases (such as real-time multiplayer games) introduces undesirable performance overhead from having to frequently set up new connections for client-server interactions.
 Developers have since adopted other communication transport abstractions over HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy low-latency full-duplex client-server communication in their applications over a single persistent connection.
-Enabling more complex communication patterns caters for more interactive use cases (such as real-time multiplayer games), but introduces additional correctness concerns to the developer.
+Enabling more complex communication patterns caters for more interactive use cases, but introduces additional correctness concerns to the developer.
 

### 2020-01-24 PLACES_2020/main.tex
More FZ comments, latex file formatting

@@
 \usepackage{listings}
+\usepackage[noabbrev, capitalise]{cleveref}
 \lstdefinelanguage{Scribble}{
@@
 Advancements in mobile device compute power have made interactive web
-applications possible, allowing the web browser to render contents dynamically and support low-latency communication with the server.
-This comes at a cost to the developer, who now needs to reason more about the communication correctness of their application as they support more complex communication patterns.
+applications possible, allowing the web browser to render contents dynamically
+and support low-latency communication with the server.
+This comes at a cost to the developer, who now needs to reason more about the
+communication correctness of their application as they support more complex
+communication patterns.
 
 Multiparty session types (MPST) provide a framework for verifying conformance
-of implementations to their prescribed communication protocol. Existing
-proposals for applying the MPST framework in application development either
-neglect the event-driven nature of web applications, or lack compatibility with
-industry tools and practices which discourages mainstream adoption by web developers.
+of implementations to their prescribed communication protocol.
+Existing proposals for applying the MPST framework in application development
+either neglect the event-driven nature of web applications, or lack
+compatibility with industry tools and practices, which discourages mainstream
+adoption by web developers.
 
-In this paper, we present an implementation of the MPST framework for developing interactive web applications using familiar industry tools in TypeScript and the \textit{React.js} framework.
-The global communication protocol is specified in the Scribble protocol language and checked for well-formedness by the Scribble toolchain.
-We obtain the local protocol for each role, which describes the interactions of the global communication protocols observed by the role.
-We encode the local protocol into TypeScript types, catering for server-side and client-side targets separately, and show that our encoding guarantees that only implementations which conform to the protocol will type-check.
-We demonstrate the effectiveness of our approach through a web-based implementation of the classic \textit{Noughts and Crosses} game from a MPST formalism of the game logic.
+In this paper, we present an implementation of the MPST framework for
+developing interactive web applications using familiar industry tools in
+TypeScript and the \textit{React.js} framework.
+The developer can use the Scribble protocol
+language to specify the protocol and use the Scribble toolchain to validate
+and obtain the local protocol for each role.
+The local protocol describes the interactions of the global communication
+protocols observed by the role.
+We encode the local protocol into TypeScript types, catering for server-side
+and client-side targets separately, and show that our encoding guarantees that
+only implementations which conform to the protocol can type-check.
+We demonstrate the effectiveness of our approach through a web-based
+implementation of the classic \textit{Noughts and Crosses} game from a MPST
+formalism of the game logic.
 
@@
 
-Modern interactive web applications aim to provide a highly responsive user experience by minimising the communication latency between clients and servers.
-Whilst the HTTP request-response model is sufficient for retrieving static assets, applying the same stateless communication approach for interactive use cases (such as real-time multiplayer games) introduces undesirable performance overhead from having to frequently set up new connections for client-server interactions.
-Developers have since adopted other communication transport abstractions over HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy low-latency full-duplex client-server communication in their applications over a single persistent connection.
-Enabling more complex communication patterns caters for more interactive use cases, but introduces additional correctness concerns to the developer.
+Modern interactive web applications aim to provide a highly responsive user
+experience by minimising the communication latency between clients and servers.
+Whilst the HTTP request-response model is sufficient for retrieving static
+assets, applying the same stateless communication approach for interactive use
+cases (such as real-time multiplayer games) introduces undesirable performance
+overhead.
+Developers have since adopted other communication transport abstractions over
+HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy
+low-latency full-duplex client-server communication in their applications over
+a single persistent connection.
+Enabling more complex communication patterns caters for more interactive use
+cases, but introduces additional correctness concerns to the developer.
 
 % Example: noughts and crosses
-Consider a classic turn-based board game of \textit{Noughts and Crosses} between two players.
-Both players are identified by either noughts or crosses respectively, and take turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player wins (when their markers form one straight line on the board) or a stalemate is reached (when all cells are occupied).
-A web-based implementation may involve players connected to a game server via WebSocket connections and interacting with the game from their web browser, which serve a \textit{single-page application} (SPA) of the game client written in a popular framework like \textit{React.js} \cite{React}.
-SPAs feature a single HTML page and dynamically renders content via JavaScript in the browser.
-Players take turns to make a move on the game board and the server implements the game logic to progress the game forward until a result (either a win/loss or draw) can be declared.
+Consider a classic turn-based board game of \textit{Noughts and Crosses}
+between two players.
+Both players, identified by either noughts or crosses respectively, take
+turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player
+wins (when their markers form one straight line on the board) or a stalemate is
+reached (when all cells are occupied).
+A web-based implementation may involve players connected to a game server via
+WebSocket connections.
+The players interact with the game from their web browser, which shows a
+\textit{single-page application} (SPA) of the game client written in a popular
+framework like \textit{React.js} \cite{React}.
+SPAs feature a single HTML page and dynamically rendered content via JavaScript
+in the browser.
+Players take turns to make a move on the game board, which sends a message to
+the server, and the server implements the game logic to progress the game
+forward until a result (either a win/loss or draw) can be declared, where
+either the move of the other player or the game result is sent to players.
 
-Whilst WebSockets make this web-based implementation possible, it introduces the developer to a new family of communication errors, even for this simple game.
-In addition to the usual testing for game logic correctness, the developer needs to test against \textit{deadlocks} (e.g.
-both players waiting for each other to make a move at the same time) and \textit{communication mismatches} (e.g.
-player 1 sending a boolean to the game server instead of the board coordinates).
-The complexity of these errors correlate to the complexity of related tests required, scale with the complexity of communication patterns involved.
+Whilst WebSockets make this web-based implementation possible, it introduces
+the developer to a new family of communication errors, even for this simple
+game.
+In addition to the usual testing for game logic correctness, the
+developer needs to test against \textit{deadlocks} (e.g. both players waiting
+for each other to make a move at the same time) and \textit{communication
+ mismatches} (e.g. player 1 sending a boolean to the game server instead of
+the board coordinates).
+The complexity of these errors correlate to the
+complexity of related tests required, scale with the complexity of
+communication patterns involved.
 
-\textit{Multiparty Session Types} (MPST) \cite{MPST} provide a framework for formally specifying a structured communication pattern between concurrent processes and verifying implementations for correctness with respect to the communications aspect.
-By specifying the client-server interactions of our game as a protocol and verifying the implementations against the protocol for well-formedness, MPST theory guarantees well-formed implementations to be free from communication errors.
+\textit{Multiparty Session Types} (MPST) \cite{MPST} provide a framework for
+formally specifying a structured communication pattern between concurrent
+processes and verifying implementations for correctness with respect to the
+communications aspect.
+By specifying the client-server interactions of our game as an MPST protocol
+and verifying the implementations against the protocol for well-formedness,
+MPST theory guarantees well-formed implementations to be free from
+communication errors.
 
-We see the application of the MPST methodology to verifying
-interactive web applications written in TypeScript to be an
-interesting design space -- to what extent can the MPST methodology be
-applied such that we leverage TypeScript language features to maximise
-static safety guarantees whilst preserving a flexible, natural and
-idiomatic workflow for TypeScript developers? Such a workflow would
-ultimately decrease the overhead for incorporating MPST into
-mainstream web system, which reduces development time by
-programmatically
+We see the application of the MPST methodology to verifying interactive web
+applications written in TypeScript to be an interesting design space -- to what
+extent can the MPST methodology be applied such that we leverage TypeScript
+language features to maximise static safety guarantees whilst preserving a
+flexible, natural and idiomatic workflow for TypeScript developers?
+Such a workflow would ultimately decrease the overhead for incorporating MPST
+into mainstream web system, which reduces development time by programmatically
 verifying communication correctness of the implementation.
@@
 \paragraph{Contributions}
-This paper presents a workflow for developing type-safe interactive SPAs motivated by the MPST framework: \textbf{(1)} An endpoint API code generation workflow targeting TypeScript-based web applications for multiparty sessions; \textbf{(2)} An encoding of session types in server-side TypeScript that enforces static linearity; and \textbf{(3)} An encoding of session types in browser-side TypeScript using the React framework that guarantees affine usage of communication channels.
+This paper presents a workflow for developing type-safe interactive SPAs
+motivated by the MPST framework:
+\textbf{(1)} An endpoint API code generation
+workflow targeting TypeScript-based web applications for multiparty sessions;
+\textbf{(2)} An encoding of session types in server-side TypeScript that
+enforces static linearity;
+and \textbf{(3)} An encoding of session types in
+browser-side TypeScript using the React framework that guarantees affine usage
+of communication channels.
 
@@
 
-Development begins by specifying the permitted communications between participants as a \textit{global protocol} in Scribble \cite{Scribble}, a MPST-based protocol specification language and code generation toolchain.
-We specify the \textit{Noughts and Crosses} game as a Scribble protocol in Listing \ref{lst:game}.
+Development begins by specifying the permitted communications between
+participants as a \textit{global protocol} in Scribble \cite{Scribble}, a
+MPST-based protocol specification language and code generation toolchain.
+We specify the \textit{Noughts and Crosses} game as a Scribble protocol in
+\cref{lst:game}.
 
@@
     Update(Point) from Svr to P2; Update(Point) from Svr to P1;
-    do Game(Svr, P2, P1);
+    do Game(Svr, P2, P1); // Continue the game with player roles swapped
   }
@@
 \fz{Cite the featherweight scribble paper for the correspondence?}
-a Scribble protocol maps to some \textit{global type}, and protocol verification implements the algorithmic projection
-defined in \cite{MPST} to derive valid local type \textit{projections} for all participants.  We obtain a set of \textit{endpoint protocols}
--- one for each role from a well-formed global protocol.
-An endpoint protocol only preserves the interactions defined by the global protocol in which the target role is involved, and corresponds to an equivalent \textit{Endpoint Finite State Machine} (EFSM) in the Scribble toolchain.
-We use the EFSMs as a basis for API generation and adopt the formalisms in \cite{Hybrid2016}.
+a Scribble protocol maps to some \textit{global type}, and protocol
+verification implements the algorithmic projection defined in \cite{MPST} to
+derive valid local type \textit{projections} for all participants.
+% Note: Scribble Projection is not MPST projection, because you can have
+% protocols with choices that send to different roles.
+We obtain a set of \textit{endpoint protocols} -- one for each role from a
+well-formed global protocol.
+An endpoint protocol only preserves the interactions defined by the global
+protocol in which the target role is involved, and corresponds to an equivalent
+\textit{Endpoint Finite State Machine} (EFSM) \cite{ICALP13CFSM}.
+We use the EFSMs as a basis for API generation and adopt the formalisms in
+\cite{Hybrid2016}.
 
@@
 
-Developers can implement their application using APIs generated from the EFSM to guarantee correctness by construction.
+Developers can implement their application using APIs generated from the EFSM
+to guarantee correctness by construction.
 Our approach integrates the EFSM into the development workflow by encoding
 session types as TypeScript types.
-There are inherent differences between front- and back-end web development: business logic tends to initiate communication in the back-end, whilst communication in the front-end is mostly driven by user interactions with the browser.
-This motivates our design choice of generating different session types encodings for server (\sectionref{section:server}) and browser targets (\sectionref{section:browser}).
+There are inherent differences between front- and back-end web development:
+business logic tends to initiate communication in the back-end, whilst
+communication in the front-end is mostly driven by user interactions with the
+browser. This motivates our design choice of generating different session types
+encodings for server (\cref{section:server}) and browser targets
+(\cref{section:browser}).
 
@@
 \begin{wrapfigure}{r}{0.5\textwidth}
+  \vspace{-5mm}
   \begin{center}
@@
 
+  \vspace{-5mm}
   \captionof{figure}{EFSM for \texttt{Svr}.}
   \label{fig:efsmsvr}
+\vspace{-1cm}
 \end{wrapfigure}
 
-We refer to the \texttt{Svr} EFSM (fig.~\ref{fig:efsmsvr}) as a
-running example in this section. For server-side targets, we encode
-EFSM states as types, considering
-receive and send states separately. Each TypeScript encoding is
-assigned to its state identifier from the EFSM as its type alias, thus providing syntactic sugar when referring to the successor state in the TypeScript encoding of the current state.
-For any state $S$ in the EFSM, we refer to the TypeScript type alias of its encoding as $\llbracket S \rrbracket$.
+We refer to the \texttt{Svr} EFSM (\cref{fig:efsmsvr}) as a running example in
+this section.
+For server-side targets, we encode EFSM states into TypeScript types and
+considering receive and send states separately.
+Each TypeScript encoding is assigned to its state identifier from the EFSM as
+its type alias, thus providing syntactic sugar when referring to the successor
+state in the TypeScript encoding of the current state.
+For any state $S$ in the EFSM, we refer to the TypeScript type alias of its
+encoding as $\llbracket S \rrbracket$.
 
-We make a key design decision \textit{not} to expose communication channels in the TypeScript session type encoding to provide \textit{static} linearity guarantees (\sectionref{section:serverlinear}).
-Our encoding sufficiently exposes seams for the developer to inject their business logic, whilst the generated session API (\sectionref{section:serversessionapi}) handles the sending and receiving of messages; as a result, the encoding does not concern with the role involved in the send/receive action.
-We outline the encoding below using examples from the \textit{Noughts and Crosses} game (Listing \ref{lst:svr}).
+\paragraph{Branching state}
+We consider a receive state as a unary branching state for conciseness.
+A branching state is encoded as an \textit{object literal}
+\cite{TypeScriptSpec}, with each branch $i \in I$, $I$ denoting set of all
+branches, corresponding to a member field.
+A branch expecting to receive a message labelled $\texttt{label}_i$ carrying
+payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as an
+\textit{member field} named $\texttt{label}_i$ of function type
+$(payload:\texttt{T}_i) \to \llbracket S_i \rrbracket$.
+The developer implements a branching operation by passing callbacks for each
+branch, parameterised by the expected message payload type for that branch.
 
-\paragraph{Branching state} We consider a receive state as a unary branching state for conciseness.
-A branching state is encoded as an \textit{object literal} \cite{TypeScriptSpec}, with each branch $i \in I$, $I$ denoting set of all branches, corresponding to a member field.
-A branch expecting to receive a message labelled $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as an \textit{member field} named $\texttt{label}_i$ of function type \texttt{(payload:$\texttt{T}_i$) => $\llbracket S_i \rrbracket$}.
-The developer implements a branching operation by passing callbacks for each branch, parameterised by the expected message payload type for that branch.
+\paragraph{Selection state}
+We consider a send state as a unary selection state for conciseness.
+A selection state is encoded as a \textit{union type}
+\cite{TypeScriptSpec} of internal choice encodings: each internal choice $i \in
+I$, $I$ denoting set of all choices, sending a message labelled
+$\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state
+$S_i$ is encoded as a \textit{tuple type} of \texttt{[Labels.label$_i$, T$_i$,
+  $\llbracket S_i \rrbracket$]}.
+The developer implements a selection operation by passing the selected label
+and payload to send in the message.
+We generate a \textit{string enum} (named \texttt{Labels}) wrapping the labels
+in the protocol.
 
-\paragraph{Selection state} We consider a send state as a unary selection state for conciseness.
-A selection state is encoded as a \textit{union type} \cite{TypeScriptSpec} of internal choice encodings: each internal choice $i \in I$, $I$ denoting set of all choices, sending a message labelled $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as a \textit{tuple type} of \texttt{[Labels.label$_i$, T$_i$, $\llbracket S_i \rrbracket$]}.
-The developer implements a selection operation by passing the selected label and payload to send in the message.
-We generate a \textit{string enum} (named \texttt{Labels}) wrapping the labels in the protocol.
-
-\begin{figure}[!h]
+\begin{figure}[ht]
 \begin{lstlisting}[language=JavaScript]
 export type S13 = { Pos: (payload: Point) => S15 };
-export type S15 = [Labels.Lose, Point, S16]
-                | [Labels.Draw, Point, S17]
-                | [Labels.Update, Point, S18];
+export type S15 = [ Labels.Lose, Point, S16 ]
+                | [ Labels.Draw, Point, S17 ]
+                | [ Labels.Update, Point, S18 ];
 \end{lstlisting}
@@
 
+We make a key design decision \textit{not} to expose communication channels in
+the TypeScript session type encoding to provide \textit{static} linearity
+guarantees (\cref{section:serverlinear}).
+Our encoding sufficiently exposes seams for the developer to inject their
+business logic, whilst the generated session API
+(\cref{section:serversessionapi}) handles the sending and receiving of
+messages;
+as a result, the encoding does not concern with the role involved in the
+send/receive action. We outline the encoding below using examples from the
+\textit{Noughts and Crosses} game (\cref{lst:svr}).
+
 \subsubsection{Session runtime}
@@
 
-Our session runtime performs communication in a protocol-conformant manner, but does not expose these IO actions to the developer by delegating the aforementioned responsibilities to an inner class.
-The runtime listens to message (receive) events on the communication channel, invokes the corresponding callback to obtain the value to send next, and performs the send.
-The developer instantiates the session by constructing an instance of the session runtime class, providing the WebSocket endpoint URL (to open the connection) and initial state (to execute the EFSM).
+Our session runtime performs communication in a protocol-conformant manner, but
+does not expose these IO actions to the developer by delegating the
+aforementioned responsibilities to an inner class.
+The runtime listens to message (receive) events on the communication channel,
+invokes the corresponding callback to obtain the value to send next, and
+performs the send.
+The developer instantiates the session by constructing an instance of the
+session runtime class, providing the WebSocket endpoint URL (to open the
+connection) and initial state (to execute the EFSM).
 
@@
 \label{section:serverlinear}
-Channel linearity is guaranteed based on the correctness of our library design and session runtime implementation, which means a faulty implementation could violate this, but is up to the library author to verify once rather than the end user.
-Our library design prevent the two properties below:
+Channel linearity is guaranteed based on the correctness of our library design
+and session runtime implementation, which means a faulty implementation could
+violate this, but is up to the library author to verify once rather than the
+end user. Our library design prevent the two properties below:
 
-\paragraph{Repeat use} We do not expose channels to the programmer, which makes \textit{reuse} impossible.
-For example, to send a message, the generated API only requires the payload that needs to be sent, and the session runtime performs the send internally, guaranteeing this action is done \textit{exactly once} by construction.
+\paragraph{Repeated use}
+We do not expose channels to the developer, which makes \textit{reuse}
+impossible.
+For example, to send a message, the generated API only requires the payload
+that needs to be sent, and the session runtime performs the send internally,
+guaranteeing this action is done \textit{exactly once} by construction.
 
-\paragraph{Unused} The initial state must be supplied to the session runtime
-constructor in order to instantiate the session; this initial state is defined
+\paragraph{Unused}
+The initial state must be supplied to the session runtime
+constructor in order to instantiate the session;
+this initial state is defined
 in terms of the successor states, which in turn has references to its
-successors and so forth. This encoding approach covers the terminal state (if it exists), and the session runtime guarantees this terminal state, if it exists, will be reached by construction.
+successors and so forth.
+This encoding approach covers the terminal state (if it exists), and the
+session runtime guarantees this terminal state, if it exists, will be reached
+by construction.
 
@@
 \label{section:browser}
-
-We refer to the \texttt{P1} EFSM (figure \ref{fig:efsmp1}) as a
-running example in this section. Preserving behavioural typing and
-channel linearity are known to be challenging for browser-side
-applications due to EFSM transitions being triggered by user events:
-in the case of \textit{Noughts and Crosses}, once the user makes a move
-by clicking on a cell on the game board, this click event must be
-deactivated until the user's next turn, otherwise the user can click
-again and violate channel linearity. Our design goal is to enforce
-this statically through the generated APIs.
-
-For browser-side targets, we extend the approach presented in \cite{MVU2019} on \textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU) architecture.
-Each state in the EFSM is a model type and uniquely defines a \textit{view function}, set of \textit{messages} and \textit{update function}: the view function defines what to render to the DOM, the set of messages define the possible (IO) actions available at that state, and the update function defines which successor state to transition to, given some supported IO action at this state.
 
@@
 
+We refer to the \texttt{P1} EFSM (\cref{fig:efsmp1}) as a running example in
+this section.
+Preserving behavioural typing and channel linearity are known to be challenging
+for browser-side applications due to EFSM transitions being triggered by user
+events:
+in the case of \textit{Noughts and Crosses}, once the user makes a move by
+clicking on a cell on the game board, this click event must be deactivated
+until the user's next turn, otherwise the user can click again and violate
+channel linearity.
+Our design goal is to enforce this statically through the generated APIs.
+
+For browser-side targets, we extend the approach presented in \cite{MVU2019} on
+\textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU)
+architecture.
+Each state in the EFSM is a \emph{model type} and uniquely defines a
+\textit{view function}, set of \textit{messages} and \textit{update function}:
+the view function defines what to render to the DOM, the set of messages define
+the possible (IO) actions available at that state, and the update function
+defines which successor state to transition to, given some supported IO action
+at this state.
+
 \subsubsection{The React Framework}
@@
 
-We apply the multiple model types approach in \cite{MVU2019} to the React framework \cite{React} developed by Facebook, albeit originally intended for the \textit{Model-View-Controller} (MVC) architecture.
-React is widely used in industry to create scalable single-page TypeScript applications, and we intend for our proposed workflow to be beneficial in an industrial context.
+We apply the multiple model types approach in \cite{MVU2019} to the React
+framework \cite{React} developed by Facebook, albeit originally intended for
+the \textit{Model-View-Controller} (MVC) architecture.
+React is widely used in industry to create scalable single-page TypeScript
+applications, and we intend for our proposed workflow to be beneficial in an
+industrial context.
 We introduce the key features of the framework.
 
-\paragraph{Components} A component is a reusable UI element which
-contains its own markup and logic. Components implement a
-\texttt{render()} function which returns a React element, the smallest
-building blocks of a React application. Components can keep
-\textit{state} and the \texttt{render()} function is invoked upon a
-change of state. A simple counter can be implemented as a component,
+\paragraph{Components}
+A component is a reusable UI element which
+contains its own markup and logic.
+Components implement a \texttt{render()} function which returns a React
+element, the smallest building blocks of a React application, analogous to the
+view function in the MVU architecture.
+Components can keep \textit{state} and the \texttt{render()} function is
+invoked upon a change of state.
+A simple counter can be implemented as a component,
 with \texttt{count} stored as state, a button which increments \texttt{count}
 when clicked and a \texttt{div} that renders the current
-\texttt{count}. Components can also render other components, which gives rise
-to a parent/child relationship between components. Parents can pass
-data to children as \textit{props}: going back to the aforementioned
-example, the the counter component could render a child component
-\texttt{<StyledDiv count=\{this.state.count\} />} in its
-\texttt{render()} function, propagating the \texttt{count} from its
-state to the child. This enables reusability, and for our use case,
-gives a control to the parent on what data to pass to its children (e.g. pass the payload of a received message to a child to render).
+\texttt{count}.
+Components can also render other components, which gives rise
+to a parent/child relationship between components.
+Parents can pass data to children as \textit{props} (short for properties).
+Going back to the aforementioned example, the the counter component could
+render a child component \texttt{<StyledDiv count=\{this.state.count\} />} in
+its \texttt{render()} function, propagating the \texttt{count} from its state
+to the child.
+This enables reusability, and for our use case, gives a control to the parent
+on what data to pass to its children (e.g. pass the payload of a received
+message to a child to render).
 
 \fz{Is VDOM needed to be covered here?}
-\paragraph{Virtual DOM (VDOM)} React components are rendered on a logical abstraction of the DOM, which in turn performs a \texttt{diff} on the current browser DOM and patches the delta accordingly.
-This allows the programmer to directly specify what should be rendered without having to worry about which elements to append or remove from the browser DOM.
+\paragraph{Virtual DOM (VDOM)}
+React components are rendered on a logical
+abstraction of the DOM, which in turn performs a \texttt{diff} on the current
+browser DOM and patches the delta accordingly.
+This allows the developer to
+directly specify what should be rendered without having to worry about which
+elements to append or remove from the browser DOM.
 
@@
 
-\paragraph{State} An EFSM state is encoded as an \textit{abstract} React component.
-This is an abstract class to require the programmer to provide their own view function, which translates conveniently to the \texttt{render()} function of React components.
-Our session runtime (\sectionref{section:clientruntime}) ``executes'' the EFSM and renders the current state, so upon transitioning to a successor state, the successor's view function will be invoked, as per the semantics expressed in \cite{MVU2019}.
+\paragraph{State}
+An EFSM state is encoded as an \textit{abstract} React
+component.
+This is an abstract class to require the developer to provide their
+own view function, which translates conveniently to the \texttt{render()}
+function of React components.
+Our session runtime (\cref{section:clientruntime}) ``executes'' the EFSM and
+renders the current state.
+Upon transitioning to a successor state, the successor's view function will be
+invoked, as per the semantics expressed in \cite{MVU2019}.
 
-\paragraph{Model transitions} Transitions are encoded as React component props onto the encoded states by the session runtime (\sectionref{section:clientruntime}).
-We motivate the design choice of not exposing channel resources to provide guarantees on channel usage.
-React components in TypeScript are \textit{generic} \cite{TypeScriptSpec} which are parameterised by the permitted types of prop and state, which allows us to leverage the TypeScript compiler to verify that the props for model transitions stay local to the state they are defined for.
+\paragraph{Model transitions}
+Transitions are encoded as React component props onto the encoded states by the
+session runtime (\cref{section:clientruntime}).
+We motivate the design choice of not exposing channel resources to provide
+guarantees on channel usage.
+React components in TypeScript are
+\textit{generic} \cite{TypeScriptSpec}, parameterised by the permitted
+types of prop and state.
+The parameters allow us to leverage the TypeScript compiler to
+verify that the props for model transitions stay local to the state they are
+defined for.
 The model transitions for EFSMs are message send and receive.
 
-\subparagraph{Send} We make the assumption that message sending is triggered by some user-driven UI event (e.g.
-clicking a button, pressing a key on the keyboard) which interacts with some DOM element.
-We could pass a \texttt{send()} function as a prop to the sending state, but the programmer would be free to call the function multiple times which makes channel reuse possible.
-Instead, we pass a \textit{factory function} as a prop, which will, given an HTML event and a event handler function, return a fresh React component that binds the sending action on construction, so once the bound event is triggered, our session runtime executes the event handler function to obtain the payload to send, perform the send \textit{exactly once} and transition to (which, in practice, means render) the successor state.
+\subparagraph{Send}
+We make the assumption that message sending is triggered by
+some user-driven UI event (e.g. clicking a button, pressing a key on the
+keyboard) which interacts with some DOM element.
+We could pass a
+\texttt{send()} function as a prop to the sending state, but the developer
+would be free to call the function multiple times which makes channel reuse
+possible.
+Instead, we pass a \textit{factory function} as a prop, which will,
+given an HTML event and a event handler function, return a fresh React
+component that binds the sending action on construction.
+So once the bound event is triggered, our session runtime executes the event
+handler function to obtain the payload to send, perform the send
+\textit{exactly once} and transition to (which, in practice, means render) the
+successor state.
 
@@
 
-We clarify the semantics using the \textit{Noughts and Crosses} example in Listing \ref{lst:clientapp}.
-\texttt{this.props.Pos} is the factory function prop passed from the session runtime.
-For each x-y coordinate on the game board, we create a \texttt{SelectPoint} React component from the factory function (which reads ``build a React component that sends the \texttt{Pos} message with x-y coordinates as payload when the user clicks on it'') and we wrap a table cell (the game board is rendered as a HTML table) inside the \texttt{SelectPoint} component to bind the click event on the table cell.
+We demonstrate the semantics using the \textit{Noughts and Crosses} example in
+\cref{lst:clientapp}.
+\texttt{this.props.Pos} is the factory function prop
+passed from the session runtime.
+For each x-y coordinate on the game board, we
+create a \texttt{SelectPoint} React component from the factory function (which
+reads ``build a React component that sends the \texttt{Pos} message with x-y
+coordinates as payload when the user clicks on it'') and we wrap a table cell
+(the game board is rendered as a HTML table) inside the \texttt{SelectPoint}
+component to bind the click event on the table cell.
 
-\subparagraph{Receive} The payload of a message receive transition is passed as a prop to the successor state and can be handled in the successor's view function.
-Upon a message receive event, the session runtime renders the successor of the receive state and propagates the payload to the successor accordingly.
+\subparagraph{Receive}
+The payload of a message receive transition is passed as
+a prop to the successor state and can be handled in the successor's view
+function.
+Upon a message receive event, the session runtime renders the
+successor of the receive state and propagates the payload to the successor
+accordingly.
 
@@
 
-Our session runtime can be interpreted as an abstraction on top of the React VDOM that implements the EFSM by construction.
-The session runtime itself is a React component too, named after the endpoint role identifier: it opens the WebSocket connection to the server, keeps track of the current EFSM state as part of its React component state, and most importantly, renders the React component encoding of the active EFSM state.
-Channel communications are managed by the runtime, which allows it to render the successor of a receive state upon receiving a message from the channel.
-Similarly, the session runtime is responsible for passing the required props for model transitions to EFSM state React components.
-The session runtime component is rendered by the programmer and requires, as props, the \textit{endpoint URL} (so it can open the connection) and a list of \textit{concrete state components}.
+Our session runtime can be interpreted as an abstraction on top of the React
+VDOM that implements the EFSM by construction.
+The session runtime itself is a React component too, named after the endpoint
+role identifier:
+it opens the WebSocket connection to the server, keeps track of the current
+EFSM state as part of its React component state, and most importantly, renders
+the React component encoding of the active EFSM state.
+Channel communications are managed by the runtime, which allows it to render
+the successor of a receive state upon receiving a message from the channel.
+Similarly, the session runtime is responsible for passing the required props
+for model transitions to EFSM state React components.
+The session runtime component is rendered by the developer and requires, as
+props, the \textit{endpoint URL} (so it can open the connection) and a list of
+\textit{concrete state components}.
 
-The programmer writes their own implementation of each state (mainly to customise how the state is rendered and inject business logic into state transitions) by extending the abstract React class components.
-The session runtime requires references to these concrete components order to render the user implementation accordingly.
+The developer writes their own implementation of each state (mainly to
+customise how the state is rendered and inject business logic into state
+transitions) by extending the abstract React class components.
+The session runtime requires references to these concrete components order to
+render the user implementation accordingly.
 
 \subsubsection{Affine channel usage}
-A limitation of our browser-side session type encoding is only being able to guarantee that channel resources are used \textit{at most once} as supposed to \textit{exactly once}.
+A limitation of our browser-side session type encoding is only being able to
+guarantee that channel resources are used \textit{at most once} as supposed to
+\textit{exactly once}.
 
-Communication channels are not exposed to the programmer so multiple sends are impossible.
-This does not restrict the programmer from binding the send action to exactly one UI event: for \textit{Noughts and Crosses}, we bind the \texttt{Pos(Point)} send action to each unoccupied cell on the game board, but the generated runtime ensures that, once the cell is clicked, the send is only performed once and the successor state is rendered on the DOM, so the channel resource used to send becomes unavailable.
+Communication channels are not exposed to the developer so multiple sends are
+impossible.
+This does not restrict the developer from binding the send action to exactly
+one UI event: for \textit{Noughts and Crosses}, we bind the \texttt{Pos(Point)}
+send action to each unoccupied cell on the game board, but the generated
+runtime ensures that, once the cell is clicked, the send is only performed once
+and the successor state is rendered on the DOM, so the channel resource used to
+send becomes unavailable.
 
-However, our approach \textit{does not} statically detect whether all transitions in a certain state are bound to some UI event.
-This means that it is possible for an implementation to \textit{not} handle transitions to a terminal state but still type-check, so we cannot prevent unused states.
+However, our approach \textit{does not} statically detect whether all
+transitions in a certain state are bound to some UI event.
+This means that it is possible for an implementation to \textit{not} handle
+transitions to a terminal state but still type-check, so we cannot prevent
+unused states.
 \fz{A client may also close the browser, hence dropping the connection?}
@@
 
-We apply our framework to implement a web-based implementation of the \textit{Noughts and Crosses} running example in TypeScript; the interested reader can find the full implementation in \cite{NoughtsAndCrosses}.
-In addition to MPST-safety, we show that our library design welcomes idiomatic JavaScript practices in the user implementation and is interoperable with common front- and back-end frameworks.
+We apply our framework to implement a web-based implementation of the
+\textit{Noughts and Crosses} running example in TypeScript;
+the interested reader can find the full implementation in
+\cite{NoughtsAndCrosses}.
+In addition to MPST-safety, we show that our library design welcomes idiomatic
+JavaScript practices in the user implementation and is interoperable with
+common front- and back-end frameworks.
 
-\paragraph{Game server} We set up the WebSocket server inside an Express.js \cite{ExpressJS} application on top of a Node.js \cite{NodeJS} runtime.
-We define our own game logic in a \texttt{Board} class to keep track of the game state and expose methods to query the result.
-This custom logic is integrated into our \texttt{Receive\_P1} and \texttt{Receive\_P2} handlers (Listing \ref{lst:svrprotocol}), so the session runtime can handle \texttt{Pos(Point)} messages from players and transition according to the injected game logic.
+\paragraph{Game server}
+We set up the WebSocket server inside an Express.js \cite{ExpressJS}
+application on top of a Node.js \cite{NodeJS} runtime.
+We define our own game logic in a \texttt{Board} class to keep track of the
+game state and expose methods to query the result.
+This custom logic is integrated into our \texttt{Receive\_P1} and
+\texttt{Receive\_P2} handlers (\cref{lst:svrprotocol}), so the session runtime
+can handle \texttt{Pos(Point)} messages from players and transition according
+to the injected game logic.
 
-Note that, by the structural typing nature of TypeScript, replacing \texttt{handleP2Move} on line 6 with \texttt{handleP1Move} would be type-correct - this allows for better code reuse as supposed to defining additionl abstractions to work around the limitations of nominal typing in \cite{Hybrid2016}.
-There is also full type erasure when transpiling to JavaScript to run the server code, so state space explosions is never a runtime consideration.
+\fz{Where is Line 6?}
+Note that, by the structural typing nature of TypeScript, replacing
+\texttt{handleP2Move} on line 6 with \texttt{handleP1Move} would be
+type-correct -- this allows for better code reuse as supposed to defining
+additional abstractions to work around the limitations of nominal typing in
+\cite{Hybrid2016}.
+There is also full type erasure when transpiling to JavaScript to run the
+server code, so state space explosions is never a runtime consideration.
 
-\paragraph{Game clients} We implement the game client for \texttt{P1} and \texttt{P2} by extending from the generated abstract React (EFSM state) components and registering those to the session runtime component.
+\paragraph{Game clients}
+We implement the game client for \texttt{P1} and
+\texttt{P2} by extending from the generated abstract React (EFSM state)
+components and registering those to the session runtime component.
 
-For the sake of code reuse for demonstration purposes, \cite{NoughtsAndCrosses} uses \textit{higher-order components} (HOC) to build the correct state implementations depending on which player the user chooses to be.
-We leverage the \textit{Redux} state management library to keep track of game state, thus showing the flexibility of our library design in being interoperable with other libraries and idiomatic JavaScript practices.
+For the sake of code reuse for demonstration purposes, \cite{NoughtsAndCrosses}
+uses \textit{higher-order components} (HOC) to build the correct state
+implementations depending on which player the user chooses to be.
+We leverage the \textit{Redux} state management library to keep track of game
+state, thus showing the flexibility of our library design in being
+interoperable with other libraries and idiomatic JavaScript practices.
 
@@
 
-Native language support for first-class linear channel resources \cite{ATS} and code generation are two main approaches for incorporating MPST into application development.
-The latter closely relates to our proposal; we highlight two areas of existing works under this approach that motivate our design choice.
+Native language support for first-class linear channel resources \cite{ATS} and
+code generation are two main approaches for incorporating MPST into application
+development.
+The latter closely relates to our proposal;
+we highlight two areas of existing works under this approach that motivate our
+design choice.
 
-\paragraph{Endpoint API generation} \cite{Python2017} targeted Python applications and the generation of runtime monitors to dynamically verify communication patterns.
-Whilst the same approach could be applied to JavaScript, we can provide more static guarantees with TypeScript's gradual typing system and compiler.
-\cite{Hybrid2016} targeted Java applications and proposed to encode the EFSM states and transitions as classes and instance methods respectively, with behavioural typing achieved statically by the type system and channel linearity guarantees achieved dynamically since channel resources are exposed and aliasing is not monitored.
+\paragraph{Endpoint API generation}
+\cite{Python2017} targeted Python applications and the generation of runtime
+monitors to dynamically verify communication patterns.
+Whilst the same approach could be applied to JavaScript, we can provide more
+static guarantees with TypeScript's gradual typing system and compiler.
+\cite{Hybrid2016} targeted Java applications and proposed to encode the EFSM
+states and transitions as classes and instance methods respectively, with
+behavioural typing achieved statically by the type system and channel linearity
+guarantees achieved dynamically since channel resources are exposed and
+aliasing is not monitored.
 
-\paragraph{Session types in web development} \cite{PureScript2019} targeted web development in PureScript using the \textit{Concur UI} framework and proposed a type-level encoding of EFSMs as multi-parameter type classes.
-However, it presents a trade-off between achieving static linearity guarantees from the type-level EFSM encoding under the expressive type system and providing an intuitive development experience to programmers, especially given the prevalence of JavaScript and TypeScript applications in industry.
-\cite{MVU2019} focused on applying binary session types in front-end web development and presented approaches that tackle the challenge of guaranteeing linearity in the event-driven environment.
+\paragraph{Session types in web development}
+\cite{PureScript2019} targeted web development in PureScript using the
+\textit{Concur UI} framework and proposed a type-level encoding of EFSMs as
+multi-parameter type classes.
+However, it presents a trade-off between achieving static linearity guarantees
+from the type-level EFSM encoding under the expressive type system and
+providing an intuitive development experience to developers, especially given
+the prevalence of JavaScript and TypeScript applications in industry.
+\cite{MVU2019} focused on applying binary session types in front-end web
+development and presented approaches that tackle the challenge of guaranteeing
+linearity in the event-driven environment.
 
-Our work applies the aforementioned approaches in a \textit{multiparty} context using industrial tools and practices to ultimately encourage MPST-safe web application development workflows in industry.
+Our work applies the aforementioned approaches in a \textit{multiparty} context
+using industrial tools and practices to ultimately encourage MPST-safe web
+application development workflows in industry.
 
@@
 \section{Conclusion and Future Works}
-We have presented a MPST-based framework for developing full-stack interactive TypeScript applications with WebSocket communications that conform to a protocol, statically providing linear channel usage guarantees and affine channel usage guarantees for back-end and front-end targets respectively.
+We have presented a MPST-based framework for developing full-stack interactive
+TypeScript applications with WebSocket communications that conform to a
+protocol, statically providing linear channel usage guarantees and affine
+channel usage guarantees for back-end and front-end targets respectively.
 
-Future works include incorporating \textit{explicit connection actions} introduced in \cite{ExplicitConnections} in our API generation to better model real-world communication protocols that may feature in interactive web applications.
-Whilst our approach supports multiparty sessions, the nature of WebSockets require some server-based role in the communication protocol and clients to interact via the server.
-Extending support to WebRTC would cater for peer-to-peer communication between browsers, which further opens up possibilities for communication protocols supported by our approach.
+Future works include incorporating \textit{explicit connection actions}
+introduced in \cite{ExplicitConnections} in our API generation to better model
+real-world communication protocols that may feature in interactive web
+applications.
+Whilst our approach supports multiparty sessions, the nature of
+WebSockets require some server-based role in the communication protocol and
+clients to interact via the server.
+Extending support to WebRTC \fz{cite it or we don't know what it is} would
+cater for peer-to-peer communication between browsers, which further opens up
+possibilities for communication protocols supported by our approach.
 

### 2020-01-24 PLACES_2020/abstract.tex
Split tex files

--- 2020-01-24 PLACES_2020/main.tex
@@
-\documentclass[submission,copyright,creativecommons]{eptcs}
-\providecommand{\event}{PLACES 2020} % Name of the event you are submitting to
-%\usepackage{breakurl}             % Not needed if you use pdflatex only.
-
-\usepackage{wrapfig}
-\usepackage{graphicx}
-\usepackage{subcaption}
-\usepackage{stmaryrd}
-\usepackage[strings]{underscore}
-\usepackage{color}
-\usepackage{etoolbox}
-\newtoggle{draft}
-\definecolor{purple}{rgb}{0.65, 0.12, 0.82}
-\definecolor{darkgray}{rgb}{.4,.4,.4}
-
-\usepackage{listings}
-\usepackage[noabbrev, capitalise]{cleveref}
-\lstdefinelanguage{Scribble}{
-  keywords={module, type, from, as, global,protocol, role, to, choice, at, or, do},
-  keywordstyle=\color{blue}\bfseries,
-  identifierstyle=\color{black},
-  sensitive=false,
-  comment=[l]{//},
-  morecomment=[s]{/*}{*/},
-  commentstyle=\color{darkgray}\ttfamily,
-  stringstyle=\color{purple}\ttfamily,
-  morestring=[b]',
-  morestring=[b]",
-}
-\lstdefinelanguage{JavaScript}{
-  keywords={typeof, new, true, false, catch, function, return, null, catch, switch, var, if, in, while, do, else, case, break, class, export, boolean, throw, implements, import, this, const, let, extends},
-  keywordstyle=\color{blue}\bfseries,
-%  ndkeywords={class, export, boolean, throw, implements, import, this},
-%  ndkeywordstyle=\color{darkgray}\bfseries,
-  identifierstyle=\color{black},
-  sensitive=false,
-  comment=[l]{//},
-  morecomment=[s]{/*}{*/},
-  commentstyle=\color{darkgray}\ttfamily,
-  stringstyle=\color{purple}\ttfamily,
-  morestring=[b]',
-  morestring=[b]"
-}
-\lstset{
-	basicstyle=\footnotesize\ttfamily,
-	numbers=left,
-	frame=single,
-	tabsize=4
-}
-\usepackage{caption}
-
-\newcommand{\sectionref}[1]{\textsection~\ref{#1}}
-
-\toggletrue{draft}
-\newcommand{\fz}[1]{\iftoggle{draft}{{\color{red}FZ: #1}}{}}
-\newcommand{\am}[1]{\iftoggle{draft}{{\color{blue}AM: #1}}{}}
-
-\title{Verifying Interactive WebSocket Applications in TypeScript}
-\author{Anson Miu
-\institute{Imperial College London}
-\and
-Francisco Ferreira
-\institute{Imperial College London}
-\and
-Nobuko Yoshida
-\institute{Imperial College London}
-\and
-Fangyi Zhou
-\institute{Imperial College London}
-}
-\def\titlerunning{Verifying Interactive WebSocket Applications in TypeScript}
-\def\authorrunning{A. Miu, F. Ferreira, N. Yoshida \& F. Zhou}
-\begin{document}
-\maketitle
-
 \begin{abstract}
@@
 \end{abstract}
-
-\section{Introduction}
-% Situation: stateful communication over WebSocket and interactive web applications
-
-Modern interactive web applications aim to provide a highly responsive user
-experience by minimising the communication latency between clients and servers.
-Whilst the HTTP request-response model is sufficient for retrieving static
-assets, applying the same stateless communication approach for interactive use
-cases (such as real-time multiplayer games) introduces undesirable performance
-overhead.
-Developers have since adopted other communication transport abstractions over
-HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy
-low-latency full-duplex client-server communication in their applications over
-a single persistent connection.
-Enabling more complex communication patterns caters for more interactive use
-cases, but introduces additional correctness concerns to the developer.
-
-% Example: noughts and crosses
-Consider a classic turn-based board game of \textit{Noughts and Crosses}
-between two players.
-Both players, identified by either noughts or crosses respectively, take
-turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player
-wins (when their markers form one straight line on the board) or a stalemate is
-reached (when all cells are occupied).
-A web-based implementation may involve players connected to a game server via
-WebSocket connections.
-The players interact with the game from their web browser, which shows a
-\textit{single-page application} (SPA) of the game client written in a popular
-framework like \textit{React.js} \cite{React}.
-SPAs feature a single HTML page and dynamically rendered content via JavaScript
-in the browser.
-Players take turns to make a move on the game board, which sends a message to
-the server, and the server implements the game logic to progress the game
-forward until a result (either a win/loss or draw) can be declared, where
-either the move of the other player or the game result is sent to players.
-
-Whilst WebSockets make this web-based implementation possible, it introduces
-the developer to a new family of communication errors, even for this simple
-game.
-In addition to the usual testing for game logic correctness, the
-developer needs to test against \textit{deadlocks} (e.g. both players waiting
-for each other to make a move at the same time) and \textit{communication
- mismatches} (e.g. player 1 sending a boolean to the game server instead of
-the board coordinates).
-The complexity of these errors correlate to the
-complexity of related tests required, scale with the complexity of
-communication patterns involved.
-
-\textit{Multiparty Session Types} (MPST) \cite{MPST} provide a framework for
-formally specifying a structured communication pattern between concurrent
-processes and verifying implementations for correctness with respect to the
-communications aspect.
-By specifying the client-server interactions of our game as an MPST protocol
-and verifying the implementations against the protocol for well-formedness,
-MPST theory guarantees well-formed implementations to be free from
-communication errors.
-
-We see the application of the MPST methodology to verifying interactive web
-applications written in TypeScript to be an interesting design space -- to what
-extent can the MPST methodology be applied such that we leverage TypeScript
-language features to maximise static safety guarantees whilst preserving a
-flexible, natural and idiomatic workflow for TypeScript developers?
-Such a workflow would ultimately decrease the overhead for incorporating MPST
-into mainstream web system, which reduces development time by programmatically
-verifying communication correctness of the implementation.
-
-\paragraph{Contributions}
-This paper presents a workflow for developing type-safe interactive SPAs
-motivated by the MPST framework:
-\textbf{(1)} An endpoint API code generation
-workflow targeting TypeScript-based web applications for multiparty sessions;
-\textbf{(2)} An encoding of session types in server-side TypeScript that
-enforces static linearity;
-and \textbf{(3)} An encoding of session types in
-browser-side TypeScript using the React framework that guarantees affine usage
-of communication channels.
-
-%Figure \ref{fig:workflow} illustrates our proposed development workflow: we extend the Scribble framework (\sectionref{section:scribble}) and generate TypeScript endpoints for server-side (\sectionref{section:server}) and browser-side (\sectionref{section:browser}) targets.
-We will use the \textit{Noughts and Crosses} game as our running example in the rest of the paper and show how our approach is a practical compromise that combines benefits from static session typing and being compatible with common libraries and idiomatic JavaScript practices (\sectionref{section:example}).
-
-\section{The Scribble Framework}
-\label{section:scribble}
-
-Development begins by specifying the permitted communications between
-participants as a \textit{global protocol} in Scribble \cite{Scribble}, a
-MPST-based protocol specification language and code generation toolchain.
-We specify the \textit{Noughts and Crosses} game as a Scribble protocol in
-\cref{lst:game}.
-
-\begin{figure}[!h]
-\begin{lstlisting}[
-	language=Scribble
-]
-module NoughtsAndCrosses;
-type <typescript> "Coordinate" from "./Types" as Point;	// Position on board
-
-global protocol Game(role Svr, role P1, role P2) {
-  Pos(Point) from P1 to Svr;
-  choice at Svr {
-    Lose(Point) from Svr to P2; Win(Point) from Svr to P1;
-  } or {
-    Draw(Point) from Svr to P2; Draw(Point) from Svr to P1;
-  } or {
-    Update(Point) from Svr to P2; Update(Point) from Svr to P1;
-    do Game(Svr, P2, P1); // Continue the game with player roles swapped
-  }
-}
-\end{lstlisting}
-\captionof{lstlisting}{Main body of the \textit{Noughts and Crosses} protocol.}
-\label{lst:game}
-\end{figure}
-
-We leverage the Scribble toolchain to check for protocol
-well-formedness.
-This directly corresponds to multiparty session
-type theory \cite{MPST}:
-\fz{Cite the featherweight scribble paper for the correspondence?}
-a Scribble protocol maps to some \textit{global type}, and protocol
-verification implements the algorithmic projection defined in \cite{MPST} to
-derive valid local type \textit{projections} for all participants.
-% Note: Scribble Projection is not MPST projection, because you can have
-% protocols with choices that send to different roles.
-We obtain a set of \textit{endpoint protocols} -- one for each role from a
-well-formed global protocol.
-An endpoint protocol only preserves the interactions defined by the global
-protocol in which the target role is involved, and corresponds to an equivalent
-\textit{Endpoint Finite State Machine} (EFSM) \cite{ICALP13CFSM}.
-We use the EFSMs as a basis for API generation and adopt the formalisms in
-\cite{Hybrid2016}.
-
-\section{Encoding Session Types in TypeScript}
-% Design choices
-
-Developers can implement their application using APIs generated from the EFSM
-to guarantee correctness by construction.
-Our approach integrates the EFSM into the development workflow by encoding
-session types as TypeScript types.
-There are inherent differences between front- and back-end web development:
-business logic tends to initiate communication in the back-end, whilst
-communication in the front-end is mostly driven by user interactions with the
-browser. This motivates our design choice of generating different session types
-encodings for server (\cref{section:server}) and browser targets
-(\cref{section:browser}).
-
-\fz{I think there is an important thing to note here, ``Server'' side is
-  basically the side which accepts the connection and ``Client'' side is the
-side which makes the connection. But this is orthogonal to whether it runs on
-Node or on browser? (Correct me if I'm wrong)}
-
-\am{Not for the case of WebSockets, because only WebSocket clients on browser can initiate connections and only WebSocket Server on Node can accept; will make this distinction clear.}
-
-\subsection{Server-side API generation}
-\fz{Could you remove the state names in figures?}
-\label{section:server}
-
-\begin{wrapfigure}{r}{0.5\textwidth}
-  \vspace{-5mm}
-  \begin{center}
-    \includegraphics[width=0.48\textwidth]{figures/efsm_svr.png}
-  \end{center}
-
-  \vspace{-5mm}
-  \captionof{figure}{EFSM for \texttt{Svr}.}
-  \label{fig:efsmsvr}
-\vspace{-1cm}
-\end{wrapfigure}
-
-We refer to the \texttt{Svr} EFSM (\cref{fig:efsmsvr}) as a running example in
-this section.
-For server-side targets, we encode EFSM states into TypeScript types and
-considering receive and send states separately.
-Each TypeScript encoding is assigned to its state identifier from the EFSM as
-its type alias, thus providing syntactic sugar when referring to the successor
-state in the TypeScript encoding of the current state.
-For any state $S$ in the EFSM, we refer to the TypeScript type alias of its
-encoding as $\llbracket S \rrbracket$.
-
-\paragraph{Branching state}
-We consider a receive state as a unary branching state for conciseness.
-A branching state is encoded as an \textit{object literal}
-\cite{TypeScriptSpec}, with each branch $i \in I$, $I$ denoting set of all
-branches, corresponding to a member field.
-A branch expecting to receive a message labelled $\texttt{label}_i$ carrying
-payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as an
-\textit{member field} named $\texttt{label}_i$ of function type
-$(payload:\texttt{T}_i) \to \llbracket S_i \rrbracket$.
-The developer implements a branching operation by passing callbacks for each
-branch, parameterised by the expected message payload type for that branch.
-
-\paragraph{Selection state}
-We consider a send state as a unary selection state for conciseness.
-A selection state is encoded as a \textit{union type}
-\cite{TypeScriptSpec} of internal choice encodings: each internal choice $i \in
-I$, $I$ denoting set of all choices, sending a message labelled
-$\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state
-$S_i$ is encoded as a \textit{tuple type} of \texttt{[Labels.label$_i$, T$_i$,
-  $\llbracket S_i \rrbracket$]}.
-The developer implements a selection operation by passing the selected label
-and payload to send in the message.
-We generate a \textit{string enum} (named \texttt{Labels}) wrapping the labels
-in the protocol.
-
-\begin{figure}[ht]
-\begin{lstlisting}[language=JavaScript]
-export type S13 = { Pos: (payload: Point) => S15 };
-export type S15 = [ Labels.Lose, Point, S16 ]
-                | [ Labels.Draw, Point, S17 ]
-                | [ Labels.Update, Point, S18 ];
-\end{lstlisting}
-\captionof{lstlisting}{Example encodings from \textit{Noughts and Crosses} \texttt{Svr} EFSM.}
-\label{lst:svr}
-\end{figure}
-
-We make a key design decision \textit{not} to expose communication channels in
-the TypeScript session type encoding to provide \textit{static} linearity
-guarantees (\cref{section:serverlinear}).
-Our encoding sufficiently exposes seams for the developer to inject their
-business logic, whilst the generated session API
-(\cref{section:serversessionapi}) handles the sending and receiving of
-messages;
-as a result, the encoding does not concern with the role involved in the
-send/receive action. We outline the encoding below using examples from the
-\textit{Noughts and Crosses} game (\cref{lst:svr}).
-
-\subsubsection{Session runtime}
-\label{section:serversessionapi}
-
-Our session runtime performs communication in a protocol-conformant manner, but
-does not expose these IO actions to the developer by delegating the
-aforementioned responsibilities to an inner class.
-The runtime listens to message (receive) events on the communication channel,
-invokes the corresponding callback to obtain the value to send next, and
-performs the send.
-The developer instantiates the session by constructing an instance of the
-session runtime class, providing the WebSocket endpoint URL (to open the
-connection) and initial state (to execute the EFSM).
-
-\subsubsection{Linear channel usage}
-\label{section:serverlinear}
-Channel linearity is guaranteed based on the correctness of our library design
-and session runtime implementation, which means a faulty implementation could
-violate this, but is up to the library author to verify once rather than the
-end user. Our library design prevent the two properties below:
-
-\paragraph{Repeated use}
-We do not expose channels to the developer, which makes \textit{reuse}
-impossible.
-For example, to send a message, the generated API only requires the payload
-that needs to be sent, and the session runtime performs the send internally,
-guaranteeing this action is done \textit{exactly once} by construction.
-
-\paragraph{Unused}
-The initial state must be supplied to the session runtime
-constructor in order to instantiate the session;
-this initial state is defined
-in terms of the successor states, which in turn has references to its
-successors and so forth.
-This encoding approach covers the terminal state (if it exists), and the
-session runtime guarantees this terminal state, if it exists, will be reached
-by construction.
-
-\subsection{Browser-side API generation}
-\label{section:browser}
-
-\begin{wrapfigure}{r}{0.4\textwidth}
-  \begin{center}
-    \includegraphics[width=0.4\textwidth]{figures/efsm_p1.png}
-  \end{center}
-
-  \captionof{figure}{EFSM for \texttt{P1}.}
-  \label{fig:efsmp1}
-\end{wrapfigure}
-
-We refer to the \texttt{P1} EFSM (\cref{fig:efsmp1}) as a running example in
-this section.
-Preserving behavioural typing and channel linearity are known to be challenging
-for browser-side applications due to EFSM transitions being triggered by user
-events:
-in the case of \textit{Noughts and Crosses}, once the user makes a move by
-clicking on a cell on the game board, this click event must be deactivated
-until the user's next turn, otherwise the user can click again and violate
-channel linearity.
-Our design goal is to enforce this statically through the generated APIs.
-
-For browser-side targets, we extend the approach presented in \cite{MVU2019} on
-\textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU)
-architecture.
-Each state in the EFSM is a \emph{model type} and uniquely defines a
-\textit{view function}, set of \textit{messages} and \textit{update function}:
-the view function defines what to render to the DOM, the set of messages define
-the possible (IO) actions available at that state, and the update function
-defines which successor state to transition to, given some supported IO action
-at this state.
-
-\subsubsection{The React Framework}
-
-\fz{This subsubsection is a bit misplaced, probably move it up}
-
-We apply the multiple model types approach in \cite{MVU2019} to the React
-framework \cite{React} developed by Facebook, albeit originally intended for
-the \textit{Model-View-Controller} (MVC) architecture.
-React is widely used in industry to create scalable single-page TypeScript
-applications, and we intend for our proposed workflow to be beneficial in an
-industrial context.
-We introduce the key features of the framework.
-
-\paragraph{Components}
-A component is a reusable UI element which
-contains its own markup and logic.
-Components implement a \texttt{render()} function which returns a React
-element, the smallest building blocks of a React application, analogous to the
-view function in the MVU architecture.
-Components can keep \textit{state} and the \texttt{render()} function is
-invoked upon a change of state.
-A simple counter can be implemented as a component,
-with \texttt{count} stored as state, a button which increments \texttt{count}
-when clicked and a \texttt{div} that renders the current
-\texttt{count}.
-Components can also render other components, which gives rise
-to a parent/child relationship between components.
-Parents can pass data to children as \textit{props} (short for properties).
-Going back to the aforementioned example, the the counter component could
-render a child component \texttt{<StyledDiv count=\{this.state.count\} />} in
-its \texttt{render()} function, propagating the \texttt{count} from its state
-to the child.
-This enables reusability, and for our use case, gives a control to the parent
-on what data to pass to its children (e.g. pass the payload of a received
-message to a child to render).
-
-\fz{Is VDOM needed to be covered here?}
-\paragraph{Virtual DOM (VDOM)}
-React components are rendered on a logical
-abstraction of the DOM, which in turn performs a \texttt{diff} on the current
-browser DOM and patches the delta accordingly.
-This allows the developer to
-directly specify what should be rendered without having to worry about which
-elements to append or remove from the browser DOM.
-
-\subsubsection{Model types in React}
-
-\paragraph{State}
-An EFSM state is encoded as an \textit{abstract} React
-component.
-This is an abstract class to require the developer to provide their
-own view function, which translates conveniently to the \texttt{render()}
-function of React components.
-Our session runtime (\cref{section:clientruntime}) ``executes'' the EFSM and
-renders the current state.
-Upon transitioning to a successor state, the successor's view function will be
-invoked, as per the semantics expressed in \cite{MVU2019}.
-
-\paragraph{Model transitions}
-Transitions are encoded as React component props onto the encoded states by the
-session runtime (\cref{section:clientruntime}).
-We motivate the design choice of not exposing channel resources to provide
-guarantees on channel usage.
-React components in TypeScript are
-\textit{generic} \cite{TypeScriptSpec}, parameterised by the permitted
-types of prop and state.
-The parameters allow us to leverage the TypeScript compiler to
-verify that the props for model transitions stay local to the state they are
-defined for.
-The model transitions for EFSMs are message send and receive.
-
-\subparagraph{Send}
-We make the assumption that message sending is triggered by
-some user-driven UI event (e.g. clicking a button, pressing a key on the
-keyboard) which interacts with some DOM element.
-We could pass a
-\texttt{send()} function as a prop to the sending state, but the developer
-would be free to call the function multiple times which makes channel reuse
-possible.
-Instead, we pass a \textit{factory function} as a prop, which will,
-given an HTML event and a event handler function, return a fresh React
-component that binds the sending action on construction.
-So once the bound event is triggered, our session runtime executes the event
-handler function to obtain the payload to send, perform the send
-\textit{exactly once} and transition to (which, in practice, means render) the
-successor state.
-
-\begin{figure}[!h]
-\begin{lstlisting}[language=JavaScript, tabsize=4]
-// Inside some render() function..
-{board.map((row, x) => (
-	row.map((col, y) => {
-		const SelectPoint = this.props.Pos('click', (event: UIEvent) => {
-			event.preventDefault();
-			return { x: x, y: y };}
-		return <SelectPoint><td>.</td></SelectPoint>;
-});}
-\end{lstlisting}
-\captionof{lstlisting}{Model transition for message sending in
-\textit{Noughts and Crosses} \texttt{P1} implementation.}
-\label{lst:clientapp}
-\end{figure}
-
-We demonstrate the semantics using the \textit{Noughts and Crosses} example in
-\cref{lst:clientapp}.
-\texttt{this.props.Pos} is the factory function prop
-passed from the session runtime.
-For each x-y coordinate on the game board, we
-create a \texttt{SelectPoint} React component from the factory function (which
-reads ``build a React component that sends the \texttt{Pos} message with x-y
-coordinates as payload when the user clicks on it'') and we wrap a table cell
-(the game board is rendered as a HTML table) inside the \texttt{SelectPoint}
-component to bind the click event on the table cell.
-
-\subparagraph{Receive}
-The payload of a message receive transition is passed as
-a prop to the successor state and can be handled in the successor's view
-function.
-Upon a message receive event, the session runtime renders the
-successor of the receive state and propagates the payload to the successor
-accordingly.
-
-\subsubsection{Session runtime}
-\label{section:clientruntime}
-
-Our session runtime can be interpreted as an abstraction on top of the React
-VDOM that implements the EFSM by construction.
-The session runtime itself is a React component too, named after the endpoint
-role identifier:
-it opens the WebSocket connection to the server, keeps track of the current
-EFSM state as part of its React component state, and most importantly, renders
-the React component encoding of the active EFSM state.
-Channel communications are managed by the runtime, which allows it to render
-the successor of a receive state upon receiving a message from the channel.
-Similarly, the session runtime is responsible for passing the required props
-for model transitions to EFSM state React components.
-The session runtime component is rendered by the developer and requires, as
-props, the \textit{endpoint URL} (so it can open the connection) and a list of
-\textit{concrete state components}.
-
-The developer writes their own implementation of each state (mainly to
-customise how the state is rendered and inject business logic into state
-transitions) by extending the abstract React class components.
-The session runtime requires references to these concrete components order to
-render the user implementation accordingly.
-
-\subsubsection{Affine channel usage}
-A limitation of our browser-side session type encoding is only being able to
-guarantee that channel resources are used \textit{at most once} as supposed to
-\textit{exactly once}.
-
-Communication channels are not exposed to the developer so multiple sends are
-impossible.
-This does not restrict the developer from binding the send action to exactly
-one UI event: for \textit{Noughts and Crosses}, we bind the \texttt{Pos(Point)}
-send action to each unoccupied cell on the game board, but the generated
-runtime ensures that, once the cell is clicked, the send is only performed once
-and the successor state is rendered on the DOM, so the channel resource used to
-send becomes unavailable.
-
-However, our approach \textit{does not} statically detect whether all
-transitions in a certain state are bound to some UI event.
-This means that it is possible for an implementation to \textit{not} handle
-transitions to a terminal state but still type-check, so we cannot prevent
-unused states.
-\fz{A client may also close the browser, hence dropping the connection?}
-
-\section{Case Study}
-\label{section:example}
-
-\begin{figure}
-\begin{lstlisting}[language=JavaScript, tabsize=4]
-const handleP1Move: Receive_P1 = (move: Point) => {
-	board.P1(move);			// User logic
-	if (board.won()) { return [Labels.Lose, move, move]; }
-	else if (board.draw()) { return [Labels.Draw, move, move]; }
-	else { return [Labels.Update, move, handleP2Move]; }}
-
-// Instantiate session - `handleP2Move` defined similarly
-new NoughtsAndCrosses.Svr(webSocketServer, handleP1Move, handleP2Move);
-\end{lstlisting}
-\captionof{lstlisting}{Session runtime instantiation for \textit{Noughts and Crosses} \texttt{Svr}.}
-\label{lst:svrprotocol}
-\end{figure}
-
-We apply our framework to implement a web-based implementation of the
-\textit{Noughts and Crosses} running example in TypeScript;
-the interested reader can find the full implementation in
-\cite{NoughtsAndCrosses}.
-In addition to MPST-safety, we show that our library design welcomes idiomatic
-JavaScript practices in the user implementation and is interoperable with
-common front- and back-end frameworks.
-
-\paragraph{Game server}
-We set up the WebSocket server inside an Express.js \cite{ExpressJS}
-application on top of a Node.js \cite{NodeJS} runtime.
-We define our own game logic in a \texttt{Board} class to keep track of the
-game state and expose methods to query the result.
-This custom logic is integrated into our \texttt{Receive\_P1} and
-\texttt{Receive\_P2} handlers (\cref{lst:svrprotocol}), so the session runtime
-can handle \texttt{Pos(Point)} messages from players and transition according
-to the injected game logic.
-
-\fz{Where is Line 6?}
-Note that, by the structural typing nature of TypeScript, replacing
-\texttt{handleP2Move} on line 6 with \texttt{handleP1Move} would be
-type-correct -- this allows for better code reuse as supposed to defining
-additional abstractions to work around the limitations of nominal typing in
-\cite{Hybrid2016}.
-There is also full type erasure when transpiling to JavaScript to run the
-server code, so state space explosions is never a runtime consideration.
-
-\paragraph{Game clients}
-We implement the game client for \texttt{P1} and
-\texttt{P2} by extending from the generated abstract React (EFSM state)
-components and registering those to the session runtime component.
-
-For the sake of code reuse for demonstration purposes, \cite{NoughtsAndCrosses}
-uses \textit{higher-order components} (HOC) to build the correct state
-implementations depending on which player the user chooses to be.
-We leverage the \textit{Redux} state management library to keep track of game
-state, thus showing the flexibility of our library design in being
-interoperable with other libraries and idiomatic JavaScript practices.
-
-\section{Related Works}
-\fz{Try to avoid using {[X]} at the beginning of a sentence.}
-
-Native language support for first-class linear channel resources \cite{ATS} and
-code generation are two main approaches for incorporating MPST into application
-development.
-The latter closely relates to our proposal;
-we highlight two areas of existing works under this approach that motivate our
-design choice.
-
-\paragraph{Endpoint API generation}
-\cite{Python2017} targeted Python applications and the generation of runtime
-monitors to dynamically verify communication patterns.
-Whilst the same approach could be applied to JavaScript, we can provide more
-static guarantees with TypeScript's gradual typing system and compiler.
-\cite{Hybrid2016} targeted Java applications and proposed to encode the EFSM
-states and transitions as classes and instance methods respectively, with
-behavioural typing achieved statically by the type system and channel linearity
-guarantees achieved dynamically since channel resources are exposed and
-aliasing is not monitored.
-
-\paragraph{Session types in web development}
-\cite{PureScript2019} targeted web development in PureScript using the
-\textit{Concur UI} framework and proposed a type-level encoding of EFSMs as
-multi-parameter type classes.
-However, it presents a trade-off between achieving static linearity guarantees
-from the type-level EFSM encoding under the expressive type system and
-providing an intuitive development experience to developers, especially given
-the prevalence of JavaScript and TypeScript applications in industry.
-\cite{MVU2019} focused on applying binary session types in front-end web
-development and presented approaches that tackle the challenge of guaranteeing
-linearity in the event-driven environment.
-
-Our work applies the aforementioned approaches in a \textit{multiparty} context
-using industrial tools and practices to ultimately encourage MPST-safe web
-application development workflows in industry.
-
-%\paragraph{Typestate programming} \dots
-
-\section{Conclusion and Future Works}
-We have presented a MPST-based framework for developing full-stack interactive
-TypeScript applications with WebSocket communications that conform to a
-protocol, statically providing linear channel usage guarantees and affine
-channel usage guarantees for back-end and front-end targets respectively.
-
-Future works include incorporating \textit{explicit connection actions}
-introduced in \cite{ExplicitConnections} in our API generation to better model
-real-world communication protocols that may feature in interactive web
-applications.
-Whilst our approach supports multiparty sessions, the nature of
-WebSockets require some server-based role in the communication protocol and
-clients to interact via the server.
-Extending support to WebRTC \fz{cite it or we don't know what it is} would
-cater for peer-to-peer communication between browsers, which further opens up
-possibilities for communication protocols supported by our approach.
-
-% Future work
-% - generalise to encode protocols as typestates (e.g. Mungo example)
-% - WebRTC for full MPST (since current limitation is w.r.t. MPST protocols with 1-server many-client roles
-
-\nocite{*}
-\bibliographystyle{eptcs}
-\bibliography{generic}
-\end{document}

### 2020-01-24 PLACES_2020/casestudy.tex
Split tex files

--- 2020-01-24 PLACES_2020/main.tex
@@
-\documentclass[submission,copyright,creativecommons]{eptcs}
-\providecommand{\event}{PLACES 2020} % Name of the event you are submitting to
-%\usepackage{breakurl}             % Not needed if you use pdflatex only.
-
-\usepackage{wrapfig}
-\usepackage{graphicx}
-\usepackage{subcaption}
-\usepackage{stmaryrd}
-\usepackage[strings]{underscore}
-\usepackage{color}
-\usepackage{etoolbox}
-\newtoggle{draft}
-\definecolor{purple}{rgb}{0.65, 0.12, 0.82}
-\definecolor{darkgray}{rgb}{.4,.4,.4}
-
-\usepackage{listings}
-\usepackage[noabbrev, capitalise]{cleveref}
-\lstdefinelanguage{Scribble}{
-  keywords={module, type, from, as, global,protocol, role, to, choice, at, or, do},
-  keywordstyle=\color{blue}\bfseries,
-  identifierstyle=\color{black},
-  sensitive=false,
-  comment=[l]{//},
-  morecomment=[s]{/*}{*/},
-  commentstyle=\color{darkgray}\ttfamily,
-  stringstyle=\color{purple}\ttfamily,
-  morestring=[b]',
-  morestring=[b]",
-}
-\lstdefinelanguage{JavaScript}{
-  keywords={typeof, new, true, false, catch, function, return, null, catch, switch, var, if, in, while, do, else, case, break, class, export, boolean, throw, implements, import, this, const, let, extends},
-  keywordstyle=\color{blue}\bfseries,
-%  ndkeywords={class, export, boolean, throw, implements, import, this},
-%  ndkeywordstyle=\color{darkgray}\bfseries,
-  identifierstyle=\color{black},
-  sensitive=false,
-  comment=[l]{//},
-  morecomment=[s]{/*}{*/},
-  commentstyle=\color{darkgray}\ttfamily,
-  stringstyle=\color{purple}\ttfamily,
-  morestring=[b]',
-  morestring=[b]"
-}
-\lstset{
-	basicstyle=\footnotesize\ttfamily,
-	numbers=left,
-	frame=single,
-	tabsize=4
-}
-\usepackage{caption}
-
-\newcommand{\sectionref}[1]{\textsection~\ref{#1}}
-
-\toggletrue{draft}
-\newcommand{\fz}[1]{\iftoggle{draft}{{\color{red}FZ: #1}}{}}
-\newcommand{\am}[1]{\iftoggle{draft}{{\color{blue}AM: #1}}{}}
-
-\title{Verifying Interactive WebSocket Applications in TypeScript}
-\author{Anson Miu
-\institute{Imperial College London}
-\and
-Francisco Ferreira
-\institute{Imperial College London}
-\and
-Nobuko Yoshida
-\institute{Imperial College London}
-\and
-Fangyi Zhou
-\institute{Imperial College London}
-}
-\def\titlerunning{Verifying Interactive WebSocket Applications in TypeScript}
-\def\authorrunning{A. Miu, F. Ferreira, N. Yoshida \& F. Zhou}
-\begin{document}
-\maketitle
-
-\begin{abstract}
-Advancements in mobile device compute power have made interactive web
-applications possible, allowing the web browser to render contents dynamically
-and support low-latency communication with the server.
-This comes at a cost to the developer, who now needs to reason more about the
-communication correctness of their application as they support more complex
-communication patterns.
-
-Multiparty session types (MPST) provide a framework for verifying conformance
-of implementations to their prescribed communication protocol.
-Existing proposals for applying the MPST framework in application development
-either neglect the event-driven nature of web applications, or lack
-compatibility with industry tools and practices, which discourages mainstream
-adoption by web developers.
-
-In this paper, we present an implementation of the MPST framework for
-developing interactive web applications using familiar industry tools in
-TypeScript and the \textit{React.js} framework.
-The developer can use the Scribble protocol
-language to specify the protocol and use the Scribble toolchain to validate
-and obtain the local protocol for each role.
-The local protocol describes the interactions of the global communication
-protocols observed by the role.
-We encode the local protocol into TypeScript types, catering for server-side
-and client-side targets separately, and show that our encoding guarantees that
-only implementations which conform to the protocol can type-check.
-We demonstrate the effectiveness of our approach through a web-based
-implementation of the classic \textit{Noughts and Crosses} game from a MPST
-formalism of the game logic.
-
-\end{abstract}
-
-\section{Introduction}
-% Situation: stateful communication over WebSocket and interactive web applications
-
-Modern interactive web applications aim to provide a highly responsive user
-experience by minimising the communication latency between clients and servers.
-Whilst the HTTP request-response model is sufficient for retrieving static
-assets, applying the same stateless communication approach for interactive use
-cases (such as real-time multiplayer games) introduces undesirable performance
-overhead.
-Developers have since adopted other communication transport abstractions over
-HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy
-low-latency full-duplex client-server communication in their applications over
-a single persistent connection.
-Enabling more complex communication patterns caters for more interactive use
-cases, but introduces additional correctness concerns to the developer.
-
-% Example: noughts and crosses
-Consider a classic turn-based board game of \textit{Noughts and Crosses}
-between two players.
-Both players, identified by either noughts or crosses respectively, take
-turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player
-wins (when their markers form one straight line on the board) or a stalemate is
-reached (when all cells are occupied).
-A web-based implementation may involve players connected to a game server via
-WebSocket connections.
-The players interact with the game from their web browser, which shows a
-\textit{single-page application} (SPA) of the game client written in a popular
-framework like \textit{React.js} \cite{React}.
-SPAs feature a single HTML page and dynamically rendered content via JavaScript
-in the browser.
-Players take turns to make a move on the game board, which sends a message to
-the server, and the server implements the game logic to progress the game
-forward until a result (either a win/loss or draw) can be declared, where
-either the move of the other player or the game result is sent to players.
-
-Whilst WebSockets make this web-based implementation possible, it introduces
-the developer to a new family of communication errors, even for this simple
-game.
-In addition to the usual testing for game logic correctness, the
-developer needs to test against \textit{deadlocks} (e.g. both players waiting
-for each other to make a move at the same time) and \textit{communication
- mismatches} (e.g. player 1 sending a boolean to the game server instead of
-the board coordinates).
-The complexity of these errors correlate to the
-complexity of related tests required, scale with the complexity of
-communication patterns involved.
-
-\textit{Multiparty Session Types} (MPST) \cite{MPST} provide a framework for
-formally specifying a structured communication pattern between concurrent
-processes and verifying implementations for correctness with respect to the
-communications aspect.
-By specifying the client-server interactions of our game as an MPST protocol
-and verifying the implementations against the protocol for well-formedness,
-MPST theory guarantees well-formed implementations to be free from
-communication errors.
-
-We see the application of the MPST methodology to verifying interactive web
-applications written in TypeScript to be an interesting design space -- to what
-extent can the MPST methodology be applied such that we leverage TypeScript
-language features to maximise static safety guarantees whilst preserving a
-flexible, natural and idiomatic workflow for TypeScript developers?
-Such a workflow would ultimately decrease the overhead for incorporating MPST
-into mainstream web system, which reduces development time by programmatically
-verifying communication correctness of the implementation.
-
-\paragraph{Contributions}
-This paper presents a workflow for developing type-safe interactive SPAs
-motivated by the MPST framework:
-\textbf{(1)} An endpoint API code generation
-workflow targeting TypeScript-based web applications for multiparty sessions;
-\textbf{(2)} An encoding of session types in server-side TypeScript that
-enforces static linearity;
-and \textbf{(3)} An encoding of session types in
-browser-side TypeScript using the React framework that guarantees affine usage
-of communication channels.
-
-%Figure \ref{fig:workflow} illustrates our proposed development workflow: we extend the Scribble framework (\sectionref{section:scribble}) and generate TypeScript endpoints for server-side (\sectionref{section:server}) and browser-side (\sectionref{section:browser}) targets.
-We will use the \textit{Noughts and Crosses} game as our running example in the rest of the paper and show how our approach is a practical compromise that combines benefits from static session typing and being compatible with common libraries and idiomatic JavaScript practices (\sectionref{section:example}).
-
-\section{The Scribble Framework}
-\label{section:scribble}
-
-Development begins by specifying the permitted communications between
-participants as a \textit{global protocol} in Scribble \cite{Scribble}, a
-MPST-based protocol specification language and code generation toolchain.
-We specify the \textit{Noughts and Crosses} game as a Scribble protocol in
-\cref{lst:game}.
-
-\begin{figure}[!h]
-\begin{lstlisting}[
-	language=Scribble
-]
-module NoughtsAndCrosses;
-type <typescript> "Coordinate" from "./Types" as Point;	// Position on board
-
-global protocol Game(role Svr, role P1, role P2) {
-  Pos(Point) from P1 to Svr;
-  choice at Svr {
-    Lose(Point) from Svr to P2; Win(Point) from Svr to P1;
-  } or {
-    Draw(Point) from Svr to P2; Draw(Point) from Svr to P1;
-  } or {
-    Update(Point) from Svr to P2; Update(Point) from Svr to P1;
-    do Game(Svr, P2, P1); // Continue the game with player roles swapped
-  }
-}
-\end{lstlisting}
-\captionof{lstlisting}{Main body of the \textit{Noughts and Crosses} protocol.}
-\label{lst:game}
-\end{figure}
-
-We leverage the Scribble toolchain to check for protocol
-well-formedness.
-This directly corresponds to multiparty session
-type theory \cite{MPST}:
-\fz{Cite the featherweight scribble paper for the correspondence?}
-a Scribble protocol maps to some \textit{global type}, and protocol
-verification implements the algorithmic projection defined in \cite{MPST} to
-derive valid local type \textit{projections} for all participants.
-% Note: Scribble Projection is not MPST projection, because you can have
-% protocols with choices that send to different roles.
-We obtain a set of \textit{endpoint protocols} -- one for each role from a
-well-formed global protocol.
-An endpoint protocol only preserves the interactions defined by the global
-protocol in which the target role is involved, and corresponds to an equivalent
-\textit{Endpoint Finite State Machine} (EFSM) \cite{ICALP13CFSM}.
-We use the EFSMs as a basis for API generation and adopt the formalisms in
-\cite{Hybrid2016}.
-
-\section{Encoding Session Types in TypeScript}
-% Design choices
-
-Developers can implement their application using APIs generated from the EFSM
-to guarantee correctness by construction.
-Our approach integrates the EFSM into the development workflow by encoding
-session types as TypeScript types.
-There are inherent differences between front- and back-end web development:
-business logic tends to initiate communication in the back-end, whilst
-communication in the front-end is mostly driven by user interactions with the
-browser. This motivates our design choice of generating different session types
-encodings for server (\cref{section:server}) and browser targets
-(\cref{section:browser}).
-
-\fz{I think there is an important thing to note here, ``Server'' side is
-  basically the side which accepts the connection and ``Client'' side is the
-side which makes the connection. But this is orthogonal to whether it runs on
-Node or on browser? (Correct me if I'm wrong)}
-
-\am{Not for the case of WebSockets, because only WebSocket clients on browser can initiate connections and only WebSocket Server on Node can accept; will make this distinction clear.}
-
-\subsection{Server-side API generation}
-\fz{Could you remove the state names in figures?}
-\label{section:server}
-
-\begin{wrapfigure}{r}{0.5\textwidth}
-  \vspace{-5mm}
-  \begin{center}
-    \includegraphics[width=0.48\textwidth]{figures/efsm_svr.png}
-  \end{center}
-
-  \vspace{-5mm}
-  \captionof{figure}{EFSM for \texttt{Svr}.}
-  \label{fig:efsmsvr}
-\vspace{-1cm}
-\end{wrapfigure}
-
-We refer to the \texttt{Svr} EFSM (\cref{fig:efsmsvr}) as a running example in
-this section.
-For server-side targets, we encode EFSM states into TypeScript types and
-considering receive and send states separately.
-Each TypeScript encoding is assigned to its state identifier from the EFSM as
-its type alias, thus providing syntactic sugar when referring to the successor
-state in the TypeScript encoding of the current state.
-For any state $S$ in the EFSM, we refer to the TypeScript type alias of its
-encoding as $\llbracket S \rrbracket$.
-
-\paragraph{Branching state}
-We consider a receive state as a unary branching state for conciseness.
-A branching state is encoded as an \textit{object literal}
-\cite{TypeScriptSpec}, with each branch $i \in I$, $I$ denoting set of all
-branches, corresponding to a member field.
-A branch expecting to receive a message labelled $\texttt{label}_i$ carrying
-payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as an
-\textit{member field} named $\texttt{label}_i$ of function type
-$(payload:\texttt{T}_i) \to \llbracket S_i \rrbracket$.
-The developer implements a branching operation by passing callbacks for each
-branch, parameterised by the expected message payload type for that branch.
-
-\paragraph{Selection state}
-We consider a send state as a unary selection state for conciseness.
-A selection state is encoded as a \textit{union type}
-\cite{TypeScriptSpec} of internal choice encodings: each internal choice $i \in
-I$, $I$ denoting set of all choices, sending a message labelled
-$\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state
-$S_i$ is encoded as a \textit{tuple type} of \texttt{[Labels.label$_i$, T$_i$,
-  $\llbracket S_i \rrbracket$]}.
-The developer implements a selection operation by passing the selected label
-and payload to send in the message.
-We generate a \textit{string enum} (named \texttt{Labels}) wrapping the labels
-in the protocol.
-
-\begin{figure}[ht]
-\begin{lstlisting}[language=JavaScript]
-export type S13 = { Pos: (payload: Point) => S15 };
-export type S15 = [ Labels.Lose, Point, S16 ]
-                | [ Labels.Draw, Point, S17 ]
-                | [ Labels.Update, Point, S18 ];
-\end{lstlisting}
-\captionof{lstlisting}{Example encodings from \textit{Noughts and Crosses} \texttt{Svr} EFSM.}
-\label{lst:svr}
-\end{figure}
-
-We make a key design decision \textit{not} to expose communication channels in
-the TypeScript session type encoding to provide \textit{static} linearity
-guarantees (\cref{section:serverlinear}).
-Our encoding sufficiently exposes seams for the developer to inject their
-business logic, whilst the generated session API
-(\cref{section:serversessionapi}) handles the sending and receiving of
-messages;
-as a result, the encoding does not concern with the role involved in the
-send/receive action. We outline the encoding below using examples from the
-\textit{Noughts and Crosses} game (\cref{lst:svr}).
-
-\subsubsection{Session runtime}
-\label{section:serversessionapi}
-
-Our session runtime performs communication in a protocol-conformant manner, but
-does not expose these IO actions to the developer by delegating the
-aforementioned responsibilities to an inner class.
-The runtime listens to message (receive) events on the communication channel,
-invokes the corresponding callback to obtain the value to send next, and
-performs the send.
-The developer instantiates the session by constructing an instance of the
-session runtime class, providing the WebSocket endpoint URL (to open the
-connection) and initial state (to execute the EFSM).
-
-\subsubsection{Linear channel usage}
-\label{section:serverlinear}
-Channel linearity is guaranteed based on the correctness of our library design
-and session runtime implementation, which means a faulty implementation could
-violate this, but is up to the library author to verify once rather than the
-end user. Our library design prevent the two properties below:
-
-\paragraph{Repeated use}
-We do not expose channels to the developer, which makes \textit{reuse}
-impossible.
-For example, to send a message, the generated API only requires the payload
-that needs to be sent, and the session runtime performs the send internally,
-guaranteeing this action is done \textit{exactly once} by construction.
-
-\paragraph{Unused}
-The initial state must be supplied to the session runtime
-constructor in order to instantiate the session;
-this initial state is defined
-in terms of the successor states, which in turn has references to its
-successors and so forth.
-This encoding approach covers the terminal state (if it exists), and the
-session runtime guarantees this terminal state, if it exists, will be reached
-by construction.
-
-\subsection{Browser-side API generation}
-\label{section:browser}
-
-\begin{wrapfigure}{r}{0.4\textwidth}
-  \begin{center}
-    \includegraphics[width=0.4\textwidth]{figures/efsm_p1.png}
-  \end{center}
-
-  \captionof{figure}{EFSM for \texttt{P1}.}
-  \label{fig:efsmp1}
-\end{wrapfigure}
-
-We refer to the \texttt{P1} EFSM (\cref{fig:efsmp1}) as a running example in
-this section.
-Preserving behavioural typing and channel linearity are known to be challenging
-for browser-side applications due to EFSM transitions being triggered by user
-events:
-in the case of \textit{Noughts and Crosses}, once the user makes a move by
-clicking on a cell on the game board, this click event must be deactivated
-until the user's next turn, otherwise the user can click again and violate
-channel linearity.
-Our design goal is to enforce this statically through the generated APIs.
-
-For browser-side targets, we extend the approach presented in \cite{MVU2019} on
-\textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU)
-architecture.
-Each state in the EFSM is a \emph{model type} and uniquely defines a
-\textit{view function}, set of \textit{messages} and \textit{update function}:
-the view function defines what to render to the DOM, the set of messages define
-the possible (IO) actions available at that state, and the update function
-defines which successor state to transition to, given some supported IO action
-at this state.
-
-\subsubsection{The React Framework}
-
-\fz{This subsubsection is a bit misplaced, probably move it up}
-
-We apply the multiple model types approach in \cite{MVU2019} to the React
-framework \cite{React} developed by Facebook, albeit originally intended for
-the \textit{Model-View-Controller} (MVC) architecture.
-React is widely used in industry to create scalable single-page TypeScript
-applications, and we intend for our proposed workflow to be beneficial in an
-industrial context.
-We introduce the key features of the framework.
-
-\paragraph{Components}
-A component is a reusable UI element which
-contains its own markup and logic.
-Components implement a \texttt{render()} function which returns a React
-element, the smallest building blocks of a React application, analogous to the
-view function in the MVU architecture.
-Components can keep \textit{state} and the \texttt{render()} function is
-invoked upon a change of state.
-A simple counter can be implemented as a component,
-with \texttt{count} stored as state, a button which increments \texttt{count}
-when clicked and a \texttt{div} that renders the current
-\texttt{count}.
-Components can also render other components, which gives rise
-to a parent/child relationship between components.
-Parents can pass data to children as \textit{props} (short for properties).
-Going back to the aforementioned example, the the counter component could
-render a child component \texttt{<StyledDiv count=\{this.state.count\} />} in
-its \texttt{render()} function, propagating the \texttt{count} from its state
-to the child.
-This enables reusability, and for our use case, gives a control to the parent
-on what data to pass to its children (e.g. pass the payload of a received
-message to a child to render).
-
-\fz{Is VDOM needed to be covered here?}
-\paragraph{Virtual DOM (VDOM)}
-React components are rendered on a logical
-abstraction of the DOM, which in turn performs a \texttt{diff} on the current
-browser DOM and patches the delta accordingly.
-This allows the developer to
-directly specify what should be rendered without having to worry about which
-elements to append or remove from the browser DOM.
-
-\subsubsection{Model types in React}
-
-\paragraph{State}
-An EFSM state is encoded as an \textit{abstract} React
-component.
-This is an abstract class to require the developer to provide their
-own view function, which translates conveniently to the \texttt{render()}
-function of React components.
-Our session runtime (\cref{section:clientruntime}) ``executes'' the EFSM and
-renders the current state.
-Upon transitioning to a successor state, the successor's view function will be
-invoked, as per the semantics expressed in \cite{MVU2019}.
-
-\paragraph{Model transitions}
-Transitions are encoded as React component props onto the encoded states by the
-session runtime (\cref{section:clientruntime}).
-We motivate the design choice of not exposing channel resources to provide
-guarantees on channel usage.
-React components in TypeScript are
-\textit{generic} \cite{TypeScriptSpec}, parameterised by the permitted
-types of prop and state.
-The parameters allow us to leverage the TypeScript compiler to
-verify that the props for model transitions stay local to the state they are
-defined for.
-The model transitions for EFSMs are message send and receive.
-
-\subparagraph{Send}
-We make the assumption that message sending is triggered by
-some user-driven UI event (e.g. clicking a button, pressing a key on the
-keyboard) which interacts with some DOM element.
-We could pass a
-\texttt{send()} function as a prop to the sending state, but the developer
-would be free to call the function multiple times which makes channel reuse
-possible.
-Instead, we pass a \textit{factory function} as a prop, which will,
-given an HTML event and a event handler function, return a fresh React
-component that binds the sending action on construction.
-So once the bound event is triggered, our session runtime executes the event
-handler function to obtain the payload to send, perform the send
-\textit{exactly once} and transition to (which, in practice, means render) the
-successor state.
-
-\begin{figure}[!h]
-\begin{lstlisting}[language=JavaScript, tabsize=4]
-// Inside some render() function..
-{board.map((row, x) => (
-	row.map((col, y) => {
-		const SelectPoint = this.props.Pos('click', (event: UIEvent) => {
-			event.preventDefault();
-			return { x: x, y: y };}
-		return <SelectPoint><td>.</td></SelectPoint>;
-});}
-\end{lstlisting}
-\captionof{lstlisting}{Model transition for message sending in
-\textit{Noughts and Crosses} \texttt{P1} implementation.}
-\label{lst:clientapp}
-\end{figure}
-
-We demonstrate the semantics using the \textit{Noughts and Crosses} example in
-\cref{lst:clientapp}.
-\texttt{this.props.Pos} is the factory function prop
-passed from the session runtime.
-For each x-y coordinate on the game board, we
-create a \texttt{SelectPoint} React component from the factory function (which
-reads ``build a React component that sends the \texttt{Pos} message with x-y
-coordinates as payload when the user clicks on it'') and we wrap a table cell
-(the game board is rendered as a HTML table) inside the \texttt{SelectPoint}
-component to bind the click event on the table cell.
-
-\subparagraph{Receive}
-The payload of a message receive transition is passed as
-a prop to the successor state and can be handled in the successor's view
-function.
-Upon a message receive event, the session runtime renders the
-successor of the receive state and propagates the payload to the successor
-accordingly.
-
-\subsubsection{Session runtime}
-\label{section:clientruntime}
-
-Our session runtime can be interpreted as an abstraction on top of the React
-VDOM that implements the EFSM by construction.
-The session runtime itself is a React component too, named after the endpoint
-role identifier:
-it opens the WebSocket connection to the server, keeps track of the current
-EFSM state as part of its React component state, and most importantly, renders
-the React component encoding of the active EFSM state.
-Channel communications are managed by the runtime, which allows it to render
-the successor of a receive state upon receiving a message from the channel.
-Similarly, the session runtime is responsible for passing the required props
-for model transitions to EFSM state React components.
-The session runtime component is rendered by the developer and requires, as
-props, the \textit{endpoint URL} (so it can open the connection) and a list of
-\textit{concrete state components}.
-
-The developer writes their own implementation of each state (mainly to
-customise how the state is rendered and inject business logic into state
-transitions) by extending the abstract React class components.
-The session runtime requires references to these concrete components order to
-render the user implementation accordingly.
-
-\subsubsection{Affine channel usage}
-A limitation of our browser-side session type encoding is only being able to
-guarantee that channel resources are used \textit{at most once} as supposed to
-\textit{exactly once}.
-
-Communication channels are not exposed to the developer so multiple sends are
-impossible.
-This does not restrict the developer from binding the send action to exactly
-one UI event: for \textit{Noughts and Crosses}, we bind the \texttt{Pos(Point)}
-send action to each unoccupied cell on the game board, but the generated
-runtime ensures that, once the cell is clicked, the send is only performed once
-and the successor state is rendered on the DOM, so the channel resource used to
-send becomes unavailable.
-
-However, our approach \textit{does not} statically detect whether all
-transitions in a certain state are bound to some UI event.
-This means that it is possible for an implementation to \textit{not} handle
-transitions to a terminal state but still type-check, so we cannot prevent
-unused states.
-\fz{A client may also close the browser, hence dropping the connection?}
-
 \section{Case Study}
@@
 interoperable with other libraries and idiomatic JavaScript practices.
-
-\section{Related Works}
-\fz{Try to avoid using {[X]} at the beginning of a sentence.}
-
-Native language support for first-class linear channel resources \cite{ATS} and
-code generation are two main approaches for incorporating MPST into application
-development.
-The latter closely relates to our proposal;
-we highlight two areas of existing works under this approach that motivate our
-design choice.
-
-\paragraph{Endpoint API generation}
-\cite{Python2017} targeted Python applications and the generation of runtime
-monitors to dynamically verify communication patterns.
-Whilst the same approach could be applied to JavaScript, we can provide more
-static guarantees with TypeScript's gradual typing system and compiler.
-\cite{Hybrid2016} targeted Java applications and proposed to encode the EFSM
-states and transitions as classes and instance methods respectively, with
-behavioural typing achieved statically by the type system and channel linearity
-guarantees achieved dynamically since channel resources are exposed and
-aliasing is not monitored.
-
-\paragraph{Session types in web development}
-\cite{PureScript2019} targeted web development in PureScript using the
-\textit{Concur UI} framework and proposed a type-level encoding of EFSMs as
-multi-parameter type classes.
-However, it presents a trade-off between achieving static linearity guarantees
-from the type-level EFSM encoding under the expressive type system and
-providing an intuitive development experience to developers, especially given
-the prevalence of JavaScript and TypeScript applications in industry.
-\cite{MVU2019} focused on applying binary session types in front-end web
-development and presented approaches that tackle the challenge of guaranteeing
-linearity in the event-driven environment.
-
-Our work applies the aforementioned approaches in a \textit{multiparty} context
-using industrial tools and practices to ultimately encourage MPST-safe web
-application development workflows in industry.
-
-%\paragraph{Typestate programming} \dots
-
-\section{Conclusion and Future Works}
-We have presented a MPST-based framework for developing full-stack interactive
-TypeScript applications with WebSocket communications that conform to a
-protocol, statically providing linear channel usage guarantees and affine
-channel usage guarantees for back-end and front-end targets respectively.
-
-Future works include incorporating \textit{explicit connection actions}
-introduced in \cite{ExplicitConnections} in our API generation to better model
-real-world communication protocols that may feature in interactive web
-applications.
-Whilst our approach supports multiparty sessions, the nature of
-WebSockets require some server-based role in the communication protocol and
-clients to interact via the server.
-Extending support to WebRTC \fz{cite it or we don't know what it is} would
-cater for peer-to-peer communication between browsers, which further opens up
-possibilities for communication protocols supported by our approach.
-
-% Future work
-% - generalise to encode protocols as typestates (e.g. Mungo example)
-% - WebRTC for full MPST (since current limitation is w.r.t. MPST protocols with 1-server many-client roles
-
-\nocite{*}
-\bibliographystyle{eptcs}
-\bibliography{generic}
-\end{document}

### 2020-01-24 PLACES_2020/conclusion.tex
Split tex files

--- 2020-01-24 PLACES_2020/main.tex
@@
-\documentclass[submission,copyright,creativecommons]{eptcs}
-\providecommand{\event}{PLACES 2020} % Name of the event you are submitting to
-%\usepackage{breakurl}             % Not needed if you use pdflatex only.
-
-\usepackage{wrapfig}
-\usepackage{graphicx}
-\usepackage{subcaption}
-\usepackage{stmaryrd}
-\usepackage[strings]{underscore}
-\usepackage{color}
-\usepackage{etoolbox}
-\newtoggle{draft}
-\definecolor{purple}{rgb}{0.65, 0.12, 0.82}
-\definecolor{darkgray}{rgb}{.4,.4,.4}
-
-\usepackage{listings}
-\usepackage[noabbrev, capitalise]{cleveref}
-\lstdefinelanguage{Scribble}{
-  keywords={module, type, from, as, global,protocol, role, to, choice, at, or, do},
-  keywordstyle=\color{blue}\bfseries,
-  identifierstyle=\color{black},
-  sensitive=false,
-  comment=[l]{//},
-  morecomment=[s]{/*}{*/},
-  commentstyle=\color{darkgray}\ttfamily,
-  stringstyle=\color{purple}\ttfamily,
-  morestring=[b]',
-  morestring=[b]",
-}
-\lstdefinelanguage{JavaScript}{
-  keywords={typeof, new, true, false, catch, function, return, null, catch, switch, var, if, in, while, do, else, case, break, class, export, boolean, throw, implements, import, this, const, let, extends},
-  keywordstyle=\color{blue}\bfseries,
-%  ndkeywords={class, export, boolean, throw, implements, import, this},
-%  ndkeywordstyle=\color{darkgray}\bfseries,
-  identifierstyle=\color{black},
-  sensitive=false,
-  comment=[l]{//},
-  morecomment=[s]{/*}{*/},
-  commentstyle=\color{darkgray}\ttfamily,
-  stringstyle=\color{purple}\ttfamily,
-  morestring=[b]',
-  morestring=[b]"
-}
-\lstset{
-	basicstyle=\footnotesize\ttfamily,
-	numbers=left,
-	frame=single,
-	tabsize=4
-}
-\usepackage{caption}
-
-\newcommand{\sectionref}[1]{\textsection~\ref{#1}}
-
-\toggletrue{draft}
-\newcommand{\fz}[1]{\iftoggle{draft}{{\color{red}FZ: #1}}{}}
-\newcommand{\am}[1]{\iftoggle{draft}{{\color{blue}AM: #1}}{}}
-
-\title{Verifying Interactive WebSocket Applications in TypeScript}
-\author{Anson Miu
-\institute{Imperial College London}
-\and
-Francisco Ferreira
-\institute{Imperial College London}
-\and
-Nobuko Yoshida
-\institute{Imperial College London}
-\and
-Fangyi Zhou
-\institute{Imperial College London}
-}
-\def\titlerunning{Verifying Interactive WebSocket Applications in TypeScript}
-\def\authorrunning{A. Miu, F. Ferreira, N. Yoshida \& F. Zhou}
-\begin{document}
-\maketitle
-
-\begin{abstract}
-Advancements in mobile device compute power have made interactive web
-applications possible, allowing the web browser to render contents dynamically
-and support low-latency communication with the server.
-This comes at a cost to the developer, who now needs to reason more about the
-communication correctness of their application as they support more complex
-communication patterns.
-
-Multiparty session types (MPST) provide a framework for verifying conformance
-of implementations to their prescribed communication protocol.
-Existing proposals for applying the MPST framework in application development
-either neglect the event-driven nature of web applications, or lack
-compatibility with industry tools and practices, which discourages mainstream
-adoption by web developers.
-
-In this paper, we present an implementation of the MPST framework for
-developing interactive web applications using familiar industry tools in
-TypeScript and the \textit{React.js} framework.
-The developer can use the Scribble protocol
-language to specify the protocol and use the Scribble toolchain to validate
-and obtain the local protocol for each role.
-The local protocol describes the interactions of the global communication
-protocols observed by the role.
-We encode the local protocol into TypeScript types, catering for server-side
-and client-side targets separately, and show that our encoding guarantees that
-only implementations which conform to the protocol can type-check.
-We demonstrate the effectiveness of our approach through a web-based
-implementation of the classic \textit{Noughts and Crosses} game from a MPST
-formalism of the game logic.
-
-\end{abstract}
-
-\section{Introduction}
-% Situation: stateful communication over WebSocket and interactive web applications
-
-Modern interactive web applications aim to provide a highly responsive user
-experience by minimising the communication latency between clients and servers.
-Whilst the HTTP request-response model is sufficient for retrieving static
-assets, applying the same stateless communication approach for interactive use
-cases (such as real-time multiplayer games) introduces undesirable performance
-overhead.
-Developers have since adopted other communication transport abstractions over
-HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy
-low-latency full-duplex client-server communication in their applications over
-a single persistent connection.
-Enabling more complex communication patterns caters for more interactive use
-cases, but introduces additional correctness concerns to the developer.
-
-% Example: noughts and crosses
-Consider a classic turn-based board game of \textit{Noughts and Crosses}
-between two players.
-Both players, identified by either noughts or crosses respectively, take
-turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player
-wins (when their markers form one straight line on the board) or a stalemate is
-reached (when all cells are occupied).
-A web-based implementation may involve players connected to a game server via
-WebSocket connections.
-The players interact with the game from their web browser, which shows a
-\textit{single-page application} (SPA) of the game client written in a popular
-framework like \textit{React.js} \cite{React}.
-SPAs feature a single HTML page and dynamically rendered content via JavaScript
-in the browser.
-Players take turns to make a move on the game board, which sends a message to
-the server, and the server implements the game logic to progress the game
-forward until a result (either a win/loss or draw) can be declared, where
-either the move of the other player or the game result is sent to players.
-
-Whilst WebSockets make this web-based implementation possible, it introduces
-the developer to a new family of communication errors, even for this simple
-game.
-In addition to the usual testing for game logic correctness, the
-developer needs to test against \textit{deadlocks} (e.g. both players waiting
-for each other to make a move at the same time) and \textit{communication
- mismatches} (e.g. player 1 sending a boolean to the game server instead of
-the board coordinates).
-The complexity of these errors correlate to the
-complexity of related tests required, scale with the complexity of
-communication patterns involved.
-
-\textit{Multiparty Session Types} (MPST) \cite{MPST} provide a framework for
-formally specifying a structured communication pattern between concurrent
-processes and verifying implementations for correctness with respect to the
-communications aspect.
-By specifying the client-server interactions of our game as an MPST protocol
-and verifying the implementations against the protocol for well-formedness,
-MPST theory guarantees well-formed implementations to be free from
-communication errors.
-
-We see the application of the MPST methodology to verifying interactive web
-applications written in TypeScript to be an interesting design space -- to what
-extent can the MPST methodology be applied such that we leverage TypeScript
-language features to maximise static safety guarantees whilst preserving a
-flexible, natural and idiomatic workflow for TypeScript developers?
-Such a workflow would ultimately decrease the overhead for incorporating MPST
-into mainstream web system, which reduces development time by programmatically
-verifying communication correctness of the implementation.
-
-\paragraph{Contributions}
-This paper presents a workflow for developing type-safe interactive SPAs
-motivated by the MPST framework:
-\textbf{(1)} An endpoint API code generation
-workflow targeting TypeScript-based web applications for multiparty sessions;
-\textbf{(2)} An encoding of session types in server-side TypeScript that
-enforces static linearity;
-and \textbf{(3)} An encoding of session types in
-browser-side TypeScript using the React framework that guarantees affine usage
-of communication channels.
-
-%Figure \ref{fig:workflow} illustrates our proposed development workflow: we extend the Scribble framework (\sectionref{section:scribble}) and generate TypeScript endpoints for server-side (\sectionref{section:server}) and browser-side (\sectionref{section:browser}) targets.
-We will use the \textit{Noughts and Crosses} game as our running example in the rest of the paper and show how our approach is a practical compromise that combines benefits from static session typing and being compatible with common libraries and idiomatic JavaScript practices (\sectionref{section:example}).
-
-\section{The Scribble Framework}
-\label{section:scribble}
-
-Development begins by specifying the permitted communications between
-participants as a \textit{global protocol} in Scribble \cite{Scribble}, a
-MPST-based protocol specification language and code generation toolchain.
-We specify the \textit{Noughts and Crosses} game as a Scribble protocol in
-\cref{lst:game}.
-
-\begin{figure}[!h]
-\begin{lstlisting}[
-	language=Scribble
-]
-module NoughtsAndCrosses;
-type <typescript> "Coordinate" from "./Types" as Point;	// Position on board
-
-global protocol Game(role Svr, role P1, role P2) {
-  Pos(Point) from P1 to Svr;
-  choice at Svr {
-    Lose(Point) from Svr to P2; Win(Point) from Svr to P1;
-  } or {
-    Draw(Point) from Svr to P2; Draw(Point) from Svr to P1;
-  } or {
-    Update(Point) from Svr to P2; Update(Point) from Svr to P1;
-    do Game(Svr, P2, P1); // Continue the game with player roles swapped
-  }
-}
-\end{lstlisting}
-\captionof{lstlisting}{Main body of the \textit{Noughts and Crosses} protocol.}
-\label{lst:game}
-\end{figure}
-
-We leverage the Scribble toolchain to check for protocol
-well-formedness.
-This directly corresponds to multiparty session
-type theory \cite{MPST}:
-\fz{Cite the featherweight scribble paper for the correspondence?}
-a Scribble protocol maps to some \textit{global type}, and protocol
-verification implements the algorithmic projection defined in \cite{MPST} to
-derive valid local type \textit{projections} for all participants.
-% Note: Scribble Projection is not MPST projection, because you can have
-% protocols with choices that send to different roles.
-We obtain a set of \textit{endpoint protocols} -- one for each role from a
-well-formed global protocol.
-An endpoint protocol only preserves the interactions defined by the global
-protocol in which the target role is involved, and corresponds to an equivalent
-\textit{Endpoint Finite State Machine} (EFSM) \cite{ICALP13CFSM}.
-We use the EFSMs as a basis for API generation and adopt the formalisms in
-\cite{Hybrid2016}.
-
-\section{Encoding Session Types in TypeScript}
-% Design choices
-
-Developers can implement their application using APIs generated from the EFSM
-to guarantee correctness by construction.
-Our approach integrates the EFSM into the development workflow by encoding
-session types as TypeScript types.
-There are inherent differences between front- and back-end web development:
-business logic tends to initiate communication in the back-end, whilst
-communication in the front-end is mostly driven by user interactions with the
-browser. This motivates our design choice of generating different session types
-encodings for server (\cref{section:server}) and browser targets
-(\cref{section:browser}).
-
-\fz{I think there is an important thing to note here, ``Server'' side is
-  basically the side which accepts the connection and ``Client'' side is the
-side which makes the connection. But this is orthogonal to whether it runs on
-Node or on browser? (Correct me if I'm wrong)}
-
-\am{Not for the case of WebSockets, because only WebSocket clients on browser can initiate connections and only WebSocket Server on Node can accept; will make this distinction clear.}
-
-\subsection{Server-side API generation}
-\fz{Could you remove the state names in figures?}
-\label{section:server}
-
-\begin{wrapfigure}{r}{0.5\textwidth}
-  \vspace{-5mm}
-  \begin{center}
-    \includegraphics[width=0.48\textwidth]{figures/efsm_svr.png}
-  \end{center}
-
-  \vspace{-5mm}
-  \captionof{figure}{EFSM for \texttt{Svr}.}
-  \label{fig:efsmsvr}
-\vspace{-1cm}
-\end{wrapfigure}
-
-We refer to the \texttt{Svr} EFSM (\cref{fig:efsmsvr}) as a running example in
-this section.
-For server-side targets, we encode EFSM states into TypeScript types and
-considering receive and send states separately.
-Each TypeScript encoding is assigned to its state identifier from the EFSM as
-its type alias, thus providing syntactic sugar when referring to the successor
-state in the TypeScript encoding of the current state.
-For any state $S$ in the EFSM, we refer to the TypeScript type alias of its
-encoding as $\llbracket S \rrbracket$.
-
-\paragraph{Branching state}
-We consider a receive state as a unary branching state for conciseness.
-A branching state is encoded as an \textit{object literal}
-\cite{TypeScriptSpec}, with each branch $i \in I$, $I$ denoting set of all
-branches, corresponding to a member field.
-A branch expecting to receive a message labelled $\texttt{label}_i$ carrying
-payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as an
-\textit{member field} named $\texttt{label}_i$ of function type
-$(payload:\texttt{T}_i) \to \llbracket S_i \rrbracket$.
-The developer implements a branching operation by passing callbacks for each
-branch, parameterised by the expected message payload type for that branch.
-
-\paragraph{Selection state}
-We consider a send state as a unary selection state for conciseness.
-A selection state is encoded as a \textit{union type}
-\cite{TypeScriptSpec} of internal choice encodings: each internal choice $i \in
-I$, $I$ denoting set of all choices, sending a message labelled
-$\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state
-$S_i$ is encoded as a \textit{tuple type} of \texttt{[Labels.label$_i$, T$_i$,
-  $\llbracket S_i \rrbracket$]}.
-The developer implements a selection operation by passing the selected label
-and payload to send in the message.
-We generate a \textit{string enum} (named \texttt{Labels}) wrapping the labels
-in the protocol.
-
-\begin{figure}[ht]
-\begin{lstlisting}[language=JavaScript]
-export type S13 = { Pos: (payload: Point) => S15 };
-export type S15 = [ Labels.Lose, Point, S16 ]
-                | [ Labels.Draw, Point, S17 ]
-                | [ Labels.Update, Point, S18 ];
-\end{lstlisting}
-\captionof{lstlisting}{Example encodings from \textit{Noughts and Crosses} \texttt{Svr} EFSM.}
-\label{lst:svr}
-\end{figure}
-
-We make a key design decision \textit{not} to expose communication channels in
-the TypeScript session type encoding to provide \textit{static} linearity
-guarantees (\cref{section:serverlinear}).
-Our encoding sufficiently exposes seams for the developer to inject their
-business logic, whilst the generated session API
-(\cref{section:serversessionapi}) handles the sending and receiving of
-messages;
-as a result, the encoding does not concern with the role involved in the
-send/receive action. We outline the encoding below using examples from the
-\textit{Noughts and Crosses} game (\cref{lst:svr}).
-
-\subsubsection{Session runtime}
-\label{section:serversessionapi}
-
-Our session runtime performs communication in a protocol-conformant manner, but
-does not expose these IO actions to the developer by delegating the
-aforementioned responsibilities to an inner class.
-The runtime listens to message (receive) events on the communication channel,
-invokes the corresponding callback to obtain the value to send next, and
-performs the send.
-The developer instantiates the session by constructing an instance of the
-session runtime class, providing the WebSocket endpoint URL (to open the
-connection) and initial state (to execute the EFSM).
-
-\subsubsection{Linear channel usage}
-\label{section:serverlinear}
-Channel linearity is guaranteed based on the correctness of our library design
-and session runtime implementation, which means a faulty implementation could
-violate this, but is up to the library author to verify once rather than the
-end user. Our library design prevent the two properties below:
-
-\paragraph{Repeated use}
-We do not expose channels to the developer, which makes \textit{reuse}
-impossible.
-For example, to send a message, the generated API only requires the payload
-that needs to be sent, and the session runtime performs the send internally,
-guaranteeing this action is done \textit{exactly once} by construction.
-
-\paragraph{Unused}
-The initial state must be supplied to the session runtime
-constructor in order to instantiate the session;
-this initial state is defined
-in terms of the successor states, which in turn has references to its
-successors and so forth.
-This encoding approach covers the terminal state (if it exists), and the
-session runtime guarantees this terminal state, if it exists, will be reached
-by construction.
-
-\subsection{Browser-side API generation}
-\label{section:browser}
-
-\begin{wrapfigure}{r}{0.4\textwidth}
-  \begin{center}
-    \includegraphics[width=0.4\textwidth]{figures/efsm_p1.png}
-  \end{center}
-
-  \captionof{figure}{EFSM for \texttt{P1}.}
-  \label{fig:efsmp1}
-\end{wrapfigure}
-
-We refer to the \texttt{P1} EFSM (\cref{fig:efsmp1}) as a running example in
-this section.
-Preserving behavioural typing and channel linearity are known to be challenging
-for browser-side applications due to EFSM transitions being triggered by user
-events:
-in the case of \textit{Noughts and Crosses}, once the user makes a move by
-clicking on a cell on the game board, this click event must be deactivated
-until the user's next turn, otherwise the user can click again and violate
-channel linearity.
-Our design goal is to enforce this statically through the generated APIs.
-
-For browser-side targets, we extend the approach presented in \cite{MVU2019} on
-\textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU)
-architecture.
-Each state in the EFSM is a \emph{model type} and uniquely defines a
-\textit{view function}, set of \textit{messages} and \textit{update function}:
-the view function defines what to render to the DOM, the set of messages define
-the possible (IO) actions available at that state, and the update function
-defines which successor state to transition to, given some supported IO action
-at this state.
-
-\subsubsection{The React Framework}
-
-\fz{This subsubsection is a bit misplaced, probably move it up}
-
-We apply the multiple model types approach in \cite{MVU2019} to the React
-framework \cite{React} developed by Facebook, albeit originally intended for
-the \textit{Model-View-Controller} (MVC) architecture.
-React is widely used in industry to create scalable single-page TypeScript
-applications, and we intend for our proposed workflow to be beneficial in an
-industrial context.
-We introduce the key features of the framework.
-
-\paragraph{Components}
-A component is a reusable UI element which
-contains its own markup and logic.
-Components implement a \texttt{render()} function which returns a React
-element, the smallest building blocks of a React application, analogous to the
-view function in the MVU architecture.
-Components can keep \textit{state} and the \texttt{render()} function is
-invoked upon a change of state.
-A simple counter can be implemented as a component,
-with \texttt{count} stored as state, a button which increments \texttt{count}
-when clicked and a \texttt{div} that renders the current
-\texttt{count}.
-Components can also render other components, which gives rise
-to a parent/child relationship between components.
-Parents can pass data to children as \textit{props} (short for properties).
-Going back to the aforementioned example, the the counter component could
-render a child component \texttt{<StyledDiv count=\{this.state.count\} />} in
-its \texttt{render()} function, propagating the \texttt{count} from its state
-to the child.
-This enables reusability, and for our use case, gives a control to the parent
-on what data to pass to its children (e.g. pass the payload of a received
-message to a child to render).
-
-\fz{Is VDOM needed to be covered here?}
-\paragraph{Virtual DOM (VDOM)}
-React components are rendered on a logical
-abstraction of the DOM, which in turn performs a \texttt{diff} on the current
-browser DOM and patches the delta accordingly.
-This allows the developer to
-directly specify what should be rendered without having to worry about which
-elements to append or remove from the browser DOM.
-
-\subsubsection{Model types in React}
-
-\paragraph{State}
-An EFSM state is encoded as an \textit{abstract} React
-component.
-This is an abstract class to require the developer to provide their
-own view function, which translates conveniently to the \texttt{render()}
-function of React components.
-Our session runtime (\cref{section:clientruntime}) ``executes'' the EFSM and
-renders the current state.
-Upon transitioning to a successor state, the successor's view function will be
-invoked, as per the semantics expressed in \cite{MVU2019}.
-
-\paragraph{Model transitions}
-Transitions are encoded as React component props onto the encoded states by the
-session runtime (\cref{section:clientruntime}).
-We motivate the design choice of not exposing channel resources to provide
-guarantees on channel usage.
-React components in TypeScript are
-\textit{generic} \cite{TypeScriptSpec}, parameterised by the permitted
-types of prop and state.
-The parameters allow us to leverage the TypeScript compiler to
-verify that the props for model transitions stay local to the state they are
-defined for.
-The model transitions for EFSMs are message send and receive.
-
-\subparagraph{Send}
-We make the assumption that message sending is triggered by
-some user-driven UI event (e.g. clicking a button, pressing a key on the
-keyboard) which interacts with some DOM element.
-We could pass a
-\texttt{send()} function as a prop to the sending state, but the developer
-would be free to call the function multiple times which makes channel reuse
-possible.
-Instead, we pass a \textit{factory function} as a prop, which will,
-given an HTML event and a event handler function, return a fresh React
-component that binds the sending action on construction.
-So once the bound event is triggered, our session runtime executes the event
-handler function to obtain the payload to send, perform the send
-\textit{exactly once} and transition to (which, in practice, means render) the
-successor state.
-
-\begin{figure}[!h]
-\begin{lstlisting}[language=JavaScript, tabsize=4]
-// Inside some render() function..
-{board.map((row, x) => (
-	row.map((col, y) => {
-		const SelectPoint = this.props.Pos('click', (event: UIEvent) => {
-			event.preventDefault();
-			return { x: x, y: y };}
-		return <SelectPoint><td>.</td></SelectPoint>;
-});}
-\end{lstlisting}
-\captionof{lstlisting}{Model transition for message sending in
-\textit{Noughts and Crosses} \texttt{P1} implementation.}
-\label{lst:clientapp}
-\end{figure}
-
-We demonstrate the semantics using the \textit{Noughts and Crosses} example in
-\cref{lst:clientapp}.
-\texttt{this.props.Pos} is the factory function prop
-passed from the session runtime.
-For each x-y coordinate on the game board, we
-create a \texttt{SelectPoint} React component from the factory function (which
-reads ``build a React component that sends the \texttt{Pos} message with x-y
-coordinates as payload when the user clicks on it'') and we wrap a table cell
-(the game board is rendered as a HTML table) inside the \texttt{SelectPoint}
-component to bind the click event on the table cell.
-
-\subparagraph{Receive}
-The payload of a message receive transition is passed as
-a prop to the successor state and can be handled in the successor's view
-function.
-Upon a message receive event, the session runtime renders the
-successor of the receive state and propagates the payload to the successor
-accordingly.
-
-\subsubsection{Session runtime}
-\label{section:clientruntime}
-
-Our session runtime can be interpreted as an abstraction on top of the React
-VDOM that implements the EFSM by construction.
-The session runtime itself is a React component too, named after the endpoint
-role identifier:
-it opens the WebSocket connection to the server, keeps track of the current
-EFSM state as part of its React component state, and most importantly, renders
-the React component encoding of the active EFSM state.
-Channel communications are managed by the runtime, which allows it to render
-the successor of a receive state upon receiving a message from the channel.
-Similarly, the session runtime is responsible for passing the required props
-for model transitions to EFSM state React components.
-The session runtime component is rendered by the developer and requires, as
-props, the \textit{endpoint URL} (so it can open the connection) and a list of
-\textit{concrete state components}.
-
-The developer writes their own implementation of each state (mainly to
-customise how the state is rendered and inject business logic into state
-transitions) by extending the abstract React class components.
-The session runtime requires references to these concrete components order to
-render the user implementation accordingly.
-
-\subsubsection{Affine channel usage}
-A limitation of our browser-side session type encoding is only being able to
-guarantee that channel resources are used \textit{at most once} as supposed to
-\textit{exactly once}.
-
-Communication channels are not exposed to the developer so multiple sends are
-impossible.
-This does not restrict the developer from binding the send action to exactly
-one UI event: for \textit{Noughts and Crosses}, we bind the \texttt{Pos(Point)}
-send action to each unoccupied cell on the game board, but the generated
-runtime ensures that, once the cell is clicked, the send is only performed once
-and the successor state is rendered on the DOM, so the channel resource used to
-send becomes unavailable.
-
-However, our approach \textit{does not} statically detect whether all
-transitions in a certain state are bound to some UI event.
-This means that it is possible for an implementation to \textit{not} handle
-transitions to a terminal state but still type-check, so we cannot prevent
-unused states.
-\fz{A client may also close the browser, hence dropping the connection?}
-
-\section{Case Study}
-\label{section:example}
-
-\begin{figure}
-\begin{lstlisting}[language=JavaScript, tabsize=4]
-const handleP1Move: Receive_P1 = (move: Point) => {
-	board.P1(move);			// User logic
-	if (board.won()) { return [Labels.Lose, move, move]; }
-	else if (board.draw()) { return [Labels.Draw, move, move]; }
-	else { return [Labels.Update, move, handleP2Move]; }}
-
-// Instantiate session - `handleP2Move` defined similarly
-new NoughtsAndCrosses.Svr(webSocketServer, handleP1Move, handleP2Move);
-\end{lstlisting}
-\captionof{lstlisting}{Session runtime instantiation for \textit{Noughts and Crosses} \texttt{Svr}.}
-\label{lst:svrprotocol}
-\end{figure}
-
-We apply our framework to implement a web-based implementation of the
-\textit{Noughts and Crosses} running example in TypeScript;
-the interested reader can find the full implementation in
-\cite{NoughtsAndCrosses}.
-In addition to MPST-safety, we show that our library design welcomes idiomatic
-JavaScript practices in the user implementation and is interoperable with
-common front- and back-end frameworks.
-
-\paragraph{Game server}
-We set up the WebSocket server inside an Express.js \cite{ExpressJS}
-application on top of a Node.js \cite{NodeJS} runtime.
-We define our own game logic in a \texttt{Board} class to keep track of the
-game state and expose methods to query the result.
-This custom logic is integrated into our \texttt{Receive\_P1} and
-\texttt{Receive\_P2} handlers (\cref{lst:svrprotocol}), so the session runtime
-can handle \texttt{Pos(Point)} messages from players and transition according
-to the injected game logic.
-
-\fz{Where is Line 6?}
-Note that, by the structural typing nature of TypeScript, replacing
-\texttt{handleP2Move} on line 6 with \texttt{handleP1Move} would be
-type-correct -- this allows for better code reuse as supposed to defining
-additional abstractions to work around the limitations of nominal typing in
-\cite{Hybrid2016}.
-There is also full type erasure when transpiling to JavaScript to run the
-server code, so state space explosions is never a runtime consideration.
-
-\paragraph{Game clients}
-We implement the game client for \texttt{P1} and
-\texttt{P2} by extending from the generated abstract React (EFSM state)
-components and registering those to the session runtime component.
-
-For the sake of code reuse for demonstration purposes, \cite{NoughtsAndCrosses}
-uses \textit{higher-order components} (HOC) to build the correct state
-implementations depending on which player the user chooses to be.
-We leverage the \textit{Redux} state management library to keep track of game
-state, thus showing the flexibility of our library design in being
-interoperable with other libraries and idiomatic JavaScript practices.
-
-\section{Related Works}
-\fz{Try to avoid using {[X]} at the beginning of a sentence.}
-
-Native language support for first-class linear channel resources \cite{ATS} and
-code generation are two main approaches for incorporating MPST into application
-development.
-The latter closely relates to our proposal;
-we highlight two areas of existing works under this approach that motivate our
-design choice.
-
-\paragraph{Endpoint API generation}
-\cite{Python2017} targeted Python applications and the generation of runtime
-monitors to dynamically verify communication patterns.
-Whilst the same approach could be applied to JavaScript, we can provide more
-static guarantees with TypeScript's gradual typing system and compiler.
-\cite{Hybrid2016} targeted Java applications and proposed to encode the EFSM
-states and transitions as classes and instance methods respectively, with
-behavioural typing achieved statically by the type system and channel linearity
-guarantees achieved dynamically since channel resources are exposed and
-aliasing is not monitored.
-
-\paragraph{Session types in web development}
-\cite{PureScript2019} targeted web development in PureScript using the
-\textit{Concur UI} framework and proposed a type-level encoding of EFSMs as
-multi-parameter type classes.
-However, it presents a trade-off between achieving static linearity guarantees
-from the type-level EFSM encoding under the expressive type system and
-providing an intuitive development experience to developers, especially given
-the prevalence of JavaScript and TypeScript applications in industry.
-\cite{MVU2019} focused on applying binary session types in front-end web
-development and presented approaches that tackle the challenge of guaranteeing
-linearity in the event-driven environment.
-
-Our work applies the aforementioned approaches in a \textit{multiparty} context
-using industrial tools and practices to ultimately encourage MPST-safe web
-application development workflows in industry.
-
-%\paragraph{Typestate programming} \dots
-
 \section{Conclusion and Future Works}
@@
 % - WebRTC for full MPST (since current limitation is w.r.t. MPST protocols with 1-server many-client roles
-
-\nocite{*}
-\bibliographystyle{eptcs}
-\bibliography{generic}
-\end{document}

### 2020-01-24 PLACES_2020/encoding.tex
Split tex files

--- 2020-01-24 PLACES_2020/main.tex
@@
-\documentclass[submission,copyright,creativecommons]{eptcs}
-\providecommand{\event}{PLACES 2020} % Name of the event you are submitting to
-%\usepackage{breakurl}             % Not needed if you use pdflatex only.
-
-\usepackage{wrapfig}
-\usepackage{graphicx}
-\usepackage{subcaption}
-\usepackage{stmaryrd}
-\usepackage[strings]{underscore}
-\usepackage{color}
-\usepackage{etoolbox}
-\newtoggle{draft}
-\definecolor{purple}{rgb}{0.65, 0.12, 0.82}
-\definecolor{darkgray}{rgb}{.4,.4,.4}
-
-\usepackage{listings}
-\usepackage[noabbrev, capitalise]{cleveref}
-\lstdefinelanguage{Scribble}{
-  keywords={module, type, from, as, global,protocol, role, to, choice, at, or, do},
-  keywordstyle=\color{blue}\bfseries,
-  identifierstyle=\color{black},
-  sensitive=false,
-  comment=[l]{//},
-  morecomment=[s]{/*}{*/},
-  commentstyle=\color{darkgray}\ttfamily,
-  stringstyle=\color{purple}\ttfamily,
-  morestring=[b]',
-  morestring=[b]",
-}
-\lstdefinelanguage{JavaScript}{
-  keywords={typeof, new, true, false, catch, function, return, null, catch, switch, var, if, in, while, do, else, case, break, class, export, boolean, throw, implements, import, this, const, let, extends},
-  keywordstyle=\color{blue}\bfseries,
-%  ndkeywords={class, export, boolean, throw, implements, import, this},
-%  ndkeywordstyle=\color{darkgray}\bfseries,
-  identifierstyle=\color{black},
-  sensitive=false,
-  comment=[l]{//},
-  morecomment=[s]{/*}{*/},
-  commentstyle=\color{darkgray}\ttfamily,
-  stringstyle=\color{purple}\ttfamily,
-  morestring=[b]',
-  morestring=[b]"
-}
-\lstset{
-	basicstyle=\footnotesize\ttfamily,
-	numbers=left,
-	frame=single,
-	tabsize=4
-}
-\usepackage{caption}
-
-\newcommand{\sectionref}[1]{\textsection~\ref{#1}}
-
-\toggletrue{draft}
-\newcommand{\fz}[1]{\iftoggle{draft}{{\color{red}FZ: #1}}{}}
-\newcommand{\am}[1]{\iftoggle{draft}{{\color{blue}AM: #1}}{}}
-
-\title{Verifying Interactive WebSocket Applications in TypeScript}
-\author{Anson Miu
-\institute{Imperial College London}
-\and
-Francisco Ferreira
-\institute{Imperial College London}
-\and
-Nobuko Yoshida
-\institute{Imperial College London}
-\and
-Fangyi Zhou
-\institute{Imperial College London}
-}
-\def\titlerunning{Verifying Interactive WebSocket Applications in TypeScript}
-\def\authorrunning{A. Miu, F. Ferreira, N. Yoshida \& F. Zhou}
-\begin{document}
-\maketitle
-
-\begin{abstract}
-Advancements in mobile device compute power have made interactive web
-applications possible, allowing the web browser to render contents dynamically
-and support low-latency communication with the server.
-This comes at a cost to the developer, who now needs to reason more about the
-communication correctness of their application as they support more complex
-communication patterns.
-
-Multiparty session types (MPST) provide a framework for verifying conformance
-of implementations to their prescribed communication protocol.
-Existing proposals for applying the MPST framework in application development
-either neglect the event-driven nature of web applications, or lack
-compatibility with industry tools and practices, which discourages mainstream
-adoption by web developers.
-
-In this paper, we present an implementation of the MPST framework for
-developing interactive web applications using familiar industry tools in
-TypeScript and the \textit{React.js} framework.
-The developer can use the Scribble protocol
-language to specify the protocol and use the Scribble toolchain to validate
-and obtain the local protocol for each role.
-The local protocol describes the interactions of the global communication
-protocols observed by the role.
-We encode the local protocol into TypeScript types, catering for server-side
-and client-side targets separately, and show that our encoding guarantees that
-only implementations which conform to the protocol can type-check.
-We demonstrate the effectiveness of our approach through a web-based
-implementation of the classic \textit{Noughts and Crosses} game from a MPST
-formalism of the game logic.
-
-\end{abstract}
-
-\section{Introduction}
-% Situation: stateful communication over WebSocket and interactive web applications
-
-Modern interactive web applications aim to provide a highly responsive user
-experience by minimising the communication latency between clients and servers.
-Whilst the HTTP request-response model is sufficient for retrieving static
-assets, applying the same stateless communication approach for interactive use
-cases (such as real-time multiplayer games) introduces undesirable performance
-overhead.
-Developers have since adopted other communication transport abstractions over
-HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy
-low-latency full-duplex client-server communication in their applications over
-a single persistent connection.
-Enabling more complex communication patterns caters for more interactive use
-cases, but introduces additional correctness concerns to the developer.
-
-% Example: noughts and crosses
-Consider a classic turn-based board game of \textit{Noughts and Crosses}
-between two players.
-Both players, identified by either noughts or crosses respectively, take
-turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player
-wins (when their markers form one straight line on the board) or a stalemate is
-reached (when all cells are occupied).
-A web-based implementation may involve players connected to a game server via
-WebSocket connections.
-The players interact with the game from their web browser, which shows a
-\textit{single-page application} (SPA) of the game client written in a popular
-framework like \textit{React.js} \cite{React}.
-SPAs feature a single HTML page and dynamically rendered content via JavaScript
-in the browser.
-Players take turns to make a move on the game board, which sends a message to
-the server, and the server implements the game logic to progress the game
-forward until a result (either a win/loss or draw) can be declared, where
-either the move of the other player or the game result is sent to players.
-
-Whilst WebSockets make this web-based implementation possible, it introduces
-the developer to a new family of communication errors, even for this simple
-game.
-In addition to the usual testing for game logic correctness, the
-developer needs to test against \textit{deadlocks} (e.g. both players waiting
-for each other to make a move at the same time) and \textit{communication
- mismatches} (e.g. player 1 sending a boolean to the game server instead of
-the board coordinates).
-The complexity of these errors correlate to the
-complexity of related tests required, scale with the complexity of
-communication patterns involved.
-
-\textit{Multiparty Session Types} (MPST) \cite{MPST} provide a framework for
-formally specifying a structured communication pattern between concurrent
-processes and verifying implementations for correctness with respect to the
-communications aspect.
-By specifying the client-server interactions of our game as an MPST protocol
-and verifying the implementations against the protocol for well-formedness,
-MPST theory guarantees well-formed implementations to be free from
-communication errors.
-
-We see the application of the MPST methodology to verifying interactive web
-applications written in TypeScript to be an interesting design space -- to what
-extent can the MPST methodology be applied such that we leverage TypeScript
-language features to maximise static safety guarantees whilst preserving a
-flexible, natural and idiomatic workflow for TypeScript developers?
-Such a workflow would ultimately decrease the overhead for incorporating MPST
-into mainstream web system, which reduces development time by programmatically
-verifying communication correctness of the implementation.
-
-\paragraph{Contributions}
-This paper presents a workflow for developing type-safe interactive SPAs
-motivated by the MPST framework:
-\textbf{(1)} An endpoint API code generation
-workflow targeting TypeScript-based web applications for multiparty sessions;
-\textbf{(2)} An encoding of session types in server-side TypeScript that
-enforces static linearity;
-and \textbf{(3)} An encoding of session types in
-browser-side TypeScript using the React framework that guarantees affine usage
-of communication channels.
-
-%Figure \ref{fig:workflow} illustrates our proposed development workflow: we extend the Scribble framework (\sectionref{section:scribble}) and generate TypeScript endpoints for server-side (\sectionref{section:server}) and browser-side (\sectionref{section:browser}) targets.
-We will use the \textit{Noughts and Crosses} game as our running example in the rest of the paper and show how our approach is a practical compromise that combines benefits from static session typing and being compatible with common libraries and idiomatic JavaScript practices (\sectionref{section:example}).
-
-\section{The Scribble Framework}
-\label{section:scribble}
-
-Development begins by specifying the permitted communications between
-participants as a \textit{global protocol} in Scribble \cite{Scribble}, a
-MPST-based protocol specification language and code generation toolchain.
-We specify the \textit{Noughts and Crosses} game as a Scribble protocol in
-\cref{lst:game}.
-
-\begin{figure}[!h]
-\begin{lstlisting}[
-	language=Scribble
-]
-module NoughtsAndCrosses;
-type <typescript> "Coordinate" from "./Types" as Point;	// Position on board
-
-global protocol Game(role Svr, role P1, role P2) {
-  Pos(Point) from P1 to Svr;
-  choice at Svr {
-    Lose(Point) from Svr to P2; Win(Point) from Svr to P1;
-  } or {
-    Draw(Point) from Svr to P2; Draw(Point) from Svr to P1;
-  } or {
-    Update(Point) from Svr to P2; Update(Point) from Svr to P1;
-    do Game(Svr, P2, P1); // Continue the game with player roles swapped
-  }
-}
-\end{lstlisting}
-\captionof{lstlisting}{Main body of the \textit{Noughts and Crosses} protocol.}
-\label{lst:game}
-\end{figure}
-
-We leverage the Scribble toolchain to check for protocol
-well-formedness.
-This directly corresponds to multiparty session
-type theory \cite{MPST}:
-\fz{Cite the featherweight scribble paper for the correspondence?}
-a Scribble protocol maps to some \textit{global type}, and protocol
-verification implements the algorithmic projection defined in \cite{MPST} to
-derive valid local type \textit{projections} for all participants.
-% Note: Scribble Projection is not MPST projection, because you can have
-% protocols with choices that send to different roles.
-We obtain a set of \textit{endpoint protocols} -- one for each role from a
-well-formed global protocol.
-An endpoint protocol only preserves the interactions defined by the global
-protocol in which the target role is involved, and corresponds to an equivalent
-\textit{Endpoint Finite State Machine} (EFSM) \cite{ICALP13CFSM}.
-We use the EFSMs as a basis for API generation and adopt the formalisms in
-\cite{Hybrid2016}.
-
 \section{Encoding Session Types in TypeScript}
@@
 \fz{A client may also close the browser, hence dropping the connection?}
-
-\section{Case Study}
-\label{section:example}
-
-\begin{figure}
-\begin{lstlisting}[language=JavaScript, tabsize=4]
-const handleP1Move: Receive_P1 = (move: Point) => {
-	board.P1(move);			// User logic
-	if (board.won()) { return [Labels.Lose, move, move]; }
-	else if (board.draw()) { return [Labels.Draw, move, move]; }
-	else { return [Labels.Update, move, handleP2Move]; }}
-
-// Instantiate session - `handleP2Move` defined similarly
-new NoughtsAndCrosses.Svr(webSocketServer, handleP1Move, handleP2Move);
-\end{lstlisting}
-\captionof{lstlisting}{Session runtime instantiation for \textit{Noughts and Crosses} \texttt{Svr}.}
-\label{lst:svrprotocol}
-\end{figure}
-
-We apply our framework to implement a web-based implementation of the
-\textit{Noughts and Crosses} running example in TypeScript;
-the interested reader can find the full implementation in
-\cite{NoughtsAndCrosses}.
-In addition to MPST-safety, we show that our library design welcomes idiomatic
-JavaScript practices in the user implementation and is interoperable with
-common front- and back-end frameworks.
-
-\paragraph{Game server}
-We set up the WebSocket server inside an Express.js \cite{ExpressJS}
-application on top of a Node.js \cite{NodeJS} runtime.
-We define our own game logic in a \texttt{Board} class to keep track of the
-game state and expose methods to query the result.
-This custom logic is integrated into our \texttt{Receive\_P1} and
-\texttt{Receive\_P2} handlers (\cref{lst:svrprotocol}), so the session runtime
-can handle \texttt{Pos(Point)} messages from players and transition according
-to the injected game logic.
-
-\fz{Where is Line 6?}
-Note that, by the structural typing nature of TypeScript, replacing
-\texttt{handleP2Move} on line 6 with \texttt{handleP1Move} would be
-type-correct -- this allows for better code reuse as supposed to defining
-additional abstractions to work around the limitations of nominal typing in
-\cite{Hybrid2016}.
-There is also full type erasure when transpiling to JavaScript to run the
-server code, so state space explosions is never a runtime consideration.
-
-\paragraph{Game clients}
-We implement the game client for \texttt{P1} and
-\texttt{P2} by extending from the generated abstract React (EFSM state)
-components and registering those to the session runtime component.
-
-For the sake of code reuse for demonstration purposes, \cite{NoughtsAndCrosses}
-uses \textit{higher-order components} (HOC) to build the correct state
-implementations depending on which player the user chooses to be.
-We leverage the \textit{Redux} state management library to keep track of game
-state, thus showing the flexibility of our library design in being
-interoperable with other libraries and idiomatic JavaScript practices.
-
-\section{Related Works}
-\fz{Try to avoid using {[X]} at the beginning of a sentence.}
-
-Native language support for first-class linear channel resources \cite{ATS} and
-code generation are two main approaches for incorporating MPST into application
-development.
-The latter closely relates to our proposal;
-we highlight two areas of existing works under this approach that motivate our
-design choice.
-
-\paragraph{Endpoint API generation}
-\cite{Python2017} targeted Python applications and the generation of runtime
-monitors to dynamically verify communication patterns.
-Whilst the same approach could be applied to JavaScript, we can provide more
-static guarantees with TypeScript's gradual typing system and compiler.
-\cite{Hybrid2016} targeted Java applications and proposed to encode the EFSM
-states and transitions as classes and instance methods respectively, with
-behavioural typing achieved statically by the type system and channel linearity
-guarantees achieved dynamically since channel resources are exposed and
-aliasing is not monitored.
-
-\paragraph{Session types in web development}
-\cite{PureScript2019} targeted web development in PureScript using the
-\textit{Concur UI} framework and proposed a type-level encoding of EFSMs as
-multi-parameter type classes.
-However, it presents a trade-off between achieving static linearity guarantees
-from the type-level EFSM encoding under the expressive type system and
-providing an intuitive development experience to developers, especially given
-the prevalence of JavaScript and TypeScript applications in industry.
-\cite{MVU2019} focused on applying binary session types in front-end web
-development and presented approaches that tackle the challenge of guaranteeing
-linearity in the event-driven environment.
-
-Our work applies the aforementioned approaches in a \textit{multiparty} context
-using industrial tools and practices to ultimately encourage MPST-safe web
-application development workflows in industry.
-
-%\paragraph{Typestate programming} \dots
-
-\section{Conclusion and Future Works}
-We have presented a MPST-based framework for developing full-stack interactive
-TypeScript applications with WebSocket communications that conform to a
-protocol, statically providing linear channel usage guarantees and affine
-channel usage guarantees for back-end and front-end targets respectively.
-
-Future works include incorporating \textit{explicit connection actions}
-introduced in \cite{ExplicitConnections} in our API generation to better model
-real-world communication protocols that may feature in interactive web
-applications.
-Whilst our approach supports multiparty sessions, the nature of
-WebSockets require some server-based role in the communication protocol and
-clients to interact via the server.
-Extending support to WebRTC \fz{cite it or we don't know what it is} would
-cater for peer-to-peer communication between browsers, which further opens up
-possibilities for communication protocols supported by our approach.
-
-% Future work
-% - generalise to encode protocols as typestates (e.g. Mungo example)
-% - WebRTC for full MPST (since current limitation is w.r.t. MPST protocols with 1-server many-client roles
-
-\nocite{*}
-\bibliographystyle{eptcs}
-\bibliography{generic}
-\end{document}

### 2020-01-24 PLACES_2020/intro.tex
Split tex files

--- 2020-01-24 PLACES_2020/main.tex
@@
-\documentclass[submission,copyright,creativecommons]{eptcs}
-\providecommand{\event}{PLACES 2020} % Name of the event you are submitting to
-%\usepackage{breakurl}             % Not needed if you use pdflatex only.
-
-\usepackage{wrapfig}
-\usepackage{graphicx}
-\usepackage{subcaption}
-\usepackage{stmaryrd}
-\usepackage[strings]{underscore}
-\usepackage{color}
-\usepackage{etoolbox}
-\newtoggle{draft}
-\definecolor{purple}{rgb}{0.65, 0.12, 0.82}
-\definecolor{darkgray}{rgb}{.4,.4,.4}
-
-\usepackage{listings}
-\usepackage[noabbrev, capitalise]{cleveref}
-\lstdefinelanguage{Scribble}{
-  keywords={module, type, from, as, global,protocol, role, to, choice, at, or, do},
-  keywordstyle=\color{blue}\bfseries,
-  identifierstyle=\color{black},
-  sensitive=false,
-  comment=[l]{//},
-  morecomment=[s]{/*}{*/},
-  commentstyle=\color{darkgray}\ttfamily,
-  stringstyle=\color{purple}\ttfamily,
-  morestring=[b]',
-  morestring=[b]",
-}
-\lstdefinelanguage{JavaScript}{
-  keywords={typeof, new, true, false, catch, function, return, null, catch, switch, var, if, in, while, do, else, case, break, class, export, boolean, throw, implements, import, this, const, let, extends},
-  keywordstyle=\color{blue}\bfseries,
-%  ndkeywords={class, export, boolean, throw, implements, import, this},
-%  ndkeywordstyle=\color{darkgray}\bfseries,
-  identifierstyle=\color{black},
-  sensitive=false,
-  comment=[l]{//},
-  morecomment=[s]{/*}{*/},
-  commentstyle=\color{darkgray}\ttfamily,
-  stringstyle=\color{purple}\ttfamily,
-  morestring=[b]',
-  morestring=[b]"
-}
-\lstset{
-	basicstyle=\footnotesize\ttfamily,
-	numbers=left,
-	frame=single,
-	tabsize=4
-}
-\usepackage{caption}
-
-\newcommand{\sectionref}[1]{\textsection~\ref{#1}}
-
-\toggletrue{draft}
-\newcommand{\fz}[1]{\iftoggle{draft}{{\color{red}FZ: #1}}{}}
-\newcommand{\am}[1]{\iftoggle{draft}{{\color{blue}AM: #1}}{}}
-
-\title{Verifying Interactive WebSocket Applications in TypeScript}
-\author{Anson Miu
-\institute{Imperial College London}
-\and
-Francisco Ferreira
-\institute{Imperial College London}
-\and
-Nobuko Yoshida
-\institute{Imperial College London}
-\and
-Fangyi Zhou
-\institute{Imperial College London}
-}
-\def\titlerunning{Verifying Interactive WebSocket Applications in TypeScript}
-\def\authorrunning{A. Miu, F. Ferreira, N. Yoshida \& F. Zhou}
-\begin{document}
-\maketitle
-
-\begin{abstract}
-Advancements in mobile device compute power have made interactive web
-applications possible, allowing the web browser to render contents dynamically
-and support low-latency communication with the server.
-This comes at a cost to the developer, who now needs to reason more about the
-communication correctness of their application as they support more complex
-communication patterns.
-
-Multiparty session types (MPST) provide a framework for verifying conformance
-of implementations to their prescribed communication protocol.
-Existing proposals for applying the MPST framework in application development
-either neglect the event-driven nature of web applications, or lack
-compatibility with industry tools and practices, which discourages mainstream
-adoption by web developers.
-
-In this paper, we present an implementation of the MPST framework for
-developing interactive web applications using familiar industry tools in
-TypeScript and the \textit{React.js} framework.
-The developer can use the Scribble protocol
-language to specify the protocol and use the Scribble toolchain to validate
-and obtain the local protocol for each role.
-The local protocol describes the interactions of the global communication
-protocols observed by the role.
-We encode the local protocol into TypeScript types, catering for server-side
-and client-side targets separately, and show that our encoding guarantees that
-only implementations which conform to the protocol can type-check.
-We demonstrate the effectiveness of our approach through a web-based
-implementation of the classic \textit{Noughts and Crosses} game from a MPST
-formalism of the game logic.
-
-\end{abstract}
-
 \section{Introduction}
@@
 We will use the \textit{Noughts and Crosses} game as our running example in the rest of the paper and show how our approach is a practical compromise that combines benefits from static session typing and being compatible with common libraries and idiomatic JavaScript practices (\sectionref{section:example}).
-
-\section{The Scribble Framework}
-\label{section:scribble}
-
-Development begins by specifying the permitted communications between
-participants as a \textit{global protocol} in Scribble \cite{Scribble}, a
-MPST-based protocol specification language and code generation toolchain.
-We specify the \textit{Noughts and Crosses} game as a Scribble protocol in
-\cref{lst:game}.
-
-\begin{figure}[!h]
-\begin{lstlisting}[
-	language=Scribble
-]
-module NoughtsAndCrosses;
-type <typescript> "Coordinate" from "./Types" as Point;	// Position on board
-
-global protocol Game(role Svr, role P1, role P2) {
-  Pos(Point) from P1 to Svr;
-  choice at Svr {
-    Lose(Point) from Svr to P2; Win(Point) from Svr to P1;
-  } or {
-    Draw(Point) from Svr to P2; Draw(Point) from Svr to P1;
-  } or {
-    Update(Point) from Svr to P2; Update(Point) from Svr to P1;
-    do Game(Svr, P2, P1); // Continue the game with player roles swapped
-  }
-}
-\end{lstlisting}
-\captionof{lstlisting}{Main body of the \textit{Noughts and Crosses} protocol.}
-\label{lst:game}
-\end{figure}
-
-We leverage the Scribble toolchain to check for protocol
-well-formedness.
-This directly corresponds to multiparty session
-type theory \cite{MPST}:
-\fz{Cite the featherweight scribble paper for the correspondence?}
-a Scribble protocol maps to some \textit{global type}, and protocol
-verification implements the algorithmic projection defined in \cite{MPST} to
-derive valid local type \textit{projections} for all participants.
-% Note: Scribble Projection is not MPST projection, because you can have
-% protocols with choices that send to different roles.
-We obtain a set of \textit{endpoint protocols} -- one for each role from a
-well-formed global protocol.
-An endpoint protocol only preserves the interactions defined by the global
-protocol in which the target role is involved, and corresponds to an equivalent
-\textit{Endpoint Finite State Machine} (EFSM) \cite{ICALP13CFSM}.
-We use the EFSMs as a basis for API generation and adopt the formalisms in
-\cite{Hybrid2016}.
-
-\section{Encoding Session Types in TypeScript}
-% Design choices
-
-Developers can implement their application using APIs generated from the EFSM
-to guarantee correctness by construction.
-Our approach integrates the EFSM into the development workflow by encoding
-session types as TypeScript types.
-There are inherent differences between front- and back-end web development:
-business logic tends to initiate communication in the back-end, whilst
-communication in the front-end is mostly driven by user interactions with the
-browser. This motivates our design choice of generating different session types
-encodings for server (\cref{section:server}) and browser targets
-(\cref{section:browser}).
-
-\fz{I think there is an important thing to note here, ``Server'' side is
-  basically the side which accepts the connection and ``Client'' side is the
-side which makes the connection. But this is orthogonal to whether it runs on
-Node or on browser? (Correct me if I'm wrong)}
-
-\am{Not for the case of WebSockets, because only WebSocket clients on browser can initiate connections and only WebSocket Server on Node can accept; will make this distinction clear.}
-
-\subsection{Server-side API generation}
-\fz{Could you remove the state names in figures?}
-\label{section:server}
-
-\begin{wrapfigure}{r}{0.5\textwidth}
-  \vspace{-5mm}
-  \begin{center}
-    \includegraphics[width=0.48\textwidth]{figures/efsm_svr.png}
-  \end{center}
-
-  \vspace{-5mm}
-  \captionof{figure}{EFSM for \texttt{Svr}.}
-  \label{fig:efsmsvr}
-\vspace{-1cm}
-\end{wrapfigure}
-
-We refer to the \texttt{Svr} EFSM (\cref{fig:efsmsvr}) as a running example in
-this section.
-For server-side targets, we encode EFSM states into TypeScript types and
-considering receive and send states separately.
-Each TypeScript encoding is assigned to its state identifier from the EFSM as
-its type alias, thus providing syntactic sugar when referring to the successor
-state in the TypeScript encoding of the current state.
-For any state $S$ in the EFSM, we refer to the TypeScript type alias of its
-encoding as $\llbracket S \rrbracket$.
-
-\paragraph{Branching state}
-We consider a receive state as a unary branching state for conciseness.
-A branching state is encoded as an \textit{object literal}
-\cite{TypeScriptSpec}, with each branch $i \in I$, $I$ denoting set of all
-branches, corresponding to a member field.
-A branch expecting to receive a message labelled $\texttt{label}_i$ carrying
-payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as an
-\textit{member field} named $\texttt{label}_i$ of function type
-$(payload:\texttt{T}_i) \to \llbracket S_i \rrbracket$.
-The developer implements a branching operation by passing callbacks for each
-branch, parameterised by the expected message payload type for that branch.
-
-\paragraph{Selection state}
-We consider a send state as a unary selection state for conciseness.
-A selection state is encoded as a \textit{union type}
-\cite{TypeScriptSpec} of internal choice encodings: each internal choice $i \in
-I$, $I$ denoting set of all choices, sending a message labelled
-$\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state
-$S_i$ is encoded as a \textit{tuple type} of \texttt{[Labels.label$_i$, T$_i$,
-  $\llbracket S_i \rrbracket$]}.
-The developer implements a selection operation by passing the selected label
-and payload to send in the message.
-We generate a \textit{string enum} (named \texttt{Labels}) wrapping the labels
-in the protocol.
-
-\begin{figure}[ht]
-\begin{lstlisting}[language=JavaScript]
-export type S13 = { Pos: (payload: Point) => S15 };
-export type S15 = [ Labels.Lose, Point, S16 ]
-                | [ Labels.Draw, Point, S17 ]
-                | [ Labels.Update, Point, S18 ];
-\end{lstlisting}
-\captionof{lstlisting}{Example encodings from \textit{Noughts and Crosses} \texttt{Svr} EFSM.}
-\label{lst:svr}
-\end{figure}
-
-We make a key design decision \textit{not} to expose communication channels in
-the TypeScript session type encoding to provide \textit{static} linearity
-guarantees (\cref{section:serverlinear}).
-Our encoding sufficiently exposes seams for the developer to inject their
-business logic, whilst the generated session API
-(\cref{section:serversessionapi}) handles the sending and receiving of
-messages;
-as a result, the encoding does not concern with the role involved in the
-send/receive action. We outline the encoding below using examples from the
-\textit{Noughts and Crosses} game (\cref{lst:svr}).
-
-\subsubsection{Session runtime}
-\label{section:serversessionapi}
-
-Our session runtime performs communication in a protocol-conformant manner, but
-does not expose these IO actions to the developer by delegating the
-aforementioned responsibilities to an inner class.
-The runtime listens to message (receive) events on the communication channel,
-invokes the corresponding callback to obtain the value to send next, and
-performs the send.
-The developer instantiates the session by constructing an instance of the
-session runtime class, providing the WebSocket endpoint URL (to open the
-connection) and initial state (to execute the EFSM).
-
-\subsubsection{Linear channel usage}
-\label{section:serverlinear}
-Channel linearity is guaranteed based on the correctness of our library design
-and session runtime implementation, which means a faulty implementation could
-violate this, but is up to the library author to verify once rather than the
-end user. Our library design prevent the two properties below:
-
-\paragraph{Repeated use}
-We do not expose channels to the developer, which makes \textit{reuse}
-impossible.
-For example, to send a message, the generated API only requires the payload
-that needs to be sent, and the session runtime performs the send internally,
-guaranteeing this action is done \textit{exactly once} by construction.
-
-\paragraph{Unused}
-The initial state must be supplied to the session runtime
-constructor in order to instantiate the session;
-this initial state is defined
-in terms of the successor states, which in turn has references to its
-successors and so forth.
-This encoding approach covers the terminal state (if it exists), and the
-session runtime guarantees this terminal state, if it exists, will be reached
-by construction.
-
-\subsection{Browser-side API generation}
-\label{section:browser}
-
-\begin{wrapfigure}{r}{0.4\textwidth}
-  \begin{center}
-    \includegraphics[width=0.4\textwidth]{figures/efsm_p1.png}
-  \end{center}
-
-  \captionof{figure}{EFSM for \texttt{P1}.}
-  \label{fig:efsmp1}
-\end{wrapfigure}
-
-We refer to the \texttt{P1} EFSM (\cref{fig:efsmp1}) as a running example in
-this section.
-Preserving behavioural typing and channel linearity are known to be challenging
-for browser-side applications due to EFSM transitions being triggered by user
-events:
-in the case of \textit{Noughts and Crosses}, once the user makes a move by
-clicking on a cell on the game board, this click event must be deactivated
-until the user's next turn, otherwise the user can click again and violate
-channel linearity.
-Our design goal is to enforce this statically through the generated APIs.
-
-For browser-side targets, we extend the approach presented in \cite{MVU2019} on
-\textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU)
-architecture.
-Each state in the EFSM is a \emph{model type} and uniquely defines a
-\textit{view function}, set of \textit{messages} and \textit{update function}:
-the view function defines what to render to the DOM, the set of messages define
-the possible (IO) actions available at that state, and the update function
-defines which successor state to transition to, given some supported IO action
-at this state.
-
-\subsubsection{The React Framework}
-
-\fz{This subsubsection is a bit misplaced, probably move it up}
-
-We apply the multiple model types approach in \cite{MVU2019} to the React
-framework \cite{React} developed by Facebook, albeit originally intended for
-the \textit{Model-View-Controller} (MVC) architecture.
-React is widely used in industry to create scalable single-page TypeScript
-applications, and we intend for our proposed workflow to be beneficial in an
-industrial context.
-We introduce the key features of the framework.
-
-\paragraph{Components}
-A component is a reusable UI element which
-contains its own markup and logic.
-Components implement a \texttt{render()} function which returns a React
-element, the smallest building blocks of a React application, analogous to the
-view function in the MVU architecture.
-Components can keep \textit{state} and the \texttt{render()} function is
-invoked upon a change of state.
-A simple counter can be implemented as a component,
-with \texttt{count} stored as state, a button which increments \texttt{count}
-when clicked and a \texttt{div} that renders the current
-\texttt{count}.
-Components can also render other components, which gives rise
-to a parent/child relationship between components.
-Parents can pass data to children as \textit{props} (short for properties).
-Going back to the aforementioned example, the the counter component could
-render a child component \texttt{<StyledDiv count=\{this.state.count\} />} in
-its \texttt{render()} function, propagating the \texttt{count} from its state
-to the child.
-This enables reusability, and for our use case, gives a control to the parent
-on what data to pass to its children (e.g. pass the payload of a received
-message to a child to render).
-
-\fz{Is VDOM needed to be covered here?}
-\paragraph{Virtual DOM (VDOM)}
-React components are rendered on a logical
-abstraction of the DOM, which in turn performs a \texttt{diff} on the current
-browser DOM and patches the delta accordingly.
-This allows the developer to
-directly specify what should be rendered without having to worry about which
-elements to append or remove from the browser DOM.
-
-\subsubsection{Model types in React}
-
-\paragraph{State}
-An EFSM state is encoded as an \textit{abstract} React
-component.
-This is an abstract class to require the developer to provide their
-own view function, which translates conveniently to the \texttt{render()}
-function of React components.
-Our session runtime (\cref{section:clientruntime}) ``executes'' the EFSM and
-renders the current state.
-Upon transitioning to a successor state, the successor's view function will be
-invoked, as per the semantics expressed in \cite{MVU2019}.
-
-\paragraph{Model transitions}
-Transitions are encoded as React component props onto the encoded states by the
-session runtime (\cref{section:clientruntime}).
-We motivate the design choice of not exposing channel resources to provide
-guarantees on channel usage.
-React components in TypeScript are
-\textit{generic} \cite{TypeScriptSpec}, parameterised by the permitted
-types of prop and state.
-The parameters allow us to leverage the TypeScript compiler to
-verify that the props for model transitions stay local to the state they are
-defined for.
-The model transitions for EFSMs are message send and receive.
-
-\subparagraph{Send}
-We make the assumption that message sending is triggered by
-some user-driven UI event (e.g. clicking a button, pressing a key on the
-keyboard) which interacts with some DOM element.
-We could pass a
-\texttt{send()} function as a prop to the sending state, but the developer
-would be free to call the function multiple times which makes channel reuse
-possible.
-Instead, we pass a \textit{factory function} as a prop, which will,
-given an HTML event and a event handler function, return a fresh React
-component that binds the sending action on construction.
-So once the bound event is triggered, our session runtime executes the event
-handler function to obtain the payload to send, perform the send
-\textit{exactly once} and transition to (which, in practice, means render) the
-successor state.
-
-\begin{figure}[!h]
-\begin{lstlisting}[language=JavaScript, tabsize=4]
-// Inside some render() function..
-{board.map((row, x) => (
-	row.map((col, y) => {
-		const SelectPoint = this.props.Pos('click', (event: UIEvent) => {
-			event.preventDefault();
-			return { x: x, y: y };}
-		return <SelectPoint><td>.</td></SelectPoint>;
-});}
-\end{lstlisting}
-\captionof{lstlisting}{Model transition for message sending in
-\textit{Noughts and Crosses} \texttt{P1} implementation.}
-\label{lst:clientapp}
-\end{figure}
-
-We demonstrate the semantics using the \textit{Noughts and Crosses} example in
-\cref{lst:clientapp}.
-\texttt{this.props.Pos} is the factory function prop
-passed from the session runtime.
-For each x-y coordinate on the game board, we
-create a \texttt{SelectPoint} React component from the factory function (which
-reads ``build a React component that sends the \texttt{Pos} message with x-y
-coordinates as payload when the user clicks on it'') and we wrap a table cell
-(the game board is rendered as a HTML table) inside the \texttt{SelectPoint}
-component to bind the click event on the table cell.
-
-\subparagraph{Receive}
-The payload of a message receive transition is passed as
-a prop to the successor state and can be handled in the successor's view
-function.
-Upon a message receive event, the session runtime renders the
-successor of the receive state and propagates the payload to the successor
-accordingly.
-
-\subsubsection{Session runtime}
-\label{section:clientruntime}
-
-Our session runtime can be interpreted as an abstraction on top of the React
-VDOM that implements the EFSM by construction.
-The session runtime itself is a React component too, named after the endpoint
-role identifier:
-it opens the WebSocket connection to the server, keeps track of the current
-EFSM state as part of its React component state, and most importantly, renders
-the React component encoding of the active EFSM state.
-Channel communications are managed by the runtime, which allows it to render
-the successor of a receive state upon receiving a message from the channel.
-Similarly, the session runtime is responsible for passing the required props
-for model transitions to EFSM state React components.
-The session runtime component is rendered by the developer and requires, as
-props, the \textit{endpoint URL} (so it can open the connection) and a list of
-\textit{concrete state components}.
-
-The developer writes their own implementation of each state (mainly to
-customise how the state is rendered and inject business logic into state
-transitions) by extending the abstract React class components.
-The session runtime requires references to these concrete components order to
-render the user implementation accordingly.
-
-\subsubsection{Affine channel usage}
-A limitation of our browser-side session type encoding is only being able to
-guarantee that channel resources are used \textit{at most once} as supposed to
-\textit{exactly once}.
-
-Communication channels are not exposed to the developer so multiple sends are
-impossible.
-This does not restrict the developer from binding the send action to exactly
-one UI event: for \textit{Noughts and Crosses}, we bind the \texttt{Pos(Point)}
-send action to each unoccupied cell on the game board, but the generated
-runtime ensures that, once the cell is clicked, the send is only performed once
-and the successor state is rendered on the DOM, so the channel resource used to
-send becomes unavailable.
-
-However, our approach \textit{does not} statically detect whether all
-transitions in a certain state are bound to some UI event.
-This means that it is possible for an implementation to \textit{not} handle
-transitions to a terminal state but still type-check, so we cannot prevent
-unused states.
-\fz{A client may also close the browser, hence dropping the connection?}
-
-\section{Case Study}
-\label{section:example}
-
-\begin{figure}
-\begin{lstlisting}[language=JavaScript, tabsize=4]
-const handleP1Move: Receive_P1 = (move: Point) => {
-	board.P1(move);			// User logic
-	if (board.won()) { return [Labels.Lose, move, move]; }
-	else if (board.draw()) { return [Labels.Draw, move, move]; }
-	else { return [Labels.Update, move, handleP2Move]; }}
-
-// Instantiate session - `handleP2Move` defined similarly
-new NoughtsAndCrosses.Svr(webSocketServer, handleP1Move, handleP2Move);
-\end{lstlisting}
-\captionof{lstlisting}{Session runtime instantiation for \textit{Noughts and Crosses} \texttt{Svr}.}
-\label{lst:svrprotocol}
-\end{figure}
-
-We apply our framework to implement a web-based implementation of the
-\textit{Noughts and Crosses} running example in TypeScript;
-the interested reader can find the full implementation in
-\cite{NoughtsAndCrosses}.
-In addition to MPST-safety, we show that our library design welcomes idiomatic
-JavaScript practices in the user implementation and is interoperable with
-common front- and back-end frameworks.
-
-\paragraph{Game server}
-We set up the WebSocket server inside an Express.js \cite{ExpressJS}
-application on top of a Node.js \cite{NodeJS} runtime.
-We define our own game logic in a \texttt{Board} class to keep track of the
-game state and expose methods to query the result.
-This custom logic is integrated into our \texttt{Receive\_P1} and
-\texttt{Receive\_P2} handlers (\cref{lst:svrprotocol}), so the session runtime
-can handle \texttt{Pos(Point)} messages from players and transition according
-to the injected game logic.
-
-\fz{Where is Line 6?}
-Note that, by the structural typing nature of TypeScript, replacing
-\texttt{handleP2Move} on line 6 with \texttt{handleP1Move} would be
-type-correct -- this allows for better code reuse as supposed to defining
-additional abstractions to work around the limitations of nominal typing in
-\cite{Hybrid2016}.
-There is also full type erasure when transpiling to JavaScript to run the
-server code, so state space explosions is never a runtime consideration.
-
-\paragraph{Game clients}
-We implement the game client for \texttt{P1} and
-\texttt{P2} by extending from the generated abstract React (EFSM state)
-components and registering those to the session runtime component.
-
-For the sake of code reuse for demonstration purposes, \cite{NoughtsAndCrosses}
-uses \textit{higher-order components} (HOC) to build the correct state
-implementations depending on which player the user chooses to be.
-We leverage the \textit{Redux} state management library to keep track of game
-state, thus showing the flexibility of our library design in being
-interoperable with other libraries and idiomatic JavaScript practices.
-
-\section{Related Works}
-\fz{Try to avoid using {[X]} at the beginning of a sentence.}
-
-Native language support for first-class linear channel resources \cite{ATS} and
-code generation are two main approaches for incorporating MPST into application
-development.
-The latter closely relates to our proposal;
-we highlight two areas of existing works under this approach that motivate our
-design choice.
-
-\paragraph{Endpoint API generation}
-\cite{Python2017} targeted Python applications and the generation of runtime
-monitors to dynamically verify communication patterns.
-Whilst the same approach could be applied to JavaScript, we can provide more
-static guarantees with TypeScript's gradual typing system and compiler.
-\cite{Hybrid2016} targeted Java applications and proposed to encode the EFSM
-states and transitions as classes and instance methods respectively, with
-behavioural typing achieved statically by the type system and channel linearity
-guarantees achieved dynamically since channel resources are exposed and
-aliasing is not monitored.
-
-\paragraph{Session types in web development}
-\cite{PureScript2019} targeted web development in PureScript using the
-\textit{Concur UI} framework and proposed a type-level encoding of EFSMs as
-multi-parameter type classes.
-However, it presents a trade-off between achieving static linearity guarantees
-from the type-level EFSM encoding under the expressive type system and
-providing an intuitive development experience to developers, especially given
-the prevalence of JavaScript and TypeScript applications in industry.
-\cite{MVU2019} focused on applying binary session types in front-end web
-development and presented approaches that tackle the challenge of guaranteeing
-linearity in the event-driven environment.
-
-Our work applies the aforementioned approaches in a \textit{multiparty} context
-using industrial tools and practices to ultimately encourage MPST-safe web
-application development workflows in industry.
-
-%\paragraph{Typestate programming} \dots
-
-\section{Conclusion and Future Works}
-We have presented a MPST-based framework for developing full-stack interactive
-TypeScript applications with WebSocket communications that conform to a
-protocol, statically providing linear channel usage guarantees and affine
-channel usage guarantees for back-end and front-end targets respectively.
-
-Future works include incorporating \textit{explicit connection actions}
-introduced in \cite{ExplicitConnections} in our API generation to better model
-real-world communication protocols that may feature in interactive web
-applications.
-Whilst our approach supports multiparty sessions, the nature of
-WebSockets require some server-based role in the communication protocol and
-clients to interact via the server.
-Extending support to WebRTC \fz{cite it or we don't know what it is} would
-cater for peer-to-peer communication between browsers, which further opens up
-possibilities for communication protocols supported by our approach.
-
-% Future work
-% - generalise to encode protocols as typestates (e.g. Mungo example)
-% - WebRTC for full MPST (since current limitation is w.r.t. MPST protocols with 1-server many-client roles
-
-\nocite{*}
-\bibliographystyle{eptcs}
-\bibliography{generic}
-\end{document}

### 2020-01-24 PLACES_2020/main.tex
Split tex files

--- 2020-01-23 PLACES_2020/main.tex
@@
 \maketitle
-
-\begin{abstract}
-Advancements in mobile device compute power have made interactive web
-applications possible, allowing the web browser to render contents dynamically and support low-latency communication with the server.
-This comes at a cost to the developer, who now needs to reason more about the communication correctness of their application as they support more complex communication patterns.
-
-Multiparty session types (MPST) provide a framework for verifying conformance
-of implementations to their prescribed communication protocol. Existing
-proposals for applying the MPST framework in application development either
-neglect the event-driven nature of web applications, or lack compatibility with
-industry tools and practices which discourages mainstream adoption by web developers.
-
-In this paper, we present an implementation of the MPST framework for developing interactive web applications using familiar industry tools in TypeScript and the \textit{React.js} framework.
-The global communication protocol is specified in the Scribble protocol language and checked for well-formedness by the Scribble toolchain.
-We obtain the local protocol for each role, which describes the interactions of the global communication protocols observed by the role.
-We encode the local protocol into TypeScript types, catering for server-side and client-side targets separately, and show that our encoding guarantees that only implementations which conform to the protocol will type-check.
-We demonstrate the effectiveness of our approach through a web-based implementation of the classic \textit{Noughts and Crosses} game from a MPST formalism of the game logic.
-
-\end{abstract}
-
-\section{Introduction}
-% Situation: stateful communication over WebSocket and interactive web applications
-
-Modern interactive web applications aim to provide a highly responsive user experience by minimising the communication latency between clients and servers.
-Whilst the HTTP request-response model is sufficient for retrieving static assets, applying the same stateless communication approach for interactive use cases (such as real-time multiplayer games) introduces undesirable performance overhead from having to frequently set up new connections for client-server interactions.
-Developers have since adopted other communication transport abstractions over HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy low-latency full-duplex client-server communication in their applications over a single persistent connection.
-Enabling more complex communication patterns caters for more interactive use cases, but introduces additional correctness concerns to the developer.
-
-% Example: noughts and crosses
-Consider a classic turn-based board game of \textit{Noughts and Crosses} between two players.
-Both players are identified by either noughts or crosses respectively, and take turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player wins (when their markers form one straight line on the board) or a stalemate is reached (when all cells are occupied).
-A web-based implementation may involve players connected to a game server via WebSocket connections and interacting with the game from their web browser, which serve a \textit{single-page application} (SPA) of the game client written in a popular framework like \textit{React.js} \cite{React}.
-SPAs feature a single HTML page and dynamically renders content via JavaScript in the browser.
-Players take turns to make a move on the game board and the server implements the game logic to progress the game forward until a result (either a win/loss or draw) can be declared.
-
-Whilst WebSockets make this web-based implementation possible, it introduces the developer to a new family of communication errors, even for this simple game.
-In addition to the usual testing for game logic correctness, the developer needs to test against \textit{deadlocks} (e.g.
-both players waiting for each other to make a move at the same time) and \textit{communication mismatches} (e.g.
-player 1 sending a boolean to the game server instead of the board coordinates).
-The complexity of these errors correlate to the complexity of related tests required, scale with the complexity of communication patterns involved.
-
-\textit{Multiparty Session Types} (MPST) \cite{MPST} provide a framework for formally specifying a structured communication pattern between concurrent processes and verifying implementations for correctness with respect to the communications aspect.
-By specifying the client-server interactions of our game as a protocol and verifying the implementations against the protocol for well-formedness, MPST theory guarantees well-formed implementations to be free from communication errors.
-
-We see the application of the MPST methodology to verifying
-interactive web applications written in TypeScript to be an
-interesting design space -- to what extent can the MPST methodology be
-applied such that we leverage TypeScript language features to maximise
-static safety guarantees whilst preserving a flexible, natural and
-idiomatic workflow for TypeScript developers? Such a workflow would
-ultimately decrease the overhead for incorporating MPST into
-mainstream web system, which reduces development time by
-programmatically
-verifying communication correctness of the implementation.
-
-\paragraph{Contributions}
-This paper presents a workflow for developing type-safe interactive SPAs motivated by the MPST framework: \textbf{(1)} An endpoint API code generation workflow targeting TypeScript-based web applications for multiparty sessions; \textbf{(2)} An encoding of session types in server-side TypeScript that enforces static linearity; and \textbf{(3)} An encoding of session types in browser-side TypeScript using the React framework that guarantees affine usage of communication channels.
-
-%Figure \ref{fig:workflow} illustrates our proposed development workflow: we extend the Scribble framework (\sectionref{section:scribble}) and generate TypeScript endpoints for server-side (\sectionref{section:server}) and browser-side (\sectionref{section:browser}) targets.
-We will use the \textit{Noughts and Crosses} game as our running example in the rest of the paper and show how our approach is a practical compromise that combines benefits from static session typing and being compatible with common libraries and idiomatic JavaScript practices (\sectionref{section:example}).
-
-\section{The Scribble Framework}
-\label{section:scribble}
-
-Development begins by specifying the permitted communications between participants as a \textit{global protocol} in Scribble \cite{Scribble}, a MPST-based protocol specification language and code generation toolchain.
-We specify the \textit{Noughts and Crosses} game as a Scribble protocol in Listing \ref{lst:game}.
-
-\begin{figure}[!h]
-\begin{lstlisting}[
-	language=Scribble
-]
-module NoughtsAndCrosses;
-type <typescript> "Coordinate" from "./Types" as Point;	// Position on board
-
-global protocol Game(role Svr, role P1, role P2) {
-  Pos(Point) from P1 to Svr;
-  choice at Svr {
-    Lose(Point) from Svr to P2; Win(Point) from Svr to P1;
-  } or {
-    Draw(Point) from Svr to P2; Draw(Point) from Svr to P1;
-  } or {
-    Update(Point) from Svr to P2; Update(Point) from Svr to P1;
-    do Game(Svr, P2, P1);
-  }
-}
-\end{lstlisting}
-\captionof{lstlisting}{Main body of the \textit{Noughts and Crosses} protocol.}
-\label{lst:game}
-\end{figure}
-
-We leverage the Scribble toolchain to check for protocol
-well-formedness.
-This directly corresponds to multiparty session
-type theory \cite{MPST}:
-\fz{Cite the featherweight scribble paper for the correspondence?}
-a Scribble protocol maps to some \textit{global type}, and protocol verification implements the algorithmic projection
-defined in \cite{MPST} to derive valid local type \textit{projections} for all participants.  We obtain a set of \textit{endpoint protocols}
--- one for each role from a well-formed global protocol.
-An endpoint protocol only preserves the interactions defined by the global protocol in which the target role is involved, and corresponds to an equivalent \textit{Endpoint Finite State Machine} (EFSM) in the Scribble toolchain.
-We use the EFSMs as a basis for API generation and adopt the formalisms in \cite{Hybrid2016}.
-
-\section{Encoding Session Types in TypeScript}
-% Design choices
-
-Developers can implement their application using APIs generated from the EFSM to guarantee correctness by construction.
-Our approach integrates the EFSM into the development workflow by encoding
-session types as TypeScript types.
-There are inherent differences between front- and back-end web development: business logic tends to initiate communication in the back-end, whilst communication in the front-end is mostly driven by user interactions with the browser.
-This motivates our design choice of generating different session types encodings for server (\sectionref{section:server}) and browser targets (\sectionref{section:browser}).
-
-\fz{I think there is an important thing to note here, ``Server'' side is
-  basically the side which accepts the connection and ``Client'' side is the
-side which makes the connection. But this is orthogonal to whether it runs on
-Node or on browser? (Correct me if I'm wrong)}
-
-\am{Not for the case of WebSockets, because only WebSocket clients on browser can initiate connections and only WebSocket Server on Node can accept; will make this distinction clear.}
-
-\subsection{Server-side API generation}
-\fz{Could you remove the state names in figures?}
-\label{section:server}
-
-\begin{wrapfigure}{r}{0.5\textwidth}
-  \begin{center}
-    \includegraphics[width=0.48\textwidth]{figures/efsm_svr.png}
-  \end{center}
-
-  \captionof{figure}{EFSM for \texttt{Svr}.}
-  \label{fig:efsmsvr}
-\end{wrapfigure}
-
-We refer to the \texttt{Svr} EFSM (fig.~\ref{fig:efsmsvr}) as a
-running example in this section. For server-side targets, we encode
-EFSM states as types, considering
-receive and send states separately. Each TypeScript encoding is
-assigned to its state identifier from the EFSM as its type alias, thus providing syntactic sugar when referring to the successor state in the TypeScript encoding of the current state.
-For any state $S$ in the EFSM, we refer to the TypeScript type alias of its encoding as $\llbracket S \rrbracket$.
-
-We make a key design decision \textit{not} to expose communication channels in the TypeScript session type encoding to provide \textit{static} linearity guarantees (\sectionref{section:serverlinear}).
-Our encoding sufficiently exposes seams for the developer to inject their business logic, whilst the generated session API (\sectionref{section:serversessionapi}) handles the sending and receiving of messages; as a result, the encoding does not concern with the role involved in the send/receive action.
-We outline the encoding below using examples from the \textit{Noughts and Crosses} game (Listing \ref{lst:svr}).
-
-\paragraph{Branching state} We consider a receive state as a unary branching state for conciseness.
-A branching state is encoded as an \textit{object literal} \cite{TypeScriptSpec}, with each branch $i \in I$, $I$ denoting set of all branches, corresponding to a member field.
-A branch expecting to receive a message labelled $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as an \textit{member field} named $\texttt{label}_i$ of function type \texttt{(payload:$\texttt{T}_i$) => $\llbracket S_i \rrbracket$}.
-The developer implements a branching operation by passing callbacks for each branch, parameterised by the expected message payload type for that branch.
-
-\paragraph{Selection state} We consider a send state as a unary selection state for conciseness.
-A selection state is encoded as a \textit{union type} \cite{TypeScriptSpec} of internal choice encodings: each internal choice $i \in I$, $I$ denoting set of all choices, sending a message labelled $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as a \textit{tuple type} of \texttt{[Labels.label$_i$, T$_i$, $\llbracket S_i \rrbracket$]}.
-The developer implements a selection operation by passing the selected label and payload to send in the message.
-We generate a \textit{string enum} (named \texttt{Labels}) wrapping the labels in the protocol.
-
-\begin{figure}[!h]
-\begin{lstlisting}[language=JavaScript]
-export type S13 = { Pos: (payload: Point) => S15 };
-export type S15 = [Labels.Lose, Point, S16]
-                | [Labels.Draw, Point, S17]
-                | [Labels.Update, Point, S18];
-\end{lstlisting}
-\captionof{lstlisting}{Example encodings from \textit{Noughts and Crosses} \texttt{Svr} EFSM.}
-\label{lst:svr}
-\end{figure}
-
-\subsubsection{Session runtime}
-\label{section:serversessionapi}
-
-Our session runtime performs communication in a protocol-conformant manner, but does not expose these IO actions to the developer by delegating the aforementioned responsibilities to an inner class.
-The runtime listens to message (receive) events on the communication channel, invokes the corresponding callback to obtain the value to send next, and performs the send.
-The developer instantiates the session by constructing an instance of the session runtime class, providing the WebSocket endpoint URL (to open the connection) and initial state (to execute the EFSM).
-
-\subsubsection{Linear channel usage}
-\label{section:serverlinear}
-Channel linearity is guaranteed based on the correctness of our library design and session runtime implementation, which means a faulty implementation could violate this, but is up to the library author to verify once rather than the end user.
-Our library design prevent the two properties below:
-
-\paragraph{Repeat use} We do not expose channels to the programmer, which makes \textit{reuse} impossible.
-For example, to send a message, the generated API only requires the payload that needs to be sent, and the session runtime performs the send internally, guaranteeing this action is done \textit{exactly once} by construction.
-
-\paragraph{Unused} The initial state must be supplied to the session runtime
-constructor in order to instantiate the session; this initial state is defined
-in terms of the successor states, which in turn has references to its
-successors and so forth. This encoding approach covers the terminal state (if it exists), and the session runtime guarantees this terminal state, if it exists, will be reached by construction.
-
-\subsection{Browser-side API generation}
-\label{section:browser}
-
-We refer to the \texttt{P1} EFSM (figure \ref{fig:efsmp1}) as a
-running example in this section. Preserving behavioural typing and
-channel linearity are known to be challenging for browser-side
-applications due to EFSM transitions being triggered by user events:
-in the case of \textit{Noughts and Crosses}, once the user makes a move
-by clicking on a cell on the game board, this click event must be
-deactivated until the user's next turn, otherwise the user can click
-again and violate channel linearity. Our design goal is to enforce
-this statically through the generated APIs.
-
-For browser-side targets, we extend the approach presented in \cite{MVU2019} on \textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU) architecture.
-Each state in the EFSM is a model type and uniquely defines a \textit{view function}, set of \textit{messages} and \textit{update function}: the view function defines what to render to the DOM, the set of messages define the possible (IO) actions available at that state, and the update function defines which successor state to transition to, given some supported IO action at this state.
-
-\begin{wrapfigure}{r}{0.4\textwidth}
-  \begin{center}
-    \includegraphics[width=0.4\textwidth]{figures/efsm_p1.png}
-  \end{center}
-
-  \captionof{figure}{EFSM for \texttt{P1}.}
-  \label{fig:efsmp1}
-\end{wrapfigure}
-
-\subsubsection{The React Framework}
-
-\fz{This subsubsection is a bit misplaced, probably move it up}
-
-We apply the multiple model types approach in \cite{MVU2019} to the React framework \cite{React} developed by Facebook, albeit originally intended for the \textit{Model-View-Controller} (MVC) architecture.
-React is widely used in industry to create scalable single-page TypeScript applications, and we intend for our proposed workflow to be beneficial in an industrial context.
-We introduce the key features of the framework.
-
-\paragraph{Components} A component is a reusable UI element which
-contains its own markup and logic. Components implement a
-\texttt{render()} function which returns a React element, the smallest
-building blocks of a React application. Components can keep
-\textit{state} and the \texttt{render()} function is invoked upon a
-change of state. A simple counter can be implemented as a component,
-with \texttt{count} stored as state, a button which increments \texttt{count}
-when clicked and a \texttt{div} that renders the current
-\texttt{count}. Components can also render other components, which gives rise
-to a parent/child relationship between components. Parents can pass
-data to children as \textit{props}: going back to the aforementioned
-example, the the counter component could render a child component
-\texttt{<StyledDiv count=\{this.state.count\} />} in its
-\texttt{render()} function, propagating the \texttt{count} from its
-state to the child. This enables reusability, and for our use case,
-gives a control to the parent on what data to pass to its children (e.g. pass the payload of a received message to a child to render).
-
-\fz{Is VDOM needed to be covered here?}
-\paragraph{Virtual DOM (VDOM)} React components are rendered on a logical abstraction of the DOM, which in turn performs a \texttt{diff} on the current browser DOM and patches the delta accordingly.
-This allows the programmer to directly specify what should be rendered without having to worry about which elements to append or remove from the browser DOM.
-
-\subsubsection{Model types in React}
-
-\paragraph{State} An EFSM state is encoded as an \textit{abstract} React component.
-This is an abstract class to require the programmer to provide their own view function, which translates conveniently to the \texttt{render()} function of React components.
-Our session runtime (\sectionref{section:clientruntime}) ``executes'' the EFSM and renders the current state, so upon transitioning to a successor state, the successor's view function will be invoked, as per the semantics expressed in \cite{MVU2019}.
-
-\paragraph{Model transitions} Transitions are encoded as React component props onto the encoded states by the session runtime (\sectionref{section:clientruntime}).
-We motivate the design choice of not exposing channel resources to provide guarantees on channel usage.
-React components in TypeScript are \textit{generic} \cite{TypeScriptSpec} which are parameterised by the permitted types of prop and state, which allows us to leverage the TypeScript compiler to verify that the props for model transitions stay local to the state they are defined for.
-The model transitions for EFSMs are message send and receive.
-
-\subparagraph{Send} We make the assumption that message sending is triggered by some user-driven UI event (e.g.
-clicking a button, pressing a key on the keyboard) which interacts with some DOM element.
-We could pass a \texttt{send()} function as a prop to the sending state, but the programmer would be free to call the function multiple times which makes channel reuse possible.
-Instead, we pass a \textit{factory function} as a prop, which will, given an HTML event and a event handler function, return a fresh React component that binds the sending action on construction, so once the bound event is triggered, our session runtime executes the event handler function to obtain the payload to send, perform the send \textit{exactly once} and transition to (which, in practice, means render) the successor state.
-
-\begin{figure}[!h]
-\begin{lstlisting}[language=JavaScript, tabsize=4]
-// Inside some render() function..
-{board.map((row, x) => (
-	row.map((col, y) => {
-		const SelectPoint = this.props.Pos('click', (event: UIEvent) => {
-			event.preventDefault();
-			return { x: x, y: y };}
-		return <SelectPoint><td>.</td></SelectPoint>;
-});}
-\end{lstlisting}
-\captionof{lstlisting}{Model transition for message sending in
-\textit{Noughts and Crosses} \texttt{P1} implementation.}
-\label{lst:clientapp}
-\end{figure}
-
-We clarify the semantics using the \textit{Noughts and Crosses} example in Listing \ref{lst:clientapp}.
-\texttt{this.props.Pos} is the factory function prop passed from the session runtime.
-For each x-y coordinate on the game board, we create a \texttt{SelectPoint} React component from the factory function (which reads ``build a React component that sends the \texttt{Pos} message with x-y coordinates as payload when the user clicks on it'') and we wrap a table cell (the game board is rendered as a HTML table) inside the \texttt{SelectPoint} component to bind the click event on the table cell.
-
-\subparagraph{Receive} The payload of a message receive transition is passed as a prop to the successor state and can be handled in the successor's view function.
-Upon a message receive event, the session runtime renders the successor of the receive state and propagates the payload to the successor accordingly.
-
-\subsubsection{Session runtime}
-\label{section:clientruntime}
-
-Our session runtime can be interpreted as an abstraction on top of the React VDOM that implements the EFSM by construction.
-The session runtime itself is a React component too, named after the endpoint role identifier: it opens the WebSocket connection to the server, keeps track of the current EFSM state as part of its React component state, and most importantly, renders the React component encoding of the active EFSM state.
-Channel communications are managed by the runtime, which allows it to render the successor of a receive state upon receiving a message from the channel.
-Similarly, the session runtime is responsible for passing the required props for model transitions to EFSM state React components.
-The session runtime component is rendered by the programmer and requires, as props, the \textit{endpoint URL} (so it can open the connection) and a list of \textit{concrete state components}.
-
-The programmer writes their own implementation of each state (mainly to customise how the state is rendered and inject business logic into state transitions) by extending the abstract React class components.
-The session runtime requires references to these concrete components order to render the user implementation accordingly.
-
-\subsubsection{Affine channel usage}
-A limitation of our browser-side session type encoding is only being able to guarantee that channel resources are used \textit{at most once} as supposed to \textit{exactly once}.
-
-Communication channels are not exposed to the programmer so multiple sends are impossible.
-This does not restrict the programmer from binding the send action to exactly one UI event: for \textit{Noughts and Crosses}, we bind the \texttt{Pos(Point)} send action to each unoccupied cell on the game board, but the generated runtime ensures that, once the cell is clicked, the send is only performed once and the successor state is rendered on the DOM, so the channel resource used to send becomes unavailable.
-
-However, our approach \textit{does not} statically detect whether all transitions in a certain state are bound to some UI event.
-This means that it is possible for an implementation to \textit{not} handle transitions to a terminal state but still type-check, so we cannot prevent unused states.
-\fz{A client may also close the browser, hence dropping the connection?}
-
-\section{Case Study}
-\label{section:example}
-
-\begin{figure}
-\begin{lstlisting}[language=JavaScript, tabsize=4]
-const handleP1Move: Receive_P1 = (move: Point) => {
-	board.P1(move);			// User logic
-	if (board.won()) { return [Labels.Lose, move, move]; }
-	else if (board.draw()) { return [Labels.Draw, move, move]; }
-	else { return [Labels.Update, move, handleP2Move]; }}
-
-// Instantiate session - `handleP2Move` defined similarly
-new NoughtsAndCrosses.Svr(webSocketServer, handleP1Move, handleP2Move);
-\end{lstlisting}
-\captionof{lstlisting}{Session runtime instantiation for \textit{Noughts and Crosses} \texttt{Svr}.}
-\label{lst:svrprotocol}
-\end{figure}
-
-We apply our framework to implement a web-based implementation of the \textit{Noughts and Crosses} running example in TypeScript; the interested reader can find the full implementation in \cite{NoughtsAndCrosses}.
-In addition to MPST-safety, we show that our library design welcomes idiomatic JavaScript practices in the user implementation and is interoperable with common front- and back-end frameworks.
-
-\paragraph{Game server} We set up the WebSocket server inside an Express.js \cite{ExpressJS} application on top of a Node.js \cite{NodeJS} runtime.
-We define our own game logic in a \texttt{Board} class to keep track of the game state and expose methods to query the result.
-This custom logic is integrated into our \texttt{Receive\_P1} and \texttt{Receive\_P2} handlers (Listing \ref{lst:svrprotocol}), so the session runtime can handle \texttt{Pos(Point)} messages from players and transition according to the injected game logic.
-
-Note that, by the structural typing nature of TypeScript, replacing \texttt{handleP2Move} on line 6 with \texttt{handleP1Move} would be type-correct - this allows for better code reuse as supposed to defining additionl abstractions to work around the limitations of nominal typing in \cite{Hybrid2016}.
-There is also full type erasure when transpiling to JavaScript to run the server code, so state space explosions is never a runtime consideration.
-
-\paragraph{Game clients} We implement the game client for \texttt{P1} and \texttt{P2} by extending from the generated abstract React (EFSM state) components and registering those to the session runtime component.
-
-For the sake of code reuse for demonstration purposes, \cite{NoughtsAndCrosses} uses \textit{higher-order components} (HOC) to build the correct state implementations depending on which player the user chooses to be.
-We leverage the \textit{Redux} state management library to keep track of game state, thus showing the flexibility of our library design in being interoperable with other libraries and idiomatic JavaScript practices.
-
-\section{Related Works}
-\fz{Try to avoid using {[X]} at the beginning of a sentence.}
-
-Native language support for first-class linear channel resources \cite{ATS} and code generation are two main approaches for incorporating MPST into application development.
-The latter closely relates to our proposal; we highlight two areas of existing works under this approach that motivate our design choice.
-
-\paragraph{Endpoint API generation} \cite{Python2017} targeted Python applications and the generation of runtime monitors to dynamically verify communication patterns.
-Whilst the same approach could be applied to JavaScript, we can provide more static guarantees with TypeScript's gradual typing system and compiler.
-\cite{Hybrid2016} targeted Java applications and proposed to encode the EFSM states and transitions as classes and instance methods respectively, with behavioural typing achieved statically by the type system and channel linearity guarantees achieved dynamically since channel resources are exposed and aliasing is not monitored.
-
-\paragraph{Session types in web development} \cite{PureScript2019} targeted web development in PureScript using the \textit{Concur UI} framework and proposed a type-level encoding of EFSMs as multi-parameter type classes.
-However, it presents a trade-off between achieving static linearity guarantees from the type-level EFSM encoding under the expressive type system and providing an intuitive development experience to programmers, especially given the prevalence of JavaScript and TypeScript applications in industry.
-\cite{MVU2019} focused on applying binary session types in front-end web development and presented approaches that tackle the challenge of guaranteeing linearity in the event-driven environment.
-
-Our work applies the aforementioned approaches in a \textit{multiparty} context using industrial tools and practices to ultimately encourage MPST-safe web application development workflows in industry.
-
-%\paragraph{Typestate programming} \dots
-
-\section{Conclusion and Future Works}
-We have presented a MPST-based framework for developing full-stack interactive TypeScript applications with WebSocket communications that conform to a protocol, statically providing linear channel usage guarantees and affine channel usage guarantees for back-end and front-end targets respectively.
-
-Future works include incorporating \textit{explicit connection actions} introduced in \cite{ExplicitConnections} in our API generation to better model real-world communication protocols that may feature in interactive web applications.
-Whilst our approach supports multiparty sessions, the nature of WebSockets require some server-based role in the communication protocol and clients to interact via the server.
-Extending support to WebRTC would cater for peer-to-peer communication between browsers, which further opens up possibilities for communication protocols supported by our approach.
-
-% Future work
-% - generalise to encode protocols as typestates (e.g. Mungo example)
-% - WebRTC for full MPST (since current limitation is w.r.t. MPST protocols with 1-server many-client roles
-
+\input{abstract.tex}
+\input{intro.tex}
+\input{scribble.tex}
+\input{encoding.tex}
+\input{casestudy.tex}
+\input{related.tex}
+\input{conclusion.tex}
 \nocite{*}

### 2020-01-24 PLACES_2020/related.tex
Split tex files

\section{Related Works}
\fz{Try to avoid using {[X]} at the beginning of a sentence.}

Native language support for first-class linear channel resources \cite{ATS} and
code generation are two main approaches for incorporating MPST into application
development.
The latter closely relates to our proposal;
we highlight two areas of existing works under this approach that motivate our
design choice.

\paragraph{Endpoint API generation}
\cite{Python2017} targeted Python applications and the generation of runtime
monitors to dynamically verify communication patterns.
Whilst the same approach could be applied to JavaScript, we can provide more
static guarantees with TypeScript's gradual typing system and compiler.
\cite{Hybrid2016} targeted Java applications and proposed to encode the EFSM
states and transitions as classes and instance methods respectively, with
behavioural typing achieved statically by the type system and channel linearity
guarantees achieved dynamically since channel resources are exposed and
aliasing is not monitored.

\paragraph{Session types in web development}
\cite{PureScript2019} targeted web development in PureScript using the
\textit{Concur UI} framework and proposed a type-level encoding of EFSMs as
multi-parameter type classes.
However, it presents a trade-off between achieving static linearity guarantees
from the type-level EFSM encoding under the expressive type system and
providing an intuitive development experience to developers, especially given
the prevalence of JavaScript and TypeScript applications in industry.
\cite{MVU2019} focused on applying binary session types in front-end web
development and presented approaches that tackle the challenge of guaranteeing
linearity in the event-driven environment.

Our work applies the aforementioned approaches in a \textit{multiparty} context
using industrial tools and practices to ultimately encourage MPST-safe web
application development workflows in industry.

%\paragraph{Typestate programming} \dots

### 2020-01-24 PLACES_2020/scribble.tex
Split tex files

\section{The Scribble Framework}
\label{section:scribble}

Development begins by specifying the permitted communications between
participants as a \textit{global protocol} in Scribble \cite{Scribble}, a
MPST-based protocol specification language and code generation toolchain.
We specify the \textit{Noughts and Crosses} game as a Scribble protocol in
\cref{lst:game}.

\begin{figure}[!h]
\begin{lstlisting}[
	language=Scribble
]
module NoughtsAndCrosses;
type <typescript> "Coordinate" from "./Types" as Point;	// Position on board

global protocol Game(role Svr, role P1, role P2) {
  Pos(Point) from P1 to Svr;
  choice at Svr {
    Lose(Point) from Svr to P2; Win(Point) from Svr to P1;
  } or {
    Draw(Point) from Svr to P2; Draw(Point) from Svr to P1;
  } or {
    Update(Point) from Svr to P2; Update(Point) from Svr to P1;
    do Game(Svr, P2, P1); // Continue the game with player roles swapped
  }
}
\end{lstlisting}
\captionof{lstlisting}{Main body of the \textit{Noughts and Crosses} protocol.}
\label{lst:game}
\end{figure}

We leverage the Scribble toolchain to check for protocol
well-formedness.
This directly corresponds to multiparty session
type theory \cite{MPST}:
\fz{Cite the featherweight scribble paper for the correspondence?}
a Scribble protocol maps to some \textit{global type}, and protocol
verification implements the algorithmic projection defined in \cite{MPST} to
derive valid local type \textit{projections} for all participants.
% Note: Scribble Projection is not MPST projection, because you can have
% protocols with choices that send to different roles.
We obtain a set of \textit{endpoint protocols} -- one for each role from a
well-formed global protocol.
An endpoint protocol only preserves the interactions defined by the global
protocol in which the target role is involved, and corresponds to an equivalent
\textit{Endpoint Finite State Machine} (EFSM) \cite{ICALP13CFSM}.
We use the EFSMs as a basis for API generation and adopt the formalisms in
\cite{Hybrid2016}.

### 2020-01-26 PLACES_2020/casestudy.tex
PLACES: address FZ comments

@@
 
-\fz{Where is Line 6?}
 Note that, by the structural typing nature of TypeScript, replacing
-\texttt{handleP2Move} on line 6 with \texttt{handleP1Move} would be
+\texttt{handleP2Move} on line 5 with \texttt{handleP1Move} would be
 type-correct -- this allows for better code reuse as supposed to defining

### 2020-01-26 PLACES_2020/conclusion.tex
PLACES: address FZ comments

@@
 clients to interact via the server.
-Extending support to WebRTC \fz{cite it or we don't know what it is} would
+Extending support to WebRTC \cite{WebRTC} would
 cater for peer-to-peer communication between browsers, which further opens up

### 2020-01-26 PLACES_2020/encoding.tex
Prettify the figures
PLACES: address FZ comments

@@
 \subsection{Server-side API generation}
-\fz{Could you remove the state names in figures?}
 \label{section:server}
@@
 
+\subsection{The React Framework}
+Our browser-side session type encodings for browser-side targets build upon the React framework, developed by Facebook \cite{React} for the \textit{Model-View-Controller} (MVC) architecture.
+React is widely used in industry to create scalable single-page TypeScript
+applications, and we intend for our proposed workflow to be beneficial in an
+industrial context.
+We introduce the key features of the framework.
+
+\paragraph{Components}
+A component is a reusable UI element which
+contains its own markup and logic.
+Components implement a \texttt{render()} function which returns a React
+element, the smallest building blocks of a React application, analogous to the
+view function in the MVU architecture.
+Components can keep \textit{state} and the \texttt{render()} function is
+invoked upon a change of state.
+A simple counter can be implemented as a component,
+with \texttt{count} stored as state, a button which increments \texttt{count}
+when clicked and a \texttt{div} that renders the current
+\texttt{count}.
+Components can also render other components, which gives rise
+to a parent/child relationship between components.
+Parents can pass data to children as \textit{props} (short for properties).
+Going back to the aforementioned example, the the counter component could
+render a child component \texttt{<StyledDiv count=\{this.state.count\} />} in
+its \texttt{render()} function, propagating the \texttt{count} from its state
+to the child.
+This enables reusability, and for our use case, gives a control to the parent
+on what data to pass to its children (e.g. pass the payload of a received
+message to a child to render).
+
+\fz{Is VDOM needed to be covered here?}
+
+\am{Intended to motivate how the session runtime implementation renders the required state and the VDOM deals with the diffs - at the same time, unsure whether this is important for the reader to understand (or just leave the DOM rendering as a black box)?}
+\paragraph{Virtual DOM (VDOM)}
+React components are rendered on a logical
+abstraction of the DOM, which in turn performs a \texttt{diff} on the current
+browser DOM and patches the delta accordingly.
+This allows the developer to
+directly specify what should be rendered without having to worry about which
+elements to append or remove from the browser DOM.
+
 \subsection{Browser-side API generation}
+
 \label{section:browser}
@@
 at this state.
-
-\subsubsection{The React Framework}
-
-\fz{This subsubsection is a bit misplaced, probably move it up}
-
-We apply the multiple model types approach in \cite{MVU2019} to the React
-framework \cite{React} developed by Facebook, albeit originally intended for
-the \textit{Model-View-Controller} (MVC) architecture.
-React is widely used in industry to create scalable single-page TypeScript
-applications, and we intend for our proposed workflow to be beneficial in an
-industrial context.
-We introduce the key features of the framework.
-
-\paragraph{Components}
-A component is a reusable UI element which
-contains its own markup and logic.
-Components implement a \texttt{render()} function which returns a React
-element, the smallest building blocks of a React application, analogous to the
-view function in the MVU architecture.
-Components can keep \textit{state} and the \texttt{render()} function is
-invoked upon a change of state.
-A simple counter can be implemented as a component,
-with \texttt{count} stored as state, a button which increments \texttt{count}
-when clicked and a \texttt{div} that renders the current
-\texttt{count}.
-Components can also render other components, which gives rise
-to a parent/child relationship between components.
-Parents can pass data to children as \textit{props} (short for properties).
-Going back to the aforementioned example, the the counter component could
-render a child component \texttt{<StyledDiv count=\{this.state.count\} />} in
-its \texttt{render()} function, propagating the \texttt{count} from its state
-to the child.
-This enables reusability, and for our use case, gives a control to the parent
-on what data to pass to its children (e.g. pass the payload of a received
-message to a child to render).
-
-\fz{Is VDOM needed to be covered here?}
-\paragraph{Virtual DOM (VDOM)}
-React components are rendered on a logical
-abstraction of the DOM, which in turn performs a \texttt{diff} on the current
-browser DOM and patches the delta accordingly.
-This allows the developer to
-directly specify what should be rendered without having to worry about which
-elements to append or remove from the browser DOM.
 
@@
 unused states.
+
 \fz{A client may also close the browser, hence dropping the connection?}

### 2020-01-26 PLACES_2020/related.tex
PLACES: address FZ comments

@@
 \section{Related Works}
-\fz{Try to avoid using {[X]} at the beginning of a sentence.}
-
 Native language support for first-class linear channel resources \cite{ATS} and
@@
 \paragraph{Endpoint API generation}
-\cite{Python2017} targeted Python applications and the generation of runtime
-monitors to dynamically verify communication patterns.
+Neykova and Yoshida targeted Python applications and the generation of runtime
+monitors \cite{Python2017} to dynamically verify communication patterns.
 Whilst the same approach could be applied to JavaScript, we can provide more
-static guarantees with TypeScript's gradual typing system and compiler.
-\cite{Hybrid2016} targeted Java applications and proposed to encode the EFSM
+static guarantees with TypeScript's gradual typing system and compiler. Scribble-Java \cite{Hybrid2016} proposed to encode the EFSM
 states and transitions as classes and instance methods respectively, with
@@
 \paragraph{Session types in web development}
-\cite{PureScript2019} targeted web development in PureScript using the
+King et al. \cite{PureScript2019} targeted web development in PureScript using the
 \textit{Concur UI} framework and proposed a type-level encoding of EFSMs as
@@
 providing an intuitive development experience to developers, especially given
-the prevalence of JavaScript and TypeScript applications in industry.
-\cite{MVU2019} focused on applying binary session types in front-end web
+the prevalence of JavaScript and TypeScript applications in industry. Fowler \cite{MVU2019} focused on applying binary session types in front-end web
 development and presented approaches that tackle the challenge of guaranteeing

### 2020-01-26 PLACES_2020/scribble.tex
PLACES: address FZ comments

@@
 This directly corresponds to multiparty session
-type theory \cite{MPST}:
-\fz{Cite the featherweight scribble paper for the correspondence?}
+type theory \cite{FeatherweightScribble}:
 a Scribble protocol maps to some \textit{global type}, and protocol

### 2020-01-27 PLACES_2020/casestudy.tex
PLACES: address FZ comments

@@
 
-Note that, by the structural typing nature of TypeScript, replacing
+Note that, by TypeScript's structural typing \cite{TypeScriptSpec}, replacing
 \texttt{handleP2Move} on line 5 with \texttt{handleP1Move} would be
@@
 There is also full type erasure when transpiling to JavaScript to run the
-server code, so state space explosions is never a runtime consideration.
+server code, so state space explosion is not a runtime consideration.
 
@@
 implementations depending on which player the user chooses to be.
-We leverage the \textit{Redux} state management library to keep track of game
+We leverage the \textit{Redux} \cite{Redux} state management library to keep track of game
 state, thus showing the flexibility of our library design in being

### 2020-01-27 PLACES_2020/encoding.tex
PLACES: address FZ comments

@@
 this section.
-Preserving behavioural typing and channel linearity are known to be challenging
+Preserving behavioural typing and channel linearity is challenging
 for browser-side applications due to EFSM transitions being triggered by user
@@
 transitions to a terminal state but still type-check, so we cannot prevent
-unused states.
-
-\fz{A client may also close the browser, hence dropping the connection?}
+unused states. Equally, our approach cannot prevent a client closing the browser, which would drop the connection.

### 2020-01-27 PLACES_2020/related.tex
PLACES: address FZ comments

@@
 behavioural typing achieved statically by the type system and channel linearity
-guarantees achieved dynamically since channel resources are exposed and
+guarantees achieved dynamically since channels are exposed and
 aliasing is not monitored.

### 2020-01-28 PLACES_2020/encoding.tex
PLACES: address FZ comments

@@
 session types as TypeScript types.
-There are inherent differences between front- and back-end web development:
-business logic tends to initiate communication in the back-end, whilst
-communication in the front-end is mostly driven by user interactions with the
-browser. This motivates our design choice of generating different session types
+Communication over the WebSocket protocol introduces additional constriants: 
+communication is always initiated in the front-end and driven by user interactions, 
+whilst back-end roles can only accept connections.
+This motivates our design choice of generating different session types
 encodings for server (\cref{section:server}) and browser targets
@@
 For server-side targets, we encode EFSM states into TypeScript types and
-considering receive and send states separately.
+consider receive and send states separately.
 Each TypeScript encoding is assigned to its state identifier from the EFSM as
@@
 Parents can pass data to children as \textit{props} (short for properties).
-Going back to the aforementioned example, the the counter component could
+Going back to the aforementioned example, the counter component could
 render a child component \texttt{<StyledDiv count=\{this.state.count\} />} in
@@
 to the child.
-This enables reusability, and for our use case, gives a control to the parent
+This enables reusability, and for our use case, gives control to the parent
 on what data to pass to its children (e.g. pass the payload of a received

### 2020-01-31 PLACES_2020/encoding.tex
Address comments from Francisco

@@
 whilst back-end roles can only accept connections.
-This motivates our design choice of generating different session types
-encodings for server (\cref{section:server}) and browser targets
-(\cref{section:browser}).
+This motivates our design of encoding the session types differently for server (\cref{section:server}) and client (\cref{section:browser}) targets.
 
@@
 \label{section:server}
+
+% Francisco: mention that it corresponds to the messages for the Noughts and Crosses example too
 
@@
 violate this, but is up to the library author to verify once rather than the
-end user. Our library design prevent the two properties below:
+end user. Our library design prevents the two properties below:
 
@@
 
-Our session runtime can be interpreted as an abstraction on top of the React
+The session runtime can be interpreted as an abstraction on top of the React
 VDOM that implements the EFSM by construction.
@@
 transitions) by extending the abstract React class components.
-The session runtime requires references to these concrete components order to
+The session runtime requires references to these concrete components in order to
 render the user implementation accordingly.
@@
 transitions to a terminal state but still type-check, so we cannot prevent
-unused states. Equally, our approach cannot prevent a client closing the browser, which would drop the connection.
+unused states. Equally, our approach does not prevent a client closing the browser, which would drop the connection.

### 2020-01-31 PLACES_2020/scribble.tex
Address comments from Francisco

@@
 
-Development begins by specifying the permitted communications between
+Development begins by specifying the expected communications between
 participants as a \textit{global protocol} in Scribble \cite{Scribble}, a

### 2020-02-02 PLACES_2020/abstract.tex

@@
 Advancements in mobile device compute power have made interactive web
-applications possible, allowing the web browser to render contents dynamically
+applications possible, allowing the web browser to render content dynamically
 and support low-latency communication with the server.

### 2020-02-02 PLACES_2020/conclusion.tex

@@
-\section{Conclusion and Future Works}
+\section{Conclusion and Future Work}
 We have presented a MPST-based framework for developing full-stack interactive

### 2020-02-02 PLACES_2020/intro.tex
Address comments from Francisco

@@
 framework like \textit{React.js} \cite{React}.
-SPAs feature a single HTML page and dynamically rendered content via JavaScript
+SPAs feature a single HTML page and dynamically render content via JavaScript
 in the browser.
@@
 The complexity of these errors correlate to the
-complexity of related tests required, scale with the complexity of
+complexity of the required tests and scale with the complexity of
 communication patterns involved.

### 2020-02-02 PLACES_2020/related.tex

@@
-\section{Related Works}
+\section{Related Work}
 Native language support for first-class linear channel resources \cite{ATS} and
@@
 The latter closely relates to our proposal;
-we highlight two areas of existing works under this approach that motivate our
+we highlight two areas of existing work under this approach that motivate our
 design choice.

### 2020-02-02 interim/sections/background/scribble.tex

@@
 
-\cite{Hybrid2016} also identifies properties of EFSMs for well-formed protocol specifications: \textbf{(1)} there is exactly one initial state, \textbf{(2)} there is at most one terminal state, and \textbf{(3)} every non-terminal state is either an input state or output state.
+We adopt the properties of EFSMs for well-formed protocol specifications from \cite{Hybrid2016} \textbf{(1)} there is exactly one initial state, \textbf{(2)} there is at most one terminal state, and \textbf{(3)} every non-terminal state is either an input state or output state.
 There are further restrictions on input and output states, namely these states must only send to or receive from the same role.

### 2020-02-02 interim/sections/evaluationplan.tex

\chapter{Evaluation Plan}

\section{Correctness}
% Qualitatively argue how the approach achieves behavioural typing and preserves channel linearity
We will qualitatively argue how our TypeScript encodings of session types achieve behavioural typing and preserves channel linearity.

% May explore formalisms if time permits
If time permits, we may explore possible formalisms of our implementation as a calculus upon which to prove correctness. This may be related to the formal language presented in \cite{UnderstandingTypeScript}, but we will need to extend this to be specific to our TypeScript encodings of session types.

\section{Productivity} 
% Compare with PureScript implementation to talk about developer productivity when using PureScript APIs with ConcurUI compared to TypeScript APIs when using React
By adopting tools and practices used in industry, we wish to assess the extent to which our implementation boosts developer productivity by providing APIs that guarantee communication protocol conformance in a way that is compatible with existing workflows. Comparing the development workflow of implementing the \textit{Battleships} game using the PureScript workflow in \cite{PureScript2019} and our TypeScript-based proposal would yield interesting findings.

% Survey - qualitative and quantitative feedback
This can be carried out as a survey to be completed by web developers familiar with both languages. We would include questions that allow us to collect both qualitative and quantitative feedback to make effective comparisons between our approach and existing proposals.

\section{Performance} Examples of performance metrics include: lines of code to be written by the developer (as a result of using the generated APIs), the size of the transpiled JavaScript assets to be loaded on the server and client browsers. We aim to extract these metrics from our approach and compare them against metrics extracted from a baseline implementation (i.e. without using the MPST methodology) and evaluate the performance overhead (if any) of our MPST-based approach.

### 2020-02-02 interim/sections/introduction.tex

@@
 Modern interactive web applications aim to provide a highly responsive user experience by minimising the communication latency between clients and servers. 
-Whilst the HTTP request-response model is sufficient for retrieving static assets, applying the same stateless communication approach for interactive use cases (such as a multiplayer game) introduces undesirable performance overhead from having to frequently set up new connections for client-server interactions.
+Whilst the HTTP request-response model is sufficient for retrieving static assets, applying the same stateless communication approach for interactive use cases (such as multiplayer games) introduces undesirable performance overhead from having to frequently set up new connections for client-server interactions.
 Developers have since adopted other communication transport abstractions over HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy low-latency full-duplex client-server communication in their applications over a single persistent connection.
-Enabling more complex communication patterns caters for more interactive use cases (such as real-time multiplayer games), but introduces additional concerns to the developer with respect to implementation correctness.
+Enabling more complex communication patterns caters for more interactive use cases, but introduces additional concerns to the developer with respect to implementation correctness.
 
@@
 
-We present a workflow for developing type-safe interactive SPAs motivated by the MPST framework: \textbf{(1)} An endpoint API code generation workflow targeting TypeScript-based web applications for multiparty sessions; \textbf{(2)} An encoding of session types in server-side TypeScript that enforces static linearity; and \textbf{(3)} An encoding of session types in browser-side TypeScript using the React framework that guarantees affine usage of communication channels.
+\paragraph{Contributions} We present a workflow for developing type-safe interactive SPAs motivated by the MPST framework:
+
+\begin{enumerate}
+\item An endpoint API code generation workflow targeting TypeScript-based web applications for multiparty sessions;
+\item An encoding of session types in server-side TypeScript that enforces static linearity;
+\item An encoding of session types in browser-side TypeScript using the React framework that guarantees affine usage of communication channels.
+\end{enumerate}

### 2020-02-03 PLACES_2020/main.tex
Title change

--- 2020-01-23 PLACES_2020/main.tex
@@
 \usepackage{color}
-\usepackage{etoolbox}
-\newtoggle{draft}
 \definecolor{purple}{rgb}{0.65, 0.12, 0.82}
@@
   morestring=[b]',
-  morestring=[b]",
+  morestring=[b]"
 }
@@
 	frame=single,
-	tabsize=4
 }
@@
 
-\toggletrue{draft}
-\newcommand{\fz}[1]{\iftoggle{draft}{{\color{red}FZ: #1}}{}}
-\newcommand{\am}[1]{\iftoggle{draft}{{\color{blue}AM: #1}}{}}
-
-\title{Verifying Interactive WebSocket Applications in TypeScript}
+\title{Generating Interactive WebSocket Applications in TypeScript}
 \author{Anson Miu
@@
 }
-\def\titlerunning{Verifying Interactive WebSocket Applications in TypeScript}
+\def\titlerunning{Generating Interactive WebSocket Applications in TypeScript}
 \def\authorrunning{A. Miu, F. Ferreira, N. Yoshida \& F. Zhou}
@@
 \begin{abstract}
-Advancements in mobile device compute power have made interactive web
-applications possible, allowing the web browser to render contents dynamically and support low-latency communication with the server.
+Advancements in mobile device compute power has made interactive web applications possible, allowing the web browser to render content dynamically and support low-latency communication with the server.
 This comes at a cost to the developer, who now needs to reason more about the communication correctness of their application as they support more complex communication patterns.
 
-Multiparty session types (MPST) provide a framework for verifying conformance
-of implementations to their prescribed communication protocol. Existing
-proposals for applying the MPST framework in application development either
-neglect the event-driven nature of web applications, or lack compatibility with
-industry tools and practices which discourages mainstream adoption by web developers.
+Multiparty session types (MPST) provide a framework for verifying implementations to conform to a prescribed communication protocol.
+Existing proposals for applying the MPST framework in application development either neglect the event-driven nature of web applications, or lack compatibility with industry tools and practices which discourage mainstream adoption by web developers.
 
@@
 Modern interactive web applications aim to provide a highly responsive user experience by minimising the communication latency between clients and servers. 
-Whilst the HTTP request-response model is sufficient for retrieving static assets, applying the same stateless communication approach for interactive use cases (such as real-time multiplayer games) introduces undesirable performance overhead from having to frequently set up new connections for client-server interactions.
+Whilst the HTTP request-response model is sufficient for retrieving static assets, applying the same stateless communication approach for interactive use cases (such as a multiplayer game) introduces undesirable performance overhead from having to frequently set up new connections for client-server interactions.
 Developers have since adopted other communication transport abstractions over HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy low-latency full-duplex client-server communication in their applications over a single persistent connection.
-Enabling more complex communication patterns caters for more interactive use cases, but introduces additional correctness concerns to the developer.
+Enabling more complex communication patterns caters for more interactive use cases (such as real-time multiplayer games), but introduces additional concerns to the developer with respect to implementation correctness.
 
@@
 player 1 sending a boolean to the game server instead of the board coordinates).
-The complexity of these errors correlate to the complexity of related tests required, scale with the complexity of communication patterns involved.
+The complexity of these errors, which correlate to the complexity of tests required against these errors, scale with the complexity of the communication patterns involved.
 
@@
 
-Development begins by specifying the permitted communications between participants as a \textit{global protocol} in Scribble \cite{Scribble}, a MPST-based protocol specification language and code generation toolchain.
+Development begins by specifying the permitted communications between participants as a \textit{global protocol} in Scribble, a MPST-based framework introduced in \cite{Scribble} which features a protocol specification language and code generation toolchain.
+This means that a protocol specified in Scribble maps to some \textit{global type} in multiparty session type theory \cite{MPST}.
 We specify the \textit{Noughts and Crosses} game as a Scribble protocol in Listing \ref{lst:game}.
@@
   choice at Svr {
-    Lose(Point) from Svr to P2; Win(Point) from Svr to P1;
+    Lose(Point) from Svr to P2;
+    Win(Point) from Svr to P1;
   } or {
-    Draw(Point) from Svr to P2; Draw(Point) from Svr to P1;
+    Draw(Point) from Svr to P2;
+    Draw(Point) from Svr to P1;
   } or {
-    Update(Point) from Svr to P2; Update(Point) from Svr to P1;
+    Update(Point) from Svr to P2;
+    Update(Point) from Svr to P1;
     do Game(Svr, P2, P1);
@@
 We leverage the Scribble toolchain to check for protocol
-well-formedness.
-This directly corresponds to multiparty session
-type theory \cite{MPST}:
-\fz{Cite the featherweight scribble paper for the correspondence?}
-a Scribble protocol maps to some \textit{global type}, and protocol verification implements the algorithmic projection
-defined in \cite{MPST} to derive valid local type \textit{projections} for all participants.  We obtain a set of \textit{endpoint protocols}
+well-formedness. This directly corresponds to the multiparty session
+type theory of verifying valid local type \textit{projections} for all
+participants of a global type. The implementation of this validation
+by the toolchain follows from the algorithmic projection procedures
+defined in \cite{MPST}. We obtain a set of \textit{endpoint protocols}
 -- one for each role from a well-formed global protocol.
@@
 
-Developers can implement their application using APIs generated from the EFSM to guarantee correctness by construction.
-Our approach integrates the EFSM into the development workflow by encoding
+Developers can validate the communication aspects of their
+implementation against their EFSM to verify protocol conformance. Our
+approach integrates the EFSM into the development workflow by encoding
 session types as TypeScript types.
 There are inherent differences between front- and back-end web development: business logic tends to initiate communication in the back-end, whilst communication in the front-end is mostly driven by user interactions with the browser.
-This motivates our design choice of generating different session types encodings for server (\sectionref{section:server}) and browser targets (\sectionref{section:browser}).
-
-\fz{I think there is an important thing to note here, ``Server'' side is
-  basically the side which accepts the connection and ``Client'' side is the
-side which makes the connection. But this is orthogonal to whether it runs on
-Node or on browser? (Correct me if I'm wrong)}
-
-\am{Not for the case of WebSockets, because only WebSocket clients on browser can initiate connections and only WebSocket Server on Node can accept; will make this distinction clear.}
+This motivates our design choice of generating different session types encodings for the target application on the server (\sectionref{section:server}) and in the browser (\sectionref{section:browser}).
 
 \subsection{Server-side API generation}
-\fz{Could you remove the state names in figures?}
 \label{section:server}
@@
 
-  \captionof{figure}{EFSM for \texttt{Svr}.}
+  \captionof{figure}{EFSM for \texttt{Svr} in \textit{Noughts and Crosses}.}
   \label{fig:efsmsvr}
@@
 
-We refer to the \texttt{Svr} EFSM (fig.~\ref{fig:efsmsvr}) as a
+We will refer to the \texttt{Svr} EFSM (fig.~\ref{fig:efsmsvr}) as a
 running example in this section. For server-side targets, we encode
-EFSM states as types, considering
+EFSM states as TypeScript types depending on the type of state. We
+adopt the EFSM definition presented in \cite{Hybrid2016} to consider
 receive and send states separately. Each TypeScript encoding is
@@
 
-We make a key design decision \textit{not} to expose communication channels in the TypeScript session type encoding to provide \textit{static} linearity guarantees (\sectionref{section:serverlinear}).
+We make a key design decision \textit{not} to expose communication channels in the TypeScript session type encoding to provide linearity guarantees (\sectionref{section:serverlinear}).
 Our encoding sufficiently exposes seams for the developer to inject their business logic, whilst the generated session API (\sectionref{section:serversessionapi}) handles the sending and receiving of messages; as a result, the encoding does not concern with the role involved in the send/receive action.
-We outline the encoding below using examples from the \textit{Noughts and Crosses} game (Listing \ref{lst:svr}).
+We outline the encoding below using examples from the \textit{Noughts and Crosses} game server (Listing \ref{lst:svr}).
 
 \paragraph{Branching state} We consider a receive state as a unary branching state for conciseness.
-A branching state is encoded as an \textit{object literal} \cite{TypeScriptSpec}, with each branch $i \in I$, $I$ denoting set of all branches, corresponding to a member field.
+A branching state is encoded as an \textit{object literal} \cite{TypeScriptSpec}, with each branch corresponding to a member field.
 A branch expecting to receive a message labelled $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as an \textit{member field} named $\texttt{label}_i$ of function type \texttt{(payload:$\texttt{T}_i$) => $\llbracket S_i \rrbracket$}.
@@
 \paragraph{Selection state} We consider a send state as a unary selection state for conciseness.
-A selection state is encoded as a \textit{union type} \cite{TypeScriptSpec} of internal choice encodings: each internal choice $i \in I$, $I$ denoting set of all choices, sending a message labelled $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as a \textit{tuple type} of \texttt{[Labels.label$_i$, T$_i$, $\llbracket S_i \rrbracket$]}.
+A selection state is encoded as a \textit{union type} \cite{TypeScriptSpec} of internal choice encodings: each internal choice sending a message labelled $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as a \textit{tuple type} of \texttt{[Labels.label$_i$, T$_i$, $\llbracket S_i \rrbracket$]}.
 The developer implements a selection operation by passing the selected label and payload to send in the message.
-We generate a \textit{string enum} (named \texttt{Labels}) wrapping the labels in the protocol.
+We generate a \textit{string enum} (named \texttt{Labels}) wrapping the labels in the protocol, hence the enum member access in the first element of the tuple type.
 
@@
 \begin{lstlisting}[language=JavaScript]
-export type S13 = { Pos: (payload: Point) => S15 };
+import { Coordinate as Point } from "./Types";
+import { Labels } from "./Constants";
+
+export type S13 = { Pos: (payload: Point) => S15 }
 export type S15 = [Labels.Lose, Point, S16]
 		| [Labels.Draw, Point, S17]
-                | [Labels.Update, Point, S18];
+		| [Labels.Update, Point, S18]
 \end{lstlisting}  
@@
 
-Our session runtime performs communication in a protocol-conformant manner, but does not expose these IO actions to the developer by delegating the aforementioned responsibilities to an inner class.
+Our session runtime performs communication in a way that conforms to the protocol specification, but does not expose these IO actions to the developer by delegating the aforementioned responsibilities to an inner class.
 The runtime listens to message (receive) events on the communication channel, invokes the corresponding callback to obtain the value to send next, and performs the send.
-The developer instantiates the session by constructing an instance of the session runtime class, providing the WebSocket endpoint URL (to open the connection) and initial state (to execute the EFSM).
+The developer instantiates the session by constructing an instance of the session runtime class.
 
@@
 \label{section:serverlinear}
-Channel linearity is guaranteed based on the correctness of our library design and session runtime implementation, which means a faulty implementation could violate this, but is up to the library author to verify once rather than the end user.
-Our library design prevent the two properties below:
+Our callback-oriented design for server-side API generation provides guarantees on state channel linearity by preventing the two properties detailed below.
+Here, channel linearity is guaranteed based on the correctness of our library design and session runtime implementation, which means a faulty implementation could violate this, but is up to the library author to verify once rather than the end user.
 
@@
 
-\paragraph{Unused} The initial state must be supplied to the session runtime
-constructor in order to instantiate the session; this initial state is defined
-in terms of the successor states, which in turn has references to its
-successors and so forth. This encoding approach covers the terminal state (if it exists), and the session runtime guarantees this terminal state, if it exists, will be reached by construction.
+\paragraph{Unused} The initial state must be supplied to the session runtime constructor in order to instantiate the session; this initial state is defined in terms of the successor states, which in turn has references to its successors and so forth.
+This encoding approach will cover the terminal state (if it exists in the EFSM, as \cite{Hybrid2016} notes that an EFSM contains at most one terminal state), and the session runtime guarantees this terminal state, if it exists, will be reached by construction.
 
@@
 
-We refer to the \texttt{P1} EFSM (figure \ref{fig:efsmp1}) as a
+We will refer to the \texttt{P1} EFSM (figure \ref{fig:efsmp1}) as a
 running example in this section. Preserving behavioural typing and
 channel linearity are known to be challenging for browser-side
-applications due to EFSM transitions being triggered by user events:
-in the case of \textit{Noughts and Crosses}, once the user makes a move
+applications due to the event-driven nature of user interaction: in
+the case of \textit{Noughts and Crosses}, once the user makes a move
 by clicking on a cell on the game board, this click event must be
@@
 
-\begin{wrapfigure}{r}{0.4\textwidth}
+\begin{wrapfigure}{r}{0.5\textwidth}
   \begin{center}
-    \includegraphics[width=0.4\textwidth]{figures/efsm_p1.png}
+    \includegraphics[width=0.48\textwidth]{figures/efsm_p1.png}
   \end{center}
 
-  \captionof{figure}{EFSM for \texttt{P1}.}
+  \captionof{figure}{EFSM for \texttt{P1} in \textit{Noughts and Crosses}.}
   \label{fig:efsmp1}
@@
 \subsubsection{The React Framework}
-
-\fz{This subsubsection is a bit misplaced, probably move it up}
-
-We apply the multiple model types approach in \cite{MVU2019} to the React framework \cite{React} developed by Facebook, albeit originally intended for the \textit{Model-View-Controller} (MVC) architecture.
+Rather than using the LINKS web programming language from \cite{MVU2019}, we apply the multiple model types approach to the React framework \cite{React} developed by Facebook.
 React is widely used in industry to create scalable single-page TypeScript applications, and we intend for our proposed workflow to be beneficial in an industrial context.
-We introduce the key features of the framework.
+We introduce the key features of the framework below.
 
@@
 change of state. A simple counter can be implemented as a component,
-with \texttt{count} stored as state, a button which increments \texttt{count}
+with the count stored as state, a button which increments the count
 when clicked and a \texttt{div} that renders the current
-\texttt{count}. Components can also render other components, which gives rise
+count. Components can also render other components, which gives rise
 to a parent/child relationship between components. Parents can pass
@@
 
-\fz{Is VDOM needed to be covered here?}
 \paragraph{Virtual DOM (VDOM)} React components are rendered on a logical abstraction of the DOM, which in turn performs a \texttt{diff} on the current browser DOM and patches the delta accordingly.
@@
 \subsubsection{Model types in React}
+We leverage React to express the model types approach in \cite{MVU2019} rather than the \textit{Model-View-Controller} (MVC) architecture it was intended for.
 
@@
 We motivate the design choice of not exposing channel resources to provide guarantees on channel usage.
-React components in TypeScript are \textit{generic} \cite{TypeScriptSpec} which are parameterised by the permitted types of prop and state, which allows us to leverage the TypeScript compiler to verify that the props for model transitions stay local to the state they are defined for.
+React components in TypeScript are \textit{generics} \cite{TypeScriptSpec} which are parameterised by the permitted types of prop and state, which allows us to leverage the TypeScript compiler to verify that the props for model transitions stay local to the state they are defined for.
 The model transitions for EFSMs are message send and receive.
@@
 \begin{lstlisting}[language=JavaScript, tabsize=4]
-// Inside some render() function..
+render() {
+	...
 	{board.map((row, x) => (
-	row.map((col, y) => {
+		board.map((col, y) => {
 			const SelectPoint = this.props.Pos('click', (event: UIEvent) => {
-			event.preventDefault();
-			return { x: x, y: y };}
-		return <SelectPoint><td>.</td></SelectPoint>;
-});}
+				event.preventDefault()
+				return { x, y }
+			}
+			return <SelectPoint><td>.</td></SelectPoint>
+		}
+	)}
+	...	
+}
 \end{lstlisting}  
@@
 \subparagraph{Receive} The payload of a message receive transition is passed as a prop to the successor state and can be handled in the successor's view function.
-Upon a message receive event, the session runtime renders the successor of the receive state and propagates the payload to the successor accordingly.
+Upon receiving the message from the channel, the session runtime renders the successor of the receive state and propagates the payload as the prop accordingly.
 
@@
 Similarly, the session runtime is responsible for passing the required props for model transitions to EFSM state React components.
-The session runtime component is rendered by the programmer and requires, as props, the \textit{endpoint URL} (so it can open the connection) and a list of \textit{concrete state components}.
+The session runtime component is rendered by the programmer and requires the following props:
 
-The programmer writes their own implementation of each state (mainly to customise how the state is rendered and inject business logic into state transitions) by extending the abstract React class components.
+\paragraph{Endpoint URL} The session runtime needs this in order to open the WebSocket connection.
+
+\paragraph{Concrete state components} The programmer writes their own implementation of each state (mainly to customise how the state is rendered and inject business logic into state transitions) by extending the abstract React class components.
 The session runtime requires references to these concrete components order to render the user implementation accordingly.
@@
 Communication channels are not exposed to the programmer so multiple sends are impossible.
-This does not restrict the programmer from binding the send action to exactly one UI event: for \textit{Noughts and Crosses}, we bind the \texttt{Pos(Point)} send action to each unoccupied cell on the game board, but the generated runtime ensures that, once the cell is clicked, the send is only performed once and the successor state is rendered on the DOM, so the channel resource used to send becomes unavailable.
+This does not restrict the programmer from binding the send action to exactly one UI event: for the \textit{Noughts and Crosses} game, we bind the \texttt{Pos(Point)} send action to each unoccupied cell on the game board, but the generated runtime ensures that, once an event bound to the send action is triggered, the send is only performed once and the successor state is rendered on the DOM to prevent multiple sends initiated by the user.
 
@@
 This means that it is possible for an implementation to \textit{not} handle transitions to a terminal state but still type-check, so we cannot prevent unused states.
-\fz{A client may also close the browser, hence dropping the connection?}
 
@@
 const handleP1Move: Receive_P1 = (move: Point) => {
-	board.P1(move);			// User logic
-	if (board.won()) { return [Labels.Lose, move, move]; }
-	else if (board.draw()) { return [Labels.Draw, move, move]; }
-	else { return [Labels.Update, move, handleP2Move]; }}
+	// User-implemented logic
+	board.P1(move)
+	if (board.won()) { return [Labels.Lose, move, move] }
+	else if (board.draw()) { return [Labels.Draw, move, move] }
+	else { return [Labels.Draw, move, handleP2Move] } 
+}
 
-// Instantiate session - `handleP2Move` defined similarly
-new NoughtsAndCrosses.Svr(webSocketServer, handleP1Move, handleP2Move);
+const handleP2Move: Receive_P2 = ...
+
+// Instantiate session
+new NoughtsAndCrosses.Svr(webSocketServer, handleP1Move, handleP2Move)
 \end{lstlisting}  
-\captionof{lstlisting}{Session runtime instantiation for \textit{Noughts and Crosses} \texttt{Svr}.}
+\captionof{lstlisting}{Example fragment of \textit{Noughts and Crosses} \texttt{Svr} implementation.}
 \label{lst:svrprotocol}
@@
 
-\paragraph{Game clients} We implement the game client for \texttt{P1} and \texttt{P2} by extending from the generated abstract React (EFSM state) components and registering those to the session runtime component.
+\paragraph{Game clients} We implement the game client for players as a React application - extending from the generated abstract React components, registering our custom implementation as props for the session runtime component.
 
@@
 \section{Related Works}
-\fz{Try to avoid using {[X]} at the beginning of a sentence.}
-
-Native language support for first-class linear channel resources \cite{ATS} and code generation are two main approaches for incorporating MPST into application development.
-The latter closely relates to our proposal; we highlight two areas of existing works under this approach that motivate our design choice.
+%\paragraph{Languages with native session type support} \dots
+% New languages written  (e.g. ATS)
+%	- addresses the challenge of resource linearity with a liner type system
+%	- problem with webdev: not compatible with libraries, technologies, flexibility of JavaScript (ultimately the language of the browser)
 
@@
 
-Our work applies the aforementioned approaches in a \textit{multiparty} context using industrial tools and practices to ultimately encourage MPST-safe web application development workflows in industry.
-
-%\paragraph{Typestate programming} \dots
+Our work applies the aforementioned approaches in a \textit{multiparty} context in a manner that is compatible with industrial tools and practices to ultimately invite the notion of MPST-safety into interactive web application development in industry.
 
@@
 Future works include incorporating \textit{explicit connection actions} introduced in \cite{ExplicitConnections} in our API generation to better model real-world communication protocols that may feature in interactive web applications.
-Whilst our approach supports multiparty sessions, the nature of WebSockets require some server-based role in the communication protocol and clients to interact via the server.
+Whilst our approach does support multiparty sessions, protocols are required to have exactly one role running on the server and client roles must interact with each other via the server to respect the properties of WebSocket connections, which is the transport abstraction we support.
 Extending support to WebRTC would cater for peer-to-peer communication between browsers, which further opens up possibilities for communication protocols supported by our approach.

### 2020-03-03 PLACES_2020/casestudy.tex
Address comments from Francisco
PLACES: address typo comments

@@
 \paragraph{Game server}
-We set up the WebSocket server inside an Express.js \cite{ExpressJS}
+We set up the WebSocket server as an Express.js \cite{ExpressJS}
 application on top of a Node.js \cite{NodeJS} runtime.
@@
 \texttt{handleP2Move} on line 5 with \texttt{handleP1Move} would be
-type-correct -- this allows for better code reuse as supposed to defining
+type-correct -- this allows for better code reuse as opposed to defining
 additional abstractions to work around the limitations of nominal typing in

### 2020-03-03 PLACES_2020/encoding.tex
PLACES: final modifications
PLACES: address typo comments

@@
 For any state $S$ in the EFSM, we refer to the TypeScript type alias of its
-encoding as $\llbracket S \rrbracket$.
+encoding as $\llbracket S \rrbracket$. We outline the encoding below using examples from the
+\textit{Noughts and Crosses} game (\cref{lst:svr}).
 
@@
 as a result, the encoding does not concern with the role involved in the
-send/receive action. We outline the encoding below using examples from the
-\textit{Noughts and Crosses} game (\cref{lst:svr}).
+send/receive action.
 
@@
 violate this, but is up to the library author to verify once rather than the
-end user. Our library design prevents the two properties below:
+end user. Our library design prevents the two conditions detailed below:
 
@@
 Instead, we pass a \textit{factory function} as a prop, which will,
-given an HTML event and a event handler function, return a fresh React
+given a HTML event and an event handler function, return a fresh React
 component that binds the sending action on construction.
@@
 A limitation of our browser-side session type encoding is only being able to
-guarantee that channel resources are used \textit{at most once} as supposed to
+guarantee that channel resources are used \textit{at most once} as opposed to
 \textit{exactly once}.

### 2020-03-03 PLACES_2020/encoding.tex
PLACES: address reviewer comments

@@
 
+% REVIEW:
+% Also, when describing Listing 2, it would be helpful to say a bit more explicitly that the client is supposed to implement S13, and the code in S13 returns an object of type S15.
+Then you can point to Listing 4 if they are coordinated.
+In the case of \cref{lst:svr}, the developer is expected to implement 
+\texttt{S13} which handles the \texttt{Pos} message sent by P1,
+and the code in \texttt{S13} returns a value of type \texttt{S15}, which corresponds to a selection of messages to send to P2.
+\cref{lst:svrprotocol} illustrates how the developer may implement these types.
+
 We make a key design decision \textit{not} to expose communication channels in
@@
 (\cref{section:serversessionapi}) handles the sending and receiving of
-messages;
-as a result, the encoding does not concern with the role involved in the
-send/receive action.
+messages.
 

### 2020-03-03 PLACES_2020/encoding.tex
an HTML

@@
 Instead, we pass a \textit{factory function} as a prop, which will,
-given a HTML event and an event handler function, return a fresh React
+given an HTML event and an event handler function, return a fresh React
 component that binds the sending action on construction.
@@
 coordinates as payload when the user clicks on it'') and we wrap a table cell
-(the game board is rendered as a HTML table) inside the \texttt{SelectPoint}
+(the game board is rendered as an HTML table) inside the \texttt{SelectPoint}
 component to bind the click event on the table cell.

### 2020-03-03 PLACES_2020/casestudy.tex

@@
 \begin{lstlisting}[language=JavaScript, tabsize=4]
-const handleP1Move: Receive_P1 = (move: Point) => {
+const handleP1Move: S13 = (move: Point) => {
 	board.P1(move);			// User logic
@@
 
-// Instantiate session - `handleP2Move` defined similarly
-new NoughtsAndCrosses.Svr(webSocketServer, handleP1Move, handleP2Move);
+// Instantiate session - `handleP2Move` defined similarly as S18
+new NoughtsAndCrosses.Svr(webSocketServer, handleP1Move);
 \end{lstlisting}
@@
 game state and expose methods to query the result.
-This custom logic is integrated into our \texttt{Receive\_P1} and
-\texttt{Receive\_P2} handlers (\cref{lst:svrprotocol}), so the session runtime
+This custom logic is integrated into our \texttt{handleP1Move} and
+\texttt{handleP2Move} handlers (\cref{lst:svrprotocol}), so the session runtime
 can handle \texttt{Pos(Point)} messages from players and transition according

### 2020-03-03 PLACES_2020/intro.tex

@@
 Such a workflow would ultimately decrease the overhead for incorporating MPST
-into mainstream web system, which reduces development time by programmatically
+into mainstream web development, which reduces development time by programmatically
 verifying communication correctness of the implementation.

### 2020-03-05 PLACES_2020/casestudy.tex
Minor changes and comments

@@
 
-\paragraph{Game server}
+\paragraph{Game Server}
 We set up the WebSocket server as an Express.js \cite{ExpressJS}
@@
 
-\paragraph{Game clients}
+\paragraph{Game Clients}
 We implement the game client for \texttt{P1} and

### 2020-03-05 PLACES_2020/encoding.tex
Increase figure sizes
Minor changes and comments

@@
 
-\subsection{Server-side API generation}
+\subsection{Server-Side API Generation}
 \label{section:server}
@@
 
-\begin{wrapfigure}{r}{0.5\textwidth}
+\begin{wrapfigure}{r}{0.6\textwidth}
   \vspace{-5mm}
   \begin{center}
-    \includegraphics[width=0.48\textwidth]{figures/efsm_svr.png}
+    \includegraphics[width=0.58\textwidth]{figures/efsm_svr.png}
   \end{center}
@@
 
-\paragraph{Branching state}
+\paragraph{Branching State}
 We consider a receive state as a unary branching state for conciseness.
@@
 
-\paragraph{Selection state}
+\paragraph{Selection State}
 We consider a send state as a unary selection state for conciseness.
@@
 
-\subsubsection{Session runtime}
+\subsubsection{Session Runtime}
 \label{section:serversessionapi}
@@
 
-\subsubsection{Linear channel usage}
+\subsubsection{Linear Channel Usage}
 \label{section:serverlinear}
@@
 
-\paragraph{Repeated use}
+\paragraph{Repeated Usage}
 We do not expose channels to the developer, which makes \textit{reuse}
@@
 
-\paragraph{Unused}
+\paragraph{Unused Channels}
 The initial state must be supplied to the session runtime
@@
 
-\subsection{Browser-side API generation}
+\subsection{Browser-Side API Generation}
 
@@
 
-\begin{wrapfigure}{r}{0.4\textwidth}
+\begin{wrapfigure}{r}{0.5\textwidth}
   \begin{center}
-    \includegraphics[width=0.4\textwidth]{figures/efsm_p1.png}
+    \includegraphics[width=0.5\textwidth]{figures/efsm_p1.png}
   \end{center}
@@
 
-\subsubsection{Model types in React}
+\subsubsection{Model Types in React}
 
@@
 
-\subparagraph{Send}
+\subparagraph{Sending}
 We make the assumption that message sending is triggered by
@@
 
-\subparagraph{Receive}
+\subparagraph{Receiving}
 The payload of a message receive transition is passed as
@@
 
-\subsubsection{Session runtime}
+\subsubsection{Session Runtime}
 \label{section:clientruntime}
@@
 
-\subsubsection{Affine channel usage}
+\subsubsection{Affine Channel Usage}
 A limitation of our browser-side session type encoding is only being able to

### 2020-03-05 PLACES_2020/intro.tex
Minor changes and comments

@@
 By specifying the client-server interactions of our game as an MPST protocol
-and verifying the implementations against the protocol for well-formedness,
+and verifying the implementations against the protocol for conformance,
 MPST theory guarantees well-formed implementations to be free from
@@
 
+\fz{This paragraph needs to be modified, to emphasis that our generated APIs
+are correct by construction, and not ``verifying'' existing ones}
 We see the application of the MPST methodology to verifying interactive web

### 2020-03-05 PLACES_2020/related.tex
Minor changes and comments

@@
 
-\paragraph{Endpoint API generation}
+\paragraph{Endpoint API Generation}
 Neykova and Yoshida targeted Python applications and the generation of runtime
@@
 
-\paragraph{Session types in web development}
+\paragraph{Session Types in Web Development}
 King et al. \cite{PureScript2019} targeted web development in PureScript using the

### 2020-03-08 PLACES_2020/casestudy.tex
PLACES: address reviewers/FZ comments

@@
 	else if (board.draw()) { return [Labels.Draw, move, move]; }
-	else { return [Labels.Update, move, handleP2Move]; }}
+	else { return [Labels.Update, move, move, handleP2Move]; }
+}
 
@@
 \texttt{handleP2Move} handlers (\cref{lst:svrprotocol}), so the session runtime
-can handle \texttt{Pos(Point)} messages from players and transition according
-to the injected game logic.
+can handle \texttt{Pos(Point)} messages from players and transition to the 
+permitted successor states (\cref{lst:game}) according
+to the injected game logic: if \texttt{P1} played a winning move (line 3), 
+\texttt{Svr} sends a \texttt{Lose} message to \texttt{P2} with the winning 
+move, and also sends a \texttt{Win} message to \texttt{P1} 
+(note that we do not need to specify the label for a single choice); 
+if \texttt{P1}'s move resulted 
+in a draw (line 4), \texttt{Svr} sends \texttt{Draw} messages to 
+both \texttt{P2} and \texttt{P1}; otherwise, the game continues (line 5), 
+so \texttt{Svr} updates both \texttt{P2} and \texttt{P1} with the latest 
+move and proceeds to handle \texttt{P2}'s turn.
 
 Note that, by TypeScript's structural typing \cite{TypeScriptSpec}, replacing
-\texttt{handleP2Move} on line 5 with \texttt{handleP1Move} would be
+\texttt{handleP2Move} on line 5 with a recursion of \texttt{handleP1Move} would be
 type-correct -- this allows for better code reuse as opposed to defining
@@
 interoperable with other libraries and idiomatic JavaScript practices.
+% REVIEW: tight-coupling?
+Our approach encourages the separation of concerns between the 
+communication logic and program logic -- the generated session runtime
+keeps track of the state of the EFSM to ensure protocol conformance by 
+construction, whilst \textit{Redux} solely manages our game state.

### 2020-03-08 PLACES_2020/encoding.tex
PLACES: make wrapfig float to avoid overflow
PLACES: address reviewers/FZ comments

@@
 
-Our session runtime performs communication in a protocol-conformant manner, but
+The generated code for our session runtime performs communication in a protocol-conformant manner, but
 does not expose these IO actions to the developer by delegating the
 aforementioned responsibilities to an inner class.
+% REVIEW: it is not clear what is
+% generated on the server-side and what actually is the session runtime
+The runtime executes the EFSM by keeping track of 
+the current state (similar to the generated code in \cite{javatypestate}) and only permitting the allowed IO actions at the current state.
 The runtime listens to message (receive) events on the communication channel,
@@
 
-\begin{wrapfigure}{r}{0.5\textwidth}
+\begin{wrapfigure}{R}{0.5\textwidth}
   \begin{center}
@@
 architecture.
-Each state in the EFSM is a \emph{model type} and uniquely defines a
-\textit{view function}, set of \textit{messages} and \textit{update function}:
-the view function defines what to render to the DOM, the set of messages define
-the possible (IO) actions available at that state, and the update function
-defines which successor state to transition to, given some supported IO action
-at this state.
+% FEEDBACK: introduce MVU and model type
+An MVU application features a \textit{model} encapsulating application 
+state, a \textit{view function} rendering the state on the DOM, and an 
+\textit{update function} handling \textit{messages} produced by the 
+rendered model to produce a new model.
+The concept of model types express type dependencies between these 
+components: a \emph{model type} uniquely defines a \textit{view function}, 
+set of \textit{messages} and \textit{update function} -- rather than 
+producing a new model, the update function defines valid transitions to 
+other model types.
+We leverage the correspondence between model types and states in the EFSM:
+each state in the EFSM is a model type, the set of messages represent
+the possible (IO) actions available at that state,
+and the update function defines which successor state to transition to, 
+given the supported IO actions at this state.
 

### 2020-03-08 PLACES_2020/intro.tex
PLACES: address reviewers/FZ comments

@@
 
-\fz{This paragraph needs to be modified, to emphasis that our generated APIs
-are correct by construction, and not ``verifying'' existing ones}
-We see the application of the MPST methodology to verifying interactive web
-applications written in TypeScript to be an interesting design space -- to what
-extent can the MPST methodology be applied such that we leverage TypeScript
-language features to maximise static safety guarantees whilst preserving a
-flexible, natural and idiomatic workflow for TypeScript developers?
+We see the application of the MPST methodology to generating interactive 
+TypeScript web applications to be an interesting design space 
+-- to what extent can the MPST methodology be applied to
+deliver a workflow where developers use the generated TypeScript APIs in
+their application to guarantee protocol conformance by construction?
 Such a workflow would ultimately decrease the overhead for incorporating MPST

### 2020-03-08 PLACES_2020/casestudy.tex
PLACES: update case study code listing

--- 2020-03-05 PLACES_2020/casestudy.tex
@@
 	board.P1(move);			// User logic
-	if (board.won()) { return [Labels.Lose, move, move]; }
-	else if (board.draw()) { return [Labels.Draw, move, move]; }
-	else { return [Labels.Update, move, handleP2Move]; }}
+	if (board.won()) {
+		return [Labels.Lose, move, [Labels.Win, move]]; (*@\label{linelose}@*)
+	} else if (board.draw()) {
+		return [Labels.Draw, move, [Labels.Draw, move]]; (*@\label{linedraw}@*)
+	} else {
+		return [Labels.Update, move, [Labels.Update, move, handleP2Move]]; 	(*@\label{lineupdate}@*)
+	}
+}
 
@@
 \texttt{handleP2Move} handlers (\cref{lst:svrprotocol}), so the session runtime
-can handle \texttt{Pos(Point)} messages from players and transition according
-to the injected game logic.
+can handle \texttt{Pos(Point)} messages from players and transition to the 
+permitted successor states (\cref{lst:game}) according
+to the injected game logic: if \texttt{P1} played a winning move (\cref{linelose}), 
+\texttt{Svr} sends a \texttt{Lose} message to \texttt{P2} with the winning 
+move, and also sends a \texttt{Win} message to \texttt{P1}; 
+if \texttt{P1}'s move resulted 
+in a draw (\cref{linedraw}), \texttt{Svr} sends \texttt{Draw} messages to 
+both \texttt{P2} and \texttt{P1}; otherwise, the game continues (\cref{lineupdate}), 
+so \texttt{Svr} updates both \texttt{P2} and \texttt{P1} with the latest 
+move and proceeds to handle \texttt{P2}'s turn.
 
 Note that, by TypeScript's structural typing \cite{TypeScriptSpec}, replacing
-\texttt{handleP2Move} on line 5 with \texttt{handleP1Move} would be
+\texttt{handleP2Move} on \cref{lineupdate} with a recursion of \texttt{handleP1Move} would be
 type-correct -- this allows for better code reuse as opposed to defining

### 2020-03-08 PLACES_2020/main.tex
Revert "Title change"
This reverts commit da4bd3602c8f414b269ba7d373a81aff0bd72977.

Revert "change"

This reverts commit 1481d262fe516097ad80b0fdc8bc2d55127abd8c, reversing
changes made to a7ddeea4f6e88eaf60b73a2668464374dd7a2e94.
Try to fix main
PLACES: address typo comments
PLACES: add acks
Minor changes and comments
PLACES: update case study code listing

--- 2020-01-24 PLACES_2020/main.tex
@@
 	frame=single,
-	tabsize=4
+	tabsize=4,
+	escapeinside={(*@}{@*)}
 }
@@
 
-\title{Verifying Interactive WebSocket Applications in TypeScript}
+\title{Generating Interactive WebSocket Applications in TypeScript}
 \author{Anson Miu
@@
 \input{conclusion.tex}
+\input{acks.tex}
 \nocite{*}

### 2020-03-09 PLACES_2020/abstract.tex
Proofread first 2 pages

@@
 \begin{abstract}
-Advancements in mobile device compute power have made interactive web
-applications possible, allowing the web browser to render content dynamically
+Advancements in mobile device computing power have made interactive web
+applications possible, allowing the web browser to render contents dynamically
 and support low-latency communication with the server.
-This comes at a cost to the developer, who now needs to reason more about the
-communication correctness of their application as they support more complex
-communication patterns.
+This comes at a cost to the developer, who now needs to reason more about
+correctness of communication patterns in their application as web applications
+support more complex communication patterns.
 
@@
 of implementations to their prescribed communication protocol.
-Existing proposals for applying the MPST framework in application development
+Existing proposals for applying the MPST framework in application developments
 either neglect the event-driven nature of web applications, or lack
@@
 In this paper, we present an implementation of the MPST framework for
-developing interactive web applications using familiar industry tools in
+developing interactive web applications using familiar industry tools using
 TypeScript and the \textit{React.js} framework.
@@
 language to specify the protocol and use the Scribble toolchain to validate
-and obtain the local protocol for each role.
+and obtain the \emph{local protocol} for each role.
 The local protocol describes the interactions of the global communication
-protocols observed by the role.
+protocol observed by the role.
 We encode the local protocol into TypeScript types, catering for server-side
-and client-side targets separately, and show that our encoding guarantees that
-only implementations which conform to the protocol can type-check.
+and client-side targets separately.
+We show that our encoding guarantees that only implementations which conform to
+the protocol can type-check.
 We demonstrate the effectiveness of our approach through a web-based

### 2020-03-09 PLACES_2020/intro.tex
Proofread first 2 pages

@@
 between two players.
-Both players, identified by either noughts or crosses respectively, take
+Both players, identified by either \emph{noughts (O's)} or \emph{crosses (X's)} respectively, take
 turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player
 wins (when their markers form one straight line on the board) or a stalemate is
-reached (when all cells are occupied).
+reached (when all cells are occupied and no one wins).
 A web-based implementation may involve players connected to a game server via
@@
 Players take turns to make a move on the game board, which sends a message to
-the server, and the server implements the game logic to progress the game
+the server.
+The server implements the game logic to progress the game
 forward until a result (either a win/loss or draw) can be declared, where
@@
 
-Whilst WebSockets make this web-based implementation possible, it introduces
+Whilst WebSockets make this web-based implementation possible, they introduce
 the developer to a new family of communication errors, even for this simple
@@
 In addition to the usual testing for game logic correctness, the
-developer needs to test against \textit{deadlocks} (e.g. both players waiting
+developer needs to test against \textit{deadlocks} (e.g.\ both players waiting
 for each other to make a move at the same time) and \textit{communication
- mismatches} (e.g. player 1 sending a boolean to the game server instead of
+ mismatches} (e.g.\ player 1 sending a boolean to the game server instead of
 the board coordinates).
-The complexity of these errors correlate to the
-complexity of the required tests and scale with the complexity of
+The complexity of these errors correlates to the
+complexity of the required tests and scales with the complexity of
 communication patterns involved.
@@
 TypeScript web applications to be an interesting design space
--- to what extent can the MPST methodology be applied to
+--- to what extent can the MPST methodology be applied to
 deliver a workflow where developers use the generated TypeScript APIs in

### 2020-03-09 PLACES_2020/scribble.tex
PLACES: address reviewers/FZ comments
Proofread first 2 pages

@@
 
-Development begins by specifying the expected communications between
+Development begins with specifying the expected communications between
 participants as a \textit{global protocol} in Scribble \cite{Scribble}, a
@@
 \cref{lst:game}.
+In the protocol, the role \texttt{Svr} stands for the Server, and the roles
+\texttt{P1} and \texttt{P2} stand for the two Players respectively.
 
@@
 \end{lstlisting}
-\captionof{lstlisting}{Main body of the \textit{Noughts and Crosses} protocol.}
+\captionof{lstlisting}{\textit{Noughts and Crosses} in a Scribble protocol.}
 \label{lst:game}
@@
 type theory \cite{FeatherweightScribble}:
-a Scribble protocol maps to some \textit{global type}, and protocol
-verification implements the algorithmic projection defined in \cite{MPST} to
+a Scribble protocol maps to some \textit{global type}, and the Scribble
+toolchain implements the algorithmic projection defined in \cite{MPST} to
 derive valid local type \textit{projections} for all participants.
@@
 % protocols with choices that send to different roles.
-We obtain a set of \textit{endpoint protocols} -- one for each role from a
+We obtain a set of \textit{endpoint protocols} (corresponds to \emph{local
+  types}) --- one for each role from a
 well-formed global protocol.
@@
 \textit{Endpoint Finite State Machine} (EFSM) \cite{ICALP13CFSM}.
+% Review: The meaning and hte usage of EFSM should be explained
+The EFSM holds information about the permitted IO actions for the role.
 We use the EFSMs as a basis for API generation and adopt the formalisms in

### 2020-03-09 PLACES_2020/encoding.tex
WIP Proofreading Sec. 3

@@
 session types as TypeScript types.
-Communication over the WebSocket protocol introduces additional constriants:
+Communication over the WebSocket protocol introduces additional constraints:
 communication is always initiated in the front-end and driven by user interactions,
 whilst back-end roles can only accept connections.
-This motivates our design of encoding the session types differently for server (\cref{section:server}) and client (\cref{section:browser}) targets.
-
-\fz{I think there is an important thing to note here, ``Server'' side is
-  basically the side which accepts the connection and ``Client'' side is the
-side which makes the connection. But this is orthogonal to whether it runs on
-Node or on browser? (Correct me if I'm wrong)}
-
-\am{Not for the case of WebSockets, because only WebSocket clients on browser can initiate connections and only WebSocket Server on Node can accept; will make this distinction clear.}
+This motivates our design of encoding the session types differently for server
+(\cref{section:server}) and client (\cref{section:browser}) targets.
 
@@
 For server-side targets, we encode EFSM states into TypeScript types and
-consider receive and send states separately.
-Each TypeScript encoding is assigned to its state identifier from the EFSM as
-its type alias, thus providing syntactic sugar when referring to the successor
-state in the TypeScript encoding of the current state.
+consider branching (receiving) and selection (sending) states separately.
+We assign TypeScript encodings of states to their state identifiers in the
+EFSM, providing syntactic sugar when referring to the successor state when
+encoding the current state.
 For any state $S$ in the EFSM, we refer to the TypeScript type alias of its
-encoding as $\llbracket S \rrbracket$. We outline the encoding below using examples from the
+encoding as $\llbracket S \rrbracket$.
+We outline the encoding below using examples from the
 \textit{Noughts and Crosses} game (\cref{lst:svr}).
@@
 \paragraph{Branching State}
-We consider a receive state as a unary branching state for conciseness.
+We consider a receiving state as a unary branching state for conciseness.
 A branching state is encoded as an \textit{object literal}
-\cite{TypeScriptSpec}, with each branch $i \in I$, $I$ denoting set of all
-branches, corresponding to a member field.
+\cite{TypeScriptSpec} (a record type), with each branch $i \in I$ ($I$ denoting set of all
+branches), corresponding to a member field.
 A branch expecting to receive a message labelled $\texttt{label}_i$ carrying
@@
 \paragraph{Selection State}
-We consider a send state as a unary selection state for conciseness.
+We consider a sending state as a unary selection state for conciseness.
 A selection state is encoded as a \textit{union type}
@@
 In the case of \cref{lst:svr}, the developer is expected to implement
-\texttt{S13} which handles the \texttt{Pos} message sent by P1,
-and the code in \texttt{S13} returns a value of type \texttt{S15}, which corresponds to a selection of messages to send to P2.
-\cref{lst:svrprotocol} illustrates how the developer may implement these types.
+\texttt{S13} which handles the \texttt{Pos} message sent by \texttt{P1},
+and the code in \texttt{S13} returns a value of type \texttt{S15}, which
+corresponds to a selection of messages to send to \texttt{P2}. \Cref{lst:svrprotocol}
+illustrates how the developer may implement these types.
 
 We make a key design decision \textit{not} to expose communication channels in
-the TypeScript session type encoding to provide \textit{static} linearity
+the TypeScript session type encodings to provide \textit{static} linearity
 guarantees (\cref{section:serverlinear}).
 Our encoding sufficiently exposes seams for the developer to inject their
-business logic, whilst the generated session API
+program logic, whilst the generated session API
 (\cref{section:serversessionapi}) handles the sending and receiving of
@@
 The runtime executes the EFSM by keeping track of
-the current state (similar to the generated code in \cite{javatypestate}) and only permitting the allowed IO actions at the current state.
-The runtime listens to message (receive) events on the communication channel,
+the current state (similar to the generated code in \cite{javatypestate})
+and only permitting the specified IO actions at the current state.
+The runtime listens to message (receiving) events on the communication channel,
 invokes the corresponding callback to obtain the value to send next, and
-performs the send.
-The developer instantiates the session by constructing an instance of the
+performs the sending.
+The developer instantiates a session by constructing an instance of the
 session runtime class, providing the WebSocket endpoint URL (to open the
-connection) and initial state (to execute the EFSM).
+connection) and the initial state (to execute the EFSM).
 
@@
 \label{section:serverlinear}
+\fz{This sentence is confusing and I don't know how to fix.}
 Channel linearity is guaranteed based on the correctness of our library design
@@
 \paragraph{Repeated Usage}
-We do not expose channels to the developer, which makes \textit{reuse}
-impossible.
+We do not expose channels to the developer, which makes \textit{reusing
+  channels} impossible.
 For example, to send a message, the generated API only requires the payload
@@
 The initial state must be supplied to the session runtime
-constructor in order to instantiate the session;
+constructor in order to instantiate a session;
 this initial state is defined
@@
 successors and so forth.
+\fz{It's not the approach, rather than the developer implementation?}
 This encoding approach covers the terminal state (if it exists), and the
@@
 \subsection{The React Framework}
-Our browser-side session type encodings for browser-side targets build upon the React framework, developed by Facebook \cite{React} for the \textit{Model-View-Controller} (MVC) architecture.
+Our browser-side session type encodings for browser-side targets build upon the
+\emph{React.js} framework, developed by Facebook \cite{React} for the
+\textit{Model-View-Controller} (MVC) architecture.
 React is widely used in industry to create scalable single-page TypeScript
@@
 view function in the MVU architecture.
-Components can keep \textit{state} and the \texttt{render()} function is
+Components can keep \textit{state}s and the \texttt{render()} function is
 invoked upon a change of state.
-A simple counter can be implemented as a component,
-with \texttt{count} stored as state, a button which increments \texttt{count}
+
+For example, a simple counter can be implemented as a component,
+with its \texttt{count} stored as state.
+When rendered, it displays a button which increments \texttt{count}
 when clicked and a \texttt{div} that renders the current
 \texttt{count}.
+If the button is clicked, the \texttt{count} is incremented, which triggers a
+re-rendering (since the state has changed), and the updated \texttt{count} is
+displayed.
+
 Components can also render other components, which gives rise

### 2020-03-09 PLACES_2020/encoding.tex
PLACES: remove VDOM paragraph

@@
 
-\fz{Is VDOM needed to be covered here?}
-
-\am{Intended to motivate how the session runtime implementation renders the required state and the VDOM deals with the diffs - at the same time, unsure whether this is important for the reader to understand (or just leave the DOM rendering as a black box)?}
-\paragraph{Virtual DOM (VDOM)}
-React components are rendered on a logical
-abstraction of the DOM, which in turn performs a \texttt{diff} on the current
-browser DOM and patches the delta accordingly.
-This allows the developer to
-directly specify what should be rendered without having to worry about which
-elements to append or remove from the browser DOM.
-
 \subsection{Browser-Side API Generation}
@@
 An MVU application features a \textit{model} encapsulating application
-state, a \textit{view function} rendering the state on the DOM, and an
+state, a \textit{view function} rendering the state on the Document Object Model (DOM), and an
 \textit{update function} handling \textit{messages} produced by the

### 2020-03-09 PLACES_2020/related.tex
PLACES: address reviewers/FZ comments
PLACES: update case study code listing
PLACES: rearrange first sentence under 'Related Work'

@@
 \section{Related Work}
-Native language support for first-class linear channel resources \cite{ATS} and
-code generation are two main approaches for incorporating MPST into application
-development.
+The two main approaches for incorporating our MPST workflow into application
+development are native language support for first-class linear channel resources \cite{ATS} and code generation.
 The latter closely relates to our proposal;
@@
 aliasing is not monitored.
+% REVIEW: could the approach of the present paper by applied to Java as well? discuss?
+Scribble-Java can generate callback-style APIs similar to the approach we 
+present, but this approach is arguably less idiomatic for Java developers.
 

### 2020-03-09 PLACES_2020/encoding.tex
PLACES: address FZ comments

@@
 \fz{This sentence is confusing and I don't know how to fix.}
-Channel linearity is guaranteed based on the correctness of our library design
-and session runtime implementation, which means a faulty implementation could
-violate this, but is up to the library author to verify once rather than the
-end user. Our library design prevents the two conditions detailed below:
+
+Developers writing their implementation using the generated APIs 
+enjoy channel linearity by construction.
+Our library design prevents the two conditions detailed below:
 
@@
 \fz{It's not the approach, rather than the developer implementation?}
-This encoding approach covers the terminal state (if it exists), and the
-session runtime guarantees this terminal state, if it exists, will be reached
+
+The developer's implementation will cover the terminal state
+(if it exists), and the
+session runtime guarantees this terminal state will be reached
 by construction.

### 2020-03-09 PLACES_2020/encoding.tex
PLACES: address FZ comments
PLACES: remove conversation thread

@@
 \label{section:serverlinear}
-\fz{This sentence is confusing and I don't know how to fix.}
-
 Developers writing their implementation using the generated APIs 
@@
 successors and so forth.
-\fz{It's not the approach, rather than the developer implementation?}
-
 The developer's implementation will cover the terminal state

### 2020-03-09 PLACES_2020/encoding.tex
Save some space

@@
 We refer to the \texttt{Svr} EFSM (\cref{fig:efsmsvr}) as a running example in
-this section.
+this subsection.
 For server-side targets, we encode EFSM states into TypeScript types and
@@
 This enables reusability, and for our use case, gives control to the parent
-on what data to pass to its children (e.g. pass the payload of a received
+on what data to pass to its children (e.g.\ pass the payload of a received
 message to a child to render).
@@
   \label{fig:efsmp1}
+  \vspace{-5mm}
 \end{wrapfigure}
@@
 We refer to the \texttt{P1} EFSM (\cref{fig:efsmp1}) as a running example in
-this section.
+this subsection.
 Preserving behavioural typing and channel linearity is challenging
@@
 channel linearity.
-Our design goal is to enforce this statically through the generated APIs.
+Our design goal is to enforce linearity statically through the generated APIs.
 
@@
 components: a \emph{model type} uniquely defines a \textit{view function},
-set of \textit{messages} and \textit{update function} -- rather than
+set of \textit{messages} and \textit{update function} --- rather than
 producing a new model, the update function defines valid transitions to
@@
 
-\paragraph{Model transitions}
+\paragraph{Model Transitions}
 Transitions are encoded as React component props onto the encoded states by the
@@
 We make the assumption that message sending is triggered by
-some user-driven UI event (e.g. clicking a button, pressing a key on the
-keyboard) which interacts with some DOM element.
+some user-driven UI event (e.g.\ clicking a button, pressing a key on the
+keyboard) which interacts with some DOM elements.
 We could pass a
@@
 role identifier:
-it opens the WebSocket connection to the server, keeps track of the current
+it opens a WebSocket connection to the server, keeps track of the current
 EFSM state as part of its React component state, and most importantly, renders
@@
 The developer writes their own implementation of each state (mainly to
-customise how the state is rendered and inject business logic into state
+customise how the state is rendered and inject program logic into state
 transitions) by extending the abstract React class components.
@@
 \subsubsection{Affine Channel Usage}
-A limitation of our browser-side session type encoding is only being able to
-guarantee that channel resources are used \textit{at most once} as opposed to
+A limitation of our browser-side session type encoding is
+that channel resources are guaranteed to be used \textit{at most once} as opposed to
 \textit{exactly once}.
@@
 transitions to a terminal state but still type-check, so we cannot prevent
-unused states. Equally, our approach does not prevent a client closing the browser, which would drop the connection.
+unused states.
+Equally, our approach does not prevent a client closing the browser, which
+would drop the connection.

### 2020-03-09 PLACES_2020/conclusion.tex
PLACES: modify review feedback
PLACES: address reviewers/FZ comments
typo in conclusion

@@
 
-Future works include incorporating \textit{explicit connection actions}
+Future work includes incorporating \textit{explicit connection actions}
 introduced in \cite{ExplicitConnections} in our API generation to better model
@@
 applications.
+% REVIEW: asynchronous
+Server-side implementations may perform asynchronous operations on the
+received messages, so supporting asynchronous values (such as JavaScript
+\textit{Promises} \cite{promise}) in IO actions would be a welcome addition.
 Whilst our approach supports multiparty sessions, the nature of
@@
 % - WebRTC for full MPST (since current limitation is w.r.t. MPST protocols with 1-server many-client roles
+
+% - support asynchronous operations???

### 2020-03-09 PLACES_2020/casestudy.tex
PLACES: proofread case study

@@
 
-Note that, by TypeScript's structural typing \cite{TypeScriptSpec}, replacing
-\texttt{handleP2Move} on \cref{lineupdate} with a recursion of \texttt{handleP1Move} would be
-type-correct -- this allows for better code reuse as opposed to defining
-additional abstractions to work around the limitations of nominal typing in
+Note that, by TypeScript's structural typing 
+\cite{TypeScriptSpec}, replacing
+\texttt{handleP2Move} on \cref{lineupdate} with a recursive 
+occurrence of \texttt{handleP1Move} would be
+type-correct -- this allows for better code reuse as opposed to 
+defining
+additional abstractions to work around the limitations of nominal 
+typing in
 \cite{Hybrid2016}.
-There is also full type erasure when transpiling to JavaScript to run the
-server code, so state space explosion is not a runtime consideration.
+There is also full type erasure when transpiling to JavaScript to 
+run the server code, so state space explosion is not a 
+runtime consideration.
 

### 2020-03-09 PLACES_2020/casestudy.tex
PLACES: proofread case study

@@
 
-For the sake of code reuse for demonstration purposes, \cite{NoughtsAndCrosses}
+For the sake of code reuse, \cite{NoughtsAndCrosses}
 uses \textit{higher-order components} (HOC) to build the correct state

### 2020-03-09 PLACES_2020/casestudy.tex
More proofreading

@@
 occurrence of \texttt{handleP1Move} would be
-type-correct -- this allows for better code reuse as opposed to 
+type-correct --- this allows for better code reuse as opposed to
 defining
@@
 There is also full type erasure when transpiling to JavaScript to
-run the server code, so state space explosion is not a 
-runtime consideration.
+run the
+server code, so the types defined in TypeScript will not appear in the
+JavaScript after typechecking.
+This means state space explosion is not a runtime consideration.
 
@@
 implementations depending on which player the user chooses to be.
-We leverage the \textit{Redux} \cite{Redux} state management library to keep track of game
-state, thus showing the flexibility of our library design in being
-interoperable with other libraries and idiomatic JavaScript practices.
+We leverage the \textit{Redux} \cite{Redux} state management library to keep
+track of the game state, thus showing the flexibility of our library design in
+being interoperable with other libraries and idiomatic JavaScript practices.
 % REVIEW: tight-coupling?
 Our approach encourages the separation of concerns between the
-communication logic and program logic -- the generated session runtime
+communication logic and program logic --- the generated session runtime
 keeps track of the state of the EFSM to ensure protocol conformance by

### 2020-03-09 PLACES_2020/conclusion.tex
More proofreading

@@
 We have presented a MPST-based framework for developing full-stack interactive
-TypeScript applications with WebSocket communications that conform to a
+TypeScript applications with WebSocket communications.
+The implementation conforms to a specified
 protocol, statically providing linear channel usage guarantees and affine

### 2020-03-09 PLACES_2020/related.tex
More proofreading

@@
 The two main approaches for incorporating our MPST workflow into application
-development are native language support for first-class linear channel resources \cite{ATS} and code generation.
+development are native language support for first-class linear channel
+resources \cite{ATS} and code generation.
 The latter closely relates to our proposal;
@@
 Whilst the same approach could be applied to JavaScript, we can provide more
-static guarantees with TypeScript's gradual typing system and compiler. Scribble-Java \cite{Hybrid2016} proposed to encode the EFSM
+static guarantees with TypeScript's gradual typing system.
+Scribble-Java \cite{Hybrid2016} proposed to encode the EFSM
 states and transitions as classes and instance methods respectively, with
@@
 \paragraph{Session Types in Web Development}
-King et al. \cite{PureScript2019} targeted web development in PureScript using the
+King et al.\ \cite{PureScript2019} targeted web development in PureScript using the
 \textit{Concur UI} framework and proposed a type-level encoding of EFSMs as
@@
 providing an intuitive development experience to developers, especially given
-the prevalence of JavaScript and TypeScript applications in industry. Fowler \cite{MVU2019} focused on applying binary session types in front-end web
+the prevalence of JavaScript and TypeScript applications in industry.
+Fowler \cite{MVU2019} focused on applying binary session types in front-end web
 development and presented approaches that tackle the challenge of guaranteeing
-linearity in the event-driven environment.
+linearity in the event-driven environment, whereas our work is applicable to
+multiparty scenarios.
 

### 2020-03-09 PLACES_2020/encoding.tex
PLACES: update receive transition explanation

--- 2020-03-09 PLACES_2020/encoding.tex
@@
 \subparagraph{Receiving}
-The payload of a message receive transition is passed as
-a prop to the successor state and can be handled in the successor's view
-function.
-Upon a message receive event, the session runtime renders the
-successor of the receive state and propagates the payload to the successor
-accordingly.
+The React component for a receiving state is required to
+define a handler for each supported branch. 
+Upon a message receive event, the session runtime invokes the
+handler of the corresponding branch with the message payload and 
+renders the successor state upon completion.
 

### 2020-03-09 PLACES_2020/encoding.tex
Consistency

@@
 \cite{TypeScriptSpec} of internal choice encodings: each internal choice $i \in
-I$, $I$ denoting set of all choices, sending a message labelled
+I$ ($I$ denoting set of all choices), sending a message labelled
 $\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state

### 2020-03-09 PLACES_2020/casestudy.tex
Consistency

@@
 server code, so the types defined in TypeScript will not appear in the
-JavaScript after typechecking.
+JavaScript after type-checking.
 This means state space explosion is not a runtime consideration.

### 2020-03-12 PLACES_2020/abstract.tex
typo

@@
 We demonstrate the effectiveness of our approach through a web-based
-implementation of the classic \textit{Noughts and Crosses} game from a MPST
+implementation of the classic \textit{Noughts and Crosses} game from an MPST
 formalism of the game logic.

### 2020-03-12 PLACES_2020/conclusion.tex
typo

@@
 \section{Conclusion and Future Work}
-We have presented a MPST-based framework for developing full-stack interactive
+We have presented an MPST-based framework for developing full-stack interactive
 TypeScript applications with WebSocket communications.

### 2020-03-12 PLACES_2020/encoding.tex
PLACES: fix warnings for figure placement and overfull
PLACES: minor updates

@@
 \cref{lst:clientapp}.
-\texttt{this.props.Pos} is the factory function prop
-passed from the session runtime.
+The session runtime passes the factory function \texttt{this.props.Pos} as a prop.
 For each x-y coordinate on the game board, we

### 2020-04-01 report/sections/acknowledgements.tex
Report: init TeX

\begin{abstract}
These are some acknowledgements.
\end{abstract}

### 2020-04-02 notes/notes.tex
Notes: init TeX

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{listings}

\title{Project Notes}
\author{Anson Miu}

\begin{document}

\nocite{*}

\maketitle

\section{Progress}

\paragraph{01/04/2020}
\begin{itemize}
\item 
Set up Node and TypeScript sandbox for testing.
The idea is to have a workflow where the
tool generates code for some protocol,
then copies into the sandbox and runs
the TypeScript compiler with the predefined
\texttt{tsconfig.json}.

\item
This guarantees that ``correct'' TypeScript is
generated with respect to the TypeScript 
language, not about correctness with respect
to session.
\end{itemize}

\paragraph{02/04/2020}
\begin{itemize}
\item
Set up React TypeScript sandbox for testing.
No need \texttt{create-react-app} at
the moment (which creates full application
boilerplate) as we are just trying to check
for TypeScript syntax.

\item
Piped the \texttt{tsfmt} into the code 
generation workflow so no need to deal with
spacing issues on Jinja template -- timesaver.

\item
The full code generation workflow ``works'' for 
both targets. The validity of the generated 
code works only for Node targets.

\item
Now focusing on formalising the React generated 
code -- taking more time than planned because
the TicTacToe example made more assumptions
about the protocol than I thought
\end{itemize}

\section{MPST Encoding}

\section{Code generation workflow}

\section{Testing}

\subsection{Unit tests}
TODO -- typical, use \texttt{unittest} module
to check for functionality of EFSM parsing
etc.

\subsection{Integration tests}
For a given protocol and specified role/target,
generate the TypeScript code, put into the
sandbox environment and run \texttt{npm test} 
(which is just a wrapper for 
\texttt{tsc --noEmit})
to expect success return code.

\section{To fix later (most important first)}

\subsection*{Receive transitions in React}
This is the current workflow -- need to verify whether the receive transition is safe?

\begin{enumerate}
\item
Assume you are on send state, user triggers
event to perform send.
\item
Runtime performs the send over the WebSocket
and renders the receive state \textbf{by
calling \texttt{setState}}. Calling setState 
schedules the state transition, just like
how a Promise resolve scheduels the call in
the high-priority callback queue.

\item When the receive state is rendered (as a 
result of \texttt{setState}), the lifecycle
method \texttt{componentDidMount} will call \texttt{register} to add the receive handlers

\item \textbf{At the same time}, message is sent from
other role to this user -- this triggers the 
onMessage event listener on the WebSocket. 
\textbf{There is no guarantee that \texttt{setState} 
is called before the message event listener.}

\end{enumerate}

If the onMessage event listener
is called before the setState,
then the implementation breaks because there is no
handler to use.

\subsubsection*{Possible solution}

Use the same approach as the backend.
\textit{Ignoring the exact TypeScript types for now,}
keep a message queue and a handler queue.

On mounting the receive state (which is the actual
transition), check whether there is a message in the 
queue.
\begin{itemize}
\item
If yes -- pop message and call handler based on label
(and cast payload accordingly).

\item
If no -- add handler to queue.
\end{itemize}

On message receive, check whether there are handlers
in the queue. 
\begin{itemize}
\item
If yes -- call the correct one based on
the label (and cast payload accordingly).

\item
If no -- add message to queue.
\end{itemize}

\begin{itemize}
\item
Do we need a queue? Is it possible for more than
one message to be queued up, since you advance to
a receive state and only expect to receive exactly
one message?

\item
Could use an optional-type variable
but need to consider the types...
\end{itemize}

\subsection*{Send transition in React}
Sends can only be event-driven --
currently not possible to do this:

\begin{enumerate}
\item Server sends browser 
\texttt{Offer(int)}
\item Browser immediately sends 
\texttt{Reject()} if payload is less than 0;
otherwise attach a \texttt{Counter(int)} 
send action to a user event.
\end{enumerate}

\section{Extensions}

\subsection*{Explicit connection actions}

See \cite{FASE2017} -- also refer to Battleships 
example from Jonathan.

\begin{lstlisting}
explicit global protocol Game(role Svr, role P1, role P2) {
  Play() connect P1 to Svr;
  Play() connect P2 to Svr;
  do OneGame(Svr, P1, P2);
  disconnect P1 and Svr;
  disconnect P2 and Svr;
 }

global protocol OneGame(role Svr, role P1, role P2)
  Pos(Point) from P1 to Svr;
  choice at Svr {
    Lose(Point) from Svr to P2;
    Win(Point) from Svr to P1;
  } or {
    Draw(Point) from Svr to P2;
    Draw(Point) from Svr to P1;
  } or {
    Update(Point) from Svr to P2;
    Update(Point) from Svr to P1;
    do OneGame(Svr, P2, P1);
  }
}

\end{lstlisting}

\bibliographystyle{acm}
\bibliography{notes}

\end{document}

### 2020-04-02 notes/notes.tex
Notes: update 02/04/2020

@@
 \begin{document}
-
-\nocite{*}
 
@@
 
+\subsection{Node}
+
+TODO
+
+\subsection{Browser}
+
+TODO
+
 \section{Code generation workflow}
@@
 to expect success return code.
+
+\texttt{npm test} is already set up -- just need to
+set up the Python script to call for testing.
 
@@
 queue.
-\begin{itemize}
-\item
 If yes -- pop message and call handler based on label
 (and cast payload accordingly).
-
-\item
 If no -- add handler to queue.
-\end{itemize}
 
@@
 in the queue. 
-\begin{itemize}
-\item
 If yes -- call the correct one based on
 the label (and cast payload accordingly).
-
-\item
 If no -- add message to queue.
-\end{itemize}
 
@@
 \end{itemize}
+
+\subsection*{Exception handling}
+
+TODO read Exceptional GV paper \cite{ExceptionalGV}.
 
@@
 
+\subsection*{Typestate in TypeScript}
+
+Look into TyoeScript transfomers -- the idea is
+to define typestate decorators, then augment the
+TypeScript compiler to transform the code into some
+intermediate layer to perform validation on
+typestate: for example, consider the example of a 
+stack that can either be full or empty.
+
+\section{Report Structure}
+
+\begin{enumerate}
+\item 
+Introduction -- as per PLACES (for now)
+
+\item
+Background -- process calculus, BST, MPST, code 
+generation, TypeScript; think about where React goes?
+
+\item
+MPST encoding in TypeScript (section 2 in these notes).
+Split into Node and browser targets, also talk about
+the different ideas building up to this (pros and
+cons of the experiments I did with Calculator in
+2019).
+
+Any of the new concepts (exception handling, etc.) will 
+go in this section too.
+
+TODO think about how to split into 
+more logical sections.
+
+\item
+Code generation workflow (section 3 in these notes)
+-- software architecture of
+toolchain implemented in Python.
+
+\end{enumerate}
+
+\nocite{*}
 \bibliographystyle{acm}

### 2020-04-03 notes/notes.tex
Notes: update 03/04/2020

@@
 the TicTacToe example made more assumptions
-about the protocol than I thought
+about the protocol than I thought.
+\end{itemize}
+
+\paragraph{03/04/2020}
+\begin{itemize}
+\item
+Build system testing pipeline -- register which
+Scribble protocols to test, specify which role
+(e.g. Bob) corresponds with which target (e.g. Node),
+then the pipeline will perform the code generation,
+check for success exit code, then invoke the
+TypeScript compiler to check the TypeScript code
+and also check for success exit code.
 \end{itemize}

### 2020-04-03 notes/notes.tex
Notes: update 03/04/2020

@@
 
+\paragraph{04/04/2020 [tomorrow]}
+\begin{itemize}
+\item
+Focus on getting React code generation to work -- 
+TicTacToe example for the receive handlers weren't
+generalised enough, go back to OneNote to sketch
+the encoding before trying it in code.
+\end{itemize}
+
 \section{MPST Encoding}
@@
 
-TODO
+TODO - on OneNote
 
@@
 
-TODO
+TODO - on OneNote
 
 \section{Code generation workflow}
+
+TODO - on OneNote
 
@@
 \subsection{Integration tests}
+TODO -- integrate with external components, namely
+scribble-java (which would be mocked in unit tests).
+
+\subsection{Systems tests}
 For a given protocol and specified role/target,

### 2020-04-05 notes/notes.tex
Notes: update 05/04/2020

@@
 
-\paragraph{04/04/2020 [tomorrow]}
+\paragraph{05/04/2020}
 \begin{itemize}
 \item
-Focus on getting React code generation to work -- 
-TicTacToe example for the receive handlers weren't
-generalised enough, go back to OneNote to sketch
-the encoding before trying it in code.
+Implemented templates for React code generation.
+The main breakthrough was to better leverage
+the type system and reduce duplication in the
+generated code. The TicTacToe example had a lot
+of duplication by hardcoding the state machine.
+The new template pushes the logic to the actual
+states and makes the runtime more lightweight.
+
+\item
+The generated code for BinaryNoPayload.scr is functional for both Node and browser targets.
 \end{itemize}

### 2020-04-06 notes/notes.tex
Notes: update 06/04/2020

@@
 The generated code for BinaryNoPayload.scr is functional for both Node and browser targets.
+\end{itemize}
+
+\paragraph{06/04/2020}
+\begin{itemize}
+\item
+Modified Node.js code generation templates to
+take inspiration from the browser ones -- namely,
+to make the runtime more lightweight by pushing
+logic to the individual states; again, this means
+no more huge switch statement with duplication.
+
+\item
+New structure for Node.js generated code generally
+works. Explicitly used partial application to
+focus the types (e.g. State $\to$ ReceiveState $\to$
+ReceiveState.S12), and define each non-terminal
+state as a separate module to internally use
+its known message/label types to parse messages
+without having one large switch statement in the
+runtime.
+
+\item
+Taking cues from Express middleware (with
+the \texttt{next} function) to allow the
+runtime to pass the send and receive functions
+into the individual states, and give a callback
+function for the individual states to pass any
+data back to the runtime.
+
+\item
+An example of this is for receive states -- 
+the receive state needs to construct the message 
+receive handler \textbf{tailored to the types/labels of 
+that particular state}. This message receive handler
+will do the parsing, invoke the correct receive
+function, and advance the EFSM to the correct
+successor state (achieved by having the runtime
+pass the state-advancing function as param).
+
+\item
+Problem -- still need a type cast for the time being
+with the handler functions; not sure why the types
+aren't specific enough , trying to make sense of the 
+error message at the moment.
 \end{itemize}

### 2020-04-09 notes/notes.tex
Notes: update 07/04/2020
Notes: update 09/04/2020

@@
 
+\paragraph{07/04/2020}
+\begin{itemize}
+\item
+Address ``technical debt'' in EFSM API -- it has
+been inconsistent through being adapted alongside
+the Node/React codegen template generation. 
+Make use of dunder methods and properties
+throughout to remove some of the internal logic
+required in the templates.
+
+\end{itemize}
+
+\paragraph{08/04/2020}
+\begin{itemize}
+\item
+Different MPST TypeSrcript encoding for Node targets --
+we consider the previous encodings (labelled callbacks for
+receive states, union types for send states) as ``handlers'',
+and the developer needs to wrap them in an ``implementation''
+class instance.
+
+\item
+The aim is for the runtime to be as lightweight as possible
+-- it only needs to concern with whether the new state
+is sending, receiving or terminal; and the state implementation
+deals with performing the corresponding channel action
+based on the particular label/payload types of that state.
+
+\item
+The class-based implementation/handler distinction allows
+the runtime to use \texttt{instanceof} to safely narrow the
+type of implementation. Each type of state essentially acts as
+an interface (but we use classes in TypeScript to take advantage
+of the \texttt{instanceof} function -- the equivalent notion
+of a Java interface can be expressed as a TypeScript abstract
+class with abstract methods and no properties defined). Because
+all receive states will expose the same method signature,
+everything type-checks.
+
+\end{itemize}
+
+\paragraph{09/04/2020}
+\begin{itemize}
+\item
+Fix label regex problem in EFSM parsing -- currently just
+allow alphabets, should allow more (i.e. alphanumeric
++ ``typical symbols'' like underscores and dashes).
+
+\item
+Could also look up source code of Scribble-Java to see what they use.
+\end{itemize}
+
 \section{MPST Encoding}
@@
 etc.
+
+\begin{itemize}
+\item EFSM parsing -- regexes, \dots
+\item EFSM API
+\end{itemize}
 

### 2020-04-10 notes/notes.tex
Notes: start define formalism for Node send/receive

@@
 \documentclass{article}
+\usepackage[margin=1in]{geometry}
 \usepackage[utf8]{inputenc}
 \usepackage{listings}
+\usepackage{amsmath}
 
@@
 Could also look up source code of Scribble-Java to see what they use.
+
+\item
+Fix message payload parsing in EFSM to correctly strip and filter out whitespace.
+
+\item
+Adapt Node/React templates for message payload types -- the Calculator example
+(binary session with recursion, selection and branching) now produces valid TypeScript
+code for both browser and client.
+
 \end{itemize}
@@
 
-TODO - on OneNote
+\subsubsection{Encoding}
+We distinguish between \textbf{handler} (the PLACES encoding which handle the
+message sending and receiving) and \textbf{implementation} (the wrapper around the handler).
+
+\paragraph{Send / Selection}
+Assume state S1 can select from a list of [Label, Payload, Successor] such that Payload and Successor
+are dependent on Label.
+
+\begin{align}
+encoding_{Send}(S) &= handler_{Send}(S), impl_{Send}(S) \\
+handler_{Send}(S) &= \bigcup_{i \in labels(S)} [Labels_i,~Payload_i,~Impl(Succ_i)] \\
+impl_{Send}(S) &= sendImpl(Handler(S),~Message(S))
+\end{align}
+
+\texttt{sendImpl} defines a subclass of \texttt{Implementation.Send} based on the handler data type
+and message data type (the union of possible messages). It implements the abstract method
+\texttt{performSend} based on these parameters, which (1) sends the message using
+the method (or channel passed in by runtime), then (2) advances to next state (again, using
+function passed in by runtime which acts as a ``send``).
+
+\paragraph{Receive / Branching}
+
+\begin{align}
+encoding_{Recv}(S) &= handler_{Recv}(S), impl_{Recv}(S) \\
+handler_{Recv}(S) &= Map_{i \in labels(S)}\left<Labels_i, Function\left<Payload_i, Impl(Succ_i)\right>\right> \\
+impl_{Recv}(S) &= recvImpl(Handler(S), Message(S))
+\end{align}
+
+\texttt{recvImpl} defines a subclass of \texttt{Implementation.Recv} based on the handler data type
+and message data type (the union of possible messages). It implements the abstract method
+\texttt{performRecv} based on these parameters, which (1) prepares the receive handler (a function that takes
+(message: any) => void), then (2) registers this handler to the runtime using the passed-in callback
+(which acts as sending the handler to the runtime), and (3) when said handler in (1) gets called by the runtime
+(essentially receiving from the runtime), actually performs the receive and (4) advances to next state.
+
+\subsubsection{Runtime as session type}
+Loosely define the protocol for message passing between the runtime and the
+corresponding state implementation.
+
+\begin{lstlisting}[breaklines=true, tabsize=4, frame=single]
+protocol Next(role Runtime, role State) {
+	// not really a `choice`, more of `based on the type of State`\dots
+	choice at Runtime {
+		// non-terminal
+		choice at Runtime {
+			// send
+			PerformSend(???) from Runtime to State;
+			Send(role: string, label: string, payload: any[]) from State to Runtime;
+
+			// runtime sends message through WebSocket
+
+			// `Successor` is the state transition function in the literature
+			do Next(Runtime, Successor<State, label>);
+
+		} or {
+			// receive
+			PrepareReceive(???) from Runtime to State;
+			do Receive(Runtime, State);
+		}
+
+	} or {
+		// terminal
+	}
+}
+
+protocol Receive(role Runtime, role State) {
+
+	/*
+	choice {
+		Message(label: string, payload: any[]) from ??? to Runtime;
+		Register(???) from State to Runtime;
+		PerfomReceive(???) from Runtime to State;
+		do Next(Runtime, Successor<State, label>);
+	} or {
+		Register(???) from State to Runtime;
+		Message(label: string, payload: any[]) from ??? to Runtime;
+		PerformReceive(???) from Runtime to State;
+		do Next(Runtime, Successor<State, label>); 
+	}
+	*/
+
+	// State sends a Channel to Runtime
+	Register(performReceive: ???) from State to Runtime;
+
+	// whenever the message arrives -- regardless whether it is before or after
+	performReceive({ label: string, payload: any[] }) from Runtime to State;
+
+	do Next(Runtime, Successor<State, label>);
+}
+\end{lstlisting}
 

### 2020-04-10 notes/notes.tex
Notes: update 10/04/2020

@@
 
+\paragraph{10/04/2020}
+\begin{itemize}
+\item
+Start formalising the encoding for states and runtime for Node targets.
+
+\end{itemize}
+
 \section{MPST Encoding}
@@
 			// send
-			PerformSend(???) from Runtime to State;
+			PerformSend(send: ???, next: ???) from Runtime to State;
 			Send(role: string, label: string, payload: any[]) from State to Runtime;
@@
 			// receive
-			PrepareReceive(???) from Runtime to State;
+			PrepareReceive(register: ???, next: ???) from Runtime to State;
 			do Receive(Runtime, State);

### 2020-04-17 notes/notes.tex
Notes: update 13/04/2020
Notes: update ≤17/04/2020

@@
 Start formalising the encoding for states and runtime for Node targets.
+\end{itemize}
 
+\paragraph{13/04/2020}
+\begin{itemize}
+\item
+Fix type signature for React event handler -- extract the 
+camelCase keys from \texttt{React.DOMAttributes} and infer
+the type of event (e.g. MouseEvent from `onClick') to correctly
+type the event handler.
+\end{itemize}
+
+\paragraph{16/04/2020}
+\begin{itemize}
+\item
+Experiment with what code should be generated
+for explicit cancellation using the Calculator example.
+
+\item
+Verify (through experimenting in code) that affine channel
+usage cannot be violated by introducing long for-loops
+in backend to simulate delay in Calculator example. 
+This will be better addressed by supporting Promises
+in the future.
+
+\end{itemize}
+
+\paragraph{17/04/2020}
+\begin{itemize}
+\item
+Exploring the theory for reasoning about global types
+specific for web services generation.
 \end{itemize}

### 2020-04-18 notes/notes.tex
Notes: update 18/04/2020

@@
 specific for web services generation.
+\end{itemize}
+
+\paragraph{18/04/2020}
+\begin{itemize}
+\item
+Implement utility tool to: (1) parse existing type declarations
+from Scribble file, (2) extract user custom types,
+(3) insert missing TypeScript primitive types, before proceeding
+with normal workflow.
+
+\item
+Adapt EFSM API -- previously, I use a metadata dictionary
+to store the protocol name and role name. Now that I need to
+propagate the custom type information too, I have a new 
+\texttt{Endpoint} API which holds the EFSM (without metadata)
+and the protocol/role name is lifted as a property of the 
+endpoint API. Jinja templates have been adapted with the
+API change.
 \end{itemize}

### 2020-04-19 notes/notes.tex
Notes: update 18/04/2020
Notes: ideas on MPST web service formalism for client-client interaction

@@
 \usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{wasysym}
+\usepackage{stmaryrd}
 
@@
 API change.
+
+\item
+Update Jinja templates to render custom type import
+statements in TypeScript.
+\end{itemize}
+
+\paragraph{19/04/2020}
+\begin{itemize}
+\item
+Draft theory of MPST for webservices over websockets 
+(on OneNote).
 \end{itemize}
@@
 
+\newpage
+
+\section{MPST for Web Services}
+
+\paragraph{Main Idea}
+Protocols for webservices implemented over WebSockets
+present additional constraints.
+
+\begin{enumerate}
+\item Exactly one server-side role.
+\item Client-side roles only \textit{directly} communicate 
+to the server-side role.
+\end{enumerate}
+
+\subsection*{Global types for webservices}
+
+Recall $G$, the global types from multiparty session types:
+
+\begin{align*}
+G ::&= \texttt{end} \\
+&\mid p \to q : \{ l_i: G_i \}_{i \in I} \\
+&\mid \mu t. G \\
+&\mid t
+\end{align*}
+
+We consider $G_{WS}$, global types for webservices over 
+websockets, defined as a tuple of the original global type $G$, 
+and a participant $p$ explicitly specified as the server role;
+all other participants $q \in pt(G)$ are assumed to take
+on browser roles:
+\begin{align*}
+G_{WS} ::= G \times p
+\end{align*}
+
+We also use the same process calculus defined before:
+
+\begin{align*}
+P ::&= 0 \mid \mu X. P \mid X \\
+&\mid p \rhd \{ l_i: P_i \}_{i \in I} \\
+&\mid p \lhd l. P \\
+&\mid \texttt{if e then } P \texttt{ else } P\\\\
+\mathcal{M} ::&= p :: P \\
+&\mid \mathcal{M} \mid \mathcal{M}
+\end{align*}
+
+The typing judgement for a multiparty session with respect
+to the webservice global type builds upon the \textsc{[MTy]}
+rule.
+
+\subsection*{Codegen}
+\texttt{codegen :: $\mathcal{M}$ $\times$ Role 
+$\to$ $\mathcal{M}$}
+
+\subsubsection*{Main rules}
+
+\begin{align*}
+&codegen~(p :: q \lhd l_j. P 
+\mid q :: p \rhd \{ l_i: Q_i \}_{i \in I} 
+\mid s :: S 
+\mid \mathcal{M})~s \\
+=~&codegen~(p :: s \lhd l. P 
+\mid q :: s \rhd \{l_i: Q_i\}_{i \in I}
+\mid s :: p \rhd )\\
+&where  \\
+\dots
+\end{align*}
+
+\subsubsection*{Trivial rules}
+
+\subsection*{Encoding}
+
+\subsection*{Preservation}
+Modifying the process through code generation preserves
+well-typedness of the session.
+
+\paragraph{To show:} let $G_{WS} = (G, p)$. 
+If $\mathcal{M}$ is well-typed with respect to $G_{WS}$,
+then \texttt{codegen($\mathcal{M},~p$)} is well-typed
+with respect to $\llbracket G_{WS} \rrbracket$
+
+\newpage
+
 \section{To fix later (most important first)}
 
-\subsection*{Receive transitions in React}
+\subsection*{[Done] Receive transitions in React}
 This is the current workflow -- need to verify whether the receive transition is safe?

### 2020-04-21 notes/notes.tex
Notes: update 21/04/2020

@@
 
+\paragraph{20/04/2020}
+\begin{itemize}
+\item
+Prove preservation of wellformedness.
+\end{itemize}
+
 \section{MPST Encoding}
@@
 
-We also use the same process calculus defined before:
+We define an encoding $\llbracket \cdot \rrbracket :: G_{WS} \to G$
+which replaces client-to-client interactions to be routed via
+the server role.
 
-\begin{align*}
-P ::&= 0 \mid \mu X. P \mid X \\
-&\mid p \rhd \{ l_i: P_i \}_{i \in I} \\
-&\mid p \lhd l. P \\
-&\mid \texttt{if e then } P \texttt{ else } P\\\\
-\mathcal{M} ::&= p :: P \\
-&\mid \mathcal{M} \mid \mathcal{M}
-\end{align*}
+\paragraph{Well-formedness} A global type is well-formed if its
+projections exist for all participants. Formally,
 
-The typing judgement for a multiparty session with respect
-to the webservice global type builds upon the \textsc{[MTy]}
-rule.
+\[ \text{wellFormed}(G) \Longleftrightarrow 
+\forall p \in \texttt{pt}(G). \exists S. (G \upharpoonright p = S) \]
 
-\subsection*{Codegen}
-\texttt{codegen :: $\mathcal{M}$ $\times$ Role 
-$\to$ $\mathcal{M}$}
+This naturally extends to $G_{WS} = (G, s)$ as we just require that
+the server role is a participant in $G$.
 
-\subsubsection*{Main rules}
+\[ \text{wellFormed}_{WS}(G, s) = \text{wellFormed}(G)
+\wedge s \in \texttt{pt}(G) \]
 
-\begin{align*}
-&codegen~(p :: q \lhd l_j. P 
-\mid q :: p \rhd \{ l_i: Q_i \}_{i \in I} 
-\mid s :: S 
-\mid \mathcal{M})~s \\
-=~&codegen~(p :: s \lhd l. P 
-\mid q :: s \rhd \{l_i: Q_i\}_{i \in I}
-\mid s :: p \rhd )\\
-&where  \\
-\dots
-\end{align*}
+\subsection*{Preservation of well-formedness}
 
-\subsubsection*{Trivial rules}
+Take $G_{WS} = 
+(G, s)$ for some global type $G$ and server role $s$.
 
-\subsection*{Encoding}
+\paragraph{Proposition}
+If $\text{wellFormed}_{WS}(G_{WS})$, 
+then $\text{wellFormed}(\llbracket G_{WS} \rrbracket)$.
 
-\subsection*{Preservation}
-Modifying the process through code generation preserves
-well-typedness of the session.
-
-\paragraph{To show:} let $G_{WS} = (G, p)$. 
-If $\mathcal{M}$ is well-typed with respect to $G_{WS}$,
-then \texttt{codegen($\mathcal{M},~p$)} is well-typed
-with respect to $\llbracket G_{WS} \rrbracket$
+\paragraph{Proof}
+By induction on structure of $G$.
 

### 2020-04-23 notes/notes.tex
Notes: update 23/04/2020

@@
 Prove preservation of wellformedness.
+\end{itemize}
+
+\paragraph{23/04/2020}
+\begin{itemize}
+\item
+Prove preservation of full-mergeability -- for global types
+$G_1, G_2$ and role $r$, if $(G_1 \upharpoonright r)
+\sqcap (G_2 \upharpoonright r)$ exists, then the projections 
+from the encodings should be mergeable too, i.e. 
+($\llbracket G_1, s \rrbracket \upharpoonright r)
+\sqcap (\llbracket G_2, s \rrbracket \upharpoonright r)$
+exists too.
+
+\item
+Implement the TwoBuyer protocol in code -- modified templates
+for routing messages in server role.
+
+\item
+Identified additional lemmas to prove for full-mergeability;
+likewise for the well-formedness proof (found a caveat with
+inductive case for proving encoding of recursion, because
+the way the encoding handles recursion is different from how
+projection handles recursion -- the latter projects depending
+on whether the role to project to is a participant).
 \end{itemize}

### 2020-05-25 report/sections/evalconcl/concl/contributions.tex
Report: detailed scaffolding

--- 2020-03-09 PLACES_2020/conclusion.tex
@@
-\section{Conclusion and Future Work}
-We have presented a MPST-based framework for developing full-stack interactive
+\section{Contributions}
+
+\paragraph{Note: from PLACES}
+
+We have presented an MPST-based framework for developing full-stack interactive
 TypeScript applications with WebSocket communications.
@@
 channel usage guarantees for back-end and front-end targets respectively.
-
-Future work includes incorporating \textit{explicit connection actions}
-introduced in \cite{ExplicitConnections} in our API generation to better model
-real-world communication protocols that may feature in interactive web
-applications.
-% REVIEW: asynchronous
-Server-side implementations may perform asynchronous operations on the
-received messages, so supporting asynchronous values (such as JavaScript
-\textit{Promises} \cite{promise}) in IO actions would be a welcome addition.
-Whilst our approach supports multiparty sessions, the nature of
-WebSockets require some server-based role in the communication protocol and
-clients to interact via the server.
-Extending support to WebRTC \cite{WebRTC} would
-cater for peer-to-peer communication between browsers, which further opens up
-possibilities for communication protocols supported by our approach.
-
-% Future work
-% - generalise to encode protocols as typestates (e.g. Mungo example)
-% - WebRTC for full MPST (since current limitation is w.r.t. MPST protocols with 1-server many-client roles
-
-% - support asynchronous operations???

### 2020-05-25 report/sections/evalconcl/concl/future.tex
Report: detailed scaffolding

--- 2020-03-09 PLACES_2020/conclusion.tex
@@
-\section{Conclusion and Future Work}
-We have presented a MPST-based framework for developing full-stack interactive
-TypeScript applications with WebSocket communications that conform to a
-protocol, statically providing linear channel usage guarantees and affine
-channel usage guarantees for back-end and front-end targets respectively.
+\section{Future Work}
 
+\paragraph{Note: from PLACES}
 Future work includes incorporating \textit{explicit connection actions}
@@
 possibilities for communication protocols supported by our approach.
-
-% Future work
-% - generalise to encode protocols as typestates (e.g. Mungo example)
-% - WebRTC for full MPST (since current limitation is w.r.t. MPST protocols with 1-server many-client roles
-
-% - support asynchronous operations???

### 2020-05-25 report/sections/evalconcl/eval.tex
Report: detailed scaffolding

\chapter{Evaluation}
%benchmark -- compile time / benchmarking

### 2020-05-25 report/sections/introduction/contributions.tex
Report: detailed scaffolding

--- 2020-02-02 interim/sections/introduction.tex
@@
-\chapter{Introduction}
-% Rise of distributed programs -> main challenges
-% How to ensure correctness in general: type system, data types
-% Type discipline for concurrent programs - behavioural types (--> session types)
-% Mainstream example of distirbuted programs: web services (microservice architecture), interactive web application; 
-% Objective:
+\section{Contributions}
 
-Modern interactive web applications aim to provide a highly responsive user experience by minimising the communication latency between clients and servers. 
-Whilst the HTTP request-response model is sufficient for retrieving static assets, applying the same stateless communication approach for interactive use cases (such as multiplayer games) introduces undesirable performance overhead from having to frequently set up new connections for client-server interactions.
-Developers have since adopted other communication transport abstractions over HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy low-latency full-duplex client-server communication in their applications over a single persistent connection.
-Enabling more complex communication patterns caters for more interactive use cases, but introduces additional concerns to the developer with respect to implementation correctness.
+\paragraph{Note: from PLACES}
 
-% Example: noughts and crosses
-Consider a classic turn-based board game of \textit{Noughts and Crosses} between two players.
-Both players are identified by either noughts or crosses respectively, and take turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player wins (when their markers form one straight line on the board) or a stalemate is reached (when all cells are occupied).
-A web-based implementation may involve players connected to a game server via WebSocket connections and interacting with the game from their web browser, which serve a \textit{single-page application} (SPA) of the game client written in a popular framework like \textit{React.js} \cite{React}.
-SPAs feature a single HTML page and dynamically renders content via JavaScript in the browser.
-Players take turns to make a move on the game board and the server implements the game logic to progress the game forward until a result (either a win/loss or draw) can be declared.
-
-Whilst WebSockets make this web-based implementation possible, it introduces the developer to a new family of communication errors, even for this simple game.
-In addition to the usual testing for game logic correctness, the developer needs to test against \textit{deadlocks} (e.g.
-both players waiting for each other to make a move at the same time) and \textit{communication mismatches} (e.g.
-player 1 sending a boolean to the game server instead of the board coordinates).
-The complexity of these errors, which correlate to the complexity of tests required against these errors, scale with the complexity of the communication patterns involved.
-
-\textit{Multiparty Session Types} (MPST) \cite{MPST} provide a framework for formally specifying a structured communication pattern between concurrent processes and verifying implementations for correctness with respect to the communications aspect.
-By specifying the client-server interactions of our game as a protocol and verifying the implementations against the protocol for well-formedness, MPST theory guarantees well-formed implementations to be free from communication errors.
-
-We see the application of the MPST methodology to verifying interactive web applications written in TypeScript to be an interesting design space - to what extent can the MPST methodology be applied such that we leverage TypeScript language features to maximise static safety guarantees whilst preserving a flexible, natural and idiomatic workflow for TypeScript developers?
-Such a workflow would ultimately decrease the overhead for incorporating MPST into mainstream web development, which reduces development time by programmatically verifying implementations for communication correctness.
-
-\paragraph{Contributions} We present a workflow for developing type-safe interactive SPAs motivated by the MPST framework:
+We present a workflow for developing type-safe interactive SPAs motivated by the MPST framework:
 

### 2020-05-25 report/sections/introduction/motivation.tex
Report: detailed scaffolding

--- 2020-02-02 interim/sections/introduction.tex
@@
-\chapter{Introduction}
+\section{Motivation}
+
+\paragraph{Note: from PLACES}
+
 % Rise of distributed programs -> main challenges
@@
 Such a workflow would ultimately decrease the overhead for incorporating MPST into mainstream web development, which reduces development time by programmatically verifying implementations for communication correctness.
-
-\paragraph{Contributions} We present a workflow for developing type-safe interactive SPAs motivated by the MPST framework:
-
-\begin{enumerate}
-\item An endpoint API code generation workflow targeting TypeScript-based web applications for multiparty sessions;
-\item An encoding of session types in server-side TypeScript that enforces static linearity;
-\item An encoding of session types in browser-side TypeScript using the React framework that guarantees affine usage of communication channels.
-\end{enumerate}

### 2020-05-25 report/sections/server/codegen.tex
Report: detailed scaffolding

\chapter{\fancyname{SessionTS}: TypeScript API Generation}

\input{sections/server/codegen/workflow}

\input{sections/server/codegen/scribble}

\input{sections/server/codegen/efsm}

\input{sections/server/codegen/apigen}

### 2020-05-25 report/sections/server/node.tex
Report: detailed scaffolding

\chapter{\fancyname{NodeMPST}: Back-End Session Type Web Development}

\section{Motivation}

\section{Approach}

\section{State Encodings}

\subsection{Send}

\subsection{Receive}

\subsection{Terminal}

\section{EFSM Execution Runtime}

\section{Alternative Designs and Limitations}

\section{Summary}

### 2020-05-25 report/sections/general/implementation.tex
Report:  theory section scaffolding

\chapter{Implementation}

\section{Extending \fancyname{NodeMPST}}

\section{Extending \fancyname{ReactMPST}}

\section{Limitations}

### 2020-05-25 report/sections/general/theory.tex
Report:  theory section scaffolding

\chapter{Theory: Routed Multiparty Session Types}

\section{Syntax}

\subsection{Global Types}

\subsection{Local Types}

\section{Labelled Transition System (LTS) Semantics}

\subsection{Trace Equivalence}

\subsection{Deadlock Freedom}

\section{Router-Parameterised Encoding}

\subsection{Preserving Well-formedness}

\subsection{Preserving Communication}

### 2020-05-27 report/sections/general/p2p.tex
Report: section notes

\chapter{Motivation: Supporting Peer-to-Peer Interactions}

\section{\fancyname{Two Buyer} Protocol}
\begin{itemize}
\item TwoBuyer example
\item Not possible with WebSocket topology
\end{itemize}

\section{Server as a Router}
\begin{itemize}
\item How to make TwoBuyer example work just with WebSockets
\item Essentially `decomposing' multiparty session into binary interactions with respect to the server
\end{itemize}

\section{Challenges}
\begin{itemize}
\item Routed communication is different from `normal' communication
- server cannot overserialise interactions
\item Existing work in decomposing multiparty sessions (see LinearDecompScala paper and the binary one it references) assume the `server' takes on the pure role as a router -- our work argues that the server can still participate in the multiparty session and concurrently route messages between clients in a way that does not overserialise the interactions and allow the routing to be transparent to other roles.
\end{itemize}

### 2020-05-27 report/sections/server/codegen/workflow.tex
Report: section notes

\section{Workflow}

\begin{itemize}
\item introduce end-to-end workflow
\item must have diagram
\item describe input scribble protocol and flags, output is typescript code
\end{itemize}

### 2020-05-28 report/sections/server/codegen/testing.tex
Report: notes on part 1

\section{Testing}
\begin{itemize}
\item unit tests for code generator executable
\item systems tests for type checking the generated output with the typescript compiler
\item sandbox environment set up on the docker container to set up different tsconfig.json files for both react/node targets
\end{itemize}

### 2020-05-28 report/sections/server/node.tex
Report: notes on part 1

@@
 \section{Motivation}
+\begin{itemize}
+\item session type implementation on the node runtime
+\item need to provide consistency under the single-threaded event loop model
+\item provide idiomatic event-driven APIs
+\item provide static guarantees where possible -- communication mismatch and channel linearity
+\end{itemize}
 
 \section{Approach}
+\begin{itemize}
+\item typescript's type system is gradual, structural and non-linear -- session type api generation examples that provide static guarantees on channel usage build upon target language type system's support for linear resources
+\item inspired by purescript approach -- hide the channel resource so they are linear by construction, simply expose callbacks and handlers
+\item the runtime executes the EFSM, performs the send and receive actions through the websocket, and invokes the correct handler as required
+\item provide overview of the files we generate (and forward reference in the correct subsections) -- EFSM, Runtime, `others' (exclude cancellation)
+\end{itemize}
 
@@
 
+\begin{itemize}
+\item callbacks => type aliases for each state (handler)
+\item need a way for the runtime to distinguish between states
+\item attempt at using conditional types, but because they handle conditional union types in a distributed way (give example), cannot be statically typed
+\item alternative => use discriminated union by wrapping in an Implementation class
+\item all non-terminal states will `return' the successor implementation -- very difficult to resolve in the type-checker in the runtime (give example of what it may look like) -- solve by giving each state implementation the `advance' function to respect the event-driven nature of everything, so they can call it on completion
+\item visualise the interaction between runtime and states as a sequence diagram for message passing
+\end{itemize}
+
 \subsection{Send}
+\begin{itemize}
+\item very short -- go over the structure and give examples
+\item show how it extends for polyadic payloads
+\end{itemize}
 
 \subsection{Receive}
+\begin{itemize}
+\item very short -- go over the structure and give examples
+\item show how it extends for polyadic payloads
+\end{itemize}
 
 \subsection{Terminal}
+\begin{itemize}
+\item basically nothing here -- can highlight the provided type alias to provide a better developer experience
+\end{itemize}
 
-\section{EFSM Execution Runtime}
+\section{Runtime}
+\begin{itemize}
+\item public API -- provide seam for websocket server and implementation
+\item private API -- carry out the session
+\end{itemize}
+
+\subsection{Managing Connections}
+\begin{itemize}
+\item use set and partial object to keep track of pending connections
+\item ws event listeners
+\item forward reference that we need to manage cancellations here too
+\end{itemize}
+
+\subsection{Executing the EFSM}
+\begin{itemize}
+\item implementation details delegated to the actual implementation states
+\item constructor binds all methods (that will be passed to other components0 to `this' because of how javascript works with respect to the scoping of `this' (give short example -- or ignore, if this is in the background for typescript)
+\item advance -- use discriminated union to figure out what to pass to the state (send -- sendMessage, receive -- register)
+\end{itemize}
+
+\subsection{Handling Message Sends}
+\begin{itemize}
+\item very short (just for completeness in the report) -- messages are serialised JS objects (in JSON notation) sent and decoded on the other end, type-correct by construction
+\end{itemize}
+
+\subsection{Handling Message Receives}
+\begin{itemize}
+\item motivate the edge case that message can arrive in the websocket (in succession) before the handler is registered
+\item motivate the edge case that message can arrive `out of protocol order' 
+\item explain the double queue system and how that provides consistency -- emphasising that this works because of the single-threaded typescript runtime
+\end{itemize}
+
+\subsection{Handling Termination}
+\begin{itemize}
+\item design choice that the client is the one that closes the connection because of the centralised role of the server
+\item forward reference that this gives opportunity to extend support for general protocols, as in those protocols, the server's terminal state doesn't imply it's terminal for everyone else
+\end{itemize}
 
@@
 
+\begin{itemize}
+\item exposing channel resources like the java example -- less idiomatic (not-event-driven), need runtime linearity checks like the java example
+\item `hardcoded' switch case runtime like the TSOP papers -- doesn't best leverage the type system
+\end{itemize}
+
 \section{Summary}
+\begin{itemize}
+\item static guarantees on channel linearity and communication mismatch (up to the use of `any)
+\item handling receives introduce consistency problems, but we deal with it properly here
+\end{itemize}

### 2020-05-28 report/sections/server/react.tex
Report: detailed scaffolding
Report: notes on part 1

--- 2020-05-25 report/sections/server/node.tex
@@
-\chapter{\fancyname{NodeMPST}: Back-End Session Type Web Development}
+\chapter{\fancyname{ReactMPST}: Front-End Session Type Web Development}
+\label{chap:react}
 
 \section{Motivation}
+\begin{itemize}
+\item challenges of session typing user interface logic -- how does one formalise channel linearity?
+(give example of how complicated this actually is -- e.g.
+how to guarantee that, if a button is triggered for sending, that the action will not be triggered at other states?)
+\item existing works use non-mainstream language -- motivate Fowler LINKS, PureScript ConcurUI, identify that it needs some form of linear typing mechanism or `sequential building' UI to provide the safety properties
+\item challenge -- provide the same static communication safety guarantees for front-end logic written in typescript, in a way that is usable and idiomatic for the average typescript developer?
+\end{itemize}
 
 \section{Approach}
+\begin{itemize}
+\item leverage existing front-end frameworks that support reactive programming -- identify the parallel ideas that the UI element rendered on the screen needs to be reactive with respect to -- not only user interaction, but also channel actions.
+\item for example, UI should react to a receive state in the EFSM, and once the message is received on the channel, transition to a different UI with the supported actions
+\item main goal -- statically ensure that whatever IO actions are `present' on the window at any given time must respect the protocol
+\end{itemize}
+
+\subsection{The React Framework}
+\begin{itemize}
+\item use a class-based Counter example to explain the terms: props, state, uni-directional data flow, virtual DOM, reconciliation
+\end{itemize}
+
+\paragraph{Note: from PLACES}
+Our browser-side session type encodings for browser-side targets build upon the
+\emph{React.js} framework, developed by Facebook \cite{React} for the
+\textit{Model-View-Controller} (MVC) architecture.
+React is widely used in industry to create scalable single-page TypeScript
+applications, and we intend for our proposed workflow to be beneficial in an
+industrial context.
+We introduce the key features of the framework.
+
+\paragraph{Components}
+A component is a reusable UI element which
+contains its own markup and logic.
+Components implement a \texttt{render()} function which returns a React
+element, the smallest building blocks of a React application, analogous to the
+view function in the MVU architecture.
+Components can keep \textit{state}s and the \texttt{render()} function is
+invoked upon a change of state.
+
+For example, a simple counter can be implemented as a component,
+with its \texttt{count} stored as state.
+When rendered, it displays a button which increments \texttt{count}
+when clicked and a \texttt{div} that renders the current
+\texttt{count}.
+If the button is clicked, the \texttt{count} is incremented, which triggers a
+re-rendering (since the state has changed), and the updated \texttt{count} is
+displayed.
+
+Components can also render other components, which gives rise
+to a parent/child relationship between components.
+Parents can pass data to children as \textit{props} (short for properties).
+Going back to the aforementioned example, the counter component could
+render a child component \texttt{<StyledDiv count=\{this.state.count\} />} in
+its \texttt{render()} function, propagating the \texttt{count} from its state
+to the child.
+This enables reusability, and for our use case, gives control to the parent
+on what data to pass to its children (e.g. pass the payload of a received
+message to a child to render).
+
+\subsection{Model Types Revisited}
+\paragraph{Note: from PLACES paper}
+For browser-side targets, we extend the approach presented in \cite{MVU2019} on
+\textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU)
+architecture.
+% FEEDBACK: introduce MVU and model type
+An MVU application features a \textit{model} encapsulating application
+state, a \textit{view function} rendering the state on the Document Object Model (DOM), and an
+\textit{update function} handling \textit{messages} produced by the
+rendered model to produce a new model.
+The concept of model types express type dependencies between these
+components: a \emph{model type} uniquely defines a \textit{view function},
+set of \textit{messages} and \textit{update function} -- rather than
+producing a new model, the update function defines valid transitions to
+other model types.
+We leverage the correspondence between model types and states in the EFSM:
+each state in the EFSM is a model type, the set of messages represent
+the possible (IO) actions available at that state,
+and the update function defines which successor state to transition to,
+given the supported IO actions at this state.
 
@@
 
+\begin{itemize}
+\item motivated by the idea of model types, each state is encoded as a react component, where the user provides a view function (i.e. render())
+\item channel actions are abstracted by the runtime and passed to state implementations as props
+\item the developer provides an implementation of the state, i.e. ``uses the APIs'', by extending the abstract base classes and providing their own implementations
+\end{itemize}
+
 \subsection{Send}
+\begin{itemize}
+\item make the assumption that channel send actions are triggered by UI elements -- for example, clicking a button or typing enter in a textbox
+\item pre-session-types -- the developer will attach an event listener
+\item enhanced factory approach -- given an event and a event handler, return a react component that binds the handler to that event, and injects the channel action; the developer can use this API by putting the UI elements that represent the event action as children.
+\item the event handler is conditionally parameterised by the event type (achievable by typescript conditional types -- eg, if event is click, then the event handler parameter is a React.MouseEvent), and the return type is the payload type
+\item the abstract base class will set up the factory methods to be used by the developer
+\end{itemize}
+
+\subsubsection{Alternative Designs}
+\begin{itemize}
+\item not sure whether to put this in the subsection above so it flows better?
+\item naive implementation is to pass the send() function as a prop for the developer to incorporate in their event listener -- this is most flexible, but we cannot provide static guarantees
+\item factory approach -- the runtime could pass a factory function which, takes an UI element and performs the binding; this is better, as channel resources are not exposed, but the adversarial-minded developer can later override the event listener in the DOM etc
+\end{itemize}
 
 \subsection{Receive}
+\begin{itemize}
+\item receive state is `shown' when waiting for a receive event
+\item base class has abstract methods -- one for each possible receive -- so the user must implement these when they extend the abstract base class
+\item each abstract method is named after the message label expected to be received, and parameterised by the payload
+\item no channel actions are passed to the state -- instead, once the receive state is mounted on the DOM (i.e.
+rendered), the abstract base class will register the event handlers to the runtime (using the register method passed as props)
+\item when the message arrives, the runtime will invoke the registered function (+ forward reference the next section with the edge case that the message arrives before handler registration occurs)
+\end{itemize}
 
 \subsection{Terminal}
+\begin{itemize}
+\item by convention and design choice, the client gracefully closes the connection with the server role
+\end{itemize}
 
-\section{EFSM Execution Runtime}
+\section{Runtime}
+\begin{itemize}
+\item public API - accept websocket endpoint URL, a mapping of state to user implementation, and UI elements to render when waiting for connection to occur, and connection failure
+\item private API - carry out the session (just like Node)
+\end{itemize}
 
-\section{Alternative Designs and Limitations}
+\subsection{Session Joining}
+\begin{itemize}
+\item separate into two react components, since the websocket connection cannot be opened in the constructor by React StrictMode
+\item `outer' class sets websocket to undefined, but upon componentDidMount, creates the actual connection
+\item renders the `inner' class component upon websocket not being undefined, so it can be typed appropriately (otherwise, need to always use bang operator)
+\item on connection, send join request message to the server, wait for serber response before executing the EFSM at the initial state
+\end{itemize}
+
+\subsection{Executing the EFSM}
+\begin{itemize}
+\item same concept as node -- implementation details delegated to the actual classes
+\item constructor binds methods
+\item advance takes the enum instead of the class (since unlike the node example, it is not instantiated) -- uses custom type guards to distinguish between the type of state
+\end{itemize}
+
+\subsection{Handling Message Sends}
+\begin{itemize}
+\item passes in the factory prop, which is a factoryfactory to be precise;
+\item sendMessage is bound in the closure and sends the message in the same way as how Node codegen does, by convention
+\item (if not already mentioned in the node section) send is asynchronous
+\end{itemize}
+
+\subsection{Handling Message Receives}
+\begin{itemize}
+\item same edge case as the back-end code generation, as message can arrive before handlers are registered
+\item use the same two-queue mechanism to keep track of `what comes first' -- i.e.
+if message comes first, then put in message queue, so handler can pop off when registered; vice-versa, if handler is registered first, put in handler queue so message can be processed when that is received
+\item importantly -- only one queue is required, as server-centric topologies only have the client talk to the server
+\end{itemize}
+
+\subsection{Handling Termination}
+\begin{itemize}
+\item close the websocket with a clean exit code
+\end{itemize}
+
+\section{Limitations}
+\begin{itemize}
+\item affine channel usage guarantees -- cannot statically guarantee that the factory is attached to an event visible on the DOM
+\item decoupling communication state management from business logic state management -- runtime design does not permit passing non-communication-related data as props; need to use other methods for propagating applicaton data (such as React Context API or Redux)
+\item the point above might not be a `limitation', but more of an encouragement to change practices because it makes sense, architecturally, to separate concerns
+\end{itemize}
 
 \section{Summary}
+\begin{itemize}
+\item static guarantees on avoiding communication mismatch
+\item static guarantees on affine channel usage
+\item leverage the react framework to express the model types approach from Fowler 2020 appropriately, to provide static guarantees, by construction, that the channel actions made available at each state respect the protocol, and by definition, multiparty session type safety
+\end{itemize}

### 2020-05-28 report/sections/introduction/contributions.tex
Report: notes on part 1

@@
 
-\paragraph{Note: from PLACES}
+We present \dots
 
-We present a workflow for developing type-safe interactive SPAs motivated by the MPST framework:
+\paragraph{\cref{part:server}} \dots
 
-\begin{enumerate}
-\item An endpoint API code generation workflow targeting TypeScript-based web applications for multiparty sessions;
-\item An encoding of session types in server-side TypeScript that enforces static linearity;
-\item An encoding of session types in browser-side TypeScript using the React framework that guarantees affine usage of communication channels.
-\end{enumerate}
+\paragraph{\cref{part:general}} \dots
+
+\paragraph{\cref{part:evalconcl}} \dots

### 2020-05-28 report/sections/server/extensions/async.tex
Report: notes on part 1

\section{Supporting Asynchronous Implementations}

\subsection{Motivation}

\begin{itemize}
\item developer implementation in javascript -- let alone typescript -- will need asynchronous operations
\item backend -- database calls are asynchronous (e.g. for a multiplayer game, server may update database before sending the game results)
\item frontend -- may invoke third-party API calls (that are not part of the protocol) which return promises
\item how to support asynchronous developer implementations
\end{itemize}

\subsubsection{Concurrency Primitives in TypeScript}
\begin{itemize}
\item use an example to go over callbacks, promise and async/await functions
\item compare promises to java futures
\end{itemize}

\subsection{API Extension}
\begin{itemize}
\item should be very short section -- use union type MaybePromise so developer can provide either
\end{itemize}

\subsection{Runtime Extension}
\begin{itemize}
\item server send actions -- use the Thunk pattern to wrap the existing logic, the thunk takes the original payload type (extracted using conditional infer types) as parameter; implementation either returns promise or not-promise, so the implementation will either pass thunk as resolve or call thunk as-is
\item server receive actions - thunk pattern as well
\item browser send action -- need a runtime flag to silently disable channel reuse (which would otherwise violate linearity) because the sending action is delayed by the Promise resolve
\item browser receive action -- thunk pattern
\end{itemize}

\subsection{Limitations}
\begin{itemize}
\item for the browser target, we technically still provide affine linearity in the code, but the user would be able to click a button twice when it is sending as the promise resolution is asynchronous; to solve this, we add the used flag which is used in java paper for detecting channel reuse, which is `technically' a runtime check, depending on how one interprets
\end{itemize}

### 2020-05-28 report/sections/server/extensions/error.tex
Report: notes on part 1

\section{Error Handling}

\subsection{Motivation}
\begin{itemize}
\item existing work extends theory based on canonical session cancellations are from exceptions raised by logic
\item refer to 2FA example in Fowler paper on exceptional session types
\item need to also consider cancellations in a web-based scenario -- server dead, client closing their browser
\item need to also offer developer a seam for handling cancellations -- this needs to be well-parameterised so the developer knows how to handle the cancellation (for example, which role cancelled -- e.g. if it is a multi-way transaction, the server needs to know which role cancelled, so they can update the database accordingly).
\end{itemize}

\subsection{API Extension}
\begin{itemize}
\item developers just need to provide cancellation handlers along with the initial state
\item we standardise on parameterising it with role and reason: the reason is typed `any' to give developers the autonomy for implementation details
\end{itemize}

\subsection{Runtime Extension}
\begin{itemize}
\item build upon websocket RFC which specifies the `base' closing (or exit) code
\item we additionally define contracts for exit codes in our use case (e.g. role occupied, logical error), and make sure it does not override any websocket `traditions' as per the RFC
\item all based on well-defined event listeners on the websocket close and error events
\item must consider cancellation in two phases
\end{itemize}

\subsubsection{Cancellations during Session Joining}

\begin{itemize}
\item server -- emits role occupied
\item client -- closes browser whilst waiting for others
\end{itemize}

\subsubsection{Cancellations during EFSM Execution}

\begin{itemize}
\item server -- dies
\item server -- logical error
\item client -- dies
\item client -- logical error
\end{itemize}

\subsection{Limitations}
\begin{itemize}
\item Cancellation handlers are not typed -- namely, we use `any';
\item exception handler blocks in typescript are not typed
\item the `logical error' idea is also very vague as all code can be effectful in typescript (statically); cannot specifically say that the implementation is expected to `send payload of type int, but possibly throw error of type Error', which would otherwise be possible in other languages(?)
\end{itemize}

### 2020-05-28 report/sections/general/implementation.tex
Report: notes on part 2

@@
 
+\begin{itemize}
+\item key property from the theory is that the server's actions are effectively transparent, so we reflect that in the implementation
+\end{itemize}
+
 \section{Extending \fancyname{NodeMPST}}
+\begin{itemize}
+\item all messages have a recipient role
+\item for messages not intended for server, simply route transparently
+\end{itemize}
 
 \section{Extending \fancyname{ReactMPST}}
-
-\section{Limitations}
+\begin{itemize}
+\item receive as before, but determine the sender from the payload since it could have originated from a non-server role
+\item the double-queue system required for sending and receiving now need to be a map for React targets because the ordering is not preserved between roles (give example: A to B.
+C to B, server can route C's message before A's but B has to process A's first by protocol).
+\end{itemize}

### 2020-05-28 report/sections/general/p2p.tex
Report: notes on part 2

@@
 \begin{itemize}
-\item TwoBuyer example
+\item TwoBuyer example -- highlight the sections that show the `arbitrary topology'
 \item Not possible with WebSocket topology
@@
 \item How to make TwoBuyer example work just with WebSockets
+\item Quickly highlight WebRtC as an option, but it has its own proprietary connection `protocol', and will still need to rely on a centralised server role to set up the peer to peer connection; including the setup in the code generation will introduce unnecessary complications in error handling which deviates from the focus of implementing typesafe web services
 \item Essentially `decomposing' multiparty session into binary interactions with respect to the server
@@
 \item Existing work in decomposing multiparty sessions (see LinearDecompScala paper and the binary one it references) assume the `server' takes on the pure role as a router -- our work argues that the server can still participate in the multiparty session and concurrently route messages between clients in a way that does not overserialise the interactions and allow the routing to be transparent to other roles.
+\item theory side -- need to prove that it is indeed possible for the server to carry out its own interactions and perform routing in a transparent way, as if the messages were directly peer to peer
+\item implementation side -- changes need to reflect the theory and be as transparent to the developer as possible
 \end{itemize}

### 2020-05-29 report/sections/evalconcl/eval.tex
Report:  theory section scaffolding
Report: section notes
Report: notes on part 3

@@
 \chapter{Evaluation}
+\label{chap:eval}
 %benchmark -- compile time / benchmarking
+
+\section{Binary Sessions: \fancyname{Calculator}}
+\begin{itemize}
+\item sample client implementation + snippet of generated code (full in appendix)
+\item show that not all the UI is being controlled by the runtime (i.e. calculator app has a header/fake navbar)
+\item show the calculator logic achievable using React Context
+\item sample server implementation + snippet of generated code (full in appendix)
+\end{itemize}
+
+\section{Multiparty Sessions: \fancyname{Noughts and Crosses}}
+\begin{itemize}
+\item sample client implementation -- show the flexibility of factory API for sending to be used in the UI part (since move selection can be be made )
+\end{itemize}
+
+\section{Routed Multiparty Sessions: \fancyname{Two Buyers}}
+\begin{itemize}
+\item show that the routing aspect is transparent to both sides
+\end{itemize}
+
+\section{Benchmarks}
+\begin{itemize}
+\item compile time: \begin{itemize}
+\item use the examples from above to compare compilation time
+\end{itemize}
+\item execution time: based on parameterised ping pong with n messages
+\begin{itemize}
+\item n = 10, 100, 1000
+\item benchmark using node timer tools in the backend as it knows when the protocol ends
+\end{itemize}
+\end{itemize}

### 2020-05-29 report/sections/introduction/contributions.tex
Report: contributions

@@
 
-We present \dots
+We present an implementation of the MPST framework for developing
+interactive web applications over WebSocket transport. 
+Related work \citep{PureScript2019, MVU2020, LINKS, Tiers} requires the 
+developer to implement their application using tools that are 
+unconventional for web development.
+To the best of our knowledge, this is the first work on multiparty session
+type-safe web development that provides the same communication safety 
+guarantees using familiar industry tools in TypeScript and React.
 
-\paragraph{\cref{part:server}} \dots
+In \cref{chap:background}, we introduce the core session type theory
+and provide an overview of related work on integrating session types into
+application development. We also introduce TypeScript and highlight some 
+interesting features and caveats in the language.
 
-\paragraph{\cref{part:general}} \dots
+The remainder of the report is organised in three parts as such:
 
-\paragraph{\cref{part:evalconcl}} \dots
+\paragraph{\cref{part:server}}
+We present an end-to-end workflow for generating TypeScript APIs for 
+\emph{server-centric} communication protocols over WebSockets. Server-centric 
+protocols feature exactly one server role that is involved in all interactions.
+
+In \cref{chap:codegen}, we introduce the system architecture for our
+code generator, \fancyname{SessionTS}. 
+The implementation integrates with the \fancyname{Scribble}
+toolchain to generate handler style APIs that the developer can use to write 
+their web applications to guarantee protocol conformance by construction.
+
+We present two code generation strategies for back-end and front-end
+TypeScript web development as \fancyname{NodeMPST} and \fancyname{ReactMPST}
+in \cref{chap:node} and \cref{chap:react} respectively.
+
+In \cref{chap:ext}, we extend \fancyname{SessionTS} to support advanced
+practices commonly found in full-stack web development in a way that
+preserves communication safety. We modify the API generation strategies
+to support asynchronous implementations and introduce how to 
+gracefully handle errors and session cancellation 
+in the context of interactive web applications.
+
+\paragraph{\cref{part:general}} 
+We present a method to relax the server-centric constraint over WebSocket
+transport in order to support peer-to-peer interactions amongst 
+client roles, through the concept of \emph{routing}.
+
+In \cref{chap:p2p}, we introduce the \fancyname{TwoBuyer} multiparty
+protocol to motivate our proposal of providing the server role with 
+routing responsibilities in addition to its existing interactions
+specified in the protocol.
+
+In \cref{chap:theory}, we introduce a variant of multiparty 
+session type theory with routing. We prove that our extensions preserve 
+properties required for communication safety, 
+session fidelity and deadlock freedom.
+We provide an encoding for global types into routed multiparty 
+session type theory, and prove the preservation of well-formedness
+and communication. Detailed lemmas and proofs in this section can
+be found in \cref{appendix:proofs}.
+
+In \cref{chap:impl}, we demonstrate how our API generation strategies
+implement our routed session type theory to support arbitrary communication
+topologies.
+
+\paragraph{\cref{part:evalconcl}}
+In \cref{chap:eval}, we evaluate our work through case studies of communication
+protocols implemented as webservices. Through examples of binary sessions,
+multiparty sessions and routed multiparty sessions, we demonstrate how the
+developer may implement the generated APIs and consider the impact of our work
+on developer productivity.
+
+Finally, we summarise our work in \cref{chap:concl} and propose potential
+areas of improvement and future work.
+
+This project was published as \emph{Generating Interactive WebSocket 
+Applications in TypeScript} in the Proceedings of the 
+12th International Workshop on Programming Language Approaches to 
+Concurrency- \& Communication-cEntric Software (PLACES 2020) 
+\cite{PLACES2020}.

### 2020-05-29 report/sections/introduction/objectives.tex
Report: contributions

\section{Objectives}

The objective of this project is to \dots

### 2020-05-30 report/sections/evalconcl/eval.tex
Report: small changes to eval
Report: start on eval section

@@
 
-\section{Benchmarks}
-\begin{itemize}
-\item compile time: \begin{itemize}
-\item use the examples from above to compare compilation time
-\end{itemize}
-\item execution time: based on parameterised ping pong with n messages
-\begin{itemize}
-\item n = 10, 100, 1000
-\item benchmark using node timer tools in the backend as it knows when the protocol ends
-\end{itemize}
-\end{itemize}
+\section{Performance Benchmarks}
+
+Whilst web applications that implement our generated APIs enjoy
+communication safety guarantees, the presence of the session runtime acts
+as an additional layer of abstraction between the application logic and the
+WebSocket transport, which is likely to present a performance trade-off.
+
+To measure the overhead of our implementation, we compare the
+execution time of an interactive web application implementing the
+Ping Pong protocol (\cref{fig:pingpong}) using our generated APIs
+against an implementation of the protocol without session types.
+
+We standardise the application logic across experiments to be 
+parameterised by the number of round trip messages, $n$.
+Upon establishing a connection, the experiment proceeds as follow:
+
+\begin{enumerate}
+
+\item \trole{Client} sends a \tmsg{PING($m$:number)} message, 
+with $m = 0$ initially.
+
+\item \trole{Server} receives the \tmsg{PING($m$:number)} message, and
+conditionally responds based on the experiment parameter $n$:
+
+\begin{enumerate}
+\item If $m + 1 < n$, then \trole{Server} responds with \tmsg{PONG($m + 1$)}.
+\trole{Client} responds to the \tmsg{PONG} message by returning to
+step 1 with $m$ set as the payload of the received \tmsg{PONG} message.
+
+\item Otherwise, $m + 1 = n$, then \trole{Server} responds with 
+\tmsg{BYE($m + 1$)}, as $n$ round trips have taken place. 
+\trole{Client} responds to the
+\tmsg{BYE} message by closing the connection, thus ending the experiment.
+\end{enumerate}
+
+\end{enumerate}
+
+%\begin{itemize}
+%\item compile time: \begin{itemize}
+%\item use the examples from above to compare compilation time
+%\end{itemize}
+%\item execution time: based on parameterised ping pong with n messages
+%\begin{itemize}
+%\item n = 10, 100, 1000
+%\item benchmark using node timer tools in the backend as it knows when the protocol ends
+%\end{itemize}
+%\end{itemize}
+
+\subsection{Setup}
+
+% mention that it is hard to simulate multi-party sessions 
+% and see the overhead with increasing roles (eg ping pong with 50 roles)
+% , because of the need to
+% mock developer implementations for those roles
+
+We run the experiments under a network of latency 0.165ms
+(64 bytes ping), and repeat each experiment 20 times.
+Compilation time and execution time measurements 
+are taken using a machine equipped with Intel i7-4850HQ CPU
+(2.3 GHz, 4 cores, 8 threads), 16 GB RAM, macOS operating system 
+version 10.15.4, Node.js runtime version 12.12.0, and
+TypeScript compiler version 3.7.4.
+
+The other packages used in all implementations are outlined in their
+corresponding \texttt{package.json} files.
+
+\paragraph{\fancyname{Bare} variant}
+\dots
+
+\paragraph{\fancyname{BareSafe} variant}
+\dots
+
+\paragraph{\fancyname{MPST} variant}
+\dots
+
+\subsection{Execution Pattern}
+
+\subsection{Analysing Overhead}

### 2020-05-30 report/sections/evalconcl/eval.tex
Report: add to eval

--- 2020-05-29 report/sections/evalconcl/eval.tex
@@
 
-\section{Benchmarks}
+\section{Performance Benchmarks}
+
+Whilst web applications that implement our generated APIs enjoy
+communication safety guarantees, the presence of the session runtime acts
+as an additional layer of abstraction between the application logic and the
+WebSocket transport, which is likely to present a performance trade-off.
+
+\begin{figure}[!ht]
+\begin{lstlisting}[language=Scribble]
+global protocol PingPong(role Client, role Svr) {
+	PING(number) from Client to Svr;
+	choice at Svr {
+		PONG(number) from Svr to Client;
+		do PingPong(Client, Svr);
+	} or {
+		BYE(number) from Svr to Client;	
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{Ping Pong Protocol}
+\label{lst:pingpong}
+\end{figure}
+
+To measure the overhead of our implementation, we compare the
+execution time of an interactive web application implementing the
+Ping Pong protocol (\cref{lst:pingpong}) using our generated APIs,
+against implementations of the protocol \textit{without} session types.
+
+We standardise the application logic across experiments to be 
+parameterised by the number of round trip messages, $n > 0$.
+Upon establishing a connection, the experiment proceeds as follow:
+
+\begin{enumerate}
+
+\item \trole{Client} sends \tmsg{PING($m$:number)} to \trole{Svr}, 
+with $m = 0$ initially.
+
+\item \trole{Svr} receives \tmsg{PING($m$:number)}, and
+conditionally responds based on $n$:
+
+\begin{enumerate}
+\item If $m + 1 < n$, then \trole{Svr} replies \tmsg{PONG($m + 1$)}.
+\trole{Client} responds to \tmsg{PONG} by returning to
+step 1 with $m$ set as the payload from \tmsg{PONG}.
+
+\item Otherwise, $m + 1 = n$, then \trole{Svr} responds with 
+\tmsg{BYE($m + 1$)}, as $n$ round trips have taken place. 
+\trole{Client} responds to \tmsg{BYE} by 
+closing the connection, thus ending the experiment.
+\end{enumerate}
+
+\end{enumerate}
+
+We note that the Ping Pong protocol implements a \textit{binary} session. 
+It would be interesting to observe the overhead in a \textit{multiparty}
+context, but due to limited time constraints, we were unable to 
+extend our benchmarking suite to support multiple browser targets.
+Benchmarking multiparty protocols would also require writing multiple
+distinct React applications using the generated APIs -- as this is currently
+a manual process, doing this for multiple roles requires more time than
+available.
+
+\subsection{Setup}
+
+% mention that it is hard to simulate multi-party sessions 
+% and see the overhead with increasing roles (eg ping pong with 50 roles)
+% , because of the need to
+% mock developer implementations for those roles
+
+In order to measure the overhead as accurately as possible,
+we outline the logic that all implementations must follow:
+
+\subparagraph{Ping Pong \trole{Client} on React:}
 \begin{itemize}
-\item compile time: \begin{itemize}
-\item use the examples from above to compare compilation time
+
+\item All \trole{Client}s implement the same user interface, rendering
+a \texttt{<button>} which triggers the send, and
+a \texttt{<div>} captioned with the number of \tmsg{PONG}s received.
+
+\item \trole{Client}s will use the React Context API to manage 
+application state, i.e. the number of \tmsg{PONG}s received. 
+Session logic will be wrapped in a \texttt{ContextProvider}.
+
+\item We use the production build generated by 
+\texttt{create-react-app} \cite{cra} for all experiments, which performs the
+transpilation into JavaScript.
+
+\item We serve the production build on \url{localhost:3000} using the
+\texttt{serve} package \cite{npmserve} available on \texttt{npm}.
+
 \end{itemize}
-\item execution time: based on parameterised ping pong with n messages
+
+\subparagraph{Ping Pong \trole{Svr} on Node}
 \begin{itemize}
-\item n = 10, 100, 1000
-\item benchmark using node timer tools in the backend as it knows when the protocol ends
+
+\item timing using console.time
+
+\item start timing when connected
+
+\item stop timing when close
+
+\item using transpiled version
+
 \end{itemize}
+
+We run the experiments under a network of latency 0.165ms
+(64 bytes ping), and repeat each experiment 20 times.
+Execution time measurements  are taken using a machine 
+equipped with Intel i7-4850HQ CPU (2.3 GHz, 4 cores, 8 threads), 
+16 GB RAM, macOS operating system version 10.15.4, 
+Node.js runtime version 12.12.0, and
+TypeScript compiler version 3.7.4.
+We standardise all packages used in the front- and back-end
+implementations across experiments. Details can be found in their
+corresponding \texttt{package.json} files (\cref{appendix:eval}).
+
+We run our experiments using three implementations of the Ping Pong protocol:
+
+\begin{itemize}
+
+\item The \texttt{bare} implementation directly interfaces with 
+WebSocket primitives for sending and receiving.
+
+\item The \texttt{bare_safe} implementation \dots
+
+\item The \texttt{mpst} implementation \dots
+
 \end{itemize}
+
+\subsection{Execution Pattern}
+
+\subsection{Analysing Overhead}

### 2020-05-31 report/sections/evalconcl/eval.tex
Report: add benchmark data

@@
 
-We standardise the application logic across experiments to be 
-parameterised by the number of round trip messages, $n > 0$.
+We parameterise the Ping Pong protocol by $n > 0$, the number of
+round-trip messages. This is standardised in the application logic
+across experiments.
 Upon establishing a connection, the experiment proceeds as follow:
@@
 
-% mention that it is hard to simulate multi-party sessions 
-% and see the overhead with increasing roles (eg ping pong with 50 roles)
-% , because of the need to
-% mock developer implementations for those roles
-
 In order to measure the overhead as accurately as possible,
@@
 
-\item All \trole{Client}s implement the same user interface, rendering
+\item All \trole{Client}s implement the same user interface 
+(\cref{fig:pingpongclient}), rendering
 a \texttt{<button>} which triggers the send, and
@@
 
-\item \trole{Client}s will use the React Context API to manage 
-application state, i.e. the number of \tmsg{PONG}s received. 
-Session logic will be wrapped in a \texttt{ContextProvider}.
+\item \trole{Client}s will use the React Context API \cite{reactcontext}
+for application state management, i.e. the number of \tmsg{PONG}s received. 
+We wrap the session logic in a \texttt{<Benchmark>} component 
+that acts as the \texttt{ContextProvider} using its component state.
+
+\item To automate the benchmark, we use the React Refs API \cite{reactrefs}
+to access the DOM \texttt{<button>} node programmatically, in order to
+simulate the click event and send a \tmsg{PING} message upon establishing
+the WebSocket connection, or upon receiving a \tmsg{PONG}.
 
@@
 \texttt{create-react-app} \cite{cra} for all experiments, which performs the
-transpilation into JavaScript.
-
-\item We serve the production build on \url{localhost:3000} using the
+transpilation into JavaScript. We serve the production build using the
 \texttt{serve} package \cite{npmserve} available on \texttt{npm}.
@@
 
-\subparagraph{Ping Pong \trole{Svr} on Node}
+\begin{figure}[!ht]
+\captionof{figure}{User interface of Ping Pong \trole{Client}}
+\label{fig:pingpongclient}
+\end{figure}
+
+\subparagraph{Ping Pong \trole{Svr} on Node:}
 \begin{itemize}
 
-\item timing using console.time
+\item We use the built-in \texttt{console.time} function to record
+the execution time of all experiments. 
+The timer starts when a WebSocket connection has been established at
+\trole{Server}, and stops when on a \texttt{CloseEvent}.
 
-\item start timing when connected
+\item To observe the execution pattern, the \trole{Svr} will log the running 
+elapsed time for every \tmsg{PING} message received.
 
-\item stop timing when close
+\item All \trole{Svr}s run the benchmarks without a real web browser,
+using headless browsing functionality from 
+the Zombie.js \cite{zombie} package.
 
-\item using transpiled version
+\item The entry point to the \trole{Svr} logic is parameterised by the
+number of messages, $n$, configured through an environment variable
+passed through the command line.
+
+\item We use the transpiled JavaScript versions of all \trole{Svr}
+for the experiments.
 
 \end{itemize}
+
+The interested reader may follow the instructions under
+\texttt{benchmarks/README.md} in the project repository \cite{repo}
+to run the benchmarks and visualise the logs using the interactive
+notebook in the same directory.
 
@@
 implementations across experiments. Details can be found in their
-corresponding \texttt{package.json} files (\cref{appendix:eval}).
+corresponding \texttt{package.json} manifests (\cref{appendix:eval}).
 
@@
 
-\begin{itemize}
+\subparagraph{bare:}
+The \texttt{bare} implementation directly interfaces with 
+WebSocket primitives for sending and receiving. 
+The implementation executes the Ping Pong protocol, but does not 
+guarantee communication safety by construction -- e.g. the user can click
+the \tmsg{PING} button multiple times before a \tmsg{PONG} message 
+is received, violating channel linearity. 
+This represents the typical developer implementation without using the
+MPST framework.
 
-\item The \texttt{bare} implementation directly interfaces with 
-WebSocket primitives for sending and receiving.
+\subparagraph{bare_safe:}
+The \texttt{bare_safe} implementation also directly interfaces with
+WebSocket primitives for communication, but assumes the developer
+implements minimal viable workarounds to address the lack of
+communication safety. Here, the developer renders an inactive version
+of the \tmsg{PING} button when the \tmsg{PING} message has been sent
+but a response has yet to be received; a \texttt{visible} boolean flag
+is used to explicitly manage which \texttt{<button>} to render.
 
-\item The \texttt{bare_safe} implementation \dots
-
-\item The \texttt{mpst} implementation \dots
-
-\end{itemize}
+\subparagraph{mpst:}
+The \texttt{mpst} implementation follows the MPST framework and
+uses the APIs generated from \fancyname{SessionTS}, so it enjoys the
+communication safety guarantees from our methodology.
 
 \subsection{Execution Pattern}
+We compare the execution patterns of exchanging 10,000 Ping-Pongs
+throughout 20 repeated experiments across the three implementations.
+We visualise the elapsed time with respect to the number of \tmsg{PING}s
+received in \cref{fig:execution}.
 
-\subsection{Analysing Overhead}
+\begin{figure}[!h]
+\centering
+\begin{subfigure}[b]{0.3\textwidth}
+\centering
+%\includegraphics[width=\textwidth]{...}
+\caption{\texttt{bare}}
+\label{fig:executionbare}
+\end{subfigure}
+\hfill
+\begin{subfigure}[b]{0.3\textwidth}
+\centering
+%\includegraphics[width=\textwidth]{...}
+\caption{\texttt{bare_safe}}
+\label{fig:executionbaresafe}
+\end{subfigure}
+\hfill
+\begin{subfigure}[b]{0.3\textwidth}
+\centering
+%\includegraphics[width=\textwidth]{...}
+\caption{\texttt{mpst}}
+\label{fig:executionmpst}
+\end{subfigure}
+\captionof{figure}{Comparison of Execution Pattern for 10,000 Ping-Pongs}
+\label{fig:execution}
+\end{figure}
+
+\subsection{Overhead}
+We compare the total execution time (\textit{Exe. Time}) 
+and execution time per round trip (\textit{Exe. Time / Ping-Pong}) -- 
+averaged over 20 repeated experiments -- across the three implementations,
+for $n \in \{10^2, 10^3, 10^4\}$.
+We summarise the results in \cref{table:overhead}.
+
+\renewcommand{\arraystretch}{1.5}
+\begin{table}[!h]
+\centering
+\begin{tabular}{||c||c|c|c||c|c|c||}
+\hline
+\multirow{2}{*}{$n$} & 
+\multicolumn{3}{c||}{Exe. Time} & 
+\multicolumn{3}{c||}{Exe. Time / Ping-Pong} \\
+\cline{2-7}
+ & \texttt{bare} & \texttt{bare_safe} & \texttt{mpst} 
+ & \texttt{bare} & \texttt{bare_safe} & \texttt{mpst} \\
+\hline\hline
+100 & 89.64ms & 107.09ms & 186.23ms & 0.90ms & 1.07ms & 1.86ms \\
+1000 & 642.92ms & 663.91ms & 1155.48ms & 0.64ms & 0.66ms & 1.16ms \\
+10000 & 3542.16ms & 3837.97ms & 7015.25ms & 0.35ms & 0.38ms & 0.70ms \\
+\hline
+\end{tabular}
+\captionof{table}{Comparison of Execution Time for 100, 1,000 and 10,000
+Ping-Pongs}
+\label{table:overhead}
+\end{table}

### 2020-05-31 report/sections/evalconcl/eval.tex
Report: extract benchmark as its own TeX file

--- 2020-05-30 report/sections/evalconcl/eval.tex
@@
 
-\section{Performance Benchmarks}
-
-Whilst web applications that implement our generated APIs enjoy
-communication safety guarantees, the presence of the session runtime acts
-as an additional layer of abstraction between the application logic and the
-WebSocket transport, which is likely to present a performance trade-off.
-
-To measure the overhead of our implementation, we compare the
-execution time of an interactive web application implementing the
-Ping Pong protocol (\cref{fig:pingpong}) using our generated APIs
-against an implementation of the protocol without session types.
-
-We standardise the application logic across experiments to be 
-parameterised by the number of round trip messages, $n$.
-Upon establishing a connection, the experiment proceeds as follow:
-
-\begin{enumerate}
-
-\item \trole{Client} sends a \tmsg{PING($m$:number)} message, 
-with $m = 0$ initially.
-
-\item \trole{Server} receives the \tmsg{PING($m$:number)} message, and
-conditionally responds based on the experiment parameter $n$:
-
-\begin{enumerate}
-\item If $m + 1 < n$, then \trole{Server} responds with \tmsg{PONG($m + 1$)}.
-\trole{Client} responds to the \tmsg{PONG} message by returning to
-step 1 with $m$ set as the payload of the received \tmsg{PONG} message.
-
-\item Otherwise, $m + 1 = n$, then \trole{Server} responds with 
-\tmsg{BYE($m + 1$)}, as $n$ round trips have taken place. 
-\trole{Client} responds to the
-\tmsg{BYE} message by closing the connection, thus ending the experiment.
-\end{enumerate}
-
-\end{enumerate}
-
-%\begin{itemize}
-%\item compile time: \begin{itemize}
-%\item use the examples from above to compare compilation time
-%\end{itemize}
-%\item execution time: based on parameterised ping pong with n messages
-%\begin{itemize}
-%\item n = 10, 100, 1000
-%\item benchmark using node timer tools in the backend as it knows when the protocol ends
-%\end{itemize}
-%\end{itemize}
-
-\subsection{Setup}
-
-% mention that it is hard to simulate multi-party sessions 
-% and see the overhead with increasing roles (eg ping pong with 50 roles)
-% , because of the need to
-% mock developer implementations for those roles
-
-We run the experiments under a network of latency 0.165ms
-(64 bytes ping), and repeat each experiment 20 times.
-Compilation time and execution time measurements 
-are taken using a machine equipped with Intel i7-4850HQ CPU
-(2.3 GHz, 4 cores, 8 threads), 16 GB RAM, macOS operating system 
-version 10.15.4, Node.js runtime version 12.12.0, and
-TypeScript compiler version 3.7.4.
-
-The other packages used in all implementations are outlined in their
-corresponding \texttt{package.json} files.
-
-\paragraph{\fancyname{Bare} variant}
-\dots
-
-\paragraph{\fancyname{BareSafe} variant}
-\dots
-
-\paragraph{\fancyname{MPST} variant}
-\dots
-
-\subsection{Execution Pattern}
-
-\subsection{Analysing Overhead}
+\input{sections/evalconcl/eval/benchmarks}

### 2020-05-31 report/sections/evalconcl/eval/benchmarks.tex
Report: extract benchmark as its own TeX file

\section{Performance Benchmarks}

Whilst web applications that implement our generated APIs enjoy
communication safety guarantees, the presence of the session runtime acts
as an additional layer of abstraction between the application logic and the
WebSocket transport, which is likely to present a performance trade-off.

\begin{figure}[!ht]
\begin{lstlisting}[language=Scribble]
global protocol PingPong(role Client, role Svr) {
	PING(number) from Client to Svr;
	choice at Svr {
		PONG(number) from Svr to Client;
		do PingPong(Client, Svr);
	} or {
		BYE(number) from Svr to Client;	
	}
}
\end{lstlisting}
\captionof{lstlisting}{Ping Pong Protocol}
\label{lst:pingpong}
\end{figure}

To measure the overhead of our implementation, we compare the
execution time of an interactive web application implementing the
Ping Pong protocol (\cref{lst:pingpong}) using our generated APIs,
against implementations of the protocol \textit{without} session types.

We parameterise the Ping Pong protocol by $n > 0$, the number of
round-trip messages. This is standardised in the application logic
across experiments.
Upon establishing a connection, the experiment proceeds as follow:

\begin{enumerate}

\item \trole{Client} sends \tmsg{PING($m$:number)} to \trole{Svr}, 
with $m = 0$ initially.

\item \trole{Svr} receives \tmsg{PING($m$:number)}, and
conditionally responds based on $n$:

\begin{enumerate}
\item If $m + 1 < n$, then \trole{Svr} replies \tmsg{PONG($m + 1$)}.
\trole{Client} responds to \tmsg{PONG} by returning to
step 1 with $m$ set as the payload from \tmsg{PONG}.

\item Otherwise, $m + 1 = n$, then \trole{Svr} responds with 
\tmsg{BYE($m + 1$)}, as $n$ round trips have taken place. 
\trole{Client} responds to \tmsg{BYE} by 
closing the connection, thus ending the experiment.
\end{enumerate}

\end{enumerate}

We note that the Ping Pong protocol implements a \textit{binary} session. 
It would be interesting to observe the overhead in a \textit{multiparty}
context, but due to limited time constraints, we were unable to 
extend our benchmarking suite to support multiple browser targets.
Benchmarking multiparty protocols would also require writing multiple
distinct React applications using the generated APIs -- as this is currently
a manual process, doing this for multiple roles requires more time than
available.

\subsection{Setup}

In order to measure the overhead as accurately as possible,
we outline the logic that all implementations must follow:

\subparagraph{Ping Pong \trole{Client} on React:}
\begin{itemize}

\item All \trole{Client}s implement the same user interface 
(\cref{fig:pingpongclient}), rendering
a \texttt{<button>} which triggers the send, and
a \texttt{<div>} captioned with the number of \tmsg{PONG}s received.

\item \trole{Client}s will use the React Context API \cite{reactcontext}
for application state management, i.e. the number of \tmsg{PONG}s received. 
We wrap the session logic in a \texttt{<Benchmark>} component 
that acts as the \texttt{ContextProvider} using its component state.

\item To automate the benchmark, we use the React Refs API \cite{reactrefs}
to access the DOM \texttt{<button>} node programmatically, in order to
simulate the click event and send a \tmsg{PING} message upon establishing
the WebSocket connection, or upon receiving a \tmsg{PONG}.

\item We use the production build generated by 
\texttt{create-react-app} \cite{cra} for all experiments, which performs the
transpilation into JavaScript. We serve the production build using the
\texttt{serve} package \cite{npmserve} available on \texttt{npm}.

\end{itemize}

\begin{figure}[!ht]
\captionof{figure}{User interface of Ping Pong \trole{Client}}
\label{fig:pingpongclient}
\end{figure}

\subparagraph{Ping Pong \trole{Svr} on Node:}
\begin{itemize}

\item We use the built-in \texttt{console.time} function to record
the execution time of all experiments. 
The timer starts when a WebSocket connection has been established at
\trole{Server}, and stops when on a \texttt{CloseEvent}.

\item To observe the execution pattern, the \trole{Svr} will log the running 
elapsed time for every \tmsg{PING} message received.

\item All \trole{Svr}s run the benchmarks without a real web browser,
using headless browsing functionality from 
the Zombie.js \cite{zombie} package.

\item The entry point to the \trole{Svr} logic is parameterised by the
number of messages, $n$, configured through an environment variable
passed through the command line.

\item We use the transpiled JavaScript versions of all \trole{Svr}
for the experiments.

\end{itemize}

The interested reader may follow the instructions under
\texttt{benchmarks/README.md} in the project repository \cite{repo}
to run the benchmarks and visualise the logs using the interactive
notebook in the same directory.

We run the experiments under a network of latency 0.165ms
(64 bytes ping), and repeat each experiment 20 times.
Execution time measurements  are taken using a machine 
equipped with Intel i7-4850HQ CPU (2.3 GHz, 4 cores, 8 threads), 
16 GB RAM, macOS operating system version 10.15.4, 
Node.js runtime version 12.12.0, and
TypeScript compiler version 3.7.4.
We standardise all packages used in the front- and back-end
implementations across experiments. Details can be found in their
corresponding \texttt{package.json} manifests (\cref{appendix:eval}).

We run our experiments using three implementations of the Ping Pong protocol:

\subparagraph{bare:}
The \texttt{bare} implementation directly interfaces with 
WebSocket primitives for sending and receiving. 
The implementation executes the Ping Pong protocol, but does not 
guarantee communication safety by construction -- e.g. the user can click
the \tmsg{PING} button multiple times before a \tmsg{PONG} message 
is received, violating channel linearity. 
This represents the typical developer implementation without using the
MPST framework.

\subparagraph{bare_safe:}
The \texttt{bare_safe} implementation also directly interfaces with
WebSocket primitives for communication, but assumes the developer
implements minimal viable workarounds to address the lack of
communication safety. Here, the developer renders an inactive version
of the \tmsg{PING} button when the \tmsg{PING} message has been sent
but a response has yet to be received; a \texttt{visible} boolean flag
is used to explicitly manage which \texttt{<button>} to render.

\subparagraph{mpst:}
The \texttt{mpst} implementation follows the MPST framework and
uses the APIs generated from \fancyname{SessionTS}, so it enjoys the
communication safety guarantees from our methodology.

\subsection{Execution Pattern}
We compare the execution patterns of exchanging 10,000 Ping-Pongs
throughout 20 repeated experiments across the three implementations.
We visualise the elapsed time with respect to the number of \tmsg{PING}s
received in \cref{fig:execution}.

\begin{figure}[!h]
\centering
\begin{subfigure}[b]{0.3\textwidth}
\centering
\includegraphics[width=\textwidth]{execbare10000}
\caption{\texttt{bare}}
\label{fig:executionbare}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.3\textwidth}
\centering
\includegraphics[width=\textwidth]{execbaresafe10000}
\caption{\texttt{bare_safe}}
\label{fig:executionbaresafe}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.3\textwidth}
\centering
\includegraphics[width=\textwidth]{execmpst10000}
\caption{\texttt{mpst}}
\label{fig:executionmpst}
\end{subfigure}
\captionof{figure}{Comparison of Execution Pattern for 10,000 Ping-Pongs}
\label{fig:execution}
\end{figure}

\subsection{Overhead}
We compare the total execution time (\textit{Exe. Time}) 
and execution time per round trip (\textit{Exe. Time / Ping-Pong}) -- 
averaged over 20 repeated experiments -- across the three implementations,
for $n \in \{10^2, 10^3, 10^4\}$.
We summarise the results in \cref{table:overhead}.

\renewcommand{\arraystretch}{1.5}
\begin{table}[!h]
\centering
\begin{tabular}{||c||c|c|c||c|c|c||}
\hline
\multirow{2}{*}{$n$} & 
\multicolumn{3}{c||}{Exe. Time} & 
\multicolumn{3}{c||}{Exe. Time / Ping-Pong} \\
\cline{2-7}
 & \texttt{bare} & \texttt{bare_safe} & \texttt{mpst} 
 & \texttt{bare} & \texttt{bare_safe} & \texttt{mpst} \\
\hline\hline
$10^2$ & 89.64ms & 107.09ms & 186.23ms & 0.90ms & 1.07ms & 1.86ms \\
$10^3$ & 642.92ms & 663.91ms & 1155.48ms & 0.64ms & 0.66ms & 1.16ms \\
$10^4$ & 3542.16ms & 3837.97ms & 7015.25ms & 0.35ms & 0.38ms & 0.70ms \\
\hline
\end{tabular}
\captionof{table}{Comparison of Execution Time for 
100, 1,000 and 10,000 Ping-Pongs}
\label{table:overhead}
\end{table}
\renewcommand{\arraystretch}{1}

We note that the addition of a session runtime for all roles in the 
\texttt{mpst} implementation \textit{does} incur a performance overhead. 
This is made apparent when looking closely at 
\textit{Exe. Time / Ping-Pong};
we visualise this in \cref{fig:timepermsg}.

\begin{figure}[!h]
\centering
\includegraphics[width=\textwidth]{timepermsg}
\captionof{figure}{Comparing Average Time per Ping-Pong 
Across Implementations}
\label{fig:timepermsg}
\end{figure}

The \texttt{mpst} implementation records greater round trip times compared
to both \texttt{bare} and \texttt{bare_safe} variants. This is expected,
as \dots

### 2020-05-31 report/sections/evalconcl/eval/benchmarks.tex
Report: benchmarks

@@
 
-\item The entry point to the \trole{Svr} logic is parameterised by the
-number of messages, $n$, configured through an environment variable
-passed through the command line.
+\item \trole{Svr} logic is parameterised by the number of round trips,
+$n$, configured through an environment variable passed through 
+the command line.
 
-\item We use the transpiled JavaScript versions of all \trole{Svr}
+\item We use the transpiled JavaScript versions of all \trole{Svr}s
 for the experiments.
@@
 
-We run our experiments using three implementations of the Ping Pong protocol:
+The benchmark compares three implementations of the Ping Pong protocol:
 
@@
 \subparagraph{mpst:}
-The \texttt{mpst} implementation follows the MPST framework and
-uses the APIs generated from \fancyname{SessionTS}, so it enjoys the
+The \texttt{mpst} implementation uses the APIs generated from 
+\fancyname{SessionTS}, so it enjoys the
 communication safety guarantees from our methodology.
@@
 \subsection{Execution Pattern}
+\label{section:execpattern}
+
 We compare the execution patterns of exchanging 10,000 Ping-Pongs
@@
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
 \centering
@@
 
+Relative to the two \texttt{bare} implementations, the
+\texttt{mpst} implementation shows more consistency.
+
 \subsection{Overhead}
-We compare the total execution time (\textit{Exe. Time}) 
-and execution time per round trip (\textit{Exe. Time / Ping-Pong}) -- 
+We compare the total execution time (\textit{Exec. Time}) 
+and execution time per round trip (\textit{Exec. Time / Ping-Pong}) -- 
 averaged over 20 repeated experiments -- across the three implementations,
@@
 \renewcommand{\arraystretch}{1.5}
-\begin{table}[!h]
+\begin{table}[!ht]
 \centering
@@
 \multirow{2}{*}{$n$} & 
-\multicolumn{3}{c||}{Exe. Time} & 
-\multicolumn{3}{c||}{Exe. Time / Ping-Pong} \\
+\multicolumn{3}{c||}{Exec. Time} & 
+\multicolumn{3}{c||}{Exec. Time / Ping-Pong} \\
 \cline{2-7}
@@
 to both \texttt{bare} and \texttt{bare_safe} variants. This is expected,
-as \dots
+as the runtime for \textit{each} role performs additional logic during
+\textit{both} the sending and receiving of messages. 
+For example, receiving a message involves performing
+(albeit $O(1)$ time complexity) operations on the message queue and
+handler queue. As for browser-side implementations, every EFSM transition
+invokes an updated \texttt{render()} on the VDOM, which requires
+reconciliation internally by React to update the browser DOM accordingly.
+
+We also observe that the (round trip time) performance gap between 
+the \texttt{mpst} implementation and the two 
+\texttt{bare} implementations narrows as the number of round trips 
+increase. This suggests that the session joining logic in our
+implementation yields greater overhead than the additional processing
+logic injected during communication, which is consistent with 
+our findings in \cref{section:execpattern}.
+
+We interpret the overhead as a trade-off between maximising performance and
+maximising the \textbf{static} communication safety guarantees for
+web applications. In particular, the event-driven nature of 
+browser-side logic makes it highly challenging to guarantee 
+communication safety, as having ``active'' channel actions on the DOM
+allows linearity to be violated by the user, even if one verifies
+linear channel usage in the source code. The \texttt{mpst} implementation
+properly addresses the problem by statically guaranteeing,
+by construction, that the UI component rendered on the DOM will only
+contain channel actions permitted at that state in the EFSM execution.
+
+One may argue that the \texttt{bare_safe} implementation also
+provides said guarantee but adds negligible overhead. 
+By inspecting the workaround in the \texttt{bare_safe} implementation
+(\cref{lst:workaround}),
+it is clear that this does not scale for more complicated protocols with
+larger EFSMs.
+
+\begin{figure}[!ht]
+\begin{lstlisting}[language=javascript, tabsize=2]
+click() {
+	this.state.ws?.send(JSON.stringify({
+		label: 'PING',
+		payload: [this.context.count],
+	}));
+	this.setState({ visible: false });
+}
+
+render() {
+	return this.state.visible 
+		? <button
+		   ref={this.state.button}
+		   onClick={this.click.bind(this)}
+		   >Ping</button>
+		: <button>Ping</button>;
+}
+\end{lstlisting}
+\captionof{lstlisting}{Preventing user-triggered linearity violation in
+\texttt{bare_safe} Ping Pong}
+\label{lst:workaround}
+\end{figure}
+
+The workaround is tailored to the Ping Pong protocol
+and toggles the \texttt{visible} flag to hide the button that triggers
+a send action when \trole{Client} transitions to a receive state.
+In fact, this ``workaround'' precisely generalises to our 
+\texttt{mpst} implementation of having some form of wrapper
+component that renders the UI component corresponding to the
+current EFSM state. Our API generation strategy formally implements
+the EFSM for browser-side logic.
+
+We also compare the server-side logic between the \texttt{base} and
+\texttt{mpst} implementations in \cref{fig:nodepingpong}, and observe
+that our generated APIs (\cref{lst:nodepingpongmpst}) 
+allow the developer to focus on the
+implementation detail, whilst the naive bare implementation 
+(\cref{lst:nodepingpongbare}) chooses to interleave communication logic
+with application logic, which arguably contributes towards a hidden
+source of bugs found when implementing more complex protocols.
+
+\begin{figure}[!ht]
+\centering
+\begin{subfigure}[b]{\textwidth}
+\centering
+\begin{lstlisting}[language=javascript,tabsize=2]
+socket.onmessage = ({ data }) => {
+	const { label, payload } = JSON.parse(data.toString());
+	if (label === 'PING') {
+		let count: number = payload[0];
+		console.timeLog(LABEL, ++count);
+		if (count === MSGS) {
+			socket.send(JSON.stringify({
+				label: 'BYE',
+				payload: [count],
+			}));
+		} else {
+			socket.send(JSON.stringify({
+				label: 'PONG',
+				payload: [count],
+			}));
+		}
+	} else {
+		throw new Error(`Unrecognised label: ${label}`)
+	}
+}
+\end{lstlisting}
+\caption{\texttt{bare}}
+\label{lst:nodepingpongbare}
+\end{subfigure}
+\hfill
+\begin{subfigure}[b]{\textwidth}
+\centering
+\begin{lstlisting}[language=javascript,tabsize=2]
+// NOTE: `Implementation' namespace omitted for brevity.
+
+const logic = new Initial({
+	PING: (count) => {
+		console.timeLog(LABEL, ++count);
+		if (count === MSGS) {
+			return new S16([Labels.S16.BYE, [count], new Terminal()]);
+		} else {
+			return new S16([Labels.S16.PONG, [count], logic]);
+		}
+ 	}
+});
+\end{lstlisting}
+\caption{\texttt{mpst}}
+\label{lst:nodepingpongmpst}
+\end{subfigure}
+\captionof{figure}{Comparison of \trole{Svr} implementations in benchmark}
+\label{fig:nodepingpong}
+\end{figure}

### 2020-05-31 report/sections/evalconcl/eval/benchmarks.tex
Report: benchmark

@@
 Relative to the two \texttt{bare} implementations, the
-\texttt{mpst} implementation shows more consistency.
+\texttt{mpst} version performs more consistently, perhaps as a result
+of the session runtimes handling all WebSocket interactions in a
+systematic way. The gradient of the graph represents the rate at which
+a Ping-Pong round trip takes place. 
+We observe a steeper gradient when the protocol begins, 
+which illustrates the overhead incurred in the 
+session joining phase in our generated APIs.
+Aside from these factors, all three implementations generally share
+similar characteristics in their protocol execution.
 
@@
 	} else {
-		throw new Error(`Unrecognised label: ${label}`)
+		throw new Error(`Unrecognised label: ${label}`);
 	}
@@
 \end{lstlisting}
-\caption{\texttt{bare}}
+\caption{\texttt{bare} implementation of Ping Pong \trole{Svr}}
 \label{lst:nodepingpongbare}
@@
 
-const logic = new Initial({
+const logic = new S14({
 	PING: (count) => {
@@
 		if (count === MSGS) {
-			return new S16([Labels.S16.BYE, [count], new Terminal()]);
+			return new S16([Labels.S16.BYE, [count], new S15()]);
 		} else {
@@
 \end{lstlisting}
-\caption{\texttt{mpst}}
+\caption{\texttt{mpst} implementation of Ping Pong \trole{Svr}}
 \label{lst:nodepingpongmpst}
 \end{subfigure}
-\captionof{figure}{Comparison of \trole{Svr} implementations in benchmark}
+\captionof{figure}{Comparison of \texttt{bare} and \texttt{mpst}
+implementations for Ping Pong \trole{Svr}}
 \label{fig:nodepingpong}

### 2020-05-31 report/sections/server/codegen.tex
Report: notes on part 1
Report: small changes to eval
Report: start chapter 3

@@
-\chapter{\fancyname{SessionTS}: TypeScript API Generation}
+\chapter{\fancyname{SessionTS}: Session Type API Generation for TypeScript}
+\label{chap:codegen}
+
+In this chapter, we present
+\fancyname{SessionTS}, an API generation toolchain for building
+web applications in TypeScript that conform to a communication protocol 
+based on multiparty session type theory.
+
+\fancyname{SessionTS} supports communication protocols that define 
+\emph{server-centric} topologies, meaning that: 
+\textbf{(1)} there is exactly one participant executed on the 
+Node.js runtime; 
+\textbf{(2)} all other participants run on the web browser; and
+\textbf{(3)} all non-server participants only communicate with the server.
+We relax this assumption in \cref{part:general}.
+
+The toolchain is publicly available at \cite{repo}.
 
@@
 \input{sections/server/codegen/apigen}
+
+\input{sections/server/codegen/testing}

### 2020-05-31 report/sections/evalconcl/eval/benchmarks.tex
Report: complete benchmarks
Report: proofread some benchmark

@@
 
-Whilst web applications that implement our generated APIs enjoy
+Whilst web applications implementing our generated APIs enjoy
 communication safety guarantees, the presence of the session runtime acts
@@
 To measure the overhead of our implementation, we compare the
-execution time of an interactive web application implementing the
-Ping Pong protocol (\cref{lst:pingpong}) using our generated APIs,
-against implementations of the protocol \textit{without} session types.
+execution time of web-based implementations of the
+Ping Pong protocol (\cref{lst:pingpong}) with and without
+our generated APIs.
 
@@
 It would be interesting to observe the overhead in a \textit{multiparty}
-context, but due to limited time constraints, we were unable to 
-extend our benchmarking suite to support multiple browser targets.
+context, but due to limited time constraints,
+our benchmarking suite does not support multiple browser targets.
 Benchmarking multiparty protocols would also require writing multiple
@@
 we outline the logic that all implementations must follow:
+
+\begin{figure}[!ht]
+\centering
+\includegraphics[width=0.5\textwidth]{pingpongclient}
+\captionof{figure}{User interface of Ping Pong \trole{Client}}
+\label{fig:pingpongclient}
+\end{figure}
 
@@
 \end{itemize}
-
-\begin{figure}[!ht]
-\captionof{figure}{User interface of Ping Pong \trole{Client}}
-\label{fig:pingpongclient}
-\end{figure}
 

### 2020-05-31 report/sections/server/codegen/workflow.tex
Report: commented notes for chapter 3

@@
 \section{Workflow}
+
+We introduce the high-level system architecture of \fancyname{SessionTS}
+in \cref{fig:workflow}.
+
+\begin{itemize}
+
+\item We integrate with the \fancyname{Scribble} toolchain to
+specify the protocol, validate its well-formedness and 
+formalise and validate the well-formedness of the communication protocol.
+
+The communication protocol is specified using \fancyname{Scribble},
+
+\end{itemize}
+
+\cref{fig:cmdline} outlines the information required at the entry
+point of \fancyname{SessionTS}:
+
+\begin{itemize}
+
+\item The protocol and role name are required to
+
+\end{itemize}
+
+The developer provides the following information at the 
+command line entry point:
+\textbf{(1)} the protocol to generate,
+\textbf{(2)} the role to generate for, 
+\textbf{(3)} the code generation target.
+We adopt different code generation strategies for back-end 
+(\cref{chap:node}) and front-end (\cref{chap:react}) implementations.
+
+, and conditionally
+\textbf{(4)} the server role.
 

### 2020-06-01 report/sections/server/codegen.tex
Report: restructure chapter 3

@@
 In this chapter, we present
-\fancyname{SessionTS}, an API generation toolchain for building
-web applications in TypeScript that conform to a communication protocol 
-based on multiparty session type theory.
+\fancyname{SessionTS},
+a toolchain for generating TypeScript APIs that developers can
+use to write web applications that conform to their specified
+communication protocol.
+The toolchain is publicly available at \cite{repo}.
 
@@
 \textbf{(3)} all non-server participants only communicate with the server.
+Examples of such protocols include the \tprotocol{Calculator} webservice
+(\cref{section:evalcalculator}) and the \tprotocol{Noughts and Crosses}
+multiplayer game (\cref{section:evalgame}).
 We relax this assumption in \cref{part:general}.
-
-The toolchain is publicly available at \cite{repo}.
 
@@
 
-\input{sections/server/codegen/scribble}
-
-\input{sections/server/codegen/efsm}
-
-\input{sections/server/codegen/apigen}
+\input{sections/server/codegen/implementation}
 

### 2020-06-01 report/sections/server/codegen/implementation.tex
Report: restructure chapter 3

\section{Implementation}

\fancyname{SessionTS} is written in Python. 
It offers flexible syntax, a rich standard library, and 
a healthy ecosystem of DOT graph parsers and template engines,
making it a suitable choice for implementing our code generator.
We provide a simple command line interface (\cref{lst:cmdline})
for the developer to supply the required information 
to use \fancyname{SessionTS}.

\begin{figure}[!ht]
\begin{lstlisting}
root@mpst_ts:/home# python3.7 -m mpst_ts --help
usage: __main__.py [-h] [-s SERVER] [-o OUTPUT]
                   filename protocol role {browser,node}

positional arguments:
  filename       Path to Scribble protocol
  protocol       Name of protocol
  role           Role to project
  {browser,node} Code generation target

optional arguments:
  -h, --help     show this help message and exit
  -s SERVER, --server SERVER
                 Server role (only applicable for browser targets)
  -o OUTPUT, --output OUTPUT
                 Output directory for generation
\end{lstlisting}
\captionof{figure}{Entry Point for \fancyname{SessionTS}}
\label{lst:cmdline}
\end{figure}

We use Docker \cite{docker} to encapsulate our code generator
and its dependencies -- we found the canonical Python virtual environment
solution to be insufficient, as the Scribble toolchain is
a standalone Java executable with non-trivial setup procedures.
The Dockerfile builds a Docker image with \fancyname{Scribble} and
the Python dependencies all pre-configured, and the provided
\texttt{build.sh} script instantiates the image as a container 
for development.
The \texttt{start.sh} script enters the Docker container
development environment and mounts the local project directory
onto the container as a volume to synchronise changes between the
two environments.

### 2020-06-01 report/sections/server/codegen/workflow.tex
Report: restructure chapter 3

--- 2020-05-27 report/sections/server/codegen/workflow.tex
@@
-\section{Workflow}
+\section{Development Workflow}
+
+% input: protocol, role, target
+% then: verify well-formedness
+% then: obtain EFSM for role
+% then: parse EFSM
+% then: generate code for said target
+% output: API
+
+We motivate our development workflow from previous work 
+\cite{PureScript2019} by extending the \fancyname{Scribble} toolchain
+and generating APIs that integrate developer implementation
+into the EFSM execution.
+
+We visualise the workflow in \cref{fig:devworkflow} 
+and provide a brief overview:
+
+\begin{enumerate}
+
+\item The developer supplies the communication protocol written in
+\fancyname{Scribble} (\cref{subsection:scribble}), 
+stating the role (hereafter \textit{endpoint})
+to generate APIs for,
+and the code generation \textit{target} 
+(i.e. whether the role runs on the server or the web browser).
+
+\item \fancyname{SessionTS} delegates to the 
+\fancyname{Scribble} toolchain for verifying the well-formedness of
+the protocol and expects to receive a DOT graph representation of
+the endpoint FSM (\cref{subsection:efsm}). 
+\fancyname{SessionTS} parses the endpoint's 
+interactions from the DOT graph and generates TypeScript APIs
+for the developer (\cref{subsection:apigen}) 
+tailored to the specified target.
+
+\item The developer implements their web application using the
+generated APIs. Implementations that pass the type-checking phase
+of the TypeScript Compiler are guaranteed to be free from 
+communication errors by session type theory.
+
+\end{enumerate}
+
+\begin{figure}[!ht]
+\centering
+\includegraphics[width=\textwidth]{DevelopmentWorkflow}
+\captionof{figure}{Overview of \fancyname{SessionTS} Development Workflow}
+\label{fig:devworkflow}
+\end{figure}
+
+\subsection{Protocol Specification with \fancyname{Scribble}}
+\label{subsection:scribble}
 
 \begin{itemize}
-\item introduce end-to-end workflow
-\item must have diagram
-\item describe input scribble protocol and flags, output is typescript code
+\item see notes from PLACES paper on scribble
+\item scribble language definition (unless already in background)
 \end{itemize}
+
+\subsection{From \fancyname{Scribble} to EFSM}
+\label{subsection:efsm}
+
+\begin{itemize}
+\item scribble validates the protocol
+\item use flag to obtain EFSM representation
+\item DOT graph output
+\item parse in python -- use open-source library for this
+\end{itemize}
+
+\subsection{Code Generation}
+\label{subsection:apigen}
+
+\begin{itemize}
+\item api generation is a function of the EFSM representation
+\item traditional methods -- visitor pattern on the EFSM and using stringbuilder to build the file
+\item templating library is more suitable to decouple the ``presentation'' (how the code should look) from the ``content'' (the EFSM which is used to build the code)
+\item strategy pattern to support the two required build targets (in node and react)
+\end{itemize}

### 2020-06-01 report/sections/evalconcl/eval.tex
Report: update captionof's
Report: add protocol listings for forward-ref

@@
 \section{Binary Sessions: \fancyname{Calculator}}
+\label{section:evalcalculator}
 \begin{itemize}
@@
 
+\begin{figure}[!ht]
+\begin{lstlisting}[language=Scribble, tabsize=2]
+module BinarySession;
+
+global protocol Calculator(role Svr, role Client) {
+  choice at Client {
+    Add(number, number) from Client to Svr;
+    Res(number)         from Svr to Client;
+    do Calculator(Svr, Client);
+  } or {
+    Mult(number, number) from Client to Svr;
+    Res(number)          from Svr to Client;
+    do Calculator(Svr, Client);
+  } or {
+    Quit()               from Client to Svr;
+    Terminate()          from Svr to Client;
+  }
+}
+\end{lstlisting}
+\captionof{lstlisting}{The \tprotocol{Calculator} Protocol}
+\label{lst:calculator}
+\end{figure}
+
 \section{Multiparty Sessions: \fancyname{Noughts and Crosses}}
+\label{section:evalgame}
 \begin{itemize}
@@
 \end{itemize}
+
+\begin{figure}[!ht]
+\begin{lstlisting}[language=Scribble, tabsize=2]
+module NoughtsAndCrosses;
+type <typescript> "Coordinate" from "./Types" as Point;
+
+global protocol Game(role Svr, role P1, role P2) {
+	Pos(Point) from P1 to Svr;
+	choice at Svr {
+		Lose(Point)   from Svr to P2;
+		Win(Point)    from Svr to P1;
+	} or {
+		Draw(Point)   from Svr to P2;
+		Draw(Point)   from Svr to P1;
+	} or {
+		Update(Point) from Svr to P2;
+		Update(Point) from Svr to P1;
+		do Game(Svr, P2, P1);
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{The \tprotocol{Noughts and Crosses} Protocol}
+\label{lst:game}
+\end{figure}
 
@@
 
+\begin{figure}[!ht]
+\begin{lstlisting}[language=Scribble, tabsize=2]
+module ClientClient;
+
+global protocol TwoBuyer(role A, role B, role S) {
+	title(string)      from A to S;
+	quote(number)      from S to A;
+	quote(number)      from S to B;
+	quoteByTwo(number) from A to B;
+	choice at B {
+		ok(string)     from B to S;
+		date(string)   from S to B;
+	} or {
+		quit()         from B to S;
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{The \tprotocol{Two Buyer} Protocol}
+\label{lst:twobuyer}
+\end{figure}
+
 \input{sections/evalconcl/eval/benchmarks}

### 2020-06-01 report/sections/evalconcl/eval/benchmarks.tex
Report: update captionof's
Report: add protocol listings for forward-ref

@@
 \end{lstlisting}
-\captionof{lstlisting}{Ping Pong Protocol}
+\captionof{lstlisting}{The \tprotocol{Ping Pong} Protocol}
 \label{lst:pingpong}
@@
 
-We parameterise the Ping Pong protocol by $n > 0$, the number of
+We parameterise the \tprotocol{Ping Pong} protocol by $n > 0$, the number of
 round-trip messages. This is standardised in the application logic
@@
 
-We note that the Ping Pong protocol implements a \textit{binary} session. 
+We note that the \tprotocol{Ping Pong} protocol 
+implements a \textit{binary} session. 
 It would be interesting to observe the overhead in a \textit{multiparty}
@@
 \includegraphics[width=0.5\textwidth]{pingpongclient}
-\captionof{figure}{User interface of Ping Pong \trole{Client}}
+\captionof{figure}{User interface of \tprotocol{Ping Pong} \trole{Client}}
 \label{fig:pingpongclient}
@@
 
-\subparagraph{Ping Pong \trole{Client} on React:}
+\subparagraph{\tprotocol{Ping Pong} \trole{Client} on React:}
 \begin{itemize}
@@
 
-\subparagraph{Ping Pong \trole{Svr} on Node:}
+\subparagraph{\tprotocol{Ping Pong} \trole{Svr} on Node:}
 \begin{itemize}
@@
 
-The benchmark compares three implementations of the Ping Pong protocol:
+The benchmark compares three implementations of the \tprotocol{Ping Pong} protocol:
 
@@
 WebSocket primitives for sending and receiving. 
-The implementation executes the Ping Pong protocol, but does not 
+The implementation executes the \tprotocol{Ping Pong} protocol, but does not 
 guarantee communication safety by construction -- e.g. the user can click
@@
 \captionof{lstlisting}{Preventing user-triggered linearity violation in
-\texttt{bare_safe} Ping Pong}
+\texttt{bare_safe} \tprotocol{Ping Pong}}
 \label{lst:workaround}
@@
 
-The workaround is tailored to the Ping Pong protocol
+The workaround is tailored to the \tprotocol{Ping Pong} protocol
 and toggles the \texttt{visible} flag to hide the button that triggers
@@
 \end{lstlisting}
-\caption{\texttt{bare} implementation of Ping Pong \trole{Svr}}
+\caption{\texttt{bare} implementation of \tprotocol{Ping Pong} \trole{Svr}}
 \label{lst:nodepingpongbare}
@@
 \end{lstlisting}
-\caption{\texttt{mpst} implementation of Ping Pong \trole{Svr}}
+\caption{\texttt{mpst} implementation of \tprotocol{Ping Pong} \trole{Svr}}
 \label{lst:nodepingpongmpst}
@@
 \captionof{figure}{Comparison of \texttt{bare} and \texttt{mpst}
-implementations for Ping Pong \trole{Svr}}
+implementations for \tprotocol{Ping Pong} \trole{Svr}}
 \label{fig:nodepingpong}

### 2020-06-01 report/sections/server/codegen/workflow.tex
Report: progress on chapter 3

@@
 \section{Development Workflow}
-
-% input: protocol, role, target
-% then: verify well-formedness
-% then: obtain EFSM for role
-% then: parse EFSM
-% then: generate code for said target
-% output: API
 
@@
 \cite{PureScript2019} by extending the \fancyname{Scribble} toolchain
-and generating APIs that integrate developer implementation
-into the EFSM execution.
+and generating APIs that integrate the developer's 
+application logic
+into the execution of the communication automata.
 
@@
 
-\begin{itemize}
-\item see notes from PLACES paper on scribble
-\item scribble language definition (unless already in background)
-\end{itemize}
+We use the \fancyname{Scribble} protocol description language, 
+as presented in
+\cite{Scribble}, for formalising the communication structure. This is
+inspired by existing work on implementing session type theory 
+in mainstream programming languages
+\cite{Hybrid2016, PureScript2019, Python2017}. 
+We use the variant of the \fancyname{Scribble} language 
+previously introduced in \cref{subsection:bgscribble}.
+
+\subparagraph{Type declaration statements}
+Specific to our TypeScript API generation toolchain,
+the developer is \textit{not} required to explicitly
+add type declaration statements for built-in types.
+\cref{lst:adder} is a \textit{syntactically correct}
+\fancyname{Scribble} protocol as far as 
+\fancyname{SessionTS} is concerned. 
+Internally, \fancyname{SessionTS} inspects the protocol file
+and parses existing type declarations using regular expressions
+(or \textit{regex}) -- this is necessary to extract any
+custom data types that will appear in the communication (for example,
+\cref{lst:game}), and allows \fancyname{SessionTS} to inject
+``boilerplate'' type declarations for built-in TypeScript types before
+calling \fancyname{Scribble}.
+
+\begin{figure}[!ht]
+\begin{lstlisting}[language=Scribble]
+module Adder;
+
+global protocol Adder(role Client, role Svr) {
+	choice at Client {
+		ADD(number, number) from Client to Svr;
+		RES(number)         from Svr to Client;	
+	} or {
+		QUIT(string) from Client to Svr;	
+		TERMINATE()  from Svr to Client;
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{The \tprotocol{Adder} Protocol}
+\label{lst:adder}
+\end{figure}
+
+We will use the \tprotocol{Adder} protocol as a running example
+to demonstrate how our work performs TypeScript API generation.
 

### 2020-06-01 report/sections/server/codegen/workflow.tex
Report: progress on chapter 3

@@
 
-\begin{itemize}
-\item scribble validates the protocol
-\item use flag to obtain EFSM representation
-\item DOT graph output
-\item parse in python -- use open-source library for this
-\end{itemize}
+Given the protocol and endpoint, we use \fancyname{Scribble}
+to validate the well-formedness of the protocol and extract
+information from the protocol relevant for the endpoint.
+The latter is expressed as a finite state machine
+where each state restricts the possible transitions, 
+and transitions between states are represented by
+communication actions, i.e. the sending or receiving of a message.
 
-\subsection{Code Generation}
+\fancyname{Scribble} expresses the EFSM using the 
+DOT graph description language \cite{dot}, with each
+communication action encoded as the label of the corresponding
+state transition. 
+\fancyname{SessionTS} uses the pydot library \cite{pydot}
+to parse the graph into an internal representation of the EFSM.
+We define an \texttt{EfsmBuilder} class with APIs designed for
+constructing the EFSM representation by iterating over the 
+state transitions from the DOT representation.
+
+\begin{figure}[!ht]
+\begin{lstlisting}[language=Python]
+@dataclass
+class Endpoint:
+    protocol: str
+    role    : str
+    server  : str
+    efsm    : EFSM
+    types   : typing.Iterable[DataType]
+\end{lstlisting}
+\captionof{lstlisting}{The Endpoint API}
+\label{lst:endpointapi}
+\end{figure}
+
+As the code generation process requires additional information,
+we define an \texttt{Endpoint} dataclass\footnote{
+A Python dataclass uses the decorator to generate
+``boilerplate'' methods, such as the constructor, based on the
+properties listed in the annotations.} (\cref{lst:endpointapi})
+to contain the \texttt{EFSM}
+representation, along with the information passed in from the
+command line (\texttt{protocol}, \texttt{role}, \texttt{server}) 
+and the custom type declarations (\texttt{types}) parsed from the
+protocol specification.
+
+\subsection{API Generation}
 \label{subsection:apigen}
+
+Formally, API generation is a function of the constructed
+\texttt{Endpoint} instance
+\textit{and} the target specified in the command line. We use a 
+different code generation strategy for implementations running on
+Node.js versus the browser -- we discuss in greater depth in 
+\cref{chap:node} and \cref{chap:react} respectively. 
+In this subsection, we explain how we perform API generation
+at a higher level of abstraction.
+
+Traditional methods of code generation involve applying the
+Visitor pattern on the internal representation. 
+In the context of the MPST framework,
+this may involve defining a Visitor class that implements a
+\texttt{generate()} operation to be performed on the EFSM states,
+such that the \texttt{generate()} implementation specialises to the
+type of EFSM state, i.e. send, receive or terminal.
+This is not straightforward in Python, as method overloading is not 
+supported, so the ``visit'' methods would need different names.
+More importantly, it is less straightforward to visualise
+the structure of the generated code, as the string interpolation
+aspect is likely to be interleaved with source code implementing
+additional logic for code generation.
+
+For \fancyname{SessionTS}, we opt to use a template engine for code
+generation, namely Jinja \cite{jinja}.
+
+\dots %what is a template, 
+\dots %template focuses on how the generated code looks (which makes it easy to debug etc), and uses lightweight markup syntax to inject dynamic content (from our endpoint API)
+
+The main advantage of using a template engine for 
+our code generation task is that it decouples the ``presentation''
+from the ``content'': in the \fancyname{SessionTS}
+% so it is easy to prototype and extend without touching the source code
 

### 2020-06-01 report/sections/introduction/contributions.tex
Report: complete chapter 3 first draft

@@
 code generator, \fancyname{SessionTS}. 
-The implementation integrates with the \fancyname{Scribble}
+The implementation integrates with the Scribble
 toolchain to generate handler style APIs that the developer can use to write 

### 2020-06-01 report/sections/server/codegen/implementation.tex
Report: progress on chapter 3
Report: complete chapter 3 first draft

@@
 making it a suitable choice for implementing our code generator.
-We provide a simple command line interface (\cref{lst:cmdline})
-for the developer to supply the required information 
-to use \fancyname{SessionTS}.
+Its rich standard library also simplifies many tasks: we use
+the \texttt{argparse} package to generate an informative
+command line interface (CLI) for developers to 
+supply the correct information (\cref{lst:cmdline})
+to use \fancyname{SessionTS},
+and \texttt{subprocess} package provides an abstraction to
+invoke external tools, such as Scribble and the TypeScript Compiler.
 
@@
 \end{lstlisting}
-\captionof{figure}{Entry Point for \fancyname{SessionTS}}
+\captionof{lstlisting}{Entry Point for \fancyname{SessionTS}}
 \label{lst:cmdline}
@@
 a standalone Java executable with non-trivial setup procedures.
-The Dockerfile builds a Docker image with \fancyname{Scribble} and
+The Dockerfile builds a Docker image with Scribble and
 the Python dependencies all pre-configured, and the provided

### 2020-06-01 report/sections/server/codegen/testing.tex
Report: complete chapter 3 first draft

@@
 \section{Testing}
-\begin{itemize}
-\item unit tests for code generator executable
-\item systems tests for type checking the generated output with the typescript compiler
-\item sandbox environment set up on the docker container to set up different tsconfig.json files for both react/node targets
-\end{itemize}
+
+The usual unit testing practices from software engineering
+apply to our code generator -- the built-in \texttt{unittest} 
+package is more than sufficient for our use case. 
+Instead, we detail our methodology for \textit{system testing} -- 
+executing \fancyname{SessionTS} end-to-end and testing
+the generated code.
+
+We implement a test suite on top of \texttt{unittest} APIs
+for verifying that \fancyname{SessionTS}
+generates \textit{valid TypeScript code}. This is especially useful as
+our templates contain both TypeScript syntax and Jinja markup, 
+so we cannot easily make sure that each template generates valid code,
+let alone checking that the collection of templates generate a valid
+TypeScript project altogether.
+
+\begin{figure}[!ht]
+\begin{lstlisting}[language=python,tabsize=4]
+def test_code_generation(self):
+	flags = [scr, protocol, role, target]
+	if svr is not None:
+		flags.append('-s')
+		flags.append(svr)
+
+	rc = mpst_ts.main(flags) (*@\label{line:testmain}@*)
+	self.assertEqual(rc, 0)  (*@\label{line:testmainrc}@*)
+
+	completion = subprocess.run(self.npm_test_cmd, shell=True) (*@\label{line:testtsc}@*)
+	self.assertEqual(completion.returncode, 0) (*@\label{line:testtscrc}@*)
+
+	shutil.rmtree(self.output_dir)
+\end{lstlisting}
+\captionof{lstlisting}{Main logic in \fancyname{SessionTS} system testing
+test case}
+\label{lst:systemtest}
+\end{figure}
+
+We provide a collection of Scribble protocols under \texttt{protocols/}
+to generate test cases, one per protocol participant.
+For each test case, the test suite (\cref{lst:systemtest}) will:
+
+\begin{enumerate}
+\item Invoke \fancyname{SessionTS} to generate the TypeScript project 
+(\cref{line:testmain}), 
+expecting a zero exit code (\cref{line:testmainrc});
+\item Run the TypeScript Compiler on the generated directory 
+(\cref{line:testtsc}),
+passing the \texttt{noEmit} flag to purely perform type-checking,
+and expecting a zero exit code (\cref{line:testtscrc}).
+\end{enumerate}
+
+As the generated TypeScript code makes assumptions about the
+environment in which it is used (for example, having the \texttt{ws}
+WebSocket package installed on server-side targets), we require a
+sandbox environment to type-check the generated code. 
+The sandbox contains the minimal boilerplate required
+for testing -- this involves having the WebSocket package installed
+for server-side targets, the React.js framework (\cref{section:react})
+instantiated for browser-side targets,
+and corresponding \texttt{tsconfig.json} files for both targets to
+be picked up by the TypeScript Compiler.
+
+For convenience, we extend the
+Dockerfile and \texttt{build.sh} script 
+to set up the sandbox environments. 
+We also make use of the optional
+\texttt{--output} flag exposed by the \fancyname{SessionTS} CLI
+to redirect the generated code to the correct sandbox environment
+to simply the testing process.

### 2020-06-01 report/sections/server/codegen/workflow.tex
Report: complete chapter 3 first draft

@@
 We motivate our development workflow from previous work 
-\cite{PureScript2019} by extending the \fancyname{Scribble} toolchain
+\cite{PureScript2019} by extending the Scribble toolchain
 and generating APIs that integrate the developer's 
@@
 \item The developer supplies the communication protocol written in
-\fancyname{Scribble} (\cref{subsection:scribble}), 
+Scribble (\cref{subsection:scribble}), 
 stating the role (hereafter \textit{endpoint})
@@
 \item \fancyname{SessionTS} delegates to the 
-\fancyname{Scribble} toolchain for verifying the well-formedness of
+Scribble toolchain for verifying the well-formedness of
 the protocol and expects to receive a DOT graph representation of
@@
 
-\subsection{Protocol Specification with \fancyname{Scribble}}
+\subsection{Protocol Specification with Scribble}
 \label{subsection:scribble}
 
-We use the \fancyname{Scribble} protocol description language, 
+We use the Scribble protocol description language, 
 as presented in
@@
 \cite{Hybrid2016, PureScript2019, Python2017}. 
-We use the variant of the \fancyname{Scribble} language 
+We use the variant of the Scribble language 
 previously introduced in \cref{subsection:bgscribble}.
@@
 \cref{lst:adder} is a \textit{syntactically correct}
-\fancyname{Scribble} protocol as far as 
+Scribble protocol as far as 
 \fancyname{SessionTS} is concerned. 
@@
 ``boilerplate'' type declarations for built-in TypeScript types before
-calling \fancyname{Scribble}.
+calling Scribble.
 
@@
 
-\subsection{From \fancyname{Scribble} to EFSM}
+\subsection{From Scribble to EFSM}
 \label{subsection:efsm}
 
-Given the protocol and endpoint, we use \fancyname{Scribble}
+Given the protocol and endpoint, we use Scribble
 to validate the well-formedness of the protocol and extract
@@
 
-\fancyname{Scribble} expresses the EFSM using the 
+Scribble expresses the EFSM using the 
 DOT graph description language \cite{dot}, with each
@@
 different code generation strategy for implementations running on
-Node.js versus the browser -- we discuss in greater depth in 
-\cref{chap:node} and \cref{chap:react} respectively. 
+the servre (\cref{chap:node}) versus the browser (\cref{chap:react}).
 In this subsection, we explain how we perform API generation
@@
 
-For \fancyname{SessionTS}, we opt to use a template engine for code
-generation, namely Jinja \cite{jinja}.
+For \fancyname{SessionTS}, we leverage the Jinja \cite{jinja} 
+template engine library for code generation. 
+We first construct templates for the TypeScript files we wish to generate,
+specifying placeholders for dynamic content (to be extracted
+from the \texttt{Endpoint} object); 
+we then provide Jinja with the template path and the 
+\texttt{Endpoint} object, and the template engine renders the
+TypeScript code by filling in the dynamic placeholders. 
+We show an example in \cref{lst:jinja}.
 
-\dots %what is a template, 
-\dots %template focuses on how the generated code looks (which makes it easy to debug etc), and uses lightweight markup syntax to inject dynamic content (from our endpoint API)
+\begin{figure}[!ht]
+\begin{lstlisting}[language=javascript, title=efsm.ts.j2]
+export namespace Message {
+{% for state in endpoint.efsm.nonterminal_states %}
+  {% for action in state.actions -%}
+  export type S{{ state ~ action.label }} = {
+    label: Labels.S{{ state }}.{{ action.label }},
+    payload: [{{ action.payloads|join(', ') }}],
+  };
+  {% endfor %}
+  export type S{{ state }} = {% for action in state.actions -%}
+  | S{{ state ~ action.label }}{% endfor %};
+{% endfor %}
+\end{lstlisting}
+\captionof{lstlisting}{Example Jinja Template for
+\fancyname{SessionTS} API Generation}
+\label{lst:jinja}
+\end{figure}
 
-The main advantage of using a template engine for 
-our code generation task is that it decouples the ``presentation''
-from the ``content'': in the \fancyname{SessionTS}
-% so it is easy to prototype and extend without touching the source code
+Jinja provides lightweight syntax for injecting content and
+markup for simple control structure: \texttt{\{\{ state \}\}} 
+denotes a placeholder for Jinja to render the \texttt{state} variable,
+and the \texttt{\{\% \%\}} syntax  is used for conditionals and
+control structures 
+(such as for loops, to dynamically render the enclosing ``sub-template''
+by iterating over a collection).
+The main advantage that Jinja brings is that 
+it decouples the ``presentation''
+from the ``content'' and makes it quick to prototype and extend
+the generated code, \textit{usually} without modifications to the 
+code generator.
 
-\begin{itemize}
-\item api generation is a function of the EFSM representation
-\item traditional methods -- visitor pattern on the EFSM and using stringbuilder to build the file
-\item templating library is more suitable to decouple the ``presentation'' (how the code should look) from the ``content'' (the EFSM which is used to build the code)
-\item strategy pattern to support the two required build targets (in node and react)
-\end{itemize}
+\begin{figure}[!ht]
+\begin{lstlisting}[language=python]
+class CodeGenerationStrategy(ABC):
+
+    target_to_strategy = {} (*@\label{line:mapping}@*)
+
+    def __init__(self):
+        super().__init__()
+
+    @classmethod
+    def __init_subclass__(cls, *, target):
+        CodeGenerationStrategy.target_to_strategy[target] = cls (*@\label{line:register}@*)
+        return super().__init_subclass__()
+
+    @abstractmethod
+    def generate(self, endpoint: Endpoint):
+        pass
+\end{lstlisting}
+\captionof{lstlisting}{Implementing \texttt{CodeGenerationStrategy}}
+\label{lst:codegenerationstrategy}
+\end{figure}
+
+As we generate a different set of TypeScript artefacts depending
+on the specified target, we structure the different code generators
+using the Strategy design pattern. Each target extends the
+abstract base class \texttt{CodeGenerationStrategy} 
+and implements its own \texttt{generate()}
+method to return a list of \texttt{(path, content)} tuples. 
+We define a \texttt{CodeGenerator} class that is parameterised 
+by \texttt{target}: when instantiated, it will select and perform 
+the specialised \texttt{generate()} method based on \texttt{target}, 
+before formatting and committing the generated code 
+to the file system.
+We implement a subclass hook in \texttt{CodeGenerationStrategy}
+(\cref{lst:codegenerationstrategy}),
+such that each derived class must provide the target name to
+``register'' with the base class (\cref{line:register}),
+and the base class keeps an internal
+mapping of the concrete strategies (\cref{line:mapping}); 
+\texttt{CodeGenerator} accesses
+this mapping to select the appropriate strategy.

### 2020-06-01 report/sections/evalconcl/eval.tex
Report: start chapter 7

@@
 
-\begin{figure}[!ht]
-\begin{lstlisting}[language=Scribble, tabsize=2]
-module ClientClient;
-
-global protocol TwoBuyer(role A, role B, role S) {
-	title(string)      from A to S;
-	quote(number)      from S to A;
-	quote(number)      from S to B;
-	quoteByTwo(number) from A to B;
-	choice at B {
-		ok(string)     from B to S;
-		date(string)   from S to B;
-	} or {
-		quit()         from B to S;
-	}
-}
-\end{lstlisting}
-\captionof{lstlisting}{The \tprotocol{Two Buyer} Protocol}
-\label{lst:twobuyer}
-\end{figure}
-
 \input{sections/evalconcl/eval/benchmarks}

### 2020-06-01 report/sections/general/p2p.tex
Report: notes on part 2
Report: start chapter 7

@@
 \chapter{Motivation: Supporting Peer-to-Peer Interactions}
+\label{chap:p2p}
 
 \section{\fancyname{Two Buyer} Protocol}
+
+We present the \tprotocol{Two Buyer} protocol in \cref{lst:twobuyer}.
+
+\begin{figure}[!ht]
+\begin{lstlisting}[language=Scribble, tabsize=2]
+module ClientClient;
+
+global protocol TwoBuyer(role A, role B, role S) {
+	title(string)      from A to S;
+	quote(number)      from S to A;
+	quote(number)      from S to B;
+	quoteByTwo(number) from A to B;
+	choice at B {
+		ok()           from B to S;
+		date(string)   from S to B;
+	} or {
+		quit()         from B to S;
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{The \tprotocol{Two Buyer} Protocol}
+\label{lst:twobuyer}
+\end{figure}
+
+Assume \trole{S} is the server, and roles \trole{A}, \trole{B} run on the 
+browser:
+
+\begin{enumerate}
+\item \trole{A} sends to \trole{S} the title of an item to sell;
+\item \trole{S} sends a quote for said item to \trole{A} and \trole{B};
+\item \trole{A} privately sends a separate quote to \trole{B};
+\item \trole{B} makes a choice:
+\begin{itemize}
+\item if \trole{B} prefers the offer from \trole{S}, 
+\trole{B} replies \tmsg{ok()} to \trole{S}, and \trole{S} informs
+\trole{B} of the delivery date;
+
+\item otherwise, \trole{B} takes the offer from \trole{A} and simply
+responds \tmsg{quit()} to \trole{S}.
+\end{itemize} 
+\end{enumerate}
+
 \begin{itemize}

### 2020-06-02 report/sections/server/codegen/workflow.tex
Report: minor changes  to background and workflow

@@
 We motivate our development workflow from previous work 
-\cite{PureScript2019} by extending the Scribble toolchain
+\cite{PureScript2019, Hybrid2016} by extending the Scribble toolchain
 and generating APIs that integrate the developer's 

### 2020-06-02 report/sections/general/p2p.tex
Report: progress on chapter 7

--- 2020-05-27 report/sections/general/p2p.tex
@@
 \chapter{Motivation: Supporting Peer-to-Peer Interactions}
+\label{chap:p2p}
 
-\section{\fancyname{Two Buyer} Protocol}
-\begin{itemize}
-\item TwoBuyer example
-\item Not possible with WebSocket topology
-\end{itemize}
+In this chapter,
+we demonstrate how to support peer-to-peer interactions
+between web browser endpoints over WebSocket transport,
+thus relaxing the server-centric topology assumption from 
+\cref{part:server}.
 
-\section{Server as a Router}
-\begin{itemize}
-\item How to make TwoBuyer example work just with WebSockets
-\item Essentially `decomposing' multiparty session into binary interactions with respect to the server
-\end{itemize}
+We motivate the problem using the 
+\tprotocol{Two Buyer} protocol (\cref{section:twobuyer}),
+outline our approach (\cref{section:routerapproach}),
+and discuss the challenges we need to address 
+(\cref{section:routerchallenges}) in order to validate the
+correctness of our approach.
+
+\section{\tprotocol{Two Buyer} Protocol}
+\label{section:twobuyer}
+
+We present the \tprotocol{Two Buyer} protocol in \cref{lst:twobuyer},
+which describes a protocol between two buyers \trole{A}, \trole{B}
+and a \trole{S}eller. We assume that the \trole{S}eller runs on
+the server, and the buyers run on the browser.
+
+\begin{figure}[!ht]
+\begin{lstlisting}[language=scribble]
+module ClientClient;
+
+global protocol TwoBuyer(role A, role B, role S) {
+	title(string)   from A to S;
+	quote(number)   from S to A;
+	quote(number)   from S to B;
+	split(number)   from A to B; (*@\label{line:twobuyersplit}@*)
+	choice at B {
+		accept()    from B to A; (*@\label{line:twobuyerok}@*)
+		buy()       from A to S;
+	} or {
+		reject()    from B to A; (*@\label{line:twobuyerno}@*)
+		cancel()    from A to S;
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{The \tprotocol{Two Buyer} Protocol}
+\label{lst:twobuyer}
+\end{figure}
+
+In the protocol, \trole{A} asks the \trole{S}eller for the quote
+of a book \tmsg{title}. \trole{S}eller sends the \tmsg{quote}
+to both buyers \trole{A} and \trole{B}. \trole{A} privately asks
+proposes a \tmsg{split} with \trole{B}. Now, \trole{B} must make
+a choice: 
+either \trole{B} can \tmsg{accept} the split, where \trole{A} confirms
+to \tmsg{buy} from the \trole{S}eller;
+otherwise, \trole{B} will \tmsg{reject} the split, which
+\tmsg{cancel}s the purchase.
+
+\cref{line:twobuyersplit,line:twobuyerok,line:twobuyerno}
+illustrate that the \tprotocol{Two Buyer} protocol 
+does not implement a server-centric communication topology,
+since these lines reference communication actions between
+the two client browsers. 
+As a result, \fancyname{SessionTS} cannot correctly generate
+APIs that support the specified interactions 
+for buyers \trole{A} and \trole{B}.
+
+\section{Proposal: Server as a Router}
+\label{section:routerapproach}
+
+We acknowledge that peer-to-peer browser interactions
+could be achieved via different transport abstractions, 
+such as the WebRTC framework \cite{WebRTC}. 
+However, WebRTC has its own proprietary ``protocol'' 
+for establishing connections and exchanging network information,
+which isn't formalised in session type theory. This setup phase
+generally requires a \textit{signalling server} \cite{WebRTCSignaling},
+which brings our problem full circle, as it can be implemented
+over WebSocket transport. Extending API generation to also include
+the boilerplate for initialising WebRTC connections could be an option,
+but would introduce complications in the session runtime and
+error handling (e.g. how to handle cancellations during the connection
+phase?).
+We believe that adopting WebRTC deviates from the focus of project in 
+implementing type-safe webservices.
+
+Hence, for the purpose of our work in this part,
+we establish that WebSocket transport is an invariant, 
+and proceed to build up our approach. 
+We start by breaking down the concept of 
+``peer-to-peer communication''.
+Focusing on \cref{lst:twobuyer} \cref{line:twobuyersplit},
+we can say that the interaction satisfies three properties:
+
+\begin{enumerate}
+\item 
+As far as \trole{A} is concerned, \trole{A} is sending
+a message that will be received by \trole{B}.
+
+\item
+As far as \trole{B} is concerned, \trole{B} is receiving
+a message that was sent by \trole{A}.
+
+\item
+The \trole{S}eller is not involved in this interaction.
+\end{enumerate}
+
+We proceed to relax point 3: for the \trole{S}eller to be
+not involved in the interaction, it suffices to guarantee that
+the \trole{S}eller cannot intervene or hijack the communication.
+This allows \trole{S}eller to ``oversee'' the communication.
+
+On this basis, we motivate our approach to empower the 
+server endpoint to act as a \textit{router}. 
+That is to say, if the server endpoint receives a message
+of which it isn't the intended recipient for, 
+it routes the message to the intended recipient. 
+Based on the WebSocket transport invariant,
+all browser endpoints must join the session
+through the server. This means that the server has information about
+all browser endpoints, 
+and is capable to perform the routing.
+
+Returning to the \tprotocol{Two Buyer} example,
+\trole{A} sends the \tmsg{split} message through its existing
+WebSocket connection (with \trole{S}eller on the other side),
+noting that \trole{B} is the intended recipient;
+\trole{S}eller receives this through the WebSocket bound to
+\trole{A}, and routes the message to \trole{B}, who receives it through
+its existing connection with \trole{S}eller.
+It is trivial that the three properties listed previously are still
+satisfied: considering the relaxed version of point 3,
+the semantics of routing a message corresponds to that of
+overseeing the communication, in the sense that \trole{S}eller 
+does not tamper with the payload or drop the message, so it is simply
+overseeing the message being passed through.
 
 \section{Challenges}
+\label{section:routerchallenges}
+
+% routing needs to be treated very carefully in the theory,
+% we want to preserve communication, so the routed constructs must not % overserialise
+% example: A->B: M1; S->B: M2.
+% a valid trace is SB!M2, AB!M1;
+% 
+% if we treat routing messages with the same constructs,
+% we get A->S: M1; S->B:M1; S->B:M2;
+% then S
+%}
+
 \begin{itemize}

### 2020-06-02 report/sections/general/p2p.tex
Report: complete chapter 7  first draft

@@
 
-Returning to the \tprotocol{Two Buyer} example,
+Returning to \cref{lst:twobuyer} \cref{line:twobuyersplit},
 \trole{A} sends the \tmsg{split} message through its existing
@@
 its existing connection with \trole{S}eller.
-It is trivial that the three properties listed previously are still
-satisfied: considering the relaxed version of point 3,
-the semantics of routing a message corresponds to that of
-overseeing the communication, in the sense that \trole{S}eller 
-does not tamper with the payload or drop the message, so it is simply
-overseeing the message being passed through.
+It is trivial that the three properties\footnote{Here, we use the relaxed
+version of point 3 mentioned previously.} listed previously are still
+satisfied:
+
+\begin{enumerate}
+\item
+\textit{As far as \trole{A} is concerned, \trole{A} is sending
+a message that will be received by \trole{B}} -- 
+the intended recipient is specified in the message.
+
+\item
+\textit{As far as \trole{B} is concerned, \trole{B} is receiving
+a message that was sent by \trole{A}} --
+the original sender is specified in the message.
+
+\item
+\textit{\trole{S}eller cannot intervene or hijack the communication} --
+this remains true, as \trole{S}eller simply receives the message
+from the WebSocket bound to \trole{A} and sends it through the
+WebSocket bound to \trole{B}.
+
+\end{enumerate}
 
@@
 
-% routing needs to be treated very carefully in the theory,
-% we want to preserve communication, so the routed constructs must not % overserialise
-% example: A->B: M1; S->B: M2.
-% a valid trace is SB!M2, AB!M1;
-% 
-% if we treat routing messages with the same constructs,
-% we get A->S: M1; S->B:M1; S->B:M2;
-% then S
-%}
+We need to formalise our concept of routing by extending session
+type theory, such that, given a communication protocol implementing 
+an arbitrary topology, we can encode it into our extended theory and
+prove that safety properties (e.g. well-formedness, deadlock freedom)
+are preserved, and that communication traces are preserved.
+This must be done \textit{very carefully}, as
+naive definitions risk over-serialising the original communication,
+thereby not preserving all communication traces.
 
-\begin{itemize}
-\item Routed communication is different from `normal' communication
-- server cannot overserialise interactions
-\item Existing work in decomposing multiparty sessions (see LinearDecompScala paper and the binary one it references) assume the `server' takes on the pure role as a router -- our work argues that the server can still participate in the multiparty session and concurrently route messages between clients in a way that does not overserialise the interactions and allow the routing to be transparent to other roles.
-\item theory side -- need to prove that it is indeed possible for the server to carry out its own interactions and perform routing in a transparent way, as if the messages were directly peer to peer
-\item implementation side -- changes need to reflect the theory and be as transparent to the developer as possible
-\end{itemize}
+For example, consider the global type
+\[
+A \to B: \tmsg{M1}. S \to B: \tmsg{M2}. \texttt{end}
+\]
+By existing LTS semantics over
+global types (\cref{subsubsection:lts}), the sequence 
+$[SB!\tmsg{M2}, AB!\tmsg{M1}]$
+is a prefix of a valid execution trace.
+If we define our routing construct simply by replacing every single
+$A \to B: \tmsg{M}$ interaction with $A \to S: \tmsg{M}.
+S \to B: \tmsg{M}$, we transform the global type into
+\[
+A \to S: \tmsg{M1}. S \to B: \tmsg{M1}. S \to B: \tmsg{M2}. \texttt{end}
+\]
+our naive definition of routing has over-serialised the protocol
+since $SB!\tmsg{M2}$ can no longer occur before $AB!\tmsg{M1}$.
+This implies that we need to handle routing interactions differently
+from ``normal'' send/receive actions.
+
+Existing work in decomposing multiparty sessions into a collection
+of binary sessions do resemble our routing proposal, but their approach
+involves having an endpoint to exclusively carry out
+routing responsibilities and act as a centralised orchestrator 
+for the multiparty session.
+Our proposal recognises that, in the context of web applications,
+there is a need for server-side endpoints to implement business logic
+so enforcing the server to be an exclusive orchestrator makes our work
+incompatible with webservice protocols; 
+we also argue that the server can still participate in the 
+multiparty session and concurrently perform its routing duties
+in a way that is transparent to non-server endpoints, and crucially,
+does not overserialise the permitted interactions. 
+We introduce our extension as \textit{Routed Session Types},
+and proceed to prove these claims in \cref{chap:theory}.
+
+Likewise, the implementation of routed session types 
+in \fancyname{SessionTS} needs to accurately reflect the theory
+and be as transparent to the developer as possible to minimise
+the learning curve overhead demanded 
+by complicating the generated APIs -- we document our extensions
+to \fancyname{SessionTS} in \cref{chap:impl}.

### 2020-06-02 report/sections/general/p2p.tex
Report: minor style change to chapter 7

@@
 
-For example, consider the global type
+\begin{example}[Naive definition of routing in global types]
+Consider the global type
 \[
@@
 By existing LTS semantics over
-global types (\cref{subsubsection:lts}), the sequence 
-$[SB!\tmsg{M2}, AB!\tmsg{M1}]$
+global types (\cref{subsubsection:lts}), then $SB!\tmsg{M2}$
 is a prefix of a valid execution trace.
@@
 our naive definition of routing has over-serialised the protocol
-since $SB!\tmsg{M2}$ can no longer occur before $AB!\tmsg{M1}$.
+since $SB!\tmsg{M2}$ can no longer occur first because \trole{S}
+is forced to send \tmsg{M1} first, which it cannot do so until receiving
+from \trole{A}.
 This implies that we need to handle routing interactions differently
 from ``normal'' send/receive actions.
+\end{example}
 

### 2020-06-02 report/sections/general/p2p.tex
Report: restructure chapter 7

--- 2020-06-02 report/sections/general/p2p.tex
@@
 
-\section{\tprotocol{Two Buyer} Protocol}
-\label{section:twobuyer}
+\input{sections/general/p2p/twobuyer}
 
-We present the \tprotocol{Two Buyer} protocol in \cref{lst:twobuyer},
-which describes a protocol between two buyers \trole{A}, \trole{B}
-and a \trole{S}eller. We assume that the \trole{S}eller runs on
-the server, and the buyers run on the browser.
+\input{sections/general/p2p/approach}
 
-\begin{figure}[!ht]
-\begin{lstlisting}[language=scribble]
-module ClientClient;
-
-global protocol TwoBuyer(role A, role B, role S) {
-	title(string)   from A to S;
-	quote(number)   from S to A;
-	quote(number)   from S to B;
-	split(number)   from A to B; (*@\label{line:twobuyersplit}@*)
-	choice at B {
-		accept()    from B to A; (*@\label{line:twobuyerok}@*)
-		buy()       from A to S;
-	} or {
-		reject()    from B to A; (*@\label{line:twobuyerno}@*)
-		cancel()    from A to S;
-	}
-}
-\end{lstlisting}
-\captionof{lstlisting}{The \tprotocol{Two Buyer} Protocol}
-\label{lst:twobuyer}
-\end{figure}
-
-In the protocol, \trole{A} asks the \trole{S}eller for the quote
-of a book \tmsg{title}. \trole{S}eller sends the \tmsg{quote}
-to both buyers \trole{A} and \trole{B}. \trole{A} privately asks
-proposes a \tmsg{split} with \trole{B}. Now, \trole{B} must make
-a choice: 
-either \trole{B} can \tmsg{accept} the split, where \trole{A} confirms
-to \tmsg{buy} from the \trole{S}eller;
-otherwise, \trole{B} will \tmsg{reject} the split, which
-\tmsg{cancel}s the purchase.
-
-\cref{line:twobuyersplit,line:twobuyerok,line:twobuyerno}
-illustrate that the \tprotocol{Two Buyer} protocol 
-does not implement a server-centric communication topology,
-since these lines reference communication actions between
-the two client browsers. 
-As a result, \fancyname{SessionTS} cannot correctly generate
-APIs that support the specified interactions 
-for buyers \trole{A} and \trole{B}.
-
-\section{Proposal: Server as a Router}
-\label{section:routerapproach}
-
-We acknowledge that peer-to-peer browser interactions
-could be achieved via different transport abstractions, 
-such as the WebRTC framework \cite{WebRTC}. 
-However, WebRTC has its own proprietary ``protocol'' 
-for establishing connections and exchanging network information,
-which isn't formalised in session type theory. This setup phase
-generally requires a \textit{signalling server} \cite{WebRTCSignaling},
-which brings our problem full circle, as it can be implemented
-over WebSocket transport. Extending API generation to also include
-the boilerplate for initialising WebRTC connections could be an option,
-but would introduce complications in the session runtime and
-error handling (e.g. how to handle cancellations during the connection
-phase?).
-We believe that adopting WebRTC deviates from the focus of project in 
-implementing type-safe webservices.
-
-Hence, for the purpose of our work in this part,
-we establish that WebSocket transport is an invariant, 
-and proceed to build up our approach. 
-We start by breaking down the concept of 
-``peer-to-peer communication''.
-Focusing on \cref{lst:twobuyer} \cref{line:twobuyersplit},
-we can say that the interaction satisfies three properties:
-
-\begin{enumerate}
-\item 
-As far as \trole{A} is concerned, \trole{A} is sending
-a message that will be received by \trole{B}.
-
-\item
-As far as \trole{B} is concerned, \trole{B} is receiving
-a message that was sent by \trole{A}.
-
-\item
-The \trole{S}eller is not involved in this interaction.
-\end{enumerate}
-
-We proceed to relax point 3: for the \trole{S}eller to be
-not involved in the interaction, it suffices to guarantee that
-the \trole{S}eller cannot intervene or hijack the communication.
-This allows \trole{S}eller to ``oversee'' the communication.
-
-On this basis, we motivate our approach to empower the 
-server endpoint to act as a \textit{router}. 
-That is to say, if the server endpoint receives a message
-of which it isn't the intended recipient for, 
-it routes the message to the intended recipient. 
-Based on the WebSocket transport invariant,
-all browser endpoints must join the session
-through the server. This means that the server has information about
-all browser endpoints, 
-and is capable to perform the routing.
-
-Returning to the \tprotocol{Two Buyer} example,
-\trole{A} sends the \tmsg{split} message through its existing
-WebSocket connection (with \trole{S}eller on the other side),
-noting that \trole{B} is the intended recipient;
-\trole{S}eller receives this through the WebSocket bound to
-\trole{A}, and routes the message to \trole{B}, who receives it through
-its existing connection with \trole{S}eller.
-It is trivial that the three properties listed previously are still
-satisfied: considering the relaxed version of point 3,
-the semantics of routing a message corresponds to that of
-overseeing the communication, in the sense that \trole{S}eller 
-does not tamper with the payload or drop the message, so it is simply
-overseeing the message being passed through.
-
-\section{Challenges}
-\label{section:routerchallenges}
-
-% routing needs to be treated very carefully in the theory,
-% we want to preserve communication, so the routed constructs must not % overserialise
-% example: A->B: M1; S->B: M2.
-% a valid trace is SB!M2, AB!M1;
-% 
-% if we treat routing messages with the same constructs,
-% we get A->S: M1; S->B:M1; S->B:M2;
-% then S
-%}
-
-\begin{itemize}
-\item Routed communication is different from `normal' communication
-- server cannot overserialise interactions
-\item Existing work in decomposing multiparty sessions (see LinearDecompScala paper and the binary one it references) assume the `server' takes on the pure role as a router -- our work argues that the server can still participate in the multiparty session and concurrently route messages between clients in a way that does not overserialise the interactions and allow the routing to be transparent to other roles.
-\item theory side -- need to prove that it is indeed possible for the server to carry out its own interactions and perform routing in a transparent way, as if the messages were directly peer to peer
-\item implementation side -- changes need to reflect the theory and be as transparent to the developer as possible
-\end{itemize}
+\input{sections/general/p2p/challenges}

### 2020-06-02 report/sections/general/p2p/approach.tex
Report: restructure chapter 7

\section{Proposal: Server as a Router}
\label{section:routerapproach}

We acknowledge that peer-to-peer browser interactions
could be achieved via different transport abstractions, 
such as the WebRTC framework \cite{WebRTC}. 
However, WebRTC has its own proprietary ``protocol'' 
for establishing connections and exchanging network information,
which isn't formalised in session type theory. This setup phase
generally requires a \textit{signalling server} \cite{WebRTCSignaling},
which brings our problem full circle, as it can be implemented
over WebSocket transport. Extending API generation to also include
the boilerplate for initialising WebRTC connections could be an option,
but would introduce complications in the session runtime and
error handling (e.g. how to handle cancellations during the connection
phase?).
We believe that adopting WebRTC deviates from the focus of project in 
implementing type-safe webservices.

Hence, for the purpose of our work in this part,
we establish that WebSocket transport is an invariant, 
and proceed to build up our approach. 
We start by breaking down the concept of 
``peer-to-peer communication''.
Focusing on \cref{lst:twobuyer} \cref{line:twobuyersplit},
we can say that the interaction satisfies three properties:

\begin{enumerate}
\item 
As far as \trole{A} is concerned, \trole{A} is sending
a message that will be received by \trole{B}.

\item
As far as \trole{B} is concerned, \trole{B} is receiving
a message that was sent by \trole{A}.

\item
The \trole{S}eller is not involved in this interaction.
\end{enumerate}

We proceed to relax point 3: for the \trole{S}eller to be
not involved in the interaction, it suffices to guarantee that
the \trole{S}eller cannot intervene or hijack the communication.
This allows \trole{S}eller to ``oversee'' the communication.

On this basis, we motivate our approach to empower the 
server endpoint to act as a \textit{router}. 
That is to say, if the server endpoint receives a message
of which it isn't the intended recipient for, 
it routes the message to the intended recipient. 
Based on the WebSocket transport invariant,
all browser endpoints must join the session
through the server. This means that the server has information about
all browser endpoints, 
and is capable to perform the routing.

Returning to \cref{lst:twobuyer} \cref{line:twobuyersplit},
\trole{A} sends the \tmsg{split} message through its existing
WebSocket connection (with \trole{S}eller on the other side),
noting that \trole{B} is the intended recipient;
\trole{S}eller receives this through the WebSocket bound to
\trole{A}, and routes the message to \trole{B}, who receives it through
its existing connection with \trole{S}eller.
It is trivial that the three properties\footnote{Here, we use the relaxed
version of point 3 mentioned previously.} listed previously are still
satisfied:

\begin{enumerate}
\item
\textit{As far as \trole{A} is concerned, \trole{A} is sending
a message that will be received by \trole{B}} -- 
the intended recipient is specified in the message.

\item
\textit{As far as \trole{B} is concerned, \trole{B} is receiving
a message that was sent by \trole{A}} --
the original sender is specified in the message.

\item
\textit{\trole{S}eller cannot intervene or hijack the communication} --
this remains true, as \trole{S}eller simply receives the message
from the WebSocket bound to \trole{A} and sends it through the
WebSocket bound to \trole{B}.

\end{enumerate}

### 2020-06-02 report/sections/general/p2p/challenges.tex
Report: restructure chapter 7

\section{Challenges}
\label{section:routerchallenges}

We need to formalise our concept of routing by extending session
type theory, such that, given a communication protocol implementing 
an arbitrary topology, we can encode it into our extended theory and
prove that safety properties (e.g. well-formedness, deadlock freedom)
are preserved, and that communication traces are preserved.
This must be done \textit{very carefully}, as
naive definitions risk over-serialising the original communication,
thereby not preserving all communication traces.

\begin{example}[Naive definition of routing in global types]
Consider the global type
\[
A \to B: \tmsg{M1}. S \to B: \tmsg{M2}. \texttt{end}
\]
By existing LTS semantics over
global types (\cref{subsubsection:lts}), then $SB!\tmsg{M2}$
is a prefix of a valid execution trace.
If we define our routing construct simply by replacing every single
$A \to B: \tmsg{M}$ interaction with $A \to S: \tmsg{M}.
S \to B: \tmsg{M}$, we transform the global type into
\[
A \to S: \tmsg{M1}. S \to B: \tmsg{M1}. S \to B: \tmsg{M2}. \texttt{end}
\]
our naive definition of routing has over-serialised the protocol
since $SB!\tmsg{M2}$ can no longer occur first because \trole{S}
is forced to send \tmsg{M1} first, which it cannot do so until receiving
from \trole{A}.
This implies that we need to handle routing interactions differently
from ``normal'' send/receive actions.
\end{example}

Existing work in decomposing multiparty sessions into a collection
of binary sessions do resemble our routing proposal, but their approach
involves having an endpoint to exclusively carry out
routing responsibilities and act as a centralised orchestrator 
for the multiparty session.
Our proposal recognises that, in the context of web applications,
there is a need for server-side endpoints to implement business logic
so enforcing the server to be an exclusive orchestrator makes our work
incompatible with webservice protocols; 
we also argue that the server can still participate in the 
multiparty session and concurrently perform its routing duties
in a way that is transparent to non-server endpoints, and crucially,
does not overserialise the permitted interactions. 
We introduce our extension as \textit{Routed Session Types},
and proceed to prove these claims in \cref{chap:theory}.

Likewise, the implementation of routed session types 
in \fancyname{SessionTS} needs to accurately reflect the theory
and be as transparent to the developer as possible to minimise
the learning curve overhead demanded 
by complicating the generated APIs -- we document our extensions
to \fancyname{SessionTS} in \cref{chap:impl}.

### 2020-06-02 report/sections/general/p2p/twobuyer.tex
Report: restructure chapter 7

\section{\tprotocol{Two Buyer} Protocol}
\label{section:twobuyer}

We present the \tprotocol{Two Buyer} protocol in \cref{lst:twobuyer},
which describes a protocol between two buyers \trole{A}, \trole{B}
and a \trole{S}eller. We assume that the \trole{S}eller runs on
the server, and the buyers run on the browser.

\begin{figure}[!ht]
\begin{lstlisting}[language=scribble]
module ClientClient;

global protocol TwoBuyer(role A, role B, role S) {
	title(string)   from A to S;
	quote(number)   from S to A;
	quote(number)   from S to B;
	split(number)   from A to B; (*@\label{line:twobuyersplit}@*)
	choice at B {
		accept()    from B to A; (*@\label{line:twobuyerok}@*)
		buy()       from A to S;
	} or {
		reject()    from B to A; (*@\label{line:twobuyerno}@*)
		cancel()    from A to S;
	}
}
\end{lstlisting}
\captionof{lstlisting}{The \tprotocol{Two Buyer} Protocol}
\label{lst:twobuyer}
\end{figure}

In the protocol, \trole{A} asks the \trole{S}eller for the quote
of a book \tmsg{title}. \trole{S}eller sends the \tmsg{quote}
to both buyers \trole{A} and \trole{B}. \trole{A} privately asks
proposes a \tmsg{split} with \trole{B}. Now, \trole{B} must make
a choice: 
either \trole{B} can \tmsg{accept} the split, where \trole{A} confirms
to \tmsg{buy} from the \trole{S}eller;
otherwise, \trole{B} will \tmsg{reject} the split, which
\tmsg{cancel}s the purchase.

\cref{line:twobuyersplit,line:twobuyerok,line:twobuyerno}
illustrate that the \tprotocol{Two Buyer} protocol 
does not implement a server-centric communication topology,
since these lines reference communication actions between
the two client browsers. 
As a result, \fancyname{SessionTS} cannot correctly generate
APIs that support the specified interactions 
for buyers \trole{A} and \trole{B}.

### 2020-06-02 report/sections/server/node.tex
Report: notes on part 1
Report: add figures to Node chapter

@@
 \chapter{\fancyname{NodeMPST}: Back-End Session Type Web Development}
+\label{chap:node}
+
+In this chapter, 
+we \dots
 
@@
 \begin{itemize}
-\item callbacks => type aliases for each state (handler)
+\item callbacks -- type aliases for each state (handler)
 \item need a way for the runtime to distinguish between states
 \item attempt at using conditional types, but because they handle conditional union types in a distributed way (give example), cannot be statically typed
-\item alternative => use discriminated union by wrapping in an Implementation class
+\item alternative -- use discriminated union by wrapping in an Implementation class
 \item all non-terminal states will `return' the successor implementation -- very difficult to resolve in the type-checker in the runtime (give example of what it may look like) -- solve by giving each state implementation the `advance' function to respect the event-driven nature of everything, so they can call it on completion
@@
 
+\begin{figure}[!ht]
+\centering
+\includegraphics[width=0.5\textwidth]{NodeRuntimeEFSM}
+\captionof{figure}{Informal Abstraction of EFSM Execution for
+Server Endpoints}
+\label{fig:noderuntimeefsm}
+\end{figure}
+
 \subsection{Handling Message Sends}
@@
 \end{itemize}
+
+\begin{figure}[!ht]
+\centering
+\begin{subfigure}[b]{0.8\textwidth}
+\centering
+\includegraphics[width=\textwidth]{NodeRuntimeReceive2}
+\caption{Message processed before transitioning to receive state}
+\label{subfig:nodereceivemsgfirst}
+\end{subfigure}
+\hfill
+\begin{subfigure}[b]{0.8\textwidth}
+\centering
+\includegraphics[width=\textwidth]{NodeRuntimeReceive1}
+\caption{Message processed after transitioning to receive state}
+\label{subfig:nodereceivehandlefirst}
+\end{subfigure}
+\captionof{figure}{Possible Orderings for Handling Message Event and Preparing
+Receive State}
+\label{fig:nodereceivecompare}
+\end{figure}
 

### 2020-06-02 report/sections/server/codegen/testing.tex
Report: make wording more consistent

@@
 environment in which it is used (for example, having the \texttt{ws}
-WebSocket package installed on server-side targets), we require a
+WebSocket package installed on server-side endpoints), we require a
 sandbox environment to type-check the generated code. 
@@
 for testing -- this involves having the WebSocket package installed
-for server-side targets, the React.js framework (\cref{section:react})
-instantiated for browser-side targets,
+for server-side endpoints, the React.js framework (\cref{section:react})
+instantiated for browser-side endpoints,
 and corresponding \texttt{tsconfig.json} files for both targets to

### 2020-06-02 report/sections/server/codegen/workflow.tex
Report: make wording more consistent

@@
 different code generation strategy for implementations running on
-the servre (\cref{chap:node}) versus the browser (\cref{chap:react}).
+the server (\cref{chap:node}) versus the browser (\cref{chap:react}),
+hereafter referred to as \textit{server-side endpoints} and 
+\textit{browser-side endpoints} respectively.
 In this subsection, we explain how we perform API generation

### 2020-06-02 report/sections/server/node.tex
Report: make wording more consistent
Report: start on chapter 4

@@
 
-In this chapter, 
-we \dots
+In this chapter we present \fancyname{NodeMPST},
+our session type API generation strategy for 
+server-side endpoints implemented on the Node.js runtime \cite{node}.
+We continue to use the running example of the \tprotocol{Adder}
+protocol, and refer to the FSM of the \trole{Svr} endpoint
+(\cref{fig:addersvrfsm}) throughout this chapter.
 
-\section{Motivation}
+\begin{figure}[!b]
+\centering
+%\includegraphics[width=0.5\textwidth]{•}
+\captionof{figure}{\dots}
+\label{fig:addersvrfsm}
+\end{figure}
+
+\section{Challenges}
+\label{section:nodechallenges}
 \begin{itemize}
@@
 \item provide static guarantees where possible -- communication mismatch and channel linearity
+\item typescript's type system is gradual, structural and non-linear -- session type api generation examples that provide static guarantees on channel usage build upon target language type system's support for linear resources
 \end{itemize}
 
+As outlined in \cref{section:typescript}, TypeScript's type system
+is neither affine nor linear, so we need to work around the 
+limitations of the language to enforce linear usage of channel resources.
+Similarly, we need to 
+the language does not provide
+built-in constructs for enforcing linear channel usage.
+
 \section{Approach}
+\label{section:nodeapproach}
+
+We motivate our approach from \cite{PureScript2019, HuJava}
+to generate \textbf{handler-style APIs} to be implemented
+by developers.
+This aligns with idiomatic TypeScript practices of defining
+\textit{callbacks} in the application logic.
+By strictly specifying handlers for send and receive actions,
+we do not expose send and receive APIs to the developer,
+making it impossible for the developer to reuse channels.
+
 \begin{itemize}
-\item typescript's type system is gradual, structural and non-linear -- session type api generation examples that provide static guarantees on channel usage build upon target language type system's support for linear resources
+
 \item inspired by purescript approach -- hide the channel resource so they are linear by construction, simply expose callbacks and handlers
@@
 
+When executing \fancyname{SessionTS} to generate code for the
+\trole{Svr} endpoint specifying the \texttt{node} target,
+we obtain the following files:
+
+\begin{itemize}
+\item \textbf{EFSM.ts:} \dots
+\item \textbf{Svr.ts:} \dots
+\end{itemize}
+
 \section{State Encodings}
+\label{section:nodeefsm}
 
@@
 \item callbacks -- type aliases for each state (handler)
-\item need a way for the runtime to distinguish between states
+\item need a way for the runtime to distinguish between states (implementation)
 \item attempt at using conditional types, but because they handle conditional union types in a distributed way (give example), cannot be statically typed
@@
 
+We generalise deterministic send actions as a trivial \textit{selection}, 
+as motivated from the theory (\dots).
+
 \subsection{Receive}
@@
 \end{itemize}
+
+As with send states,
+we generalise deterministic receive actions as a trivial \textit{branch}, 
+as motivated from the theory (\dots).
 
@@
 \section{Runtime}
+\label{section:noderuntime}
+
 \begin{itemize}
@@
 \includegraphics[width=0.5\textwidth]{NodeRuntimeEFSM}
-\captionof{figure}{Informal Abstraction of EFSM Execution for
+\captionof{figure}{``Message Passing'' Abstraction of EFSM Execution for
 Server Endpoints}
@@
 
-\section{Alternative Designs and Limitations}
+\section{Alternative Designs}
+\label{section:nodealt}
+
+The main alternative EFSM encoding would be similar to those
+presented in \cite{Hybrid2016}, which encodes
+each EFSM state as a separate class and expose communication APIs
+(e.g. send and/or receive) that respect the permitted transitions at
+that state.
+
+\begin{lstlisting}[language=scribble]
+global protocol OneAdder(role Client, role Svr) {
+	NUM1(number) from Client to Svr;
+	NUM2(number) from Client to Svr;
+	SUM(number)  from Svr to Client;
+}
+\end{lstlisting}
+
+\begin{figure}[!h]
+\centering
+\begin{subfigure}{\textwidth}
+\begin{lstlisting}[language=javascript, tabsize=2]
+const logic = async (init) => {
+	const [x, num2] = await init.receive();
+ 	const [y, sum] = await num2.receive();
+	return sum.send(x + y);
+}
+\end{lstlisting}
+\caption{\dots}
+%\label{...}
+\end{subfigure}
+\hfill
+\begin{subfigure}{\textwidth}
+\begin{lstlisting}[language=javascript]
+const logic = new S4({
+	NUM1: (x) => new S6({
+		NUM2: (y) => ([Labels.S7.SUM, [x + y], new S5()]),
+	}),
+});
+\end{lstlisting}
+\caption{\dots}
+%\label{...}
+\end{subfigure}
+\captionof{figure}{\dots}
+%\label{...}
+\end{figure}
 
@@
 \item `hardcoded' switch case runtime like the TSOP papers -- doesn't best leverage the type system
+\end{itemize}
+
+\section{Limitations}
+\label{section:nodelimitations}
+
+\begin{itemize}
+\item verbose
+\item state identifiers are in the API
+\item performance aspect with class-based handlers
 \end{itemize}

### 2020-06-03 report/sections/server/codegen/implementation.tex
Report: progress on chapter 4

@@
 to use \fancyname{SessionTS},
-and \texttt{subprocess} package provides an abstraction to
+and the \texttt{subprocess} package to 
 invoke external tools, such as Scribble and the TypeScript Compiler.

### 2020-06-03 report/sections/server/codegen/workflow.tex
Report: progress on chapter 4

@@
 \begin{lstlisting}[language=Scribble]
-module Adder;
-
 global protocol Adder(role Client, role Svr) {
@@
 		RES(number)         from Svr to Client;
+		do Adder(Client, Svr);
 	} or {
 		QUIT(string) from Client to Svr;
+		choice at Svr {
+			THANKS()    from Svr to Client;
+		} or {
 			TERMINATE() from Svr to Client;
+		}
 	}
@@
 to demonstrate how our work performs TypeScript API generation.
+This describes a binary session between \trole{Client} and \trole{Svr}:
+if \trole{Client} asks \trole{Svr} to \tmsg{ADD} two numbers,
+\trole{Svr} responds with the \tmsg{RES}ult; this loop continues
+until \trole{Client} kindly asks to \tmsg{QUIT} with a message,
+where \trole{Svr} either responds kindly with \tmsg{THANKS},
+or just says \tmsg{TERMINATE}, but both choices will close the session.
 
@@
 we define an \texttt{Endpoint} dataclass\footnote{
-A Python dataclass uses the decorator to generate
+A Python dataclass uses the \texttt{@dataclass} decorator to generate
 ``boilerplate'' methods, such as the constructor, based on the

### 2020-06-03 report/sections/server/node.tex
Report: progress on chapter 4

@@
 \centering
-%\includegraphics[width=0.5\textwidth]{•}
-\captionof{figure}{\dots}
+\includegraphics[width=0.5\textwidth]{AdderSvrFSM}
+\captionof{figure}{\trole{Svr} Endpoint FSM
+in \tprotocol{Adder} protocol}
 \label{fig:addersvrfsm}
@@
 \textit{callbacks} in the application logic.
+We type the parameters and return values of the handlers to
+reflect the message types specified in the protocol.
 By strictly specifying handlers for send and receive actions,
@@
 making it impossible for the developer to reuse channels.
+
+The responsibility of guaranteeing linear channel usage now falls
+under the runtime that executes the EFSM.
+As we generate this for the developer,
+we can provide \textbf{static}
+linearity guarantees by construction.
 
@@
 \trole{Svr} endpoint specifying the \texttt{node} target,
-we obtain the following files:
+the developer obtains the following files:
 
 \begin{itemize}
-\item \textbf{EFSM.ts:} \dots
-\item \textbf{Svr.ts:} \dots
+\item \textbf{EFSM.ts:} TypeScript types constructed for 
+the EFSM encoding (\cref{section:nodeefsm});
+\item \textbf{Svr.ts:} Session runtime for executing the EFSM 
+(\cref{section:noderuntime}).
 \end{itemize}
 
-\section{State Encodings}
+\section{EFSM Encoding}
 \label{section:nodeefsm}
+
+In addition to states and transitions, we generate 
+TypeScript constructs
+for relevant metadata in the state machine: this includes
+\textit{roles}, \textit{labels} and \textit{message structures}.
+
+The EFSM contains more than just states and transitions
+
+The generated \textbf{EFSM.ts} file encodes all information
+available
+
+\begin{figure}
+\begin{lstlisting}[language=javascript,tabsize=2]
+export namespace Roles {...};
+export namespace Labels {...};
+export namespace Message {...};
+
+export namespace Handler {...};
+
+abstract class ISend {...};
+abstract class IReceive {...};
+abstract class ITerminal {...};
+export namespace Implementation {...};
+
+export type EfsmTransitionHandler =
+	(implementation: Implementation.Type) => void;
+export type MessageHandler = (message: any) => void;
+\end{lstlisting}
+\captionof{lstlisting}{Structure of generated EFSM.ts file 
+for server endpoints}
+\label{lst:nodeefsmfile}
+\end{figure}
 
@@
 
-\subsection{Send}
+\begin{figure}[!ht]
+\begin{lstlisting}[language=javascript, tabsize=2]
+// Inside the Message namespace...
+export type S54THANKS = {
+	label: Labels.S54.THANKS,
+	payload: [string],
+};
+export type S54TERMINATE = {
+	label: Labels.S54.TERMINATE,
+	payload: [],
+};
+
+export type S54 = | S54THANKS | S54TERMINATE;
+\end{lstlisting}
+%\label{}
+\captionof{lstlisting}{Generated Message Type Definition for State 54}
+\end{figure}
+
+\subsection{Handler APIs}
+
+We collect the APIs that the developer needs to implement
+under the \texttt{Handler} namespace. 
+As a design choice, we \textit{do not} generate handlers for
+terminal states, because the semantics of inactivity mean
+there is nothing to handle.
+We outline how the generated API differs between sending
+and receiving states.
+
+\paragraph{Send}
+We generalise deterministic send actions as a trivial \textit{selection}, 
+as motivated from the theory (\dots).
+
+\begin{figure}[!ht]
+\begin{lstlisting}[language=javascript, tabsize=2]
+// Inside the Handler namespace...
+export type S54 = 
+	| [Labels.S54.THANKS, Message.S54THANKS['payload'],
+		Implementation.S52] 
+	| [Labels.S54.TERMINATE, Message.S54TERMINATE['payload'], 
+		Implementation.S52];
+\end{lstlisting}
+\captionof{lstlisting}{Generated Type for \trole{Svr} Send State
+in \tprotocol{Adder} protocol}
+\end{figure}
+
+\paragraph{Receive}
 \begin{itemize}
@@
 
-We generalise deterministic send actions as a trivial \textit{selection}, 
-as motivated from the theory (\dots).
-
-\subsection{Receive}
-\begin{itemize}
-\item very short -- go over the structure and give examples
-\item show how it extends for polyadic payloads
-\end{itemize}
+\begin{figure}[!ht]
+\begin{lstlisting}[language=javascript,tabsize=2]
+// Inside the Handler namespace...
+export type S51 = {
+	[Labels.S51.ADD]: (...payload: Message.S51ADD['payload']) =>
+		Implementation.S53,
+	[Labels.S51.QUIT]: (...payload: Message.S51QUIT['payload']) => 
+		Implementation.S54,
+}
+\end{lstlisting}
+\captionof{lstlisting}{Generated Type for \trole{Svr} Receive State
+in \tprotocol{Adder} protocol}
+\end{figure}
 

### 2020-06-03 report/sections/server/node.tex
Report: reorganise directory for chapter 4

--- 2020-06-02 report/sections/server/node.tex
@@
 
-In this chapter, 
-we \dots
+In this chapter we present \fancyname{NodeMPST},
+our session type API generation strategy for 
+server-side endpoints implemented on the Node.js runtime \cite{node}.
+We continue to use the running example of the \tprotocol{Adder}
+protocol, and refer to the FSM of the \trole{Svr} endpoint
+(\cref{fig:addersvrfsm}) throughout this chapter.
 
-\section{Motivation}
-\begin{itemize}
-\item session type implementation on the node runtime
-\item need to provide consistency under the single-threaded event loop model
-\item provide idiomatic event-driven APIs
-\item provide static guarantees where possible -- communication mismatch and channel linearity
-\end{itemize}
-
-\section{Approach}
-\begin{itemize}
-\item typescript's type system is gradual, structural and non-linear -- session type api generation examples that provide static guarantees on channel usage build upon target language type system's support for linear resources
-\item inspired by purescript approach -- hide the channel resource so they are linear by construction, simply expose callbacks and handlers
-\item the runtime executes the EFSM, performs the send and receive actions through the websocket, and invokes the correct handler as required
-\item provide overview of the files we generate (and forward reference in the correct subsections) -- EFSM, Runtime, `others' (exclude cancellation)
-\end{itemize}
-
-\section{State Encodings}
-
-\begin{itemize}
-\item callbacks -- type aliases for each state (handler)
-\item need a way for the runtime to distinguish between states
-\item attempt at using conditional types, but because they handle conditional union types in a distributed way (give example), cannot be statically typed
-\item alternative -- use discriminated union by wrapping in an Implementation class
-\item all non-terminal states will `return' the successor implementation -- very difficult to resolve in the type-checker in the runtime (give example of what it may look like) -- solve by giving each state implementation the `advance' function to respect the event-driven nature of everything, so they can call it on completion
-\item visualise the interaction between runtime and states as a sequence diagram for message passing
-\end{itemize}
-
-\subsection{Send}
-\begin{itemize}
-\item very short -- go over the structure and give examples
-\item show how it extends for polyadic payloads
-\end{itemize}
-
-\subsection{Receive}
-\begin{itemize}
-\item very short -- go over the structure and give examples
-\item show how it extends for polyadic payloads
-\end{itemize}
-
-\subsection{Terminal}
-\begin{itemize}
-\item basically nothing here -- can highlight the provided type alias to provide a better developer experience
-\end{itemize}
-
-\section{Runtime}
-\begin{itemize}
-\item public API -- provide seam for websocket server and implementation
-\item private API -- carry out the session
-\end{itemize}
-
-\subsection{Managing Connections}
-\begin{itemize}
-\item use set and partial object to keep track of pending connections
-\item ws event listeners
-\item forward reference that we need to manage cancellations here too
-\end{itemize}
-
-\subsection{Executing the EFSM}
-\begin{itemize}
-\item implementation details delegated to the actual implementation states
-\item constructor binds all methods (that will be passed to other components0 to `this' because of how javascript works with respect to the scoping of `this' (give short example -- or ignore, if this is in the background for typescript)
-\item advance -- use discriminated union to figure out what to pass to the state (send -- sendMessage, receive -- register)
-\end{itemize}
-
-\begin{figure}[!ht]
+\begin{figure}[!b]
 \centering
-\includegraphics[width=0.5\textwidth]{NodeRuntimeEFSM}
-\captionof{figure}{Informal Abstraction of EFSM Execution for
-Server Endpoints}
-\label{fig:noderuntimeefsm}
+\includegraphics[width=0.5\textwidth]{AdderSvrFSM}
+\captionof{figure}{\trole{Svr} Endpoint FSM
+in \tprotocol{Adder} protocol}
+\label{fig:addersvrfsm}
 \end{figure}
 
-\subsection{Handling Message Sends}
-\begin{itemize}
-\item very short (just for completeness in the report) -- messages are serialised JS objects (in JSON notation) sent and decoded on the other end, type-correct by construction
-\end{itemize}
-
-\subsection{Handling Message Receives}
-\begin{itemize}
-\item motivate the edge case that message can arrive in the websocket (in succession) before the handler is registered
-\item motivate the edge case that message can arrive `out of protocol order' 
-\item explain the double queue system and how that provides consistency -- emphasising that this works because of the single-threaded typescript runtime
-\end{itemize}
-
-\begin{figure}[!ht]
-\centering
-\begin{subfigure}[b]{0.8\textwidth}
-\centering
-\includegraphics[width=\textwidth]{NodeRuntimeReceive2}
-\caption{Message processed before transitioning to receive state}
-\label{subfig:nodereceivemsgfirst}
-\end{subfigure}
-\hfill
-\begin{subfigure}[b]{0.8\textwidth}
-\centering
-\includegraphics[width=\textwidth]{NodeRuntimeReceive1}
-\caption{Message processed after transitioning to receive state}
-\label{subfig:nodereceivehandlefirst}
-\end{subfigure}
-\captionof{figure}{Possible Orderings for Handling Message Event and Preparing
-Receive State}
-\label{fig:nodereceivecompare}
-\end{figure}
-
-\subsection{Handling Termination}
-\begin{itemize}
-\item design choice that the client is the one that closes the connection because of the centralised role of the server
-\item forward reference that this gives opportunity to extend support for general protocols, as in those protocols, the server's terminal state doesn't imply it's terminal for everyone else
-\end{itemize}
-
-\section{Alternative Designs and Limitations}
-
-\begin{itemize}
-\item exposing channel resources like the java example -- less idiomatic (not-event-driven), need runtime linearity checks like the java example
-\item `hardcoded' switch case runtime like the TSOP papers -- doesn't best leverage the type system
-\end{itemize}
-
-\section{Summary}
-\begin{itemize}
-\item static guarantees on channel linearity and communication mismatch (up to the use of `any)
-\item handling receives introduce consistency problems, but we deal with it properly here
-\end{itemize}
+\input{sections/server/node/challenges}
+\input{sections/server/node/approach}
+\input{sections/server/node/efsm}
+\input{sections/server/node/runtime}
+\input{sections/server/node/alt}
+\input{sections/server/node/limitations}

### 2020-06-03 report/sections/server/node/alt.tex
Report: reorganise directory for chapter 4

\section{Alternative Designs}
\label{section:nodealt}

The main alternative EFSM encoding would be similar to those
presented in \cite{Hybrid2016}, which encodes
each EFSM state as a separate class and expose communication APIs
(e.g. send and/or receive) that respect the permitted transitions at
that state.

\begin{lstlisting}[language=scribble]
global protocol OneAdder(role Client, role Svr) {
	NUM1(number) from Client to Svr;
	NUM2(number) from Client to Svr;
	SUM(number)  from Svr to Client;
}
\end{lstlisting}

\begin{figure}[!h]
\centering
\begin{subfigure}{\textwidth}
\begin{lstlisting}[language=javascript, tabsize=2]
const logic = async (init) => {
	const [x, num2] = await init.receive();
 	const [y, sum] = await num2.receive();
 	// init.receive();	// this would type-check but violate linearity
	return sum.send(x + y);
}
\end{lstlisting}
\caption{\dots}
%\label{...}
\end{subfigure}
\hfill
\begin{subfigure}{\textwidth}
\begin{lstlisting}[language=javascript]
const logic = new S4({
	NUM1: (x) => new S6({
		NUM2: (y) => ([Labels.S7.SUM, [x + y], new S5()]),
	}),
});
\end{lstlisting}
\caption{\dots}
%\label{...}
\end{subfigure}
\captionof{figure}{\dots}
%\label{...}
\end{figure}

\begin{itemize}
\item exposing channel resources like the java example -- less idiomatic (not-event-driven), need runtime linearity checks like the java example
\item `hardcoded' switch case runtime like the TSOP papers -- doesn't best leverage the type system
\end{itemize}

### 2020-06-03 report/sections/server/node/approach.tex
Report: reorganise directory for chapter 4

\section{Approach}
\label{section:nodeapproach}

We motivate our approach from \cite{PureScript2019, HuJava}
to generate \textbf{handler-style APIs} to be implemented
by developers.
This aligns with idiomatic TypeScript practices of defining
\textit{callbacks} in the application logic.
We type the parameters and return values of the handlers to
reflect the message types specified in the protocol.
By strictly specifying handlers for send and receive actions,
we do not expose send and receive APIs to the developer,
making it impossible for the developer to reuse channels.

The responsibility of guaranteeing linear channel usage now falls
under the runtime that executes the EFSM.
As we generate this for the developer,
we can provide \textbf{static}
linearity guarantees by construction.

When executing \fancyname{SessionTS} to generate code for the
\trole{Svr} endpoint specifying the \texttt{node} target,
the developer obtains the following files:

\begin{itemize}
\item \textbf{EFSM.ts:} TypeScript types constructed for 
the EFSM encoding (\cref{section:nodeefsm});
\item \textbf{Svr.ts:} Session runtime for executing the EFSM 
(\cref{section:noderuntime}).
\end{itemize}

### 2020-06-03 report/sections/server/node/challenges.tex
Report: reorganise directory for chapter 4

\section{Challenges}
\label{section:nodechallenges}
\begin{itemize}
\item session type implementation on the node runtime
\item need to provide consistency under the single-threaded event loop model
\item provide idiomatic event-driven APIs
\item provide static guarantees where possible -- communication mismatch and channel linearity
\item typescript's type system is gradual, structural and non-linear -- session type api generation examples that provide static guarantees on channel usage build upon target language type system's support for linear resources
\end{itemize}

As outlined in \cref{section:typescript}, TypeScript's type system
is neither affine nor linear, so we need to work around the 
limitations of the language to enforce linear usage of channel resources.
Similarly, we need to 
the language does not provide
built-in constructs for enforcing linear channel usage.

### 2020-06-03 report/sections/server/node/efsm.tex
Report: reorganise directory for chapter 4

\section{EFSM Encoding}
\label{section:nodeefsm}

In addition to states and transitions, we generate 
TypeScript constructs
for relevant metadata in the state machine: this includes
\textit{roles}, \textit{labels} and \textit{message structures}.

The EFSM contains more than just states and transitions

The generated \textbf{EFSM.ts} file encodes all information
available

\begin{figure}
\begin{lstlisting}[language=javascript,tabsize=2]
export namespace Roles {...};
export namespace Labels {...};
export namespace Message {...};

export namespace Handler {...};

abstract class ISend {...};
abstract class IReceive {...};
abstract class ITerminal {...};
export namespace Implementation {...};

export type EfsmTransitionHandler =
	(implementation: Implementation.Type) => void;
export type MessageHandler = (message: any) => void;
\end{lstlisting}
\captionof{lstlisting}{Structure of generated EFSM.ts file 
for server endpoints}
\label{lst:nodeefsmfile}
\end{figure}

\begin{itemize}
\item callbacks -- type aliases for each state (handler)
\item need a way for the runtime to distinguish between states (implementation)
\item attempt at using conditional types, but because they handle conditional union types in a distributed way (give example), cannot be statically typed
\item alternative -- use discriminated union by wrapping in an Implementation class
\item all non-terminal states will `return' the successor implementation -- very difficult to resolve in the type-checker in the runtime (give example of what it may look like) -- solve by giving each state implementation the `advance' function to respect the event-driven nature of everything, so they can call it on completion
\item visualise the interaction between runtime and states as a sequence diagram for message passing
\end{itemize}

\begin{figure}[!ht]
\begin{lstlisting}[language=javascript, tabsize=2]
// Inside the Message namespace...
export type S54THANKS = {
	label: Labels.S54.THANKS,
	payload: [string],
};
export type S54TERMINATE = {
	label: Labels.S54.TERMINATE,
	payload: [],
};

export type S54 = | S54THANKS | S54TERMINATE;
\end{lstlisting}
%\label{}
\captionof{lstlisting}{Generated Message Type Definition for State 54}
\end{figure}

\subsection{Handler APIs}

We collect the APIs that the developer needs to implement
under the \texttt{Handler} namespace. 
As a design choice, we \textit{do not} generate handlers for
terminal states, because the semantics of inactivity mean
there is nothing to handle.
We outline how the generated API differs between sending
and receiving states.

\paragraph{Send}
We generalise deterministic send actions as a trivial \textit{selection}, 
as motivated from the theory (\dots).

\begin{figure}[!ht]
\begin{lstlisting}[language=javascript, tabsize=2]
// Inside the Handler namespace...
export type S54 = 
	| [Labels.S54.THANKS, Message.S54THANKS['payload'],
		Implementation.S52] 
	| [Labels.S54.TERMINATE, Message.S54TERMINATE['payload'], 
		Implementation.S52];
\end{lstlisting}
\captionof{lstlisting}{Generated Type for \trole{Svr} Send State
in \tprotocol{Adder} protocol}
\end{figure}

\paragraph{Receive}
\begin{itemize}
\item very short -- go over the structure and give examples
\item show how it extends for polyadic payloads
\end{itemize}

\begin{figure}[!ht]
\begin{lstlisting}[language=javascript,tabsize=2]
// Inside the Handler namespace...
export type S51 = {
	[Labels.S51.ADD]: (...payload: Message.S51ADD['payload']) =>
		Implementation.S53,
	[Labels.S51.QUIT]: (...payload: Message.S51QUIT['payload']) => 
		Implementation.S54,
}
\end{lstlisting}
\captionof{lstlisting}{Generated Type for \trole{Svr} Receive State
in \tprotocol{Adder} protocol}
\end{figure}

As with send states,
we generalise deterministic receive actions as a trivial \textit{branch}, 
as motivated from the theory (\dots).

\subsection{Terminal}
\begin{itemize}
\item basically nothing here -- can highlight the provided type alias to provide a better developer experience
\end{itemize}

### 2020-06-03 report/sections/server/node/limitations.tex
Report: reorganise directory for chapter 4

\section{Limitations}
\label{section:nodelimitations}

\begin{itemize}
\item verbose
\item state identifiers are in the API
\item scoping achieved through nested objects
\item performance aspect with class-based handlers
\end{itemize}

\begin{lstlisting}[language=javascript]
const first = new S4({ NUM1: (x) => secondOperand, });
const second = new S6({ NUM2: (y) => sum, });
const sum = [Labels.S7.SUM, ???, new S5()];
\end{lstlisting}

### 2020-06-03 report/sections/server/node/runtime.tex
Report: reorganise directory for chapter 4

\section{Runtime}
\label{section:noderuntime}

\begin{itemize}
\item public API -- provide seam for websocket server and implementation
\item private API -- carry out the session
\end{itemize}

\subsection{Managing Connections}
\begin{itemize}
\item use set and partial object to keep track of pending connections
\item ws event listeners
\item forward reference that we need to manage cancellations here too
\end{itemize}

\subsection{Executing the EFSM}
\begin{itemize}
\item implementation details delegated to the actual implementation states
\item constructor binds all methods (that will be passed to other components0 to `this' because of how javascript works with respect to the scoping of `this' (give short example -- or ignore, if this is in the background for typescript)
\item advance -- use discriminated union to figure out what to pass to the state (send -- sendMessage, receive -- register)
\end{itemize}

\begin{figure}[!ht]
\centering
\includegraphics[width=0.5\textwidth]{NodeRuntimeEFSM}
\captionof{figure}{``Message Passing'' Abstraction of EFSM Execution for
Server Endpoints}
\label{fig:noderuntimeefsm}
\end{figure}

\subsection{Handling Message Sends}
\begin{itemize}
\item very short (just for completeness in the report) -- messages are serialised JS objects (in JSON notation) sent and decoded on the other end, type-correct by construction
\end{itemize}

\subsection{Handling Message Receives}
\begin{itemize}
\item motivate the edge case that message can arrive in the websocket (in succession) before the handler is registered
\item motivate the edge case that message can arrive `out of protocol order' 
\item explain the double queue system and how that provides consistency -- emphasising that this works because of the single-threaded typescript runtime
\end{itemize}

\begin{figure}[!ht]
\centering
\begin{subfigure}[b]{0.8\textwidth}
\centering
\includegraphics[width=\textwidth]{NodeRuntimeReceive2}
\caption{Message processed before transitioning to receive state}
\label{subfig:nodereceivemsgfirst}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.8\textwidth}
\centering
\includegraphics[width=\textwidth]{NodeRuntimeReceive1}
\caption{Message processed after transitioning to receive state}
\label{subfig:nodereceivehandlefirst}
\end{subfigure}
\captionof{figure}{Possible Orderings for Handling Message Event and Preparing
Receive State}
\label{fig:nodereceivecompare}
\end{figure}

\subsection{Handling Termination}
\begin{itemize}
\item design choice that the client is the one that closes the connection because of the centralised role of the server
\item forward reference that this gives opportunity to extend support for general protocols, as in those protocols, the server's terminal state doesn't imply it's terminal for everyone else
\end{itemize}

### 2020-06-03 report/sections/general/p2p/twobuyer.tex
Report: minor change to TwoBuyer

@@
 \begin{lstlisting}[language=scribble]
-module ClientClient;
-
 global protocol TwoBuyer(role A, role B, role S) {

### 2020-06-03 report/sections/server/node/efsm.tex
Report: progress on chapter 4

@@
 \label{section:nodeefsm}
+
+We show the structure of the generated EFSM.ts file in
+\cref{lst:nodeefsmfile}.
+Note that the formal definition of the EFSM in 
+\cref{section:scribbleefsm}
+contains more than just states and the state transition function,
+so we encode the additional information as well.
+Each type of information is grouped into their own
+\textit{namespace}, and are collectively exported in
+the EFSM \textit{module} for the developer to use.
 
@@
 \item callbacks -- type aliases for each state (handler)
-\item need a way for the runtime to distinguish between states (implementation)
-\item attempt at using conditional types, but because they handle conditional union types in a distributed way (give example), cannot be statically typed
-\item alternative -- use discriminated union by wrapping in an Implementation class
-\item all non-terminal states will `return' the successor implementation -- very difficult to resolve in the type-checker in the runtime (give example of what it may look like) -- solve by giving each state implementation the `advance' function to respect the event-driven nature of everything, so they can call it on completion
-\item visualise the interaction between runtime and states as a sequence diagram for message passing
 \end{itemize}
@@
 // Inside the Message namespace...
-export type S54THANKS = {
+export interface S54THANKS {
 	label: Labels.S54.THANKS,
@@
 };
-export type S54TERMINATE = {
+export interface S54TERMINATE {
 	label: Labels.S54.TERMINATE,
@@
 \end{lstlisting}
-%\label{}
+\label{lst:addersvrmsg}
 \captionof{lstlisting}{Generated Message Type Definition for State 54}
 \end{figure}
+
+By expressing the payload type as a \textit{tuple}\footnote{
+In TypeScript, a tuple is an array with fixed size
+and known types for elements at each position.
+},
+we easily generalise our type definition to polyadic payloads.
 
@@
 there is nothing to handle.
+%Hence, all generated handler APIs are for non-terminal states,
+%and will hold some reference to the successor state encoding.
+%The reader will notice that, in the listings below, the successor
+%state is stated to be under the 
 We outline how the generated API differs between sending
@@
 
-\paragraph{Send}
+\subparagraph{Send}
+We model selections using \textit{union types} to
+encapsulate the possible send actions. 
+Each send action is encoded as a tuple of
+the label, the payload, and the successor state.
+We see some benefits from defining Message Types as interfaces:
+with deterministic properties, we can index into the message type
+to access the property's type, 
+e.g. \texttt{Message.S54THANKS['payload']} would resolve to
+\texttt{[string]}, based on \cref{lst:addersvrmsg}.
+We provide an example in \cref{lst:addersvrsendhandler}.
+
 We generalise deterministic send actions as a trivial \textit{selection}, 
-as motivated from the theory (\dots).
+as motivated from the theory (\cref{fig:globaltypes}),
+so the encoding for State 53 in the \trole{Svr} FSM would be
+the union of a single tuple.
 
@@
 in \tprotocol{Adder} protocol}
+\label{lst:addersvrsendhandler}
 \end{figure}
 
-\paragraph{Receive}
+\subparagraph{Receive}
 \begin{itemize}
@@
 we generalise deterministic receive actions as a trivial \textit{branch},
-as motivated from the theory (\dots).
+which would be an interface with one property.
 
-\subsection{Terminal}
+\subsection{Wrapping Handlers in ``Implementations''}
 \begin{itemize}
-\item basically nothing here -- can highlight the provided type alias to provide a better developer experience
+\item need a way for the runtime to distinguish between states (implementation)
+\item attempt at using conditional types, but because they handle conditional union types in a distributed way (give example), cannot be statically typed
+\item alternative -- use discriminated union by wrapping in an Implementation class
+\item all non-terminal states will `return' the successor implementation -- very difficult to resolve in the type-checker in the runtime (give example of what it may look like) -- solve by giving each state implementation the `advance' function to respect the event-driven nature of everything, so they can call it on completion
+\item visualise the interaction between runtime and states as a sequence diagram for message passing
 \end{itemize}

### 2020-06-03 report/sections/server/node/alt.tex
Report: progress on chapter 4

@@
 \centering
-\begin{subfigure}{\textwidth}
-\begin{lstlisting}[language=javascript, tabsize=2]
+\begin{subfigure}{0.49\textwidth}
+\begin{lstlisting}[language=javascript,tabsize=2]
 const logic = async (init) => {
-	const [x, num2] = await init.receive();
- 	const [y, sum] = await num2.receive();
- 	// init.receive();	// this would type-check but violate linearity
+	const [x, num2] = 
+		await init.receive();
+ 	const [y, sum] = 
+ 		await num2.receive();
+
+ 	// init.receive(); (*@\label{line:nodebreaklinear}@*)
 	return sum.send(x + y);
@@
 \end{lstlisting}
-\caption{\dots}
+\caption{Alternative Version}
 %\label{...}
@@
 \hfill
-\begin{subfigure}{\textwidth}
-\begin{lstlisting}[language=javascript]
+\begin{subfigure}{0.49\textwidth}
+\begin{lstlisting}[language=javascript,tabsize=2]
 const logic = new S4({
 	NUM1: (x) => new S6({
-		NUM2: (y) => ([Labels.S7.SUM, [x + y], new S5()]),
+		NUM2: (y) => ([
+			Labels.S7.SUM,
+			[x + y],
+			new S5(),
+		]),
 	}),
 });
+
 \end{lstlisting}
-\caption{\dots}
+\caption{Current Version}
 %\label{...}

### 2020-06-03 report/sections/server/node/efsm.tex
Report: progress on chapter 4

@@
 
-In addition to states and transitions, we generate 
-TypeScript constructs
-for relevant metadata in the state machine: this includes
-\textit{roles}, \textit{labels} and \textit{message structures}.
-
-The EFSM contains more than just states and transitions
-
-The generated \textbf{EFSM.ts} file encodes all information
-available
-
 \begin{figure}
-\begin{lstlisting}[language=javascript,tabsize=2]
+\begin{lstlisting}[language=javascript,tabsize=2,title=EFSM.ts]
+// (*@\cref{subsection:nodeefsmroleslabelsmsg}@*)
 export namespace Roles {...};
@@
 
+// (*@\cref{subsection:nodeefsmhandlers}@*)
 export namespace Handler {...};
 
+// (*@\cref{subsection:nodeefsmimplementation}@*)
 abstract class ISend {...};
@@
 \end{lstlisting}
-\captionof{lstlisting}{Structure of generated EFSM.ts file 
-for server endpoints}
+\captionof{lstlisting}{Structure of Generated EFSM Encoding 
+for Server Endpoint}
 \label{lst:nodeefsmfile}
@@
 
-\begin{itemize}
-\item callbacks -- type aliases for each state (handler)
-\end{itemize}
+\subsection{Roles, Labels, Messages}
+\label{subsection:nodeefsmroleslabelsmsg}
+
+We generate TypeScript constructs for these pieces of information
+so they can be reused throughout the generated code, 
+and in particular, the runtime.
+
+\subparagraph{Roles}
+The runtime needs to know the identifiers of participants involved
+in the session, and who to send/receive from 
+depending on the EFSM state.
+We generate string enumerations, or \textit{enums}, for each 
+participant in the protocol, \textit{excluding} the 
+first person endpoint. 
+The enum appropriately groups the collection
+of participants involved and scales for multiparty sessions,
+whilst making it simple to derive other types, e.g. a mapping from
+participants (indexed by the enum) to WebSockets.
+
+\subparagraph{Labels}
+The runtime needs to decide which handler to invoke, based
+on the label of the received message. Similarly, the developer needs
+to provide handlers specifying their internal choice (e.g. which
+message label to send) and how to handle external choice (e.g. 
+how to handle received message with particular label).
+For the same reason, we also generate string enums for message labels,
+one enum per state. Enums are compatible with switch statements,
+which can be used to dispatch messages to the correct handlers
+in the runtime based on the message label. 
+We give an example in \cref{lst:nodeefsmlabels}.
+
+\begin{figure}[!ht]
+\begin{lstlisting}[language=javascript,tabsize=2]
+// Inside the Labels namespace...
+export enum S51 { ADD = "ADD", QUIT = "QUIT", };
+export enum S53 { RES = "RES", };
+export enum S54 { THANKS = "THANKS", TERMINATE = "TERMINATE", };
+\end{lstlisting}
+\captionof{lstlisting}{Generated Label Enums for \trole{Svr} endpoint}
+\label{lst:nodeefsmlabels}
+\end{figure}
+
+\subparagraph{Messages}
 
@@
 \end{lstlisting}
+\captionof{lstlisting}{Generated Message Type Definition for State 54}
 \label{lst:addersvrmsg}
-\captionof{lstlisting}{Generated Message Type Definition for State 54}
 \end{figure}
@@
 \subsection{Handler APIs}
+\label{subsection:nodeefsmhandlers}
 
@@
 \subparagraph{Send}
-We model selections using \textit{union types} to
-encapsulate the possible send actions. 
+We model selections using a \textit{union type} to
+encapsulate the possible send actions, as shown in 
+\cref{lst:addersvrsendhandler}.
 Each send action is encoded as a tuple of
-the label, the payload, and the successor state.
+the label, the payload, and the successor state encoding.
 We see some benefits from defining Message Types as interfaces:
@@
 to access the property's type, 
-e.g. \texttt{Message.S54THANKS['payload']} would resolve to
-\texttt{[string]}, based on \cref{lst:addersvrmsg}.
-We provide an example in \cref{lst:addersvrsendhandler}.
-
-We generalise deterministic send actions as a trivial \textit{selection}, 
-as motivated from the theory (\cref{fig:globaltypes}),
-so the encoding for State 53 in the \trole{Svr} FSM would be
-the union of a single tuple.
+e.g. \lstonelinejs{Message.S54THANKS['payload']} 
+would resolve to
+\lstonelinejs{[string]},
+based on \cref{lst:addersvrmsg}.
 
@@
 
+We generalise deterministic send actions as a trivial \textit{selection}, 
+as motivated from the theory (\cref{fig:globaltypes}),
+so the encoding for State 53 in the \trole{Svr} FSM would be
+the union of a single tuple.
+
 \subparagraph{Receive}
-\begin{itemize}
-\item very short -- go over the structure and give examples
-\item show how it extends for polyadic payloads
-\end{itemize}
+We model branching using an interface to 
+enumerate the possible branches, as shown in
+\cref{lst:addersvrreceivehandler}.
+As with send states,
+we generalise deterministic receive actions as a trivial \textit{branch},
+which would be an interface with one property.
 
@@
 in \tprotocol{Adder} protocol}
+\label{lst:addersvrreceivehandler}
 \end{figure}
 
-As with send states,
-we generalise deterministic receive actions as a trivial \textit{branch},
-which would be an interface with one property.
+The interface properties are defined by the 
+labels of the permitted receive actions:
+the square-bracket notation means that the property name
+is derived from the value of the enclosing variable,
+so \lstonelinejs{[Labels.S51.ADD]} resolves to the
+\lstonelinejs{'ADD'} string.
+
+The interface values are functions parameterised by
+the message payload, and must return the successor state encoding.
+We see another benefit of defining the payload in Message Types
+as a tuple: we can define the receive handler parameter
+using the \textit{spread syntax}, which allows the tuple
+expression to be expanded into a list of arguments. 
+More concretely, as shown in \cref{lst:nodeefsmspread}
+it allows the developer to pattern match on the
+individual payload values (\cref{line:yesspread}) 
+rather than defining their function to expect a tuple 
+and manually destructing it (\cref{line:nospread}),
+so the former is more intuitive.
+
+\begin{figure}[!ht]
+\begin{lstlisting}[language=javascript,tabsize=2]
+const withSpread = (x: number, y: number) => {...} (*@\label{line:yesspread}@*)
+const withoutSpread = (payload: [number, number]) => {...} (*@\label{line:nospread}@*)
+
+const handler1: Handler.S51 = { ADD: withSpread		, ... };	// OK
+const handler2: Handler.S51 = { ADD: withoutSpread, ... };	// OK
+\end{lstlisting}
+\captionof{lstlisting}{Example Handler Signature 
+Compatible with Spread Syntax}
+\label{lst:nodeefsmspread}
+\end{figure}
 
 \subsection{Wrapping Handlers in ``Implementations''}
+\label{subsection:nodeefsmimplementation}
 \begin{itemize}

### 2020-06-03 report/sections/general/p2p/challenges.tex
Report: add citation to chapter 7

@@
 
-Existing work in decomposing multiparty sessions into a collection
-of binary sessions do resemble our routing proposal, but their approach
-involves having an endpoint to exclusively carry out
+There is existing work \cite{LinearDecomp, BinaryDecomp, BinaryDuality}
+that resembles our routing proposal through
+\textit{decomposing} multiparty sessions into a collection
+of binary sessions,
+but these approaches
+specify an endpoint to exclusively carry out
 routing responsibilities and act as a centralised orchestrator 

### 2020-06-03 report/sections/server/node/alt.tex
Report: progress on chapter 4

@@
 presented in \cite{Hybrid2016}, which encodes
-each EFSM state as a separate class and expose communication APIs
-(e.g. send and/or receive) that respect the permitted transitions at
-that state.
+each EFSM state in its own class 
+and expose communication APIs (e.g. send and/or receive) 
+that respect the permitted state transitions as instance methods.
 
+\begin{figure}[!h]
+\centering
+\begin{subfigure}{\textwidth}
 \begin{lstlisting}[language=scribble]
@@
 \end{lstlisting}
-
-\begin{figure}[!h]
-\centering
+\caption{The \tprotocol{OneAdder} Protocol}
+\end{subfigure}
 \begin{subfigure}{0.49\textwidth}
@@
 \end{lstlisting}
-\caption{Alternative Version}
+\caption{Alternative API Design}
 %\label{...}
@@
 \end{lstlisting}
-\caption{Current Version}
+\caption{Current API Design}
 %\label{...}

### 2020-06-03 report/sections/server/node/efsm.tex
Report: progress on chapter 4

@@
 \subparagraph{Messages}
+The handler APIs that we generate for developers
+need to refer to the label identifier and payload type: 
+we refer to this as the message structure, and encode this as a
+Message Type. 
+Each message is expressed as an interface with
+properties for the label and payload.
+These interfaces are grouped based on the EFSM state
+they belong using \textit{union types}.
+We illustrate this in \cref{lst:addersvrmsg}.
+By expressing the payload type as a \textit{tuple}\footnote{
+In TypeScript, a tuple is an array with fixed size
+and known types for elements at each position.
+},
+we easily generalise our type definition to polyadic payloads.
 
@@
 
-By expressing the payload type as a \textit{tuple}\footnote{
-In TypeScript, a tuple is an array with fixed size
-and known types for elements at each position.
-},
-we easily generalise our type definition to polyadic payloads.
-
 \subsection{Handler APIs}
@@
 there is nothing to handle.
-%Hence, all generated handler APIs are for non-terminal states,
-%and will hold some reference to the successor state encoding.
-%The reader will notice that, in the listings below, the successor
-%state is stated to be under the 
-We outline how the generated API differs between sending
-and receiving states.
+We introduce the generated handlers for sending and receiving states.
+These are non-terminal states that will involve the encoding of its
+successor state. The reader will notice that, in the listings below,
+the successor state is stated to be under the
+\texttt{Implementation} namespace: we explain in 
+\cref{subsection:nodeefsmimplementation}, but for now,
+it is sufficient to acknowledge that those refer to the encoding
+of the successor state.
 
 \subparagraph{Send}
-We model selections using a \textit{union type} to
+We model selections using a union type to
 encapsulate the possible send actions, as shown in 
@@
 using the \textit{spread syntax}, which allows the tuple
-expression to be expanded into a list of arguments. 
-More concretely, as shown in \cref{lst:nodeefsmspread}
+expression to be expanded into a list of function arguments. 
+More concretely, as shown in \cref{lst:nodeefsmspread},
 it allows the developer to pattern match on the
@@
 \label{subsection:nodeefsmimplementation}
+
+The behaviour of the runtime is dependent on the current state,
+so it needs a way to distinguish between
+all the different states -- one can think of this as implementing
+the state transition function from the theory, which is analogue to 
+overloading a \texttt{next()} method for each state.
+Due to limitations in the TypeScript language, this would have to
+be some sort of switch statement, with the \texttt{next()}
+method parameterised by some base type assignable to all states.
+Currently, the state is only determined by the handler
+to be implemented by the developer, so the switch statement
+and base type would have to be defined on the handler APIs.
+
+Unfortunately, this is not practical. 
+Handlers for send states are union types and
+handlers for receive states are interfaces,
+both of which are not supported 
+by the \lstonelinejs{instanceof}
+operator.
+
+We attempt to address this by defining an enum of state identifiers
+for each type of state (i.e. an enum for send states, 
+an enum for receive states)
+upon which to execute the EFSM, which solves the switch statement
+problem.
+Now, we are left with defining a mapping between the 
+state identifier enum to the handler type. This construct would be
+analogue to \textit{dependent types}, which again, is not a feature
+of the TypeScript type system.
+
 \begin{itemize}

### 2020-06-03 report/sections/server/node/efsm.tex
Report: complete EFSM ssubsection for chapter 4

@@
 We see some benefits from defining Message Types as interfaces:
-with deterministic properties, we can index into the message type
-to access the property's type, 
-e.g. \lstonelinejs{Message.S54THANKS['payload']} 
-would resolve to
-\lstonelinejs{[string]},
-based on \cref{lst:addersvrmsg}.
+TypeScript supports \textit{index type queries} to extract
+named property types, so
+\lstonelinejs{Message.S54THANKS['payload']} 
+would resolve to \lstonelinejs{[string]},
+based on the interface definition \cref{lst:addersvrmsg}.
 
@@
 
+\subsubsection{Distinguishing Handlers using Conditional Types}
+
 We attempt to address this by defining an enum of state identifiers
@@
 
-\begin{itemize}
-\item need a way for the runtime to distinguish between states (implementation)
-\item attempt at using conditional types, but because they handle conditional union types in a distributed way (give example), cannot be statically typed
-\item alternative -- use discriminated union by wrapping in an Implementation class
-\item all non-terminal states will `return' the successor implementation -- very difficult to resolve in the type-checker in the runtime (give example of what it may look like) -- solve by giving each state implementation the `advance' function to respect the event-driven nature of everything, so they can call it on completion
-\item visualise the interaction between runtime and states as a sequence diagram for message passing
-\end{itemize}
+We try to define type dependencies using
+\textit{conditional types} in TypeScript.
+A conditional types is a type-level expression
+\begin{lstlisting}[language=javascript,numbers=none]
+T extends U ? X : Y;
+\end{lstlisting}
+which reads, \textit{if \texttt{T} is assignable to \texttt{U},
+then the type is \texttt{X}; otherwise, the type is \texttt{Y}}.
+
+Combined with \textit{generic constraints}\footnote{
+\lstonelinejs{<T extends U>} defines a generic type \texttt{T}
+and enforces that it must be a type assignable to \texttt{U}.
+},
+we can approximate the dependency between the state identifier
+enum and the generated handler API using something
+similar to \cref{lst:conditionaltypes}.
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript,tabsize=2]
+enum SendState { S1, S3, ... };
+enum ReceiveState { S2, S4, ... };
+type State = SendState | ReceiveState;
+
+type SendHandler<S extends SendState> = 
+	S extends SendState.S1 ? Handler.S1 :
+	S extends SendState.S3 ? Handler.S3 : ... ;
+
+type ReceiveHandler<S extends ReceiveState> = 
+	S extends ReceiveState.S2 ? Handler.S2 :
+	S extends ReceiveState.S4 ? Handler.S4 : ... ;
+\end{lstlisting}
+\captionof{lstlisting}{Approximating Type Dependency
+using \textit{Conditional Types}}
+\label{lst:conditionaltypes}
+\end{figure}
+
+We intend to use this construct when defining the EFSM transition
+function for the runtime, for each type of state,
+so the method signature for transitioning to send states
+would resemble \cref{lst:conditionaltransitionfunction}
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript,numbers=none]
+declare function transitionToSend<S extends SendState>(
+	stateId: S, handler: SendHandler<S>
+);
+\end{lstlisting}
+\captionof{lstlisting}{EFSM Transition Function 
+using Conditional Types}
+\label{lst:conditionaltransitionfunction}
+\end{figure}
+
+Unfortunately, this approach does not work for the simple fact that
+conditional types were not designed to be exploited in this manner.
+The main limitation of conditional types is its 
+\textit{distributivity} when the type ``parameter'' is an
+union type (which is the case for enums, as
+\texttt{S = SendState.S1 | SendState.S3 | ...}), where
+
+\begin{lstlisting}[language=javascript,numbers=none]
+(T1 | T2) extends U ? X : Y
+\end{lstlisting}
+
+results in the conditional type being \textit{distributed}
+among each constituent,
+
+\begin{lstlisting}[language=javascript,numbers=none]
+(T1 extends U ? X : Y) | (T2 extends U ? X : Y)
+\end{lstlisting}
+
+so the type expression returns to an union type,
+
+\begin{lstlisting}[language=javascript,numbers=none]
+X | Y
+\end{lstlisting}
+
+Returning to \cref{lst:conditionaltransitionfunction}, 
+the type of \texttt{handler} will end up being a union type, 
+rather than the ``dependent type''
+construct we were hoping for.
+
+\subsubsection{Distinguishing Handlers using Discriminated Unions}
+
+Instead, we leverage \textit{discriminated unions}: all members
+of the union type share a common property (the \textit{discriminant})
+of which they each define an
+unique value for, so that the TypeScript Compiler can refine the union
+to the specific constituent upon checking the value of the discriminant
+(e.g. applying a switch statement).
+
+For the time being,
+it is sufficient to understand that for each EFSM state,
+in addition to the API defined under the
+\texttt{Handler} namespace, it also has a wrapper API defined under
+the \texttt{Implementation} namespace (\cref{lst:nodeefsmimplementation}),
+which defines the \texttt{type}
+discriminant property internally. This explains why the successor
+state encodings in 
+\cref{lst:addersvrsendhandler,lst:addersvrreceivehandler}
+were defined as such.
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript,tabsize=2]
+abstract class ISend { type: 'Send' = 'Send'; ... }
+abstract class IReceive { type: 'Receive' = 'Receive'; ... }
+
+export namespace Implementation {
+
+	export class S51 extends IReceive {
+		constructor(private handler: Handler.S51) { super(); }
+		...
+	}
+	...
+};
+\end{lstlisting}
+\captionof{lstlisting}{Discriminated Unions in EFSM 
+for Server-Side Endpoints}
+\label{lst:nodeefsmimplementation}
+\end{figure}
+
+We discuss our runtime implementation shortly 
+(\cref{section:noderuntime}), where we disclose more details regarding
+the role of the \texttt{Implemenation} wrapper API in
+the runtime.

### 2020-06-03 report/sections/server/node/runtime.tex
Report: complete EFSM ssubsection for chapter 4

@@
 \label{section:noderuntime}
+
+We define the session runtime in Svr.ts,
+named after the endpoint.
+It exposes (through the module export) a public API 
+(\cref{lst:noderuntimepublicapi})
+for the developer to pass in the WebSocket server and application logic
+(i.e. their handler implementations).
 
@@
 \end{itemize}
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript,tabsize=2,title=Svr.ts]
+export class Svr {
+	constructor(wss: WebSocket.Server,
+							initialState: Implementation.S51) { ... }
+	...
+}
+\end{lstlisting}
+\captionof{lstlisting}{Public API of Session Runtime for Server Endpoint}
+\label{lst:noderuntimepublicapi}
+\end{figure}
 
@@
 \item implementation details delegated to the actual implementation states
+\item all non-terminal states will `return' the successor implementation -- very difficult to resolve in the type-checker in the runtime (give example of what it may look like) -- solve by giving each state implementation the `advance' function to respect the event-driven nature of everything, so they can call it on completion
+\item visualise the interaction between runtime and states as a sequence diagram for message passing
 \item constructor binds all methods (that will be passed to other components0 to `this' because of how javascript works with respect to the scoping of `this' (give short example -- or ignore, if this is in the background for typescript)

### 2020-06-03 report/sections/server/node/runtime.tex
Report: start on runtime section in chapter 4

@@
 We define the session runtime in Svr.ts,
-named after the endpoint.
-It exposes (through the module export) a public API 
+named after the \trole{Svr} endpoint.
+It exposes a \textbf{public API}
 (\cref{lst:noderuntimepublicapi})
-for the developer to pass in the WebSocket server and application logic
+with seams for the developer to pass in the WebSocket server 
+and application logic
 (i.e. their handler implementations).
-
-\begin{itemize}
-\item public API -- provide seam for websocket server and implementation
-\item private API -- carry out the session
-\end{itemize}
+It is the responsibility of the developer to construct the
+WebSocket server and set it up to listen for incoming connections.
+Internally, it keeps a \textit{private API}
+for executing the EFSM, when all participants have 
+joined the session.
+The role of the public API is to manage incoming connections and
+wait for all participants to join the session 
+(\cref{subsection:noderuntimepublic}), before
+handing off to the private API to execute the EFSM
+(\cref{subsection:noderuntimeprivate}).
 
@@
 \subsection{Managing Connections}
+\label{subsection:noderuntimepublic}
+
 \begin{itemize}
@@
 \subsection{Executing the EFSM}
+\label{subsection:noderuntimeprivate}
+
+We revisit the concept of using discriminated unions to distinguish
+the type of EFSM state. The discriminated union is implemented by the
+\texttt{Implementation} API which wraps the developer's
+implementation of the  \texttt{Handler} API.
+
+It is sufficient for the runtime to only discriminate the type of state
+rather than the specific state. 
+we can define \textit{state-specific}
+details within the discriminated union
+
 \begin{itemize}
@@
 
-\subsection{Handling Message Sends}
-\begin{itemize}
-\item very short (just for completeness in the report) -- messages are serialised JS objects (in JSON notation) sent and decoded on the other end, type-correct by construction
-\end{itemize}
+\subsection{Sending Messages}
+Message Types are defined as interfaces, which
+are represented by objects. By convention,
+we serialise messages into \textit{JavaScript Object Notation}
+(or JSON) \cite{json} using the built-in \texttt{JSON.stringify()}
+method.
+
+They are decoded using the same interface schema on the receiving end
+using \texttt{JSON.parse()}: whilst the method return type is the
+dynamic \lstonelinejs{any} type, 
+we guarantee type safety by construction
+as we performed the serialisation in the first place, so
+we can safely interpret the deserialised content 
+using a concrete type.
 
@@
 \subsection{Handling Termination}
-\begin{itemize}
-\item design choice that the client is the one that closes the connection because of the centralised role of the server
-\item forward reference that this gives opportunity to extend support for general protocols, as in those protocols, the server's terminal state doesn't imply it's terminal for everyone else
-\end{itemize}
+WebSocket connections should be closed when the session terminates,
+and both the browser endpoint and the server endpoint are capable
+of closing connection. As we generate code for both, we define
+a convention that the browser endpoint will initiate the WebSocket
+close event.

### 2020-06-04 report/sections/server/node/efsm.tex
Report: progress on chapter 4 runtime

@@
 which defines the \texttt{type}
-discriminant property internally. This explains why the successor
+discriminant property internally (via inheritance).
+This explains why the successor
 state encodings in 
@@
 were defined as such.
+The type \texttt{Implementation.Type} is the discriminated union,
+which is used by the runtime.
 
@@
 abstract class IReceive { type: 'Receive' = 'Receive'; ... }
+abstract class ITerminal { type: 'Terminal' = 'Terminal'; ... }
 
 export namespace Implementation {
+	export type Type = ISend | IReceive | ITerminal;
 
@@
 	...
+
 };

### 2020-06-04 report/sections/server/node/runtime.tex
Report: progress on chapter 4 runtime

@@
 
-We define the session runtime in Svr.ts,
-named after the \trole{Svr} endpoint.
+We define the session runtime for the \trole{Svr} endpoint
+of the \tprotocol{Adder} protocol in Svr.ts,
+named after the endpoint.
 It exposes a \textbf{public API}
@@
 joined the session.
+
 The role of the public API is to manage incoming connections and
@@
 \begin{figure}[!h]
-\begin{lstlisting}[language=javascript,tabsize=2,title=Svr.ts]
+\begin{lstlisting}[language=javascript,title=Svr.ts]
 export class Svr {
@@
 }
+
+// Not exported to developer
+class Session {
+	private wss: WebSocket.Server;
+	private initialState: Implementation.S51;
+	private roleToSocket: RoleToSocket;
+	...
+}
 \end{lstlisting}
-\captionof{lstlisting}{Public API of Session Runtime for Server Endpoint}
+\captionof{lstlisting}{Public and Private API of Session Runtime for Server Endpoint}
 \label{lst:noderuntimepublicapi}
@@
 
-We revisit the concept of using discriminated unions to distinguish
-the type of EFSM state. The discriminated union is implemented by the
-\texttt{Implementation} API which wraps the developer's
-implementation of the  \texttt{Handler} API.
+The \texttt{Session} class executes the EFSM.
+We define a transition function, \texttt{next()}, 
+parameterised by the current state,
+which invokes the handler defined by the develoepr
+and performs the required channel actions for non-terminal states:
 
-It is sufficient for the runtime to only discriminate the type of state
-rather than the specific state. 
-we can define \textit{state-specific}
-details within the discriminated union
+\begin{itemize}
+\item 
+For \textbf{send} states, the handler will
+return the label and payload to be sent, along with the
+successor state implementation. The transition function
+should construct and send the message, 
+and transition to the successor state.
+
+\item
+For \textbf{receive} states, we change the
+message event listener on the WebSocket to
+pass the incoming message to the handler.
+The handler will return the successor state,
+which the runtime can transition to.
+
+\end{itemize}
+
+We conceptualise this in \cref{lst:noderuntimesimple}.
+The discriminated union lets the runtime figure out
+the type of the current state.
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript,tabsize=2]
+next(state: Implementation.Type) {
+	switch (state.type) {
+		case 'Send': {
+			const [label, payload, succ]: (*@\hl{???}@*) = state.handler;	 (*@\label{line:noderuntimebadtype1}@*)
+			this.send((*@\hl{???}@*), label, payload); (*@\label{line:noderuntimeunknownrole}@*)		
+			return this.next(succ);
+		}
+		case 'Receive': {
+			this.wss.onmessage = ({ data }) => {
+				const { label, payload } = JSON.parse(data) as (*@\hl{???}@*); (*@\label{line:noderuntimebadtype2}@*)
+				const succ: (*@\hl{???}@*) = state.handler[label](...payload); (*@\label{line:noderuntimebadtype3}@*)
+				return this.next(succ);
+			}
+		}
+		case 'Terminate': { return; }
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{Conceptual EFSM Transition Function 
+for Server-Side Endpoint}
+\label{lst:noderuntimesimple}
+\end{figure}
+
+However, we still face problems with resolving types, as highlighted.
+Just because we know that the current state is a send state,
+we do not know \textit{which} particular state it is,
+so we cannot accurately type the handler (\cref{line:noderuntimebadtype1}).
+The same problem is amplified for the receive state:
+we need to know the specific receive state in order
+to correctly serialise the message (\cref{line:noderuntimebadtype2})
+and interpret the successor state (\cref{line:noderuntimebadtype3}).
+We see another problem with handling send states:
+because we do not know the specific send state, we do not know which
+participant to send the message to (\cref{line:noderuntimeunknownrole}).
+
+These all reduce to the same core problem: the runtime
+needs to know the specific state at compile-time\footnote{
+Whether TypeScript ``compiles'' or ``transpiles''
+(or even \textit{``transcompiles''} \cite{transcompiles}) to JavaScript
+is not relevant to our work; we stick with compilation
+and keep our terminology consistent.}.
+We solve this through \textit{runtime polymorphism} instead, since
+the specific type of \texttt{state} is known at runtime.
+For each type of state, we define a common API that can be invoked
+by the EFSM transition function. To achieve runtime polymorphism,
+each concrete state must provide a specific implementation: 
+this motivates our design for defining the discriminated union
+using abstract classes with abstract methods 
+(\cref{lst:nodeabstractclass}).
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript,tabsize=2]
+abstract class ISend {
+	type: 'Send' = 'Send';
+	abstract performSend(
+		next: EfsmTransitionHandler,
+		sendMessage: (role: Roles.Peers, label: string, payload: any[])
+			=> void,
+	): void;
+};
+
+abstract class IReceive {
+	type: 'Receive' = 'Receive';
+	abstract prepareReceive(
+		next: EfsmTransitionHandler,
+		register: (from: Roles.Peers, messageHandler: MessageHandler)
+			=> void,
+	): void;
+};
+\end{lstlisting}
+\captionof{lstlisting}{\dots}
+\label{lst:nodeabstractclass}
+\end{figure}
+
+Using this approach,
+we can pass the transition function and channel actions 
+from the \texttt{Session} runtime class
+to the individual state \texttt{Implementation} classes:
+these are also generated by \fancyname{NodeTS}, so we
+guarantee linear usage of channel resources by construction as well.
+This \textit{significantly} simplifies the design of the runtime
+(\cref{lst:noderuntime}),
+because the transition function no longer needs to know the type 
+of the specific state at compile-time.
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript,tabsize=2]
+next(state: Implementation.Type) {
+	switch (state.type) {
+		case 'Send':
+			return state.performSend(this.next, this.send);
+		case 'Receive':
+			return state.prepareReceive(this.next, this.register);
+		case 'Terminal':
+			return;
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{Final EFSM Transition Function for Server-Side
+Endpoint}
+\label{lst:noderuntime}
+\end{figure}
+
+By passing the specific functions defined in the runtime
+to the individual EFSM states, we can visualise our runtime
+as a \textit{message passing} abstraction 
+(\cref{fig:noderuntimeefsm}): the runtime uses the common
+\texttt{performSend()} or \texttt{prepareReceive()} API
+to delegate to the specialised implementation, which will in turn
+ask the runtime to perform specialised channel actions using
+the parameterised methods, and finally delegate back to
+the runtime to transition to the specific successor state.
+
+We illustrate
 

### 2020-06-04 report/sections/server/node/runtime.tex
Report: progress on chapter 4 runtime

@@
 
-We illustrate
+We elaborate on how this mechanism
+handles the sending and receiving of messages in
+\cref{subsection:noderuntimesend,subsection:noderuntimereceive}
+respectively.
 
@@
 \subsection{Sending Messages}
+\label{subsection:noderuntimesend}
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript,tabsize=2]
+export class S54 extends ISend {
+	constructor(private handler: Handler.S54) { super(); }
+
+	performSend(
+		next: EfsmTransitionHandler,
+		sendMessage: (role: Roles.Peers, label: string, payload: any[])
+			=> void
+	) {
+		const [label, payload, successor] = this.handler;
+
+		sendMessage(Roles.Peers.Client, label, payload);
+		return next(successor);
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{\dots}
+\label{lst:nodesend}
+\end{figure}
+
+\subparagraph{Sending through WebSockets}
 Message Types are defined as interfaces, which
@@
 method.
+
+\begin{lstlisting}[language=javascript]
+sendMessage(role: Roles.Peers, label: string, payload: any[]) {
+	this.roleToSocket[role].send(JSON.stringify({
+		label, payload
+	});
+}
+\end{lstlisting}
 
@@
 
-\subsection{Handling Message Receives}
+\subsection{Receiving Messages}
+\label{subsection:noderuntimereceive}
 \begin{itemize}
@@
 close event.
+As a result, we do nothing for this state at \cref{lst:noderuntme}.

### 2020-06-04 report/sections/server/node/runtime.tex
Report: progress on chapter 4 runtime

@@
 		next: EfsmTransitionHandler,
-		sendMessage: (role: Roles.Peers, label: string, payload: any[])
+		send: (role: Roles.Peers, label: string, payload: any[]) 
 			=> void,
@@
 
+\begin{remark}
+We need to be careful when passing \textit{instance} methods
+as \textit{function} arguments 
+-- namely, the semantics of \lstonelinejs{this} is different.
+In short, we have to \textit{explicitly}
+\texttt{bind()} the \texttt{Session} object
+to the instance methods that we pass as function arguments:
+
+\begin{lstlisting}[language=javascript,tabsize=2]
+// Inside Session class constructor...
+this.next = this.next.bind(this);
+this.send = this.send.bind(this);
+this.register = this.register.bind(this);
+\end{lstlisting}
+
+Not doing so will result in \lstonelinejs{this} taking
+a different value 
+(either the global object or \lstonelinejs{undefined}).
+\end{remark}
+
 We elaborate on how this mechanism
@@
 respectively.
-
-\begin{itemize}
-\item implementation details delegated to the actual implementation states
-\item all non-terminal states will `return' the successor implementation -- very difficult to resolve in the type-checker in the runtime (give example of what it may look like) -- solve by giving each state implementation the `advance' function to respect the event-driven nature of everything, so they can call it on completion
-\item visualise the interaction between runtime and states as a sequence diagram for message passing
-\item constructor binds all methods (that will be passed to other components0 to `this' because of how javascript works with respect to the scoping of `this' (give short example -- or ignore, if this is in the background for typescript)
-\item advance -- use discriminated union to figure out what to pass to the state (send -- sendMessage, receive -- register)
-\end{itemize}
 
@@
 
+This is rather straightforward: 
+we show the generated code in \cref{lst:nodesend}.
+
 \begin{figure}[!h]
@@
 		next: EfsmTransitionHandler,
-		sendMessage: (role: Roles.Peers, label: string, payload: any[])
+		send: (role: Roles.Peers, label: string, payload: any[]) 
 			=> void
@@
 		const [label, payload, successor] = this.handler;
-
-		sendMessage(Roles.Peers.Client, label, payload);
+		send(Roles.Peers.Client, label, payload); (*@\label{line:nodesendrole}@*)
 		return next(successor);
@@
 
+We get the label, payload and successor implementation
+directly from the handler implemented by the developer, 
+accurately typed by how we define the handler API in EFSM.ts.
+The developer does not need to specify which role
+to send to: this is a sensible design choice, as we know this
+from the Scribble protocol, so we do not need the developer 
+to specify separately.
+As a result, we generate the code to send the message
+to the correct role (\cref{line:nodesendrole}).
+We use the \texttt{send()} method passed down
+by the runtime to commit our communication action:
+the runtime will handle how to serialise the message and perform
+the send. We guarantee that \texttt{send()} is called
+\textbf{exactly once} by construction, thus channel linearity
+is never violated.
+Finally, we use the parameterised EFSM transition handler
+to notify the rutime which specific state to transition to.
+
 \subparagraph{Sending through WebSockets}
-Message Types are defined as interfaces, which
-are represented by objects. By convention,
+We define message structures as interfaces, which
+are represented by objects. 
+By convention in \fancyname{SessionTS},
 we serialise messages into \textit{JavaScript Object Notation}
@@
 \begin{lstlisting}[language=javascript]
-sendMessage(role: Roles.Peers, label: string, payload: any[]) {
+send(role: Roles.Peers, label: string, payload: any[]) {
 	this.roleToSocket[role].send(JSON.stringify({
@@
 \label{subsection:noderuntimereceive}
+
+We need to update the message event listener on the WebSocket
+to use the developer's handler -- 
+specific to the \textit{current state} -- 
+to process the message.
+Our approach is to keep the WebSocket message event listener
+untouched, but define it in a way that allows \textit{dynamic} behaviour.
+We walk through the concept implemented in (\cref{lst:noderuntimewsmsg}):
+
+\begin{enumerate}
+\item 
+\texttt{Session} keeps track of the \textit{current}
+message receive handler (\cref{line:noderuntimehandler}).
+The \texttt{?} syntax denotes it is an \textit{optional}
+type: not every state is a receive state, so there does not \textit{have} to
+be an active message handler.
+
+\item
+The receive handler does \textit{not} need a specialised type
+(\cref{line:msghandler}). The receive handler is defined in
+the \texttt{Implementation} class of the concrete receive state,
+so it will deserialise the message to the correct form.
+
+\item
+The \texttt{register()} method (\cref{line:register})
+is passed to the \texttt{Implementation} class of the concrete
+receive state, which will construct the message handler
+around the developer's handler implementation and register it
+with the runtime.
+
+\item
+When a message is received from the channel,
+we dynamically process it with 
+the current registered handler (\cref{line:callhandler}).
+By MPST-safety, the incoming message must correspond
+to the structure expected by the current receive state,
+so correctness is preserved.
+We encapsulate this dynamic behaviour in an instance method
+and bind it as an event listener (\cref{line:wsbind})
+for the WebSocket connection
+of each non-server endpoint.
+\end{enumerate}
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript,tabsize=2]
+type MessageHandler = (message: any) => void; (*@\label{line:msghandler}@*)
+
+class Session {
+	private handler?: MessageHandler; (*@\label{line:noderuntimehandler}@*)
+
+	constructor(...) {
+		...
+		Object.values(this.roleToSocket).
+			.forEach(ws => ws.onmessage = this.receive.bind(this)); (*@\label{line:wsbind}@*)
+		this.handler = undefined;
+	}
+
+	register(handler: MessageHandler) { this.handler = handler; } (*@\label{line:register}@*)
+
+	receive({ data }: WebSocketMessage) {
+		const handler = this.handler;
+		this.handler = undefined;
+		handler?.(data); (*@\label{line:callhandler}@*)
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{Approach to Dynamic WebSocket Message Event Listener}
+\label{lst:noderuntimewsmsg}
+\end{figure}
+
+We also show the generated code for the \texttt{Implementation}
+class of the receive state in \cref{lst:nodereceive}
+-- this should appear consistent
+with the explanation above.
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript,tabsize=2]
+export class S51 extends IReceive {
+
+	constructor(private handler: Handler.S51) { super(); }
+
+	prepareReceive(
+		next: EfsmTransitionHandler,
+		register: (from: Roles.Peers, messageHandler: MessageHandler)
+			=> void
+	) {
+		const messageHandler = (message: any) => {
+			const decoded = JSON.parse(message) as Message.S51;
+			switch (decoded.label) {
+				case Labels.S51.ADD: {
+					const successor = 
+						this.handler[decoded.label](...decoded.payload);
+					return next(successor);
+				}
+				case Labels.S51.QUIT: {
+					// Same as above...		
+				}
+			}            
+		}
+		register(Roles.Peers.Client, messageHandler);
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{\dots}
+\label{lst:nodereceive}
+\end{figure}
+
+Ideally, a more succinct (and direct) representation would be
+\begin{lstlisting}[language=javascript]
+const successor = this.handler[decoded.label](...decoded.payload);
+return next(successor);
+\end{lstlisting}
+
+But this expresses a type dependency between \texttt{label}
+and \texttt{payload}, which as discussed, cannot be implemented.
+But message structures \textit{precisely} define a discriminated
+union, so we handle this with a switch statement,
+at the cost of having the same code in each case -- but
+TypeScript does infer different types in each case statement,
+so the duplicated code does serve a functional purpose.
+
+Returning to \cref{lst:noderuntimewsmsg},
+note that the type of the \texttt{handler} property is \textit{optional}
+-- this hints at a consistency problem: 
+\textit{how do we know that \texttt{this.handler} is set when
+a message is received?}
+
 \begin{itemize}
@@
 close event.
-As a result, we do nothing for this state at \cref{lst:noderuntme}.
+As a result, we do nothing for this state at \cref{lst:noderuntime}.

### 2020-06-04 report/sections/server/node/runtime.tex
Report: progress on chapter 4 receive edge case

@@
 Ideally, a more succinct (and direct) representation would be
-\begin{lstlisting}[language=javascript]
+\begin{lstlisting}[language=javascript,tabsize=2]
+(message: any) => {
+	const decoded = JSON.parse(message) as Message.S51;
 	const successor =  this.handler[decoded.label](...decoded.payload);
 	return next(successor);
+}
 \end{lstlisting}
@@
 But this expresses a type dependency between \texttt{label}
-and \texttt{payload}, which as discussed, cannot be implemented.
-But message structures \textit{precisely} define a discriminated
-union, so we handle this with a switch statement,
-at the cost of having the same code in each case -- but
-TypeScript does infer different types in each case statement,
-so the duplicated code does serve a functional purpose.
+and \texttt{payload}, which as discussed 
+(\cref{subsubsection:dependenttypes}), cannot be implemented.
+However message structures \textit{precisely} 
+define a discriminated union 
+(the label acts as the discriminant 
+to distinguish between payload types),
+so we handle this with a switch statement,
+at the cost of having the same code in each case --
+TypeScript does infer the correct specific type in each case statement,
+so code duplication here does serve a functional purpose.
 
@@
 note that the type of the \texttt{handler} property is \textit{optional}
--- this hints at a consistency problem: 
+-- this hints at a problem: 
 \textit{how do we know that \texttt{this.handler} is set when
-a message is received?}
+a message is received?} The types imply that we \textit{do not},
+and this is indeed the case.
+In fact, when we consider a \textit{multiparty} context,
+our approach with receive handler registration 
+using an optional value actually \textit{fails} to guarantee correctness.
+We motivate the problem with a worked
+example.
+
+\begin{example}[``Out-of-order'' message receives]
+Recall that Node.js is a single-threaded event loop runtime, so
+when a message arrives, 
+the \texttt{onmessage} event is \textit{queued},
+and current execution is \textit{not pre-empted}.
+
+Now consider a multiparty session specified by the global type
+
+\[ A \to S: \tmsg{M1(string)}. B \to S: \tmsg{M2(number)}. \texttt{end} \]
+
+Given that \trole{S} is the server endpoint,
+we describe a possible execution flow for the protocol
+that breaks our implementation:
+
+\begin{enumerate}
+\item 
+\trole{S} transitions to its initial state, ``receive from \trole{A}''.
+The receive handler for \tmsg{M1} is registered.
+
+\item
+\tmsg{M2} arrives at \trole{S}, so the \texttt{onmessage} handler
+is queued. This is \textit{perfectly plausible}: there
+is no causal relation between \tmsg{M1} and \tmsg{M2}.
+
+\item
+\tmsg{M1} arrives at \trole{S}, so the \texttt{onmessage} handler
+is queued.
+
+\item
+The \texttt{onmessage} handler for \tmsg{M2} is executed.
+The registered handler expects \tmsg{M1(string)},
+but it is callled with \tmsg{M2(number)}, which raises
+a runtime type error.
+\end{enumerate}
+
+This exposes two problems: 
+\textbf{(1)} the 
+\end{example}
 

### 2020-06-04 report/sections/introduction/contributions.tex
Report: small fixes
Report: minor changes

@@
 
-In \cref{chap:p2p}, we introduce the \fancyname{TwoBuyer} multiparty
+In \cref{chap:p2p}, we introduce the \tprotocol{Two Buyer} multiparty
 protocol to motivate our proposal of providing the server role with 
@@
 \paragraph{\cref{part:evalconcl}}
-In \cref{chap:eval}, we evaluate our work through case studies of communication
-protocols implemented as webservices. Through examples of binary sessions,
+In \cref{chap:eval}, we evaluate our work through 
+case studies of communication protocols 
+implemented as webservices. Through examples of binary sessions,
 multiparty sessions and routed multiparty sessions, we demonstrate how the

### 2020-06-04 report/sections/server/node/runtime.tex
Report: complete chapter  4 runtime

@@
 \end{lstlisting}
-\captionof{lstlisting}{\dots}
+\captionof{lstlisting}{Class Definitions for \texttt{Implementation} API}
 \label{lst:nodeabstractclass}
@@
 \begin{lstlisting}[language=javascript,tabsize=2]
-// Inside Session class constructor...
+// Inside the Session class constructor...
 this.next = this.next.bind(this);
@@
 \end{lstlisting}
-\captionof{lstlisting}{\dots}
+\captionof{lstlisting}{Generated Code for \texttt{Implementation} API for
+Send State}
 \label{lst:nodesend}
@@
 the current registered handler (\cref{line:callhandler}).
-By MPST-safety, the incoming message must correspond
-to the structure expected by the current receive state,
-so correctness is preserved.
 We encapsulate this dynamic behaviour in an instance method
@@
 	receive({ data }: WebSocketMessage) {
-		const handler = this.handler;
-		this.handler = undefined;
+		const handler = this.handler; (*@\label{line:callhandlerstart}@*)
+		this.handler = undefined; (*@\label{line:callhandler2}@*)
 		handler?.(data); (*@\label{line:callhandler}@*)
@@
 \end{lstlisting}
-\captionof{lstlisting}{\dots}
+\captionof{lstlisting}{Generated Code for \texttt{Implementation} API for
+Receive State}
 \label{lst:nodereceive}
@@
 Ideally, a more succinct (and direct) representation would be
-\begin{lstlisting}[language=javascript,tabsize=2]
+\begin{lstlisting}[language=javascript]
 (message: any) => {
 	const decoded = JSON.parse(message) as Message.S51;
-	const successor =  this.handler[decoded.label](...decoded.payload);
+	const successor = 
+		this.handler[decoded.label](...decoded.payload);
 	return next(successor);
@@
 But this expresses a type dependency between \texttt{label}
-and \texttt{payload}, which as discussed 
+and \texttt{payload} which, as discussed 
 (\cref{subsubsection:dependenttypes}), cannot be implemented.
@@
 
-\[ A \to S: \tmsg{M1(string)}. B \to S: \tmsg{M2(number)}. \texttt{end} \]
+\[
+A \to S: \text{\lstonelinejs{M1(string)}}. 
+B \to S: \text{\lstonelinejs{M2(number)}}. \texttt{end} 
+\]
 
-Given that \trole{S} is the server endpoint,
-we describe a possible execution flow for the protocol
+Suppose \trole{S} is the server endpoint.
+We describe a possible execution flow for the protocol
 that breaks our implementation:
@@
 \item 
-\trole{S} transitions to its initial state, ``receive from \trole{A}''.
+\trole{S} transitions to its initial state, 
+``receive \lstonelinejs{M1(string)} from \trole{A}''.
 The receive handler for \tmsg{M1} is registered.
@@
 The \texttt{onmessage} handler for \tmsg{M2} is executed.
-The registered handler expects \tmsg{M1(string)},
-but it is callled with \tmsg{M2(number)}, which raises
+The registered handler expects \lstonelinejs{M1(string)},
+but it is called with \lstonelinejs{M2(number)}, which raises
 a runtime type error.
@@
 
-This exposes two problems: 
-\textbf{(1)} the 
+This exposes a problem: 
+the order of message \textit{arrivals} may not
+correspond with the order of \textit{receiving} messages as
+specified in the protocol,
+\textit{so} the message may arrive before its
+corresponding handler is registered.
 \end{example}
 
-\begin{itemize}
-\item motivate the edge case that message can arrive in the websocket (in succession) before the handler is registered
-\item motivate the edge case that message can arrive `out of protocol order' 
-\item explain the double queue system and how that provides consistency -- emphasising that this works because of the single-threaded typescript runtime
-\end{itemize}
+We observe that message arrivals do not have to be causally related.
+However, if we consider a similar \textit{binary} example
+\[
+A \to S: \text{\lstonelinejs{M1(string)}}. 
+A \to S: \text{\lstonelinejs{M2(number)}}. \texttt{end}
+\]
+then \tmsg{M1} \textit{must} arrive before \textit{M2},
+since this is sent through the WebSocket connection between 
+\trole{A} and \trole{S}, and FIFO guarantees are respected 
+for each individual WebSocket connection. 
+We visualise the possible orders of message receive events
+in \cref{fig:nodereceivecompare}.
 
-\begin{figure}[!ht]
+\begin{figure}[!h]
 \centering
@@
 
+We observe that defining the handler using an optional type
+is insufficient. We need a similar mechanism for handling messages
+waiting for handlers, and we cannot assume an ordering on the arrival
+of messages that are not causally related.
+
+We proceed to generalise our approach from one optional-type
+handler to two mappings:
+\textbf{(1)} a mapping from endpoint to message queues\footnote{
+TypeScript arrays have built-in $O(1)$ time complexity
+\texttt{shift()} and \texttt{push()} operations, which
+can be used as a queue.}, and
+\textbf{(2)} a mapping from endpoint to handler queues.
+To simply put, if an incoming message is waiting for its handler,
+it gets enqueued in the message queue labelled by the sender of the message;
+when the handler is created, it pops the message off the queue 
+and directly processes it; the same logic applies for a handler waiting
+for its message.
+We could have used a mapping from endpoint to optional type,
+but queue operations elegantly hides the mechanics of
+\cref{line:callhandlerstart,line:callhandler2,line:callhandler}
+in \cref{lst:noderuntimewsmsg}.
+
+We outline the changes made to the \texttt{Session} class,
+as shown in \cref{lst:nodesession}:
+
+\begin{enumerate}
+\item We construct types for these two mappings 
+(\cref{line:mapped1,,line:mapped2}), 
+using a generic mapped typed defined in
+\cref{lst:noderolesmappedtypes}.
+
+\item Empty queues are initialised for both mappings
+in the constructor (\cref{line:queue1,,line:queue2}).
+
+\item Each endpoint has a different \texttt{onmessage}
+event listener, which will interact with the message queue
+and handler queue corresponding to that endpoint. 
+We achieve this by changing the \texttt{receive()} method
+to be parameterised on the role instead 
+(\cref{line:receivefunc}),
+so it \textit{generates} an event listener 
+(\cref{line:receivefuncret})
+tailored for receiving messages
+from that particular role.
+
+\item
+The \texttt{register()} method now also takes the role
+as a parameter (\cref{line:newregister}) in order to check the 
+corresponding pair of queues.
+\end{enumerate}
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript,tabsize=2]
+type RoleToMessageQueue = Roles.PeersToMapped<any[]>; (*@\label{line:mapped1}@*)
+type RoleToHandlerQueue = Roles.PeersToMapped<MessageHandler[]>; (*@\label{line:mapped2}@*)
+type RoleToSocket = Roles.PeersToMapped<WebSocket>;
+
+class Session {
+	...
+	private roleToSocket: RoleToSocket;
+	private messageQueue: RoleToMessageQueue;
+	private handlerQueue: RoleToHandlerQueue;
+
+	constructor(...) {
+		...
+		Object.values(Roles.Peers).forEach(role => {
+			const socket = this.roleToSocket[role];
+			socket.onmessage = this.receive(role).bind(this);
+		});
+
+		this.messageQueue = { [Roles.Peers.Client]: [], }; (*@\label{line:queue1}@*)
+		this.handlerQueue = { [Roles.Peers.Client]: [], }; (*@\label{line:queue2}@*)
+
+		this.next(initialState);	
+	}
+
+	receive(from: Roles) { (*@\label{line:receivefunc}@*)
+		return ({ data }) => { (*@\label{line:receivefuncret}@*)
+			// Array.shift() can return undefined if empty
+			const handler = this.handlerQueue[from].shift();
+			if (handler !== undefined) {
+				handler(data);
+			} else {
+				this.messageQueue[from].push(data);			
+			}
+		}
+	}
+
+	register(from: Roles.Peers, handler: MessageHandler) { (*@\label{line:newregister}@*)
+		const message = this.messageQueue[from].shift();
+		if (message !== undefined) {
+			handler(message);		
+		} else {
+			this.handlerQueue[from].push(data);		
+		}
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{Modified \texttt{Session} class to correctly
+handle message receive events}
+\label{lst:nodesession}
+\end{figure}
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript]
+// Inside the Roles namespace...
+export type PeersToMapped<Value> = { [Role in Peers]: Value };
+\end{lstlisting}
+\captionof{lstlisting}{Generic Mapped Type for Roles}
+\label{lst:noderolesmappedtypes}
+\end{figure}
+
+Each endpoint must have consistency between
+its message queue and handler queue.
+We get the consistency from the simple fact that Node.js 
+is a single-threaded runtime and execution is never pre-empted,
+so there is no need to worry about atomic queue operations or
+locking data structures.
+
+We walk through how this design addresses the problems
+in the previous example.
+
+\begin{example}[Revisiting ``out-of-order'' message receives]
+Consider the same multiparty session specified by the global type
+
+\[
+A \to S: \text{\lstonelinejs{M1(string)}}. 
+B \to S: \text{\lstonelinejs{M2(number)}}. \texttt{end} 
+\]
+
+We show that our modified implementation addresses the problem:
+
+\begin{enumerate}
+\item 
+\trole{S} transitions to its initial state,
+``receive \lstonelinejs{M1(string)} from \trole{A}''.
+\textit{The message queue for \trole{A} is empty,}
+so the receive handler for \tmsg{M1} is registered 
+\textit{under the handler queue for \trole{A}.}
+
+\item
+\tmsg{M2} arrives at \trole{S}, so the \texttt{onmessage}
+handler is queued.
+
+\item
+\tmsg{M1} arrives at \trole{S}, so the \texttt{onmessage} handler
+is queued.
+
+\item
+The \texttt{onmessage} handler for \tmsg{M2} is executed.
+\textit{The handler queue for \trole{B} is empty, 
+so \tmsg{M2} is added to the message queue for \trole{B}}.
+
+\item
+The \texttt{onmessage} handler for \tmsg{M1} is executed.
+\textit{The handler queue for \trole{A} is non-empty,
+so the handler is popped off the front of the queue 
+and processes \tmsg{M1}.}
+
+\item
+\trole{S} transitions to the successor state, 
+``receive \lstonelinejs{M2(number)} from \trole{B}''.
+\textit{The message for \trole{B} is non-empty,
+so \tmsg{M2} is popped off the queue and 
+processed by the handler.}
+\end{enumerate}
+
+This execution is free of communication mismatches.
+\end{example}
+
 \subsection{Handling Termination}
@@
 of closing connection. As we generate code for both, we define
-a convention that the browser endpoint will initiate the WebSocket
-close event.
-As a result, we do nothing for this state at \cref{lst:noderuntime}.
+a convention that the browser endpoint will close 
+the WebSocket connection,
+so we do nothing for this state at \cref{lst:noderuntime}.

### 2020-06-04 report/sections/server/node/alt.tex
Report: complete chapter 4 alternative designs

@@
 and expose communication APIs (e.g. send and/or receive) 
-that respect the permitted state transitions as instance methods.
+for permitted state transitions.
+We would still a session runtime to provide
+abstractions over the WebSocket APIs and handle the 
+``out-of-order'' arrival of non-causally-related messages.
+
+We conceptualise this alternative design (\cref{fig:nodealt})
+using the \tprotocol{One Adder} protocol (\cref{subfig:oneadder}), 
+a simplified version of the \tprotocol{Adder} protocol 
+but with \textit{curried} addition and without recursion.
+An implementation using the alternative API design
+(\cref{subfig:oneadderalt}) could have the state transition 
+API return a tuple of payload and continuation,
+and we could express the event-driven nature of a receive action
+using built-in async/await concurrency primitives.
+But ultimately, exposing channel resources without a linear type system
+means that the developer could violate channel linearity
+(i.e. by un-commenting \cref{line:nodebreaklinear}), 
+so we would need runtime checks as part of the session runtime.
 
@@
 \caption{The \tprotocol{OneAdder} Protocol}
+\label{subfig:oneadder}
 \end{subfigure}
@@
 \caption{Alternative API Design}
-%\label{...}
+\label{subfig:oneadderalt}
 \end{subfigure}
@@
 });
-
 \end{lstlisting}
 \caption{Current API Design}
-%\label{...}
 \end{subfigure}
-\captionof{figure}{\dots}
-%\label{...}
+\captionof{figure}{Comparing Alternative \fancyname{NodeMPST} 
+API Design using \tprotocol{One Adder} Protocol}
+\label{fig:nodealt}
 \end{figure}
 
-\begin{itemize}
-\item exposing channel resources like the java example -- less idiomatic (not-event-driven), need runtime linearity checks like the java example
-\item `hardcoded' switch case runtime like the TSOP papers -- doesn't best leverage the type system
-\end{itemize}
+Looking at alternative designs for the runtime,
+we could adopt the approach presented in \cite{TSOP},
+which eliminates the need for our \texttt{Implementation} wrapper API
+by pushing the workload back to the runtime in the form of a
+dense switch statement enumerating all EFSM states.
+This may simply the developer API (by removing the need for
+the \texttt{Implementation} wrapper), but results in a
+session runtime that increases in complexity as the size of the EFSM
+increases.

### 2020-06-04 report/sections/server/node/challenges.tex
Report: complete chapter 4 sans connection

@@
 \label{section:nodechallenges}
-\begin{itemize}
-\item session type implementation on the node runtime
-\item need to provide consistency under the single-threaded event loop model
-\item provide idiomatic event-driven APIs
-\item provide static guarantees where possible -- communication mismatch and channel linearity
-\item typescript's type system is gradual, structural and non-linear -- session type api generation examples that provide static guarantees on channel usage build upon target language type system's support for linear resources
-\end{itemize}
 
+Our goal with \fancyname{NodeMPST} is to
+generate session type implementations that
+provide as much \textbf{static} communication safety guarantees
+as possible,
+in a manner suitable for both the TypeScript language 
+and the single-threaded event loop model in Node.js
 As outlined in \cref{section:typescript}, TypeScript's type system
 is neither affine nor linear, so we need to work around the 
-limitations of the language to enforce linear usage of channel resources.
-Similarly, we need to 
-the language does not provide
-built-in constructs for enforcing linear channel usage.
+language's limitations to enforce linear channel usage.
+
+We also aim to provide the developer with
+idiomatic event-driven APIs commonly found in
+server-side TypeScript development.

### 2020-06-04 report/sections/server/node/limitations.tex
Report: complete chapter 4 sans connection

@@
 
-\begin{itemize}
-\item verbose
-\item state identifiers are in the API
-\item scoping achieved through nested objects
-\item performance aspect with class-based handlers
-\end{itemize}
+Providing \textbf{static} communication safety guarantees have
+come at a cost of providing relatively verbose APIs to the developer,
+which introduces a learning curve. In particular,
+the distinction between the \texttt{Implementation} and
+\texttt{Handler} API should be an internal detail, but
+the developer still needs to use the \texttt{Implementation} API
+in their logic.
+
+Our APIs also rely on state identifiers, which also
+does not help code readability, as it is neither apparent nor
+self-documenting what \texttt{S51} refers to -- some existing work
+\cite{Hybrid2016} augment the state identifiers with the channel
+action and labels involved (e.g. 
+\texttt{S51_Receive__Add_Number__Quit_String}).
+
+Using handler-style APIs also mean that we require
+nested scoping to propagate values between states,
+which does not scale elegantly as the levels of nesting
+increase. For example, the following implementation of
+the \tprotocol{One Adder} protocol would be invalid if
+we define the handlers for the continuations separately,
+because the received payload are no longer in scope
+(\cref{line:notinscope}).
 
@@
 const second = new S6({ NUM2: (y) => sum, });
-const sum = [Labels.S7.SUM, ???, new S5()];
+const sum = [Labels.S7.SUM, (*@\label{line:notinscope}\hl{???}@*), new S5()];
 \end{lstlisting}
+
+However, the developer may choose to propagate values
+between states through persistent storage APIs (such as
+storing values in a database and retrieving them for
+handlers of continuation states), so our generated APIs
+do not strictly enforce scoping to be the only way to
+propagate values in the application logic.

### 2020-06-04 report/sections/server/node/runtime.tex
Report: complete chapter 4 sans connection

@@
 				case Labels.S51.QUIT: {
-					// Same as above...		
+					const successor = 
+						this.handler[decoded.label](...decoded.payload);
+					return next(successor);	
 				}

### 2020-06-05 report/sections/server/node/limitations.tex
Report: complete chapter 4

@@
 handlers of continuation states), so our generated APIs
-do not strictly enforce scoping to be the only way to
+do not \textit{strictly} enforce scoping to be the only way to
 propagate values in the application logic.

### 2020-06-05 report/sections/server/node/runtime.tex
Report: complete chapter 4

@@
 It exposes a \textbf{public API}
-(\cref{lst:noderuntimepublicapi})
 with seams for the developer to pass in the WebSocket server 
@@
 (i.e. their handler implementations).
-It is the responsibility of the developer to construct the
+It is developer's responsibility to construct the
 WebSocket server and set it up to listen for incoming connections.
@@
 
-The role of the public API is to manage incoming connections and
-wait for all participants to join the session 
-(\cref{subsection:noderuntimepublic}), before
-handing off to the private API to execute the EFSM
-(\cref{subsection:noderuntimeprivate}).
-
-\begin{figure}[!h]
-\begin{lstlisting}[language=javascript,title=Svr.ts]
+\begin{lstlisting}[language=javascript]
 export class Svr {
@@
 \end{lstlisting}
-\captionof{lstlisting}{Public and Private API of Session Runtime for Server Endpoint}
-\label{lst:noderuntimepublicapi}
-\end{figure}
+
+The role of the public API is to manage incoming connections and
+wait for all participants to join the session 
+(\cref{subsection:noderuntimepublic}), before
+handing off to the private API to execute the EFSM
+(\cref{subsection:noderuntimeprivate}).
 
@@
 
-\begin{itemize}
-\item use set and partial object to keep track of pending connections
-\item ws event listeners
-\item forward reference that we need to manage cancellations here too
-\end{itemize}
+The constructor of the public API class sets up
+the framework for mapping incoming WebSocket connections to
+participants. 
+The main challenge is to wait for all participants to 
+connect to the server endpoint before EFSM execution begins.
+We address this by defining an internal protocol for managing
+session joining -- since we generate the runtime for both server 
+and browser endpoints, we can implement this in a way that is 
+transparent to the developer.
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript,tabsize=2]
+const waiting: Set<Roles.Peers> = new Set([Roles.Peers.Client]); (*@\label{line:initwaiting}@*)
+
+const roleToSocket: Partial<RoleToSocket> = { (*@\label{line:initroletosocket}@*)
+	[Roles.Peers.Client]: undefined,
+};
+
+const onSubscribe = ({ data, target: socket }) => {
+	const { connect: role } = JSON.parse(data) 
+		as Message.ConnectRequest; (*@\label{line:connectionrequest}@*)
+	if (!waiting.has(role)) { return socket.close(); } (*@\label{line:occupied}@*)
+
+	roleToSocket[role] = socket; (*@\label{line:connectwsbind}@*)
+	waiting.delete(role);
+
+	if (waiting.size === 0)
+		new Session(wss, roleToSocket as RoleToSocket, initialState); (*@\label{line:newsession}@*)
+};
+
+wss.addEventListener('connection', ws => {
+	ws.onmessage = onSubscribe; (*@\label{line:wslisteneroverride}@*)
+});
+\end{lstlisting}
+\captionof{lstlisting}{\dots}
+\label{lst:nodeconnect}
+\end{figure}
+
+We show this in \cref{lst:nodeconnect} and walk through
+the main parts:
+
+\begin{enumerate}
+\item
+The server keeps track of the participants that 
+have yet to join the session -- this is initialised to the
+complete set of
+non-server endpoints at the start (\cref{line:initrolesocket}).
+
+\item
+Browser endpoints request to join the session
+by sending a \textit{connection request} with the role
+identifier as payload (\cref{line:connectionrequest}). 
+We generate role enums for browser targets
+in the same way, so the server can correctly
+interpret the message.
+We listen to connection requests by overriding the
+\texttt{onmessage} event listener for every new connection
+(\cref{line:wslisteneroverride}).
+
+\item
+If the role is already occupied, then the server
+responds by closing the connection (\cref{line:occupied}).
+
+\item
+Otherwise, the role is not occupied, so the server binds
+the WebSocket (which the message was received from) to the role
+(\cref{line:connectwsbind}).
+This is accumulated in an interface type mapping each
+role to an \textit{optional}\footnote{
+TypeScript provides utility types for common type transformations:
+\texttt{Partial<T>} constructs a type with all properties
+of \texttt{T} set to optional.
+} 
+WebSocket property
+-- for roles who have
+not yet connected, we do not know the WebSocket binding, so the
+WebSocket for these roles are \lstonelinejs{undefined}.
+
+\item
+When the server is no longer waiting for any participants,
+it notifies all other roles through the bounded WebSocket
+connections that the session will start, and delegates
+EFSM execution to the private API by constructing
+an instance of the \texttt{Session} class (\cref{line:newsession}).
+The notification process is managed by the \texttt{Session} class.
+
+\begin{lstlisting}[language=javascript,tabsize=2,numbers=none]
+Object.values(roleToSocket).forEach(socket => {
+	socket.send(JSON.stringify(Message.ConnectConfirm));
+});
+\end{lstlisting}
+
+\item
+We define the interfaces and factories for connection messages 
+in the generated EFSM.ts file under
+the \texttt{Message} namespace.
+
+\begin{lstlisting}[language=javascript,tabsize=2,numbers=none]
+// Inside the Message namespace...
+export interface ConnectRequest { connect: Role.Peers };
+export const ConnectConfirm = { connected: true };
+\end{lstlisting}
+\end{enumerate}
 
@@
 parameterised by the current state,
-which invokes the handler defined by the develoepr
-and performs the required channel actions for non-terminal states:
+The constructor of the \texttt{Session} class
+explicitly calls \texttt{next()} with the 
+initial state implementation provided by the developer
+to start EFSM execution.
+\texttt{next()} invokes the handler defined by the developer
+and performs the required channel actions for non-terminal states.
 
@@
 
-These all reduce to the same core problem: the runtime
-needs to know the specific state at compile-time\footnote{
-Whether TypeScript ``compiles'' or ``transpiles''
-(or even \textit{``transcompiles''} \cite{transcompiles}) to JavaScript
-is not relevant to our work; we stick with compilation
-and keep our terminology consistent.}.
-We solve this through \textit{runtime polymorphism} instead, since
-the specific type of \texttt{state} is known at runtime.
-For each type of state, we define a common API that can be invoked
-by the EFSM transition function. To achieve runtime polymorphism,
-each concrete state must provide a specific implementation: 
-this motivates our design for defining the discriminated union
-using abstract classes with abstract methods 
-(\cref{lst:nodeabstractclass}).
-
 \begin{figure}[!h]
@@
 
-Using this approach,
-we can pass the transition function and channel actions 
-from the \texttt{Session} runtime class
-to the individual state \texttt{Implementation} classes:
-these are also generated by \fancyname{NodeTS}, so we
-guarantee linear usage of channel resources by construction as well.
-This \textit{significantly} simplifies the design of the runtime
-(\cref{lst:noderuntime}),
-because the transition function no longer needs to know the type 
-of the specific state at compile-time.
+These all reduce to the same core problem: the runtime
+needs to know the specific state at compile-time\footnote{
+Whether TypeScript ``compiles'' or ``transpiles''
+(or even \textit{``transcompiles''} \cite{transcompiles}) to JavaScript
+is not relevant to our work; we stick with compilation
+and keep our terminology consistent.}.
+We solve this through \textit{runtime polymorphism} instead, since
+the specific type of \texttt{state} is known at runtime.
+For each type of state, we define a common API that can be invoked
+by the EFSM transition function. To achieve runtime polymorphism,
+each concrete state must provide a specific implementation: 
+this motivates our design for defining the discriminated union
+using abstract classes with abstract methods 
+(\cref{lst:nodeabstractclass}).
 
@@
 \end{figure}
+
+Using this approach,
+we can pass the transition function and channel actions 
+from the \texttt{Session} runtime class
+to the individual state \texttt{Implementation} classes:
+these are also generated by \fancyname{NodeTS}, so we
+guarantee linear usage of channel resources by construction as well.
+This \textit{significantly} simplifies the design of the runtime
+(\cref{lst:noderuntime}),
+because the transition function no longer needs to know the type 
+of the specific state at compile-time.
 
@@
 \item Empty queues are initialised for both mappings
-in the constructor (\cref{line:queue1,,line:queue2}).
+(\cref{line:queue1,,line:queue2}).
 

### 2020-06-05 report/sections/server/node/runtime.tex
Report: complete chapter 4

@@
 \end{lstlisting}
-\captionof{lstlisting}{\dots}
+\captionof{lstlisting}{Handling Connections in Server Endpoint}
 \label{lst:nodeconnect}
@@
 complete set of
-non-server endpoints at the start (\cref{line:initrolesocket}).
+non-server endpoints at the start (\cref{line:initwaiting}).
 
@@
 (\cref{line:connectwsbind}).
-This is accumulated in an interface type mapping each
+This is accumulated in an interface type, mapping each
 role to an \textit{optional}\footnote{
@@
 not yet connected, we do not know the WebSocket binding, so the
-WebSocket for these roles are \lstonelinejs{undefined}.
+WebSocket for these roles are \lstonelinejs{undefined},
+as initialised at the start (\cref{line:initroletosocket}).
 

### 2020-06-05 report/sections/introduction/contributions.tex
Report: minor structural changes

@@
 and communication. Detailed lemmas and proofs in this section can
-be found in \cref{appendix:proofs}.
+be found in \cref{section:proofs}.
 
@@
 
+\begin{remark}
 This project was published as \emph{Generating Interactive WebSocket 
@@
 \cite{PLACES2020}.
+\end{remark}

### 2020-06-05 report/sections/server/react.tex
Report: restructure directory for React section

--- 2020-05-25 report/sections/server/node.tex
@@
-\chapter{\fancyname{NodeMPST}: Back-End Session Type Web Development}
+\chapter{\fancyname{ReactMPST}: Front-End Session Type Web Development}
+\label{chap:react}
 
-\section{Motivation}
+In this chapter, we
+\dots
 
-\section{Approach}
-
-\section{State Encodings}
-
-\subsection{Send}
-
-\subsection{Receive}
-
-\subsection{Terminal}
-
-\section{EFSM Execution Runtime}
-
-\section{Alternative Designs and Limitations}
+\input{sections/server/react/challenges}
+\input{sections/server/react/approach}
+\input{sections/server/react/efsm}
+\input{sections/server/react/runtime}
+\input{sections/server/react/alt}
+\input{sections/server/react/limitations}
 
 \section{Summary}
+\begin{itemize}
+\item static guarantees on avoiding communication mismatch
+\item static guarantees on affine channel usage
+\item leverage the react framework to express the model types approach from Fowler 2020 appropriately, to provide static guarantees, by construction, that the channel actions made available at each state respect the protocol, and by definition, multiparty session type safety
+\end{itemize}

### 2020-06-05 report/sections/server/react/approach.tex
Report: restructure directory for React section

--- 2020-03-12 PLACES_2020/encoding.tex
@@
-\section{Encoding Session Types in TypeScript}
-% Design choices
-
-Developers can implement their application using APIs generated from the EFSM
-to guarantee correctness by construction.
-Our approach integrates the EFSM into the development workflow by encoding
-session types as TypeScript types.
-Communication over the WebSocket protocol introduces additional constraints:
-communication is always initiated in the front-end and driven by user interactions,
-whilst back-end roles can only accept connections.
-This motivates our design of encoding the session types differently for server
-(\cref{section:server}) and client (\cref{section:browser}) targets.
-
-\subsection{Server-Side API Generation}
-\label{section:server}
-
-% Francisco: mention that it corresponds to the messages for the Noughts and Crosses example too
-
-\begin{wrapfigure}{r}{0.6\textwidth}
-  \vspace{-5mm}
-  \begin{center}
-    \includegraphics[width=0.58\textwidth]{figures/efsm_svr.png}
-  \end{center}
-
-  \vspace{-5mm}
-  \captionof{figure}{EFSM for \texttt{Svr}.}
-  \label{fig:efsmsvr}
-\vspace{-1cm}
-\end{wrapfigure}
-
-We refer to the \texttt{Svr} EFSM (\cref{fig:efsmsvr}) as a running example in
-this section.
-For server-side targets, we encode EFSM states into TypeScript types and
-consider branching (receiving) and selection (sending) states separately.
-We assign TypeScript encodings of states to their state identifiers in the
-EFSM, providing syntactic sugar when referring to the successor state when
-encoding the current state.
-For any state $S$ in the EFSM, we refer to the TypeScript type alias of its
-encoding as $\llbracket S \rrbracket$.
-We outline the encoding below using examples from the
-\textit{Noughts and Crosses} game (\cref{lst:svr}).
-
-\paragraph{Branching State}
-We consider a receiving state as a unary branching state for conciseness.
-A branching state is encoded as an \textit{object literal}
-\cite{TypeScriptSpec} (a record type), with each branch $i \in I$ ($I$ denoting set of all
-branches), corresponding to a member field.
-A branch expecting to receive a message labelled $\texttt{label}_i$ carrying
-payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as an
-\textit{member field} named $\texttt{label}_i$ of function type
-$(payload:\texttt{T}_i) \to \llbracket S_i \rrbracket$.
-The developer implements a branching operation by passing callbacks for each
-branch, parameterised by the expected message payload type for that branch.
-
-\paragraph{Selection State}
-We consider a sending state as a unary selection state for conciseness.
-A selection state is encoded as a \textit{union type}
-\cite{TypeScriptSpec} of internal choice encodings: each internal choice $i \in
-I$ ($I$ denoting set of all choices), sending a message labelled
-$\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state
-$S_i$ is encoded as a \textit{tuple type} of \texttt{[Labels.label$_i$, T$_i$,
-  $\llbracket S_i \rrbracket$]}.
-The developer implements a selection operation by passing the selected label
-and payload to send in the message.
-We generate a \textit{string enum} (named \texttt{Labels}) wrapping the labels
-in the protocol.
-
-\begin{figure}[ht]
-\begin{lstlisting}[language=JavaScript]
-export type S13 = { Pos: (payload: Point) => S15 };
-export type S15 = [ Labels.Lose, Point, S16 ]
-                | [ Labels.Draw, Point, S17 ]
-                | [ Labels.Update, Point, S18 ];
-\end{lstlisting}
-\captionof{lstlisting}{Example encodings from \textit{Noughts and Crosses} \texttt{Svr} EFSM.}
-\label{lst:svr}
-\end{figure}
-
-% REVIEW:
-% Also, when describing Listing 2, it would be helpful to say a bit more explicitly that the client is supposed to implement S13, and the code in S13 returns an object of type S15.
-Then you can point to Listing 4 if they are coordinated.
-In the case of \cref{lst:svr}, the developer is expected to implement
-\texttt{S13} which handles the \texttt{Pos} message sent by \texttt{P1},
-and the code in \texttt{S13} returns a value of type \texttt{S15}, which
-corresponds to a selection of messages to send to \texttt{P2}. \Cref{lst:svrprotocol}
-illustrates how the developer may implement these types.
-
-We make a key design decision \textit{not} to expose communication channels in
-the TypeScript session type encodings to provide \textit{static} linearity
-guarantees (\cref{section:serverlinear}).
-Our encoding sufficiently exposes seams for the developer to inject their
-program logic, whilst the generated session API
-(\cref{section:serversessionapi}) handles the sending and receiving of
-messages.
-
-\subsubsection{Session Runtime}
-\label{section:serversessionapi}
-
-The generated code for our session runtime performs communication in a protocol-conformant manner, but
-does not expose these IO actions to the developer by delegating the
-aforementioned responsibilities to an inner class.
-% REVIEW: it is not clear what is
-% generated on the server-side and what actually is the session runtime
-The runtime executes the EFSM by keeping track of
-the current state (similar to the generated code in \cite{javatypestate})
-and only permitting the specified IO actions at the current state.
-The runtime listens to message (receiving) events on the communication channel,
-invokes the corresponding callback to obtain the value to send next, and
-performs the sending.
-The developer instantiates a session by constructing an instance of the
-session runtime class, providing the WebSocket endpoint URL (to open the
-connection) and the initial state (to execute the EFSM).
-
-\subsubsection{Linear Channel Usage}
-\label{section:serverlinear}
-Developers writing their implementation using the generated APIs 
-enjoy channel linearity by construction.
-Our library design prevents the two conditions detailed below:
-
-\paragraph{Repeated Usage}
-We do not expose channels to the developer, which makes \textit{reusing
-  channels} impossible.
-For example, to send a message, the generated API only requires the payload
-that needs to be sent, and the session runtime performs the send internally,
-guaranteeing this action is done \textit{exactly once} by construction.
-
-\paragraph{Unused Channels}
-The initial state must be supplied to the session runtime
-constructor in order to instantiate a session;
-this initial state is defined
-in terms of the successor states, which in turn has references to its
-successors and so forth.
-The developer's implementation will cover the terminal state
-(if it exists), and the
-session runtime guarantees this terminal state will be reached
-by construction.
+\section{Approach}
+\begin{itemize}
+\item leverage existing front-end frameworks that support reactive programming -- identify the parallel ideas that the UI element rendered on the screen needs to be reactive with respect to -- not only user interaction, but also channel actions.
+\item for example, UI should react to a receive state in the EFSM, and once the message is received on the channel, transition to a different UI with the supported actions
+\item main goal -- statically ensure that whatever IO actions are `present' on the window at any given time must respect the protocol
+\end{itemize}
 
 \subsection{The React Framework}
+\begin{itemize}
+\item use a class-based Counter example to explain the terms: props, state, uni-directional data flow, virtual DOM, reconciliation
+\end{itemize}
+
+\paragraph{Note: from PLACES}
 Our browser-side session type encodings for browser-side targets build upon the
@@
 
-\subsection{Browser-Side API Generation}
-
-\label{section:browser}
-
-\begin{wrapfigure}{R}{0.5\textwidth}
-  \begin{center}
-    \includegraphics[width=0.5\textwidth]{figures/efsm_p1.png}
-  \end{center}
-
-  \captionof{figure}{EFSM for \texttt{P1}.}
-  \label{fig:efsmp1}
-\end{wrapfigure}
-
-We refer to the \texttt{P1} EFSM (\cref{fig:efsmp1}) as a running example in
-this section.
-Preserving behavioural typing and channel linearity is challenging
-for browser-side applications due to EFSM transitions being triggered by user
-events:
-in the case of \textit{Noughts and Crosses}, once the user makes a move by
-clicking on a cell on the game board, this click event must be deactivated
-until the user's next turn, otherwise the user can click again and violate
-channel linearity.
-Our design goal is to enforce this statically through the generated APIs.
-
+\subsection{Model Types Revisited}
+\paragraph{Note: from PLACES paper}
 For browser-side targets, we extend the approach presented in \cite{MVU2019} on
@@
 given the supported IO actions at this state.
-
-\subsubsection{Model Types in React}
-
-\paragraph{State}
-An EFSM state is encoded as an \textit{abstract} React
-component.
-This is an abstract class to require the developer to provide their
-own view function, which translates conveniently to the \texttt{render()}
-function of React components.
-Our session runtime (\cref{section:clientruntime}) ``executes'' the EFSM and
-renders the current state.
-Upon transitioning to a successor state, the successor's view function will be
-invoked, as per the semantics expressed in \cite{MVU2019}.
-
-\paragraph{Model Transitions}
-Transitions are encoded as React component props onto the encoded states by the
-session runtime (\cref{section:clientruntime}).
-We motivate the design choice of not exposing channel resources to provide
-guarantees on channel usage.
-React components in TypeScript are
-\textit{generic} \cite{TypeScriptSpec}, parameterised by the permitted
-types of prop and state.
-The parameters allow us to leverage the TypeScript compiler to
-verify that the props for model transitions stay local to the state they are
-defined for.
-The model transitions for EFSMs are message send and receive.
-
-\subparagraph{Sending}
-We make the assumption that message sending is triggered by
-some user-driven UI event (e.g. clicking a button, pressing a key on the
-keyboard) which interacts with some DOM element.
-We could pass a
-\texttt{send()} function as a prop to the sending state, but the developer
-would be free to call the function multiple times which makes channel reuse
-possible.
-Instead, we pass a \textit{factory function} as a prop, which will,
-given an HTML event and an event handler function, return a fresh React
-component that binds the sending action on construction.
-So once the bound event is triggered, our session runtime executes the event
-handler function to obtain the payload to send, perform the send
-\textit{exactly once} and transition to (which, in practice, means render) the
-successor state.
-
-\begin{figure}[!h]
-\begin{lstlisting}[language=JavaScript, tabsize=4]
-// Inside some render() function..
-{board.map((row, x) => (
-	row.map((col, y) => {
-		const SelectPoint = this.props.Pos('click', (event: UIEvent) => {
-			event.preventDefault();
-			return { x: x, y: y };}
-		return <SelectPoint><td>.</td></SelectPoint>;
-});}
-\end{lstlisting}
-\captionof{lstlisting}{Model transition for message sending in
-\textit{Noughts and Crosses} \texttt{P1} implementation.}
-\label{lst:clientapp}
-\end{figure}
-
-We demonstrate the semantics using the \textit{Noughts and Crosses} example in
-\cref{lst:clientapp}.
-The session runtime passes the factory function \texttt{this.props.Pos} as a prop.
-For each x-y coordinate on the game board, we
-create a \texttt{SelectPoint} React component from the factory function (which
-reads ``build a React component that sends the \texttt{Pos} message with x-y
-coordinates as payload when the user clicks on it'') and we wrap a table cell
-(the game board is rendered as an HTML table) inside the \texttt{SelectPoint}
-component to bind the click event on the table cell.
-
-\subparagraph{Receiving}
-The React component for a receiving state is required to
-define a handler for each supported branch. 
-Upon a message receive event, the session runtime invokes the
-handler of the corresponding branch with the message payload and 
-renders the successor state upon completion.
-
-\subsubsection{Session Runtime}
-\label{section:clientruntime}
-
-The session runtime can be interpreted as an abstraction on top of the React
-VDOM that implements the EFSM by construction.
-The session runtime itself is a React component too, named after the endpoint
-role identifier:
-it opens the WebSocket connection to the server, keeps track of the current
-EFSM state as part of its React component state, and most importantly, renders
-the React component encoding of the active EFSM state.
-Channel communications are managed by the runtime, which allows it to render
-the successor of a receive state upon receiving a message from the channel.
-Similarly, the session runtime is responsible for passing the required props
-for model transitions to EFSM state React components.
-The session runtime component is rendered by the developer and requires, as
-props, the \textit{endpoint URL} (so it can open the connection) and a list of
-\textit{concrete state components}.
-
-The developer writes their own implementation of each state (mainly to
-customise how the state is rendered and inject business logic into state
-transitions) by extending the abstract React class components.
-The session runtime requires references to these concrete components in order to
-render the user implementation accordingly.
-
-\subsubsection{Affine Channel Usage}
-A limitation of our browser-side session type encoding is only being able to
-guarantee that channel resources are used \textit{at most once} as opposed to
-\textit{exactly once}.
-
-Communication channels are not exposed to the developer so multiple sends are
-impossible.
-This does not restrict the developer from binding the send action to exactly
-one UI event: for \textit{Noughts and Crosses}, we bind the \texttt{Pos(Point)}
-send action to each unoccupied cell on the game board, but the generated
-runtime ensures that, once the cell is clicked, the send is only performed once
-and the successor state is rendered on the DOM, so the channel resource used to
-send becomes unavailable.
-
-However, our approach \textit{does not} statically detect whether all
-transitions in a certain state are bound to some UI event.
-This means that it is possible for an implementation to \textit{not} handle
-transitions to a terminal state but still type-check, so we cannot prevent
-unused states. Equally, our approach does not prevent a client closing the browser, which would drop the connection.

### 2020-06-05 report/sections/server/react/challenges.tex
Report: restructure directory for React section

\section{Challenges}
\begin{itemize}
\item challenges of session typing user interface logic -- how does one formalise channel linearity? (give example of how complicated this actually is -- e.g. how to guarantee that, if a button is triggered for sending, that the action will not be triggered at other states?)
\item existing works use non-mainstream language -- motivate Fowler LINKS, PureScript ConcurUI, identify that it needs some form of linear typing mechanism or `sequential building' UI to provide the safety properties
\item challenge -- provide the same static communication safety guarantees for front-end logic written in typescript, in a way that is usable and idiomatic for the average typescript developer?
\end{itemize}

### 2020-06-05 report/sections/server/react/efsm.tex
Report: restructure directory for React section

\section{State Encodings}

\begin{itemize}
\item motivated by the idea of model types, each state is encoded as a react component, where the user provides a view function (i.e. render())
\item channel actions are abstracted by the runtime and passed to state implementations as props
\item the developer provides an implementation of the state, i.e. ``uses the APIs'', by extending the abstract base classes and providing their own implementations
\end{itemize}

\subsection{Send}
\begin{itemize}
\item make the assumption that channel send actions are triggered by UI elements -- for example, clicking a button or typing enter in a textbox
\item pre-session-types -- the developer will attach an event listener
\item enhanced factory approach -- given an event and a event handler, return a react component that binds the handler to that event, and injects the channel action; the developer can use this API by putting the UI elements that represent the event action as children.
\item the event handler is conditionally parameterised by the event type (achievable by typescript conditional types -- eg, if event is click, then the event handler parameter is a React.MouseEvent), and the return type is the payload type
\item the abstract base class will set up the factory methods to be used by the developer
\end{itemize}

\subsubsection{Alternative Designs}
\begin{itemize}
\item not sure whether to put this in the subsection above so it flows better?
\item naive implementation is to pass the send() function as a prop for the developer to incorporate in their event listener -- this is most flexible, but we cannot provide static guarantees
\item factory approach -- the runtime could pass a factory function which, takes an UI element and performs the binding; this is better, as channel resources are not exposed, but the adversarial-minded developer can later override the event listener in the DOM etc
\end{itemize}

\subsection{Receive}
\begin{itemize}
\item receive state is `shown' when waiting for a receive event
\item base class has abstract methods -- one for each possible receive -- so the user must implement these when they extend the abstract base class
\item each abstract method is named after the message label expected to be received, and parameterised by the payload
\item no channel actions are passed to the state -- instead, once the receive state is mounted on the DOM (i.e. rendered), the abstract base class will register the event handlers to the runtime (using the register method passed as props)
\item when the message arrives, the runtime will invoke the registered function (+ forward reference the next section with the edge case that the message arrives before handler registration occurs)
\end{itemize}

\subsection{Terminal}
\begin{itemize}
\item by convention and design choice, the client gracefully closes the connection with the server role
\end{itemize}

### 2020-06-05 report/sections/server/react/limitations.tex
Report: restructure directory for React section

\section{Limitations}
\begin{itemize}
\item affine channel usage guarantees -- cannot statically guarantee that the factory is attached to an event visible on the DOM
\item decoupling communication state management from business logic state management -- runtime design does not permit passing non-communication-related data as props; need to use other methods for propagating applicaton data (such as React Context API or Redux)
\item the point above might not be a `limitation', but more of an encouragement to change practices because it makes sense, architecturally, to separate concerns
\end{itemize}

### 2020-06-05 report/sections/server/react/runtime.tex
Report: restructure directory for React section

\section{Runtime}
\begin{itemize}
\item public API - accept websocket endpoint URL, a mapping of state to user implementation, and UI elements to render when waiting for connection to occur, and connection failure
\item private API - carry out the session (just like Node)
\end{itemize}

\subsection{Session Joining}
\begin{itemize}
\item separate into two react components, since the websocket connection cannot be opened in the constructor by React StrictMode
\item `outer' class sets websocket to undefined, but upon componentDidMount, creates the actual connection
\item renders the `inner' class component upon websocket not being undefined, so it can be typed appropriately (otherwise, need to always use bang operator)
\item on connection, send join request message to the server, wait for serber response before executing the EFSM at the initial state
\end{itemize}

\subsection{Executing the EFSM}
\begin{itemize}
\item same concept as node -- implementation details delegated to the actual classes
\item constructor binds methods
\item advance takes the enum instead of the class (since unlike the node example, it is not instantiated) -- uses custom type guards to distinguish between the type of state
\end{itemize}

\subsection{Handling Message Sends}
\begin{itemize}
\item passes in the factory prop, which is a factoryfactory to be precise;
\item sendMessage is bound in the closure and sends the message in the same way as how Node codegen does, by convention
\item (if not already mentioned in the node section) send is asynchronous
\end{itemize}

\subsection{Handling Message Receives}
\begin{itemize}
\item same edge case as the back-end code generation, as message can arrive before handlers are registered
\item use the same two-queue mechanism to keep track of `what comes first' -- i.e. if message comes first, then put in message queue, so handler can pop off when registered; vice-versa, if handler is registered first, put in handler queue so message can be processed when that is received
\item importantly -- only one queue is required, as server-centric topologies only have the client talk to the server
\end{itemize}

\subsection{Handling Termination}
\begin{itemize}
\item close the websocket with a clean exit code
\end{itemize}

### 2020-06-05 report/sections/server/node/limitations.tex
Report: minor changes to chapter 4

@@
 does not help code readability, as it is neither apparent nor
-self-documenting what \texttt{S51} refers to -- some existing work
-\cite{Hybrid2016} augment the state identifiers with the channel
+self-documenting what \texttt{S51} refers to -- Hu and Yoshida
+\cite{Hybrid2016} augments the state identifiers with the channel
 action and labels involved (e.g. 

### 2020-06-05 report/sections/server/react.tex
Report: progress on chapter 5 challenges

@@
 
-In this chapter, we
-\dots
+In this chapter, we present \fancyname{ReactMPST},
+our session type API generation strategy for browser-side endpoints
+implemented using React \cite{React}.
+We will refer to the FSM of the \trole{Client}
+endpoint (\cref{fig:adderclientfsm})
+from the \tprotocol{Adder} protocol 
+throughout this chapter.
+
+\begin{figure}[!b]
+\centering
+\includegraphics[width=0.55\textwidth]{AdderClientFSM}
+\captionof{figure}{\trole{Client} Endpoint FSM
+in \tprotocol{Adder} protocol}
+\label{fig:adderclientfsm}
+\end{figure}
 

### 2020-06-05 report/sections/server/react/challenges.tex
Report: progress on chapter 5 challenges

@@
 \section{Challenges}
-\begin{itemize}
-\item challenges of session typing user interface logic -- how does one formalise channel linearity?
-(give example of how complicated this actually is -- e.g.
-how to guarantee that, if a button is triggered for sending, that the action will not be triggered at other states?)
-\item existing works use non-mainstream language -- motivate Fowler LINKS, PureScript ConcurUI, identify that it needs some form of linear typing mechanism or `sequential building' UI to provide the safety properties
-\item challenge -- provide the same static communication safety guarantees for front-end logic written in typescript, in a way that is usable and idiomatic for the average typescript developer?
-\end{itemize}
+Our goal with \fancyname{ReactMPST} is to
+generate session type implementations for the web browser.
+Session typing the user interface (UI) logic is inherently difficult --
+assuming channel actions are bound to user interface events 
+(i.e. clicking a button sends a message), how does one formalise
+channel linearity? How do we guarantee that, if the button
+triggers a send at some EFSM state, that it triggers not more than
+one channel action, \textit{and} the user cannot trigger the action
+at another EFSM state?
+
+Existing work (\cref{subsection:sessiontypewebdev})
+do tackle browser-side session typing.
+Fowler introduced the concept of model types to prevent
+channel linearity violation in his proposal for
+integrating session types with
+graphical user interface (GUI) programming \cite{MVU2020}, but
+it requires the linear type system available from
+the proprietary Links web programming language \cite{LINKS}.
+The session type-safe web development framework presented
+by King et al. \cite{PureScript2019} generates APIs for
+a functional target language in PureScript, and relies on
+the Concur UI framework that constructs UIs sequentially.
+
+We find these proposals to come at the cost of
+limiting developer productivity by adopting unconventional practices
+that may require a learning curve.
+Through our work, we aim to distil the key
+concepts from existing work that provide session type safety
+for web-based GUI programming, and implement them using
+mainstream front-end web development tools (e.g. TypeScript
+and React), to provide developers with an intuitive way to implement
+browser-side endpoints that guarantee communication safety.

### 2020-06-05 report/sections/server/react/approach.tex
Report: progress on chapter 5

@@
 \subsection{The React Framework}
-\begin{itemize}
-\item use a class-based Counter example to explain the terms: props, state, uni-directional data flow, virtual DOM, reconciliation
-\end{itemize}
+Our browser-side API generation strategy builds upon the 
+\emph{React.js} framework developed by Facebook \cite{React} for the
+\textit{Model-View-Controller} (MVC) architecture.
+React is widely used in industry to create scalable single-page
+TypeScript applications, so this makes our workflow beneficial in an
+industrial context.
 
-\paragraph{Note: from PLACES}
-Our browser-side session type encodings for browser-side targets build upon the
-\emph{React.js} framework, developed by Facebook \cite{React} for the
-\textit{Model-View-Controller} (MVC) architecture.
-React is widely used in industry to create scalable single-page TypeScript
-applications, and we intend for our proposed workflow to be beneficial in an
-industrial context.
-We introduce the key features of the framework.
+We introduce the key features of the framework
+through illustrating a web-based counter in \cref{lst:counter}.
+The browser shows a counter (initialised to zero) 
+and an ``Increment'' button:
+when the user clicks on the ``Increment'' button,
+the count is incremented and the UI shows the updated count.
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript,tabsize=2]
+type Props = { count: number };
+class Count extends React.Component<Props>{
+	render() { (*@\label{line:childrender}@*)
+		return <strong>{this.props.count}</strong>; (*@\label{line:childprops}@*)
+	}
+}
+
+type State = { count: number };
+class App extends React.Component<{}, State>{
+	constructor(props: {}) {
+		super(props);
+		this.state = { count: 0 }; (*@\label{line:parentstate}@*)
+	}
+
+	increment() { this.setState({ count: this.state.count + 1 }); (*@\label{line:parentsetstate}@*)
+
+	render() { (*@\label{line:parentrender}@*)
+		return (<div>
+			<button onClick={this.increment.bind(this)}>
+				Increment
+			</button>
+			<Count count={this.state.count} /> (*@\label{line:childcomponent}@*)
+		</div>);	
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{Simple Counter in React}
+\label{lst:counter}
+\end{figure}
+
+\subparagraph{Components}
+A \textit{component} is a reusable UI element which
+contains its own mark-up and logic.
+Components implement a \texttt{render()} method which returns
+a ReactNode, the smallest building blocks of a React application.
+This is analogue to the \textit{view} function in the MVU architecture.
+React uses \textit{JSX} syntax to interpolate TypeScript logic 
+(enclosed in curly braces)
+within HTML mark-up: 
+in \cref{line:childrender}, the \texttt{Count} component
+evaluates the TypeScript expression 
+\lstonelinejs{this.props.count} and renders it in bold on the web page.
+Components can render other components, which give rise to
+a tree of UI elements. \cref{line:parentrender} shows that our
+\texttt{Count} component is rendered by 
+another component, \texttt{App}.
+
+\subparagraph{Uni-directional Data Flow}
+User-defined components are derived from the abstract
+\texttt{React.Component} generic base class
+
+\subparagraph{Virtual DOM (VDOM) and Reconciliation}
+\dots
 
@@
 
+\begin{itemize}
+\item use a class-based Counter example to explain the terms: props, state, uni-directional data flow, virtual DOM, reconciliation
+\end{itemize}
+
+\paragraph{Note: from PLACES}
+Our browser-side session type encodings for browser-side targets build upon the
+\emph{React.js} framework, developed by Facebook \cite{React} for the
+\textit{Model-View-Controller} (MVC) architecture.
+React is widely used in industry to create scalable single-page TypeScript
+applications, and we intend for our proposed workflow to be beneficial in an
+industrial context.
+We introduce the key features of the framework.
+
 \subsection{Model Types Revisited}

### 2020-06-05 report/sections/server/react/approach.tex
Report: progress on chapter 5 approach

@@
 \section{Approach}
-\begin{itemize}
-\item leverage existing front-end frameworks that support reactive programming -- identify the parallel ideas that the UI element rendered on the screen needs to be reactive with respect to -- not only user interaction, but also channel actions.
-\item for example, UI should react to a receive state in the EFSM, and once the message is received on the channel, transition to a different UI with the supported actions
-\item main goal -- statically ensure that whatever IO actions are `present' on the window at any given time must respect the protocol
-\end{itemize}
+We motivate our approach from \cite{MVU2020} by extending
+their work on \textit{multiple model types} motivated by the
+\textit{Model-View-Update} architecture (MVU).
+An MVU application features a \textit{model} encapsulating application
+state, a \textit{view function} rendering the state on the 
+Document Object Model (DOM), and an
+\textit{update function} handling \textit{messages} produced by the
+rendered model to produce a new model.
+The concept of model types express type dependencies between these
+components: a \emph{model type} uniquely defines a \textit{view function},
+set of \textit{messages} and \textit{update function} -- rather than
+producing a new model, the update function defines valid transitions to
+other model types.
+
+We leverage the correspondence between model types and states in the EFSM:
+each state in the EFSM is a model type, the set of messages represent
+the possible (IO) actions available at that state,
+and the update function defines which successor state to transition to,
+given the supported IO actions at this state.
 
 \subsection{The React Framework}
-Our browser-side API generation strategy builds upon the 
-\emph{React.js} framework developed by Facebook \cite{React} for the
-\textit{Model-View-Controller} (MVC) architecture.
+We implement model types for the EFSM on top of the 
+\emph{React.js} framework developed by Facebook \cite{React}.
 React is widely used in industry to create scalable single-page
@@
 industrial context. 
+The framework defines a way for data to flow
+between UI elements, and empowers the UI to subscribe and
+``react'' to data changes.
+We aim to implement similar behaviour with respect to the EFSM:
+the UI should react to EFSM state transitions,
+so we can ensure -- preferably statically -- that the
+channel actions ``present'' on the browser at any given time
+are those permitted by the current EFSM state.
 
@@
 
-\subparagraph{Components}
+\paragraph{Components}
 A \textit{component} is a reusable UI element which
@@
 Components implement a \texttt{render()} method which returns
-a ReactNode, the smallest building blocks of a React application.
+a React element, the smallest building blocks of a React application.
 This is analogue to the \textit{view} function in the MVU architecture.
-React uses \textit{JSX} syntax to interpolate TypeScript logic 
+React uses the \textit{JSX} syntax extension \cite{JSX}
+to interpolate TypeScript logic 
 (enclosed in curly braces)
@@
 
-\subparagraph{Uni-directional Data Flow}
-User-defined components are derived from the abstract
-\texttt{React.Component} generic base class
+\paragraph{Uni-directional Data Flow}
+User-defined components derive from the abstract base class
+\texttt{React.Component<P, S>},
+which is an abstract base class with generic type parameters
+\texttt{<P, S>} for \textit{props} (short for properties) and 
+\textit{state} respectively.
 
-\subparagraph{Virtual DOM (VDOM) and Reconciliation}
-\dots
+The \texttt{App} component maintains \texttt{count} in its
+state (\cref{line:parentstate}). Clicking on the increment button
+updates the state (\cref{line:parentsetstate}), which invokes
+a re-render, so the UI ``reacts'' to state change.
 
-\paragraph{Components}
-A component is a reusable UI element which
-contains its own markup and logic.
-Components implement a \texttt{render()} function which returns a React
-element, the smallest building blocks of a React application, analogous to the
-view function in the MVU architecture.
-Components can keep \textit{state}s and the \texttt{render()} function is
-invoked upon a change of state.
+Data flows from parent components down to their children,
+in the form of props. 
+The \texttt{App} component passes the \texttt{count}
+from its state to the \texttt{Count} component 
+(\cref{line:childcomponent}), which accesses it via
+\lstonelinejs{this.props}. 
+Because \texttt{App} is re-rendered when the count is incremented,
+the \texttt{Count} child component will also be re-rendered
+with updated props.
 
-For example, a simple counter can be implemented as a component,
-with its \texttt{count} stored as state.
-When rendered, it displays a button which increments \texttt{count}
-when clicked and a \texttt{div} that renders the current
-\texttt{count}.
-If the button is clicked, the \texttt{count} is incremented, which triggers a
-re-rendering (since the state has changed), and the updated \texttt{count} is
-displayed.
+\paragraph{Virtual DOM (VDOM) and Reconciliation}
+The \texttt{render()} methods give the developer a declarative API
+to specify what should be rendered. React uses a 
+\textit{virtual DOM} abstraction, where the tree of React elements
+are rendered on the virtual DOM, and React internally
+runs a \textit{reconciliation} algorithm to update the browser DOM
+accordingly using minimal operations.
 
-Components can also render other components, which gives rise
-to a parent/child relationship between components.
-Parents can pass data to children as \textit{props} (short for properties).
-Going back to the aforementioned example, the counter component could
-render a child component \texttt{<StyledDiv count=\{this.state.count\} />} in
-its \texttt{render()} function, propagating the \texttt{count} from its state
-to the child.
-This enables reusability, and for our use case, gives control to the parent
-on what data to pass to its children (e.g. pass the payload of a received
-message to a child to render).
-
-\begin{itemize}
-\item use a class-based Counter example to explain the terms: props, state, uni-directional data flow, virtual DOM, reconciliation
-\end{itemize}
-
-\paragraph{Note: from PLACES}
-Our browser-side session type encodings for browser-side targets build upon the
-\emph{React.js} framework, developed by Facebook \cite{React} for the
-\textit{Model-View-Controller} (MVC) architecture.
-React is widely used in industry to create scalable single-page TypeScript
-applications, and we intend for our proposed workflow to be beneficial in an
-industrial context.
-We introduce the key features of the framework.
-
-\subsection{Model Types Revisited}
-\paragraph{Note: from PLACES paper}
-For browser-side targets, we extend the approach presented in \cite{MVU2019} on
-\textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU)
-architecture.
-% FEEDBACK: introduce MVU and model type
-An MVU application features a \textit{model} encapsulating application
-state, a \textit{view function} rendering the state on the Document Object Model (DOM), and an
-\textit{update function} handling \textit{messages} produced by the
-rendered model to produce a new model.
-The concept of model types express type dependencies between these
-components: a \emph{model type} uniquely defines a \textit{view function},
-set of \textit{messages} and \textit{update function} -- rather than
-producing a new model, the update function defines valid transitions to
-other model types.
-We leverage the correspondence between model types and states in the EFSM:
-each state in the EFSM is a model type, the set of messages represent
-the possible (IO) actions available at that state,
-and the update function defines which successor state to transition to,
-given the supported IO actions at this state.
+For example, the \texttt{<button>} will not be re-rendered
+on the browser DOM on every counter increment as it does not
+depend on the updated state.

### 2020-06-05 report/sections/server/react.tex
Report: complete chapter 5 approach

@@
 
-\section{Summary}
-\begin{itemize}
-\item static guarantees on avoiding communication mismatch
-\item static guarantees on affine channel usage
-\item leverage the react framework to express the model types approach from Fowler 2020 appropriately, to provide static guarantees, by construction, that the channel actions made available at each state respect the protocol, and by definition, multiparty session type safety
-\end{itemize}
+%\section{Summary}
+%\begin{itemize}
+%\item static guarantees on avoiding communication mismatch
+%\item static guarantees on affine channel usage
+%\item leverage the react framework to express the model types approach from Fowler 2020 appropriately, to provide static guarantees, by construction, that the channel actions made available at each state respect the protocol, and by definition, multiparty session type safety
+%\end{itemize}

### 2020-06-05 report/sections/server/react/approach.tex
Report: complete chapter 5 approach

@@
 each state in the EFSM is a model type, the set of messages represent
-the possible (IO) actions available at that state,
+the possible channel actions available at that state,
 and the update function defines which successor state to transition to,
-given the supported IO actions at this state.
+given the supported channel actions at this state.
 
-\subsection{The React Framework}
 We implement model types for the EFSM on top of the 
@@
 the UI should react to EFSM state transitions,
-so we can ensure -- preferably statically -- that the
+so we can \textbf{statically} ensure that the
 channel actions ``present'' on the browser at any given time
 are those permitted by the current EFSM state.
+We introduce the framework in \cref{subsection:react}.
+
+When executing \fancyname{SessionTS} to generate code
+for the \trole{Client} endpoint specifying the
+\texttt{browser} target, the developer obtains the following 
+groups of files:
+
+\begin{itemize}
+
+\item 
+\textbf{S[40-43].tsx:}
+Developer APIs for implementing EFSM states 
+(\cref{section:reactefsm});
+
+\begin{itemize}
+\item
+\textbf{EFSM.ts}, \textbf{Message.ts}, \textbf{Roles.ts:}
+Utility types for EFSM encoding;
+\end{itemize}
+
+\item 
+\textbf{Client.tsx:} 
+Session runtime for executing the EFSM 
+(\cref{section:reactruntime});
+
+\begin{itemize}
+\item
+\textbf{Session.ts}, \textbf{Types.ts:}
+Utility types for session runtime.
+\end{itemize}
+
+\end{itemize}
+
+\subsection{The React Framework}
+\label{subsection:react}
 
@@
 \lstonelinejs{this.props.count} and renders it in bold on the web page.
+
 Components can render other components, which give rise to
@@
 \texttt{Count} component is rendered by 
-another component, \texttt{App}.
+the \texttt{App} component.
 

### 2020-06-05 report/sections/server/react/alt.tex
Report: progress on chapter 5 limitations

--- 2020-06-05 report/sections/server/react/efsm.tex
@@
-\section{State Encodings}
-
-\begin{itemize}
-\item motivated by the idea of model types, each state is encoded as a react component, where the user provides a view function (i.e. render())
-\item channel actions are abstracted by the runtime and passed to state implementations as props
-\item the developer provides an implementation of the state, i.e. ``uses the APIs'', by extending the abstract base classes and providing their own implementations
-\end{itemize}
-
-\subsection{Send}
-\begin{itemize}
-\item make the assumption that channel send actions are triggered by UI elements -- for example, clicking a button or typing enter in a textbox
-\item pre-session-types -- the developer will attach an event listener
-\item enhanced factory approach -- given an event and a event handler, return a react component that binds the handler to that event, and injects the channel action; the developer can use this API by putting the UI elements that represent the event action as children.
-\item the event handler is conditionally parameterised by the event type (achievable by typescript conditional types -- eg, if event is click, then the event handler parameter is a React.MouseEvent), and the return type is the payload type
-\item the abstract base class will set up the factory methods to be used by the developer
-\end{itemize}
-
-\subsubsection{Alternative Designs}
+\section{Alternative Designs}
 \begin{itemize}
@@
 \end{itemize}
-
-\subsection{Receive}
-\begin{itemize}
-\item receive state is `shown' when waiting for a receive event
-\item base class has abstract methods -- one for each possible receive -- so the user must implement these when they extend the abstract base class
-\item each abstract method is named after the message label expected to be received, and parameterised by the payload
-\item no channel actions are passed to the state -- instead, once the receive state is mounted on the DOM (i.e.
-rendered), the abstract base class will register the event handlers to the runtime (using the register method passed as props)
-\item when the message arrives, the runtime will invoke the registered function (+ forward reference the next section with the edge case that the message arrives before handler registration occurs)
-\end{itemize}
-
-\subsection{Terminal}
-\begin{itemize}
-\item by convention and design choice, the client gracefully closes the connection with the server role
-\end{itemize}

### 2020-06-05 report/sections/server/react/efsm.tex
Report: progress on chapter 5 approach
Report: complete chapter 5 approach
Report: progress on chapter 5 limitations

@@
-\section{State Encodings}
+\section{EFSM Encoding}
+\label{section:reactefsm}
 
@@
 
-\subsubsection{Alternative Designs}
-\begin{itemize}
-\item not sure whether to put this in the subsection above so it flows better?
-\item naive implementation is to pass the send() function as a prop for the developer to incorporate in their event listener -- this is most flexible, but we cannot provide static guarantees
-\item factory approach -- the runtime could pass a factory function which, takes an UI element and performs the binding; this is better, as channel resources are not exposed, but the adversarial-minded developer can later override the event listener in the DOM etc
-\end{itemize}
-
 \subsection{Receive}

### 2020-06-05 report/sections/server/react/limitations.tex
Report: progress on chapter 5 limitations

@@
 \section{Limitations}
-\begin{itemize}
-\item affine channel usage guarantees -- cannot statically guarantee that the factory is attached to an event visible on the DOM
-\item decoupling communication state management from business logic state management -- runtime design does not permit passing non-communication-related data as props; need to use other methods for propagating applicaton data (such as React Context API or Redux)
-\item the point above might not be a `limitation', but more of an encouragement to change practices because it makes sense, architecturally, to separate concerns
-\end{itemize}
+As discussed, our EFSM encoding provides \textbf{affine} channel
+usage guarantees for send states as supposed to \textit{linear} usage
+guarantees. Suppose we bind a send action to a
+button click event -- strictly speaking,
+even if we guarantee that the button can be clicked exactly once,
+we cannot guarantee that the button will ever be clicked by the user.
+However, existing work \cite{PureScript2019,MVU2020} 
+share the same limitation, and we view this as an
+inherent limitation in front-end development, since there is
+a limit to the types of assumptions we can make about how the
+user interacts with the browser.
+Redefining the definition of channel linearity 
+for GUI programming use cases
+would be an interesting discussion.
+
+Our runtime implementation also limits the developer from
+passing data into their EFSM state component through props.
+This is because the runtime takes full control of which
+state component to render, and by definition, what props 
+are passed to the EFSM state components.
+However, we equally view this as a way to encourage developers
+to decouple communication state management from
+business logic state management, in the spirit
+of separating concerns.
+In \cref{chap:eval},
+we present examples of developer implementations
+that use separate abstractions (such as the React Context API or
+the third-party Redux library) for propagating 
+stateful application data in \cref{chap:eval}.

### 2020-06-05 report/sections/server/react/alt.tex
Report: progress on chapter 5 alternative designs

@@
 \section{Alternative Designs}
-\begin{itemize}
-\item not sure whether to put this in the subsection above so it flows better?
-\item naive implementation is to pass the send() function as a prop for the developer to incorporate in their event listener -- this is most flexible, but we cannot provide static guarantees
-\item factory approach -- the runtime could pass a factory function which, takes an UI element and performs the binding; this is better, as channel resources are not exposed, but the adversarial-minded developer can later override the event listener in the DOM etc
-\end{itemize}
+
+For send states, a simpler approach would be
+to provide the developer with a \texttt{send()} function 
+for each permitted selection.
+The factory approach would still apply -- the runtime could pass something
+of the form
+
+\begin{lstlisting}[language=javascript, numbers=none]
+declare function buildSend<T>(label: string): (payload: T) => void
+\end{lstlisting}
+
+as a prop to the send state component.
+By passing the payload type and label, the send state component
+can build a handler (bound to the payload type) that performs a
+send over the WebSocket when called.
+This provides more flexibility for the developer's implementation,
+but this clearly exposes channel resources and 
+comes at the cost of not being able to provide guarantees
+on affine channel usage.
+
+For receive states, \dots
+
+%
+%\begin{itemize}
+%\item not sure whether to put this in the subsection above so it flows better?
+%\item naive implementation is to pass the send() function as a prop for the developer to incorporate in their event listener -- this is most flexible, but we cannot provide static guarantees
+%\item factory approach -- the runtime could pass a factory function which, takes an UI element and performs the binding; this is better, as channel resources are not exposed, but the adversarial-minded developer can later override the event listener in the DOM etc
+%\end{itemize}

### 2020-06-06 report/sections/server/react/alt.tex
Report: complete chapter 5 alternative designs

@@
 as a prop to the send state component.
-By passing the payload type and label, the send state component
-can build a handler (bound to the payload type) that performs a
-send over the WebSocket when called.
+This allows the send state component to build a handler,
+bound to the payload type and label identifier, that performs a 
+send action when called with the payload value.
 This provides more flexibility for the developer's implementation,
@@
 
-For receive states, \dots
-
-%
-%\begin{itemize}
-%\item not sure whether to put this in the subsection above so it flows better?
-%\item naive implementation is to pass the send() function as a prop for the developer to incorporate in their event listener -- this is most flexible, but we cannot provide static guarantees
-%\item factory approach -- the runtime could pass a factory function which, takes an UI element and performs the binding; this is better, as channel resources are not exposed, but the adversarial-minded developer can later override the event listener in the DOM etc
-%\end{itemize}
+An alternative design for receive states
+would be to define message receive handlers on the
+\textit{successor} state.
+Using \cref{fig:adderclientfsm} as an example,
+the receive handler for \lstonelinejs{RES(number)}
+would be defined on state S51 instead -- in fact, it wouldn't
+be defined as a callback, but the runtime would simply pass
+the received payload to S51 via props.
+This allows the developer to render UI changes
+based on received messages more easily, but comes at the
+cost of less robust typing:
+the \tmsg{RES} prop would have to be an optional value,
+as an EFSM state can be preceded by more than one receive state,
+so the developer has to implement logic (i.e. checking
+for \lstonelinejs{undefined} props) to determine whether
+the predecessor state was a receive state (and if so,
+\textit{which} receive state) and render accordingly.

### 2020-06-06 report/sections/server/codegen/workflow.tex
Report: progress on chapter 5 EFSMs

@@
     def __init_subclass__(cls, *, target):
-        CodeGenerationStrategy.target_to_strategy[target] = cls (*@\label{line:register}@*)
+        CodeGenerationStrategy.target_to_strategy[target] = cls (*@\label{line:registerstrategy}@*)
         return super().__init_subclass__()
@@
 such that each derived class must provide the target name to
-``register'' with the base class (\cref{line:register}),
+``register'' with the base class (\cref{line:registerstrategy}),
 and the base class keeps an internal

### 2020-06-06 report/sections/server/react/alt.tex
Report: progress on chapter 5 EFSMs

@@
 be defined as a callback, but the runtime would simply pass
-the received payload to S51 via props.
+the received payload to S40 via props.
 This allows the developer to render UI changes

### 2020-06-06 report/sections/server/react/efsm.tex
Report: progress on chapter 5 EFSMs

@@
 
-\begin{itemize}
-\item motivated by the idea of model types, each state is encoded as a react component, where the user provides a view function (i.e. render())
-\item channel actions are abstracted by the runtime and passed to state implementations as props
-\item the developer provides an implementation of the state, i.e. ``uses the APIs'', by extending the abstract base classes and providing their own implementations
-\end{itemize}
+We encode each EFSM state as a React component.
+This encoding is consistent with the semantics of model types,
+as outlined in \cref{table:efsmmodeltypes}.
+\fancyname{ReactMPST} generates an abstract React component class
+for each EFSM state. The developer implements the generated API
+by extending the abstract class to define their own
+view function and implement any abstract methods
+required by the EFSM state.
 
-\subsection{Send}
-\begin{itemize}
-\item make the assumption that channel send actions are triggered by UI elements -- for example, clicking a button or typing enter in a textbox
-\item pre-session-types -- the developer will attach an event listener
-\item enhanced factory approach -- given an event and a event handler, return a react component that binds the handler to that event, and injects the channel action; the developer can use this API by putting the UI elements that represent the event action as children.
-\item the event handler is conditionally parameterised by the event type (achievable by typescript conditional types -- eg, if event is click, then the event handler parameter is a React.MouseEvent), and the return type is the payload type
-\item the abstract base class will set up the factory methods to be used by the developer
-\end{itemize}
+\renewcommand{\arraystretch}{1.5}
+\begin{table}[!ht]
+\centering
+\begin{tabularx}{\linewidth}{||c|X||}
+\hline
+Model Type Property & \multicolumn{1}{c||}{Equivalent Abstraction for EFSM} \\
+\hline\hline
+View Function & 
+Each EFSM state is a React component 
+with its own \texttt{render()} method. \\ 
+\hline
+Set of Messages & The permitted channel actions are defined either as 
+component props (for send states) 
+or abstract instance methods (for receive states). \\
+\hline
+Update Function & EFSM state components are rendered by a session runtime,
+so they notify the runtime to trigger a transition. \\
+\hline
+\end{tabularx}
+\captionof{table}{Implementing Model Types as React Components}
+\label{table:efsmmodeltypes}
+\end{table}
+\renewcommand{\arraystretch}{1}
 
-\subsection{Receive}
+Our EFSM encoding for browser-side endpoints remain
+consistent with those for server-side endpoints, in the sense that
+channel resources are abstracted away from the APIs generated
+for the developer, so it is impossible to trigger
+channel actions \textit{more than once} at any given EFSM state,
+by construction.
+
+\subsection{Send States}
+
+For React components implementing send states,
+we need to guarantee that the channel actions on the DOM
+precisely correspond to the permitted transitions.
+We require that send actions can only be invoked
+by user-triggered events on UI elements, such as clicking a button or
+typing in a textbox -- this generally aligns
+with idiomatic front-end web development practices.
+
+Without \fancyname{ReactMPST},
+an implementation of state S40 from \cref{fig:adderclientfsm}
+may resemble the following, where the developer attaches
+event listeners on UI elements to trigger channel actions:
+
+\begin{lstlisting}[language=javascript,numbers=none]
+<button onClick={ev => (*@\hl{Add}@*)(this.state.num1, this.state.num2)}>
+	Add
+</button>
+<button onMouseOver={ev => (*@\hl{Quit}@*)(this.state.message)}>
+	Quit
+</button>
+\end{lstlisting}
+
+If we directly provide the highlighted channel actions 
+(i.e. via props), we cannot guarantee linear usage in the code 
+without a linear type system.
+
+What we want to work towards is something similar to the following,
+where the highlighted component encapsulates the required information: 
+\textbf{(1)} what event to react to, 
+and \textbf{(2)} what payload to send when the event is triggered.
+\texttt{<AddOnClick>} should guarantee, by construction,
+that clicking on any child element will send the \tmsg{ADD}
+message with the specified payload.
+
+\begin{lstlisting}[language=javascript,numbers=none]
+(*@\hl{<AddOnClick payload=\{[this.state.num1, this.state.num2]\}>}@*)
+	<button>Add</button>
+(*@\hl{</AddOnClick>}@*);
+\end{lstlisting}
+
+This elegantly hides channel actions, but at the same time,
+we cannot make excessive assumptions about the 
+event (\texttt{onClick}) and UI element (\texttt{<button>})
+that the send action will be attached to --
+these should be left as implementation details that 
+the developer can customise for their application.
+
+Instead, we present a novel approach for binding
+channel actions to UI elements in a way that \textit{never
+exposes channel resources}:
+
+\begin{enumerate}
+\item
+The abstract send state React component defines
+\textbf{factory properties}
+that allow the developer to \textit{generate React components}
+for performing send actions, such that the API lets the developer
+specify which UI event should trigger the channel action,
+and the payload to send. It defines one factory property per
+permitted send action at that state.
+
+\item
+The runtime provides a \textit{factory method}
+which allows the abstract send state component to
+generate the aforementioned factory properties
+for the channel actions supported at that state.
+\end{enumerate}
+
+We walk through the implementation details by working backwards:
+showing how the developer would use the API first,
+before deriving the type signatures that we need to
+generate for the abstract send state component inherited by the developer.
+
+\subsubsection{Developer API}
+
+We show the developer's implementation 
+of a send state for the \trole{Client} endpoint
+from the \tprotocol{Adder} protocol in \cref{lst:adderclientsenddev}.
+Component state is an implementation detail 
+that is not related to EFSM execution, so we allow
+the developer to customise this through specifying an 
+interface (\cref{line:clientsendstatestate}).
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript,tabsize=2]
+type State = { num1: number, num2: number, message: string }; (*@\label{line:clientsendstatestate}@*)
+class InputWindow extends S40<State> {
+	render() {
+		const { num1, num2, message } = this.state;
+		const (*@\hl{Add}@*) = this.ADD('onClick', (*@\label{line:clientsendfactory1}@*)
+			(ev: React.MouseEvent) => [num1, num2] (*@\label{line:clientsendpayload1}@*)
+		);
+		const (*@\hl{Quit}@*) = this.QUIT('onMouseOver', (*@\label{line:clientsendfactory2}@*)
+			(ev: React.MouseEvent) => [message] (*@\label{line:clientsendpayload2}@*)
+		);		
+		return (<div>
+			// Omitting <input>s for entering numbers, message
+			(*@\hl{<Add>}@*)<button>Sum</button>(*@\hl{</Add>}@*)
+			(*@\hl{<Quit>}@*)<button>Quit</button>(*@\hl{</Quit>}@*)
+		</div>);
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{\dots}
+\label{lst:adderclientsenddev}
+\end{figure}
+
+The factory API for binding send actions to UI elements appear on
+\cref{line:clientsendfactory1,line:clientsendfactory2}.
+For example, \cref{line:clientsendfactory1} reads: 
+\textit{build a React component that sends the \tmsg{ADD}
+message with \texttt{[num1, num2]} as payload,
+when the user clicks on it}.
+\tmsg{ADD} is the factory API we generate: it takes an event identifier
+and an event handler which must return the payload 
+-- \textit{appropriately typed with respect to the Scribble protocol} -- 
+for sending the \tmsg{ADD} message.
+Swapping \cref{line:clientsendpayload1,line:clientsendpayload2}
+will result in a compile-time type error due to invalid 
+payload types, which is the intended behaviour.
+
+We will define the factory API implementation in a way that
+calls the event handler to obtain the payload,
+sends the payload message, and transitions to the successor state,
+such that the send action is never performed more than once.
+
+\subsubsection{Typing the Abstract Send Component}
+Referring back to \cref{lst:adderclientsenddev},
+a first attempt for typing \tmsg{ADD} could be
+
+\begin{lstlisting}[language=javascript,numbers=none]
+(ev: string, handler: (e: any) => [number, number]) 
+	=> Constructor<React.Component>
+\end{lstlisting}
+
+However, the type of \texttt{ev} depends on the value of
+\texttt{event} -- for example, an \texttt{onClick} event
+is handled by a \texttt{React.MouseEvent} handler.
+React defines the supported event handlers under the
+\texttt{React.DOMAttributes} interface (\cref{lst:reactdomattrs}).
+We can use an index type query to express the dependency,
+and type inference in conditional types\footnote{
+We can \lstonelinejs{infer} type variables inside the
+\lstonelinejs{extends} clause of a conditional type.
+Specifically, \lstonelinejs{FunctionArguments<T> = 
+T extends (...args: infer R) ? R : never}.
+}
+to extract the function argument types:
+
+\begin{lstlisting}[language=javascript,numbers=none]
+<K extends keyof React.DOMAttributes>
+(ev: K, handler: (e: FunctionArguments<React.DOMAttributes[K]>) 
+		=> [number, number])
+	=> Constructor<React.Component>)
+\end{lstlisting}
+
+Contrary to our discussions about modelling dependent types in
+\cref{subsection:nodeefsmimplementation}, we can do this here
+because: \textbf{(1)} we are using index type queries instead
+of conditional types, and \textbf{(2)} we know the type
+of \texttt{ev} at compile-time from the developer's implementation,
+so we are not resolving index type queries with union types.
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript,numbers=none]
+interface DOMAttributes<T> {
+	children?: ReactNode; (*@\label{line:reactdomattrsnonfunc}@*)
+	onChange?: FormEventHandler<T>;
+	onClick?: MouseEventHandler<T>;
+	...
+};
+\end{lstlisting}
+\captionof{lstlisting}{Snippet of the \texttt{React.DOMAttributes} Interface}
+\label{lst:reactdomattrs}
+\end{figure}
+
+However, \texttt{React.DOMAttributes} also contains
+non-function properties -- on \cref{lst:reactdomattrs}
+\cref{line:reactdomattrsnonfunc},  ``children'' is not a valid event.
+We define utility types under Types.ts 
+to extract function properties from an interface 
+(\cref{lst:extractfuncprops}): we first extract
+the \textit{names} of function properties into an union type 
+by mapping property names of non-function properties to 
+\lstonelinejs{never} (\cref{line:extractfuncprops1})
+and performing an index type query (\cref{line:extractfuncprops2})
+to obtain an union type of function property names\footnote{
+The union type technically contains \lstonelinejs{never}
+as the names for non-function properties, but 
+TypeScript removes \lstonelinejs{never} from union types
+in the same way that $A \vee \bot \equiv A$.
+}
+which removes all \lstonelinejs{never} constituents;
+then extract the properties from the interface that are
+indexed by the filtered names (\cref{line:extractfuncprops3})
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript,tabsize=2]
+type FunctionPropertyNames<T> = {
+	[K in keyof T]: T[K] extends Function | undefined ? K : never; (*@\label{line:extractfuncprops1}@*)
+}[keyof T]; (*@\label{line:extractfuncprops2}@*)
+type FunctionProperties<T> = Pick<T, FunctionPropertyNames<T>>; (*@\label{line:extractfuncprops3}@*)
+type DOMEvents = FunctionProperties<React.DOMAttributes<any>>;
+\end{lstlisting}
+\captionof{lstlisting}{\dots}
+\label{lst:extractfuncprops}
+\end{figure}
+
+We construct a generic type parameterised on the payload type, 
+which lets us define the \tmsg{ADD} and \tmsg{QUIT}
+factory property easily.
+
+\begin{lstlisting}[language=javascript,numbers=none]
+type EventHandler<Payload, K extends keyof DOMEvents> = 
+	(event: FunctionArguments<DOMEvents[K]>) => Payload;
+
+type SendComponentFactory<Payload> = <K extends keyof DOMEvents>
+	(event: K, handler: EventHandler<Payload, K>) => 
+		Constructor<React.Component>;
+\end{lstlisting}
+
+The factories are instantiated by the abstract class
+as \lstonelinejs{protected} properties, in order to allow 
+access by the developer's subclass implementation.
+Channel resources are managed by the runtime, so
+we need the runtime to pass a ``higher-order factory''
+(a factory method that generates a factory)
+which lets the send state component generate the required
+component factories (i.e. \tmsg{ADD} and \tmsg{QUIT})
+with channel actions pre-injected into the component.
+We discuss the runtime in \cref{section:reactruntime},
+but for now, it is sufficient to understand that
+the generated send component abstract class receives
+the higher-order factory as a prop, and that this
+prop binds the channel action into the factory that it returns.
+
+Now we are in a position to accurately type the abstract
+send component \texttt{S40} in \cref{lst:reactsendcomponent}.
+We hint at the usage of the higher-order factory on
+\cref{line:factory1,line:factory2}:
+just using \cref{line:factory1} as an example,
+the send component asks the runtime to generate a 
+component factory for sending a message 
+labelled \lstonelinejs{'ADD'}
+with payload typed
+\lstonelinejs{[number, number]},
+then transition to the \texttt{ReceiveState.S42}
+successor state.
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript,tabsize=2]
+type Props = { factory: SendComponentFactoryFactory };
+abstract class S40<State> extends React.Component<Props, State> {
+	protected ADD: SendComponentFactory<[number, number]>;
+	protected QUIT: SendComponentFactory<[string]>;
+
+	constructor(props: Props) {
+		super(props);
+		this.ADD = props.factory<[number, number]>( (*@\label{line:factory1}@*)
+			'ADD', ReceiveState.S42);
+		this.QUIT = props.factory<[string]>( (*@\label{line:factory2}@*)
+			'QUIT', ReceiveState.S43);
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{Generated Type for \trole{Client} Send State
+in \tprotocol{Adder} protocol}
+\label{lst:reactsendcomponent}
+\end{figure}
+
+\subsection{Receive States}
 \begin{itemize}
@@
 
-\subsection{Terminal}
+\subsection{Terminal States}
 \begin{itemize}

### 2020-06-06 report/sections/server/react/efsm.tex
Report: progress on EFSM

@@
 		const (*@\hl{Add}@*) = this.ADD('onClick', (*@\label{line:clientsendfactory1}@*)
-			(ev: React.MouseEvent) => [num1, num2] (*@\label{line:clientsendpayload1}@*)
-		);
+			(ev: React.MouseEvent) => [num1, num2]); (*@\label{line:clientsendpayload1}@*)
 		const (*@\hl{Quit}@*) = this.QUIT('onMouseOver', (*@\label{line:clientsendfactory2}@*)
-			(ev: React.MouseEvent) => [message] (*@\label{line:clientsendpayload2}@*)
-		);		
+			(ev: React.MouseEvent) => [message]); (*@\label{line:clientsendpayload2}@*)
 		return (<div>
@@
 \end{lstlisting}
-\captionof{lstlisting}{\dots}
+\captionof{lstlisting}{Developer Implementation for \trole{Client}
+Send State in \tprotocol{Adder} protocol}
 \label{lst:adderclientsenddev}
@@
 \end{lstlisting}
-\captionof{lstlisting}{\dots}
+\captionof{lstlisting}{Extracting Function Properties from 
+TypeScript Interface}
 \label{lst:extractfuncprops}
@@
 \subsection{Receive States}
+
+We also encode receive states as abstract React components,
+and leverage props to ensure channel resources are not exposed
+to the developer.
+
+According to the EFSM, a receive state means
+the endpoint is ``waiting for the message'',
+and the actual receive event represents a transition
+to a successor state.
+This means that the developer's implementation will define
+what to render on the DOM when the endpoint is waiting
+for the message, rather than updating the DOM with the
+received message.
+Referring to \cref{fig:adderclientfsm} again,
+S42 will be rendered when the endpoint is waiting for
+\tmsg{RES}, but once received, S40 will be immediately
+rendered. The developer will need to manage the flow
+of stateful application data between UI components separately
+-- \fancyname{ReactMPST} only generates code for managing
+the communication automaton.
+
+We generate an abstract method for each permitted receive
+transition, which is used as the receive handler.
+This allows the developer to intercept the
+message and perform operations -- such as updating
+application state -- before the successor state is rendered.
+Specifying receive handlers as abstract methods mandate
+the developer to provide a concrete implementation.
+
+\begin{lstlisting}[language=javascript,tabsize=2]
+enum Labels {
+	THANKS = 'THANKS', TERMINATE = 'TERMINATE',
+};
+
+type THANKSMessage = {
+    label: Labels.THANKS,
+    payload: [string],
+};
+type TERMINATEMessage = {
+    label: Labels.TERMINATE,
+    payload: [],
+};
+
+type Message = | THANKSMessage | TERMINATEMessage;
+\end{lstlisting}
+
+\begin{lstlisting}[language=javascript,tabsize=2]
+type Props = {
+	register: (role: Roles.Peers, handle: ReceiveHandler) => void
+};
+abstract class S42<S> extends React.Component<Props, S> {
+	abstract THANKS(payload1: string): void;
+	abstract TERMINATE(): void;
+
+	componentDidMount() {
+		this.props.register(this.handle.bind(this));	
+	}
+
+	handle(message: any): State {
+		const parsedMessage = JSON.parse(message) as Message;
+		switch (parsedMessage.label) {
+			case Labels.THANKS:
+				return this.THANKS(...parsedMessage.payload);			
+			case Labels.TERMINATE:
+				return this.TERMINATE(...parsedMessage.payload);
+		}
+	}
+}
+\end{lstlisting}
+
 \begin{itemize}
@@
 \end{itemize}
+
+\begin{lstlisting}[language=javascript,tabsize=2]
+type Props = { terminate: () => void };
+abstract class S41<S> extends React.Component<Props, S> {
+	componentDidMount() { this.props.terminate(); }	
+}
+\end{lstlisting}

### 2020-06-07 report/sections/server/react/efsm.tex
Report: complete chapter 5 EFSM

@@
 \begin{figure}[!h]
-\begin{lstlisting}[language=javascript,tabsize=2]
+\begin{lstlisting}[language=javascript,tabsize=2,title=S40.tsx]
 type Props = { factory: SendComponentFactoryFactory };
@@
 
-We generate an abstract method for each permitted receive
+We generate an \textit{abstract method} 
+for each permitted receive
 transition, which is used as the receive handler.
@@
 
+We demonstrate below how the developer could implement
+receive state S42 -- here, the developer manages
+stateful application data using the
+\textit{sessionStorage} API \cite{SessionStorage}
+available on the browser.
+
+\begin{lstlisting}[language=javascript]
+class WaitScreen extends S42 {
+	render() { return <h1>Waiting</h1>; }
+	RES(n: number) {
+		sessionStorage.setItem('result', n);
+	}
+}
+\end{lstlisting}
+
+Receive states in \fancyname{ReactMPST}
+work in a similar way as with \fancyname{NodeMPST}.
+The runtime will invoke the \tmsg{RES} method
+implemented by the developer upon receiving the message
+through the WebSocket.
+In order to do so, the S42 component 
+needs to \textit{register}
+the handler with the runtime.
+
+Recall that React supports an uni-directional data flow,
+so there is no direct way to pass the handler
+back up to the parent component.
+We overcome this limitation by allowing the
+runtime to pass a \texttt{register} function to the
+receive state component as a prop; once the receive state
+component is mounted on the DOM, it proactively 
+registers the handlers with the runtime.
+
+This raises questions about the type signature for the
+\texttt{register} function, as the receive handler
+identifiers and type signatures differ between receive states.
+We adopt the same strategy as with \fancyname{NodeMPST}:
+the receive handler registered by the receive state
+to the runtime also performs the message parsing,
+so all receive states register a ``general handle function'' typed
+
+\begin{lstlisting}[language=javascript,numbers=none]
+type MessageHandler = (message: any) => void;
+\end{lstlisting}
+
+and define their own \texttt{MessageHandler} to handle
+state-specific message parsing and processing.
+We construct similar types for the state-specific
+labels and message structures.
+This is illustrated in \cref{lst:reactreceivetypes}) --
+we use state S43 from \cref{fig:adderclientfsm} hereafter
+to show how our approach scales with multiple branches.
+
+\begin{figure}[!h]
 \begin{lstlisting}[language=javascript,tabsize=2]
-enum Labels {
-	THANKS = 'THANKS', TERMINATE = 'TERMINATE',
-};
-
-type THANKSMessage = {
+enum Labels { THANKS = 'THANKS', TERMINATE = 'TERMINATE' };
+interface THANKSMessage {
     label: Labels.THANKS,
@@
 };
-type TERMINATEMessage = {
+interface TERMINATEMessage {
     label: Labels.TERMINATE,
@@
 };
-
 type Message = | THANKSMessage | TERMINATEMessage;
 \end{lstlisting}
+\captionof{lstlisting}{Generated Label and Message Types
+in \fancyname{ReactMPST}}
+\label{lst:reactreceivetypes}
+\end{figure}
 
+We generate the \texttt{handle()} method for each
+abstract receive state component for state-specific message
+parsing and processing. We show the full generated code
+for state S42 in \cref{lst:reactreceivestate}.
+Unlike \fancyname{NodeMPST},
+the \texttt{handle()} method (\cref{line:reactreceivehandle})
+returns the successor state identifier to the runtime
+to make the transition.
+We leverage the discriminated union from our
+generated message types (\cref{lst:reactreceivetypes})
+to invoke the correct receive handler (\cref{line:reactreceivediscriminate}).
+
+Again, due to language limitations, we are unable to 
+simplify the message processing step as
+
+\begin{lstlisting}[language=javascript,numbers=none]
+this[parsedMessage.label](...parsedMessage.payload);
+\end{lstlisting}
+
+because this is implicitly expressing a type dependency
+over union types,
+and \texttt{parsedMessage.label} is unknown at compile-time.
+
+\begin{figure}[!h]
 \begin{lstlisting}[language=javascript,tabsize=2]
 type Props = {
-	register: (role: Roles.Peers, handle: ReceiveHandler) => void
+	register: (handle: ReceiveHandler) => void
 };
@@
 
-	handle(message: any): State {
+	handle(message: any): State { (*@\label{line:reactreceivehandle}@*)
 		const parsedMessage = JSON.parse(message) as Message;
-		switch (parsedMessage.label) {
+		switch (parsedMessage.label) { (*@\label{line:reactreceivediscriminate}@*)
 			case Labels.THANKS:
-				return this.THANKS(...parsedMessage.payload);			
+				this.THANKS(...parsedMessage.payload);
+				return TerminalState.S41;
 			case Labels.TERMINATE:
-				return this.TERMINATE(...parsedMessage.payload);
+				this.TERMINATE(...parsedMessage.payload);
+				return TerminalState.S41;
 		}
@@
 \end{lstlisting}
-
-\begin{itemize}
-\item receive state is `shown' when waiting for a receive event
-\item base class has abstract methods -- one for each possible receive -- so the user must implement these when they extend the abstract base class
-\item each abstract method is named after the message label expected to be received, and parameterised by the payload
-\item no channel actions are passed to the state -- instead, once the receive state is mounted on the DOM (i.e.
-rendered), the abstract base class will register the event handlers to the runtime (using the register method passed as props)
-\item when the message arrives, the runtime will invoke the registered function (+ forward reference the next section with the edge case that the message arrives before handler registration occurs)
-\end{itemize}
+\captionof{lstlisting}{Generated Code for \trole{Client} 
+Receive State S42 in \tprotocol{Adder} protocol}
+\label{lst:reactreceivestate}
+\end{figure}
 
 \subsection{Terminal States}
-\begin{itemize}
-\item by convention and design choice, the client gracefully closes the connection with the server role
-\end{itemize}
+To be consistent with the conventions defined for \fancyname{NodeMPST},
+the browser endpoint proactively closes the connection at its terminal state.
+We generate an abstract React component for terminal states 
+to allow the developer to customise what to
+\texttt{render} upon protocol termination.
 
-\begin{lstlisting}[language=javascript,tabsize=2]
+\begin{lstlisting}[language=javascript]
 type Props = { terminate: () => void };
-abstract class S41<S> extends React.Component<Props, S> {
+abstract class S41<State> extends React.Component<Props, State> {
 	componentDidMount() { this.props.terminate(); }	
@@
 \end{lstlisting}
+
+When the component is mounted on the DOM, it notifies the runtime
+(via the received \texttt{terminate} prop) to terminate the session.

### 2020-06-07 report/sections/server/react/runtime.tex
Report: complete chapter 5 approach
Report: progress on chapter 5 runtime

@@
 \section{Runtime}
+\label{section:reactruntime}
+
+There are many parallels between the runtime we generate
+for \fancyname{ReactMPST} compared to that of \fancyname{NodeMPST}.
+We focus on the details unique to browser-side endpoints.
+
+We define the public API for the session runtime
+as a React component. The developer provides implementation details via
+props.
+We show the public API for the \trole{Client} endpoint
+of the \tprotocol{Adder} protocol below.
+
+\begin{lstlisting}[language=javascript]
+type Props = {
+	endpoint: string,
+	states: {
+		S40: Constructor<S40>,
+		S42: Constructor<S42>,
+		S43: Constructor<S43>,
+		S41: Constructor<S41>,
+	},
+	waiting: React.ReactNode,
+	connectFailed: React.ReactNode,
+};
+\end{lstlisting}
+
+The developer provides the WebSocket URL for the runtime
+to instantiate the connection.
+As for the EFSM state implementations derived from the abstract
+classes, the developer constructs an object to map state identifiers
+to concrete implementations. 
+The developer only needs to pass the \textit{constructor} function,
+as the runtime will instantiate the React components separately
+and pass the required props based on the type of state.
+Additionally, the developer needs to define what to render
+whilst \texttt{waiting} for the session to begin, and possibly
+an error screen if the connection has failed.
+
+When the session starts, the private API takes over and executes
+the EFSM by rendering the React component corresponding to the
+current EFSM state, as well as performing channel actions
+that adhere to the protocol.
+
+\subsection{Connecting to the Session}
+We connect to the session by creating a new \texttt{WebSocket}.
+Traditionally, this would be done in the constructor,
+but for React components, the constructor may be invoked more
+than once depending on how reconciliation works.
+For this reason, we create the \texttt{WebSocket} in the
+\texttt{componentDidMount} method as it is guaranteed
+to be only called once.
+In fact, under React's \textit{Strict Mode},
+constructors are \textit{explicitly invoked twice} to 
+prevent impure constructors \cite{ReactStrictMode}.
+
+This means the WebSocket must have optional type as it is 
+strictly undefined on construction.
+To avoid having to deal with an optional WebSocket value
+in the EFSM execution (even though we know for sure that it
+has been instantiated when we use it),
+we define the public API (the React component named 
+\texttt{Session}) to manage the (possibly \lstonelinejs{undefined})
+WebSocket in its state, and the private API (the React component
+named after the role -- in this case, \trole{Client})after
+as a separate non-exported class
+with the WebSocket available via props. Once the WebSocket
+is instantiated in the \trole{Session} component, it renders the
+\trole{Client} component, passing the non-optional WebSocket value via props.
+
+\begin{lstlisting}[language=javascript,tabsize=2]
+type Transport = { ws: WebSocket };
+class Session extends React.Component<Props, Partial<Transport>> {
+	componentDidMount() {
+		this.setState({ ws: new WebSocket(this.props.endpoint) });
+	}
+
+	render() {
+		const { ws } = 	this.state;
+		return ws === undefined ? this.props.waiting 
+			: <Client ws={ws} {...this.props} />;
+	}
+}	
+export default Session;
+\end{lstlisting}
+
+The connection phase is managed in the same way as \fancyname{NodeMPST}:
+when the \texttt{Client} component is mounted,
+it sends a connection request to the server and overrides the
+\texttt{onmessage} event handler to listen for the connection confirmation,
+before advancing the EFSM to the initial state to begin executing the protocol.
+
+\subsection{Executing the EFSM}
+Unlike \fancyname{NodeMPST}, our EFSM state encoding does not define
+continuation states using the actual implementation.
+This is because state encodings are React components
+that need to be instantiated by the runtime.
+Instead, as previewed in \cref{lst:reactsendcomponent,lst:reactreceivecomponent},
+we define an enum-based abstraction for specifying
+successor state, and this is also used for executing the EFSM.
+Our enum-based abstraction defines a string enum for each type of EFSM state,
+and collects all state identifiers in an union type.
+
+\begin{lstlisting}[language=javascript,numbers=none]
+enum SendState { S40 = 'S40' };
+enum ReceiveState { S42 = 'S42', S43 = 'S43' };
+enum TerminalState { S41 = 'S41' };
+type State = ReceiveState | SendState | TerminalState;
+\end{lstlisting}
+
+The React runtime component also defines a transition function
+parameterised by the current state's enum. The runtime renders
+the current EFSM state component with different props depending
+on the type of the current state.
+
 \begin{itemize}
-\item public API - accept websocket endpoint URL, a mapping of state to user implementation, and UI elements to render when waiting for connection to occur, and connection failure
-\item private API - carry out the session (just like Node)
+\item
+For \textbf{send} states, the runtime provides the higher-order factory
+method to allow the send state to construct its own send component
+factory properties for each permitted send transition.
+
+\item
+For \textbf{receive} states, the runtime provides the \texttt{register}
+function to let the developer pass back the receive handlers
+for each permitted receive transition, which is used
+to process incoming messages received via the WebSocket.
+
+\item
+For the \textbf{terminal} state, the runtime provides the
+\texttt{terminate} function to close the WebSocket connection.
 \end{itemize}
 
-\subsection{Session Joining}
-\begin{itemize}
-\item separate into two react components, since the websocket connection cannot be opened in the constructor by React StrictMode
-\item `outer' class sets websocket to undefined, but upon componentDidMount, creates the actual connection
-\item renders the `inner' class component upon websocket not being undefined, so it can be typed appropriately (otherwise, need to always use bang operator)
-\item on connection, send join request message to the server, wait for serber response before executing the EFSM at the initial state
-\end{itemize}
+Enum unions cannot be used in the same manner as discriminated unions
+to distinguish between \texttt{SendState} and \texttt{ReceiveState},
+so we provide utility functions under EFSM.ts that define
+\textit{type guards} to narrow the type of \texttt{State}.
 
-\subsection{Executing the EFSM}
-\begin{itemize}
-\item same concept as node -- implementation details delegated to the actual classes
-\item constructor binds methods
-\item advance takes the enum instead of the class (since unlike the node example, it is not instantiated) -- uses custom type guards to distinguish between the type of state
-\end{itemize}
+\begin{lstlisting}[language=javascript,numbers=none]
+function isReceiveState(state: State): state is ReceiveState {
+	return Object.values(ReceiveState).includes(state)
+}
+\end{lstlisting}
 
-\subsection{Handling Message Sends}
+We use these type guards in the transition function,
+as shown in \cref{lst:reacttransitionfunc}.
+The choice of a string enum for state identifiers
+allow us to use it to index into the EFSM state component mapping
+provided by the developer, as seen on 
+\cref{line:reactadvance1,line:reactadvance2,line:reactadvance3}.
+Note that the value of \texttt{state} can be an union type
+(depending on the possible values in the enum),
+so the type of \texttt{View} may also be union types.
+We can still deterministically pass props and render the component,
+as all states of each particular type agree on what props to receive.
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript]
+private advance(state: State) {
+	if (isSendState(state)) {
+		const View = this.props.states[state]; (*@\label{line:reactadvance1}@*)
+		this.setState({	
+			elem: <View factory={this.buildFactory} />
+		});
+	} else if (isReceiveState(state)) {
+		const View = this.props.states[state]; (*@\label{line:reactadvance2}@*)
+		this.setState({
+			elem: <View register={this.registerReceiveHandler} />
+		});
+	} else if (isTerminalState(state)) {
+		const View = this.props.states[state]; (*@\label{line:reactadvance3}@*)
+		this.setState({
+			elem: <View terminate={this.terminate} />
+		});
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{EFSM Transition Function for Browser-Side Endpoint}
+\label{lst:reacttransitionfunc}
+\end{figure}
+
+We elaborate on \dots
+
+\subsection{Sending Messages}
 \begin{itemize}
@@
 
-\subsection{Handling Message Receives}
+\begin{lstlisting}[language=javascript,tabsize=2]
+private buildFactory<T>(label: string, successor: State) {
+	return <K extends keyof DOMEvents>
+				 (ev: K, handler: EventHandler<E, K> => {
+		const send = (payload: T) => 
+			this.sendMessage(label, payload, successor);
+		return class extends React.Component {
+			render() {
+				const props = {
+					[ev as string]: (e: FunctionArguments<DOMEvents[K]>) => {
+						const payload = handler(e);
+						send(payload);
+					}
+				};
+				return <div {...props}>{this.props.children}</div>
+			}		
+		}	
+	}
+}
+\end{lstlisting}
+
+\subsection{Receiving Messages}
 \begin{itemize}

### 2020-06-07 report/sections/server/react/efsm.tex
Report: progress on chapter 5 runtime
Report: complete chapter 5

@@
 \begin{lstlisting}[language=javascript,numbers=none]
-type MessageHandler = (message: any) => void;
+type MessageHandler = (message: any) => State;
 \end{lstlisting}
@@
 parsing and processing. We show the full generated code
-for state S42 in \cref{lst:reactreceivestate}.
+for state S42 in \cref{lst:reactreceivecomponent}.
 Unlike \fancyname{NodeMPST},
@@
 Receive State S42 in \tprotocol{Adder} protocol}
-\label{lst:reactreceivestate}
+\label{lst:reactreceivecomponent}
 \end{figure}

### 2020-06-07 report/sections/server/react/limitations.tex
Report: complete chapter 5

@@
 \section{Limitations}
-As discussed, our EFSM encoding provides \textbf{affine} channel
+Our EFSM encoding provides \textbf{affine} channel
 usage guarantees for send states as supposed to \textit{linear} usage
@@
 we present examples of developer implementations
-that use separate abstractions (such as the React Context API or
-the third-party Redux library) for propagating 
+that use separate abstractions (e.g. the \textit{React Context API} or
+the third-party \textit{Redux} state management framework) for propagating 
 stateful application data in \cref{chap:eval}.

### 2020-06-07 report/sections/server/react/runtime.tex
Report: complete chapter 5

@@
 
-We elaborate on \dots
+We keep track of the current EFSM state in the state of the
+private \texttt{Client} component.
+This enables the UI to be re-rendered, and therefore ``react'',
+to EFSM state transitions, which is precisely the goal of \fancyname{ReactMPST}.
 
 \subsection{Sending Messages}
-\begin{itemize}
-\item passes in the factory prop, which is a factoryfactory to be precise;
-\item sendMessage is bound in the closure and sends the message in the same way as how Node codegen does, by convention
-\item (if not already mentioned in the node section) send is asynchronous
-\end{itemize}
+Messages are sent through the WebSocket in the same way 
+as \fancyname{NodeMPST} -- objects marked with label and payload
+are serialised into JSON. The process of sending is always followed
+by a transition to the successor state, so we define the
+\texttt{sendMessage()} method on the runtime to also take
+the successor state as a function argument, so it can make that transition
+immediately after.
 
+\begin{lstlisting}[language=javascript,numbers=none]
+private sendMessage(label: string, payload: any, next: State) {
+	this.props.ws.send(JSON.stringify({ label, payload }));
+	this.advance(next);
+}
+\end{lstlisting}
+
+We focus on the higher-order factory that the runtime passes
+to the send component as a prop. The implementation shown in
+\cref{lst:reactsendfactory} is less straightforward,
+so we will walk through the steps.
+
+The prop allows the send component to build factory functions
+for sending messages with a particular \texttt{label}
+followed by a transition to some \texttt{successor} state
+(\cref{line:sendfactoryparams}).
+The higher-order factory relies on \textit{closures}
+to bind the channel action into the generated component.
+Here, \texttt{buildFactory} returns a (factory) function that
+builds an anonymous React component which wraps the 
+component's children into a \texttt{<div>} with special props 
+(\cref{line:sendfactorydiv}).
+The props for this anonymous React component comprises of a new
+event listener bound to the UI event (\texttt{ev})
+specified by the developer (such as \lstonelinejs{'onClick'}).
+This new event listener is defined in terms of the
+\texttt{handler} from the developer's implementation:
+\textbf{(1)} it invokes the handler to obtain the payload to send 
+(\cref{line:sendfactoryhandler}), which is constrained by
+the generic type parameter to ensure it matches the protocol specification;
+then, \textbf{(2)} it sends the message, which triggers the EFSM
+to advance to the successor state.
+We cannot directly call \lstonelinejs{this.sendMessage} inside
+the generated React component because \lstonelinejs{this} refers
+to the anonymous class; instead, we define a partially applied
+version of the \texttt{sendMessage} method as a local variable 
+(\cref{line:sendfactorycurrysend}).
+
+\begin{figure}[!h]
 \begin{lstlisting}[language=javascript,tabsize=2]
-private buildFactory<T>(label: string, successor: State) {
+private buildFactory<T>(label: string, successor: State) { (*@\label{line:sendfactoryparams}@*)
 	return <K extends keyof DOMEvents>
 				 (ev: K, handler: EventHandler<E, K> => {
-		const send = (payload: T) => 
+		const send = (payload: T) => (*@\label{line:sendfactorycurrysend}@*)
 			this.sendMessage(label, payload, successor);
@@
 					[ev as string]: (e: FunctionArguments<DOMEvents[K]>) => {
-						const payload = handler(e);
+						const payload = handler(e); (*@\label{line:sendfactoryhandler}@*)
 						send(payload);
@@
 				};
-				return <div {...props}>{this.props.children}</div>
+				return <div {...props}>{this.props.children}</div> (*@\label{line:sendfactorydiv}@*)
 			}		
@@
 \end{lstlisting}
+\captionof{lstlisting}{Higher-Order Factory Function in 
+Browser-Side Runtime}
+\label{lst:reactsendfactory}
+\end{figure}
 
 \subsection{Receiving Messages}
-\begin{itemize}
-\item same edge case as the back-end code generation, as message can arrive before handlers are registered
-\item use the same two-queue mechanism to keep track of `what comes first' -- i.e.
-if message comes first, then put in message queue, so handler can pop off when registered; vice-versa, if handler is registered first, put in handler queue so message can be processed when that is received
-\item importantly -- only one queue is required, as server-centric topologies only have the client talk to the server
-\end{itemize}
+
+We take the same approach as \fancyname{NodeMPST} --
+the runtime keeps track of the ``active'' receive handler,
+and the \texttt{onmessage} event listener for the WebSocket
+is defined to dynamically process the incoming message
+using the active receive handler.
+We also need to address the same consistency problem, since
+a message arrival event can be handled \textit{before}
+the receive handler is registered.
+The receive handler is registered when the receive state component
+is rendered on the DOM, which occurs when \lstonelinejs{this.setState}
+is invoked in the transition function.
+\lstonelinejs{this.setState} may be \textit{asynchronous}
+as React will internally batch multiple calls into a single update
+for performance \cite{ReactState}, so it is possible for the WebSocket message
+event to be queued (and hence, procesesd) before the receive handler
+is registered.
+
+We address this problem using a pair of queues to keep track
+of messages waiting for handlers, and handlers waiting for messages.
+Unlike \fancyname{NodeMPST}, we do \textit{not} need a mapping
+of queues because browser-side endpoints only receive from the server.
+
+\begin{lstlisting}[language=javascript,numbers=none]
+private messageQueue: any[];
+private handlerQueue: MessageHandler[];
+\end{lstlisting}
+
+This lets us define the WebSocket \texttt{onmessage} event
+listener and the \texttt{register} function (which we pass as a prop
+to the receive state component) in a similar way compared with
+\fancyname{NodeMPST} -- we show this in \cref{lst:reactreceiveruntime}.
+Recall that the message handler registered
+by the receive state component returns the successor state,
+so we invoke the transition function on the result directly 
+(\cref{line:reactreceiveadvance1,line:reactreceiveadvance2}).
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript]
+private register(handler: MessageHandler) {
+	const message = this.messageQueue.shift();
+	if (message !== undefined) { this.advance(handler(message)); } (*@\label{line:reactreceiveadvance1}@*)
+	else { this.handlerQueue.push(handler); }
+}
+
+private receive({ data }: MessageEvent) {
+	const handler = this.handlerQueue.shift();
+	if (handler !== undefined) { this.advance(handler(data); } (*@\label{line:reactreceiveadvance2}@*)
+	else { this.messageQueue.push(data); }
+}
+\end{lstlisting}
+\captionof{lstlisting}{Receive Handler Registration and Message Event
+Listener in \fancyname{ReactMPST}}
+\label{lst:reactreceiveruntime}
+\end{figure}
 
 \subsection{Handling Termination}
-\begin{itemize}
-\item close the websocket with a clean exit code
-\end{itemize}
+
+The terminal state component notifies the runtime (through
+the \texttt{terminate} function passed down as a prop)
+to close the connection. The implementation is straightforward --
+we simply invoke the \texttt{close()} method on the WebSocket.
+
+\begin{lstlisting}[language=javascript,numbers=none]
+private terminate() {
+	this.props.ws.close();
+}
+\end{lstlisting}

### 2020-06-07 report/sections/abstract.tex
Report: progress on abstract

\begin{abstract}

We present an end-to-end framework
to write interactive web applications,
communicating over WebSocket transport,
that statically conform to prescribed communication protocols.

We implement a code generation toolchain to
provide developers with TypeScript APIs generated
from a communication protocol specification 
based on multiparty session type theory.

Our toolchain generates APIs that are compatible with
the Node.js runtime and the React.js framework
for server-side and browser-side endpoints respectively.

We evaluate the expressiveness of our API generation solution
using three case studies of protocols found in web-services,
and analyse the overhead of our implementation through 
running micro-benchmarks on a simple binary session protocol.

\end{abstract}

### 2020-06-07 report/sections/abstract.tex
Report: complete abstract

@@
 
+\noindent
 We present an end-to-end framework
-to write interactive web applications,
-communicating over WebSocket transport,
-that statically conform to prescribed communication protocols.
-
+to write interactive web applications
+communicating over WebSocket transport
+that conform to prescribed communication protocols.
 We implement a code generation toolchain to
@@
 
-Our toolchain generates APIs that are compatible with
-the Node.js runtime and the React.js framework
+Our work is compatible with modern web programming
+industrial practices. The APIs
+generated by the toolchain 
+build upon TypeScript concurrency primitives,
+complement the event-driven style of programming
+in full-stack web development,
+and are compatible
+with the Node.js runtime and the React.js framework
 for server-side and browser-side endpoints respectively.
+
+We propose an approach for supporting peer-to-peer communication
+between browser-side endpoints through routing communication via the server
+in a way that avoids excessive serialisation
+and preserves communication safety. 
+We formalise this approach by extending multiparty session type theory
+and proving that properties guaranteeing communication safety
+are preserved.
 

### 2020-06-07 report/sections/evalconcl/concl/future.tex
Report: progress on conclusion

@@
 
-\paragraph{Note: from PLACES}
-Future work includes incorporating \textit{explicit connection actions}
-introduced in \cite{ExplicitConnections} in our API generation to better model
-real-world communication protocols that may feature in interactive web
-applications.
-% REVIEW: asynchronous
-Server-side implementations may perform asynchronous operations on the
-received messages, so supporting asynchronous values (such as JavaScript
-\textit{Promises} \cite{promise}) in IO actions would be a welcome addition.
-Whilst our approach supports multiparty sessions, the nature of
-WebSockets require some server-based role in the communication protocol and
-clients to interact via the server.
-Extending support to WebRTC \cite{WebRTC} would
-cater for peer-to-peer communication between browsers, which further opens up
-possibilities for communication protocols supported by our approach.
+We highlight several areas of future work
+to support a wider range of communication protocols
+for web applications and increase the 
+industrial relevance of our work.
+
+\subsection{Delegation}
+Delegation allows channels to be sent between endpoints.
+This directly addresses the limitations presented in
+\cref{section:evalatm}: in the \tprotocol{ATM} protocol,
+when the \trole{Client}
+connects to the \trole{Svr},
+the \trole{Svr} should be able to send a private channel
+for \trole{Client} to perform bank operations,
+whilst concurrently accepting new connections.
+
+The API generation solution 
+presented by Scalas et al. \cite{LinearDecomp} supports
+\textit{distributed multiparty delegation} in Scala.
+We have also shown in \cref{section:evalatm} that
+the developer can implement delegation with small changes
+to the session runtime.
+Future work can apply the findings from \cite{LinearDecomp},
+extend the Scribble protocol language for sending
+channels through messages, and formalise the workaround
+in \cref{section:evalatm} with respect to the extended 
+Scribble syntax for delegation.
+
+\subsection{Explicit Connection Actions}
+Hu and Yoshida \cite{ExplicitConnections} extends 
+multiparty session type theory
+with \textit{explicit connection actions} to support protocols 
+with optional participants.
+Supporting optional participants is a relevant concept
+for web applications, particularly for browser-side endpoints.
+
+Our work defines a way to handle connection requests
+and detect disconnection events. The implementation assumes that
+all participants join the session when the protocol begins.
+We also focus on disconnection events as a result of session
+cancellation rather than those stated in the protocol.
+
+\cite{ExplicitConnections} uses a variant of the Scribble protocol
+language that supports explicit connection actions. Future work
+can use that as a basis for code generation, and generalise
+our existing mechanisms for session connection and cancellation
+to support explicit connection actions.
+
+\subsection{Transport Abstraction}
+The MPST-based web development framework presented by 
+King et al. \cite{PureScript2019} defines the session runtime
+in a way that parameterises the transportation mechanism.
+The developer can run the session over different transportation,
+provided that the custom transport implements the behaviour
+of the \texttt{Transport} type class.
+
+Future work can generalise the session runtime for 
+\fancyname{NodeMPST} and \fancyname{ReactMPST} to provide
+similar abstractions. A successful implementation could
+possibly allow direct interactions between two Node.js-based
+endpoints, but this will also require the routed MPST
+theory to be adapted accordingly to reason about
+such interactions in a web-based context.
+
+\subsection{Generalised EFSM Encoding as Typestates}
+The concept of \textit{typestates} define the interface of
+an object to be dependent on its private state, meaning
+it can change at runtime.
+This is compatible with the EFSM abstraction from the
+MPST framework: the permitted methods (i.e. channel actions)
+of the EFSM depends on its current state.
+In fact, Kouzapas et al. \cite{StMungo} presents a way to
+generate typestate specifications for each endpoint of a
+Scribble protocol; Gay et al. \cite{ModularST} proposes
+an encoding of session types in object-oriented languages
+and discuss their approach with respect to typestates.
+
+Future work can explore implementing typestates
+in TypeScript. A linear type system would be useful towards 
+implementing typestates, but this is lacking in TypeScript.
+One interesting possibility for approximating this would
+be to combine \textit{decorators} with \textit{transformers}:
+the former lets the developer define typestate-related metadata,
+whilst the latter can parse the metadata and
+transform the source code into another TypeScript program that acts
+as an ``intermediate representation'',
+such that only implementations that respect the typestate specification
+will type-check with its intermediate representation.

### 2020-06-07 report/sections/evalconcl/eval.tex
Report: progress on conclusion

@@
 
+\section{Delegated Sessions: \fancyname{ATM}}
+\label{section:evalatm}
+
+\begin{itemize}
+\item Show that delegation is not supported
+even if the protocol ``allows''
+\end{itemize}
+
+\begin{figure}[!ht]
+\begin{lstlisting}[language=Scribble, tabsize=2]
+module Delegation;
+
+global protocol ATM(role Svr, role Client) {
+	choice at Client {
+		WITHDRAW(string, number) from Client to Svr;
+		choice at Svr {
+			BALANCE(number) from Svr to Client;		
+		} or {
+			FAIL(string) from Svr to Client;		
+		}
+		do ATM(Svr, Client);
+	} or {
+		DEPOSIT(string, number) from Client to Svr;
+		BALANCE(number) from Svr to Client;
+		do ATM(Svr, Client);
+	} or {
+		BYE(string) from Client to Svr;
+		BYE() from Svr to Client;	
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{The \tprotocol{ATM} Protocol}
+\label{lst:atm}
+\end{figure}
+
 \input{sections/evalconcl/eval/benchmarks}

### 2020-06-10 report/sections/evalconcl/concl/contributions.tex
Report: progress on conclusion
Report: complete conclusion

@@
 
-\paragraph{Note: from PLACES}
+In this project, we develop and present an
+MPST-based framework for developing full-stack
+interactive TypeScript applications over WebSocket transport.
 
-We have presented an MPST-based framework for developing full-stack interactive
-TypeScript applications with WebSocket communications.
-The implementation conforms to a specified
-protocol, statically providing linear channel usage guarantees and affine
-channel usage guarantees for back-end and front-end targets respectively.
+We motivate our API generation approach from 
+\cite{Hybrid2016,PureScript2019} to generate TypeScript APIs
+from a Scribble protocol specification.
+Developers write their full-stack applications
+by implementing the generated APIs, and can enjoy 
+communication safety guarantees by construction.
+We explain how our session type encodings target the Node.js
+server-side runtime and React.js framework in
+\cref{chap:node,chap:react} for server-side
+and browser-side endpoints respectively.
+
+In particular, we present a novel approach for integrating
+session types into web-based GUI programming based on
+translating the theory on model types \cite{MVU2020}
+to idiomatic practices on the React.js framework.
+With respect to session type theory, implementations using
+the generated APIs statically enjoy linear channel usage
+guarantees and affine channel usage guarantees for back-end
+and front-end targets respectively.
+
+Compared to previous work, we are not only able to
+\textbf{statically} provide the same level of communication safety
+guarantees from multiparty session type theory, but we do so
+using modern web programming practices to increase
+the relevance and usability of our work in industry:
+our work targets TypeScript, Node.js and React.js,
+and we explain in \cref{chap:ext} how we support advanced
+web development idioms such as asynchronous implementations.
+
+In addition, existing proposals on API generation for web development
+over WebSockets \cite{PureScript2019} only support protocols
+implementing server-centric topologies, in order to be 
+compatible with WebSocket transport.
+We formalise a theory of routed multiparty session types in 
+\cref{chap:theory} to prove that it is possible to relax the
+server-centric topology assumption over WebSocket transport in a way 
+that preserves communication along with the communication safety 
+properties inherited from canonical session type theory.
+
+% PLACES
+Our work has received positive feedback from academia.
+The project has been accepted and published in the
+Proceedings of the 12th International Workshop on
+Programming Language Approaches to Concurrency- \& 
+Communication-cEntric Software (PLACES 2020) \cite{PLACES2020}.
+We were also invited to present this work as 
+\emph{Generating Interactive WebSocket Applications in TypeScript}
+at PLACES 2020.
+
+% OVERALL
+Overall, we offer an end-to-end solution for integrating
+multiparty session type theory into modern full-stack web programming
+practices, in a way that actively encourages developers to
+design their application with communication safety in mind.
+Our formalism of routed multiparty session types enables our
+API generation tool to support a wider range of protocols, and
+further reveals the potential
+for implementing dynamic communication structures over centralised
+network topologies.

### 2020-06-10 report/sections/evalconcl/concl/future.tex
Report: complete conclusion

@@
 
-\subsection{Delegation}
+\paragraph{Delegation}
 Delegation allows channels to be sent between endpoints.
@@
 
-\subsection{Explicit Connection Actions}
+\paragraph{Explicit Connection Actions}
 Hu and Yoshida \cite{ExplicitConnections} extends 
@@
 
-\subsection{Transport Abstraction}
+\paragraph{Transport Abstraction}
 The MPST-based web development framework presented by 
@@
 
-\subsection{Generalised EFSM Encoding as Typestates}
-The concept of \textit{typestates} define the interface of
-an object to be dependent on its private state, meaning
-it can change at runtime.
-This is compatible with the EFSM abstraction from the
-MPST framework: the permitted methods (i.e. channel actions)
-of the EFSM depends on its current state.
-In fact, Kouzapas et al. \cite{StMungo} presents a way to
-generate typestate specifications for each endpoint of a
-Scribble protocol; Gay et al. \cite{ModularST} proposes
-an encoding of session types in object-oriented languages
-and discuss their approach with respect to typestates.
-
-Future work can explore implementing typestates
-in TypeScript. A linear type system would be useful towards 
-implementing typestates, but this is lacking in TypeScript.
-One interesting possibility for approximating this would
-be to combine \textit{decorators} with \textit{transformers}:
-the former lets the developer define typestate-related metadata,
-whilst the latter can parse the metadata and
-transform the source code into another TypeScript program that acts
-as an ``intermediate representation'',
-such that only implementations that respect the typestate specification
-will type-check with its intermediate representation.
+%\paragraph{EFSM Encoding as Typestates}
+%The concept of \textit{typestates} define the interface of
+%an object to be dependent on its private state, meaning
+%it can change at runtime.
+%This is compatible with the EFSM abstraction from the
+%MPST framework: the permitted methods (i.e. channel actions)
+%of the EFSM depends on its current state.
+%In fact, Kouzapas et al. \cite{StMungo} presents a way to
+%generate typestate specifications for each endpoint of a
+%Scribble protocol; Gay et al. \cite{ModularST} proposes
+%an encoding of session types in object-oriented languages
+%and discuss their approach with respect to typestates.
+%
+%Future work can explore implementing typestates
+%in TypeScript. A linear type system would be useful towards 
+%implementing typestates, but this is lacking in TypeScript.
+%One interesting possibility for approximating this would
+%be to combine \textit{decorators} with \textit{transformers}:
+%the former lets the developer define typestate-related metadata,
+%whilst the latter can parse the metadata and
+%transform the source code into another TypeScript program that acts
+%as an ``intermediate representation'',
+%such that only implementations that respect the typestate specification
+%will type-check with its intermediate representation.

### 2020-06-10 report/sections/evalconcl/eval.tex
Report: complete conclusion

@@
 %benchmark -- compile time / benchmarking
-
-\section{Binary Sessions: \fancyname{Calculator}}
-\label{section:evalcalculator}
-\begin{itemize}
-\item sample client implementation + snippet of generated code (full in appendix)
-\item show that not all the UI is being controlled by the runtime (i.e. calculator app has a header/fake navbar)
-\item show the calculator logic achievable using React Context
-\item sample server implementation + snippet of generated code (full in appendix)
-\end{itemize}
-
-\begin{figure}[!ht]
-\begin{lstlisting}[language=Scribble, tabsize=2]
-module BinarySession;
-
-global protocol Calculator(role Svr, role Client) {
-  choice at Client {
-    Add(number, number) from Client to Svr;
-    Res(number)         from Svr to Client;
-    do Calculator(Svr, Client);
-  } or {
-    Mult(number, number) from Client to Svr;
-    Res(number)          from Svr to Client;
-    do Calculator(Svr, Client);
-  } or {
-    Quit()               from Client to Svr;
-    Terminate()          from Svr to Client;
-  }
-}
-\end{lstlisting}
-\captionof{lstlisting}{The \tprotocol{Calculator} Protocol}
-\label{lst:calculator}
-\end{figure}
 
@@
 \item sample client implementation -- show the flexibility of factory API for sending to be used in the UI part (since move selection can be be made )
+\item show that not all the UI is being controlled by the runtime (i.e. calculator app has a header/fake navbar)
+\item show the calculator logic achievable using React Context
+\item sample server implementation + snippet of generated code (full in appendix)
 \end{itemize}

### 2020-06-10 report/sections/introduction/objectives.tex
Report: progress on introduction

@@
 
-The objective of this project is to \dots
+The objective of this project is to provide an end-to-end
+framework for writing full-stack web applications that
+\textbf{statically} conform to a communication protocol.
+Developers specify their communication protocol using
+the Scribble protocol description language \cite{Scribble}
+which is based on multiparty session type (MPST) theory \cite{MPST}.
+The Scribble protocol is used as a basis for generating
+TypeScript APIs targeted for the Node.js server-side runtime
+and React.js browser-side framework. 
+By implementing the generated APIs in their full-stack applications,
+developers can enjoy \textbf{static} guarantees with respect 
+to communication safety: there will not be any communication mismatches,
+and channel linearity will be respected.
+Finally, we extend our framework to support communication
+protocols that cannot be natively implemented over
+a server-centric network topology using WebSockets.
+This involves defining a theory of routed multiparty session types
+to formalise the preservation of communication safety properties.

### 2020-06-10 report/sections/introduction/motivation.tex
Report: complete introduction

@@
 
-Modern interactive web applications aim to provide a highly responsive user experience by minimising the communication latency between clients and servers. 
-Whilst the HTTP request-response model is sufficient for retrieving static assets, applying the same stateless communication approach for interactive use cases (such as multiplayer games) introduces undesirable performance overhead from having to frequently set up new connections for client-server interactions.
-Developers have since adopted other communication transport abstractions over HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy low-latency full-duplex client-server communication in their applications over a single persistent connection.
-Enabling more complex communication patterns caters for more interactive use cases, but introduces additional concerns to the developer with respect to implementation correctness.
+Modern interactive web applications aim to 
+provide a highly responsive user experience by 
+minimising the communication latency between clients and servers. 
+Whilst the \textit{HTTP} request-response model is 
+sufficient for retrieving static assets, applying the same 
+stateless communication approach for interactive use cases 
+(such as multiplayer games) introduces undesirable 
+performance overhead from having to frequently set up 
+new connections for client-server interactions. 
+Developers have since adopted other communication 
+transport abstractions over HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy low-latency full-duplex 
+client-server communication in their applications over 
+a single persistent connection. 
+Enabling more complex communication patterns caters for 
+more interactive use cases, but introduces additional 
+concerns to the developer with respect to implementation correctness.
 
 % Example: noughts and crosses
-Consider a classic turn-based board game of \textit{Noughts and Crosses} between two players.
-Both players are identified by either noughts or crosses respectively, and take turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player wins (when their markers form one straight line on the board) or a stalemate is reached (when all cells are occupied).
-A web-based implementation may involve players connected to a game server via WebSocket connections and interacting with the game from their web browser, which serve a \textit{single-page application} (SPA) of the game client written in a popular framework like \textit{React.js} \cite{React}.
-SPAs feature a single HTML page and dynamically renders content via JavaScript in the browser.
-Players take turns to make a move on the game board and the server implements the game logic to progress the game forward until a result (either a win/loss or draw) can be declared.
+Consider a classic turn-based board game of \textit{Noughts and Crosses} 
+between two players. Both players are identified by either noughts or crosses 
+respectively, and take turns to place a mark on an unoccupied cell 
+of a 3-by-3 grid until one player wins (when their markers form 
+one straight line on the board) or a stalemate is reached 
+(when all cells are occupied). A web-based implementation may 
+involve players connected to a game server via WebSocket connections 
+and interacting with the game from their web browser, which serve 
+a \textit{single-page application} (SPA) of the game client written 
+in a popular framework like \textit{React.js} \cite{React}. 
+SPAs feature a single HTML page and dynamically renders content 
+via JavaScript in the browser. 
+Players take turns to make a move on the game board and the server 
+implements the game logic to progress the game forward until 
+a result (either a win/loss or draw) can be declared.
 
-Whilst WebSockets make this web-based implementation possible, it introduces the developer to a new family of communication errors, even for this simple game.
-In addition to the usual testing for game logic correctness, the developer needs to test against \textit{deadlocks} (e.g.
-both players waiting for each other to make a move at the same time) and \textit{communication mismatches} (e.g.
-player 1 sending a boolean to the game server instead of the board coordinates).
-The complexity of these errors, which correlate to the complexity of tests required against these errors, scale with the complexity of the communication patterns involved.
+Whilst WebSockets make this web-based implementation possible, 
+it introduces the developer to a new family of communication errors
+(in addition to the usual testing for game logic correctness), 
+even for this simple game:
+we highlight just a few:
 
-\textit{Multiparty Session Types} (MPST) \cite{MPST} provide a framework for formally specifying a structured communication pattern between concurrent processes and verifying implementations for correctness with respect to the communications aspect.
-By specifying the client-server interactions of our game as a protocol and verifying the implementations against the protocol for well-formedness, MPST theory guarantees well-formed implementations to be free from communication errors.
+\begin{itemize}
 
-We see the application of the MPST methodology to verifying interactive web applications written in TypeScript to be an interesting design space - to what extent can the MPST methodology be applied such that we leverage TypeScript language features to maximise static safety guarantees whilst preserving a flexible, natural and idiomatic workflow for TypeScript developers?
-Such a workflow would ultimately decrease the overhead for incorporating MPST into mainstream web development, which reduces development time by programmatically verifying implementations for communication correctness.
+\item
+\textbf{Deadlocks:} how can we prevent both players waiting for 
+each other to make a move at the same time?
+
+\item 
+\textbf{Communication mismatches:} what if player 1 sends
+a boolean to the game server instead of board coordinates?
+
+\item
+\textbf{Channel linearity:} if the game server takes time
+to update the game logic and respond to the players, what if 
+player 1 clicks the same cell twice and sends the coordinates twice?
+
+\end{itemize}
+
+The complexity of these errors, which correlate to the complexity of tests 
+required against these errors, scale with the complexity of the 
+communication patterns involved. Over-reliance on integration testing
+to attempt to expose communication-related bugs will also slow the
+development process, not to mention that the time taken for these
+integration tests would scale with the number of roles involved.
+A localised, static way for verifying communication correctness would
+be highly desirable.
+
+\textit{Multiparty Session Types} (MPST) \cite{MPST} provide 
+a framework for formally specifying a structured communication pattern 
+between concurrent processes and verifying implementations for 
+correctness with respect to the communications aspect. 
+By specifying the client-server interactions of our game as a protocol 
+and verifying the implementations against the protocol for well-formedness, 
+MPST theory guarantees well-formed implementations to be 
+free from communication errors.
+
+Existing work \cite{MVU2020,PureScript2019} that 
+adapt the MPST framework for web development
+acknowledge the limitations of JavaScript -- the language of the browser --
+in providing static type-level guarantees, and proceed to use
+languages equipped with stronger type systems that compile to JavaScript.
+This comes at the cost of a learning curve for the developer,
+and limits its utility in the mainstream web development space.
+
+Among the many languages that compile to JavaScript,
+TypeScript stands out as the most intuitive to use as it is defined
+to be a \textit{superset} of JavaScript. It provides developers
+with type-safety through its gradual, structural type system.
+Whilst some \cite{MVU2020} point out this limits its usability for
+encoding multiparty session types, we believe that the language
+offers sufficient features that we can use to provide developers with
+communication safety guarantees whilst preserving a flexible, natural
+and idiomatic workflow. By building our work upon TypeScript, we
+work towards incorporating MPST into mainstream web development,
+which reduces development time by programmatically verifying 
+implementations for communication correctness.

### 2020-06-10 report/sections/evalconcl/concl/contributions.tex
Report: progress on chapter 6.1

@@
 Our work has received positive feedback from academia.
-The project has been accepted and published in the
-Proceedings of the 12th International Workshop on
+The project has been accepted and published as 
+\emph{Generating Interactive WebSocket Applications in TypeScript}
+in Volume 314 of the Electronic Proceedings in Theoretical
+Computer Science \cite{PLACES2020}; we were also invited to present this work
+at the 12th International Workshop on
 Programming Language Approaches to Concurrency- \& 
-Communication-cEntric Software (PLACES 2020) \cite{PLACES2020}.
-We were also invited to present this work as 
-\emph{Generating Interactive WebSocket Applications in TypeScript}
-at PLACES 2020.
+Communication-cEntric Software (PLACES 2020).
 

### 2020-06-10 report/sections/introduction/contributions.tex
Report: progress on chapter 6.1

@@
 interactive web applications over WebSocket transport. 
-Related work \citep{PureScript2019, MVU2020, LINKS, Tiers} requires the 
+Related work \citep{PureScript2019, MVU2020, LINKS, Exceptional} requires the 
 developer to implement their application using tools that are 
@@
 This project was published as \emph{Generating Interactive WebSocket 
-Applications in TypeScript} in the Proceedings of the 
-12th International Workshop on Programming Language Approaches to 
-Concurrency- \& Communication-cEntric Software (PLACES 2020) 
-\cite{PLACES2020}.
+Applications in TypeScript} in Volume 314 of the Electronic Proceedings
+in Theoretical Computer Science (EPTCS) \cite{PLACES2020}.
 \end{remark}

### 2020-06-10 report/sections/server/codegen.tex
Report: progress on chapter 6.1

@@
 \textbf{(3)} all non-server participants only communicate with the server.
-Examples of such protocols include the \tprotocol{Calculator} webservice
-(\cref{section:evalcalculator}) and the \tprotocol{Noughts and Crosses}
-multiplayer game (\cref{section:evalgame}).
+Examples of such protocols include the \tprotocol{Noughts and Crosses}
+multiplayer game (\cref{section:evalgame}) and the \tprotocol{ATM} protocol.
 We relax this assumption in \cref{part:general}.

### 2020-06-10 report/sections/server/extensions/async.tex
Report: progress on chapter 6.1

@@
 \section{Supporting Asynchronous Implementations}
+\label{section:async}
 
@@
 
-\begin{itemize}
-\item developer implementation in javascript -- let alone typescript -- will need asynchronous operations
-\item backend -- database calls are asynchronous (e.g. for a multiplayer game, server may update database before sending the game results)
-\item frontend -- may invoke third-party API calls (that are not part of the protocol) which return promises
-\item how to support asynchronous developer implementations
-\end{itemize}
+Consider a basic example of two-factor authentication 
+(\tprotocol{2FA}) in \cref{lst:2FA} motivated from \cite{Exceptional}:
+\trole{Client} tries to log in to \trole{Svr}, which either
+authorises or denies the login attempt. If \trole{Client} is logging 
+in from a new device, the \trole{Svr} sends a challenge key
+and waits for a response code before proceeding.
 
-\subsubsection{Concurrency Primitives in TypeScript}
-\begin{itemize}
-\item use an example to go over callbacks, promise and async/await functions
-\item compare promises to java futures
-\end{itemize}
+\begin{figure}[!h]
+\begin{lstlisting}[language=scribble]
+global protocol 2FA(role Svr, role Client) {
+	Login(string, string) from Client to Svr;
+	choice at Svr { (*@\label{line:2fachoice1}@*)
+		Authorised() from Svr to Client;
+		do Main(Svr, Client);		// Details are not important
+	} or {
+		Challenge(string)	from Svr to Client;
+		Response(number) 	from Client to Svr;
+		choice at Svr { (*@\label{line:2fachoice2}@*)
+			Authorised() from Svr to Client;
+			do Main(Svr, Client);
+		} or {
+			AccessDenied() from Svr to Client;
+		}
+	} or {
+		AccessDenied() from Svr to Client;	
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{The \tprotocol{Two Factor Authentication} Protocol}
+\label{lst:2FA}
+\end{figure}
+
+The developer's implementation on 
+\cref{line:2fachoice1,line:2fachoice2} is likely to query
+a database for credential verification.
+Database operations are implemented using \textit{asynchronous},
+\textit{non-blocking} APIs; it is
+recommended that file operations do not block the main execution,
+so a JavaScript database API may require a \textit{callback}
+to propagate the return value when it is available.
+A server endpoint implementation of \cref{lst:2FA}
+may resemble the following:
+
+\begin{lstlisting}[language=javascript]
+new Implementation.S40({
+	Login: (username: string, password: string) => {
+		db.lookup(username, password, (result) => {
+			// returns Implementation.S42 here
+		});
+	}
+});
+\end{lstlisting}
+
+Callbacks are commonplace in dealing with concurrency in JavaScript
+\cite{CallbackHell}; we will go over other widely-used
+concurrency primitives shortly.
+Whatever the case may be, asynchronous developer implementations
+using our generated APIs will not type-check as the function
+returns immediately (hence, has return type \lstonelinejs{void}),
+even though the callback function has the expected return type.
+The developer will encounter the following:
+
+\begin{lstlisting}[tabsize=2,numbers=none]
+Type '(username: string, password: string) => void' is not 
+assignable to type '(payload_0: string, payload_1: string) => S42'.
+\end{lstlisting}
+
+This is also possible for browser-side implementations:
+when the user invokes a UI event, the browser-side logic
+may first consult a third-party API for additional information
+(which is not part of the communication protocol)
+before performing a send action; the \textit{Fetch} API \cite{Fetch}
+for making API calls is asynchronous.
+
+In order to make \fancyname{SessionTS} more relevant and compatible
+with modern web programming practices, it is important to
+support asynchronous developer implementations as part of our
+generated APIs and runtime.
+
+This means we should support
+the concurrency primitives built into TypeScript:
+
+\subparagraph{Callbacks}
+Callback-based APIs are higher-order functions -- they take functions
+as parameters and invoke them when the asynchronous operation is complete.
+The \texttt{setTimeout(fn, delay)} function available on the browser
+waits for \texttt{delay} milliseconds before invoking \texttt{fn()} --
+or more precisely, before queuing the execution of \texttt{fn()}.
+As a result, the code listing below prints \lstonelinejs{'Second!'}
+before \lstonelinejs{'First!'}, even though \texttt{setTimeout}
+was called with 0ms delay.
+
+\begin{lstlisting}[language=javascript,numbers=none]
+setTimeout(() => console.log('First!', 0);
+console.log('Second!');
+\end{lstlisting}
+
+\subparagraph{Promises}
+A Promise formalises the ``completion'' callback
+and its error handling construct. The developer passes a ``success''
+handler and ``error'' handler, conventionally named
+``then'' and ``catch'' respectively, to construct a Promise.
+This is analogue to the \texttt{Future} interface in Java \cite{JavaFuture}.
+The Promise will invoke the correct handler on completion.
+The benefits of Promises in our context is that TypeScript allows
+us to specify the type of payload that the Promise resolves to.
+
+\begin{lstlisting}[language=javascript,numbers=none]
+declare function promise(n: number): Promise<number>
+const promise = (n: number) => 
+	new Promise((resolve, reject) => {
+		if (n < 10)  resolve(n));
+		else reject('Number too big!');
+	});
+
+declare function handleNumber(n: number) { ... } 
+declare function handleError(err: string) { ... }
+
+promise(10).then(handleNumber).catch(handleError)
+\end{lstlisting}
+
+\subparagraph{Async/Await}
+The \lstonelinejs{async}/\lstonelinejs{await} construct
+is syntactic sugar for Promises. An \lstonelinejs{async}
+function always returns a Promise, and marking \lstonelinejs{await}
+for the identifier of the return value for a function returning a Promise
+``unwraps'' the resolved payload. A downside is that error handling
+reverts to the \lstonelinejs{try ... catch} construct.
+
+\begin{lstlisting}[language=javascript,numbers=none]
+const asyncSum = async (x: number, y: number) => {
+	return (await promise(x)) + (await promise(y));
+}
+
+asyncSum(1, 2);  // returns Promise<number> that resolves to 3
+\end{lstlisting}
 
@@
 
+\begin{lstlisting}[language=javascript,numbers=none]
+type MaybePromise<T> = T | Promise<T>;
+type FromPromise<T> = T extends MaybePromise<infer R> ? R : never;
+\end{lstlisting}
+
+\begin{lstlisting}[language=javascript,tabsize=2]
+new Implementation.S40({
+	Login: (*@\hl{async}@*) (username: string, password: string) => {
+		const result = (*@\hl{await}@*) db.lookup(username, password);
+		if (result.valid)
+			return new Implementation.S42('Authorised', ...);
+		else if (result.unauthorised)
+			return new Implementation.S42('AccessDenied', ...);
+		else
+			return new Implementation.S42('Challenge', ..., challenger);
+	}
+});
+
+const challenger = (key: number) => {
+	return new (*@\hl{Promise}@*)((resolve, reject) => {
+		keychain.check(key, ok => {
+			if (ok)
+				resolve(new Implementation.S44('Authorised', ...);
+			else 
+				resolve(new Implementation.S44('AccessDenied', ...);	
+		});
+	});
+}
+\end{lstlisting}
+
 \subsection{Runtime Extension}
+
+Generally, minimal changes are required for the runtime.
+The key observation is that the concrete types for the any
+handlers are the same -- they just might be wrapped in a Promise
+construct.
+To handle potential Promises, we wrap existing logic
+as a delayed computation, colloquially known as a \textit{thunk}
+\cite{Thunk}.
+
 \begin{itemize}
@@
 
+\begin{figure}[!h]
+\begin{subfigure}{\textwidth}
+\begin{lstlisting}[language=javascript,tabsize=2]
+const [	label, payload, succ] = this.handler;
+send(label, payload);
+next(succ);
+\end{lstlisting}
+\caption{\dots}
+\end{subfigure}
+\hfill
+\begin{subfigure}{\textwidth}
+\begin{lstlisting}[language=javascript,tabsize=2]
+const thunk = (label, payload, succ) => {
+	send(label, payload);
+	next(succ);	
+};
+
+if (this.handler instanceof Promise)
+	this.handler.then(thunk);
+else 
+	thunk(this.handler);
+\end{lstlisting}
+\caption{\dots}
+\end{subfigure}
+\captionof{figure}{\dots}
+\end{figure}
+
+\begin{lstlisting}[language=javascript,tabsize=2]
+(*@\hl{let used = false;}@*)
+...
+return class extends React.Component {
+	render() {
+		const props = {
+			[eventLabel as string]: (event) => {
+				(*@\hl{if (used) return;}@*)
+				(*@\hl{used = true;}@*)
+
+				const result = handler(event);
+				if (result instanceof Promise) {
+					result.then(send);				
+				} else {
+					send(result);				
+				}
+			}		
+		}
+	}
+}
+\end{lstlisting}
+
 \subsection{Limitations}
-\begin{itemize}
-\item for the browser target, we technically still provide affine linearity in the code, but the user would be able to click a button twice when it is sending as the promise resolution is asynchronous; to solve this, we add the used flag which is used in java paper for detecting channel reuse, which is `technically' a runtime check, depending on how one interprets
-\end{itemize}
+
+Supporting asynchronous logic on top of our
+existing implementation in \fancyname{ReactMPST} comes at a cost
+of losing affine linearity guarantees.
+If the send action handler is asynchronous, then the UI event handler
+returns before the runtime transitions, so the UI event that can trigger
+the send action is still active, which means the user can trigger
+a channel linearity violation. This means we need a boolean flag
+to keep track of channel usage for browser-side endpoints, 
+as done in \cite{Hybrid2016}.
+
+If we decouple the \texttt{sendMessage()} 
+from the \texttt{advance()} function, we could advance to the successor state
+immediately \textit{before} handling the asynchronous return value.
+This restores affine channel linearity guarantees, but 
+results in the UI being inconsistent with the channel
+actions.

### 2020-06-11 report/sections/evalconcl/eval/benchmarks.tex
Report: make graphs bigger
Report: fix bad subfig numbering

@@
 We visualise the elapsed time with respect to the number of \tmsg{PING}s
-received in \cref{fig:execution}.
+received in \cref{fig:execution}: each line represents one
+execution of the benchmark.
 
@@
 \centering
-\begin{subfigure}[b]{0.3\textwidth}
+\begin{subfigure}[b]{0.49\textwidth}
 \centering
@@
 \hfill
-\begin{subfigure}[b]{0.3\textwidth}
+\begin{subfigure}[b]{0.49\textwidth}
 \centering
@@
 \end{subfigure}
-\captionof{figure}{Comparison of Execution Pattern for 10,000 Ping-Pongs}
+\caption{Comparison of Execution Pattern for 10,000 Ping-Pongs}
 \label{fig:execution}
@@
 \end{subfigure}
-\captionof{figure}{Comparison of \texttt{bare} and \texttt{mpst}
+\caption{Comparison of \texttt{bare} and \texttt{mpst}
 implementations for \tprotocol{Ping Pong} \trole{Svr}}

### 2020-06-11 report/sections/server/extensions/async.tex
Report: complete section 6.1 on async

@@
 
+Asynchronous APIs are commonplace in modern web programming,
+ranging from database operations to third-party API calls.
+We motivate the need for supporting asynchronous operations
+through an authentication protocol in
+\cref{subsection:asyncmotivation},
+demonstrate how we support this through extending
+the generated APIs and runtime in
+\cref{subsection:asyncapi,subsection:asyncruntime}
+respectively, 
+and highlight the limitations of our approach in
+\cref{subsection:asynclimit}.
+
 \subsection{Motivation}
+\label{subsection:asyncmotivation}
 
@@
 \subsection{API Extension}
-\begin{itemize}
-\item should be very short section -- use union type MaybePromise so developer can provide either
-\end{itemize}
+\label{subsection:asyncapi}
+
+This is rather straightforward.
+We construct a generic union type, \texttt{MaybePromise<T>},
+to specify that the type parameter might be wrapped in a Promise,
+and leverage type inference in conditional types to
+define a generic type, \texttt{FromPromise<T>}, to extract
+the wrapped type.
 
@@
 
+Now we change the type signatures of our generated APIs
+to use \texttt{MaybePromise<T>}.
+For example, if the handler of a receive state generated by
+\fancyname{NodeMPST} must return \texttt{Implementation.S42},
+it is now permitted to return a \texttt{Promise} that resolves
+to a value of type texttt{Implementation.S42}.
+The same logic applies to other types of EFSM states
+in both \fancyname{NodeMPST} and \fancyname{ReactMPST}.
+
+Hence, the asynchronous \trole{Svr} implementation
+for the 2FA protocol shown in \cref{lst:new2fa}
+now type-checks. Note that, because 
+\lstonelinejs{async} functions return \texttt{Promise}s,
+we also support this newer piece of syntactic sugar.
+
+\begin{figure}[!h]
 \begin{lstlisting}[language=javascript,tabsize=2]
@@
 \end{lstlisting}
+\captionof{lstlisting}{\tprotocol{2FA} \trole{Svr} Implementation using 
+Callbacks, \texttt{Promise}s and
+\lstonelinejs{async}/\lstonelinejs{await}}
+\label{lst:new2fa}
+\end{figure}
 
 \subsection{Runtime Extension}
+\label{subsection:asyncruntime}
 
@@
 \cite{Thunk}.
+We show an example in \cref{fig:thunk} of how we
+adapt the \texttt{performSend()} function generated by
+\fancyname{NodeMPST} to support optional \texttt{Promise}s:
+
+\begin{enumerate}
+\item
+We wrap the original logic in \cref{subfig:originalthunk} into a
+function (\cref{subfig:newthunk} \cref{line:thunk})
+to delay the computation.
+
+\item
+We distinguish asynchronous implementations
+from synchronous ones using the \lstonelinejs{instanceof} operator.
 
 \begin{itemize}
-\item server send actions -- use the Thunk pattern to wrap the existing logic, the thunk takes the original payload type (extracted using conditional infer types) as parameter; implementation either returns promise or not-promise, so the implementation will either pass thunk as resolve or call thunk as-is
-\item server receive actions - thunk pattern as well
-\item browser send action -- need a runtime flag to silently disable channel reuse (which would otherwise violate linearity) because the sending action is delayed by the Promise resolve
-\item browser receive action -- thunk pattern
+\item
+If the implementation is a \texttt{Promise}, we pass
+the thunk as the resolve function invoked on completion.
+
+\item
+Otherwise, we directly invoke the computation.
 \end{itemize}
+
+\end{enumerate}
 
@@
 \begin{lstlisting}[language=javascript,tabsize=2]
-const [	label, payload, succ] = this.handler;
+const [label, payload, succ] = this.handler;
 send(label, payload);
@@
 \end{lstlisting}
-\caption{\dots}
+\caption{Original \texttt{performSend()} Implementation}
+\label{subfig:originalthunk}
 \end{subfigure}
@@
 \begin{lstlisting}[language=javascript,tabsize=2]
-const thunk = (label, payload, succ) => {
+const thunk = (label, payload, succ) => { (*@\label{line:thunk}@*)
 	send(label, payload);
@@
 \end{lstlisting}
-\caption{\dots}
+\caption{New \texttt{performSend()} for Optional \texttt{Promise}s}
+\label{subfig:newthunk}
 \end{subfigure}
-\captionof{figure}{\dots}
+\caption{Adapting Send Operations in \fancyname{NodeMPST}
+using a Thunk}
+\label{fig:thunk}
 \end{figure}
 
+Similar changes apply to other types of states in both
+\fancyname{NodeMPST} and \fancyname{ReactMPST}.
+For \fancyname{ReactMPST}, channel-related functions in the runtime
+return the successor state identifier. Hence, aside from
+refactoring the functions to delay computation using thunks,
+the functions themselves need to
+return \texttt{MaybePromise<State>} instead
+to propagate the asynchrony
+to the execution of the EFSM -- the transition occurs
+\textit{after} the developer's handler finishes execution.
+
+\subsection{Limitations}
+\label{subsection:asynclimit}
+
+Supporting asynchronous logic on top of our
+existing implementation in \fancyname{ReactMPST} comes at a cost
+of losing affine linearity guarantees.
+If the send action handler is asynchronous, then the UI event handler
+returns before the runtime transitions, so the UI event that can trigger
+the send action is still active, which means the user can trigger
+a channel linearity violation. This means we need a boolean flag
+to keep track of channel usage, as done in \cite{Hybrid2016}.
+We illustrate this in \cref{lst:asynclinearcheck},
+with the linearity checks highlighted.
+
+\begin{figure}[!h]
 \begin{lstlisting}[language=javascript,tabsize=2]
@@
 				const result = handler(event);
-				if (result instanceof Promise) {
+				if (result instanceof Promise)
 					result.then(send);
-				} else {
+				else
 					send(result);
@@
 }
-}
 \end{lstlisting}
-
-\subsection{Limitations}
-
-Supporting asynchronous logic on top of our
-existing implementation in \fancyname{ReactMPST} comes at a cost
-of losing affine linearity guarantees.
-If the send action handler is asynchronous, then the UI event handler
-returns before the runtime transitions, so the UI event that can trigger
-the send action is still active, which means the user can trigger
-a channel linearity violation. This means we need a boolean flag
-to keep track of channel usage for browser-side endpoints, 
-as done in \cite{Hybrid2016}.
+\captionof{lstlisting}{\dots}
+\label{lst:asynclinearcheck}
+\end{figure}
 

### 2020-06-11 report/sections/server/extensions/async.tex
Report: minor fixes to section 6.1

@@
 
-if (this.handler instanceof Promise)
+if (this.handler instanceof Promise) {
 	this.handler.then(thunk);
-else
+} else {
 	thunk(this.handler);
+}
 \end{lstlisting}
@@
 \begin{figure}[!h]
-\begin{lstlisting}[language=javascript,tabsize=2]
+\begin{lstlisting}[language=javascript]
 (*@\hl{let used = false;}@*)
@@
 				const result = handler(event);
-				if (result instanceof Promise)
+				if (result instanceof Promise) {
 					result.then(send);
-				else
+				} else {
 					send(result);
@@
 	}
+}
 \end{lstlisting}
-\captionof{lstlisting}{\dots}
+\captionof{lstlisting}{Generating Runtime Linearity Checks for
+React Send State Components}
 \label{lst:asynclinearcheck}

### 2020-06-11 report/sections/evalconcl/concl/future.tex
Report: address some NY comments

--- 2020-06-07 report/sections/evalconcl/concl/future.tex
@@
 
-\subsection{Explicit Connection Actions}
+\paragraph{``Serverless'' Server-Side Endpoints}
+\dots
+
+\paragraph{Explicit Connection Actions}
 Hu and Yoshida \cite{ExplicitConnections} extends 
@@
 
-\subsection{Generalised EFSM Encoding as Typestates}
+\paragraph{EFSM Encoding as Typestates}
 The concept of \textit{typestates} define the interface of

### 2020-06-11 report/sections/introduction/contributions.tex
Report: address some NY comments

@@
 
-In \cref{chap:background}, we introduce the core session type theory
+In \textbf{\cref{chap:background}}, 
+we introduce the core session type theory
 and provide an overview of related work on integrating session types into
@@
 We present an end-to-end workflow for generating TypeScript APIs for 
-\emph{server-centric} communication protocols over WebSockets. Server-centric 
-protocols feature exactly one server role that is involved in all interactions.
+\emph{server-centric} communication protocols over WebSockets. 
+Server-centric 
+protocols feature exactly one server role that is involved in all 
+interactions.
 
-In \cref{chap:codegen}, we introduce the system architecture for our
+In \textbf{\cref{chap:codegen}}, 
+we introduce the system architecture for our
 code generator, \fancyname{SessionTS}. 
@@
 TypeScript web development as \fancyname{NodeMPST} and \fancyname{ReactMPST}
-in \cref{chap:node} and \cref{chap:react} respectively.
+in \textbf{\cref{chap:node}} and \textbf{\cref{chap:react}} respectively.
 
-In \cref{chap:ext}, we extend \fancyname{SessionTS} to support advanced
+In \textbf{\cref{chap:ext}},
+we extend \fancyname{SessionTS} to support advanced
 practices commonly found in full-stack web development in a way that
 preserves communication safety. We modify the API generation strategies
-to support asynchronous implementations and introduce how to 
-gracefully handle errors and session cancellation 
+to support \textit{asynchronous implementations} 
+and introduce how to 
+gracefully handle errors and \textit{session cancellation} 
 in the context of interactive web applications.
@@
 We present a method to relax the server-centric constraint over WebSocket
-transport in order to support peer-to-peer interactions amongst 
+transport in order to support \textit{peer-to-peer interactions} 
+amongst 
 client roles, through the concept of \emph{routing}.
 
-In \cref{chap:p2p}, we introduce the \tprotocol{Two Buyer} multiparty
+In \textbf{\cref{chap:p2p}}, 
+we introduce the \tprotocol{Two Buyer} multiparty
 protocol to motivate our proposal of providing the server role with 
@@
 
-In \cref{chap:theory}, we introduce a variant of multiparty 
+In \textbf{\cref{chap:theory}}, 
+we introduce a variant of multiparty 
 session type theory with routing. We prove that our extensions preserve 
@@
 
-In \cref{chap:impl}, we demonstrate how our API generation strategies
+In \textbf{\cref{chap:impl}}, 
+we demonstrate how our API generation strategies
 implement our routed session type theory to support arbitrary communication
@@
 \paragraph{\cref{part:evalconcl}}
-In \cref{chap:eval}, we evaluate our work through 
+In \textbf{\cref{chap:eval}}, 
+we evaluate our work through 
 case studies of communication protocols 
@@
 
-Finally, we summarise our work in \cref{chap:concl} and propose potential
+Finally, we summarise our work in \textbf{\cref{chap:concl}} 
+and propose potential
 areas of improvement and future work.

### 2020-06-11 report/sections/server/react/runtime.tex
Report: address some NY comments

@@
 \subsection{Sending Messages}
+\label{subsection:reactruntimesend}
+
 Messages are sent through the WebSocket in the same way 
@@
 	return <K extends keyof DOMEvents>
-				 (ev: K, handler: EventHandler<E, K> => {
+				 (ev: K, handler: EventHandler<E, K>) => {
 		const send = (payload: T) => (*@\label{line:sendfactorycurrysend}@*)

### 2020-06-11 report/sections/server/extensions/async.tex
Report: progress on chapter 6 error handling

@@
 
-If we decouple the \texttt{sendMessage()} 
+Referring back to the \texttt{sendMessage()} method
+defined in \cref{subsection:reactruntimesend},
+if we decouple the \texttt{send()} 
 from the \texttt{advance()} function, we could advance to the successor state

### 2020-06-11 report/sections/server/extensions/error.tex
Report: progress on chapter 6.1
Report: progress on chapter 6 error handling

@@
 \section{Error Handling}
+\label{section:error}
+
+A robust error handling framework is critical
+in a distributed setting.
+It is naive to assume that endpoint implementations
+are entirely free from exceptions.
+As Fowler pointed out in \cite{Exceptional},
+session type implementations that do not account for failure
+are of limited use in distributed programming.
+
+We motivate the need for handling a variety exceptions in
+modern web programming in \cref{subsection:errormotivation},
+demonstrate how we implement a structured error handling
+framework through extending the generated APIs and 
+runtime in \cref{subsection:errorapi,subsection:errorruntime}
+respectively, and explain the limitations of our framework
+in \cref{subsection:errorlimit}.
 
 \subsection{Motivation}
-\begin{itemize}
-\item existing work extends theory based on canonical session cancellations are from exceptions raised by logic
-\item refer to 2FA example in Fowler paper on exceptional session types
-\item need to also consider cancellations in a web-based scenario -- server dead, client closing their browser
-\item need to also offer developer a seam for handling cancellations -- this needs to be well-parameterised so the developer knows how to handle the cancellation (for example, which role cancelled -- e.g.
-if it is a multi-way transaction, the server needs to know which role cancelled, so they can update the database accordingly).
-\end{itemize}
+\label{subsection:errormotivation}
+
+Using the \tprotocol{2FA} example from
+\cref{lst:2FA}, the database lookup function
+might rely on a cloud service, and will throw an exception
+if the cloud service cannot be reached.
+This may cause \trole{Svr} to crash.
+From the perspective of session types, we need to
+ensure that \trole{Client} isn't blindly
+waiting for the session to continue if \trole{Svr}
+threw an exception.
+
+We classify this as an example where \trole{Svr}
+terminates the session because of a \textit{logical error}
+in its implementation.
+Existing work \cite{Exceptional,AffineSessions} 
+on error handling for session types address this
+through extending the process calculus with an
+\textit{explicit cancellation} operator to ensure that
+channel resources are closed during exception handling.
+
+However, the context of modern web programming
+introduces new possibilities for session cancellation
+that are not addressed by existing work.
+Browser-side endpoints can disconnect from the session due
+to network connectivity issues or simply by closing
+the browser. Similarly, the server may also face connectivity
+issues. In order to deliver a robust error handling
+framework tailored for integrating session types in 
+modern web programming, we must also take these variants
+of session cancellation into account.
+
+We also need a novel approach for defining cancellation handlers.
+The work by Fowler et al. \cite{Exceptional} integrates 
+exception handling for session types in a 
+\textit{functional language}, where their design of
+exception handlers build upon existing work on effect handlers.
+Our work targets a non-functional language in TypeScript,
+which means any piece of code can be ``effectful'',
+so the exception handler API we generate for developers
+need to respect this property.
+Exception handlers should also be well parameterised
+to empower the developer to handle cancellations more effectively
+-- for instance, if the developer knows which endpoint
+emitted the cancellation, appropriate clean-up operations
+(such as reverting database changes or application-specific
+logging) can be performed.
 
 \subsection{API Extension}
+\label{subsection:errorapi}
+
 \begin{itemize}
@@
 
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript]
+new ATMSvr(wss, logic, async (role: Roles.All, reason: any) => {
+	if (role === Roles.Self) {
+		// Handle internal server error	
+	} else {
+		// Handle client error:
+		// revert incomplete transactions from database
+		await db.revertTransactions(role);
+	}
+});
+\end{lstlisting}
+\captionof{lstlisting}{Example Cancellation Handler for Server Endpoint}
+\label{lst:cancelsvr}
+\end{figure}
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript]
+<ATMClient
+	endpoint='ws://localhost:8080'
+	...
+	cancellation={(role: Roles.All, reason: any) => (
+		<div>
+			{role === Roles.Self
+				? <h1>Internal Error</h1>
+				: <h1>Server Error</h1>}
+			<p>{reason}</p>
+		</div>
+	)}	
+/>
+\end{lstlisting}
+\captionof{lstlisting}{Example Cancellation Handler for Browser Endpoint}
+\label{lst:cancelclient}
+\end{figure}
+
 \subsection{Runtime Extension}
+\label{subsection:errorruntime}
+
+The generated session runtime \dots
+
 \begin{itemize}
@@
 \subsection{Limitations}
+\label{subsection:errorlimit}
+
+The expressiveness of our error handling framework is
+limited by the expressiveness of the 
+\lstonelinejs{try} / \lstonelinejs{catch} mechanism
+in TypeScript.
+
 \begin{itemize}

### 2020-06-11 report/sections/server/extensions/error.tex
Report: progress on chapter 6 error

@@
 
-\begin{itemize}
-\item developers just need to provide cancellation handlers along with the initial state
-\item we standardise on parameterising it with role and reason: the reason is typed `any' to give developers the autonomy for implementation details
-\end{itemize}
+We extend the generated APIs to
+allow developers to provide a \textbf{session cancellation
+handler}. The runtime will invoke this
+to handle \textit{local} exceptions (e.g. exception
+thrown by application logic) and \textit{global} session
+cancellations (e.g. other endpoint disconnecting).
+
+The session cancellation handler is uniformly
+parameterised by the \texttt{role} which emitted the cancellation,
+and some arbitrary \texttt{reason}. For exceptions thrown
+by local application logic, the \texttt{reason} argument
+is populated with the thrown exception. Because the 
+\lstonelinejs{try} / \lstonelinejs{catch} block is not typed,
+we give the \texttt{reason} parameter the dynamic type of
+\lstonelinejs{any} to give developers the autonomy for 
+handling the exceptions they throw locally.
+
+Consider a binary session protocol specifying interactions
+between an \trole{ATMSvr} and \trole{ATMClient}.
+We demonstrate how the developer may choose to define
+the session cancellation handlers in the context of this protocol.
+
+\subparagraph{Server (\cref{lst:cancelsvr}):}
+The \trole{ATMSvr}
+may need to handle internal server errors 
+(\cref{line:cancelsvrinternal}) as a result of
+exceptions thrown by its implementation. 
+If the \trole{ATMClient} disconnects prematurely, 
+\trole{ATMSvr} can also notify the database 
+to revert the transaction (\cref{line:cancelsvrexternal}). 
+Note that session cancellation handlers can be 
+\textit{asynchronous}, as we appreciate that any required
+``clean-up'' operations may need to use asynchronous APIs
+provided by databases or other persistent storage solutions.
+The cancellation handler is passed into the runtime via
+the constructor of the public API.
 
@@
 	if (role === Roles.Self) {
-		// Handle internal server error	
+		// Handle internal server error	(*@\label{line:cancelsvrinternal}@*)
 	} else {
@@
 		// revert incomplete transactions from database
-		await db.revertTransactions(role);
+		await db.revertTransactions(role); (*@\label{line:cancelsvrexternal}@*)
 	}
@@
 
+\subparagraph{Client (\cref{lst:cancelclient})}
+Session cancellation handlers for \fancyname{ReactMPST}
+are also parameterised by the \texttt{role} and \texttt{reason} for
+cancellation, but they must return a \texttt{ReactNode}
+for the runtime to render upon session cancellation, which
+guarantees that no channel actions will be rendered when a session
+is cancelled. Here, the developer renders a header depending
+on which role emitted the cancellation, and directly shows
+the reason in a paragraph. 
+The cancellation handler is passed into the runtime via a 
+prop defined on the public API.
+
 \begin{figure}[!h]
@@
 <ATMClient
-	endpoint='ws://localhost:8080'
 	...
@@
 		<div>
+			// Customise heading to show which role cancelled
 			{role === Roles.Self
@@
 				: <h1>Server Error</h1>}
-			<p>{reason}</p>
+			<p>{reason}</p>		// Render reason in paragraph
 		</div>
@@
 
-The generated session runtime \dots
+The generated session runtime is extended
+with the capability of \textit{emitting} session
+cancellation and \textit{handling} session cancellation.
+We define session cancellation over WebSocket transport
+by invoking the \texttt{close()} method to emit cancellations,
+and overriding the \texttt{onclose} event listener to handle
+cancellations.
 
-\begin{itemize}
-\item build upon websocket RFC which specifies the `base' closing (or exit) code
-\item we additionally define contracts for exit codes in our use case (e.g. role occupied, logical error), and make sure it does not override any websocket `traditions' as per the RFC
-\item all based on well-defined event listeners on the websocket close and error events
-\item must consider cancellation in two phases
-\end{itemize}
+The WebSocket \texttt{close()} method accepts a 
+\textit{close code}, which is analogue to exit codes for
+executables. We extend this mechanism to distinguish
+between the different variants of session cancellation,
+as motivated in \cref{subsection:errormotivation}.
+As shown in \cref{fig:cancelenum},
+we define unique close codes for different types
+of session cancellation under \filename{Cancellation.ts}
+for both server-side (\cref{subfig:cancelenumsvr})
+and browser-side endpoints (\cref{subfig:cancelenumclient}) 
+respectively. 
+To respect the convention specified in the WebSocket protocol
+RFC \cite{WebSocketRFC} that the 1xxx close codes 
+are reserved for internal use,
+we define our custom close codes from 4000 onwards.
+The enum values are named in a self-explanatory way.
 
-\subsubsection{Cancellations during Session Joining}
+\begin{figure}[!h]
+\begin{subfigure}{0.48\textwidth}
+\begin{lstlisting}[language=javascript,tabsize=2]
+export enum Receive {
+	NORMAL = 1000,
+	CLIENT_BROWSER_CLOSED = 1001,
+	LOGICAL_ERROR = 4001,
+};
 
-\begin{itemize}
-\item server -- emits role occupied
-\item client -- closes browser whilst waiting for others
-\end{itemize}
+export enum Emit {
+	CLIENT_BROWSER_CLOSED = 4000,
+	LOGICAL_ERROR = 4001,
+	ROLE_OCCUPIED = 4002,
+};
+\end{lstlisting}
+\caption{Close Codes for \fancyname{NodeMPST}}
+\label{subfig:cancelenumsvr}
+\end{subfigure}
+\hfill
+\begin{subfigure}{0.48\textwidth}
+\begin{lstlisting}[language=javascript,tabsize=2]
+export enum Receive {
+	NORMAL = 1000,
+	SERVER_DISCONNECT = 1006,
+	CLIENT_DISCONNECT = 4000,
+	LOGICAL_ERROR = 4001,
+	ROLE_OCCUPIED = 4002,
+};
 
-\subsubsection{Cancellations during EFSM Execution}
+export enum Emit {
+	NORMAL = 1000,
+	LOGICAL_ERROR = 4001,
+};
+\end{lstlisting}
+\caption{Close Codes for \fancyname{ReactMPST}}
+\label{subfig:cancelenumclient}
+\end{subfigure}
+\caption{WebSocket Close Codes for Session Cancellation}
+\label{fig:cancelenum}
+\end{figure}
 
-\begin{itemize}
-\item server -- dies
-\item server -- logical error
-\item client -- dies
-\item client -- logical error
-\end{itemize}
+The session runtime features two phases:
+\textbf{(1)} the \textit{connection} phase, when the server
+accepts incoming connections from session participants;
+and \textbf{(2)} the \textit{execution phase}, when all
+participants execute their EFSM.
+Cancellation can occur in both phases, and we outline
+how the runtime handles cancellation in each phase.
+
+\subsubsection{Cancellations during \textit{Connection} Phase}
+As the global protocol has not began at this phase,
+any cancellations that occur here do \textit{not} terminate
+the connection phase; the endpoints that have already
+joined just simply continue waiting for the remaining participants
+to connect.
+
+The server endpoint can emit a \texttt{ROLE_OCCUPIED} close event
+to the browser endpoint if the role is already taken.
+We know if a role is occupied if the payload of the
+incoming connection request message is not one of the roles
+that the server endpoint is waiting on.
+
+It is possible for a browser endpoint to have already joined
+the session but is still in the connection phase, because the
+server endpoint is waiting on other participants. If this browser
+endpoint disconnects prematurely (e.g. the user closes
+the browser), the server endpoint detects this and adds the role
+back to the set of roles it is waiting on. To do this, we keep
+a mapping of socket to role identifier; this is populated
+when the endpoint sends the connection request message,
+so if the endpoint disconnects prematurely, the server knows
+which role to wait for again. We show the changes made to 
+the connection management logic generated by \fancyname{NodeMPST}
+in \cref{lst:cancelconnectsvr}.
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript]
+// Inside constructor of session runtime for server-side endpoint
+
+const socketToRole = new Map<WebSocket, Roles.Peers>();
+
+const onClose = ({ target: socket }) => {
+	// Wait for the role again
+	waiting.add(socketToRole.get(socket));
+}
+
+const onSubscribe = (event) => {
+	...
+	// Update role-WebSocket mapping.
+	roleToSocket[role] = socket;
+	socketToRole.set(socket, role);
+	waiting.delete(role);
+	...
+}
+\end{lstlisting}
+\captionof{lstlisting}{\dots}
+\label{lst:cancelconnectsvr}
+\end{figure}
+
+\subsubsection{Cancellations during \textit{Execution} Phase}
+Any cancellation event that occurs during the execution of the
+EFSM will terminate the protocol for all roles.
+We see the concept of recovery to be an implementation detail
+unique to each application, so we do not build recovery
+mechanisms in our session cancellation framework.
+There are four possible ways for a cancellation event
+to be emitted during this phase.
+
+\subparagraph{Server endpoint experiences logical error:}
+This is when the developer's implementation for the server
+endpoint throws an uncaught exception. 
+We wrap the runtime EFSM execution methods in
+\lstonelinejs{try}/\lstonelinejs{catch}
+blocks, and handle the error with a new \texttt{cancel()}
+method we generate for the runtime in \fancyname{NodeMPST}.
+The \texttt{cancel()} method invokes
+the \texttt{close()} method on all browser endpoints,
+then executes the user-defined cancellation handler.
+
+Browser endpoints detect this through the \texttt{onclose}
+WebSocket event listener. We extend the runtime
+generated by \fancyname{ReactMPST} to handle this event
+in the corresponding case of a switch statement
+defined on the close code, as shown below.
+
+\begin{lstlisting}[language=javascript,tabsize=2]
+private onClose({ code, reason }: CloseEvent) {
+	switch (code) {
+		case Cancellation.Receive.NORMAL: {
+			// Normal, clean cancellation
+			return;
+		}
+		case Cancellation.Receive.SERVER_DISCONNECT: {
+			// Server role disconnected
+			this.processCancellation(Roles.Server, 'server disconnected')
+			return;
+		}
+		case Cancellation.Receive.CLIENT_DISCONNECT: { ... }
+		case Cancellation.Receive.LOGICAL_ERROR: { ... }
+		default: { ... }
+	}
+}
+\end{lstlisting}
+
+\begin{lstlisting}[language=javascript,tabsize=2]
+private processCancellation(role: Roles.All, reason?: any) {
+	const cancelUI = ...
+	this.setState({
+		elem: this.props.cancellation(role, reason !== undefined ? String(reason) : reason),
+        });
+    }
+\end{lstlisting}
+
+\subparagraph{Server endpoint disconnects:}
+This follows the same logic as the above,
+except for the fact that the server does not
+\textit{emit} the cancellation through code, since
+it simply disconnects.
+
+\subparagraph{Browser endpoint experiences logical error:}
+% emit event to server
+% server needs to propagate cancellation
+% other browser endpoints receive in the same wayy
+
+\subparagraph{Browser endpoint disconnects:}
+% same as above, except the `emit' part is implicit
+This follows the same logic as the above,
+except for the fact that the browser does not
+\textit{emit} the cancellation through code, since
+it simply disconnects.
+
+\subsection{Caveats with Asynchronous Operations}
+Implementing session cancellation whilst
+preserving support for asynchronous operations introduces
+subtle caveats.
+
+Consider the case where the developer implements
+some asynchronous logic for a send state on the server endpoint. 
+Whilst the \texttt{Promise} is being resolved, a cancellation
+event occurs concurrently. The session will be cancelled, but
+the asynchronous logic will \textit{not} be pre-empted; it
+will run to completion, and proceed to try perform the send
+action on a WebSocket that has already been closed.
+
+Traditionally, this will throw an error. However,
+the semantics of cancellation should dominate any asynchronous
+operations from the developer, so we need to work around
+the lack of pre-emption. We do this by defining a custom
+error handler for the WebSocket send action: in short,
+if a send action fails, and it is due to the session
+being cancelled prior, then there is nothing to worry about.
 
@@
 in TypeScript.
+Unlike languages like Java, we cannot enforce the type
+of exceptions being thrown, so we have to resort to
+typing the \texttt{reason} of the cancellation as 
+\lstonelinejs{any} to be compatible with 
+exception handling constructs defined by the developer.
 
-\begin{itemize}
-\item Cancellation handlers are not typed -- namely, we use `any';
-\item exception handler blocks in typescript are not typed
-\item the `logical error' idea is also very vague as all code can be effectful in typescript (statically); cannot specifically say that the implementation is expected to `send payload of type int, but possibly throw error of type Error', which would otherwise be possible in other languages(?)
-\end{itemize}
+We could change the session cancellation handler
+to an object with specific handlers for each of the
+custom WebSocket close codes we defined, but it
+clutters the API which increases the learning
+curve for the developer, and also makes it
+more difficult to share cancellation handling logic
+across different WebSocket close codes.

### 2020-06-11 report/sections/server/extensions/error.tex
Report: complete chapter 6

@@
 \end{lstlisting}
-\captionof{lstlisting}{\dots}
+\captionof{lstlisting}{Modified Connection Management in
+Server Endpoint with Cancellation}
 \label{lst:cancelconnectsvr}
@@
 \lstonelinejs{try}/\lstonelinejs{catch}
-blocks, and handle the error with a new \texttt{cancel()}
+blocks, and handle the error with a \texttt{cancel()}
 method we generate for the runtime in \fancyname{NodeMPST}.
@@
 in the corresponding case of a switch statement
-defined on the close code, as shown below.
+defined on the close code, 
+as shown in \cref{lst:canceloncloseclient}.
 
+\begin{figure}[!h]
 \begin{lstlisting}[language=javascript,tabsize=2]
@@
 	switch (code) {
-		case Cancellation.Receive.NORMAL: {
-			// Normal, clean cancellation
-			return;
-		}
+		case Cancellation.Receive.NORMAL: { ... }
 		case Cancellation.Receive.SERVER_DISCONNECT: {
@@
 \end{lstlisting}
+\captionof{lstlisting}{WebSocket Close Event Listener in
+Browser Endpoints}
+\label{lst:canceloncloseclient}
+\end{figure}
+
+Browser endpoints process the session cancellation
+by simply rendering the UI element generated
+by the user-supplied session cancellation handler.
 
@@
 private processCancellation(role: Roles.All, reason?: any) {
-	const cancelUI = ...
 	this.setState({
-		elem: this.props.cancellation(role, reason !== undefined ? String(reason) : reason),
+		elem: this.props.cancellation(role, reason);
 	});
@@
 \subparagraph{Browser endpoint experiences logical error:}
-% emit event to server
-% server needs to propagate cancellation
-% other browser endpoints receive in the same wayy
+This is when the developer's implementation for the browser
+endpoint throws an uncaught exception. 
+The runtime for the browser endpoint will catch
+the exception and emit the cancellation to the server
+using the corresponding close code.
+
+The server endpoint will detect the cancellation event
+using an \texttt{onclose} WebSocket event listener
+similar to that defined in \cref{lst:canceloncloseclient},
+Then, the server propagates the cancellation
+to the other endpoints.
+We present the propagation implementation
+in \cref{lst:cancelpropagate}.
+The ``payload'' of the propagated cancellation is also structured
+as an object with the role and reason as properties,
+so the browser endpoint can parse and handle using
+the same framework presented in \cref{lst:canceloncloseclient}.
+After propagating the cancellation to other endpoints,
+runtime invokes the user-defined session cancellation
+handler, then \textit{restarts} the runtime to accept
+new connections once again. 
+Any exceptions thrown by the developer's session cancellation
+handler will be ignored.
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript,tabsize=2]
+private propagateCancellation(cancelled: Roles.Peers, reason?: any)
+{
+	// Construct cancellation message as object
+	// with role and reason as properties
+	const message = Cancellation.toChannel(cancelled, reason);
+
+	// Emit cancellation to other roles
+	Object.entries(this.roleToSocket)
+		.filter(([role, _]) => role !== cancelled)
+		.forEach(([_, socket]) => {
+			socket.removeAllListeners();
+			socket.close(Cancellation.Emit.LOGICAL_ERROR,
+				JSON.stringify(message));
+		});
+
+	try {
+		// Execute user-defined cancellation handler
+		const doCancel = this.cancellation(cancelled, reason);
+		if (doCancel instanceof Promise) {
+			doCancel.then(this.restart).catch(this.restart);		
+		} else {
+			this.restart();
+		}
+	} catch {
+		this.restart();
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{Propagating Session Cancellation Events in
+Server Endpoint}
+\label{lst:cancelpropagate}
+\end{figure}
 
@@
 error handler for the WebSocket send action: in short,
-if a send action fails, and it is due to the session
+if a send action fails due to the session
 being cancelled prior, then there is nothing to worry about.

### 2020-06-12 report/sections/general/implementation.tex
Report: add chapter labels
Report: complete chapter 9

--- 2020-05-25 report/sections/general/implementation.tex
@@
 \chapter{Implementation}
+\label{chap:impl}
 
-\section{Extending \fancyname{NodeMPST}}
+In this chapter, we
+explain how \codegen
+is extended to implement \newtheory.
+As motivated in \cref{section:routerapproach},
+the routing mechanism is intended to be transparent
+to the developer.
+Hence, there are \textit{no changes} made to the
+generated APIs; we only need to adapt
+how the session runtime of the respective
+endpoints perform send and receive actions.
+We present the changes made to the session
+runtime generated by \nodecodegen
+and \reactcodegen in 
+\cref{section:routednode,section:routedreact}
+respectively.
 
-\section{Extending \fancyname{ReactMPST}}
+\section{Extending \nodecodegen}
+\label{section:routednode}
 
-\section{Limitations}
+By enabling peer-to-peer communication
+between browser endpoints through the server,
+not all messages received by the server are intended
+for the server role.
+We distinguish these by requiring all messages sent
+by browser endpoints to specify the intended recipient
+under the \texttt{role} property.
+We specify this through constructing an interface
+defining the structure of a 
+``general'' message received by the 
+server.
+Since this interface must match \textit{both}
+routed messages and messages intended for the server,
+we have to respect the fact that the server
+does not need to know the contents of routed messages,
+so the existing \texttt{label}
+and \texttt{payload} properties must be loosely typed
+to respect this notion of privacy.
+
+\begin{lstlisting}[language=javascript]
+// Inside the Message namespace...
+export interface Channel {
+	role: Roles.All, label: string, payload: any[]
+};
+
+// Build message object to send through WebSocket
+export const toChannel = 
+	(role: Roles.All, label: string, payload: any[]) => ({
+	role, label, payload
+});
+\end{lstlisting}
+
+We can now adapt the \texttt{onmessage}
+WebSocket event listener defined at the server endpoint
+to route messages not intended for the server role,
+as shown in \cref{lst:noderoutereceive}.
+By construction from our API generation strategy,
+the incoming message must match the interface defined
+above, so it is a safe deserialisation step.
+This preliminary deserialisation allows the server
+to identify \textit{routed messages}, and proceed to
+send it to the intended recipient.
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript]
+// Top-level parsing to distinguish between routed messages
+const { role, label, payload } = 
+	JSON.parse(data) as Message.Channel;
+if (role !== Roles.Self) {
+	// Route message
+	this.send(role, label, payload, from);
+} else {
+	// Invoke message handler as before, see (*@\cref{subsection:noderuntimereceive}@*)
+}
+\end{lstlisting}
+\captionof{lstlisting}{Modified \texttt{onmessage} Event Listener for
+\nodecodegen to Support Routing}
+\label{lst:noderoutereceive}
+\end{figure}
+
+Browser endpoints also expect to receive messages
+with the original sender explicitly marked under 
+the \texttt{role} property, so we modify the 
+\texttt{send()} method generated for the server endpoint
+to attach this extra piece of information.
+To maximise compatibility with the existing runtime
+implementation, we define the server as the default 
+argument for the \texttt{from} parameter
+of the \texttt{send()} method, as highlighted below.
+
+\begin{lstlisting}[language=javascript]
+send(to: Roles.Peers, label: string, payload: any[], 
+	(*@\hl{from: Roles.All = Roles.Self}@*)) {
+	const message = Message.toChannel(from, label, payload);
+	this.roleToSocket[to].send(JSON.stringify(message));
+}
+\end{lstlisting}
+
+\newtheory introduces an additional concern for managing
+the terminal state of the server's EFSM:
+when the server's EFSM reaches its terminal state,
+it does not mean other endpoints have reached their terminal state,
+as they can still interact amongst themselves.
+The server will need to remain connected to perform its routing duties.
+As our existing design choice already specifies that browser
+endpoints initiate the WebSocket close event when they reach
+their terminal state, we do not need to worry about this
+additional concern as the server will remain connected
+when it reaches its terminal state to carry out its routing
+duties anyway.
+
+\section{Extending \reactcodegen}
+\label{section:routedreact}
+
+Similarly, the session runtime for browser endpoints need
+to specify the intended recipient when sending a message through the
+WebSocket connection, in order for the server endpoint to route the message accordingly.
+Likewise, when registering receive handlers, the browser endpoint also 
+needs to specify the role to receive from in addition to the handler.
+
+As explained in \cref{subsection:reactruntimereceive},
+the current implementation of \reactcodegen has a pair of
+queues for messages waiting for handlers and vice-versa.
+Because \newtheory allows browser endpoints to interact
+with non-server endpoints, this existing framework may
+lead to communication mismatch.
+
+For example, consider the sequence of routed interactions between
+server endpoint $S$ and \textit{browser} endpoints $A$, $B$ and $C$:
+\[
+A \xrightarrow[S]{} B : \text{\lstonelinejs{M1(number)}}. 
+C \xrightarrow[S]{} B : \text{\lstonelinejs{M2(string)}}. \texttt{end}
+\]
+
+From the perspective of $B$, $B$ will register a receive handler expecting
+to process \lstonelinejs{M1(number)} from $A$ first.
+However, communication will be routed by
+some server endpoint $S$, and by \newtheory theory, $S$ can route 
+$C$'s \lstonelinejs{M2(string)} message
+to $B$ before $A$'s \lstonelinejs{M1(number)} message, 
+so the existing message receiving framework
+will use the receive handler defined for \lstonelinejs{M1(number)}
+to process the received \lstonelinejs{M2(string)}, which is a mismatch.
+
+Hence, we need to implement
+the message receiving framework outlined in \cref{subsection:noderuntimereceive}
+instead, which generalises the pair of message/handler queues to two mappings
+to provide a pair of message/handler queues for each endpoint.
+The runtime instantiates the two mappings of queues,
+and receive handler registration now targets the pair of queues
+specified by the role from which to receive from.
+We illustrate this in \cref{lst:reactrouterreceive}
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript]
+class Client extends React.Component<...> {
+	private messageQueue: RoleToMessageQueue;
+    private handlerQueue: RoleToHandlerQueue;
+
+    // Omitting unchanged methods
+
+	private registerReceiveHandler(
+		role: Roles.Peers, handle: ReceiveHandler) { 
+		const message = this.messageQueue[role].shift();
+		if (message !== undefined) {
+			// Message received already -- process as before
+		} else {
+			// No message received -- enqueue handler
+			this.handlerQueue[role].push(handle);
+		}
+	}
+}
+\end{lstlisting}
+\captionof{lstlisting}{Runtime Extensions for \reactcodegen to implement \newtheory}
+\label{lst:reactrouterreceive}
+\end{figure}
+
+Referring to the previous example under this framework,
+$S$ can route $C$'s message to $B$ first, but \lstonelinejs{M2(string)} will be stored
+under the message queue mapped to $C$, so it will not be processed until $B$ registers
+the receive handler for processing $C$'s message. This cannot happen until $B$ processes
+\lstonelinejs{M1(number)} from $A$, so the order of communication 
+specified in the global protocol is respected.

### 2020-06-12 report/sections/evalconcl/eval.tex
Report: complete chapter 10.1

--- 2020-06-01 report/sections/evalconcl/eval.tex
@@
 \label{chap:eval}
-%benchmark -- compile time / benchmarking
 
-\section{Binary Sessions: \fancyname{Calculator}}
-\label{section:evalcalculator}
+In this chapter, we
+evaluate the expressiveness
+and performance of our API generation
+toolchain.
+
+%TODO: add when finalised
+%with respect to the main objective
+%of \dots
+
+To evaluate the expressiveness
+of our work,
+we use three case studies
+of protocols found in webservices
+to demonstrate strengths and weaknesses
+of our work:
+
 \begin{itemize}
-\item sample client implementation + snippet of generated code (full in appendix)
-\item show that not all the UI is being controlled by the runtime (i.e. calculator app has a header/fake navbar)
-\item show the calculator logic achievable using React Context
-\item sample server implementation + snippet of generated code (full in appendix)
+
+\item
+\textbf{\tprotocol{Noughts and Crosses} Game:}
+
+In \cref{section:evalgame},
+we implement a classic multiplayer game which involves a game server
+and two players interacting with the game via the browser,
+to show that our work is compatible with multiparty sessions,
+which isn't the case in \cite{MVU2020,Exceptional}.
+We show how our generated APIs are compatible with the
+state management solution used by the game, and give examples of 
+how the generated APIs empower the developer to intuitively 
+implement the game interface.
+
+\item
+\textbf{\tprotocol{Two Buyers} Protocol, as introduced in \cref{section:twobuyer}:}
+
+In \cref{section:evaltwobuyer},
+we walk through an implementation of the \tprotocol{Two Buyers}
+protocol. This \textit{cannot} be implemented in
+\cite{PureScript2019,MVU2020}, which illustrates the novelty
+of our \newtheory theory on routed multiparty session types.
+We emphasise how the routing mechanism is transparent to the
+developer implementing the protocol, which demonstrates
+the practicality of our extensions to \codegen to support
+\newtheory.
+
+\item
+\textbf{\tprotocol{ATM} Protocol:}
+
+In \cref{section:evalatm},
+we implement an ATM web application to demonstrate
+limitations of our work. Our generated APIs do not allow
+the server endpoint (in this case, the \trole{Bank})
+to participate in concurrent sessions (i.e. interacting
+with different \trole{Client}s concurrently), even though
+the protocol is written in a way that allows this.
+However, we propose a way to address this limitation, and expand
+upon this in \cref{section:future}.
+
 \end{itemize}
 
-\begin{figure}[!ht]
-\begin{lstlisting}[language=Scribble, tabsize=2]
-module BinarySession;
+To evaluate the performance of our work,
+we run micro-benchmarks
+on a variant of the \tprotocol{Ping Pong}
+protocol parameterised by the number
+of round trips
+to analyse the overhead of our implementation,
+compared with implementations written
+without using our generated APIs,
+as the number of round trips increase.
+We describe the experiment methodology
+and comment on our findings in \cref{section:benchmarks}.
 
-global protocol Calculator(role Svr, role Client) {
-  choice at Client {
-    Add(number, number) from Client to Svr;
-    Res(number)         from Svr to Client;
-    do Calculator(Svr, Client);
-  } or {
-    Mult(number, number) from Client to Svr;
-    Res(number)          from Svr to Client;
-    do Calculator(Svr, Client);
-  } or {
-    Quit()               from Client to Svr;
-    Terminate()          from Svr to Client;
-  }
-}
-\end{lstlisting}
-\captionof{lstlisting}{The \tprotocol{Calculator} Protocol}
-\label{lst:calculator}
-\end{figure}
+\input{sections/evalconcl/eval/game}
 
-\section{Multiparty Sessions: \fancyname{Noughts and Crosses}}
-\label{section:evalgame}
-\begin{itemize}
-\item sample client implementation -- show the flexibility of factory API for sending to be used in the UI part (since move selection can be be made )
-\end{itemize}
+\input{sections/evalconcl/eval/twobuyer}
 
-\begin{figure}[!ht]
-\begin{lstlisting}[language=Scribble, tabsize=2]
-module NoughtsAndCrosses;
-type <typescript> "Coordinate" from "./Types" as Point;
-
-global protocol Game(role Svr, role P1, role P2) {
-	Pos(Point) from P1 to Svr;
-	choice at Svr {
-		Lose(Point)   from Svr to P2;
-		Win(Point)    from Svr to P1;
-	} or {
-		Draw(Point)   from Svr to P2;
-		Draw(Point)   from Svr to P1;
-	} or {
-		Update(Point) from Svr to P2;
-		Update(Point) from Svr to P1;
-		do Game(Svr, P2, P1);
-	}
-}
-\end{lstlisting}
-\captionof{lstlisting}{The \tprotocol{Noughts and Crosses} Protocol}
-\label{lst:game}
-\end{figure}
-
-\section{Routed Multiparty Sessions: \fancyname{Two Buyers}}
-\begin{itemize}
-\item show that the routing aspect is transparent to both sides
-\end{itemize}
+\input{sections/evalconcl/eval/atm}
 

### 2020-06-12 report/sections/evalconcl/eval/atm.tex
Report: complete chapter 10.1

\section{Delegated Sessions: \fancyname{ATM}}
\label{section:evalatm}

Interested readers can find the full implementation on
GitHub\footnote{\url{https://github.com/ansonmiu0214/SessionTS-Examples/ATM}}.

\begin{itemize}
\item Show that delegation is not supported
even if the protocol ``allows''
\end{itemize}

\begin{figure}[!ht]
\begin{lstlisting}[language=Scribble, tabsize=2]
module ATM;

global protocol ATM(role Bank, role Client) {
	choice at Client {
		// The string contains the client ID
		WITHDRAW(string, number) from Client to Bank;
		choice at Svr {
			BALANCE(number) from Bank to Client;		
		} or {
			FAIL(string) from Bank to Client;		
		}
		do ATM(Bank, Client);
	} or {
		DEPOSIT(string, number) from Client to Bank;
		BALANCE(number) from Bank to Client;
		do ATM(Bank, Client);
	} or {
		BYE(string) from Client to Bank;
		BYE() from Bank to Client;	
	}
}
\end{lstlisting}
\captionof{lstlisting}{The \tprotocol{ATM} Protocol}
\label{lst:atm}
\end{figure}

### 2020-06-12 report/sections/evalconcl/eval/game.tex
Report: complete chapter 10.1

\section{Multiparty Sessions: \tprotocol{Noughts and Crosses}}
\label{section:evalgame}

We implement the classic turn-based board game of \textit{Noughts and Crosses}
between two players,
as introduced in \cref{section:intro}.
We formalise the game interactions using a Scribble protocol
presented in \cref{lst:game}.
The \tprotocol{Game} protocol describes one turn:
\trole{P1} makes a move by sending the coordinates of a
vacant cell on the game board to \trole{Svr},
then \trole{Svr} reports the outcome of that move
to both players. If another round is required
to determine the game result, the \tprotocol{Game}
protocol is recursively invoked (\cref{line:gamerec}) with
roles \trole{P1} and \trole{P2} swapped.

\begin{figure}[!ht]
\begin{lstlisting}[language=Scribble]
module NoughtsAndCrosses;

// TypeScript definition:
// interface Point {x: number, y: number}
type <typescript> "Coordinate" from "./Types" as Point;

global protocol Game(role Svr, role P1, role P2) {
	Pos(Point) from P1 to Svr;
	choice at Svr {
		Lose(Point)   from Svr to P2;
		Win(Point)    from Svr to P1;
	} or {
		Draw(Point)   from Svr to P2;
		Draw(Point)   from Svr to P1;
	} or {
		Update(Point) from Svr to P2;
		Update(Point) from Svr to P1;
		do Game(Svr, P2, P1); (*@\label{line:gamerec}@*)
	}
}
\end{lstlisting}
\captionof{lstlisting}{The \tprotocol{Noughts and Crosses} Protocol}
\label{lst:game}
\end{figure}

We only highlight implementation details that
best illustrate the expressiveness of the generated APIs;
the interested reader should read the full implementation on
GitHub\footnote{
\url{https://github.com/ansonmiu0214/SessionTS-Examples/NoughtsAndCrosses}
}.

\subsection{Game Server}
We set up the WebSocket server as an Express.js \cite{Express} application
on top of the Node.js runtime.
We define our own game logic in a \texttt{Board} class
to keep track of the game state and expose methods 
to query the result.
This custom logic is integrated into the 
\texttt{handleP1Move} and \texttt{handleP2Move}
handlers implemented by the developer, defined to
handle the moves made by \trole{P1} and \trole{P2} respectively.
We illustrate this in \cref{lst:gamesvr}.

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
const handleP1Move = new Implementation.S13({
	Pos: async (move: Point) => {
		// `board` manages game state;
		// `board.P1` registers the move and returns the game result
		const result = await board.P1(move);
		switch (result) {
			case MoveResult.Win: {
				return new Implementation.S15([
					// Send losing result to P2
					[Labels.S15.Lose, [move], new Implementation.S16(
						// Send winning result to P1
						[Labels.S16.Win, [move], new Implementation.Terminal()]
					)]
				]);
			}
			case MoveResult.Draw: { ... }
			case MoveResult.Continue: {
				return new Implementation.S15(
					// Notify both players and proceed
					// with next round using `handleP2Move`
					[Labels.S15.Update, [move], new Implementation.S18(
						[Labels.S18.Update, [move], handleP2Move]
					)]
				);	
			}		
		}
	}
});

const handleP2Move = ...		// Defined similarly as handleP1Move

new Svr(wss, handleP1Move, (role, reason) => {
	console.log(`${role} cancelled because of ${reason}`);
});
\end{lstlisting}
\captionof{lstlisting}{Implementing \tprotocol{Noughts and Crosses} Game Server using
\codegen}
\label{lst:gamesvr}
\end{figure}

When the server receives a move, it notifies
the game logic to update the game state and return the game
result caused by that move.
The game logic is likely to keep track of move history
using a database; we simulate this with a delay, so 
the game result returned by the game logic is a \texttt{Promise}.
The expressiveness of our generated APIs enable the developer
to define the handlers as \lstonelinejs{async} functions
to use the asynchronous game logic API intuitively -- this is something
prevalent in modern web programming, but not directly addressed in existing
session type implementations for web development \cite{PureScript2019,MVU2020}.

\subsection{Players}
For simplicity, our game uses the same implementation for both
\trole{P1} and \trole{P2}, although they can be different in theory --
the developer could implement \trole{P1} using a GUI and provide
\trole{P2} with a text-based game experience on the browser.

The main implementation detail for players
is to make a move. 
Intuitively, the developer implements a grid and binds a handler
to the \lstonelinejs{'onClick'} event of each vacant cell to send that cell's coordinate 
in a \tmsg{Pos(Point)} message to the game server.
A common source of bugs would be not preventing the user from selecting
a second cell when waiting for the game server's response,
which violates the game rules (and the global protocol).

Our approach of providing \textit{component factories} for send states
in \reactcodegen makes this very intuitive and guarantees communication safety.
First, it gives the developer the flexibility to trigger
the same send action (in this case, \tmsg{Pos(Point)} via multiple
UI elements -- the developer can generate a send action wrapper component 
for each vacant cell on the game board.
Moreover, each generated wrapper component sends a different payload
corresponding to the coordinates of the cell:
our generated APIs support this as the handler supplied
to the send component factory can access the cell's coordinates in the closure.
Finally, the send action is always followed by a transition to the receive
state component, so the user cannot violate channel linearity by selecting
two cells.

We demonstrate how this works in \cref{lst:gamesendfactory}.
The factory function for binding the \tmsg{Pos(Point)} send action
is defined under \lstonelinejs{this.props.Pos}.
For each x-y coordinate on the game board, if the cell is vacant,
we create a \texttt{<SelectPoint>} React component from the
component factory function (which reads ``build a react
component that sends the \tmsg{Pos} message with x-y coordinate
as payload when the user clicks on it''), and we wrap
a \texttt{<td>} table cell (since the game board is rendered as an
HTML table) inside the generated component to bind the click event
to the table cell.

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
// Inside some render() function...
{board.map((row, x) =>
	<tr>
	{row.map((cell, y) => {
		if (cell === Cells.VACANT) {
			const sendPoint = (event: React.MouseEvent) => {
				// Don't refresh the screen
				event.preventDefault();
				return { x, y };
			});
			const SelectPoint = this.pros.Pos('onClick', sendPoint);
			return <SelectPoint><td>{cell}</td></SelectPoint>
		} else {
			// Render nought or cross,
			// but clicking on this cell 	will *not* send anything
			return <td>{cell}</td>
		}		
	})}
	</tr>
)}
\end{lstlisting}
\captionof{lstlisting}{\dots}
\label{lst:gamesendfactory}
\end{figure}

The developer can also leverage the session cancellation handler
to render useful messages to the player.
A different UI can be rendered depending on whether
the server or the opposition has disconnected,
and make \textit{application-specific} interpretations of the cancellation.
For example, if the opposition has disconnected,
the developer can interpret this as a forfeit and
render a winning message to the user.

\subsection{Summary}
We demonstrated how the developer can use the generated APIs
from \codegen to implement a complex multiparty protocol
which features branching, selection and recursion.
We highlighted specific features in the generated APIs for both
server and browser endpoints that allow the developer
to intuitively implement their application logic.
In particular, we observed that the extensions introduced
in \cref{chap:ext} played crucial roles in
improving the usability of the generated APIs.

### 2020-06-12 report/sections/evalconcl/eval/twobuyer.tex
Report: complete chapter 10.1

\section{Routed Multiparty Sessions: \fancyname{Two Buyers}}
\label{section:evaltwobuyer}

Interested readers can find the full implementation on
GitHub\footnote{\url{https://github.com/ansonmiu0214/SessionTS-Examples/TwoBuyer}}.

\begin{itemize}
\item show that the routing aspect is transparent to both sides
\item same flexibilities apply -- we use react context for state management
\end{itemize}

### 2020-06-12 report/sections/appendix/eval.tex
Report: progress on chapter 10

\chapter{Artefacts for Evaluation}
\label{chapter:evalcode}

\section{Implementation for \tprotocol{Noughts and Crosses}}
\label{section:evalcodegame}

% BOARD CLASS

\section{Implementation for \tprotocol{Two Buyer}}
\label{section:evalcodetwobuyer}

\section{Implementation for \tprotocol{ATM}}
\label{section:evalcodeatm}

### 2020-06-12 report/sections/evalconcl/eval/game.tex
Report: progress on chapter 10

@@
 
-We only highlight implementation details that
-best illustrate the expressiveness of the generated APIs;
-the interested reader should read the full implementation on
+We focus on the implementation details that
+best illustrate the expressiveness of our work;
+the interested reader can find the full implementation on
 GitHub\footnote{
 \url{https://github.com/ansonmiu0214/SessionTS-Examples/NoughtsAndCrosses}
-}.
+}, and consult \texttt{README.md} to navigate between the generated code
+and developer implementation.
 
@@
 to keep track of the game state and expose methods 
-to query the result.
+to query the result -- the implementation for \texttt{Board} is included
+in \cref{section:evalcodegame}.
 This custom logic is integrated into the 
@@
 
-\subsection{Players}
+\subsection{Game Players}
 For simplicity, our game uses the same implementation for both
@@
 \end{lstlisting}
-\captionof{lstlisting}{\dots}
+\captionof{lstlisting}{Safely Binding Send Actions to 
+\tprotocol{Noughts and Crosses} Game Board}
 \label{lst:gamesendfactory}
@@
 
-The developer can also leverage the session cancellation handler
-to render useful messages to the player.
-A different UI can be rendered depending on whether
+The session cancellation handler allows the developer
+to render useful messages to the player, since
+a different component can be rendered depending on whether
 the server or the opposition has disconnected,
@@
 In particular, we observed that the extensions introduced
-in \cref{chap:ext} played crucial roles in
-improving the usability of the generated APIs.
+in \cref{chap:ext} play crucial roles in
+improving the usability of the generated APIs when
+compared with existing work on session types for web development 
+\cite{PureScript2019,MVU2020,LINKS}.

### 2020-06-12 report/sections/evalconcl/eval/twobuyer.tex
Report: progress on chapter 10

@@
-\section{Routed Multiparty Sessions: \fancyname{Two Buyers}}
+\section{Routed Multiparty Sessions: \tprotocol{Two Buyers}}
 \label{section:evaltwobuyer}
+
+We implement the \tprotocol{Two Buyer} protocol
+introduced in \cref{section:twobuyer}.
 
@@
 GitHub\footnote{\url{https://github.com/ansonmiu0214/SessionTS-Examples/TwoBuyer}}.
+
+The \newtheory implementation does not affect
+the compatibility of our generated APIs with external libraries.
+The \trole{S}eller endpoint is set up as an Express.js
+application, and both buyers still use the React Context API
+for application state management.
 

### 2020-06-12 report/sections/evalconcl/eval/twobuyer.tex
Report: complete chapter 10.2 sans listings

@@
 introduced in \cref{section:twobuyer}. 
+This protocol cannot be implemented using
+existing proposals \cite{PureScript2019,MVU2020} for integrating
+session types into web development.
+\codegen overcomes the limitations of existing work
+through implementing novel theory of routed multiparty
+session types formalised in \cref{chap:theory}.
 
-Interested readers can find the full implementation on
-GitHub\footnote{\url{https://github.com/ansonmiu0214/SessionTS-Examples/TwoBuyer}}.
+We present the implementation of the \trole{S}eller
+in \cref{lst:evaltwobuyerseller}, 
+and the implementation
+of the peer-to-peer interaction by buyer \trole{A}
+in \cref{lst:evaltwobuyerA}
+The main point to note is that the routing mechanism
+is completely transparent to the developer,
+which shows the elegance of our solution.
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript]
+// two buyer seller
+\end{lstlisting}
+\captionof{lstlisting}{\dots}
+\label{lst:evaltwobuyerseller}
+\end{figure}
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript]
+// two buyer A p2p
+\end{lstlisting}
+\captionof{lstlisting}{\dots}
+\label{lst:evaltwobuyerA}
+\end{figure}
 
@@
 application, and both buyers still use the React Context API
-for application state management.
+for application state management. We show an example of
+the latter in the developer's implementation for
+buyer \trole{A} receiving and keeping track of the server's quote
+using a React Context in \cref{lst:evaltwobuyercontext}.
 
-\begin{itemize}
-\item show that the routing aspect is transparent to both sides
-\item same flexibilities apply -- we use react context for state management
-\end{itemize}
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript]
+// two buyer A context
+\end{lstlisting}
+\captionof{lstlisting}{\dots}
+\label{lst:evaltwobuyercontext}
+\end{figure}
+
+We include the implementations of the remaining EFSM states
+for buyer \trole{A} in \cref{section:evalcodetwobuyer}
+for completeness;
+interested readers can find the full implementations for the other
+endpoints alongside the generated code on
+GitHub\footnote{
+\url{https://github.com/ansonmiu0214/SessionTS-Examples/TwoBuyer}
+}.

### 2020-06-12 report/sections/evalconcl/eval/atm.tex
Report: complete chapter 10 sans code listings for 10.2 and 10.3

@@
-\section{Delegated Sessions: \fancyname{ATM}}
+\section{Session Delegation: \fancyname{ATM}}
 \label{section:evalatm}
 
-Interested readers can find the full implementation on
-GitHub\footnote{\url{https://github.com/ansonmiu0214/SessionTS-Examples/ATM}}.
+The developer may define a communication protocol
+in a way that requires implicit \textit{session delegation}
+upon connection: when the browser endpoint connects
+to the server endpoint, the server delegates a private channel
+to carry out the interactions specified by the protocol,
+whilst concurrently accepting new connections.
+We illustrate the limitations of our work with respect
+to supporting session delegation, albeit this is also
+a limiting factor in similar existing work 
+\cite{PureScript2019,MVU2020}.
 
-\begin{itemize}
-\item Show that delegation is not supported
-even if the protocol ``allows''
-\end{itemize}
+Consider an \tprotocol{ATM} protocol involving a 
+\trole{Bank} server endpoint and 
+a \trole{Client} browser endpoint,
+as presented in \cref{lst:atm}.
+This is a recursive protocol, where the \trole{Client}
+can choose to \tmsg{WITHDRAW} or \tmsg{DEPOSIT} some money
+into their account, or \tmsg{QUIT}.
+For \tmsg{WITHDRAW} and \tmsg{DEPOSIT} operations,
+the \trole{Bank} will respond with the updated \tmsg{BALANCE},
+or a \tmsg{FAIL}ure message if an operation fails (specifically,
+\tmsg{WITHDRAW}ing more money than available).
 
@@
 	choice at Client {
-		// The string contains the client ID
+		// The string contains the client's bank account ID
 		WITHDRAW(string, number) from Client to Bank;
@@
 	} or {
-		BYE(string) from Client to Bank;
+		QUIT(string) from Client to Bank;
 		BYE() 			 from Bank to Client;	
@@
 \end{figure}
+
+An important detail is that, for any operation,
+the \trole{Client} provides their bank account ID
+in the message. By specifying the communication protocol
+in this way, the developer intends to allow the
+\trole{Bank} to manage the operations of multiple
+\trole{Client}s concurrently.
+More precisely, the developer makes an assumption
+that the interactions specified in the protocol
+are delegated to a private channel upon connection.
+
+However, the runtime generated by \nodecodegen
+does not make this assumption, so we do not get
+the behaviour intended by the developer. If a second
+\trole{Client} tries to connect to the \trole{Bank},
+the connection request will be unsuccessful because
+the role is already occupied.
+Interested readers can validate this
+behaviour by running the full implementation available
+on GitHub\footnote{
+\url{https://github.com/ansonmiu0214/SessionTS-Examples/ATM}
+}.
+
+At the same time, the runtime generated by \nodecodegen
+can be adapted to ``support'' \textit{binary} session delegation.
+The generated runtime includes
+an \trole{Svr} class used to listen for
+and manage incoming connections, and a
+private \texttt{Session} class to execute the EFSM.
+When the instance of the \trole{Svr} class
+is no longer waiting for participants,
+it delegates to a fresh instance of the \texttt{Session} class
+to execute the protocol.
+To support delegation, we can simply modify this piece of 
+logic to also create a fresh instance of the \trole{Svr}
+class to concurrently manage new connections, whilst the
+instance of the \texttt{Session} class is bound to the
+participants that have already joined.
+We show this workaround in \cref{lst:atmdelegation} -- 
+specifically, on \cref{line:atmdelegation}.
+
+\begin{figure}[!h]
+\begin{lstlisting}[language=javascript]
+// ATM delegation workaround (*@\label{line:atmdelegation}@*)
+\end{lstlisting}
+\captionof{lstlisting}{Workaround for Binary Session
+Delegation in \tprotocol{ATM} protocol}
+\label{lst:atmdelegation}
+\end{figure}
+
+This workaround does not generalise well for
+\textit{multiparty} session delegation. 
+Suppose we want to do the same for the 
+\tprotocol{Noughts and Crosses} protocol discussed
+in \cref{section:evalcodegame}.
+Assume an user connects as \trole{P1} and the game server
+is still waiting for \trole{P2}. If another user attempts
+to connect as \trole{P1}, connection will be unsuccessful
+as the first game hasn't began yet -- this workaround will
+only begin accepting connections again \textit{once}
+the protocol begins executing.
+Formally supporting session delegation will also require
+extensions to the Scribble language, since we should not
+make the assumption that all protocols require
+session delegation upon connection.
+
+This would be an interesting area of future work and
+enable \codegen to support a wider range of protocols
+for webservices; we discuss further in \cref{section:future}.

### 2020-06-12 report/sections/evalconcl/eval/twobuyer.tex
Report: complete chapter 10 sans code listings for 10.2 and 10.3

@@
 The \trole{S}eller endpoint is set up as an Express.js
-application, and both buyers still use the React Context API
+application, and both buyers use the \textit{React Context API}
 for application state management. We show an example of

### 2020-06-12 report/sections/introduction/motivation.tex
Report: fix bad subfig numbering
Report: address some NY comments
Report: complete chapter 10.1
Report: complete chapter 10 sans code listings for 10.2 and 10.3

@@
 \section{Motivation}
+\label{section:intro}
 
-\paragraph{Note: from PLACES}
+\subsubsection{TODO}
+\begin{itemize}
+\item Break down into paragraphs for theory, implementation
+\item Diagrams similar to FZ
+\item Aim of introductio: motivate what is the problem 
+(session types fix web, PureScript and LINKS have two problems: 
+implementation -- not using mainstream stuff, 
+theory -- websocket transport restricts protocols)
+\end{itemize}
+
+\subsubsection{IDEA (similar to LinearDecomp structure)}
+\begin{itemize}
+\item importance of communication safety in webapps
+\item MPST framework for webapps for communication safety
+\item existing work achieves this
+\item problem 1 -- not using mainstream tools, not intuitively supporting modern practices (async logic), not a natural way for web development (event driven), limits usability, limits the benefits it brings to developers
+\item problem 2 -- rely on websocket transport (good, straightforward to map to session type implementation), which is ok, but limits the types of protocols we can support (e.g.
+TwoBuyer, can be session typed but violates websocket topology
+\item part 1 addresses problem 1 (same communication safety guarantees but targeting TS/Node/React)
+\item part 2 addresses problem 2 (RoutedMPST expresses arbitrary topology over server-centric network topology, not limitied by network topology)
+\end{itemize}
+
+% BREAK DOWN PARAGRAPHS
+
+% INTRODUCTION DIAGRAM
+
+% REFER TO FANGYI
 

### 2020-06-13 report/sections/introduction/motivation.tex
Report: note down new introduction

--- 2020-06-10 report/sections/introduction/motivation.tex
@@
+\subsubsection{New Structure (similar to ECOOP2017):}
+\begin{itemize}
+
+\item \textbf{Section 1.1:}
+
+Interactive webapps over WebSockets, harder to guarantee communication safety
+
+\item
+Example protocol similar to travel agency  -- motivate that it is difficult to manually verify communication safety
+
+\begin{itemize}
+\item traveller enquires availability, flight server replies yes/no then traveller confirms/denies.
+\item The flight server will reserve the seat until the traveller denies.
+\item Source of hidden bugs -- traveller may enquire twice (e.g. click twice), but server processes one enquiry, so one seat is held up for no reason
+\end{itemize}
+
+\item MPST framework can be applied for webapps.
+This is solved by existing work PureScript and MVU2020.
+
+\item \textbf{Problem1}:
+
+existing work provides a workflow for session typed web development, but requires the developer to compromise by adapting to programming paradigms and practices not conventional for modern web programming.
+E.g.
+functional paradigm, sequential UI framework, PureScript and LINKS not interoperable with the ever-growing universe of JavaScript libraries.
+These compromises limit the usability of MPST framework
+for web development.
+
+\item \textbf{PRoblem2}:
+
+existing work based on WebSocket transport, which
+is simple to reason about. But WebSocket transport enforces a server-centric
+network configuration, so existing work isn't expressive enough for
+more general protoocls. Suppose traveler contacts agency first to get a quote and then compare (similar to 2-buyer). Existing work cannot implement this.
+
+\item \textbf{Section 1.2: }
+
+Our objective is to present a novel variant of MPST framework
+that is compatible / supports the flexibility of modern web programming.
+We address the two problems in two parts:
+
+\item We solve problem 1 in part 1:
+
+SessionTS is API generation workflow based on MPST, which
+provides developers with same communication safety guarantees in a way that is idiomatic for modern web programming.
+We use industry tools (TS/Node/REact), compatible with async logic, error handling specific to web (e.g.
+client disconnection)
+
+\item We solve problem 2 in part 2:
+
+develop novel theory for routed MPST which lets us implement non-server-centric commuication structures over a server-centric network topology, prove
+that our theory preserves properties like communication safety, deadlock freedom, etc.
+
+\item Section 1.3: diagram similar to FZ for outline
+
+\end{itemize}
+
+% Aim of introductio: motivate what is the problem 
+%(session types fix web, PureScript and LINKS have two problems: 
+%implementation -- not using mainstream stuff, 
+%theory -- websocket transport restricts protocols)
+
 \section{Motivation}
-
-\paragraph{Note: from PLACES}
-
-% Rise of distributed programs -> main challenges
-% How to ensure correctness in general: type system, data types
-% Type discipline for concurrent programs - behavioural types (--> session types)
-% Mainstream example of distirbuted programs: web services (microservice architecture), interactive web application; 
-% Objective:
+\label{section:intro}
 

### 2020-06-13 report/sections/general/theory.tex
Report: section notes
Report: notes on part 2
Report: complete chapter 9
Report: start on chapter 8

@@
-\chapter{Theory: Routed Multiparty Session Types}
+\chapter{\newtheory: A Theory of
+Routed Multiparty Session Types}
+\label{chap:theory}
+
+In this chapter, we present
+\newtheory, an extension of the
+canonical multiparty session type (MPST)
+theory with a message routing mechanism.
+We build upon the variant of MPST theory introduced
+in \cite{LessIsMore}
+(which we have also discussed in\cref{subsection:bgmpst}), 
+and extend it
+with constructs for communication actions
+that are routed through a participant.
+We introduce the extended syntax for types in 
+\cref{section:syntax}, and give semantics of our
+syntax extensions in \cref{section:lts}.
+
+\newtheory provides the theoretical basis
+for allowing developers to implement
+protocols with browser-to-browser interactions
+(such as the \tprotocol{Two Buyer} protocol
+introduced in \cref{section:twobuyer})
+over a server-centric network topology
+which uses WebSocket transport.
+We define a router-parameterised encoding
+for MPST into \newtheory in \cref{section:encoding}
+and prove the preservation of well-formedness
+and communication.
+We proceed to extend \codegen to implement \newtheory
+in \cref{chap:impl}.
 
 \section{Syntax}
+\label{section:syntax}
 
 \subsection{Global Types}
+As motivated by \ref{section:routerchallenges},
+we need to handle \textit{routed communication} separately
+from normal send and receive actions.
+We define the syntax of global types for \newtheory 
+in \cref{fig:newsyntaxglobal}, and explain the 
+new construct for expressing routed communication.
+
+\begin{figure}[!h]
+\doublespacing
+\[
+\begin{array}{rlr}
+G ::= & \dots & \text{Global Types} \\
+\mid & \groute{p}{q}{s}{l_i: G_i}{i \in I}
+& \text{Routed Communication through $\role{s}$} \\
+\end{array}
+\]
+\singlespacing
+\captionof{figure}{Global Types in \newtheory}
+\label{fig:newsyntaxglobal}
+\end{figure}
+
+The new construct $\groute{p}{q}{s}{l_i: G_i}{i \in I}$ reads, 
+\textit{
+``the communication $\gcomm{p}{q}{l_i: G_i}{i \in I}$
+between $\role{p}$ and $\role{q}$
+is routed by $\role{s}$''
+}.
+We refer to $\role{s}$ hereafter as the \textit{router}.
+$\role{s}$ ranges over the set of roles 
+$\role{p},\role{q},\role{r},\role{s},\dots$,
+but we use $\role{s}$ by convention as the server endpoint
+is usually the router.
+We also extend the definition of the participants function,
+$\pt{G}$, in \cref{def:newpt}.
+
+\begin{definition}[Participants]
+\textit{The participants function 
+represents the set of roles involved in
+the communication specified by $G$.}
+
+\doublespacing
+\[
+\begin{array}{>{\displaystyle}rc>{\displaystyle}l}
+
+\pt{\dots} & = & \dots \\
+
+\pt{\groute{p}{q}{s}{l_i: G_i}{i \in I}} & = & 
+\left\{ \role{p}, \role{q}, \role{s} \right\} 
+\cup \bigcup_{i \in I}\pt{G_i} \\
+
+\end{array}
+\]
+\singlespacing
+\label{def:newpt}
+\end{definition}
 
 \subsection{Local Types}
+Local types are obtained from global types via
+\textit{projection}, so we also need new
+constructs to express routed communication
+from the \textit{local} perspective of individual
+endpoints.
+We explain the extensions to
+projection in \cref{subsection:newprojection}.
 
-\section{Labelled Transition System (LTS) Semantics}
+We define the syntax of local types for \newtheory 
+in \cref{fig:newsyntaxlocal}, and walk through
+the new constructs below.
+
+\begin{figure}[!h]
+\doublespacing
+\[
+\begin{array}{rlr}
+T ::= & \dots & \text{Local Types} \\
+\mid & \router{p}{q}{l_i: T_i}{i \in I}
+& \text{Routing communication from $\role{p}$ to $\role{q}$} \\
+\mid & \tselproxy{p}{q}{l_i: T_i}{i \in I}
+& \text{Selecting from $\role{p}$ via proxy $\role{q}$} \\
+\mid & \tbraproxy{p}{q}{l_i: T_i}{i \in I}
+& \text{Receiving from $\role{p}$ via proxy $\role{q}$} \\
+\end{array}
+\]
+\singlespacing
+\captionof{figure}{Global Types in \newtheory}
+\label{fig:newsyntaxlocal}
+\end{figure}
+
+\begin{itemize}
+
+\item \textbf{Routed communication:}
+
+\dots
+
+\item \textbf{Proxy selection:}
+
+\dots
+
+\item \textbf{Proxy receive:}
+
+\dots
+
+\end{itemize}
+
+% routing construct
+% proxy send and receive
+
+\subsection{Projection}
+\label{subsection:newprojection}
+
+Projection
+
+Full Merge
+
+% projection onto local types
+% full merge
+
+\subsection{Well-Formedness}
+Recall that well-formedness is a predicate defined 
+\textit{solely} on the global type
+in canonical MPST theory: a global type is well-formed 
+if a projection xists for all its participants.
+
+\[
+\dots
+\]
+
+In \newtheory, we require that there is \textit{exactly
+one} router, say $\role{s}$, in the global type, so
+we need to express that a global type to be
+well-formed \textit{with respect to the role $\role{s}$
+acting as the router}.
+
+% act as server role
+
+\begin{definition}[Centroid]
+\textit{Let} $\centroid{G}{s}$ \textit{denote \dots}
+
+\end{definition}
+
+% well-formedness
+
+\section{Labelled Transition System (LTS) Semanics}
+\label{section:lts}
+
+\begin{itemize}
+\item extend action syntax
+\end{itemize}
+
+\subsection{LTS over Global Types}
+
+\subsection{LTS over Local Types}
+
+\subsection{LTS over Collections of Local Types}
 
@@
 \section{Router-Parameterised Encoding}
+\label{section:encoding}
+
+\begin{itemize}
+\item define encoding on global types
+\item define encoding on local types
+\item define encoding on LTS actions
+\item motivate with twobuyer example
+\end{itemize}
+
+\subsection{Correspondence between Global and Local Type Encodings}
+\begin{itemize}
+\item prove proposition
+\item state corollary for subsequent proof
+\end{itemize}
 

### 2020-06-14 report/sections/general/theory.tex
Report: progress on chapter 8

--- 2020-05-25 report/sections/general/theory.tex
@@
-\chapter{Theory: Routed Multiparty Session Types}
+\chapter{\newtheory: A Theory of
+Routed Multiparty Session Types}
+\label{chap:theory}
 
-\section{Syntax}
+In this chapter, we present
+\newtheory, an extension of the
+canonical multiparty session type (MPST)
+theory with a message routing mechanism.
+We build upon the variant of MPST theory introduced
+in \cite{LessIsMore}
+(which we have also discussed in \cref{subsection:bgmpst}), 
+and extend it
+with constructs for communication actions
+that are routed through a participant.
+We introduce the extended syntax for types in 
+\cref{section:syntax}, and give semantics of our
+syntax extensions in \cref{section:lts}.
 
-\subsection{Global Types}
+\newtheory provide the theoretical basis
+for allowing developers to implement
+protocols with browser-to-browser interactions
+(such as the \tprotocol{Two Buyer} protocol
+introduced in \cref{section:twobuyer})
+over a server-centric network topology
+which uses WebSocket transport.
+We define a router-parameterised encoding
+for MPST into \newtheory in \cref{section:encoding}
+and prove the preservation of well-formedness
+and communication.
+We proceed to extend \codegen to implement \newtheory
+in \cref{chap:impl}.
 
-\subsection{Local Types}
+\input{sections/general/theory/syntax}
 
-\section{Labelled Transition System (LTS) Semantics}
+\input{sections/general/theory/lts}
 
-\subsection{Trace Equivalence}
-
-\subsection{Deadlock Freedom}
-
-\section{Router-Parameterised Encoding}
-
-\subsection{Preserving Well-formedness}
-
-\subsection{Preserving Communication}
+\input{sections/general/theory/encoding}

### 2020-06-14 report/sections/general/theory/encoding.tex
Report: progress on chapter 8

\section{Router-Parameterised Encoding}
\label{section:encoding}

\begin{itemize}
\item define encoding on global types
\item define encoding on local types
\item define encoding on LTS actions
\item motivate with twobuyer example
\end{itemize}

\subsection{Correspondence between Global and Local Type Encodings}
\begin{itemize}
\item prove proposition
\item state corollary for subsequent proof
\end{itemize}

\subsection{Preserving Well-formedness}

\subsection{Preserving Communication}

### 2020-06-14 report/sections/general/theory/lts.tex
Report: progress on chapter 8

\section{Labelled Transition System (LTS) Semantics}
\label{section:lts}

We define labelled transition system (LTS) semantics
on global types (\cref{subsection:newltsglobal}) 
and local types (\cref{subsection:newltslocal})
for \newtheory,
building upon the work of Deni\'elou and Yoshida in
\cite{characterisation}.
We show the soundness and completeness of projection
with respect to the LTSs through proving the
\textit{trace equivalence} of
a global type and the collection of local types projected
from the global type (\cref{subsection:newtraceeq}).
We then use this result to
conclude that \newtheory provide the same
communication safety guarantees from canonical
MPST theory for well-formed global types, 
namely deadlock freedom (\cref{subsection:newdeadlockfreedom}).

First, we extend the label in the LTS, as shown in 
\cref{fig:newlts}, to distinguish
the \textit{direct} sending (and reception) of a message
from the sending (and reception) of a message
\textit{via} an intermediate routing endpoint.
Labels range over $l, l', \dots$

\begin{figure}[!h]
\doublespacing
\[
\begin{array}{rlr}
l ::= & & \text{Labels} \\
	\mid & \aout{p}{q}{j} & 
		\text{Direct Send} \\
	\mid & \ain{p}{q}{j} & 
		\text{Direct Receive} \\
	\mid & \via{s}{\aout{p}{q}{j}} & 
		\text{Routed Send} \\
	\mid & \via{s}{\ain{p}{q}{j}} & 
		\text{Routed Receive} \\
\end{array}
\]
\singlespacing
\captionof{figure}{LTS Labels in \newtheory}
\label{fig:newlts}
\end{figure}

\begin{itemize}
\item \textbf{Routed send:}

The label $\via{s}{\aout{p}{q}{j}}$ represents the
\textit{sending} (performed by $\role{p}$)
of a message labelled $j$ to $\role{q}$ through
the intermediate router $\role{s}$.

\item \textbf{Routed receive:}

The label $\via{s}{\ain{p}{q}{j}}$ represents the
\textit{reception} (initiated by $\role{q}$) 
of a message labelled $j$
send from $\role{p}$ through
the intermediate router $\role{s}$.
\end{itemize}

Labels represent communication actions, so we refer
to $l$ as labels and actions interchangeably,
as is the case in the literature.

Building upon \cite{characterisation},
the \textit{subject} of a label is the role
that initiates the action. 
Intuitively, the actions for routed send and routed
receive are still initiated by the original sender and
recipient respectively;
we extend the definition
of subjects in \cref{def:newsubj} to reflect this.

\begin{definition}[Subject]
\textit{The subject of a LTS label, or $\subj{l}$, is defined as:}

\doublespacing
\[
\begin{array}{c}
\subj{\via{s}{\aout{p}{q}{j}}} = 
	\subj{\aout{p}{q}{j}} = \mrole{p} \\
\subj{\via{s}{\ain{p}{q}{j}}} = 
	\subj{\ain{p}{q}{j}} = \mrole{q} \\
\end{array}
\]
\singlespacing
\label{def:newsubj}
\end{definition}

\subsection{LTS over Global Types}
\label{subsection:newltsglobal}

The LTS semantics presented in \cite{characterisation}
models \textit{asynchronous communication},
which is consistent with our proposal.
In order to define LTS over global types for
asynchronous communication, we need to
represent intermediate states (i.e. messages in transit)
within the grammar of global types.

\cite{characterisation} added the construct
{$\gtrans{p}{q}{j}{l_i: G_i}{i \in I}$}
to represent that the message $l_j$ has been
sent by $\role{p}$ but not yet received by $\role{q}$.

We add a similar construct
{$\gtransroute{p}{q}{s}{j}{l_i: G_i}{i \in I}$}
to represent that the message $l_j$ has
been sent from $\role{p}$ to the router $\role{s}$
but not yet routed to $\role{q}$.

Because routed communication is treated differently
from normal send and receive actions, the notion
of asynchrony differs between the two types of communication
too. This definition allows us to extend
the LTS semantics defined in \cite{characterisation}
more naturally.

We define the LTS semantics 
over global types,
defined by $\treduce{G}{G'}{l}$,
in \cref{fig:newglobal}.

\begin{figure}[!h]

\begin{prooftree}
\AxiomC{}
\RightLabel{\rulename{Gr6}}
\UnaryInfC{$
\treducelong
	{\groute{p}{q}{s}{l_i: G_i}{i \in I}}
	{\gtransroute{p}{q}{s}{j}{l_i: G_i}{i \in I}}
	{\via{s}{\aout{p}{q}{j}}}
$}
\end{prooftree}

\begin{prooftree}
\AxiomC{}
\RightLabel{\rulename{Gr7}}
\UnaryInfC{$
\treducelong
	{\gtransroute{p}{q}{s}{j}{l_i: G_i}{i \in I}}
	{G_j}
	{\via{s}{\ain{p}{q}{j}}}
$}
\end{prooftree}

\begin{prooftree}
\AxiomC{$\forall i \in I. ~ \treduce{G_i}{G'_i}{l}$}
\AxiomC{$\subj{l} \notin \{\mrole{p}, \mrole{q}\}$}
\RightLabel{\rulename{Gr8}}
\BinaryInfC{$
\treducelong
	{\groute{p}{q}{s}{l_i: G_i}{i \in I}}
	{\groute{p}{q}{s}{l_i: G'_i}{i \in I}}
	{l}
$}
\end{prooftree}

\begin{prooftree}
\AxiomC{$\treduce{G_j}{G'_j}{l}$}
\AxiomC{$\subj{l} \neq \mrole{q}$}
\AxiomC{$\forall i \in I \setminus \{ j \}. ~ G'_i = G_i$}
\RightLabel{\rulename{Gr9}}
\TrinaryInfC{$
\treducelong
	{\gtransroute{p}{q}{s}{j}{l_i: G_i}{i \in I}}
	{\gtransroute{p}{q}{s}{j}{l_i: G'_i}{i \in I}}
	{l}
$}
\end{prooftree}

\captionof{figure}{LTS Semantics over Global Types in \newtheory}
\label{fig:newglobal}
\end{figure}

\begin{itemize}

\item \rulename{Gr6}: 
TODO
%The emission of a routed message $j$
%results in the message

\item \rulename{Gr7}:
TODO

\item \rulename{Gr8}:
TODO

\item \rulename{Gr9}:
TODO

\end{itemize}

\subsection{LTS over Local Types}
\label{subsection:newltslocal}

We define the LTS semantics 
over local types,
defined by $\treduce{T}{T'}{l}$,
in \cref{fig:newlocal}.

%TODO: LR1-3?

\begin{figure}[!h]

\begin{prooftree}
\AxiomC{}
\RightLabel{\rulename{Lr4}}
\UnaryInfC{$
\treducelong
	{\tselproxy{q}{s}{l_i: T_i}{i \in I}}
	{T_j}
	{\via{s}{\aout{p}{q}{j}}}
$}
\end{prooftree}

\begin{prooftree}
\AxiomC{}
\RightLabel{\rulename{Lr5}}
\UnaryInfC{$
\treducelong
	{\tbraproxy{q}{s}{l_i: T_i}{i \in I}}
	{T_j}
	{\via{s}{\ain{q}{p}{j}}}
$}
\end{prooftree}

\begin{prooftree}
\AxiomC{}
\RightLabel{\rulename{Lr6}}
\UnaryInfC{$
\treducelong
	{\router{p}{q}{l_i: T_i}{i \in I}}
	{\routertrans{p}{q}{j}{l_i: T_i}{i \in I}}
	{\via{s}{\aout{p}{q}{j}}}
$}
\end{prooftree}

\begin{prooftree}
\AxiomC{}
\RightLabel{\rulename{Lr7}}
\UnaryInfC{$
\treducelong
	{\routertrans{p}{q}{j}{l_i: T_i}{i \in I}}
	{T_j}
	{\via{s}{\ain{p}{q}{j}}}
$}
\end{prooftree}

\begin{prooftree}
\AxiomC{$\forall i \in I. ~ \treduce{T_i}{T'_i}{l}$}
\AxiomC{$\subj{l} \notin \{\mrole{p}, \mrole{q}\}$}
\RightLabel{\rulename{Lr8}}
\BinaryInfC{$
\treducelong
	{\router{p}{q}{l_i: T_i}{i \in I}}
	{\router{p}{q}{l_i: T'_i}{i \in I}}
	{l}
$}
\end{prooftree}

\begin{prooftree}
\AxiomC{$\treduce{T_j}{T'_j}{l}$}
\AxiomC{$\subj{l} \neq \mrole{q}$}
\AxiomC{$\forall i \in I \setminus \{ j \}. ~ T'_i = T_i$}
\RightLabel{\rulename{Lr9}}
\TrinaryInfC{$
\treducelong
	{\routertrans{p}{q}{j}{l_i: T_i}{i \in I}}
	{\routertrans{p}{q}{j}{l_i: T'_i}{i \in I}}
	{l}
$}
\end{prooftree}

\begin{prooftree}
\AxiomC{$l = \via{s}{\cdot}$}
\AxiomC{$\subj{l} \neq \mrole{q}$}
\AxiomC{$\forall i \in I. ~ \treduce{T_i}{T'_i}{l}$}
\RightLabel{\rulename{Lr10}}
\TrinaryInfC{$
\treducelong
	{\tsel{q}{l_i: T_i}{i \in I}}
	{\tsel{q}{l_i: T'_i}{i \in I}}
	{l}
$}
\end{prooftree}

\begin{prooftree}
\AxiomC{$l = \via{s}{\cdot}$}
\AxiomC{$\subj{l} \neq \mrole{q}$}
\AxiomC{$\forall i \in I. ~ \treduce{T_i}{T'_i}{l}$}
\RightLabel{\rulename{Lr11}}
\TrinaryInfC{$
\treducelong
	{\tbra{q}{l_i: T_i}{i \in I}}
	{\tbra{q}{l_i: T'_i}{i \in I}}
	{l}
$}
\end{prooftree}

\label{fig:newlocal}
\captionof{figure}{LTS over Local Types in \newtheory}
\end{figure}

We explain rules \rulename{Lr4} and \rulename{Lr5}
from the perspective of role $\role{p}$.

\begin{itemize}

\item \rulename{Lr4}: 
TODO

\item \rulename{Lr5}:
TODO

\end{itemize}

We explain rules \rulename{Lr6}, \rulename{Lr7},
\rulename{Lr10} and \rulename{Lr11}
from the perspective of role $\role{s}$.

\begin{itemize}

\item \rulename{Lr6}:
TODO

\item \rulename{Lr7}:
TODO

\item \rulename{Lr10}:
TODO

\item \rulename{Lr11}:
TODO

\end{itemize}

We explain the remaining inductively defined rules.

\begin{itemize}

\item \rulename{Lr8}:
TODO

\item \rulename{Lr9}:
TODO

\end{itemize}

%\subsection{LTS over Collections of Local Types}
%\label{subsection:newltssystem}

\subsection{Soundness and Completeness w.r.t. Projection}
\label{subsection:newtraceeq}

We work towards proving the soundness and completeness
of our LTS semantics with respect to projection.
Our approach is motivated from \cite{characterisation}:

\begin{enumerate}

\item We first extend the LTS semantics to
a collection of local types (hereafter referred to
as a \textit{configuration} to be consistent with
the literature);

\item Then, we extend the definition of projection to
obtain the configuration of a global type 
(hereafter referred to as the \textit{projected configuration});

\item Finally, we prove the trace equivalence between 
the global type and its projected configuration.

\end{enumerate}

\begin{theorem}[Trace Equivalence]
Let $G$ be a global type with participants 
$\mathcal{P} = \mathrm{\pt{G}}$.
Then $G \approx (\vec T, \vec \epsilon)$.
\end{theorem}

\begin{proof}
\dots
\end{proof}

\subsection{Deadlock Freedom}
\label{subsection:newdeadlockfreedom}

### 2020-06-14 report/sections/general/theory/syntax.tex
Report: progress on chapter 8

\section{Syntax}
\label{section:syntax}

We extend the syntax for global types
in \cref{subsection:newglobal} and
local types in \cref{subsection:newlocal}.
The same conventions introduced in \cref{subsection:bgmpst}
apply here: 
we omit message payload types,
and deterministic send and receive actions
are represented by single selection and
single branching constructs respectively.
With these syntax extensions, we appropriately extend
the definition of \textit{projection} and 
\textit{well-formedness} in 
\cref{subsection:newprojection,subsection:newwf}
respectively.

\subsection{Global Types}
\label{subsection:newglobal}

As motivated by \cref{section:routerchallenges},
we need to handle \textit{routed communication} separately
from normal send and receive actions.
We define the syntax of global types for \newtheory 
in \cref{fig:newsyntaxglobal}, and explain the 
new construct for expressing routed communication.
Global types range over $G, G', G_i, \dots$

\begin{figure}[!h]
\doublespacing
\[
\begin{array}{rlr}
G ::= & \dots & \text{Global Types} \\
\mid & \groute{p}{q}{s}{l_i: G_i}{i \in I}
& \text{Routed Communication through $\role{s}$} \\
\end{array}
\]
\singlespacing
\captionof{figure}{Global Types in \newtheory}
\label{fig:newsyntaxglobal}
\end{figure}

The new construct $\groute{p}{q}{s}{l_i: G_i}{i \in I}$ reads, 
\textit{
``the communication $\gcomm{p}{q}{l_i: G_i}{i \in I}$
is routed by $\role{s}$''
}.
To be more verbose,
$\role{p}$ offers $\role{q}$ a set of choices
$\left\{l_i\right\}_{i \in I}$,
$\role{q}$ sends their choice to
$\role{s}$, so $\role{p}$ receives the
selection $l_i$ made by $\role{q}$ via $\role{s}$,
and the communication system proceeds with continuation $G_i$.
We refer to $\role{s}$ hereafter as the \textit{router}.
$\role{s}$ ranges over the set of roles 
$\role{p},\role{q},\role{r},\role{s},\dots$,
but we use $\role{s}$ by convention as the server endpoint
is usually the router.

We also extend the definition of the participants function,
$\pt{G}$, in \cref{def:newpt}, to formalise that the router 
participates in the routed communication.

\begin{definition}[Participants]
\textit{The set of roles involved in the communication
interactions specified by $G$, written} $\pt{G}$
\textit{, is:}

\doublespacing
\[
\begin{array}{>{\displaystyle}rc>{\displaystyle}l}

\pt{\dots} & = & \dots \\

\pt{\groute{p}{q}{s}{l_i: G_i}{i \in I}} & = & 
\left\{ \mrole{p}, \mrole{q}, \mrole{s} \right\} 
\cup \bigcup_{i \in I}\pt{G_i} \\

\end{array}
\]
\singlespacing
\label{def:newpt}
\end{definition}

\subsection{Local Types}
\label{subsection:newlocal}

Local types are obtained from global types via
\textit{projection}, so we also need new
constructs to express routed communication
from the \textit{local} perspective of individual
endpoints.
We explain the extensions to
projection in \cref{subsection:newprojection}.

We define the syntax of local types for \newtheory 
in \cref{fig:newsyntaxlocal}, and walk through
the new constructs below from the perspective of
some arbitrary role $\role{r}$.
Local types range over $T, T', T_i, \dots$

\begin{figure}[!h]
\doublespacing
\[
\begin{array}{rlr}
T ::= & \dots & \text{Local Types} \\
\mid & \router{p}{q}{l_i: T_i}{i \in I}
& \text{Routing communication from $\role{p}$ to $\role{q}$} \\
\mid & \tselproxy{p}{q}{l_i: T_i}{i \in I}
& \text{Selecting from $\role{p}$ via proxy $\role{q}$} \\
\mid & \tbraproxy{p}{q}{l_i: T_i}{i \in I}
& \text{Receiving from $\role{p}$ via proxy $\role{q}$} \\
\end{array}
\]
\singlespacing
\captionof{figure}{Global Types in \newtheory}
\label{fig:newsyntaxlocal}
\end{figure}

\begin{itemize}

\item \textbf{Routed communication:}

If $\role{r}$ has local type $\router{p}{q}{l_i: T_i}{i \in I}$,
$\role{r}$ is routing the communication between 
$\role{p}$ and $\role{q}$.
$\role{r}$ routes the message $l_i$ sent by $\role{p}$
to the intended recipient $\role{q}$ and proceeds 
with continuation $T_i$.

\item \textbf{Proxy selection:}

If $\role{r}$ has local type $\tselproxy{p}{q}{l_i: T_i}{i \in I}$,
$\role{r}$ is making a selection from the
\textit{intended recipient} $\role{p}$,
but the selection is sent to the intermediate proxy $\role{q}$. 
$\role{r}$ sends their internal choice $l_i$ to $\role{q}$
and proceeds with continuation $T_i$.

\item \textbf{Proxy branching:}

If $\role{r}$ has local type $\tbraproxy{p}{q}{l_i: T_i}{i \in I}$,
$\role{r}$ is offering a choice to the
\textit{intended sender} $\role{p}$,
but the message is received from the intermediate proxy $\role{q}$.
$\role{r}$ receives the external choice $l_i$ from $\role{q}$
and proceeds with continuation $T_i$.

\end{itemize}

As a general comment, the new local types of proxy selection
and proxy branching should ``behave'' as normal selection and
receive types with respect to their intended recipient and sender
respectively. We keep track of the proxy role in the syntax
to allow us to distinguish between routing communications
from normal send and receive interactions.

\subsection{Projection}
\label{subsection:newprojection}

We extend the projection operator to be defined
on routed communication. Given the explanations of the
new routing constructs for both global and local types,
\cref{def:newprojection} should be self-explanatory.

\begin{definition}[Projection]
\textit{The projection of $G$ onto $\role{r}$,
written $\proj{G}{r}$, is defined as:}

\doublespacing
\[
\begin{array}{>{\displaystyle}rc>{\displaystyle}l}

\proj{
\left(\groute{p}{q}{s}{l_i: G_i}{i \in I}\right)
}{r} & = & 
\begin{cases}
\tselproxy{q}{s}{l_i: (\proj{G_i}{r})}{i \in I}
	& \text{if} ~ \mrole{r} = \mrole{p} \\
\tbraproxy{p}{s}{l_i: (\proj{G_i}{r})}{i \in I}
	& \text{if} ~ \mrole{r} = \mrole{q} \\
\router{p}{q}{l_i: (\proj{G_i}{r})}{i \in I}
	& \text{if} ~ \mrole{r} = \mrole{s} \\
\underset{i \in I}{\MERGEOP}\proj{G_i}{r}
	& \text{otherwise} \\
\end{cases}
\\

\end{array}
\]
\singlespacing

\label{def:newprojection}
\end{definition}

As shown in the last (4th) case, 
the same concept of \textit{merging}
applies for routed communication:
when projecting a routed communication
onto a non-participant, the projections of
all continuations must be ``compatible'',
namely they can be merged using the
the \textit{merging operator}, $\mergeop$.

As the merging operator is defined on local
types, we extend the merging operator to be defined
on the extended syntax in \cref{def:newmerge}.
Recall that proxy selection and proxy
branching behave in the same way as their ``non-proxified''
counterparts -- the merging operator reflects this
similarity.

\begin{definition}[Merging Operator]
\textit{The merging operator $\mergeop$ on local types is defined as:}

\doublespacing
\[
\begin{array}{rcl}
\tmerge
{(\tselproxy{p}{q}{l_i: T_i}{i \in I})}
{(\tselproxy{p}{q}{l_i: T_i}{i \in I})}
	& = & \tselproxy{p}{q}{l_i: T_i}{i \in I} \\

\tmerge
{(\tbraproxy{p}{q}{l_i: T_i}{i \in I})}
{(\tbraproxy{p}{q}{l_j: T'_j}{j \in J})}
	& = & \tbraproxy{p}{q}{l_k: T''_k}{k \in I \cup J} \\
\text{where} & & T''_k = \begin{cases}
T_k & \text{if} ~ k \in I \setminus J \\
T'_k & \text{if} ~ k \in J \setminus I \\
\tmerge{T_k}{T'_k} & \text{if} ~ k \in I \cap J \\
\end{cases} \\
\end{array}
\]
\singlespacing

\label{def:newmerge}
\end{definition}

\subsection{Well-formedness}
\label{subsection:newwf}

Recall that well-formedness is a predicate defined 
\textit{solely} on the global type
in canonical MPST theory: a global type $G$ is well-formed
if a projection exists for all its participants.

\[
\wf{G} \iff 
\forall \mrole{p} \in \pt{G}. ~ \proj{G}{p} \text{ exists}
\]

In \newtheory, we require that there is \textit{exactly
one} router, say $\role{s}$, in the global type, so
we need to express that a global type to be
well-formed \textit{with respect to the role $\role{s}$
acting as the router}.

We need to define the characteristics that $\role{s}$
must display in $G$ to prove that it is the \textit{only}
router. 
We formalise this as an \textit{inductively} defined relation,
$\centroid{G}{s}$, which reads \textit{``$\role{s}$ is the
centroid in $G$''}.
The intuition is that $\role{s}$ is at the centre
of all communication interactions.
We define what it means to be a \textit{centroid} in
\cref{def:centroid}.

% characteristics of router as centroid
\begin{definition}[Centroid]
\textit{Let} $\centroid{G}{s}$ \textit{denote \dots}

\begin{prooftree}
\AxiomC{}
\RightLabel{\rulename{$\centroidop$-End}}
\UnaryInfC{$\centroid{\tend}{s}$}
\end{prooftree}

\begin{prooftree}
\AxiomC{}
\RightLabel{\rulename{$\centroidop$-RecVar}}
\UnaryInfC{$\centroid{\trecvar}{s}$}
\end{prooftree}

\begin{prooftree}
\AxiomC{$\centroid{G}{s}$}
\RightLabel{\rulename{$\centroidop$-Rec}}
\UnaryInfC{$\centroid{\trec{G}}{s}$}
\end{prooftree}

\begin{prooftree}
\AxiomC{$\mrole{s} \in \left\{ \mrole{p}, \mrole{q} \right\}$}
\AxiomC{$\forall i \in I. ~ \centroid{G_i}{s}$}
\RightLabel{\rulename{$\centroidop$-Comm}}
\BinaryInfC{$\centroid{\gcomm{p}{q}{l_i: G_i}{i \in I}}{s}$}
\end{prooftree}

\begin{prooftree}
\AxiomC{$\mrole{r} = \mrole{s}$}
\AxiomC{$\forall i \in I. ~ \centroid{G_i}{s}$}
\RightLabel{\rulename{$\centroidop$-RouteComm}}
\BinaryInfC{$\centroid{\groute{p}{q}{r}{l_i: G_i}{i \in I}}{s}$}
\end{prooftree}

\label{def:centroid}
\end{definition}

\begin{itemize}
\item 
\rulename{$\centroidop$-End}, 
\rulename{$\centroidop$-RecVar}, 
\rulename{$\centroidop$-Rec}:
Trivial.

\item 
\rulename{$\centroidop$-Comm}:
For normal communication, $\role{s}$ must
be a participant and the centroid of all continuations.

\item 
\rulename{$\centroidop$-RoutedComm}:
For routed communication, $\role{s}$ must
be the router and be the centroid of all continuations.

\end{itemize}

Now we are in a position to formalise the definition
of well-formedness in \newtheory. We present this in
\cref{def:newwf}.

\begin{definition}[Well-formedness]
\textit{Let} $\wfnew{G}{s}$ 
\textit{denote that the global type} $G$
\textit{is well-formed with respect to the router} $\role{s}$.

\[
\wfnew{G}{s} \iff 
(\forall \role{p} \in \pt{G}. ~ \proj{G}{p} \text{ exists})
\wedge
\centroid{G}{s}
\]

\label{def:newwf}
\end{definition}

### 2020-06-14 report/sections/general/theory/lts.tex
Report: progress on chapter 8

@@
 We define labelled transition system (LTS) semantics
-on global types (\cref{subsection:newltsglobal}) 
+over global types (\cref{subsection:newltsglobal}) 
 and local types (\cref{subsection:newltslocal})
@@
 Labels range over $l, l', \dots$
+We highlight and explain the new labels.
 
@@
 	\mid & \via{s}{\aout{p}{q}{j}} & 
-		\text{Routed Send} \\
+		\text{\hl{Routed Send}} \\
 	\mid & \via{s}{\ain{p}{q}{j}} & 
-		\text{Routed Receive} \\
+		\text{\hl{Routed Receive}} \\
 \end{array}
@@
 \begin{definition}[Subject]
-\textit{The subject of a LTS label, or $\subj{l}$, is defined as:}
+The subject of a LTS label, or $\subj{l}$, is defined as:
 
@@
 
-\subsection{LTS over Global Types}
+\subsection{LTS Semantics over Global Types}
 \label{subsection:newltsglobal}
@@
 in \cref{fig:newglobal}.
+We highlight and explain the new rules.
 
@@
 \AxiomC{}
-\RightLabel{\rulename{Gr6}}
+\RightLabel{\rulename{Gr1}}
+\UnaryInfC{$
+\treducelong
+	{\gcomm{p}{q}{l_i: G_i}{i \in I}}
+	{\gtrans{p}{q}{j}{l_i: G_i}{i \in I}}
+	{\aout{p}{q}{j}}
+$}
+\end{prooftree}
+
+\begin{prooftree}
+\AxiomC{}
+\RightLabel{\rulename{Gr2}}
+\UnaryInfC{$
+\treducelong
+	{\gtrans{p}{q}{j}{l_i: G_i}{i \in I}}
+	{G_j}
+	{\ain{p}{q}{j}}
+$}
+\end{prooftree}
+
+\begin{prooftree}
+\AxiomC{$
+\treduce
+	{G[\trec{G} / \trecvar]}
+	{G'}
+	{l}
+$}
+\RightLabel{\rulename{Gr3}}
+\UnaryInfC{$
+\treduce
+	{\trec{G}}
+	{G'}
+	{l}
+$}
+\end{prooftree}
+
+\begin{prooftree}
+\AxiomC{$\forall i \in I. ~ \treduce{G_i}{G'_i}{l}$}
+\AxiomC{$\subj{l} \notin \{\mrole{p}, \mrole{q}\}$}
+\RightLabel{\rulename{Gr4}}
+\BinaryInfC{$
+\treducelong
+	{\gcomm{p}{q}{l_i: G_i}{i \in I}}
+	{\gcomm{p}{q}{l_i: G'_i}{i \in I}}
+	{l}
+$}
+\end{prooftree}
+
+\begin{prooftree}
+\AxiomC{$\treduce{G_j}{G'_j}{l}$}
+\AxiomC{$\subj{l} \neq \mrole{q}$}
+\AxiomC{$\forall i \in I \setminus \{ j \}. ~ G'_i = G_i$}
+\RightLabel{\rulename{Gr5}}
+\TrinaryInfC{$
+\treducelong
+	{\gtrans{p}{q}{j}{l_i: G_i}{i \in I}}
+	{\gtrans{p}{q}{j}{l_i: G'_i}{i \in I}}
+	{l}
+$}
+\end{prooftree}
+
+\begin{prooftree}
+\AxiomC{}
+\RightLabel{\hlrulename{Gr6}}
 \UnaryInfC{$
@@
 \AxiomC{}
-\RightLabel{\rulename{Gr7}}
+\RightLabel{\hlrulename{Gr7}}
 \UnaryInfC{$
@@
 \AxiomC{$\subj{l} \notin \{\mrole{p}, \mrole{q}\}$}
-\RightLabel{\rulename{Gr8}}
+\RightLabel{\hlrulename{Gr8}}
 \BinaryInfC{$
@@
 \AxiomC{$\forall i \in I \setminus \{ j \}. ~ G'_i = G_i$}
-\RightLabel{\rulename{Gr9}}
+\RightLabel{\hlrulename{Gr9}}
 \TrinaryInfC{$
@@
 \item \rulename{Gr6}: 
-TODO
+\hl{TODO}
 %The emission of a routed message $j$
@@
 \item \rulename{Gr7}:
-TODO
+\hl{TODO}
 
 \item \rulename{Gr8}:
-TODO
+\hl{TODO}
 
 \item \rulename{Gr9}:
-TODO
+\hl{TODO}
 
@@
 
-\subsection{LTS over Local Types}
+\subsection{LTS Semantics over Local Types}
 \label{subsection:newltslocal}
@@
 in \cref{fig:newlocal}.
-
-%TODO: LR1-3?
+We highlight and explain the new rules.
 
@@
 \AxiomC{}
-\RightLabel{\rulename{Lr4}}
+\RightLabel{\rulename{Lr1}}
+\UnaryInfC{$
+\treducelong
+	{\tsel{q}{l_i: T_i}{i \in I}}
+	{T_j}
+	{\aout{p}{q}{j}}
+$}
+\end{prooftree}
+
+\begin{prooftree}
+\AxiomC{}
+\RightLabel{\rulename{Lr2}}
+\UnaryInfC{$
+\treducelong
+	{\tbra{q}{l_i: T_i}{i \in I}}
+	{T_j}
+	{\ain{q}{p}{j}}
+$}
+\end{prooftree}
+
+\begin{prooftree}
+\AxiomC{$
+\treduce
+	{T[\trec{T} / \trecvar]}
+	{T'}
+	{l}
+$}
+\RightLabel{\rulename{Lr3}}
+\UnaryInfC{$
+\treduce
+	{\trec{T}}
+	{T'}
+	{l}
+$}
+\end{prooftree}
+
+\begin{prooftree}
+\AxiomC{}
+\RightLabel{\hlrulename{Lr4}}
 \UnaryInfC{$
@@
 \AxiomC{}
-\RightLabel{\rulename{Lr5}}
+\RightLabel{\hlrulename{Lr5}}
 \UnaryInfC{$
@@
 \AxiomC{}
-\RightLabel{\rulename{Lr6}}
+\RightLabel{\hlrulename{Lr6}}
 \UnaryInfC{$
@@
 \AxiomC{}
-\RightLabel{\rulename{Lr7}}
+\RightLabel{\hlrulename{Lr7}}
 \UnaryInfC{$
@@
 \AxiomC{$\subj{l} \notin \{\mrole{p}, \mrole{q}\}$}
-\RightLabel{\rulename{Lr8}}
+\RightLabel{\hlrulename{Lr8}}
 \BinaryInfC{$
@@
 \AxiomC{$\forall i \in I \setminus \{ j \}. ~ T'_i = T_i$}
-\RightLabel{\rulename{Lr9}}
+\RightLabel{\hlrulename{Lr9}}
 \TrinaryInfC{$
@@
 \AxiomC{$\forall i \in I. ~ \treduce{T_i}{T'_i}{l}$}
-\RightLabel{\rulename{Lr10}}
+\RightLabel{\hlrulename{Lr10}}
 \TrinaryInfC{$
@@
 \AxiomC{$\forall i \in I. ~ \treduce{T_i}{T'_i}{l}$}
-\RightLabel{\rulename{Lr11}}
+\RightLabel{\hlrulename{Lr11}}
 \TrinaryInfC{$
@@
 
+\captionof{figure}{LTS over Local Types in \newtheory}
 \label{fig:newlocal}
-\captionof{figure}{LTS over Local Types in \newtheory}
 \end{figure}
@@
 \item \rulename{Lr4}: 
-TODO
+\hl{TODO}
 
 \item \rulename{Lr5}:
-TODO
+\hl{TODO}
 
@@
 \item \rulename{Lr6}:
-TODO
+\hl{TODO}
 
 \item \rulename{Lr7}:
-TODO
+\hl{TODO}
 
 \item \rulename{Lr10}:
-TODO
+\hl{TODO}
 
 \item \rulename{Lr11}:
-TODO
+\hl{TODO}
 
@@
 \item \rulename{Lr8}:
-TODO
+\hl{TODO}
 
 \item \rulename{Lr9}:
-TODO
+\hl{TODO}
 
 \end{itemize}
-
-%\subsection{LTS over Collections of Local Types}
-%\label{subsection:newltssystem}
-
-\subsection{Soundness and Completeness w.r.t. Projection}
-\label{subsection:newtraceeq}
-
-We work towards proving the soundness and completeness
-of our LTS semantics with respect to projection.
-Our approach is motivated from \cite{characterisation}:
-
-\begin{enumerate}
-
-\item We first extend the LTS semantics to
-a collection of local types (hereafter referred to
-as a \textit{configuration} to be consistent with
-the literature);
-
-\item Then, we extend the definition of projection to
-obtain the configuration of a global type 
-(hereafter referred to as the \textit{projected configuration});
-
-\item Finally, we prove the trace equivalence between 
-the global type and its projected configuration.
-
-\end{enumerate}
-
-\begin{theorem}[Trace Equivalence]
-Let $G$ be a global type with participants 
-$\mathcal{P} = \mathrm{\pt{G}}$.
-Then $G \approx (\vec T, \vec \epsilon)$.
-\end{theorem}
-
-\begin{proof}
-\dots
-\end{proof}
-
-\subsection{Deadlock Freedom}
-\label{subsection:newdeadlockfreedom}

### 2020-06-14 report/sections/general/theory/syntax.tex
Report: progress on chapter 8

@@
 \begin{definition}[Participants]
-\textit{The set of roles involved in the communication
-interactions specified by $G$, written} $\pt{G}$
-\textit{, is:}
+The set of roles involved in the communication
+interactions specified by $G$, written $\pt{G}$, is:
 
@@
 \begin{definition}[Projection]
-\textit{The projection of $G$ onto $\role{r}$,
-written $\proj{G}{r}$, is defined as:}
+The projection of $G$ onto $\role{r}$,
+written $\proj{G}{r}$, is defined as:
 
@@
 \begin{definition}[Merging Operator]
-\textit{The merging operator $\mergeop$ on local types is defined as:}
+The merging operator $\mergeop$ on local types
+is defined as:
 
@@
 \begin{definition}[Centroid]
-\textit{Let} $\centroid{G}{s}$ \textit{denote \dots}
+Let $\centroid{G}{s}$ denote \dots
 
@@
 \begin{definition}[Well-formedness]
-\textit{Let} $\wfnew{G}{s}$ 
-\textit{denote that the global type} $G$
-\textit{is well-formed with respect to the router} $\role{s}$.
+Let $\wfnew{G}{s}$ 
+denote that the global type $G$
+is well-formed with respect to the router $\role{s}$.
 

### 2020-06-14 report/sections/general/theory/lts.tex
Report: complete chapter 8.2

@@
 
-\item \rulename{Gr6}: 
-\hl{TODO}
-%The emission of a routed message $j$
-%results in the message
+\item \rulename{Gr6} and \rulename{Gr7}
+are analogue to \rulename{Gr1} and \rulename{Gr2}
+for describing the emission and reception of
+messages in routed communication, but uses
+the ``routed in-transit'' construct instead.
 
-\item \rulename{Gr7}:
-\hl{TODO}
-
-\item \rulename{Gr8}:
-\hl{TODO}
-
-\item \rulename{Gr9}:
-\hl{TODO}
+\item \rulename{Gr8} and \rulename{Gr9}
+are analogue to \rulename{Gr4} and \rulename{Gr5}
+in the sense that we only enforce the
+syntactic order of messages for the participants
+involved in the action $l$.
 
 \end{itemize}
+
+An important observation from
+\rulename{Gr8} and \rulename{Gr9} is that,
+for the router,
+the syntactic order of routed communication
+can be freely interleaved between
+the syntactic order of direct communication.
+This is a crucial result for proving
+that the router does not over-serialise
+communication -- 
+we show this in \cref{subsection:encodepreservecomm}.
 
@@
 
-We explain rules \rulename{Lr4} and \rulename{Lr5}
+We walk through rules \rulename{Lr4} and \rulename{Lr5}
 from the perspective of role $\role{p}$.
@@
 
-\item \rulename{Lr4}: 
-\hl{TODO}
-
-\item \rulename{Lr5}:
-\hl{TODO}
+\item \rulename{Lr4} and \rulename{Lr5} are
+analogue to \rulename{Lr1} and \rulename{Lr2}
+for sending and receiving messages respectively.
+The exception is that the new rules pattern-match
+on the proxy role $\role{s}$ on the local type
+and the routed label.
 
@@
 
-We explain rules \rulename{Lr6}, \rulename{Lr7},
+We walk through rules \rulename{Lr6}, \rulename{Lr7},
 \rulename{Lr10} and \rulename{Lr11}
@@
 
-\item \rulename{Lr6}:
-\hl{TODO}
+\item \rulename{Lr6} and \rulename{Lr7} are
+analogue to \rulename{Gr1} and \rulename{Gr2}.
+Intuitively, the router $\role{s}$ holds a
+``global'' perspective on the interaction
+between $\role{p}$ and $\role{q}$, which explains
+the correspondence with the LTS semantics over global
+types.
 
-\item \rulename{Lr7}:
-\hl{TODO}
+\item \rulename{Lr10} and \rulename{Lr11} allow
+the router to perform routing actions before handling
+their own direct communication. The syntax $l = \via{s}{\cdot}$
+means that the label $l$ is ``of the form'' of a routing
+action, i.e. there exists some $\role{p}, \role{q}, j$
+such that $l = \via{s}{\aout{p}{q}{j}}$ or 
+$l = \via{s}{\ain{p}{q}{j}}$.
+The constraint of $\subj{l} \neq \mrole{q}$
+prevents the violation of the syntactic order of messages
+sent and received by $\mrole{q}$.
 
-\item \rulename{Lr10}:
-\hl{TODO}
+Curious readers can consider the examples 
+$\proj{G_1}{s}$ and $\proj{G_2}{s}$
+to see why this constraint is needed.
 
-\item \rulename{Lr11}:
-\hl{TODO}
+\begin{align*}
+G_1 &= \gcommone{s}{r}{\tmsg{M1}} \grouteone{r}{q}{s}{\tmsg{M2}} \tend \\
+\proj{G_1}{s} &= \tselone{r}{\tmsg{M1}}. ~ 
+	\routerone{r}{q}{\tmsg{M2}}{\tend} \\~\\
+G_2 &= \gcommone{s}{r}{\tmsg{M1}} \grouteone{p}{r}{s}{\tmsg{M2}} \tend \\
+\proj{G_2}{s} &= \tselone{r}{\tmsg{M1}}. ~ 
+	\routerone{p}{r}{\tmsg{M2}}{\tend} \\
+\end{align*}
 
@@
 
-We explain the remaining inductively defined rules.
-
-\begin{itemize}
-
-\item \rulename{Lr8}:
-\hl{TODO}
-
-\item \rulename{Lr9}:
-\hl{TODO}
-
-\end{itemize}
+As for the remaining rules,
+\rulename{Lr8} and \rulename{Lr9} are analogue to
+\rulename{Gr4} and \rulename{Gr5} because the router
+holds a ``global'' perspective on the communication,
+so transitions that do not violate the syntactic order
+of messages between roles $\mrole{p}$ and $\mrole{q}$
+are allowed.

### 2020-06-14 report/sections/server/codegen/testing.tex
Report: minor fix

@@
 for testing -- this involves having the WebSocket package installed
-for server-side endpoints, the React.js framework (\cref{section:react})
+for server-side endpoints, the React.js framework (\cref{chap:react})
 instantiated for browser-side endpoints,

### 2020-06-14 report/sections/general/theory/syntax.tex
Report: complete chapter 8.2
Report: complete up to deadlock freedom

@@
 \begin{array}{rlr}
-G ::= & \dots & \text{Global Types} \\
+G ::= & & \text{Global Types} \\
+\mid & \tend & \text{Termination} \\
+\mid & \trecvar & \text{Type Variable} \\
+\mid & \trec{G} & \text{Recursive Type} \\
+\mid & \gcomm{p}{q}{l_i: G_i}{i \in I} & \text{Direct Communication} \\
 \mid & \groute{p}{q}{s}{l_i: G_i}{i \in I}
-& \text{Routed Communication through $\role{s}$} \\
+	& \text{\hl{Routed Communication}} \\
 \end{array}
@@
 is usually the router.
+Just as it is assumed that $\role{p} \neq \role{q}$ for
+direct communication,
+we also assume that
+$\role{p} \neq \role{q} \neq \role{s}$ for routed communication.
 
@@
 \begin{array}{>{\displaystyle}rc>{\displaystyle}l}
-
-\pt{\dots} & = & \dots \\
 
@@
 \begin{array}{rlr}
-T ::= & \dots & \text{Local Types} \\
+T ::= &  & \text{Local Types} \\
+\mid & \tend & \text{Termination} \\
+\mid & \trecvar & \text{Type Variable} \\
+\mid & \trec{T} & \text{Recursive Type} \\
+\mid & \tsel{p}{l_i: T_i}{i \in I} & \text{Selection} \\
+\mid & \tbra{p}{l_i: T_i}{i \in I} & \text{Branching} \\
 \mid & \router{p}{q}{l_i: T_i}{i \in I}
-& \text{Routing communication from $\role{p}$ to $\role{q}$} \\
+& \text{\hl{Routing Communication}} \\
 \mid & \tselproxy{p}{q}{l_i: T_i}{i \in I}
-& \text{Selecting from $\role{p}$ via proxy $\role{q}$} \\
+& \text{\hl{Proxy Selection}} \\
 \mid & \tbraproxy{p}{q}{l_i: T_i}{i \in I}
-& \text{Receiving from $\role{p}$ via proxy $\role{q}$} \\
+& \text{\hl{Proxy Branching}} \\
 \end{array}
@@
 the \textit{merging operator}, $\mergeop$.
-
 As the merging operator is defined on local
@@
 on the extended syntax in \cref{def:newmerge}.
-Recall that proxy selection and proxy
-branching behave in the same way as their ``non-proxified''
-counterparts -- the merging operator reflects this
-similarity.
 
@@
 The merging operator $\mergeop$ on local types
-is defined as:
+is:
 
@@
 \end{definition}
+
+Recall that proxy selection and proxy
+branching behave in the same way as their ``non-proxified''
+counterparts -- the merging operator reflects this
+similarity.
 
@@
 \begin{definition}[Centroid]
-Let $\centroid{G}{s}$ denote \dots
+Let $\centroid{G}{s}$ denote that $\mrole{s}$ is the 
+centroid of $G$.
 

### 2020-06-15 report/sections/general/theory/encoding.tex
Report: complete up to deadlock freedom
Report: complete chapter 8 sans appendix proofs

@@
-\section{Router-Parameterised Encoding}
+\section{From Canonical MPST to \newtheory}
 \label{section:encoding}
 
+We present an encoding from canonical MPST theory
+to \newtheory.
+This encoding is \textit{parameterised} by the router role 
+(conventionally denoted as $\mrole{s}$);
+the intuition is that we encode all communication interactions
+to involve $\mrole{s}$.
+For example, as motivated in \cref{chap:p2p},
+we can encode the \tprotocol{Two Buyer} protocol
+into routed multiparty session types with respect to the
+\trole{S}eller role as the router.
+When we extend \codegen to implement \newtheory in \cref{chap:impl},
+the developer can implement the \tprotocol{Two Buyer} protocol
+as an interactive web application over WebSocket transport.
+
+We define the encoding in \cref{subsection:encoding}.
+More importantly, we prove that the encoded routed communications
+preserve the communication safety
+properties (\cref{subsection:encodepreservewf}) 
+and communication structure (\cref{subsection:encodepreservecomm}) 
+of the original 
+communication. 
+This directly addresses the challenges discussed in
+\cref{section:routerchallenges}
+
+\subsection{Router-Parameterised Encoding}
+\label{subsection:encoding}
+
+We define the router-parameterised encoding
+on global types, local types and LTS labels in
+the canonical MPST theory.
+
+We start with global types, 
+as presented in \cref{def:encglobal}.
+The main rule is \rulename{Enc-G-Comm}:
+if the communication did not go through $\role{s}$,
+then the encoded communication involves $\role{s}$ as the router.
+The remaining rules are self-explanatory.
+
+\begin{definition}[Encoding on Global Types]
+
+\doublespacing
+\[
+\begin{array}{rclr}
+\enc{\tend}{s} &= & \tend & \rulename{Enc-G-End} \\
+\enc{\trecvar}{s} &= & 
+	\trecvar & \rulename{Enc-G-RecVar} \\
+\enc{\trec{G}}{s} &= & 
+	\trec{\enc{G}{s}} & \rulename{Enc-G-Rec} \\
+\enc{\gcomm{p}{q}{l_i: G_i}{i \in I}}{s} &= & 
+	\begin{cases}
+	\gcomm{p}{q}{l_i: \enc{G_i}{s}}{i \in I} 
+		& \text{if } \mrole{s} \in \{ \mrole{p}, \mrole{q} \} \\
+	\groute{p}{q}{s}{l_i: \enc{G_i}{s}}{i \in I} 
+		& \text{otherwise} \\	
+	\end{cases} & \rulename{Enc-G-Comm} \\
+\end{array}
+\]
+\singlespacing
+
+\label{def:encglobal}
+\end{definition}
+
+Readers may observe striking similarities between
+the encoding on global types and the definition of a
+centroid in routed communication (\cref{def:centroid}).
+In fact, the two are the same -- the encoding on global
+types simply expresses the centroid predicate as a function.
+We arrive at \cref{lem:enccenter} -- the proof is straightforward
+by induction on global types and left as an exercise to the reader.
+
+\begin{lemma}[Encoding Defines Centroid]
+Given an encoding of global type $G$ with respect to the
+router role $\mrole{s}$,
+the router role is the centroid of the encoded communication.
+
+\[
+\centroid{\enc{G}{s}}{s}
+\]
+
+\label{lem:enccenter}
+\end{lemma}
+
+We now define the encoding on local types in \cref{def:enclocal}.
+Local types express communication from the perspective of a 
+particular role, say $\role{q}$. 
+We need to capture this information in the encoding
+to accurately encode into the routed communication,
+so the encoding on local types is parameterised by
+\textit{two} roles.
+
+\begin{definition}[Encoding on Local Types]
+The encoding of interpreting local type $T$ (from the
+perspective of role $\mrole{q}$) with respect to
+the router role $\mrole{s}$, or $\enclocal{T}{q}{s}$, is defined as:
+
+\doublespacing
+\[
+\begin{array}{rclr}
+\enclocal{\tend}{q}{s} &= & \tend & \rulename{Enc-L-End} \\
+\enclocal{\trecvar}{q}{s} &= & \trecvar & \rulename{Enc-L-RecVar} \\
+\enclocal{\trec{T}}{q}{s} &= & \trec{\enclocal{T}{q}{s}} 
+	& \rulename{Enc-L-Rec} \\
+\enclocal{\tsel{p}{l_i: T_i}{i\in I}}{q}{s} &= 
+	& \begin{cases}
+	\tsel{p}{l_i: \enclocal{T_i}{q}{s}}{i\in I}
+		& \text{if } \mrole{s} \in \{ \mrole{p}, \mrole{q} \} \\
+	\tselproxy{p}{s}{l_i: \enclocal{T_i}{q}{s}}{i\in I}
+		& \text{otherwise} \\	
+	\end{cases}
+	& \rulename{Enc-L-Sel} \\
+\enclocal{\tbra{p}{l_i: T_i}{i\in I}}{q}{s} &= 
+	& \begin{cases}
+	\tbra{p}{l_i: \enclocal{T_i}{q}{s}}{i\in I}
+		& \text{if } \mrole{s} \in \{ \mrole{p}, \mrole{q} \} \\
+	\tbraproxy{p}{s}{l_i: \enclocal{T_i}{q}{s}}{i\in I}
+		& \text{otherwise} \\	
+	\end{cases}
+	& \rulename{Enc-L-Bra} \\
+\end{array}
+\]
+\singlespacing
+
+\label{def:enclocal}
+\end{definition}
+
+Similarly, if the original communication does not involve
+the router from a local perspective, the local endpoint
+will carry out the communication through the router
+$\trole{s}$ as a proxy in the encoded local type.
+This is expressed in both \rulename{Enc-L-Sel}
+and \rulename{Enc-L-Bra}.
+
+We establish a correspondence between the encodings for
+global types and local types.
+
+\[
+\forall \mrole{r}, \mrole{s}, G. ~ \left(
+\mrole{r} \neq \mrole{s} 
+	\Longrightarrow
+\proj{\enc{G}{s}}{r} = \enclocal{\proj{G}{r}}{r}{s}
+\right)
+\]
+
+We formalise this as \cref{lem:enclink}.
+The constraint $\mrole{r} \neq \mrole{s}$ is necessary
+because we would otherwise lose information on the right-hand
+side of the equality:
+the projection of $\mrole{s}$ in the original
+communication will not contain the routed interactions,
+so applying the local type encoding cannot recover this 
+information.
+We prove \cref{lem:enclink} by induction on the structure of $G$;
+the proof is mechanical and is left in \cref{section:proofs}
+for the interested reader.
+
+This correspondence lets us prove a more useful 
+lemma about how our encoding preserves the equality of projections,
+as presented in \cref{lem:encprojeq}.
+
+\begin{lemma}
+\[
+\forall G_1, G_2, \mrole{r}, \mrole{s}. ~ \left(
+(\proj{G_1}{r}) = (\proj{G_2}{r}) \wedge \mrole{r} \neq \mrole{s}
+	\Longrightarrow
+\proj{\enc{G_1}{s}}{r} = \proj{\enc{G_2}{s}}{r}
+\right)
+\]
+
+\label{lem:encprojeq}
+\end{lemma}
+
+\begin{proof} By consequence from \cref{lem:enclink}.
+
+Take $G_1, G_2, \mrole{r}, \mrole{s}$ arbitrarily. 
+Assume $(\proj{G_1}{r}) = (\proj{G_2}{r})$ and $\mrole{r} \neq \mrole{s}$.
+
+We need to show $\proj{\enc{G_1}{s}}{r} = \proj{\enc{G_2}{s}}{r}$,
+but by \cref{lem:enclink}, it is sufficient to show
+\[
+\enclocal{\proj{G_1}{r}}{r}{s} = \enclocal{\proj{G_2}{r}}{r}{s}
+\]
+
+Define an ``inner'' function $f(z) = \enclocal{z}{r}{s}$.
+
+By definition, $\forall x, y. (x = y \Longrightarrow f(x) = f(y))$.
+
+We have $(\proj{G_1}{r}) = (\proj{G_2}{r})$ by assumption,
+so we can conclude.
+
+\begin{align*}
+(\proj{G_1}{r}) = (\proj{G_2}{r}) 
+&\Longrightarrow f(\proj{G_1}{r}) = f(\proj{G_2}{r})  \\
+&\Longrightarrow
+\enclocal{\proj{G_1}{r}}{r}{s} = \enclocal{\proj{G_2}{r}}{r}{s}
+\end{align*}
+
+%\[
+%(\proj{G_1}{r}) = (\proj{G_2}{r}) 
+%\Longrightarrow
+%f(\proj{G_1}{r}) = f(\proj{G_2}{r}) 
+%\Longrightarrow
+%\enclocal{\proj{G_1}{r}}{r}{s} = \enclocal{\proj{G_2}{r}}{r}{s}
+%\]
+
+\end{proof}
+
+We also encode LTS actions using the same idea. 
+This is presented in \cref{def:enclts}, and is required for
+proofs in \cref{subsection:encodepreservecomm}.
+
+\begin{definition}[Encoding on LTS Actions]
+
+\doublespacing
+\[
+\begin{array}{rclr}
+\enc{\aout{p}{q}{j}}{s} &= & \begin{cases}
+\aout{p}{q}{j}
+	& \text{if } \mrole{s} \in \{ \mrole{p}, \mrole{q} \} \\
+\via{s}{\aout{p}{q}{j}}
+	& \text{otherwise} \\
+\end{cases} & \rulename{Enc-L-Out} \\
+\enc{\ain{p}{q}{j}}{s} &= & \begin{cases}
+\ain{p}{q}{j}
+	& \text{if } \mrole{s} \in \{ \mrole{p}, \mrole{q} \} \\
+\via{s}{\ain{p}{q}{j}}
+	& \text{otherwise} \\
+\end{cases} & \rulename{Enc-L-In} \\
+\end{array}
+\]
+\singlespacing
+
+\label{def:enclts}
+\end{definition}
+
+\subsection{Preserving Well-formedness}
+\label{subsection:encodepreservewf}
+
+We show that our encoding preserves the well-formedness
+of global types in \cref{th:encwf}.
+Well-formedness in \newtheory defines a constraint
+on the centroid of the routed communication.
+As this directly follows from \cref{lem:enccenter},
+it is sufficient to prove that our encoding preserves
+projection (\cref{lem:encproj}), then arrive at \cref{th:encwf}
+by consequence.
+
+\begin{lemma}[Encoding Preserves Projection]
+Let $G$ be a global type.
+Take arbitrary roles $\mrole{r}, \mrole{s}$.
+
+\[
+\proj{G}{r} \text{ exists } 
+	\Longrightarrow 
+\proj{ \enc{G}{s} }{r} \text{ exists }
+\]
+
+\label{lem:encproj}
+\end{lemma}
+
+\begin{proof}
+By induction on the structure of $G$.
+
+\begin{enumerate}
+
+\item $G = \tend$, $G = \trecvar$
+
+As $\enc{G}{s} = G$, if $\proj{G}{r}$ exists, 
+so does $\proj{ \enc{G}{s} }{r}$.
+
+\item $G = \trec{G'}$
+
+By assumption, $\proj{\trec{G'}}{r}$ exists.
+Note that $\proj{G'}{r}$ exists regardless of $\mrole{r}$'s
+participation in $G'$.
+
+By induction, $\proj{\enc{G'}{s}}{r}$ exists.
+
+To show $\proj{\enc{\trec{G'}}{s}}{r}$ exists,
+consider $\mrole{r}$ by case:
+
 \begin{itemize}
-\item define encoding on global types
-\item define encoding on local types
-\item define encoding on LTS actions
-\item motivate with twobuyer example
+\item $\mrole{r} \in \pt{\enc{G'}{s}}$:
+
+\[
+\proj{\enc{\trec{G'}}{s}}{r}
+= \proj{\trec{\enc{G'}{s}}}{r}
+= \trec{\proj{\enc{G'}{s}}{r}}
+\]
+
+As $\proj{\enc{G'}{s}}{r}$ exists,
+so does $\proj{\enc{\trec{G'}}{s}}{r}$.
+
+\item $\mrole{r} \notin \pt{\enc{G'}{s}}$:
+
+\[
+\proj{\enc{\trec{G'}}{s}}{r}
+= \proj{\trec{\enc{G'}{s}}}{r}
+= \tend
+\]
+
 \end{itemize}
 
-\subsection{Correspondence between Global and Local Type Encodings}
+\item $G = \gcomm{p}{q}{l_i: G_i}{i \in I}$
+
+To determine $\enc{G}{s}$, consider $\mrole{s}$ by case:
+
 \begin{itemize}
-\item prove proposition
-\item state corollary for subsequent proof
+
+\item $\mrole{s} \in \{ \mrole{p}, \mrole{q} \}$:
+
+Then $\enc{G}{s} = \gcomm{p}{q}{l_i: \enc{G_i}{s}}{i \in I}$.
+
+To show $\proj{\enc{G}{s}}{r}$ exists,
+consider $\mrole{r}$ by case:
+
+\begin{itemize}
+\item $\mrole{r} = \mrole{p}$:
+Then $\proj{G}{p} = \tsel{q}{l_i: \proj{G_i}{p}}{i \in I}$.
+
+By induction, $\proj{\enc{G_i}{s}}{p}$ exists for $i \in I$.
+
+$\proj{\enc{G}{s}}{p} = \tsel{q}{l_i: \proj{\enc{G_i}{s}}{p}}{i \in I}$.
+
+As projections of the encoded continuations exist, 
+so does $\proj{\enc{G}{s}}{p}$.
+
+\item $\mrole{r} = \mrole{q}$: Follows similarly from above.
+
+\item $\mrole{r} \notin \{\mrole{p},\mrole{q}\}$:
+Then $\proj{G}{r} = \underset{i \in I}{\MERGEOP}\proj{G_i}{r}$,
+so the merge exists.
+
+By induction, $\proj{\enc{G_i}{s}}{r}$ exists for $i \in I$.
+
+$\proj{\enc{G}{s}}{r} = 
+	\underset{i \in I}{\MERGEOP}\proj{\enc{G_i}{s}}{r}$,
+and this merge exists by \cref{lem:encglobalpreservemerge}.
+
 \end{itemize}
 
-\subsection{Preserving Well-formedness}
+\item $\mrole{s} \notin \{ \mrole{p}, \mrole{q} \}$:
+
+Then $\enc{G}{s} = \groute{p}{q}{s}{l_i: \enc{G_i}{s}}{i \in I}$.
+
+To show $\proj{\enc{G}{s}}{r}$ exists,
+consider $\mrole{r}$ by case:
+
+\begin{itemize}
+\item $\mrole{r} = \mrole{p}$:
+Then $\proj{G}{p} = \tsel{q}{l_i: \proj{G_i}{p}}{i \in I}$.
+
+By induction, $\proj{\enc{G_i}{s}}{p}$ exists for $i \in I$.
+
+$\proj{\enc{G}{s}}{p} = 
+	\tselproxy{q}{s}{l_i: \proj{\enc{G_i}{s}}{p}}{i \in I}$.
+
+As projections of the encoded continuations exist, 
+so does $\proj{\enc{G}{s}}{p}$.
+
+\item $\mrole{r} = \mrole{q}$: Follows similarly from above.
+\item $\mrole{r} = \mrole{s}$:
+Then $\proj{G}{s} = \underset{i \in I}{\MERGEOP}\proj{G_i}{s}$.
+
+By induction, $\proj{\enc{G_i}{s}}{p}$ exists for $i \in I$.
+
+$\proj{\enc{G}{s}}{s} = 
+	\router{p}{q}{l_i: \proj{\enc{G_i}{s}}{s}}{i \in I}$.
+
+As projections of the encoded continuations exist, 
+so does $\proj{\enc{G}{s}}{s}$.
+
+\item $\mrole{r} \notin \{\mrole{p},\mrole{q},\mrole{s}\}$:
+Then $\proj{G}{r} = \underset{i \in I}{\MERGEOP}\proj{G_i}{r}$,
+so the merge exists.
+
+By induction, $\proj{\enc{G_i}{s}}{r}$ exists for $i \in I$.
+
+$\proj{\enc{G}{s}}{r} = 
+	\underset{i \in I}{\MERGEOP}\proj{\enc{G_i}{s}}{r}$,
+and this merge exists by \cref{lem:encglobalpreservemerge}.
+
+\end{itemize} % participation of r
+
+\end{itemize} % participation of s
+
+\end{enumerate} % structure of G
+
+\end{proof}
+
+\begin{theorem}[Encoding Preserves Well-Formedness]
+Let $G$ be a global type, and $\mrole{s}$ be a role.
+
+\[
+\wf{G} \Longrightarrow \wfnew{\enc{G}{s}}{s}
+\]
+
+\label{th:encwf}
+\end{theorem}
+
+\begin{proof}
+Direct consequence of \cref{lem:encproj,lem:enccenter,def:newwf}.
+\end{proof}
 
 \subsection{Preserving Communication}
+\label{subsection:encodepreservecomm}
+
+We present a crucial result that directly addresses 
+the pitfalls discussed in \cref{section:routerchallenges} --
+namely, that our encoding does not over-serialise the
+original communication.
+We show this through proving that our encoding preserves
+the LTS semantics over global types -- or more precisely,
+we can use the encodings over global types and LTS actions
+to encode all possible transitions in the LTS for
+global types in the canonical MPST theory.
+
+\begin{theorem}[Encoding Preserves Semantics]
+Let $G, G'$ be global types
+such that $\treduce{G}{G'}{l}$.
+Also take arbitrary role $\mrole{s}$.
+
+\[
+\treduce{G}{G'}{l}
+	\Longrightarrow
+\treducelong{\enc{G}{s}}{\enc{G'}{s}}{\enc{l}{s}}
+\]
+
+\end{theorem}
+
+\begin{proof}
+By induction on the structure of $\treduce{G}{G'}{l}$.
+
+\begin{itemize}
+
+\item \rulename{Gr1},
+where $G = \gcomm{p}{q}{l_i: G_i}{i \in I},
+G' = \gtrans{p}{q}{j}{l_i: G_i}{i \in I},
+l = \aout{p}{q}{j}$.
+
+To show $\treducelong
+{\enc{G}{s}}{\enc{G'}{s}}{\enc{l}{s}}$,
+consider $\mrole{s}$ by case:
+
+\begin{itemize}
+
+\item $\mrole{s} \in \{ \mrole{p}, \mrole{q} \}$:
+Then we have
+
+\begin{align*}
+\enc{G}{s} &= \gcomm{p}{q}{l_i: \enc{G_i}{s}}{i \in I} \\
+\enc{G'}{s} &= \gtrans{p}{q}{j}{l_i: \enc{G_i}{s}}{i \in I} \\
+\enc{l}{s} &= \aout{p}{q}{j} \\
+\end{align*}
+
+The encoded transition is possible using \rulename{Gr1}.
+
+\item $\mrole{s} \notin \{ \mrole{p}, \mrole{q} \}$:
+Then we have
+
+\begin{align*}
+\enc{G}{s} &= \groute{p}{q}{s}{l_i: \enc{G_i}{s}}{i \in I} \\
+\enc{G'}{s} &= \gtransroute{p}{q}{s}{j}{l_i: \enc{G_i}{s}}{i \in I} \\
+\enc{l}{s} &= \via{s}{\aout{p}{q}{j}} \\
+\end{align*}
+
+The encoded transition is possible using \rulename{Gr6}.
+
+\end{itemize}
+
+\item \rulename{Gr2},
+where $G = \gtrans{p}{q}{j}{l_i: G_i}{i \in I},
+G' = G_j,
+l = \ain{p}{q}{j}$.
+
+We know $\enc{G'}{s} = \enc{G_j}{s}$
+
+To show $\treducelong
+{\enc{G}{s}}{\enc{G'}{s}}{\enc{l}{s}}$,
+consider $\mrole{s}$ by case:
+
+\begin{itemize}
+
+\item $\mrole{s} \in \{ \mrole{p}, \mrole{q} \}$:
+Then we have
+
+\begin{align*}
+\enc{G}{s} &= \gtrans{p}{q}{j}{l_i: \enc{G_i}{s}}{i \in I} \\
+\enc{l}{s} &= \ain{p}{q}{j} \\
+\end{align*}
+
+The encoded transition is possible using \rulename{Gr2}.
+
+\item $\mrole{s} \notin \{ \mrole{p}, \mrole{q} \}$:
+Then we have
+
+\begin{align*}
+\enc{G}{s} &= \gtransroute{p}{q}{s}{j}{l_i: \enc{G_i}{s}}{i \in I} \\
+\enc{l}{s} &= \via{s}{\ain{p}{q}{j}} \\
+\end{align*}
+
+The encoded transition is possible using \rulename{Gr7}.
+
+\end{itemize}
+
+\item \rulename{Gr3},
+where $G = \trec{G''}$.
+
+By hypothesis, $\treduce{G''[\trec{G''} / \trecvar]}{G'}{l}$.
+
+By induction, $
+\treducelong
+	{\enc{G''[\trec{G''} / \trecvar]}{s}}
+	{\enc{G'}{s}}
+	{\enc{l}{s}}
+$.
+
+By \cref{lem:encsub}, 
+$\enc{G''[\trec{G''} / \trecvar]}{s} 
+= \enc{G''}{s}\left[ \trec{\enc{G''}{s}} / \trecvar \right]$.
+
+We know $\enc{G}{s} = \enc{\trec{G''}}{s} = \trec{\enc{G''}{s}}$.
+
+The encoded transition is possible using \rulename{Gr3} as shown:
+
+\begin{prooftree}
+\AxiomC{$
+\treducelong
+	{\enc{G''}{s}\left[ \trec{\enc{G''}{s}} / \trecvar \right]}
+	{\enc{G'}{s}}
+	{\enc{l}{s}}
+$}
+\RightLabel{\rulename{Gr3}}
+\UnaryInfC{$
+\treducelong{\trec{\enc{G''}{s}}}{\enc{G'}{s}}{\enc{l}{s}}
+$}
+\end{prooftree}
+
+\item \rulename{Gr4},
+where $G = \gcomm{p}{q}{l_i: G_i}{i \in I},
+G' = \gcomm{p}{q}{l_i: G'_i}{i \in I}$.
+
+By hypothesis,
+$\forall i \in I. ~ \treduce{G_i}{G'_i}{l}$
+and $\subj{l} \notin \{ \mrole{p}, \mrole{q} \}$.
+
+By induction, 
+$\forall i \in I. ~ \left(
+\treducelong{\enc{G_i}{s}}{\enc{G'_i}{s}}{\enc{l}{s}}
+\right)$.
+
+By \cref{def:newsubj}, 
+$\subj{\enc{l}{s}} \notin \{ \mrole{p}, \mrole{q} \}$.
+
+To show $\treducelong
+{\enc{G}{s}}{\enc{G'}{s}}{\enc{l}{s}}$,
+consider $\mrole{s}$ by case:
+
+\begin{itemize}
+
+\item $\mrole{s} \in \{ \mrole{p}, \mrole{q} \}$:
+Then we have
+
+\begin{align*}
+\enc{G}{s} &= \gcomm{p}{q}{l_i: \enc{G_i}{s}}{i \in I} \\
+\enc{G'}{s} &= \gcomm{p}{q}{l_i: \enc{G'_i}{s}}{i \in I} \\
+\end{align*}
+
+The encoded transition is possible using \rulename{Gr4}.
+
+\item $\mrole{s} \notin \{ \mrole{p}, \mrole{q} \}$:
+Then we have
+
+\begin{align*}
+\enc{G}{s} &= \groute{p}{q}{s}{l_i: \enc{G_i}{s}}{i \in I} \\
+\enc{G'}{s} &= \groute{p}{q}{s}{l_i: \enc{G'_i}{s}}{i \in I} \\
+\end{align*}
+
+The encoded transition is possible using \rulename{Gr8}.
+
+\end{itemize}
+
+\item \rulename{Gr5},
+where $G = \gtrans{p}{q}{j}{l_i: G_i}{i \in I},
+G' = \gtrans{p}{q}{j}{l_i: G'_i}{i \in I}$.
+
+By hypothesis,
+$\treduce{G_j}{G'_j}{l}$,
+$\mrole{p'} \neq \subj{l}$,
+and $\forall i \in I \setminus \{j\}. ~ G'_i = G_i$.
+
+By induction, 
+$\treducelong{\enc{G_j}{s}}{\enc{G'_j}{s}}{\enc{l}{s}}$.
+
+By \cref{def:newsubj}, 
+$\subj{\enc{l}{s}} \neq \mrole{q}$.
+
+To show $\treducelong
+{\enc{G}{s}}{\enc{G'}{s}}{\enc{l}{s}}$,
+consider $\mrole{s}$ by case:
+
+\begin{itemize}
+
+\item $\mrole{s} \in \{ \mrole{p}, \mrole{q} \}$:
+Then we have
+
+\begin{align*}
+\enc{G}{s} &= \gtrans{p}{q}{j}{l_i: \enc{G_i}{s}}{i \in I} \\
+\enc{G'}{s} &= \gtrans{p}{q}{j}{l_i: \enc{G'_i}{s}}{i \in I} \\
+\end{align*}
+
+The encoded transition is possible using \rulename{Gr5}.
+
+\item $\mrole{s} \notin \{ \mrole{p}, \mrole{q} \}$:
+Then we have
+
+\begin{align*}
+\enc{G}{s} &= \gtransroute{p}{q}{s}{j}{l_i: \enc{G_i}{s}}{i \in I} \\
+\enc{G'}{s} &= \gtransroute{p}{q}{s}{j}{l_i: \enc{G'_i}{s}}{i \in I} \\
+\end{align*}
+
+The encoded transition is possible using \rulename{Gr9}.
+
+\end{itemize}
+
+\end{itemize}
+
+\end{proof}

### 2020-06-15 report/sections/background/temp.tex
Report: add interim background _temporarily_

\section{Session Types}

%\subsection{Overview}
% Motivate different ways of modelling concurrency, narrow the focus to message passing
Web applications are one of many examples of distributed systems in practice. Distributed systems are built upon the interaction between concurrent processes, which can be implemented using the two main communication abstractions in {shared memory} and {message passing}.

Shared memory provides processes with the impression of a logical single large monolithic memory but requires programmers to understand consistency models in order to correctly reason about the consistency of shared state.

Message passing interprets the interaction between processes as the exchange of messages, and best describes the communication transports found in web applications, ranging from the stateless request-response client-server interactions via HTTP to full-duplex communication channels via the WebSocket protocol \cite{WebSocketRFC}.

% Outline relevance of process algebra and session types as the relevant typing discipline
The process algebra $\pi$-calculus introduced by Milner in \cite{Milner1999} provides a formalism of the message passing abstraction in terms of the basic building blocks of sending and receiving processes, along with inductively defined continuation processes. The composition of these primitives allow us to describe more complex communication sessions.
Session types define the typing discipline for the $\pi$-calculus and provide reliability guarantees for communication sessions; the latter addresses a key challenge when reasoning about the correctness of distributed systems.

% Outline existing implementations of session types in practice
Many studies are done on the practical applications of session types, from developing languages providing native session type support \cite{ATS2016} to implementing session types in existing programming languages across different paradigms.
Implementations of the latter approach differ by how they leverage the design philosophy and features provided by the programming language. For example, King et al. leveraged the expressive type system of PureScript to perform static session type checking in \cite{PureScript2019}, whilst Neykova and Yoshida introduced dynamic approaches to check the conformance of Python programs with respect to session types in \cite{Python2017}.

\subsection{Asynchronous $\pi$-calculus}\label{section_async}

% Introduce pi-calculus, monadic asynchronous pi-calculus.
The $\pi$-calculus models concurrent computation, where processes can execute in parallel and communicate via shared names.
We first consider the asynchronous $\pi$-calculus introduced by Honda and Tokoro in \cite{AsyncHonda}.
Among the many flavours of the calculus which vary depending on the application domain, we outline the variant as presented in \cite{C406Lecture}.

Figure \ref{fig:async} defines the syntax of processes in asynchronous $\pi$-calculus; the asynchrony comes from the lack of continuation in the output process.

\begin{itemize}
\item $\mathbf{0}$ is the nil process and represents inactivity.
\item $\pout{u}{v}$ is the output process that will send value $v$ on $u$.
\item $\pin{u}{x}.P$ is the input process that, upon receiving a message on $u$, will bind the message to $x$ and carry on executing $P$ under this binding.
\item $P\mid Q$ represents the parallel composition of processes executing simultaneously.
\item $!P$ represents the parallel composition of infinite instances of $P$; more specifically, $!P \equiv P \mid {!P}$.
\item $(\nu a)~P$ represents a name restriction where any occurrence of $a$ in $P$ is local and will not interfere with other names outside the scope of $P$.
\end{itemize}

\begin{figure}[!hb]
\doublespacing
\[
\begin{array}{rlr}

P,Q ::= & & \text{Processes} \\
     & \mathbf{0} & \text{Nil Process} \\
\mid & \pout{u}{v} & \text{Output} \\
\mid & \pin{u}{x}.P & \text{Input} \\     
\mid & P \mid Q & \text{Parallel Composition} \\
\mid & !P & \text{Replication} \\
\mid & (\nu a)~P & \text{Restriction} \\

u,v ::= & & \text{Identifiers} \\
     & a, b, c,~\dots & \text{Names} \\
\mid & x, y, z,~\dots & \text{Variables} \\

\end{array}
\]
\singlespacing
\caption{Syntax of Asynchronous $\pi$-calculus}
\label{fig:async}
\end{figure}

% Motivate how this models interactions between concurrent processes through the reduction rule in operational semantics.
The operational semantics model the interaction between parallel processes. Whilst \cite{C406Lecture} presents the full operational semantics, we highlight the \rulename{Comm} reduction rule which specifically models message passing:
if the parallel composition of an input process and output process share the same name, the composition reduces to the continuation of the input process, substituting the variable $x$ with the message received $v$. We omit the definitions of substitution, free variables and free names, $\alpha$-equivalence and structural congruence; the interested reader may refer to \cite{C406Lecture}.

%\[
%\infer[\mbox{\rulename{Comm}}]{\pout{a}{v} \mid \pin{a}{x}.P ~\longrightarrow ~ P[v/x] }{}
%\]

We additionally define a process $P$ to be {stuck} if $P$ is not the nil process and $P$ cannot be reduced any further. For example, the process $P = \pin{a}{x}.\mathbf{0} \mid \pout{b}{v}$ is stuck as the parallel composition of an input process and an output process that do not share the same name cannot be reduced using \rulename{Comm}. In practice, a stuck process contains communications that will never be executed.

\subsection{Binary Session Types}\label{section_bst}
% Introduce synchronous pi calculus with branching and selection
A {binary session} is a parallel composition of two processes, each representing a distinct participant.
In the context of web applications, a binary session would describe the interactions between client and server.
Without loss of generality, a {session} represents the sequence of send and receive actions of a single participant.

We introduce a {synchronous} session calculus inspired by \cite{MPST}. Figure \ref{fig:sync} defines the syntax formally; we briefly discuss the main components and how it differs from the variant introduced in \cite{C406Lecture}:

% polyadic synchronous
% communication must come with label
% equi-recursive recursion, assume guarded
\begin{itemize}
\item \textbf{Synchronous communication}: Processes that send a message will have a continuation which will be executed upon a successful send. 
\item \textbf{Polyadic communication}: A vector of values can be communicated at once.
\item \textbf{Branching and selection}: A branching process can offer a set of branches, each defined by its own label identifier and continuation process. A selection process can select a branch by sending the corresponding label identifier alongside the payload to the branching process.

\item \textbf{Labelled messages}: a label identifier is attached to all messages; the input process in {\ref{section_async}} is generalised as a branching process that offers one branch.
\end{itemize}

The \rulename{Comm} rule in the operational semantics for this calculus exemplifies these new additions: given a binary session between distinct participants $\role{p}$ and $\role{q}$ where $\role{q}$ offers a set of labelled branches, if $\role{p}$ selects a label offered by $\role{q}$ and sends a vector of expressions $e_1, \dots, e_n$ that evaluate\footnote{We adopt the operational semantics for expression evaluation $e \downarrow v$ as defined in \cite{C406Lecture}.} to the corresponding vector of values $v_1, \dots, v_n$, the session reduces to a session with the continuation from the selection process composed with the continuation from the selected branch of the branching process, the latter having the variables $x_1, \dots, x_n$ substituted with the vector of values $v_1, \dots, v_n$ received.

%\[
%\infer[\mbox{\rulename{Comm}}]{
%	\role{p} :: \sel{\role{q}}{l}{e_1 \dots e_n}.~P \mid 
%	\role{q} :: \bra{\role{p}}{l_i(x_1 \dots x_n):Q_i}_{i\in I} ~\longrightarrow~ \role{p} :: P \mid \role{q} ::Q_j[v_k/x_k]^n_{k=1}
%}{
%	\begin{array}{cccccc}
%	\exists	j \in I. l_j = l
%	&
%	e_1 \downarrow v_1 
%	& \dots 
%	& e_n \downarrow v_n
%	&
%	& \role{p} \neq \role{q}
%	\end{array}
%}
%\]

\begin{figure}[!hb]
\doublespacing
\[
\begin{array}{rlr}

%v ::= & \underline{n}~\mid~\texttt{true}~\mid~\texttt{false} & \text{Values} \\
%e, e' ::= & & \text{Expressions} \\
%	 & v & \text{Values} \\
%\mid	 & x & \text{Variables} \\
%\mid & e + e'~\mid~e - e' & \text{Arithmetic Operators} \\
%\mid & e = e'~\mid~e < e' ~\mid~e > e' & \text{Relational Operators} \\
%\mid & e \wedge e'~\mid~e \vee e' ~\mid~\neg e & \text{Logical Operators} \\
%\mid & e \oplus e' & \text{Non-Determinism} \\
%
%\role{p} ::= & \role{Client},~\role{Server}& \text{Participant} \\
%
%P,Q ::= & & \text{Processes} \\
%     & \mathbf{0} & \text{Nil Process} \\
%\mid & \bra{\role{p}}{l_i(x_1 \dots x_n):P_i}_{i\in I} & \text{Branching} \\
%\mid & \sel{\role{p}}{l}{e_1 \dots e_n}.~P & \text{Selection} \\
%\mid & \texttt{if}~e~\texttt{then}~P~\texttt{else}~Q & \text{Conditional} \\
%\mid & \mu X.~P & \text{Recursive Process} \\
%\mid & X & \text{Process Variable} \\
%
%l, l' ::= & {\tt{``str''}} & \text{Label Identifiers} \\
%
%\mathcal{M} ::= & \role{p} :: P~\mid~\role{q} :: Q & \text{Binary Session} \\
\end{array}
\]

\singlespacing
\caption{Syntax of Session Calculus with Branching, Selection and Recursion}
\label{fig:sync}
\end{figure}

Additionally, the calculus introduces:

\begin{itemize}
\item \textbf{Conditionals}: If $e \downarrow \texttt{true}$, the process $\texttt{if}~e~\texttt{then}~P~\texttt{else}~Q$ reduces to $P$; if $e \downarrow \texttt{false}$, the process $\texttt{if}~e~\texttt{then}~P~\texttt{else}~Q$ reduces to $Q$.
\item \textbf{Recursion}: Following the equirecursive approach, the occurrence of the process variable $X$ in the recursive process can be expanded into the process transparently; more specifically, $\mu X. P \equiv P[(\mu X. P)/X]$.
\end{itemize}

% Introduce binary session types
{Session types} represent the type theory for our session calculus. We define the syntax of session types for binary sessions in figure \ref{fig:bst}.

\begin{figure}[!hb]
\doublespacing
\[
\begin{array}{rlr}

U ::= & \text{\tt{int}}~\mid~\text{\tt{bool}} & \text{Sorts} \\

S ::= & & \text{Session Types} \\
     & \mathbf{end} & \text{Termination} \\
\mid & \tbra{\role{p}}{l_i(U_1 \dots U_n):S_i}{i\in I} & \text{Branching} \\
\mid & \tsel{\role{p}}{l_i(U_1 \dots U_n):S_i}{i\in I} & \text{Selection} \\
\mid & \mu \mathbf{t}.~S & \text{Recursive Type} \\
\mid & \mathbf{t} & \text{Type Variable} \\
\end{array}
\]

\singlespacing
\caption{Syntax of Session Types}
\label{fig:bst}
\end{figure}

We derive the type of a process with a {typing judgement} of the form $\Gamma \vdash P: S$, which reads, \textit{under the typing context $\Gamma$, process $P$ has session type $S$}.

The {typing context} records typing assumptions used as part of the derivation: in the case of binary session types, the context maps expressions to sorts, and process variables to session types. A typing judgement is constructed in terms of inference rules defined inductively on the structure of processes and expressions.

We present the rules for \rulename{Ty-Sel} and \rulename{Ty-Bra}, the remaining rules follow from \cite{C406Lecture} and can be trivially defined as they leverage the syntactic similarities between session types and our session calculus.

%\[
%\infer[\mbox{\rulename{Ty-Bra}}]{
%	\Gamma \vdash \bra{\role{p}}{l_i(x_1 \dots x_n):P_i}_{i\in I} : \tybra{\role{p}}{l_i(U_1 \dots U_n):S_i}_{i\in I}
%	}{
%	\begin{array}{ccc}
%	\forall i \in I.
%	&
%	&
%    \Gamma, x_1: U_1, \dots, x_n: U_n \vdash P_i : S_i		
%	\end{array}
%}
%\]
%
%\[
%\infer[\mbox{\rulename{Ty-Sel}}]{
%	\Gamma \vdash \sel{\role{p}}{l}{e_1 \dots e_n}.P : \tysel{\role{p}}{l(U_1 \dots U_n): S}
%	}{
%	\begin{array}{cccc}
%	\Gamma \vdash e_1 : U_1
%	&
%	\dots
%	&
%	\Gamma \vdash e_n : U_n
%	&
%    \Gamma \vdash P : S
%	\end{array}
%}
%\]

% Guarantees for well-typed sessions expressed through duality, also notion of subtyping.
The definition of stuck processes from {\ref{section_async}} motivate the discussion of communication errors that may occur during interactions among participants. We outline two of the main classes of errors:

\begin{itemize}
\item \textbf{Deadlock}: Progress cannot be made when the two participants expect to be receiving a message from each other at the same time.
\item \textbf{Communication mismatch}: Progress cannot be made when the selection process sends a message with a label identifier not offered by the branching process; likewise, the payload sent must be compatible with the sort expected by the branching process for the selected branch.
\end{itemize}

Session types ensure that well-typed binary sessions are guaranteed to be free from these communication errors through the concept of {duality}. Duality defines a notion of {compatibility} between processes: two session types are dual with respect to each other if the communication between them (i.e. pairs of sending and receiving actions) always match (i.e. with respect to the selected label and message payload type). We define $\overline{S}$ as the dual type of $S$ in Table \ref{table:dual}.

% Row height in tables
\renewcommand{\arraystretch}{1.2}
\begin{center}
\begin{tabular}{rcl}
%$\overline{\mathbf{end}}$ & = & $\mathbf{end}$ \\
%$\overline{\tybra{\role{p}}{l_i(U_1 \dots U_n):S_i}_{i\in I}}$ & = & $\tysel{\role{q}}{l_i(U_1 \dots U_n):\overline{S_i}}_{i\in I}$ \\
%$\overline{\tysel{\role{p}}{l_i(U_1 \dots U_n):S_i}_{i\in I}}$ & = & $\tybra{\role{q}}{l_i(U_1 \dots U_n):\overline{S_i}}_{i\in I}$ \\
%$\overline{\mu \mathbf{t}. S}$ & = & $\mu t. \overline{S}$ \\
%$\overline{\mathbf{t}}$ & = & $\mathbf{t}$ \\ 
\end{tabular}
\captionof{table}{Duality of Binary Session Types involving participants $\role{p}$ and $\role{q}$}
\label{table:dual}
\end{center}
\renewcommand{\arraystretch}{1}

Consequently, a binary session is well-typed if the participating processes are typed to be dual with respect to each other: we illustrate this in \rulename{MTy}.

%\[
%\infer[\mbox{\rulename{MTy}}]{
%	\vdash \role{p} :: P \mid 
%	\role{q} :: Q
%}{
%	\begin{array}{lr}
%	\cdot \vdash P: S
%	&
%	\cdot \vdash Q: \overline{S}
%	\end{array}
%}
%\]

The definition of duality alone restricts the definition of well-typed binary sessions to those where the two processes are derived to be \textit{exactly} dual types of one another. Consider the pair of session types below:

%\[
%\begin{array}{rl}
%S_{\text{Client}} &= \tysel{\role{Server}}{\text{Succ}(\texttt{int}): \tybra{\role{Server}}{\text{Res}(\texttt{int}): \mathbf{end}}} \\
%S_{\text{Server}} &= \tybra{\role{Client}}{\text{Succ}(\texttt{int}): \tysel{\role{Client}}{\text{Res}(\texttt{int}): \mathbf{end}},~\text{Quit}(): \mathbf{end}}
%\end{array}
%\]

Whilst $\overline{S_{\text{Client}}} \neq S_{\text{Server}}$, this pair of session types is intuitively compatible as the client is selecting a branch offered by the server, where the session types for the continuations of this branch for both participants are indeed dual.

This motivates the concept of {subtypes}, which allows a process to be typed by its ``supertype'' when required. $\leqslant$\footnote{The $\leqslant$ operator is also an overloaded relation on sorts to express subsorting.} defines the subtyping relation: $S \leqslant S'$ reads \textit{$S$ is a subtype of $S'$}, and is defined coinductively on the structure of session types.

We present the inference rules for \rulename{Sub-Bra} and \rulename{Sub-Sel} inspired by \cite{MPST} but adapted for polyadic communication; the intuition behind subtyping and subsorting is outlined below:

\begin{itemize}
\item \textbf{Branching}: The supertype of a branching process offers a subset of the branches and expects more specific types of payload; intuitively, if a process expects to receive an \texttt{int}, it can handle a \texttt{nat} payload.
\item \textbf{Selection}: The supertype of a selection process offers a superset of the internal choices and can send more generic types of payload; intuitively, if a process sends a \texttt{nat}, the payload is compatible with receivers expecting a more generic \texttt{int} payload.
\end{itemize}

\begin{prooftree}
\AxiomC{$\forall i \in I.$}
\AxiomC{$U'_1 \leqslant U_1 \dots U'_n \leqslant U_n$}
\AxiomC{$S_i \leqslant S'_i$}
\RightLabel{\mbox{\rulename{Sub-Bra}}}
\doubleLine
\TrinaryInfC{$
	\tbra{\role{p}}{l_i(U_1 \dots U_n):S_i}{i\in I \cup J}
	\leqslant
	\tbra{\role{p}}{l_i(U'_1 \dots U'_n):S'_i}{i\in I}
$}
\end{prooftree}

\begin{prooftree}
\AxiomC{$\forall i \in I.$}
\AxiomC{$U_1 \leqslant U'_1 \dots U_n \leqslant U'_n$}
\AxiomC{$S_i \leqslant S'_i$}
\RightLabel{\mbox{\rulename{Sub-Sel}}}
\doubleLine
\TrinaryInfC{$
	\tsel{\role{p}}{l_i(U_1 \dots U_n):S_i}{i\in I}
	\leqslant
	\tsel{\role{p}}{l_i(U'_1 \dots U'_n):S'_i}{i\in I \cup J}
$}
\end{prooftree}

We also introduce subsumption in \rulename{Ty-Sub} to incorporate the subtyping relation into the typing judgement.

%\[
%\infer[\mbox{\rulename{Ty-Sub}}]{
%	\Gamma \vdash P : S'
%}{
%	\begin{array}{lr}
%	\Gamma \vdash P: S
%	&
%	S \leqslant S'
%	\end{array}
%}
%\]

This allows us to construct a derivation to show that the binary session

\[
\mathcal{M} = \role{Client} :: P_\text{Client}~\mid~\role{Server} :: P_\text{Server}
\]

is well-typed, assuming $P_\text{Client}$ and $P_\text{Server}$ are typed $S_\text{Client}$ and $S_\text{Server}$ respectively.

\begin{prooftree}
\AxiomC{\vdots}
\UnaryInfC{$\cdot \vdash P_\text{Client}: S_\text{Client}$}
\AxiomC{\vdots}
\UnaryInfC{$\cdot \vdash P_\text{Server}: S_\text{Server}$}
\AxiomC{\vdots}
\doubleLine
\UnaryInfC{$S_\text{Server} \leqslant \overline{S_\text{Client}}$}

%\AxiomC{$\cdot \vdash P_\text{Client}: S_\text{Client}$}
%\AxiomC{$\cdot \vdash P_\text{Server}: S_\text{Server}$}
%\AxiomC{$S_\text{Server} \leqslant \overline{S_\text{Client}}$}
\RightLabel{\rulename{Ty-Sub}}
\BinaryInfC{$\cdot \vdash P_\text{Server}: \overline{S_\text{Client}}$}
\RightLabel{\rulename{MTy}}
\BinaryInfC{$
	\vdash \role{Client} :: P_\text{Client} \mid 
	\role{Server} :: P_\text{Server}
$}
\end{prooftree}

\subsection{Multiparty Session Types}

% Motivate the extension from BST to MPST (example of battleship game, existing distributed protocols that involve multiple participants?)
Whilst binary session types provide communication guarantees between exactly 2 participants, distributed systems generally involve more parties in practice. This is equally relevant in interactive web applications, as motivated by the \textit{Battleships} game example in \cite{PureScript2019} where the server coordinates interactions between two players.

Whilst there is a natural syntactical extension to our session calculus for describing multiparty sessions\footnote{We also adopt the shorthand $\mathcal{M} ::= \prod^n_{i=1} \role{p_i} :: P_i$ as used in the literature.} as

\[
\begin{array}{rl}
\mathcal{M} ::= & \role{p_1} :: P_1~\mid~\role{p_2} :: P_2~\mid~\dots~\mid \role{p_n} :: P_n
\end{array}
\]

the same cannot be said for the binary session typing discipline, particularly with respect to duality. The same notion of duality does not extend to the decomposition of multiparty interactions into multiple binary sessions:  \cite{C406Lecture} and \cite{MPST} both present counterexamples of well-typed binary sessions that, when composed to represent a multiparty session, results in communication errors thus violating guarantees of well-typed sessions.

Honda et al. presents {multiparty session types} in \cite{MPAST} to extend the binary session typing discipline for sessions involving more than 2 participants, whilst redefining the notion of compatibility in this multiparty context. Multiparty session types are defined in terms a {global type}, which provides a bird's eye view of the communication protocol describing the interactions between pairs of participants. Figure \ref{fig:mpst} defines the syntax of global types inspired by \cite{MPST} and adapted to be compatible with our session calculus presented in {\ref{section_bst}}.

% Introduce global type, projection
\begin{figure}[!hb]
\doublespacing
\[
\begin{array}{rlr}

G ::= & & \text{Global Types} \\
     & \mathbf{end} & \text{Termination} \\
\mid & \role{p} \to \role{q}:\left\{l_i(U_1 \dots U_n).~G_i\right\}_{i \in I} & \text{Message Exchange} \\
\mid & \mu \mathbf{t}.~G & \text{Recursive Type} \\
\mid & \mathbf{t} & \text{Type Variable} \\
\end{array}
\]

\singlespacing
\caption{Syntax of Global Types}
\label{fig:mpst}
\end{figure}

To check the conformance of a participant's process against the protocol specification described by the global type, we {project} the global type onto the participant to obtain a session type that only preserves the interactions described by the global type that pertain to the participant. Projection is defined by the $\upharpoonright$ operator, more commonly seen in literature in its infix form as $G \upharpoonright \role{p}$ describing the projection of global type $G$ for participant $\role{p}$. Intuitively, the projected local type of a participant describes the protocol from the viewpoint of the participant.

More formally, projection can be interpreted as a \textit{partial function} ${\upharpoonright} :: G \times \role{p} \rightharpoonup S$, as the projection for a participant may be undefined for an ill-formed global type; \cite{C406Lecture} presents examples of where this is the case, and \cite{MPST} presents the formal definition of projection.

% How does the notion of compatibility (which used to be duality) adapt to MPST?
The notion of compatibility in multiparty session types is still captured by \rulename{MTy}, but adapted to consider the local projections for all participants as supposed to dual types in the binary case.

%\[
%\infer[\mbox{\rulename{MTy}}]{
%	{\displaystyle \vdash \prod^{}_{i \in I} \role{p_i} :: P_i ~: G} 
%}{
%	\begin{array}{llll}
%	\forall i \in I. 
%	&
%	\cdot \vdash P_i: G \upharpoonright \role{p_i}
%	&
%	&
%	\text{pt}(G) \subseteq \left\{\role{p_i} \mid i \in I \right\}
%	\end{array}
%}
%\]

For a multiparty session $\mathcal{M} = \prod_{i \in I}\role{p_i} :: P_i$ to be well-typed by a global type $G$, we require:

\begin{enumerate}
\item All participant processes $\role{p_i} :: P_i$ to be well-typed with respect to their corresponding well-defined projection $G \upharpoonright \role{p_i}$, and 
\item $G$ does not describe interactions with participants not defined in $\mathcal{M}$ 
\end{enumerate}

Well-typed multiparty session enjoys the following communication guarantees as outlined in \cite{GentleMPST}:

\begin{itemize}
\item \textbf{Communication safety}: The types of sent and expected messages will always match.
\item \textbf{Protocol fidelity}: The exchange of messages between processes will abide by the global protocol.
\item \textbf{Progress}: Messages sent by a process will be eventually received, and a process waiting for a message will eventually receive one; this also means there will not be any sent but unreceived messages.
\end{itemize}

This motivates an elegant, decentralised solution for checking protocol conformance in practice: once the global type for the protocol is defined, local processes can verify their implementation against their corresponding projection in isolation, independent of each other.

%Figure \ref{fig:mpst_workflow}\todo{Create figure} illustrates this diagrammatically.
%
%\begin{figure}[!h]
%\caption{Type checking with Multiparty Session Types}
%\label{fig:mpst_workflow}
%\end{figure}

\section{The Scribble Protocol Language}

%\subsection{Overview}
% Introduce Scribble as an implementation of MPST
Whilst session type theory represents the type language for concurrent processes, it also forms the theoretical basis of proposals introduced to implement session types for real-world application development: the Scribble language is one such proposal.

Scribble \cite{Scribble} is a platform-independent description language for the specification of message-passing protocols. The language describes the behaviour of communicating processes at a high level of abstraction: more importantly, the description is independent from implementation details in the same way that the type signature of a function declaration is decoupled from the corresponding function definition.

% Protocol specification: show parallels between Scribble protocol and MPST global type
A Scribble protocol specification describes an agreement of how participating systems, referred to as {roles}, interact. The protocol stipulates the sequence of structured messages exchanged between roles; each message is labelled with a name and the type of payload carried by the message.

We present an example of a Scribble protocol in Figure \ref{fig:adder_scr} adapted from \cite{Hybrid2016}. The protocol specifies an arithmetic web service offered by a server to a client, represented by roles $S$ and $C$ respectively. The client is permitted to either:

\begin{itemize}
\item Send two \texttt{int}s attached to an \texttt{Add} message, where the server will respond with an \texttt{int} in a message labelled \texttt{Res}, and the protocol recurses; or,
\item Send a \texttt{Quit} message, where the server will respond with a \texttt{Terminate} message and the protocol ends.
\end{itemize}

The platform-independent nature of Scribble can be observed from the \texttt{type} declaration on Line 1: the developer has the freedom to specify message payload formats and data types from the target language of the implementation - in this case, aliasing the built-in Java integer as \texttt{int} throughout the protocol.

\begin{figure}[!h]
\begin{lstlisting}[language=Scribble]
type <java> "java.lang.Integer" from "rt.jar" as int;

global protocol Adder(role C, role S) {
	choice at C {
		Add(int, int)	from C to S;
		Res(int)		from S to C;
		do Adder(C, S);
	} or {
		Quit()		from C to S;
		Terminate()	from S to C;	
	}
}
\end{lstlisting}
\caption{Adder Protocol in Scribble}
\label{fig:adder_scr}
\end{figure}

The simplicity of the protocol specification language reflects the design goals for Scribble, as outlined in \cite{Scribble}, to be easy to read, write and maintain, even for developers who are not accustomed to the formalities in protocol specification. Moreover, we clearly observe the parallels between the Scribble language and multiparty session type (MPST) theory, from the homomorphic mapping between Scribble roles and MPST participants to the syntactic similarities between the specification in Figure \ref{fig:adder_scr} and the global type below written in the calculus.

\[
\begin{array}{rl}
G = \mu\mathbf{t}.\role{C}\to\role{S}:\{
& \text{Add}(\texttt{int}, \texttt{int}): \role{S}\to\role{C}: \{\text{Res}(\texttt{int}): \mathbf{t}\}, \\
& \text{Quit}(): \role{S}\to\role{C}: \{\text{Terminate}(): \mathbf{end}\}\\
\} &
\end{array}
\]

\subsection{Endpoint Finite State Machines}
% Show parallels between Scribble protocol-to-EFSM and MPST global-to-local projections
The protocol specification language is a component of the broader Scribble project in \cite{Scribble}; through the project, Honda et al. also aims to facilitate the development of endpoint applications that conform to user-specified protocols.

A Scribble global protocol can be projected to a role to obtain the local specification of said protocol from the role's viewpoint. This is analogous to the standard algorithmic projections in multiparty session type theory: the projected local specification, or local protocol, only preserves the interactions in which the target role is involved. Any user-defined type declarations in the global protocol will also be preserved. This allows local roles, also referred to as {endpoints}, to verify their implementation against their local protocol for conformance, independent of other endpoints. The communication safety guarantees from MPST theory also apply here: if the implementation for each endpoint is verified against its local protocol, the multiparty system as a whole will conform to the global protocol.

% Formalise syntax and properties of EFSM derived from well-formed protocols
To facilitate the verification process, a local protocol is converted into 
an {endpoint finite state machine} (EFSM). An EFSM encodes the control flow of the local protocol, where an initial state is defined and each transition from some state to a successor state corresponds to a valid IO action (i.e. sending to or receiving from another role) permitted at the endpoint at that state. Figure \ref{fig:efsmsyntax} presents the EFSM formalism adopted from \cite{Hybrid2016}.

\begin{figure}[!hb]
\doublespacing
\[
\begin{array}{rlr}

\text{EFSM} ::= & \mathbb{R} \times \mathbb{L} \times \mathbb{T} \times \Sigma \times \mathbb{S} \times \delta & \text{Endpoint FSM} \\

\mathbb{R} ::= & r,~r',~\dots & \text{Role Identifiers} \\

\mathbb{L} ::= & l,~l',~\dots & \text{Message Label Identifiers} \\

\mathbb{T} ::= & \texttt{int},~\texttt{bool},~T,~T',~\dots & \text{Payload Format Types} \\

\Sigma ::= & & \text{Actions} \\
     & r!l(\tilde{T}) \quad \text{where } \tilde{T} \subseteq \mathbb{T} & \text{Output} \\
\mid & r?l(\tilde{T}) \quad \text{where } \tilde{T} \subseteq \mathbb{T} & \text{Input} \\

\mathbb{S} ::= & S,~S',\dots & \text{State Identifiers} \\

\delta ::= & \mathbb{S} \times \Sigma \rightharpoonup \mathbb{S} & \text{State Transition Function} \\

\end{array}
\]
\singlespacing
\captionof{figure}{Syntax for Endpoint FSM}
\label{fig:efsmsyntax}
\end{figure}

We highlight the semantics of the {state transition function}. Its domain defines the set of permitted actions for each state, and the range defines the successor state. Crucially, this is a {partial function} because the EFSM restricts what can be done (in terms of send and receive actions, and what label/payload types can be sent and received).

\begin{figure}[!h]
\doublespacing
\[
\begin{array}{rl}

\texttt{initial}_{EFSM}(S) \iff & \nexists S' \in \mathbb{S},\alpha \in \Sigma.~\delta(S',\alpha) = S \\
\texttt{terminal}_{EFSM}(S) \iff & \delta(S) = \emptyset \\
\texttt{output}_{EFSM}(S) \iff & \delta(S) = \{\,\alpha \in \Sigma \mid \exists l \in \mathbb{L}, \tilde{T} \subseteq{\mathbb{T}}.~\alpha = r!l(\tilde{T})\,\}~\text{for some}~r \in \mathbb{R} \\
\texttt{input}_{EFSM}(S) \iff & \delta(S) = \{\,\alpha \in \Sigma \mid \exists l \in \mathbb{L}, \tilde{T} \subseteq{\mathbb{T}}.~\alpha = r?l(\tilde{T})\,\}~\text{for some}~r \in \mathbb{R}
\end{array}
\]
\singlespacing
\captionof{figure}{Types of EFSM States}
\label{fig:efsmstates}
\end{figure}

We adopt the properties of EFSMs for well-formed protocol specifications from \cite{Hybrid2016} \textbf{(1)} there is exactly one initial state, \textbf{(2)} there is at most one terminal state, and \textbf{(3)} every non-terminal state is either an input state or output state. There are further restrictions on input and output states, namely these states must only send to or receive from the same role. We formalise these properties in Figure \ref{fig:efsmstates}.

\section{Code Generation}
The EFSM describes the local session type and provides guidance to developers for verifying that their endpoint implementation conforms to the communication protocol. However, a direct encoding of the local session type into the target language of the implementation is usually not feasible as the EFSM assumes IO objects as first-class citizens and communication channels are linear resources, features that are left to be desired in mainstream programming languages (such as Java and Python) used by the developers' implementations. We note that languages with native session type support (e.g. ATS \cite{ATS2016}) exist, but their usage largely remains for research purposes as supposed to real-world application development.

Code generation is a common approach for verifying implementations written in the aforementioned mainstream languages against the EFSM. Approaches in the literature differ by how they leverage features in the target language (such as the type system), but generally define some interpretation of the EFSM in the target language and generate APIs which the developer can use to implement a target application that guarantees the following two properties:

\begin{itemize}
\item \textbf{Behavioural typing}: The execution trace of messages sent and received by the application is accepted by the EFSM.
\item \textbf{Channel linearity}: Each transition in the EFSM represents a channel resource. When the application transitions from some state $S$ to some successor state $S'$, it must no longer be able to access a reference (e.g. have an alias) to $S$.
\end{itemize}

We outline the different existing approaches and summarise how they verify the aforementioned properties in {\ref{section:codegencompare}}.

\subsection{Runtime Monitors}
% What
Neykova and Yoshida targeted the MPST methodology for Python programs in \cite{Python2017} and proposed to generate {runtime monitors} from the EFSM. These monitors expose APIs for sending and receiving messages, which is used by the developer in their implementation. The runtime monitor is an abstraction between the developer's implementation and the actual communication channel, and ``executes'' the EFSM internally to ensure protocol conformance. When the developer sends a message (with some label and payload) using the API, the runtime monitor checks whether this send action conforms to the current EFSM state, and if so, performs the send and advances to the successor state. Likewise, when the developer invokes a receive, the runtime monitor verifies that this is permitted at the current EFSM state before returning the received payload.

We observe that this approach complements the dynamic typing nature of the Python language, which makes it sensible to perform behavioural typing at runtime. As the send and receive IO primitives are made available to the developer, there are no ``instances'' of channel resources created, so the developer cannot explicitly hold a reference to some state in the EFSM (let alone keep aliases), so channel linearity is trivially guaranteed here.

\subsection{Type-Level Encoding}
King et al. presented an approach in \cite{PureScript2019} for integrating session types into web development using the PureScript language, which takes advantage of its expressive type system to provide static guarantees on implementation conformance with respect to the protocol. We outline the main components of their EFSM encoding:

\paragraph{Actions as type classes} The semantics of the state transition function in the EFSM formalism express that a tuple of state-action \textit{uniquely defines} a successor state. These semantics can be expressed by \textit{multi-parameter type classes (MPTC) with functional dependencies}: \texttt{class Send r s t a | s -> t r a} defines \texttt{Send} as a MPTC parameterised by recipient \texttt{r}, current state \texttt{s}, successor state \texttt{t} and payload type \texttt{a}, and \texttt{s -> t r a} expresses the functional dependency that, for an instance of this type class, the current state uniquely determines the successor state, the recipient and payload type. These type classes are independent of the EFSM.

\paragraph{Transitions as instances of type classes} By encoding states as data types, valid EFSM transitions are encoded as \textit{instances} of the type classes. If \texttt{S1} is an output state, sending an \texttt{Int} to \texttt{Svr} with successor \texttt{S2}, we would encode \texttt{instance SendS1 :: Send Svr S1 S2 Int}. Because of the functional dependency, the developer cannot instantiate an invalid transition (e.g. \texttt{Send Svr S1 S3 Bool}, since \texttt{S1} uniquely determines the other type parameters.

This proposal is relevant to the problem we are tackling, and we appreciate that the intricacy of the library design in the communication combinators to conceal the channel resources is something we can build upon in our solution. We also observe the challenges of applying session types to the front-end environment, as shown by the careful choice made in \cite{PureScript2019} to use the \textit{Concur UI} framework which builds UI elements sequentially to model sequential sessions; not doing so would require binding channel resources into event listeners on UI elements, which makes linearity violation possible (e.g. by binding a send transition to a button click event, the user might click the button twice, thus reusing the channel).

\subsection{Hybrid Session Verification}
Whilst runtime monitors and type-level encoding achieve communication safety guarantees via dynamic and static approaches respectively, Hu and Yoshida implemented a workflow in \cite{Hybrid2016} that performs hybrid verification of communication protocol conformance for Java applications. We highlight the two main components below:

\paragraph{Static session typing} States are represented as classes; supported transitions on each state are represented as instance methods, parameterised by the role, label and payload involved in the message exchange. A send method takes the payload to send as a parameter, whilst a receive method is a blocking call that requires the caller to allocate a \texttt{Buf<T>} wrapper on the stack (where \texttt{T} is the expected payload type), then the receive method populates the payload into the wrapper and returns upon receiving from the channel. These instance methods return a new instance of the successor state class.

\paragraph{Runtime linear channel usage checks} Each state keeps track of its usage in a private boolean flag and throws an exception when the instance method is called twice, which signifies channel reuse and a linearity violation. Similarly, the \texttt{SessionEndpoint} class keeps track of whether the connection is open or close, and throws an exception when program execution exits the scope of the session endpoint and a terminal state has yet to be reached, signifying a protocol violation as the session hasn't completed yet but is now out of scope.

We find that this proposal strikes a good balance between maximising static communication safety guarantees whilst providing an intuitive set of APIs for developers to efficiently write their applications. For example, the encoding of transitions as instance methods that return new instances of the successor state exposes the channel resource, and since Java does not build support for linear resources and does not monitor aliasing of variables, linear channel usage is monitored dynamically. However, this complements the imperative style of application code written in Java, and takes advantage of the type system to statically enforce valid transitions in the code.

\subsection{Comparison}
\label{section:codegencompare}

We compare how these existing code generation approaches provide communication safety guarantees in Table \ref{table:comparison}.

\begin{figure}[!h]
\centering
\begin{tabular}{l || p{0.35\textwidth} | p{0.35\textwidth}}
%\begin{tabular}{l||l|l}
Language & Behavioural typing & Channel linearity \\
\hline\hline
Python \cite{Python2017} & Dynamically enforced - runtime monitor represent the EFSM and only execute supported transitions. & Trivially guaranteed - channel resources not exposed, developer uses \texttt{send()} and \texttt{receive()} primitives. \\
\hline
PureScript \cite{PureScript2019} & Type-level encoding - EFSM states are encoded as types, IO actions are encoded as multi-parameter type classes (to express the semantics of the state transition function) and EFSM transitions are encoded as instances of said type classes. & Statically guaranteed - channels are not exposed in the provided combinators to prevent reuse, and session constructor requires a \texttt{Session} continuation parameterised by initial and terminal state to prevent incomplete sessions.  \\
\hline
Java \cite{Hybrid2016} & Statically guaranteed - states are encoded as classes; transitions are encoded as instance methods on the state class and return a new instance of the successor state class. & Checked at runtime - states keep \texttt{used} boolean flag to detect and prevent reuse, Session API implements \texttt{AutoCloseable} interface to be used in resource try-catch block to prevent unused.
\end{tabular}
\captionof{table}{Comparison between existing MPST code generation approaches.}
\label{table:comparison}
\end{figure}

### 2020-06-15 report/sections/background/interim/codegen.tex
Progress: start moving background

--- 2020-01-23 interim/sections/background/codegen.tex
@@
 However, a direct encoding of the local session type into the target language of the implementation is usually not feasible as the EFSM assumes IO objects as first-class citizens and communication channels are linear resources, features that are left to be desired in mainstream programming languages (such as Java and Python) used by the developers' implementations.
-We note that languages with native session type support \cite{ATS2016} do exist, but their usage largely remains for research purposes as supposed to real-world application development.
+We note that languages with native session type support (e.g.
+ATS \cite{ATS2016}) exist, but their usage largely remains for research purposes as supposed to real-world application development.
 

### 2020-06-15 report/sections/background/interim/typescript.tex
Progress: start moving background

\section{TypeScript}

\subsection{Overview}
\begin{itemize}
\item Background
\item Aims
\item Example code snippet
\end{itemize}

\subsection{Gradual Type System}
\begin{itemize}
\item Introduce static (Java), dynamic (Python/JS)
\item Use pros and cons of the former two systems to motivate the concept of a gradual type system
\item Shwo example of unsoundness
\end{itemize}

\subsection{Language Features}
\begin{itemize}
\item Outline features that are either (a) relevant to implementing web applications, (b) differ from the features prevalent in traditional OO (Java) or functional (PureScript) languages
\end{itemize}

\subsubsection{Promises}
\textit{Relevant for modelling `state-specific input futures' from the hybrid Java paper?}

\subsubsection{Discriminated Union}
\textit{Useful for dealing with the switch case approach for branch receives?}

\subsubsection{Index Types}
\textit{With index types, you can get the compiler to check code that uses dynamic property names.}

\subsubsection{Conditional Types}
\textit{Compare this with the type dependencies achieved in the PureScript paper?}

### 2020-06-15 report/sections/background/related.tex
Progress: start moving background

\section{Related Work}
\label{section:bgrelated}

\subsection{Scribble and Endpoint API Generation}
\label{subsection:bgscribble}

\subsection{Session Types in Web Development}
\label{subsection:sessiontypewebdev}

### 2020-06-15 report/sections/background/sessiontypes.tex
Progress: start moving background

\section{Session Types}

Web applications are one of many examples of 
distributed systems in practice. 
Distributed systems are built upon the interaction 
between concurrent processes, which can be implemented using 
the two main communication abstractions in 
\textit{shared memory} and \textit{message passing}.

Shared memory provides processes with the impression of 
a logical single large monolithic memory 
but requires programmers to understand consistency models 
in order to correctly reason about the consistency of shared state.

Message passing interprets the interaction between processes 
as the exchange of messages.
This best describes the communication transports 
found in web applications, ranging from the 
stateless request-response client-server interactions via HTTP 
to full-duplex communication channels via the WebSocket protocol 
\cite{WebSocketRFC}.

The process algebra $\pi$-calculus 
introduced by Milner in \cite{Milner1999} 
formalises the message passing abstraction in terms of 
the basic building blocks of sending and receiving processes,
along with inductively defined continuation processes. 
The composition of these primitives allow us to 
describe more complex communication sessions.
\textit{Session types} define the typing discipline 
for the $\pi$-calculus and provide reliability guarantees 
for communication sessions; 
the latter addresses a key challenge when 
reasoning about the correctness of distributed systems.

Practical application of session types in software
engineering range from
developing languages providing native session type support 
\cite{ATS2016} to implementing session types 
in existing programming languages across different paradigms.
Implementations of the latter approach differ by 
how they leverage the design philosophy and features 
provided by the programming language.
For example, King et al. leveraged the 
expressive type system of PureScript to perform 
\textit{static} session type checking in \cite{PureScript2019},
whilst Neykova and Yoshida introduced dynamic approaches to 
check the conformance of Python programs 
with respect to session types in \cite{Python2017}.
We discuss these related work, among others, in \cref{section:bgrelated}.

\subsection{Process Calculus}
\label{subsection:picalculus}

The $\pi$-calculus models concurrent computation, 
where processes can execute in parallel and communicate 
via shared names.
We first consider the \textit{asynchronous} 
$\pi$-calculus introduced by 
Honda and Tokoro in \cite{AsyncHonda}.
Among the many flavours of the calculus 
which vary depending on the application domain,
we outline the variant as presented in \cite{C406Lecture}.

We define the syntax of processes in \cref{fig:async};
the asynchrony comes from the lack of 
continuation in the output process.

\begin{itemize}
\item $\mathbf{0}$ is the nil process and represents inactivity.
\item $\pout{u}{v}$ is the output process that will send value $v$ on $u$.
\item $\pin{u}{x}.P$ is the input process that,
 upon receiving a message on $u$, 
 will bind the message to $x$ and 
 carry on executing $P$ under this binding.
\item $P\mid Q$ represents the parallel composition of 
processes executing simultaneously.
\item $!P$ represents the parallel composition of 
\textit{infinite} instances of $P$; 
more specifically, $!P \equiv P \mid {!P}$.
\item $(\nu a)~P$ represents a name restriction 
where any occurrence of $a$ in $P$ is local
 and will not interfere with other names outside the scope of $P$.
\end{itemize}

\begin{figure}[!hb]
\doublespacing
\[
\begin{array}{rlr}

P,Q ::= & & \text{Processes} \\
     & \mathbf{0} & \text{Nil Process} \\
\mid & \pout{u}{v} & \text{Output} \\
\mid & \pin{u}{x}.P & \text{Input} \\     
\mid & P \mid Q & \text{Parallel Composition} \\
\mid & !P & \text{Replication} \\
\mid & (\nu a)~P & \text{Restriction} \\

u,v ::= & & \text{Identifiers} \\
     & a, b, c,~\dots & \text{Names} \\
\mid & x, y, z,~\dots & \text{Variables} \\

\end{array}
\]
\singlespacing
\captionof{figure}{Syntax of Asynchronous $\pi$-calculus}
\label{fig:async}
\end{figure}

The operational semantics model the interaction 
between parallel processes. 
Whilst \cite{C406Lecture} presents the full operational semantics, 
we highlight the \rulename{Comm} reduction rule which 
specifically models message passing:
if the parallel composition of an input process and output process 
share the same name, the composition reduces to the 
continuation of the input process, 
substituting the variable $x$ with the message received $v$. 
We omit the definitions of substitution, free variables and free names, 
$\alpha$-equivalence and structural congruence; 
the interested reader may refer to \cite{C406Lecture}.

\begin{prooftree}
\AxiomC{}
\RightLabel{\rulename{Comm}}
\UnaryInfC{$
\pout{a}{v} \mid \pin{a}{x}.P ~\longrightarrow ~ P[v/x]
$}
\end{prooftree}

We additionally define a process $P$ to be {stuck} 
if $P$ is not the nil process and $P$ cannot be reduced any further. 
For example, the process 
$P = \pin{a}{x}.\mathbf{0} \mid \pout{b}{v}$ 
is stuck as the parallel composition of an input process 
and an output process that do not share the same name cannot 
be reduced using \rulename{Comm}. 
In practice, a stuck process contains communications 
that will never be executed.

\subsection{Binary Session Types}
\label{subsection:bgbst}

A \textit{binary session} is a parallel composition 
of two processes, each representing a distinct participant.
In the context of web applications, a binary session may
describe the interactions between client and server.
Without loss of generality, a \text{session} represents 
the sequence of send and receive actions of a single participant.

We introduce a \text{synchronous} session calculus 
in \cref{fig:sync},
inspired by \cite{MPST}. 
We briefly discuss the main components and 
how it differs from the variant introduced in \cite{C406Lecture}:

\begin{itemize}
\item \textbf{Synchronous communication}: 
Output processes have a continuation that will be executed upon
a successful send.

\item \textbf{Polyadic communication}: 
More than one values can be communicated at once.
We refer to these as a \textit{vector} of values.

\item \textbf{Branching and selection}: 
A branching process can offer a set of branches, 
each defined by its own label identifier and continuation process. 
A selection process can select a branch by 
sending the corresponding label identifier 
alongside the payload to the branching process.

\item \textbf{Labelled messages}: 
A label identifier is attached to all messages; 
the input process in \cref{fig:async}
is generalised as a branching process that offers one branch.
\end{itemize}

The \rulename{Comm} rule in the operational semantics 
for this calculus exemplifies these new additions: 
given a binary session between distinct participants 
$\role{p}$ and $\role{q}$ 
where $\role{q}$ offers a set of labelled branches,
if $\role{p}$ selects a label offered by $\role{q}$ and 
sends a vector of expressions $e_1, \dots, e_n$ 
that evaluate\footnote{
We adopt the operational semantics for 
expression evaluation $e \downarrow v$ 
as defined in \cite{C406Lecture}.
} to the corresponding vector of values 
$v_1, \dots, v_n$, 
the session reduces to a session with 
the continuation from the selection process 
composed with the continuation from the selected branch 
of the branching process.
The branching process binds the received values
$v_1, \dots, v_n$ to the variables $x_1, \dots, x_n$.

\begin{prooftree}
\AxiomC{$\exists	j \in I. l_j = l$}
\AxiomC{$e_1 \downarrow v_1 \dots e_n \downarrow v_n$}
\AxiomC{$\mrole{p} \neq \mrole{q}$}
\RightLabel{\rulename{Comm}}
\TrinaryInfC{$
\mrole{p} :: \sel{q}{l}{e_1 \dots e_n}.~P \mid 
\mrole{q} :: \bra{p}{l_i(x_1 \dots x_n): Q_i}{i\in I} 
~\longrightarrow~ 
\ptprocess{p}{P} \mid \ptprocess{q}{Q_j[v_k/x_k]^n_{k=1}}
$}
\end{prooftree}

\begin{figure}[!hb]
\doublespacing
\[
\begin{array}{rlr}
v ::= & \underline{n}~\mid~\texttt{true}~\mid~\texttt{false} 
	& \text{Values} \\
e, e' ::= & & \text{Expressions} \\
	& v & \text{Values} \\
\mid	 & x & \text{Variables} \\
\mid & e + e'~\mid~e - e' & \text{Arithmetic Operators} \\
\mid & e = e'~\mid~e < e' ~\mid~e > e' & \text{Relational Operators} \\
\mid & e \wedge e'~\mid~e \vee e' ~\mid~\neg e & \text{Logical Operators} \\
\mid & e \oplus e' & \text{Non-Determinism} \\

\mrole{p} ::= & \mrole{Client},~\mrole{Server} & \text{Participant} \\

P,Q ::= & & \text{Processes} \\
     & \mathbf{0} & \text{Nil Process} \\
\mid & \sel{p}{l}{e_1 \dots e_n}.~P & \text{Selection} \\
\mid & \bra{p}{l_i(x_1 \dots x_n): P_i}{i\in I} & \text{Branching} \\
\mid & \pcond{e}{P}{Q} & \text{Conditional} \\
\mid & \precursion{P} & \text{Recursive Process} \\
\mid & \precvar & \text{Process Variable} \\

l, l' ::= & \texttt{``str''} & \text{Label Identifiers} \\

\mathcal{M} ::= & \ptprocess{p}{P} ~\mid~ \ptprocess{q}{Q} 
	& \text{Binary Session} \\
\end{array}
\]
\singlespacing
\captionof{figure}{Syntax of Session Calculus with 
Branching, Selection and Recursion}
\label{fig:sync}
\end{figure}

Additionally, the calculus introduces:

\begin{itemize}
\item \textbf{Conditionals}: 
If $e \downarrow \texttt{true}$, 
the process $\pcond{e}{P}{Q}$ reduces to $P$; 
if $e \downarrow \texttt{false}$, 
the process $\pcond{e}{P}{Q}$ reduces to $Q$.

\item \textbf{Recursion}: 
Following the \textit{equirecursive} approach, 
the occurrence of the process variable $\precvar$ 
in the recursive process can be 
expanded into the process transparently;
more specifically, $\precursion{P} \equiv P[(\precursion{P})/X]$.
\end{itemize}

{Session types} represent the type theory for our session calculus.
We define the syntax of session types for binary sessions in
\cref{fig:bst}.

\begin{figure}[!hb]
\doublespacing
\[
\begin{array}{rlr}

U ::= & \text{\tt{int}}~\mid~\text{\tt{bool}} & \text{Sorts} \\

T ::= & & \text{Session Types} \\
     & \mathbf{end} & \text{Termination} \\
\mid & \tbra{\role{p}}{l_i(U_1 \dots U_n):T_i}{i\in I} & \text{Branching} \\
\mid & \tsel{\role{p}}{l_i(U_1 \dots U_n):T_i}{i\in I} & \text{Selection} \\
\mid & \trec{T} & \text{Recursive Type} \\
\mid & \trecvar & \text{Type Variable} \\
\end{array}
\]

\singlespacing
\caption{Syntax of Session Types}
\label{fig:bst}
\end{figure}

We derive the type of a process with a {typing judgement} 
of the form $\Gamma \vdash P: S$, which reads, 
\textit{under the typing context $\Gamma$, 
process $P$ has session type $S$}.

The \textit{typing context} records typing assumptions 
used as part of the derivation: in the case of binary session types, 
the context maps expressions to sorts, 
and process variables to session types. 
A typing judgement is constructed in terms of inference rules 
defined inductively on the structure of 
processes and expressions.

We present the rules for \rulename{Ty-Sel} and \rulename{Ty-Bra}; 
the remaining rules follow from \cite{C406Lecture} 
and can be trivially defined as 
they leverage the syntactic similarities between 
session types and our session calculus.

\begin{prooftree}
\AxiomC{$
\Gamma \vdash e_1 : U_1 \dots \Gamma \vdash e_n : U_n
$}
\AxiomC{$\Gamma \vdash P : T$}
\RightLabel{\rulename{Ty-Sel}}
\BinaryInfC{$
\Gamma \vdash \sel{p}{l}{e_1 \dots e_n}.~P : 
\tselone{p}{l(U_1 \dots U_n)}: T
$}
\end{prooftree}

\begin{prooftree}
\AxiomC{$
\forall i \in I. ~ \left(
\Gamma, x_1: U_1, \dots, x_n: U_n \vdash P_i : T_i
\right)
$}
\RightLabel{\rulename{Ty-Bra}}
\UnaryInfC{$
\Gamma \vdash \bra{p}{l_i(x_1 \dots x_n): P_i}{i\in I} : 
\tbra{p}{l_i(U_1 \dots U_n):S_i}{i\in I}
$}
\end{prooftree}

The definition of stuck processes 
from \cref{subsection:picalculus} 
motivate the discussion of communication errors 
that may occur during interactions among participants. 
We outline two of the main classes of errors:

\begin{itemize}
\item \textbf{Deadlock}: 
Progress cannot be made when the two participants 
expect to be receiving a message from each other at the same time.
\item \textbf{Communication mismatch}: 
Progress cannot be made when the selection process 
sends a message with a label identifier not 
offered by the branching process; 
likewise, the payload sent must be compatible 
with the sort expected by the branching process 
for the selected branch.
\end{itemize}

Session types ensure that 
well-typed binary sessions are guaranteed 
to be free from these communication errors 
through the concept of \textit{duality}. 
Duality defines a notion of \textit{compatibility}
 between processes: two session types are dual 
 with respect to each other if the communication 
 between them (i.e. pairs of sending and receiving actions)
 always match (i.e. with respect to the selected label 
 and message payload type). 
We define $\dual{S}$ as the dual type of $S$ in \cref{table:dual};
message payload types are omitted for brevity.

% Row height in tables
\renewcommand{\arraystretch}{1.6}
\begin{center}
\begin{tabular}{rcl}
$\dual{\tend}$ & = & $\tend$ \\
$\dual{\tbra{p}{l_i: T_i}{i\in I}}$ & = & 
	$\tsel{q}{l_i: \dual{T_i}}{i\in I}$ \\
$\dual{\tsel{p}{l_i: T_i}{i\in I}}$ & = & 
	$\tbra{q}{l_i: \dual{T_i}}{i\in I}$ \\
$\dual{\trec{T}}$ & = & $\trec{\dual{T}}$ \\
$\dual{\trecvar}$ & = & $\trecvar$ \\ 
\end{tabular}
\captionof{table}{Duality of Binary Session Types involving participants $\role{p}$ and $\role{q}$}
\label{table:dual}
\end{center}
\renewcommand{\arraystretch}{1}

Consequently, a binary session is well-typed 
if the participating processes are typed to be 
dual with respect to each other: 
we illustrate this in \rulename{MTy}.

\begin{prooftree}
\AxiomC{$
\cdot \vdash P : T
$}
\AxiomC{$
\cdot \vdash Q : \dual{T}
$}
\RightLabel{\rulename{MTy}}
\BinaryInfC{$
\vdash \ptprocess{p}{P} \mid \ptprocess{q}{Q}
$}
\end{prooftree}

The definition of duality alone 
restricts the definition of 
well-typed binary sessions to those
where the two processes are derived to be 
\textit{exactly} dual types of one another. 
Consider the pair of session types below:

\[
\begin{array}{rl}
T_{\text{Client}} &= 
	\tselone{Server}{\tmsg{Succ(nat)}}.~
	\tbraone{Server}{\tmsg{Succ(int)}}.~
	\tend \\
T_{\text{Server}} &= \tbraone{Client}{
\begin{cases}
	\tmsg{Succ(int)}: \tselone{Client}{\tmsg{Succ(int)}} \\
	\tmsg{Quit()}:  \tend \\
\end{cases}
} \\
\end{array}
\]

Whilst $\dual{T_{\text{Client}}} \neq T_{\text{Server}}$, 
this pair of session types is intuitively compatible 
as the client is selecting a branch offered by the server, 
where the session types for the continuations of 
this branch for both participants are indeed dual.
Regarding payload, the server is expecting \texttt{int},
but the client sends \texttt{nat}, which is a subtype of \texttt{int}.

This motivates the concept of \textit{subtypes} in session types,
which allows a process to be typed by its ``supertype'' when required. 
$\subtype$\footnote{
The $\subtype$ operator is also an 
overloaded relation on 
sorts to express subsorting, 
i.e. $\texttt{nat} \subtype \texttt{int}$.
} defines the subtyping relation: 
$T \subtype T'$ reads \textit{$T$ is a subtype of $T'$},
and is defined coinductively on the structure of session types.

We present the inference rules for 
\rulename{Sub-Sel} and \rulename{Sub-Bra}
inspired by \cite{MPST} but adapted for polyadic communication; 
the intuition behind subtyping and subsorting is outlined below:

\begin{itemize}
\item \textbf{Branching}: 
The supertype of a branching process offers 
a subset of the branches and 
expects more specific types of payload; 
intuitively, if a process expects to 
receive an \texttt{int}, 
it can handle a \texttt{nat} payload.

\item \textbf{Selection}: 
The supertype of a selection process offers a 
superset of the internal choices and 
can send more generic types of payload; 
intuitively, if a process sends a \texttt{nat}, 
the payload is compatible with receivers expecting 
a more generic \texttt{int} payload.
\end{itemize}

\begin{prooftree}
\AxiomC{$\forall i \in I.$}
\AxiomC{$U'_1 \subtype U_1 ~ \dots ~ U'_n \subtype U_n$}
\AxiomC{$S_i \subtype T'_i$}
\RightLabel{\rulename{Sub-Bra}}
\doubleLine
\TrinaryInfC{$
	\tbra{p}{l_i(U_1 \dots U_n):T_i}{i\in I \cup J}
	\subtype
	\tbra{p}{l_i(U'_1 \dots U'_n):T'_i}{i\in I}
$}
\end{prooftree}

\begin{prooftree}
\AxiomC{$\forall i \in I.$}
\AxiomC{$U_1 \subtype U'_1 ~ \dots ~ U_n \subtype U'_n$}
\AxiomC{$T_i \subtype T'_i$}
\RightLabel{\rulename{Sub-Sel}}
\doubleLine
\TrinaryInfC{$
	\tsel{p}{l_i(U_1 \dots U_n):T_i}{i\in I}
	\subtype
	\tsel{p}{l_i(U'_1 \dots U'_n):T'_i}{i\in I \cup J}
$}
\end{prooftree}

We also introduce \textit{subsumption} in \rulename{Ty-Sub} 
to incorporate the subtyping relation into the typing judgement.

\begin{prooftree}
\AxiomC{$\Gamma \vdash P: T$}
\AxiomC{$T \subtype T'$}
\RightLabel{\rulename{Ty-Sub}}
\BinaryInfC{$\Gamma \vdash P : T'$}
\end{prooftree}

This allows us to construct a derivation to show that the binary session

\[
\mathcal{M} = \ptprocess{Client}{P_\text{Client}}
~ \mid ~
\ptprocess{Server}{P_\text{Server}}
\]

is well-typed, assuming $P_\text{Client}$ and $P_\text{Server}$ 
are typed 
$T_\text{Client}$ and $T_\text{Server}$ respectively.

\begin{prooftree}
\AxiomC{\vdots}
\UnaryInfC{$\cdot \vdash P_\text{Client}: T_\text{Client}$}
\AxiomC{\vdots}
\UnaryInfC{$\cdot \vdash P_\text{Server}: T_\text{Server}$}
\AxiomC{\vdots}
\doubleLine
\UnaryInfC{$T_\text{Server} \subtype \dual{T_\text{Client}}$}
\RightLabel{\rulename{Ty-Sub}}
\BinaryInfC{$\cdot \vdash P_\text{Server}: \dual{T_\text{Client}}$}
\RightLabel{\rulename{MTy}}
\BinaryInfC{$
	\vdash \ptprocess{Client}{P_\text{Client}} ~ \mid ~
	\ptprocess{Server}{P_\text{Server}}
$}
\end{prooftree}

\subsection{Multiparty Session Types}
\label{subsection:bgmpst}

% need LTS semantics here too

### 2020-06-16 report/sections/background/sessiontypes.tex
Report: minor background fix

@@
 \item \textbf{Polyadic communication}: 
-More than one values can be communicated at once.
-We refer to these as a \textit{vector} of values.
+More than one value can be communicated at once.
+We refer to these as a \textit{vector} of values, $\vec v$.
 
@@
 \AxiomC{$\exists	j \in I. l_j = l$}
+\AxiomC{$\vec x_j = x_1, \dots, x_n$}
 \AxiomC{$e_1 \downarrow v_1 \dots e_n \downarrow v_n$}
@@
 \RightLabel{\rulename{Comm}}
-\TrinaryInfC{$
+\QuaternaryInfC{$
 \mrole{p} :: \sel{q}{l}{e_1 \dots e_n}.~P \mid 
-\mrole{q} :: \bra{p}{l_i(x_1 \dots x_n): Q_i}{i\in I} 
+\mrole{q} :: \bra{p}{l_i(\vec x_i): Q_i}{i\in I} 
 ~\longrightarrow~ 
@@
      & \mathbf{0} & \text{Nil Process} \\
-\mid & \sel{p}{l}{e_1 \dots e_n}.~P & \text{Selection} \\
-\mid & \bra{p}{l_i(x_1 \dots x_n): P_i}{i\in I} & \text{Branching} \\
+\mid & \sel{p}{l}{\vec e}.~P & \text{Selection} \\
+\mid & \bra{p}{l_i(\vec x_i): P_i}{i\in I} & \text{Branching} \\
 \mid & \pcond{e}{P}{Q} & \text{Conditional} \\
@@
 We define the syntax of session types for binary sessions in
-\cref{fig:bst}.
+\cref{fig:bst}. $\vec U$ denotes a vector of sorts.
 
@@
 T ::= & & \text{Session Types} \\
-     & \mathbf{end} & \text{Termination} \\
-\mid & \tbra{\role{p}}{l_i(U_1 \dots U_n):T_i}{i\in I} & \text{Branching} \\
-\mid & \tsel{\role{p}}{l_i(U_1 \dots U_n):T_i}{i\in I} & \text{Selection} \\
+     & \tend & \text{Termination} \\
+\mid & \tbra{p}{l_i(\vec U_i):T_i}{i\in I} & \text{Branching} \\
+\mid & \tsel{p}{l_i(\vec U_i):T_i}{i\in I} & \text{Selection} \\
 \mid & \trec{T} & \text{Recursive Type} \\
@@
 \AxiomC{$
-\Gamma \vdash e_1 : U_1 \dots \Gamma \vdash e_n : U_n
+\Gamma \vdash e_1 : U_1 ~ \dots ~ \Gamma \vdash e_n : U_n
 $}
@@
 \Gamma \vdash \sel{p}{l}{e_1 \dots e_n}.~P : 
-\tselone{p}{l(U_1 \dots U_n)}: T
+\tselone{p}{l(U_1, \dots, U_n)}: T
 $}
@@
 \AxiomC{$
-\forall i \in I. ~ \left(
+\forall i \in I \left(
+\vec x_i = x_1, \dots, x_n
+\qquad
+\vec U_i = U_1, \dots, U_n
+\qquad
 \Gamma, x_1: U_1, \dots, x_n: U_n \vdash P_i : T_i
-\right)
-$}
+\right)$}
 \RightLabel{\rulename{Ty-Bra}}
 \UnaryInfC{$
-\Gamma \vdash \bra{p}{l_i(x_1 \dots x_n): P_i}{i\in I} : 
-\tbra{p}{l_i(U_1 \dots U_n):S_i}{i\in I}
+\Gamma \vdash \bra{p}{l_i(\vec x_i): P_i}{i\in I} : 
+\tbra{p}{l_i(\vec U_i):S_i}{i\in I}
 $}
@@
 \end{tabular}
-\captionof{table}{Duality of Binary Session Types involving participants $\role{p}$ and $\role{q}$}
+\captionof{table}{Duality of Binary Session Types 
+Involving Participants $\role{p}$ and $\role{q}$}
 \label{table:dual}
@@
 \begin{cases}
-	\tmsg{Succ(int)}: \tselone{Client}{\tmsg{Succ(int)}} \\
+	\tmsg{Succ(int)}: \tselone{Client}{\tmsg{Succ(int)}}. \tend \\
 	\tmsg{Quit()}:  \tend \\
@@
 $T \subtype T'$ reads \textit{$T$ is a subtype of $T'$},
-and is defined coinductively on the structure of session types.
+and is defined coinductively on the structure of $T$.
 
@@
 \begin{itemize}
+\item \textbf{Selection}: 
+The supertype of a selection process offers a 
+superset of the internal choices and 
+can send more generic types of payload; 
+intuitively, if a process sends a \texttt{nat}, 
+the payload is compatible with receivers expecting 
+a more generic \texttt{int} payload.
+
 \item \textbf{Branching}: 
@@
 it can handle a \texttt{nat} payload.
-
-\item \textbf{Selection}: 
-The supertype of a selection process offers a 
-superset of the internal choices and 
-can send more generic types of payload; 
-intuitively, if a process sends a \texttt{nat}, 
-the payload is compatible with receivers expecting 
-a more generic \texttt{int} payload.
 \end{itemize}
@@
 \begin{prooftree}
-\AxiomC{$\forall i \in I.$}
-\AxiomC{$U'_1 \subtype U_1 ~ \dots ~ U'_n \subtype U_n$}
-\AxiomC{$S_i \subtype T'_i$}
-\RightLabel{\rulename{Sub-Bra}}
+\AxiomC{$\forall i \in I.\left(
+\vec U_i = U_1, \dots, U_n 
+\quad
+\vec U'_i = U'_1, \dots, U'_n
+\quad
+U_1 \subtype U'_1 ~ \dots ~ U_n \subtype U'_n
+\quad
+T_i \subtype T'_i
+\right)$}
+\RightLabel{\rulename{Sub-Sel}}
 \doubleLine
-\TrinaryInfC{$
-	\tbra{p}{l_i(U_1 \dots U_n):T_i}{i\in I \cup J}
+\UnaryInfC{$
+	\tsel{p}{l_i(\vec U_i):T_i}{i\in I}
 	\subtype
-	\tbra{p}{l_i(U'_1 \dots U'_n):T'_i}{i\in I}
+	\tsel{p}{l_i(\vec U'_i):T'_i}{i\in I \cup J}
 $}
@@
 \begin{prooftree}
-\AxiomC{$\forall i \in I.$}
-\AxiomC{$U_1 \subtype U'_1 ~ \dots ~ U_n \subtype U'_n$}
-\AxiomC{$T_i \subtype T'_i$}
-\RightLabel{\rulename{Sub-Sel}}
+\AxiomC{$\forall i \in I.\left(
+\vec U_i = U_1, \dots, U_n 
+\quad
+\vec U'_i = U'_1, \dots, U'_n
+\quad
+U'_1 \subtype U_1 ~ \dots ~ U'_n \subtype U_n
+\quad
+T_i \subtype T'_i
+\right)$}
+\RightLabel{\rulename{Sub-Bra}}
 \doubleLine
-\TrinaryInfC{$
-	\tsel{p}{l_i(U_1 \dots U_n):T_i}{i\in I}
+\UnaryInfC{$
+	\tbra{p}{l_i(\vec U_i):T_i}{i\in I \cup J}
 	\subtype
-	\tsel{p}{l_i(U'_1 \dots U'_n):T'_i}{i\in I \cup J}
+	\tbra{p}{l_i(\vec U'_i):T'_i}{i\in I}
 $}

### 2020-06-16 report/sections/introduction/contributions.tex
Report: start new introduction

@@
 We present an end-to-end workflow for generating TypeScript APIs for 
-\emph{server-centric} communication protocols over WebSockets. 
-Server-centric 
-protocols feature exactly one server role that is involved in all 
-interactions.
+server-centric communication protocols over WebSockets.
 

### 2020-06-16 report/sections/introduction/motivation.tex
Report: minor fixes
Report: start new introduction

--- 2020-06-10 report/sections/introduction/motivation.tex
@@
+%\subsubsection{New Structure (similar to ECOOP2017):}
+%\begin{itemize}
+%
+%\item \textbf{Section 1.1:}
+%
+%Interactive webapps over WebSockets, harder to guarantee communication safety
+%
+%\item
+%Example protocol similar to travel agency  -- motivate that it is difficult to manually verify communication safety
+%
+%\begin{itemize}
+%\item traveller enquires availability, flight server replies yes/no then traveller confirms/denies.
+%\item The flight server will reserve the seat until the traveller denies.
+%\item Source of hidden bugs -- traveller may enquire twice (e.g. click twice), but server processes one enquiry, so one seat is held up for no reason
+%\end{itemize} 
+%
+%\item MPST framework can be applied for webapps.
+%This is solved by existing work PureScript and MVU2020.
+%
+%\item \textbf{Problem1}: 
+%
+%existing work provides a workflow for session typed web development, but requires the developer to compromise by adapting to programming paradigms and practices not conventional for modern web programming.
+E.g.
+functional paradigm, sequential UI framework, PureScript and LINKS not interoperable with the ever-growing universe of JavaScript libraries.
+These compromises limit the usability of MPST framework
+%for web development.
+%
+%\item \textbf{PRoblem2}: 
+%
+%existing work based on WebSocket transport, which
+%is simple to reason about. But WebSocket transport enforces a server-centric
+%network configuration, so existing work isn't expressive enough for
+%more general protoocls. Suppose traveler contacts agency first to get a quote and then compare (similar to 2-buyer). Existing work cannot implement this.
+%
+%\item \textbf{Section 1.2: }
+%
+%Our objective is to present a novel variant of MPST framework
+%that is compatible / supports the flexibility of modern web programming.
+%We address the two problems in two parts:
+%
+%\item We solve problem 1 in part 1: 
+%
+%SessionTS is API generation workflow based on MPST, which
+%provides developers with same communication safety guarantees in a way that is idiomatic for modern web programming.
+We use industry tools (TS/Node/REact), compatible with async logic, error handling specific to web (e.g.
+client disconnection)
+%
+%\item We solve problem 2 in part 2:
+%
+%develop novel theory for routed MPST which lets us implement non-server-centric commuication structures over a server-centric network topology, prove
+%that our theory preserves properties like communication safety, deadlock freedom, etc.
+%
+%\item Section 1.3: diagram similar to FZ for outline
+%
+%\end{itemize}
+%
+%% Aim of introductio: motivate what is the problem 
+%%(session types fix web, PureScript and LINKS have two problems: 
+%%implementation -- not using mainstream stuff, 
+%%theory -- websocket transport restricts protocols)
+%
+
 \section{Motivation}
-
-\paragraph{Note: from PLACES}
-
-% Rise of distributed programs -> main challenges
-% How to ensure correctness in general: type system, data types
-% Type discipline for concurrent programs - behavioural types (--> session types)
-% Mainstream example of distirbuted programs: web services (microservice architecture), interactive web application; 
-% Objective:
+\label{section:intro}
 
@@
 minimising the communication latency between clients and servers. 
-Whilst the \textit{HTTP} request-response model is 
-sufficient for retrieving static assets, applying the same 
-stateless communication approach for interactive use cases 
-(such as multiplayer games) introduces undesirable 
-performance overhead from having to frequently set up 
+Whilst the \textit{HTTP} 
+request-response model is 
+sufficient for retrieving static assets, 
+applying the same 
+stateless communication approach for 
+interactive use cases 
+introduces 
+undesirable performance overhead from having to 
+frequently set up 
 new connections for client-server interactions. 
 Developers have since adopted other communication 
-transport abstractions over HTTP connections such as the WebSockets protocol \cite{WebSocketRFC} to enjoy low-latency full-duplex 
+transport abstractions over HTTP connections such 
+as the WebSockets protocol \cite{WebSocketRFC}
+to enjoy low-latency full-duplex 
 client-server communication in their applications over 
@@
 
-% Example: noughts and crosses
-Consider a classic turn-based board game of \textit{Noughts and Crosses} 
-between two players. Both players are identified by either noughts or crosses 
-respectively, and take turns to place a mark on an unoccupied cell 
-of a 3-by-3 grid until one player wins (when their markers form 
-one straight line on the board) or a stalemate is reached 
-(when all cells are occupied). A web-based implementation may 
-involve players connected to a game server via WebSocket connections 
-and interacting with the game from their web browser, which serve 
-a \textit{single-page application} (SPA) of the game client written 
+\paragraph{The need for communication safety 
+in interactive web applications}
+
+Consider a \textit{flight booking service}
+between a \trole{Traveller} and a \trole{FlightServer}.
+The \trole{Traveller} queries the \trole{FlightServer}
+about prices to a particular destination.
+If there is still availability, 
+the \trole{FlightServer} will \textit{reserve a seat}
+responds with the price,
+to which the \trole{Traveller} confirms or rejects the purchase;
+the latter choice will notify the \trole{FlightServer}
+to \textit{release the seat}.
+Otherwise, the \trole{FlightServer} notifies the
+\trole{Traveller} that all flights are full.
+The \trole{Traveller} can proceed to query for another
+destination.
+
+We can implement the flight booking service as a web
+application.
+The \trole{FlightServer} can be a web server
+running on the \textit{Node.js} JavaScript runtime \cite{nodejs}.
+The \trole{Traveller} can run on the web browser
+as a \textit{single-page application} (SPA) written
 in a popular framework like \textit{React.js} \cite{React}.
-SPAs feature a single HTML page and dynamically renders content 
-via JavaScript in the browser. 
-Players take turns to make a move on the game board and the server 
-implements the game logic to progress the game forward until 
-a result (either a win/loss or draw) can be declared.
+SPAs feature a single HTML page and dynamically renders
+content via JavaScript in the browser.
+The \trole{Traveller} connects to the \trole{FlightServer}
+using a WebSocket connection and they enjoy
+bi-directional communication to execute the flight
+booking operations outlined above.
 
@@
 it introduces the developer to a new family of communication errors
-(in addition to the usual testing for game logic correctness), 
-even for this simple game:
+(in addition to the usual testing for application logic correctness), 
+even for this simple example:
 we highlight just a few:
@@
 \item
-\textbf{Deadlocks:} how can we prevent both players waiting for 
-each other to make a move at the same time?
+\textbf{Deadlocks:} how can we prevent both sides waiting for 
+each other to respond at the same time?
 
 \item 
-\textbf{Communication mismatches:} what if player 1 sends
-a boolean to the game server instead of board coordinates?
+\textbf{Communication mismatches:} what if the 
+\trole{FlightServer} sends a string to the \trole{Traveller}
+who is expecting the price to be a numeric value?
 
 \item
-\textbf{Channel linearity:} if the game server takes time
-to update the game logic and respond to the players, what if 
-player 1 clicks the same cell twice and sends the coordinates twice?
+\textbf{Channel linearity:} if the \trole{FlightServer}
+takes time to respond to a query from the \trole{Traveller},
+what if the \trole{Traveller} sends the same query twice? 
+Given that the \trole{FlightServer} reserves a seat on
+a ``successful'' query, violating channel linearity
+would hold up seats unnecessarily.
 
@@
 
-The complexity of these errors, which correlate to the complexity of tests 
-required against these errors, scale with the complexity of the 
-communication patterns involved. Over-reliance on integration testing
-to attempt to expose communication-related bugs will also slow the
-development process, not to mention that the time taken for these
+The complexity of these errors,
+which correlate to the complexity of tests 
+required against these errors, 
+scale with the complexity of the 
+communication patterns involved. 
+Over-reliance on integration testing
+to attempt to expose communication-related bugs 
+will also slow the
+development process, not to mention that the time 
+taken for these
 integration tests would scale with the number of roles involved.
-A localised, static way for verifying communication correctness would
-be highly desirable.
+A localised, static way for verifying communication correctness
+is highly desirable.
 
-\textit{Multiparty Session Types} (MPST) \cite{MPST} provide 
-a framework for formally specifying a structured communication pattern 
-between concurrent processes and verifying implementations for 
+\paragraph{Formalising Interactions with
+Multiparty Session Types}
+\textit{Multiparty Session Types} (MPST) \cite{MPST} 
+provide a framework for formally specifying 
+a structured communication pattern 
+between concurrent processes and 
+verifying implementations for 
 correctness with respect to the communications aspect. 
-By specifying the client-server interactions of our game as a protocol 
-and verifying the implementations against the protocol for well-formedness, 
-MPST theory guarantees well-formed implementations to be 
-free from communication errors.
+By specifying the client-server interactions of 
+our flight booking service as a protocol 
+and verifying the implementations against 
+the protocol for well-formedness, 
+MPST theory guarantees well-formed 
+implementations to be free from communication errors.
 
+\paragraph{Limitations of State-of-the-Art Session-Typed Web Development}
 Existing work \cite{MVU2020,PureScript2019} that 
 adapt the MPST framework for web development
-acknowledge the limitations of JavaScript -- the language of the browser --
-in providing static type-level guarantees, and proceed to use
-languages equipped with stronger type systems that compile to JavaScript.
-This comes at the cost of a learning curve for the developer,
-and limits its utility in the mainstream web development space.
+acknowledge the limitations of JavaScript 
+-- the language of the browser --
+in providing static type-level guarantees
+for communication safety, 
+and proceed to use
+languages equipped with stronger type systems that 
+compile to JavaScript instead.
 
-Among the many languages that compile to JavaScript,
-TypeScript stands out as the most intuitive to use as it is defined
-to be a \textit{superset} of JavaScript. It provides developers
-with type-safety through its gradual, structural type system.
-Whilst some \cite{MVU2020} point out this limits its usability for
-encoding multiparty session types, we believe that the language
-offers sufficient features that we can use to provide developers with
-communication safety guarantees whilst preserving a flexible, natural
-and idiomatic workflow. By building our work upon TypeScript, we
-work towards incorporating MPST into mainstream web development,
-which reduces development time by programmatically verifying 
-implementations for communication correctness.
+With respect to the end goal of 
+offering developers a development workflow
+that provides communication safety guarantees in modern
+web programming through multiparty session types,
+we observe two main limitations from the state-of-the-art:
+
+\subparagraph{L1: 
+Incompatibility with modern web programming practices}
+
+The current state-of-the-art succeeds in \textbf{statically}
+providing communication safety guarantees,
+but requires the developer to compromise by adapting
+to programming paradigms and practices that are not
+conventional for modern web programming.
+Developers that use \cite{MVU2020} are required
+to implement their web applications in a \textit{functional} 
+paradigm using the \textit{Links} 
+web programming language \cite{LINKS}.
+\cite{PureScript2019} also endorses
+functional web programming by requiring developers to
+write PureScript \cite{PureScript} applications, but also
+enforces user interfaces to be written using 
+sequential UI frameworks.
+
+PureScript and Links are not interoperable
+with the ever-growing universe of JavaScript libraries for
+web development.
+It is also unclear whether these tools support idiomatic
+practices in event-driven programming, such as callbacks
+and asynchronous implementations.
+Whilst communication safety guarantees provided by
+\cite{PureScript2019,MVU2020} are based on
+canonical MPST theory, they are not contextualised
+for the dynamic web-based environment.
+Going back to the flight booking service as an example,
+\cite{PureScript2019,MVU2020} do not handle premature
+disconnections; if the \trole{Traveller} receives a quote
+and closes their browser prematurely and the \trole{FlightServer}
+isn't notified of this event, the reserved seat will not be
+released.
+These compromises limit the usability of the MPST framework
+in modern web programming.
+
+\subparagraph{L2: 
+Limited expressiveness of communication protocols}
+
+The current state-of-the-art leverage the WebSocket protocol
+as the communication transport between roles, which is
+simple to reason about and straightforward to relate
+to session type theory.
+However, WebSocket transport enforces a \emph{server-centric}
+network topology, so \cite{PureScript2019,MVU2020}
+enforces the same constraint on the range of communication
+protocols supported:
+server-centric protocols feature exactly one server
+role that is involved in all interactions.
+
+This limits the types set of communication topologies
+that can be implemented by session-typed web development
+proposals. 
+Suppose we extend the flight booking service as follow:
+we introduce a \trole{TravelAgent} role, which
+runs on the browser; the \trole{Traveller} first asks
+for a quote from the \trole{TravelAgent} before 
+carrying out the usual interactions with the \trole{FlightServer},
+so the \trole{Traveller} can compare between both options.
+The current state-of-the-art cannot implement this 
+extended flight booking service, as the described
+communication interactions is no longer server-centric.

### 2020-06-16 report/sections/introduction/contributions.tex
Report: new introduction

@@
 session type theory, and prove the preservation of well-formedness
-and communication. Detailed lemmas and proofs in this section can
+and communication. Detailed lemmas and proofs in this chapter can
 be found in \cref{section:proofs}.
@@
 case studies of communication protocols 
-implemented as webservices. Through examples of binary sessions,
-multiparty sessions and routed multiparty sessions, we demonstrate how the
-developer may implement the generated APIs and consider the impact of our work
+implemented as webservices. Through examples of
+multiparty sessions and routed multiparty sessions, 
+we demonstrate how the
+developer may implement the generated APIs and 
+consider the impact of our work
 on developer productivity.
@@
 Applications in TypeScript} in Volume 314 of the Electronic Proceedings
-in Theoretical Computer Science (EPTCS) \cite{PLACES2020}.
+in Theoretical Computer Science (EPTCS) \cite{PLACES2020},
+in which I am the first and lead author.
+The materials in 
+\cref{chap:ext,chap:p2p,chap:theory,chap:impl,chap:eval}
+are developed exclusively in this project by myself.
 \end{remark}

### 2020-06-16 report/sections/introduction/objectives.tex
Report: start new introduction
Report: new introduction

--- 2020-05-29 report/sections/introduction/objectives.tex
@@
 
-The objective of this project is to \dots
+The objective of this project is to provide
+developers with a development workflow
+that provides communication safety guarantees in modern
+web programming through multiparty session types.
+
+We tackle the limitations of the current
+state-of-the-art with a two-step strategy:
+
+\subparagraph{S1}
+We implement \codegen, an end-to-end framework
+for writing full-stack \textit{TypeScript} applications that
+\textbf{statically} conform to a communication protocol.
+
+Developers specify their communication protocol using
+the Scribble protocol description language \cite{Scribble}
+which is based on multiparty session type (MPST) theory \cite{MPST}.
+The Scribble protocol is used as a basis for generating
+TypeScript APIs targeted for the Node.js server-side runtime
+and React.js browser-side framework.
+By implementing the generated APIs in their full-stack applications,
+developers can enjoy static guarantees with respect 
+to communication safety: 
+there will not be any communication mismatches,
+and channel linearity will be respected.
+
+\subparagraph{S2}
+We present \newtheory, an extension of the canonical MPST
+theory that supports \textit{routed communication}.
+
+Through \newtheory, we can express interactions to be
+routed via an intermediate endpoint.
+By defining an encoding from canonical MPST theory
+to \newtheory, we show that we do not lose expressiveness
+using routed communication.
+We also extend \codegen to implement \newtheory.
+\\
+
+By step \textbf{S1}, we address the limitation \textbf{L1}.
+Our work provides the same communication safety guarantees
+as the current state-of-the-art, but 
+is compatible with modern web programming tools and idioms.
+Among the languages that compile to JavaScript,
+we select TypeScript as it is arguably the most intuitive
+to use, given it is a \textit{superset} of JavaScript 
+\cite{UnderstandingTypeScript}.
+It provides developers
+with type-safety through its gradual, structural type system,
+and is also interoperable with existing JavaScript libraries
+that the developer may incorporate into their application.
+Whilst some \cite{MVU2020} point out this limits its usability for
+encoding multiparty session types, we believe that the language
+offers sufficient features that we can use to provide developers with
+communication safety guarantees whilst preserving a flexible, natural
+and idiomatic workflow.
+The TypeScript APIs we generate also support idioms commonly
+found in modern web development (such as callbacks and
+asynchronous operations), and provides an error handling
+mechanism tailored to the dynamic web-based environment
+(where roles can disconnect prematurely).
+
+In step \textbf{S2},
+we show that routing communication interactions through an intermediate
+role, such as a web server, preserves the communication
+safety properties and the semantics from the original
+interactions.
+By extending \codegen to implement \newtheory,
+our workflow can implement protocols of
+arbitrary communication structures over a server-centric
+network topology using WebSockets; 
+this tackles limitation \textbf{L2}.

### 2020-06-16 report/sections/abstract.tex
Report: address some NY comments
Report: new introduction / abstract

--- 2020-06-07 report/sections/abstract.tex
@@
 
-We present an end-to-end framework
-to write interactive web applications,
-communicating over WebSocket transport,
-that statically conform to prescribed communication protocols.
+\noindent
+Multiparty Session Types (MPST)
+is a typing discipline for concurrent
+processes that communicate via
+message-passing. MPST theory can ensure
+communication safety properties and
+protocol conformance.
+Existing work in session-typed
+web development over WebSocket transport
+is incompatible with modern web
+programming practices and are limited
+to supporting communication protocols
+that implement the server-centric network
+topology.
 
-We implement a code generation toolchain to
-provide developers with TypeScript APIs generated
+We address limitations in the current 
+state-of-the-art by:
+\textbf{(1)} implementing \codegen, a code generation
+toolchain for session-typed web development
+over WebSocket transport using TypeScript;
+and \textbf{(2)} presenting \newtheory, a new
+multiparty session type theory that supports
+routed communications.
+
+\codegen provides
+developers with TypeScript APIs generated
 from a communication protocol specification 
 based on multiparty session type theory.
+Our work is compatible with modern web programming
+industrial practices. The generated APIs
+build upon TypeScript concurrency primitives,
+complement the event-driven style of programming
+in full-stack web development,
+and are compatible
+with the Node.js runtime and the React.js framework
+for server-side and browser-side endpoints respectively.
+We evaluate the expressiveness of \codegen for modern
+web programming
+using case studies of protocols found in web-services,
+and analyse the overhead of our implementation through 
+running micro-benchmarks against a baseline implementation.
 
-Our toolchain generates APIs that are compatible with
-the Node.js runtime and the React.js framework
-for server-side and browser-side endpoints respectively.
-
-We evaluate the expressiveness of our API generation solution
-using three case studies of protocols found in web-services,
-and analyse the overhead of our implementation through 
-running micro-benchmarks on a simple binary session protocol.
+\newtheory can express interactions to be routed
+via an intermediate participant.
+Using \newtheory, 
+we propose an approach for supporting 
+peer-to-peer communication
+between browser-side endpoints through 
+routing communication via the server
+in a way that avoids excessive serialisation
+and preserves communication safety. 
+We evaluate the correctness of \newtheory
+by proving communication safety properties, such as
+deadlock freedom.
 

### 2020-06-16 report/sections/introduction/contributions.tex
Report: new introduction / abstract

@@
 interactive web applications over WebSocket transport.
-Related work \citep{PureScript2019, MVU2020, LINKS, Exceptional} requires the 
+We also present a theory of routed multiparty session types,
+which forms the basis for supporting arbitrary communication structures
+over a server-centric WebSocket-based network topology.
+Related work \citep{PureScript2019, MVU2020, LINKS, Exceptional} 
+requires the 
 developer to implement their application using tools that are 
-unconventional for web development.
+unconventional for web development, 
+and only support communication protocols that adhere to the
+WebSocket network structure.
 To the best of our knowledge, this is the first work on multiparty session
 type-safe web development that provides the same communication safety 
-guarantees using familiar industry tools in TypeScript and React.
+guarantees using familiar industry tools in TypeScript and React,
+and supports peer-to-peer browser interactions over WebSocket transport.
 
@@
 In \textbf{\cref{chap:codegen}}, 
-we introduce the system architecture for our
-code generator, \fancyname{SessionTS}. 
+we introduce the system architecture for
+\codegen. 
 The implementation integrates with the Scribble
@@
 We present two code generation strategies for back-end and front-end
-TypeScript web development as \fancyname{NodeMPST} and \fancyname{ReactMPST}
+TypeScript web development as \nodecodegen and \reactcodegen
 in \textbf{\cref{chap:node}} and \textbf{\cref{chap:react}} respectively.
@@
 In \textbf{\cref{chap:ext}},
-we extend \fancyname{SessionTS} to support advanced
+we extend \codegen to support advanced
 practices commonly found in full-stack web development in a way that
-preserves communication safety. We modify the API generation strategies
+preserves communication safety. 
+We modify the API generation strategies
 to support \textit{asynchronous implementations} 
@@
 developer may implement the generated APIs and 
-consider the impact of our work
-on developer productivity.
+consider the compatibility of our work with respect to
+modern web programming practices.
+We perform benchmarks on a web application built using our
+generated APIs against a baseline implementation to evaluate
+the overhead of our work.
 
@@
 in which I am the first and lead author.
+
+The materials in \cref{chap:codegen,chap:node,chap:react}
+are revised from \cite{PLACES2020}.
 The materials in 

### 2020-06-16 report/sections/introduction/objectives.tex
Report: new introduction / abstract

@@
 
-\subparagraph{S1}
-We implement \codegen, an end-to-end framework
+\subparagraph{S1: \codegen}
+We implement an end-to-end framework
 for writing full-stack \textit{TypeScript} applications that
@@
 
-\subparagraph{S2}
-We present \newtheory, an extension of the canonical MPST
+\subparagraph{S2: \newtheory}
+We present an extension of the canonical MPST
 theory that supports \textit{routed communication}.

### 2020-06-16 report/sections/evalconcl/concl/contributions.tex
Report: new conclusion

@@
 
-In this project, we develop and present an
+In this project, we develop and present a novel
 MPST-based framework for developing full-stack
@@
 and browser-side endpoints respectively.
-
 In particular, we present a novel approach for integrating
@@
 
-Compared to previous work, we are not only able to
+Compared to previous work 
+\cite{Exceptional,PureScript2019,MVU2020,LINKS} on
+session-typed web development, 
+we are not only able to
 \textbf{statically} provide the same level of communication safety
@@
 and we explain in \cref{chap:ext} how we support advanced
-web development idioms such as asynchronous implementations.
+web development idioms (such as asynchronous implementations)
+and handle premature disconnections gracefully.
 
-In addition, existing proposals on API generation for web development
-over WebSockets \cite{PureScript2019} only support protocols
-implementing server-centric topologies, in order to be 
-compatible with WebSocket transport.
-We formalise a theory of routed multiparty session types in 
+We also do not limit support to protocols that describe
+a server-centric topology: we formalise 
+a theory of routed multiparty session types in 
 \cref{chap:theory} to prove that it is possible to relax the
@@
 in Volume 314 of the Electronic Proceedings in Theoretical
-Computer Science \cite{PLACES2020}; we were also invited to present this work
+Computer Science (EPTCS) \cite{PLACES2020}.
+We were also invited to present this work
 at the 12th International Workshop on

### 2020-06-16 report/sections/evalconcl/concl/future.tex
Report: complete chapter 10.1
Report: complete chapter 10 sans code listings for 10.2 and 10.3
Report: new conclusion

@@
 \section{Future Work}
+\label{section:future}
 
@@
 
-\paragraph{Delegation}
+\paragraph{Session Delegation}
 Delegation allows channels to be sent between endpoints.
-This directly addresses the limitations presented in
-\cref{section:evalatm}: in the \tprotocol{ATM} protocol,
-when the \trole{Client}
-connects to the \trole{Svr},
-the \trole{Svr} should be able to send a private channel
-for \trole{Client} to perform bank operations,
+Using the \tprotocol{Noughts and Crosses} game
+from \cref{section:evalgame}, the developer
+may want the \trole{Svr} to send a private channel
+to both players to play the game,
 whilst concurrently accepting new connections.
@@
 \textit{distributed multiparty delegation} in Scala.
-We have also shown in \cref{section:evalatm} that
-the developer can implement delegation with small changes
-to the session runtime.
-Future work can apply the findings from \cite{LinearDecomp},
-extend the Scribble protocol language for sending
-channels through messages, and formalise the workaround
-in \cref{section:evalatm} with respect to the extended 
-Scribble syntax for delegation.
-
-\paragraph{``Serverless'' Server-Side Endpoints}
-\dots
+Their work also uses the Scribble toolchain
+for protocol specification.
+Future work can apply the findings from \cite{LinearDecomp}
+and extend \codegen to support channel delegation
+in the generated handler-style APIs.
 

### 2020-06-16 report/sections/evalconcl/eval.tex
Report: progress on chapter 10
Report: new conclusion

@@
 of our work,
-we use three case studies
+we use two case studies
 of protocols found in webservices
@@
 
-\item
-\textbf{\tprotocol{ATM} Protocol:}
-
-In \cref{section:evalatm},
-we implement an ATM web application to demonstrate
-limitations of our work. Our generated APIs do not allow
-the server endpoint (in this case, the \trole{Bank})
-to participate in concurrent sessions (i.e. interacting
-with different \trole{Client}s concurrently), even though
-the protocol is written in a way that allows this.
-However, we propose a way to address this limitation, and expand
-upon this in \cref{section:future}.
-
 \end{itemize}
@@
 
-\input{sections/evalconcl/eval/atm}
+%\input{sections/evalconcl/eval/atm}
 

### 2020-06-16 report/sections/background/sessiontypes.tex
Report: complete ch2.1

@@
 
-% need LTS semantics here too
+Whilst binary session types provide communication guarantees 
+between exactly 2 participants, 
+distributed systems generally involve more parties in practice. 
+This is equally relevant in interactive web applications, 
+as motivated by the \textit{Battleships} game example 
+in \cite{PureScript2019} where the server coordinates 
+interactions between two players.
+
+Whilst there is a natural syntactical extension to 
+our session calculus for 
+describing multiparty sessions\footnote{
+We also adopt the shorthand 
+$\mathcal{M} ::= \prod^n_{i=1} \ptprocess{p_i}{P_i}$ 
+as used in the literature.} as
+
+\[
+\begin{array}{rl}
+\mathcal{M} ::= & \ptprocess{p_1}{P_1} ~\mid~
+\ptprocess{p_2}{P_1} ~\mid~\dots~\mid
+\ptprocess{p_n}{P_n}
+\end{array}
+\]
+
+the same cannot be said for the binary session typing discipline,
+particularly with respect to duality.
+The same notion of duality does not extend to the 
+decomposition of multiparty interactions into 
+multiple binary sessions: 
+\cite{C406Lecture} and \cite{MPST} both present 
+counterexamples of well-typed binary sessions that, 
+when composed to represent a multiparty session, 
+results in communication errors thus 
+violating guarantees of well-typed sessions.
+
+Honda et al. presents {multiparty session types} 
+in \cite{MPAST} to extend the 
+binary session typing discipline for 
+sessions involving more than 2 participants, 
+whilst redefining the notion of compatibility 
+in this multiparty context. 
+\textit{Multiparty session types} (MPST)
+are defined in terms a \textit{global type},
+which provides a bird's eye view of the communication protocol
+describing the interactions between pairs of participants. 
+\cref{fig:mpst} defines the syntax of global types
+inspired by \cite{MPST};
+we omit message payload types for brevity.
+
+% Introduce global type, projection
+\begin{figure}[!hb]
+\doublespacing
+\[
+\begin{array}{rlr}
+
+G ::= & & \text{Global Types} \\
+     & \tend & \text{Termination} \\
+\mid & \gcomm{p}{q}{l_i: G_i}{i \in I} & \text{Message Exchange} \\
+\mid & \trec{G} & \text{Recursive Type} \\
+\mid & \trecvar & \text{Type Variable} \\
+\end{array}
+\]
+\singlespacing
+\caption{Syntax of Global Types}
+\label{fig:mpst}
+\end{figure}
+
+To check the conformance of a participant's process 
+against the protocol specification 
+described by the global type, 
+we \textit{project} the global type onto the participant
+to obtain a session type that only preserves the 
+interactions described by the global type that 
+pertain to the participant. 
+Projection is defined by the $\upharpoonright$ operator, 
+more commonly seen in literature in its infix form 
+as $\proj{G}{p}$ describing the projection of global type 
+$G$ for participant $\role{p}$. 
+Intuitively, the projected local type of a participant 
+describes the protocol from the viewpoint of the participant.
+
+More formally, projection can be interpreted as a 
+\textit{partial function} 
+${\upharpoonright} :: G \times \mrole{p} \rightharpoonup S$,
+as the projection for a participant may be undefined for 
+an ill-formed global type; 
+\cite{C406Lecture} presents examples of where this is the case, 
+and \cite{MPST} presents the formal definition of projection.
+We focus on the definition of projection for the 
+message exchange global type given below:
+
+\[
+\proj{\gcomm{p}{q}{l_i: G_i}{i \in I}}{r} =
+\begin{cases}
+\tsel{q}{l_i: \proj{G_i}{p}}{i \in I} 
+	& \text{if } \mrole{r} = \mrole{p} \\
+\tbra{p}{l_i: \proj{G_i}{q}}{i \in I} 
+	& \text{if } \mrole{r} = \mrole{q} \\
+\underset{i \in I}{\MERGEOP}~\proj{G_i}{r}
+	& \text{otherwise} \\
+\end{cases}
+\]
+
+The third case describes that,
+for a participant not involved in the communication,
+the projections of the continuations $\proj{G_i}{r}$
+do not necessarily have to be the same, but they
+need to be ``compatible''. 
+The $\mergeop$ syntax denotes the \textit{merging operator}.
+It is a partial function on local types allows us to construct
+a projection that is compatible with the projections on
+the continuations.
+We use the \textit{inductive} definition presented by
+Scalas and Yoshida in \cite{LessIsMore}, shown in \cref{fig:merge}.
+
+\begin{figure}[!h]
+\doublespacing
+\[
+\begin{array}{rlr}
+
+\tmerge{\tend}{\tend} &= \tend 
+	& \rulename{Merge-End} \\
+\tmerge{\trecvar}{\trecvar} &= \trecvar 
+	& \rulename{Merge-RecVar} \\
+\tmerge{\trec{T}}{\trec{T'}} &= \trec{\tmerge{T}{T'}} 
+	& \rulename{Merge-Rec} \\
+\tmerge{\tsel{p}{l_i: T_i}{i \in I}}{\tsel{p}{l_i: T_i}{i \in I}} 
+	&= \tsel{p}{l_i: T_i}{i \in I}
+	& \rulename{Merge-Sel} \\
+\tmerge{\tbra{p}{l_i: T_i}{i \in I}}{\tbra{p}{l_j: T'_j}{j \in J}} 
+	&= \tbra{p}{l_k: T''_k}{k \in I \cup J}
+	& \rulename{Merge-Bra} \\
+
+	& 
+	\text{where } T''_k =
+	\begin{cases}
+	T_k & \text{if } k \in (I \setminus J) \\
+	T'_k & \text{if } k \in (J \setminus I) \\
+	\tmerge{T_k}{T'_k} & \text{otherwise} \\
+	\end{cases}
+	& \\
+
+\end{array}
+\]
+\singlespacing
+\captionof{figure}{Definition of Merging Operator}
+\label{fig:merge}
+\end{figure}
+
+The notion of compatibility in multiparty session types 
+is still captured by \rulename{MTy}, 
+but adapted to consider the local projections for 
+all participants as supposed to dual types in the binary case.
+
+\begin{prooftree}
+\AxiomC{$
+\forall i \in I. ~ \left( \cdot \vdash P_i : \proj{G}{p_i} \right)
+$}
+\AxiomC{$
+\pt{G} \subseteq \left\{ \mrole{p_i} \mid i \in I \right\}
+$}
+\RightLabel{\rulename{MTy}}
+\BinaryInfC{$
+\displaystyle \vdash \prod^{}_{i \in I} \ptprocess{p_i}{P_i} ~: G
+$}
+\end{prooftree}
+
+For a multiparty session 
+$\mathcal{M} = \prod_{i \in I} \ptprocess{p_i}{P_i}$ 
+to be well-typed by a global type $G$:
+
+\begin{enumerate}
+\item 
+All participant processes $\mrole{p_i}{P_i}$ are well-typed 
+with respect to their corresponding well-defined projection 
+$\proj{G}{p_i}$, and 
+\item 
+$G$ does not describe interactions with participants 
+not defined in $\mathcal{M}$.
+$\pt{G}$ denotes the set of participants in the global type $G$,
+defined inductively on $G$.
+
+\begin{align*}
+\pt{\tend} &= \emptyset \\
+\pt{\trecvar} &= \emptyset \\
+\pt{\trec{G}} &= \pt{G} \\
+\pt{\gcomm{p}{q}{l_i: G_i}{i \in I}} &= 
+	\{ \mrole{p}, \mrole{q} \} \cup \underset{i \in I}{\bigcup}~\pt{G_i} \\
+\end{align*}
+
+\end{enumerate}
+
+Well-typed multiparty session enjoys the following 
+guarantees as outlined in \cite{GentleMPST}:
+
+\begin{itemize}
+\item \textbf{Communication safety}: The types of sent and expected messages will always match.
+\item \textbf{Protocol fidelity}: The exchange of messages between processes will abide by the global protocol.
+\item \textbf{Progress}: Messages sent by a process will be eventually received, and a process waiting for a message will eventually receive one; this also means there will not be any sent but unreceived messages.
+\end{itemize}
+
+This motivates an elegant, decentralised solution for checking protocol conformance in practice: once the global type for the protocol is defined, local processes can verify their implementation against their corresponding projection in isolation, independent of each other.
+We illustrate this in \cref{fig:mpstworkflow}.
+
+\begin{figure}[!ht]
+\centering
+\includegraphics[width=.75\textwidth]{MpstFramework}
+\caption{Type Checking with Multiparty Session Types}
+\label{fig:mpstworkflow}
+\end{figure}

### 2020-06-16 report/sections/general/p2p/challenges.tex
Report: complete ch2.1

@@
 By existing LTS semantics over
-global types (\cref{subsubsection:lts}), then $SB!\tmsg{M2}$
+global types \cite{characterisation}, 
+then $\aout{S}{B}{\tmsg{M2}}$
 is a prefix of a valid execution trace.
@@
 our naive definition of routing has over-serialised the protocol
-since $SB!\tmsg{M2}$ can no longer occur first because \trole{S}
+since $\aout{S}{B}{\tmsg{M2}}$ can no longer occur first because \trole{S}
 is forced to send \tmsg{M1} first, which it cannot do so until receiving

### 2020-06-16 report/sections/background/related.tex
Report: progress on background

@@
 \label{section:bgrelated}
+
+Whilst session type theory represents 
+the type language for concurrent processes,
+it also forms the theoretical basis for proposals
+introduced to implement session types for 
+real-world application development:
+the \textit{Scribble} project is one such proposal.
+We discuss related work that 
+implement session types for software
+engineering using the Scribble project
+in \cref{subsection:bgscribble},
+and focus on existing work for
+session-typed web development
+in \cref{subsection:sessiontypewebdev}.
 
@@
 
+% Introduce Scribble as an implementation of MPST
+Scribble \cite{Scribble} is a
+platform-independent description language 
+for the specification of message-passing protocols.
+The language describes the behaviour of 
+communicating processes at a high level of abstraction:
+more importantly, the description is independent from 
+implementation details in the same way that 
+the type signature of a function declaration is decoupled 
+from the corresponding function definition.
+
+A Scribble \textit{protocol specification} 
+describes an agreement of how participating systems, 
+referred to as \text{roles}, interact. 
+The protocol stipulates the sequence of structured messages 
+exchanged between roles; 
+each message is labelled with a name and the
+type of payload carried by the message.
+
+We present an example of a Scribble protocol in 
+\cref{fig:adder} adapted from \cite{Hybrid2016}.
+The protocol specifies an arithmetic web service 
+offered by a \trole{S}erver to a \trole{C}lient.
+The \trole{C}lient is permitted to either:
+
+\begin{itemize}
+\item 
+Send two \texttt{int}s attached to an \tmsg{Add} message, 
+where the server will respond with an \texttt{int} in a 
+message labelled \tmsg{Res}, and the protocol restarts; or,
+\item
+Send a \tmsg{Quit} message, where the server will 
+respond with a \tmsg{Terminate} message and the protocol ends.
+\end{itemize}
+
+\begin{figure}[!ht]
+\begin{lstlisting}[language=Scribble]
+type <java> "java.lang.Integer" from "rt.jar" as int; (*@\label{line:typedecl}@*)
+
+global protocol Adder(role C, role S) {
+	choice at C {
+		Add(int, int)	from C to S;
+		Res(int)		from S to C;
+		do Adder(C, S);
+	} or {
+		Quit()		from C to S;
+		Terminate()	from S to C;	
+	}
+}
+\end{lstlisting}
+\caption{Adder Protocol in Scribble}
+\label{fig:adder}
+\end{figure}
+
+The platform-independent nature of Scribble 
+can be observed from the \texttt{type} declaration 
+on \cref{line:typedecl}: 
+the developer has the freedom to specify message payload formats 
+and data types from the target language of the implementation 
+-- in this case, aliasing the built-in Java integer as 
+\texttt{int} throughout the protocol.
+
+To observe the parallels between MPST theory
+and the Scribble language,
+we present the corresponding global type for
+the \tprotocol{Adder} protocol below.
+
+\[
+G_\text{Adder} = \trec{
+\gcommone{C}{S}{
+	\begin{cases}
+		\tmsg{Add(int, int)} & :
+			\gcommone{S}{C}{\tmsg{Res(int)}}. \trecvar \\
+		\tmsg{Quit()} & : 
+			\gcommone{S}{C}{\tmsg{Terminate()}}. \tend \\		
+	\end{cases}
+}} 
+\]
+
+The protocol specification language is a component of the broader 
+\textit{Scribble project} initiated by Honda et al. in \cite{Scribble}, 
+through which the project also facilitates the development of 
+\textit{endpoint applications} that conform to user-specified protocols.
+
 \subsection{Session Types in Web Development}
 \label{subsection:sessiontypewebdev}
+
+We discuss the current state-of-the-art proposals
+for integrating session types in web development.
+
+\subsubsection{API Generation in PureScript}
+\dots
+
+\subsubsection{MVU using Links}
+\dots

### 2020-06-16 report/sections/background/sessiontypes.tex
Report: complete ch2.1
Report: progress on background

@@
 $\pt{G}$ denotes the set of participants in the global type $G$,
-defined inductively on $G$.
+defined inductively on $G$ as shown on \cref{table:pt}.
 
-\begin{align*}
-\pt{\tend} &= \emptyset \\
-\pt{\trecvar} &= \emptyset \\
-\pt{\trec{G}} &= \pt{G} \\
-\pt{\gcomm{p}{q}{l_i: G_i}{i \in I}} &= 
-	\{ \mrole{p}, \mrole{q} \} \cup \underset{i \in I}{\bigcup}~\pt{G_i} \\
-\end{align*}
+\renewcommand{\arraystretch}{1.6}
+\begin{table}[!h]
+\centering
+\begin{tabular}{rcl}
+$\pt{\tend}$ &=& $\emptyset$ \\
+$\pt{\trecvar}$ &=& $\emptyset$ \\
+$\pt{\trec{G}}$ &=& $\pt{G}$ \\
+$\pt{\gcomm{p}{q}{l_i: G_i}{i \in I}}$ &=& 
+	$\{ \mrole{p}, \mrole{q} \} \cup \underset{i \in I}{\bigcup}~\pt{G_i}$
+	\\
+\end{tabular}
+\captionof{table}{Definition of Participants in Global Types}
+\label{table:pt}
+\end{table}
+\renewcommand{\arraystretch}{1}
 
@@
 
-Well-typed multiparty session enjoys the following 
-guarantees as outlined in \cite{GentleMPST}:
+Well-typed multiparty sessions enjoy the following 
+communication guarantees as outlined in \cite{GentleMPST}:
 
 \begin{itemize}
-\item \textbf{Communication safety}: The types of sent and expected messages will always match.
-\item \textbf{Protocol fidelity}: The exchange of messages between processes will abide by the global protocol.
-\item \textbf{Progress}: Messages sent by a process will be eventually received, and a process waiting for a message will eventually receive one; this also means there will not be any sent but unreceived messages.
+\item \textbf{Communication safety}: 
+The types of sent and expected messages will always match.
+\item \textbf{Protocol fidelity}: 
+The exchange of messages between processes will abide by the global protocol.
+\item \textbf{Progress}: 
+Messages sent by a process will be eventually received, 
+and a process waiting for a message will eventually receive one; 
+this also means there will not be any sent but unreceived messages.
 \end{itemize}
 
-This motivates an elegant, decentralised solution for checking protocol conformance in practice: once the global type for the protocol is defined, local processes can verify their implementation against their corresponding projection in isolation, independent of each other.
+This motivates an elegant, decentralised solution for 
+checking protocol conformance in practice: 
+once the global type for the protocol is defined, 
+local processes can verify their implementation 
+against their corresponding projection in isolation, 
+independent of each other. 
 We illustrate this in \cref{fig:mpstworkflow}.

### 2020-06-16 report/sections/background/related.tex
Report: progress on background

@@
 
-The protocol specification language is a component of the broader 
-\textit{Scribble project} initiated by Honda et al. in \cite{Scribble}, 
-through which the project also facilitates the development of 
-\textit{endpoint applications} that conform to user-specified protocols.
+The protocol specification language is 
+a component of the broader 
+\textit{Scribble toolchain} initiated by 
+Honda et al. in \cite{Scribble}, 
+through which the toolchain also facilitates 
+the development of 
+\textit{endpoint applications} that 
+conform to user-specified protocols.
+
+A Scribble global protocol can be projected
+to a role, or \textit{endpoint},
+to obtain a local protocol which represents
+the global protocol viewed from the perspective
+of the endpoint.
+This allows the endpoint to \textit{verify} their
+implementation against their local protocol
+for conformance, independent of other endpoints.
+The communication safety guarantees from MPST
+theory also apply here: if the implementation
+for each endpoint is verified against their
+local protocol, the distributed system as
+a whole will conform to the global protocol.
+
+The Scribble toolchain can convert the local
+protocol into an \textit{endpoint finite state machine}
+(EFSM).
+An EFSM encodes the control flow of the local
+protocol into a communication automaton:
+there is a well-defined initial \textit{state},
+and each \textit{transition} from some state to a
+successor state corresponds to a valid communication
+action (i.e. sending or receiving a message) permitted
+at that endpoint at that state.
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=.7\textwidth]{BGAdderFSM}
+\captionof{figure}{Server EFSM for Adder protocol}
+\label{fig:bgadderfsm}
+\end{figure}
+
+We show the EFSM of the \trole{Server} for the
+\tprotocol{Adder} protocol in \cref{fig:bgadderfsm}.
+We use this as an example to walk through
+the components that build up the EFSM.
+
+\begin{itemize}
+
+\item \textbf{Role Identifiers:}
+The set of roles that the endpoint interacts with.
+For \cref{fig:bgadderfsm}, this is the singleton set
+$\{~\trole{Client}~\}$.
+
+\item \textbf{Label Identifiers:}
+The set of message labels that the endpoint sends or
+expects to receive.
+For \cref{fig:bgadderfsm}, this is set
+$\{~\tmsg{Add}, \tmsg{Res}, \tmsg{Quit}, \tmsg{Terminate}~\}$.
+
+\item \textbf{Payload Types:}
+The set of message payload data types that the endpoint
+sends or expects to receive.
+For \cref{fig:bgadderfsm}, this is singleton set
+$\{~\texttt{number}~\}$.
+
+\item \textbf{Actions:}
+The set of send and receive actions permitted for the endpoint.
+An action is defined in terms of a role identifier,
+label identifier and a vector of payload types.
+For \cref{fig:bgadderfsm}, this is the set of visible
+transitions as illustrated.
+
+$\texttt{Client!Terminate()}$ denotes a \textit{send action}
+that sends the \tmsg{Terminate} message with 
+no payload to \trole{Client}.
+
+$\texttt{Client?Add(int,int)}$ denotes a \textit{receive action}
+that receives the \tmsg{Add} message with two \texttt{int}
+values as payload from \trole{Client}.
+
+\item \textbf{State Identifiers:}
+For \cref{fig:bgadderfsm}, this is the set 
+$\{~ \tmsg{S1}, \tmsg{S2}, \tmsg{S3}, \tmsg{S4}~\}$.
+
+\item \textbf{State Transition Function:}
+The \textit{partial function}, denoted $\delta$ in the literature,
+which maps state identifiers and actions to
+successor state identifiers.
+It is a partial function because not all tuples
+of state identifiers and actions constitute valid transitions
+permitted in the state machine.
+
+For \cref{fig:bgadderfsm},
+(\texttt{S1, Client?Add(int,int)}) is in the domain
+of $\delta$; the successor state is \texttt{S3}.
+Conversely, $\delta($\texttt{S1, Client!Res(int)}$)$
+is undefined.
+\end{itemize}
+
+Developers can use the EFSM to verify their implementation
+for protocol conformance with respect to an endpoint
+in the global protocol.
+Endpoint API generation is a common approach
+\cite{HuJava,Hybrid2016,Python2017,PureScript2019,LinearDecomp}:
+these proposals generate APIs in the target language
+which the developer
+can use to implement their endpoint application that guarantees
+the following properties:
+
+\begin{itemize}
+
+\item \textbf{Behavioural typing:}
+The execution trace of messages sent and received by
+the application is accepted by the EFSM, which means it
+conforms to the communication structure of the global protocol;
+
+\item \textbf{Channel linearity:}
+Each transition in the EFSM represents a channel resource.
+When the application transitions from some state $S$
+to some successor state $S'$, it must no longer
+be able to access a reference (e.g. have an alias) to $S$.
+\end{itemize}
+
+We highlight two distinct proposals for endpoint
+API generation that differ
+by how they leverage features in the target language
+to guarantee the above properties.
+
+\subsubsection{Hybrid Session Verification in Java}
+Hu and Yoshida implemented a workflow in \cite{Hybrid2016}
+that performs \textit{hybrid} verification of 
+communication protocol conformance for Java applications.
+The ``hybrid'' concept is composed of the two main components below:
+
+\subparagraph{Static session typing}
+The EFSM is interpreted under the object-oriented paradigm as follow:
+\textbf{(1)} states are represented as classes;
+\textbf{(2)} supported transitions on each state are 
+represented as instance methods,
+parameterised by the role, label and payload involved 
+in the message exchange. 
+A send method takes the payload to send as a parameter,
+whilst a receive method is a blocking call that 
+requires the caller to allocate a \texttt{Buf<T>} wrapper 
+on the stack (where \texttt{T} is the expected payload type),
+then the receive method populates the payload into 
+the wrapper and returns upon receiving from the channel. 
+These instance methods return a new instance of 
+the successor state class.
+
+\subparagraph{Runtime channel linearity checks}
+By exposing send and receive actions as instance
+methods, the developer can keep aliases of EFSM state instances
+and invoke the IO action more than once. Java does not
+have a linear type system, and hence cannot statically
+verify linear usage; this means linearity checks need to be
+done at runtime.
+Each state keeps track of its usage in a private boolean flag a
+nd throws an exception when the instance method is called twice.
+Similarly, the \texttt{SessionEndpoint} class keeps track of 
+whether the connection is open or close,
+and throws an exception when program execution exits the 
+scope of the session endpoint and a terminal state 
+has yet to be reached.\\
+
+We find that this proposal strikes a good balance 
+between maximising static communication safety guarantees 
+whilst providing an intuitive set of APIs for developers 
+to efficiently write their applications. 
+For example, the EFSM encoding takes advantage of the type system 
+to statically enforce valid transitions; whilst this
+exposes channel resources and linear usage needs to be
+monitored dynamically to compensate for the lack of linear types,
+the generated APIs complement the 
+imperative style of application code written in Java
+
+\subsubsection{Runtime Monitors in Python}
+Neykova and Yoshida targeted the MPST methodology 
+for Python programs in \cite{Python2017} 
+and proposed to generate \textit{runtime monitors} 
+from the EFSM.
+These monitors expose APIs for sending and receiving messages,
+which is used by the developer in their implementation. 
+The runtime monitor is an abstraction between the 
+developer's implementation and the actual communication channel, 
+and ``executes'' the EFSM internally to ensure protocol conformance. 
+When the developer sends a message (with some label and payload) 
+using the API, the runtime monitor checks 
+whether this send action conforms to the current EFSM state, 
+and if so, 
+performs the send and advances to the successor state. 
+Likewise, when the developer invokes a receive, 
+the runtime monitor verifies that this is permitted 
+at the current EFSM state before returning the received payload.
+
+We observe that this approach complements the 
+dynamic typing nature of the Python language, 
+which makes it sensible to perform behavioural typing at runtime.
+As the send and receive IO primitives are
+made available to the developer,
+there are no ``instances'' of channel resources created,
+so the developer cannot explicitly hold a reference to
+some state in the EFSM (let alone keep aliases),
+so channel linearity is trivially guaranteed here.
 
@@
 \subsubsection{API Generation in PureScript}
-\dots
+King et al. presented an approach in \cite{PureScript2019}
+for integrating \textit{multiparty} session types 
+into web development using 
+the PureScript language \cite{PureScript},
+PureScript is a strongly typed functional language
+that compiles to JavaScript.
+This proposal takes advantage of its expressive type system to 
+provide \textit{static} guarantees for protocol conformance.
+We outline the main components of their EFSM encoding:
+
+\subparagraph{Actions as type classes} 
+The semantics of the state transition function 
+in the EFSM express that a (state, action) tuple
+\textit{uniquely defines} a successor state. 
+These semantics can be expressed by 
+\textit{multi-parameter type classes (MPTC) 
+with functional dependencies}: 
+\lstonelineps{class Send r s t a | s -> t r a} 
+defines \texttt{Send} as a MPTC 
+parameterised by recipient \texttt{r}, 
+current state \texttt{s}, successor state \texttt{t} 
+and payload type \texttt{a}; 
+\texttt{s -> t r a} expresses the functional dependency that, 
+for an instance of this type class, 
+the current state uniquely determines the
+successor state, the recipient and payload type. 
+These type classes are independent of the EFSM.
+
+\subparagraph{Transitions as instances of type classes} 
+By encoding states as data types, 
+valid EFSM transitions are encoded as 
+\textit{instances} of the type classes. 
+If \texttt{S1} is an output state, 
+sending an \texttt{Int} to \texttt{Svr} with 
+successor state \texttt{S2}, the PureScript encoding would be
+\lstonelineps{instance SendS1 :: Send Svr S1 S2 Int}.
+Because of the functional dependency, 
+the developer cannot instantiate an invalid transition 
+(e.g. \lstonelineps{Send Svr S1 S3 Bool}, 
+since \texttt{S1} uniquely determines the other type parameters. 
+PureScript statically resolves these type constraints at
+compile-time.
+
+\subparagraph{Channel linearity by construction}
+Whilst PureScript does not support linear types,
+this proposal carefully designed the session runtime using 
+a collection of communication combinators to conceal the
+channel resource, which guarantees channel linearity by construction.\\
+
+This proposal is relevant to the problem we are tackling.
+In particular, the final point about guaranteeing channel linearity
+by construction is something we can build upon in our solution,
+since linear types is a feature left to be desired in
+``mainstream'' programming languages.
+We also observe the challenges of applying session types 
+to the front-end environment, as shown by the careful choice 
+made in \cite{PureScript2019} 
+to use the \textit{Concur UI} framework which 
+builds UI elements sequentially to model sequential sessions.
+Not doing so would require binding channel resources into 
+event listeners on UI elements, which makes linearity violation possible 
+(e.g. by binding a send transition to a button click event,
+the button may still be active after the transition is complete;
+clicking on it again will reuse the channel resource).
+
+As motivated in \cref{section:intro}, \dots
 

### 2020-06-17 report/sections/background/related.tex
Report: complete background

@@
 \subsubsection{Runtime Monitors in Python}
-Neykova and Yoshida targeted the MPST methodology 
+Neykova targeted the MPST methodology 
 for Python programs in \cite{Python2017} 
@@
 We discuss the current state-of-the-art proposals
-for integrating session types in web development.
+for session-typed web development.
 
@@
 ``mainstream'' programming languages.
-We also observe the challenges of applying session types 
-to the front-end environment, as shown by the careful choice 
+We also observe the challenges of session-typed GUI
+programming, as shown by the careful choice 
 made in \cite{PureScript2019} 
-to use the \textit{Concur UI} framework which 
+to use the \textit{Concur UI} framework \cite{ConcurUI} which 
 builds UI elements sequentially to model sequential sessions.
-Not doing so would require binding channel resources into 
-event listeners on UI elements, which makes linearity violation possible 
+Not doing so would make channel linearity violation possible 
 (e.g. by binding a send transition to a button click event,
@@
 clicking on it again will reuse the channel resource).
+Whilst \cite{PureScript2019} supports multiparty session types,
+the communication protocol must describe all interactions
+to go through a centralised server role to respect the
+network topology enforced by WebSocket transport.
 
-As motivated in \cref{section:intro}, \dots
+As motivated in \cref{section:intro}, 
+the functional paradigm and sequential UI framework
+are necessary features for the proposal to statically 
+provide communication safety guarantees, but 
+this type of workflow is not compatible with 
+modern web programming practices and compromises on usability.
 
-\subsubsection{MVU using Links}
-\dots
+\subsubsection{Session-Typed GUI Programming using Links}
+Fowler presented the first formal integration
+of session typing and GUI programming in \cite{MVU2020}.
+His work formalised the \textit{Model-View-Update} (MVU)
+architectural pattern for GUI development.
+An MVU application features a \textit{model} encapsulating application
+state, a \textit{view function} rendering the state on the 
+user interface, and an
+\textit{update function} handling
+\textit{messages} produced by the
+rendered model to produce a new model.
+Fowler introduces the concept of model types
+to express type dependencies between these components:
+a \emph{model type} uniquely defines a \textit{view} function,
+set of \textit{messages} and \textit{update function} --
+rather than producing a new model, the update function 
+defines valid transitions to other model types.
+This approach guarantees channel linearity for GUI programming:
+the update function transitions to a successor model, which only
+renders an interface with the set of messages unique to that model,
+so channel actions from the predecessor model are not rendered,
+which makes reuse impossible.
+The core MVU calculus is also extended with \textit{commands}
+and \textit{linearity} required to express communication
+actions in session type theory.
+
+\cite{MVU2020} is implemented using
+the \textit{Links} web programming language
+\cite{LINKS}. Links is a strict, typed, functional language
+that can be used to write full-stack web applications;
+and similar to PureScript, it is compiled to JavaScript.
+As it has a linear type system, session types can be
+implemented using native language features, since the
+language can monitor linear channel usage.
+Links also supports error handling for session types,
+as introduced in \cite{Exceptional}; 
+given side-effects must be made explicit
+in functional programming, \cite{Exceptional} introduces
+a \textit{session cancellation} operator to ensure
+that channel resources are closed after handling
+side-effects, such as errors.
+
+This work succeeds in 
+retaining the declarative approach for writing
+web interfaces, as the view function incorporates standard
+\textit{Hyper-Text Mark-up Language} (HTML).
+The novel approach of guaranteeing channel linearity
+in GUI programming through the model type abstraction
+is something we can build upon. 
+However, Links does not support multiparty session types,
+and similar to \cite{PureScript2019},
+it can only express server-centric communication structures
+over WebSocket transport.

### 2020-06-17 report/sections/background/typescript.tex
Report: complete background

\section{TypeScript}
\label{section:typescript}

We introduce the TypeScript language 
\cite{UnderstandingTypeScript}
as our choice of target language
for session type API generation.
Developed by \textit{Microsoft Research},
TypeScript is an extension to
JavaScript to address
the deficiencies of the latter in
\textit{developing} and \textit{maintaining}
large-scale complex applications.
Syntactically,
TypeScript is a \textit{superset} of JavaScript,
so every JavaScript program is a
TypeScript program.
The TypeScript Compiler is used
to compile a TypeScript program into JavaScript
source code; since
JavaScript is a dynamically typed langauge,
the compilation process features full type erasure.

We introduce specific language
features used to implement our API generation
solution throughout the report as needed;
here, we highlight the key properties
of the type system implemented in the language.

\paragraph{Structural Typing}
In a structural type system, type equivalence
is determined by \textit{shape} rather than by name
(which is the case in a \textit{nominal} type system).

Consider the following TypeScript code:

\begin{lstlisting}[language=javascript]
class ThisSquare {
	constructor(public side: number) { }
};

class ThatSquare {
	constructor(public side: number) { }
};

declare function area(sq: ThisSquare): number;

area(new ThisSquare(2));	// ok
area(new ThatSquare(2));	// ok (*@\label{line:structural1}@*)
area({ side: 2 });			// ok (*@\label{line:structural2}@*)
\end{lstlisting}

The \texttt{area} function takes a \texttt{ThisSquare}
as parameter.
Under a structural type system,
\cref{line:structural1,line:structural2}
will type-check, because \texttt{ThatSquare} and the object
literal created from scratch matches the \textit{shape}
of \texttt{ThisSquare} -- all of them have a \texttt{side}
property typed \lstonelinejs{number}.

In languages (e.g. Java) that use a nominal type system, 
\cref{line:structural1}
will not type-check because \texttt{ThatSquare} is not named
\texttt{ThisSquare}.

\paragraph{Gradual Typing}
In a gradual type system, a program can have parts
that are statically typed and other parts are dynamically
typed \cite{GradualTyping}.
TypeScript distinguishes dynamically typed code using
the \lstonelinejs{any} type.

\begin{lstlisting}[language=javascript]
// Invoke remote API
fetch('https://jsonplaceholder.typicode.com/todos/1')
	// Convert to JavaScript Object Notation
	.then((response: Response) => response.json())
	.then((json: any) => {
		// Up to the developer to correctly deserialise;
		// incorrect implementations will cause
		// runtime type error.
	});
\end{lstlisting}

The rationale for this decision in \cite{UnderstandingTypeScript}
is that, JavaScript programs tend to interact with
data of unspecified types (such as fetching data
from API calls over the network); these parts
need to be dynamically typed in order to give developers
a smooth transition into TypeScript, and for TypeScript
to be usable in a distributed system setting. \\

\noindent
Based on the compatibility of TypeScript
with JavaScript, 
we believe that TypeScript API generation for session-typed
web development best achieves our objective
of providing
developers with a workflow
that provides communication safety guarantees in \textit{modern
web programming}
through multiparty session types.
We argue that the type system of TypeScript,
along with other language features we introduce
throughout the course of the report, is sufficient
for implementing session type theory in a manner that
complements idiomatic web development practices.

### 2020-06-17 report/sections/server/react/approach.tex
Report: complete background

@@
 their work on \textit{multiple model types} motivated by the
-\textit{Model-View-Update} architecture (MVU).
-An MVU application features a \textit{model} encapsulating application
-state, a \textit{view function} rendering the state on the 
-Document Object Model (DOM), and an
-\textit{update function} handling \textit{messages} produced by the
-rendered model to produce a new model.
+\textit{Model-View-Update} architecture (MVU),
+introduced in \cref{section:bgrelated}.
 The concept of model types express type dependencies between these
@@
 
-We leverage the correspondence between model types and states in the EFSM:
+We leverage the correspondence between model types 
+and states in the EFSM:
 each state in the EFSM is a model type, the set of messages represent

### 2020-06-17 report/sections/abstract.tex
Report: update AIC

@@
 \noindent
+Modern web programming
+involves coordinating interactions
+between browser clients and a web server.
+Typically, the interactions are
+informally described, which makes
+it difficult to verify communication
+correctness in the distributed system.
+
 Multiparty Session Types (MPST)
@@
 industrial practices. The generated APIs
-build upon TypeScript concurrency primitives,
+build upon TypeScript concurrency practices,
 complement the event-driven style of programming
@@
 using case studies of protocols found in web-services,
-and analyse the overhead of our implementation through 
-running micro-benchmarks against a baseline implementation.
+and analyse the performance overhead
+through 
+running benchmarks against a baseline implementation.
 

### 2020-06-17 report/sections/evalconcl/concl/contributions.tex
Report: update AIC

@@
 Developers write their full-stack applications
-by implementing the generated APIs, and can enjoy 
+by implementing the generated APIs, and enjoy 
 communication safety guarantees by construction.
-We explain how our session type encodings target the Node.js
-server-side runtime and React.js framework in
-\cref{chap:node,chap:react} for server-side
-and browser-side endpoints respectively.
-In particular, we present a novel approach for integrating
+APIs generated for server-side endpoints
+are compatible with the Node.js runtime,
+as detailed in \cref{chap:node}.
+For browser-side endpoints,
+we present a novel approach for integrating
 session types into web-based GUI programming based on
 translating the theory on model types \cite{MVU2020}
-to idiomatic practices on the React.js framework.
+to idiomatic practices on the React.js framework;
+this is highlighted in \cref{chap:react}.
 With respect to session type theory, implementations using
@@
 we are not only able to
-\textbf{statically} provide the same level of communication safety
+statically provide the same level of communication safety
 guarantees from multiparty session type theory, but we do so
@@
 Computer Science (EPTCS) \cite{PLACES2020}.
-We were also invited to present this work
+We were also invited to present a part of this work
 at the 12th International Workshop on
@@
 further reveals the potential
-for implementing dynamic communication structures over centralised
+for implementing dynamic communication structures 
+over centralised
 network topologies.

### 2020-06-17 report/sections/evalconcl/concl/future.tex
Report: update AIC

@@
 \paragraph{Explicit Connection Actions}
-Hu and Yoshida \cite{ExplicitConnections} extends 
-multiparty session type theory
+Hu and Yoshida \cite{ExplicitConnections} extended 
+MPST theory
 with \textit{explicit connection actions} to support protocols 
@@
 The MPST-based web development framework presented by 
-King et al. \cite{PureScript2019} defines the session runtime
-in a way that parameterises the transportation mechanism.
+King et al. \cite{PureScript2019} parameterised
+the transport mechanism in the session runtime definition.
 The developer can run the session over different transportation,

### 2020-06-17 report/sections/introduction/contributions.tex
Report: update AIC

@@
 guarantees using familiar industry tools in TypeScript and React,
-and supports peer-to-peer browser interactions over WebSocket transport.
+and supports peer-to-peer browser interactions over WebSockets.
 
@@
 and provide an overview of related work on integrating session types into
-application development. We also introduce TypeScript and highlight some 
-interesting features and caveats in the language.
+application development. We also introduce TypeScript and highlight the 
+defining features of the language.
 

### 2020-06-17 report/sections/introduction/motivation.tex
Report: new introduction
Report: update AIC

@@
-%\subsubsection{New Structure (similar to ECOOP2017):}
-%\begin{itemize}
-%
-%\item \textbf{Section 1.1:}
-%
-%Interactive webapps over WebSockets, harder to guarantee communication safety
-%
-%\item
-%Example protocol similar to travel agency  -- motivate that it is difficult to manually verify communication safety
-%
-%\begin{itemize}
-%\item traveller enquires availability, flight server replies yes/no then traveller confirms/denies.
-%\item The flight server will reserve the seat until the traveller denies.
-%\item Source of hidden bugs -- traveller may enquire twice (e.g. click twice), but server processes one enquiry, so one seat is held up for no reason
-%\end{itemize} 
-%
-%\item MPST framework can be applied for webapps.
-%This is solved by existing work PureScript and MVU2020.
-%
-%\item \textbf{Problem1}: 
-%
-%existing work provides a workflow for session typed web development, but requires the developer to compromise by adapting to programming paradigms and practices not conventional for modern web programming.
-E.g.
-functional paradigm, sequential UI framework, PureScript and LINKS not interoperable with the ever-growing universe of JavaScript libraries.
-These compromises limit the usability of MPST framework
-%for web development.
-%
-%\item \textbf{PRoblem2}: 
-%
-%existing work based on WebSocket transport, which
-%is simple to reason about. But WebSocket transport enforces a server-centric
-%network configuration, so existing work isn't expressive enough for
-%more general protoocls. Suppose traveler contacts agency first to get a quote and then compare (similar to 2-buyer). Existing work cannot implement this.
-%
-%\item \textbf{Section 1.2: }
-%
-%Our objective is to present a novel variant of MPST framework
-%that is compatible / supports the flexibility of modern web programming.
-%We address the two problems in two parts:
-%
-%\item We solve problem 1 in part 1: 
-%
-%SessionTS is API generation workflow based on MPST, which
-%provides developers with same communication safety guarantees in a way that is idiomatic for modern web programming.
-We use industry tools (TS/Node/REact), compatible with async logic, error handling specific to web (e.g.
-client disconnection)
-%
-%\item We solve problem 2 in part 2:
-%
-%develop novel theory for routed MPST which lets us implement non-server-centric commuication structures over a server-centric network topology, prove
-%that our theory preserves properties like communication safety, deadlock freedom, etc.
-%
-%\item Section 1.3: diagram similar to FZ for outline
-%
-%\end{itemize}
-%
-%% Aim of introductio: motivate what is the problem 
-%%(session types fix web, PureScript and LINKS have two problems: 
-%%implementation -- not using mainstream stuff, 
-%%theory -- websocket transport restricts protocols)
-%
-
 \section{Motivation}
@@
 
-\paragraph{The need for communication safety 
-in interactive web applications}
+\paragraph{Communication Safety 
+in Interactive Web Applications}
 
@@
 application.
-The \trole{FlightServer} can be a web server
+\trole{FlightServer} can be a web server
 running on the \textit{Node.js} JavaScript runtime \cite{nodejs}.
-The \trole{Traveller} can run on the web browser
+\trole{Traveller} can run on the web browser
 as a \textit{single-page application} (SPA) written
@@
 booking operations outlined above.
+We visualise the interactions in \cref{fig:flightbook}
+
+\begin{figure}[!ht]
+\centering
+\includegraphics[width=0.7\textwidth]{FlightBookingService}
+\captionof{figure}{Message-Passing Architecture of Web-Based 
+Flight Booking Service}
+\label{fig:flightbook}
+\end{figure}
 
@@
 even for this simple example:
-we highlight just a few:
 
@@
 correctness with respect to the communications aspect.
+
 By specifying the client-server interactions of 
@@
 adapt the MPST framework for web development
-acknowledge the limitations of JavaScript 
+acknowledge the limitations of \textit{JavaScript} 
 -- the language of the browser --
@@
 
-The current state-of-the-art succeeds in \textbf{statically}
+The current state-of-the-art succeeds in statically
 providing communication safety guarantees,
@@
 and closes their browser prematurely and the \trole{FlightServer}
-isn't notified of this event, the reserved seat will not be
-released.
+is not notified of this event, 
+the reserved seat will not be released.
 These compromises limit the usability of the MPST framework

### 2020-06-17 report/sections/introduction/objectives.tex
Report: update AIC

@@
 developers with a development workflow
-that provides communication safety guarantees in modern
-web programming through multiparty session types.
+that \textit{provides communication safety guarantees in modern
+web programming} through multiparty session types.
 
 We tackle the limitations of the current
-state-of-the-art with a two-step strategy:
+state-of-the-art with two solutions:
 
@@
 to communication safety: 
-there will not be any communication mismatches,
-and channel linearity will be respected.
+there will \textit{not} be any \textbf{deadlocks} or
+\textbf{communication mismatches},
+and \textbf{channel linearity} will be respected.
 
@@
 theory that supports \textit{routed communication}.
+We prove that \newtheory also extends the communication safety
+properties of the canonical MPST theory, so
+processes that implement \newtheory also enjoy
+guarantees of \textbf{deadlock-freedom}, \textbf{channel linearity},
+and the absence
+of \textbf{communication mismatches}.
 
@@
 
-By step \textbf{S1}, we address the limitation \textbf{L1}.
+By solution \textbf{S1}, we address the limitation \textbf{L1}.
 Our work provides the same communication safety guarantees
@@
 
-In step \textbf{S2},
+In solution \textbf{S2},
 we show that routing communication interactions through an intermediate

### 2020-06-17 report/sections/server/codegen.tex
Report: review chapter 3

@@
-\chapter{\fancyname{SessionTS}: Session Type API Generation for TypeScript}
+\chapter{\codegen: Session Type API Generation for TypeScript}
 \label{chap:codegen}
@@
 In this chapter, we present
-\fancyname{SessionTS},
+\codegen,
 a toolchain for generating TypeScript APIs that developers can
@@
 
-\fancyname{SessionTS} supports communication protocols that define 
+\codegen supports communication protocols that define 
 \emph{server-centric} topologies, meaning that: 
@@
 \textbf{(3)} all non-server participants only communicate with the server.
-Examples of such protocols include the \tprotocol{Noughts and Crosses}
-multiplayer game (\cref{section:evalgame}) and the \tprotocol{ATM} protocol.
+An example would be the \tprotocol{Noughts and Crosses}
+multiplayer game (\cref{lst:game}).
 We relax this assumption in \cref{part:general}.

### 2020-06-17 report/sections/server/codegen/implementation.tex
Report: review chapter 3

@@
 
-\fancyname{SessionTS} is written in Python. 
+\codegen is written in Python. 
 It offers flexible syntax, a rich standard library, and 
@@
 supply the correct information (\cref{lst:cmdline})
-to use \fancyname{SessionTS},
+to use \codegen
 and the \texttt{subprocess} package to 
@@
 \end{lstlisting}
-\captionof{lstlisting}{Entry Point for \fancyname{SessionTS}}
+\captionof{lstlisting}{Entry Point for \codegen}
 \label{lst:cmdline}
@@
 We use Docker \cite{docker} to encapsulate our code generator
-and its dependencies -- we found the canonical Python virtual environment
+and its dependencies -- 
+we found the canonical Python virtual environment
 solution to be insufficient, as the Scribble toolchain is

### 2020-06-17 report/sections/server/codegen/workflow.tex
Report: review chapter 3

@@
 We motivate our development workflow from previous work 
-\cite{PureScript2019, Hybrid2016} by extending the Scribble toolchain
+\cite{PureScript2019, Hybrid2016, LinearDecomp} 
+by extending the Scribble toolchain
 and generating APIs that integrate the developer's 
@@
 and the code generation \textit{target}
-(i.e. whether the role runs on the server or the web browser).
+(i.e. whether the role runs on the back-end web server or 
+the front-end web browser).
 
-\item \fancyname{SessionTS} delegates to the 
+\item \codegen delegates to the 
 Scribble toolchain for verifying the well-formedness of
@@
 the endpoint FSM (\cref{subsection:efsm}). 
-\fancyname{SessionTS} parses the endpoint's 
+\codegen parses the endpoint's 
 interactions from the DOT graph and generates TypeScript APIs
@@
 \includegraphics[width=\textwidth]{DevelopmentWorkflow}
-\captionof{figure}{Overview of \fancyname{SessionTS} Development Workflow}
+\captionof{figure}{Overview of \codegen Development Workflow}
 \label{fig:devworkflow}
@@
 in mainstream programming languages
-\cite{Hybrid2016, PureScript2019, Python2017}. 
+\cite{Hybrid2016, PureScript2019, Python2017, LinearDecomp}. 
 We use the variant of the Scribble language 
-previously introduced in \cref{subsection:bgscribble}.
+discussed in \cref{subsection:bgscribble}.
 
 \subparagraph{Type declaration statements}
-Specific to our TypeScript API generation toolchain,
+Specific to \codegen,
 the developer is \textit{not} required to explicitly
@@
 Scribble protocol as far as 
-\fancyname{SessionTS} is concerned. 
-Internally, \fancyname{SessionTS} inspects the protocol file
+\codegen is concerned. 
+Internally, \codegen inspects the protocol file
 and parses existing type declarations using regular expressions
@@
 custom data types that will appear in the communication (for example,
-\cref{lst:game}), and allows \fancyname{SessionTS} to inject
+\cref{lst:game}), and allows \codegen to inject
 ``boilerplate'' type declarations for built-in TypeScript types before
@@
 Scribble expresses the EFSM using the 
-DOT graph description language \cite{dot}, with each
+\textit{DOT} graph description language \cite{dot}, 
+with each
 communication action encoded as the label of the corresponding
 state transition. 
-\fancyname{SessionTS} uses the pydot library \cite{pydot}
+\codegen uses the \textit{pydot} library \cite{pydot}
 to parse the graph into an internal representation of the EFSM.
@@
 
-For \fancyname{SessionTS}, we leverage the Jinja \cite{jinja} 
+For \codegen, we leverage the \textit{Jinja} \cite{jinja} 
 template engine library for code generation. 
-We first construct templates for the TypeScript files we wish to generate,
+We first construct templates for the TypeScript files 
+we wish to generate,
 specifying placeholders for dynamic content (to be extracted
@@
 control structures 
-(such as for loops, to dynamically render the enclosing ``sub-template''
+(such as for loops, to dynamically render the 
+enclosing ``sub-template''
 by iterating over a collection).
@@
 on the specified target, we structure the different code generators
-using the Strategy design pattern. Each target extends the
+using the Strategy design pattern. 
+Each target extends the
 abstract base class \texttt{CodeGenerationStrategy} 

### 2020-06-17 report/sections/evalconcl/concl/contributions.tex
Report: address NY comments

@@
 Our work has received positive feedback from academia.
-The project has been accepted and published as 
-\emph{Generating Interactive WebSocket Applications in TypeScript}
-in Volume 314 of the Electronic Proceedings in Theoretical
-Computer Science (EPTCS) \cite{PLACES2020}.
+The intitial part of this project was published as 
+\emph{Generating Interactive WebSocket 
+Applications in TypeScript} in Volume 314 of the Electronic Proceedings
+in Theoretical Computer Science (EPTCS) \cite{PLACES2020},
+in which I am the first and lead author.
+The materials in \cref{chap:codegen,chap:node,chap:react}
+are expanded from \cite{PLACES2020}.
 We were also invited to present a part of this work

### 2020-06-17 report/sections/introduction/contributions.tex
Report: address NY comments

@@
 \begin{remark}
-This project was published as \emph{Generating Interactive WebSocket 
+A part of this project was published as 
+\emph{Generating Interactive WebSocket 
 Applications in TypeScript} in Volume 314 of the Electronic Proceedings
@@
 The materials in \cref{chap:codegen,chap:node,chap:react}
-are revised from \cite{PLACES2020}.
+are expanded from \cite{PLACES2020}.
 The materials in 

### 2020-06-17 report/sections/server/node.tex
Report: update graphic placement
Report: complete chapter 4 sans connection
Report: review chapter 4

@@
-\chapter{\fancyname{NodeMPST}: Back-End Session Type Web Development}
+\chapter{\nodecodegen: Back-End Session Type Web Development}
 \label{chap:node}
 
-In this chapter we present \fancyname{NodeMPST},
+In this chapter we present \nodecodegen,
 our session type API generation strategy for 
-server-side endpoints implemented on the Node.js runtime \cite{node}.
+server-side endpoints implemented on the Node.js runtime 
+\cite{nodejs}.
 We continue to use the running example of the \tprotocol{Adder}
-protocol, and refer to the FSM of the \trole{Svr} endpoint
+protocol (\cref{lst:adder}),
+and refer to the FSM of the \trole{Svr} endpoint
 (\cref{fig:addersvrfsm}) throughout this chapter.
+
+We discuss the challenges of implementing
+session types on the Node.js runtime 
+(\cref{section:nodechallenges})
+and motivate our approach (\cref{section:nodeapproach}).
+We explain our design choices for encoding
+the EFSM for server-side endpoints (\cref{section:nodeefsm}),
+and detail the session runtime that is designed
+to execute the EFSM on Node.js (\cref{section:noderuntime}).
+We conclude by analysing some alternative designs 
+(\cref{section:nodealt}) and evaluating the limitations 
+of our approach (\cref{section:nodelimitations}).
 
@@
 \centering
-\includegraphics[width=0.5\textwidth]{AdderSvrFSM}
+\includegraphics[width=0.6\textwidth]{AdderSvrFSM}
 \captionof{figure}{\trole{Svr} Endpoint FSM
@@
 \input{sections/server/node/limitations}
+
+\section{Summary}
+In this chapter, we have presented
+our session type API generation strategy
+which targets the Node.js runtime for server-side
+endpoints.
+
+We motivated our approach of generating
+handler-style APIs that act as seams in the session runtime
+which executes the EFSM.
+By doing so, the mechanism for sending and receiving
+messages are not exposed to the developer,
+which makes channel reuse impossible by construction.
+
+We discussed our session runtime implementation in great
+depth and iterated upon the design to arrive at 
+a minimal runtime implementation that executes the EFSM
+using a message-passing abstraction internally as well.
+In particular, we motivated how to handle message receives
+in a way that preserves the message ordering specified
+in the communication protocol.
+
+Developers that implement their Node.js endpoint application
+using the generated APIs enjoy protocol conformance
+by construction: 
+the TypeScript type system will prevent type mismatches
+in the handlers defined by the developer
+(e.g. sending a \lstonelinejs{string} instead of 
+a \lstonelinejs{number}, or using a message label
+that is not defined on the EFSM state),
+and the handler-style APIs themselves conceal channel
+resources to prevent channel reuse.

### 2020-06-17 report/sections/server/node/approach.tex
Report: review chapter 4

@@
 we do not expose send and receive APIs to the developer,
-making it impossible for the developer to reuse channels.
+contrary to \cite{HuJava},
+making it impossible for the developer to reuse channels,
+hence saving the need for linearity checks.
 
@@
 
-When executing \fancyname{SessionTS} to generate code for the
+When executing \nodecodegen to generate code for the
 \trole{Svr} endpoint specifying the \texttt{node} target,

### 2020-06-17 report/sections/server/node/challenges.tex
Report: review chapter 4

@@
 
-Our goal with \fancyname{NodeMPST} is to
+Our goal with \nodecodegen is to
 generate session type implementations that
@@
 and the single-threaded event loop model in Node.js
-As outlined in \cref{section:typescript}, TypeScript's type system
+TypeScript's type system
 is neither affine nor linear, so we need to work around the 

### 2020-06-17 report/sections/server/node/efsm.tex
Report: progress on chapter 4 receive edge case
Report: complete chapter  4 runtime
Report: minor changes to chapter 4
Report: review chapter 4

@@
 
-We show the structure of the generated EFSM.ts file in
+We show the structure of the generated \filename{EFSM.ts}
+file in
 \cref{lst:nodeefsmfile}.
 Note that the formal definition of the EFSM in 
-\cref{section:scribbleefsm}
+\cref{subsection:bgscribble}
 contains more than just states and the state transition function,
@@
 // (*@\cref{subsection:nodeefsmroleslabelsmsg}@*)
-export namespace Roles {...};
-export namespace Labels {...};
-export namespace Message {...};
+export namespace Roles {...};			// for roles in protocol
+export namespace Labels {...};		// for message labels
+export namespace Message {...};		// for message structure
 
 // (*@\cref{subsection:nodeefsmhandlers}@*)
-export namespace Handler {...};
+export namespace Handler {...};		// for handler APIs
 
@@
 export namespace Implementation {...};
-
-export type EfsmTransitionHandler =
-	(implementation: Implementation.Type) => void;
-export type MessageHandler = (message: any) => void;
 \end{lstlisting}
@@
 need to refer to the label identifier and payload type: 
-we refer to this as the message structure, and encode this as a
-Message Type. 
-Each message is expressed as an interface with
+we refer to this as the message structure.
+Each message structure is expressed as an interface with
 properties for the label and payload.
@@
 \begin{figure}[!ht]
-\begin{lstlisting}[language=javascript, tabsize=2]
+\begin{lstlisting}[language=javascript]
 // Inside the Message namespace...
@@
 
-We collect the APIs that the developer needs to implement
-under the \texttt{Handler} namespace. 
-As a design choice, we \textit{do not} generate handlers for
-terminal states, because the semantics of inactivity mean
-there is nothing to handle.
+Handler APIs act as the \textit{seams} of the EFSM.
+A seam is ``a place where one can alter the
+behaviour in the program without editing in that place''
+\cite{seam}. The developer implements handlers
+to define what message to send and how to process a received
+message, without access to the send and receive APIs
+that are used to execute the EFSM.
+We collect these APIs under the \texttt{Handler} namespace. 
 We introduce the generated handlers for sending and receiving states.
@@
 of the successor state.
+As a design choice, we \textit{do not} generate handlers for
+terminal states, because the semantics of inactivity mean
+there is nothing to handle.
 
@@
 would resolve to \lstonelinejs{[string]},
-based on the interface definition \cref{lst:addersvrmsg}.
+based on the interface definition in \cref{lst:addersvrmsg}.
 
@@
 // Inside the Handler namespace...
-export type S51 = {
+export interface S51 {
 	[Labels.S51.ADD]: (...payload: Message.S51ADD['payload']) =>
@@
 the message payload, and must return the successor state encoding.
-We see another benefit of defining the payload in Message Types
+We see another benefit of defining the payload 
+in the message structure interface
 as a tuple: we can define the receive handler parameter
@@
 \begin{lstlisting}[language=javascript,tabsize=2]
+// More intuitive
 const withSpread = (x: number, y: number) => {...} (*@\label{line:yesspread}@*)
+
+// Needs manual destructuring
 const withoutSpread = (payload: [number, number]) => {...} (*@\label{line:nospread}@*)
@@
 The behaviour of the runtime is dependent on the current state,
-so it needs a way to distinguish between
+so the runtime needs a way to distinguish between
 all the different states -- one can think of this as implementing
@@
 overloading a \texttt{next()} method for each state.
-Due to limitations in the TypeScript language, this would have to
-be some sort of switch statement, with the \texttt{next()}
-method parameterised by some base type assignable to all states.
+TypeScript does not support method overloading,
+so the \texttt{next()} method would need to be defined
+on some base type assignable to all states,
+and the implementation of the \texttt{next()} method 
+can use a switch statement to distinguish between the different states.
 Currently, the state is only determined by the handler
@@
 \subsubsection{Distinguishing Handlers using Conditional Types}
+\label{subsubsection:dependenttypes}
 
@@
 an enum for receive states)
-upon which to execute the EFSM, which solves the switch statement
-problem.
+upon which to execute the EFSM. 
+This can solve the switch statement problem.
 Now, we are left with defining a mapping between the 
@@
 \begin{lstlisting}[language=javascript,tabsize=2]
-enum SendState { S1, S3, ... };
-enum ReceiveState { S2, S4, ... };
+enum SendState { S53, S54 };
+enum ReceiveState { S51 };
 type State = SendState | ReceiveState;
@@
 type SendHandler<S extends SendState> = 
-	S extends SendState.S1 ? Handler.S1 :
-	S extends SendState.S3 ? Handler.S3 : ... ;
+	S extends SendState.S53 ? Handler.S53 :
+	S extends SendState.S54 ? Handler.S54 : never ;
 
 type ReceiveHandler<S extends ReceiveState> = 
-	S extends ReceiveState.S2 ? Handler.S2 :
-	S extends ReceiveState.S4 ? Handler.S4 : ... ;
+	S extends ReceiveState.S51 ? Handler.S51 : never;
 \end{lstlisting}
@@
 union type (which is the case for enums, as
-\texttt{S = SendState.S1 | SendState.S3 | ...}), where
+\texttt{S = SendState.S53 | SendState.S54| ...}), where
 
@@
 
+We illustrate this with an example below.
+A \texttt{Shape} is either a \texttt{Circle} or \texttt{Square}.
+Both define a property name \texttt{type}, which is the discriminant
+property, along with properties unique to their shape
+(i.e. radius for circles, side lengths for squares).
+The \texttt{area()} function uses the discriminant property
+(here, in a switch statement) to narrow the type of \texttt{s}
+and perform operations specific to that shape.
+
+\begin{lstlisting}[language=javascript]
+// The `type` property is the discriminant
+interface Circle { type: 'Circle', radius: number };
+interface Square { type: 'Square', side: number };
+type Shape = Circle | Square;
+
+function area(s: Shape) {
+	switch (s.type) {
+		case 'Circle': return 3.14 * s.radius;	// s: Circle
+		case 'Square': return s.side * s.side;	// s: Square
+	}
+}
+\end{lstlisting}
+
 For the time being,
@@
 \begin{lstlisting}[language=javascript,tabsize=2]
-abstract class ISend { type: 'Send' = 'Send'; ... }
-abstract class IReceive { type: 'Receive' = 'Receive'; ... }
-abstract class ITerminal { type: 'Terminal' = 'Terminal'; ... }
+// The `type` property is the discriminant
+abstract class ISend { type: 'Send' = 'Send'; ... } (*@\label{line:nodeefsmsend}@*)
+abstract class IReceive { type: 'Receive' = 'Receive'; ... } (*@\label{line:nodeefsmreceive}@*)
+abstract class ITerminal { type: 'Terminal' = 'Terminal'; ... } (*@\label{line:nodeefsmterminal}@*)
 
@@
 	export class S51 extends IReceive {
+
+		// Stores the handler defined by the developer
+		// as a private property	
 		constructor(private handler: Handler.S51) { super(); }
@@
 	...	
-
 };

### 2020-06-17 report/sections/server/node/runtime.tex
Report: fix subfig numbering
Report: review chapter 4

@@
 We define the session runtime for the \trole{Svr} endpoint
-of the \tprotocol{Adder} protocol in Svr.ts,
+of the \tprotocol{Adder} protocol in \filename{Svr.t}s,
 named after the endpoint.
@@
 \begin{lstlisting}[language=javascript]
+// Exported to developer
 export class Svr {
@@
 
+// Mapping of roles to WebSocket connections
 const roleToSocket: Partial<RoleToSocket> = { (*@\label{line:initroletosocket}@*)
@@
 
+// Invoked when a connection request is received
 const onSubscribe = ({ data, target: socket }) => {
+
+	// Deserialise connection request message
 	const { connect: role } = JSON.parse(data) 
 		as Message.ConnectRequest; (*@\label{line:connectionrequest}@*)
+
+	// Ignore if role is aleady taken
 	if (!waiting.has(role)) { return socket.close(); } (*@\label{line:occupied}@*)
 
+	// Map the role in the connection request to this WebSocket
 	roleToSocket[role] = socket; (*@\label{line:connectwsbind}@*)
@@
 
-	if (waiting.size === 0)
+	// Start executing EFSM when all roles have joined
+	if (waiting.size === 0) {
 		new Session(wss, roleToSocket as RoleToSocket, initialState); (*@\label{line:newsession}@*)
+	}
 };
 
+// For every new connection, process message with `onSubscribe`
 wss.addEventListener('connection', ws => {
@@
 next(state: Implementation.Type) {
+	// Distinguish between states using discriminant property
 	switch (state.type) {
@@
 			const [label, payload, succ]: (*@\hl{???}@*) = state.handler;	 (*@\label{line:noderuntimebadtype1}@*)
-			this.send((*@\hl{???}@*), label, payload); (*@\label{line:noderuntimeunknownrole}@*)		
+			this.send((*@\hl{???}@*), label, payload);	// Who to send to? (*@\label{line:noderuntimeunknownrole}@*)		
 			return this.next(succ);
@@
 		case 'Receive': {
+			// Handle incoming messages using this anonymous function
 			this.wss.onmessage = ({ data }) => {
+
+				// Which message structure to use to deserialise?
 				const { label, payload } = JSON.parse(data) as (*@\hl{???}@*); (*@\label{line:noderuntimebadtype2}@*)
@@
 	switch (state.type) {
-		case 'Send':
+		case 'Send':				// recall (*@\cref{lst:nodeefsmimplementation} \cref{line:nodeefsmsend}@*)
 			return state.performSend(this.next, this.send);
-		case 'Receive':
+		case 'Receive':			// recall (*@\cref{lst:nodeefsmimplementation} \cref{line:nodeefsmreceive}@*)
 			return state.prepareReceive(this.next, this.register);
-		case 'Terminal':
+		case 'Terminal':		// recall (*@\cref{lst:nodeefsmimplementation} \cref{line:nodeefsmterminal}@*)
 			return;
@@
 directly from the handler implemented by the developer, 
-accurately typed by how we define the handler API in EFSM.ts.
+accurately typed by how we define the handler API in \filename{EFSM.ts}.
 The developer does not need to specify which role
@@
 class Session {
-	private handler?: MessageHandler; (*@\label{line:noderuntimehandler}@*)
+	// Optional type, same as `MessageHandler | undefined`
+	private handler?: MessageHandler; | undefined(*@\label{line:noderuntimehandler}@*)
 
@@
 		...
+		// Process incoming messages from each WebSocket
+		// using `this.receive()`
 		Object.values(this.roleToSocket).
 			.forEach(ws => ws.onmessage = this.receive.bind(this)); (*@\label{line:wsbind}@*)
+
+		// Initialise handler as undefined
 		this.handler = undefined;
@@
 
+	// Set the handler to be used to handle the next receive event
 	register(handler: MessageHandler) { this.handler = handler; } (*@\label{line:register}@*)
@@
 		const handler = this.handler; (*@\label{line:callhandlerstart}@*)
+
+		// `Unregister` the current receive handler
 		this.handler = undefined; (*@\label{line:callhandler2}@*)
+
+		// `handler` has an optional type, so we first
+		// check if there is a value set, then invoke the handler.
+		// Same as `if (handler !== undefined) { handler(data); }`
 		handler?.(data); (*@\label{line:callhandler}@*)
@@
 \end{lstlisting}
-\captionof{lstlisting}{Approach to Dynamic WebSocket Message Event Listener}
+\captionof{lstlisting}{Attempt to Dynamic WebSocket Message Event Listener}
 \label{lst:noderuntimewsmsg}
@@
 	) {
+		// Define dynamic WebSocket message event listener
 		const messageHandler = (message: any) => {
+
+			// Deserialise message
 			const decoded = JSON.parse(message) as Message.S51;
+
+			// Discriminate between messages using label
 			switch (decoded.label) {
 				case Labels.S51.ADD: {
+					// Invoke handler to get successor state
 					const successor = 
 						this.handler[decoded.label](...decoded.payload);
+
+					// Invoke callback to advance EFSM
 					return next(successor);
@@
 		}
+
+		// Register the message handler under the
+		// WebSocket bound to the `Client` role
 		register(Roles.Peers.Client, messageHandler);
@@
 \end{subfigure}
-\captionof{figure}{Possible Orderings for Handling Message Event and Preparing
+\caption{Possible Orderings for Handling Message Event and Preparing
 Receive State}
@@
 (\cref{line:mapped1,,line:mapped2}), 
-using a generic mapped typed defined in
-\cref{lst:noderolesmappedtypes}.
+using a generic mapped typed defined below.
+\begin{lstlisting}[language=javascript]
+// Inside the Roles namespace...
+export type PeersToMapped<Value> = { [Role in Peers]: Value };
+\end{lstlisting}
 
@@
 
+		// Set up empty queues
 		this.messageQueue = { [Roles.Peers.Client]: [], }; (*@\label{line:queue1}@*)
@@
 
-		this.next(initialState);	
+		this.next(initialState);		// Advance to initial state
 	}
@@
 	receive(from: Roles) { (*@\label{line:receivefunc}@*)
+
+		// Return a WebSocket message event listener
+		// that looks at the message/handler queue
+		// specific to the `from` role
+
 		return ({ data }) => { (*@\label{line:receivefuncret}@*)
@@
 \label{lst:nodesession}
-\end{figure}
-
-\begin{figure}[!h]
-\begin{lstlisting}[language=javascript]
-// Inside the Roles namespace...
-export type PeersToMapped<Value> = { [Role in Peers]: Value };
-\end{lstlisting}
-\captionof{lstlisting}{Generic Mapped Type for Roles}
-\label{lst:noderolesmappedtypes}
 \end{figure}

### 2020-06-17 report/sections/appendix/eval.tex
Report: new conclusion
Report: feedback

@@
 
+\begin{lstlisting}[language=javascript,tabsize=2]
+/**
+ * User implementation of game board state.
+ * Export game board instance.
+ */
+
+import { Coordinate as Point } from './Game/GameTypes';
+
+enum Cell { Empty, P1, P2 };
+export enum MoveResult { Win, Draw, Continue };
+
+class Board {
+
+	private _board: Array<Array<Cell>>;
+	private _emptyCellCount: number;
+
+	constructor() {
+		this._board = [[Cell.Empty, Cell.Empty, Cell.Empty],
+		[Cell.Empty, Cell.Empty, Cell.Empty],
+		[Cell.Empty, Cell.Empty, Cell.Empty]];
+		this._emptyCellCount = 9;
+	}
+
+	// Factory for generating a function that, given a player role,
+	// places the marker and returns the game result with respect to 
+	// that player.
+	private _makeMove = (marker: Cell) => ({ x: row, y: col }: Point) 
+		=> {
+		return new Promise<MoveResult>((resolve, _) => {
+			setTimeout(() => {
+				// Update board state
+				this._board[row][col] = marker;
+				this._emptyCellCount--;
+
+				// Check for winning move
+				if (this._board[row].every(cell => cell === marker))
+					resolve(MoveResult.Win);
+
+				/* Winning column */
+				if (this._board.every(row => row[col] === marker))
+					resolve(MoveResult.Win);
+
+				/* Placed middle marker - check corners */
+				if (row === 1 && col === 1) 
+					if ((this._board[0][0] === marker 
+						&& this._board[2][2] === marker) ||
+						(this._board[0][2] === marker 
+						&& this._board[2][0] === marker))
+						resolve(MoveResult.Win);
+
+				/* Placed corner marker - check diagonals */
+				if (row !== 1 && col !== 1) 
+					if (this._board[1][1] === marker 
+						&& this._board[2 - row][2 - col] === marker)
+						resolve(MoveResult.Win);
+
+				resolve(this._emptyCellCount === 0 ? 
+					MoveResult.Draw : MoveResult.Continue);
+			}, 2000);
+		})
+
+	}
+
+	p1(move: Point) { return this._makeMove(Cell.P1)(move); }
+	p2(move: Point) { return this._makeMove(Cell.P2)(move); }
+
+	clear() {
+		this._board = [[Cell.Empty, Cell.Empty, Cell.Empty],
+		[Cell.Empty, Cell.Empty, Cell.Empty],
+		[Cell.Empty, Cell.Empty, Cell.Empty]];
+		this._emptyCellCount = 9;
+	}
+
+}
+
+// Initialise state
+export const board = new Board();
+\end{lstlisting}
+
 % BOARD CLASS
 
-\section{Implementation for \tprotocol{Two Buyer}}
-\label{section:evalcodetwobuyer}
+%\section{Implementation for \tprotocol{Two Buyer}}
+%\label{section:evalcodetwobuyer}
 
-\section{Implementation for \tprotocol{ATM}}
-\label{section:evalcodeatm}
+\section{Package Dependencies for Benchmarks}
+\label{section:evalbenchmarkpackagejson}
+
+\begin{lstlisting}[language=javascript,tabsize=2,
+title=Packages for Ping Pong Client]
+{
+  "dependencies": {
+    "@testing-library/jest-dom": "^4.2.4",
+    "@testing-library/react": "^9.3.2",
+    "@testing-library/user-event": "^7.1.2",
+    "@types/jest": "^24.0.0",
+    "@types/node": "^12.0.0",
+    "@types/react": "^16.9.0",
+    "@types/react-dom": "^16.9.0",
+    "react": "^16.13.1",
+    "react-dom": "^16.13.1",
+    "react-scripts": "3.4.1",
+    "typescript": "~3.7.2"
+  },
+  "devDependencies": {
+    "@babel/preset-typescript": "^7.10.1",
+    "customize-cra": "^1.0.0",
+    "react-app-rewired": "^2.1.6"
+  }
+}
+\end{lstlisting}
+
+\begin{lstlisting}[language=javascript,tabsize=2,
+title=Packages for Ping Pong Server]
+{
+  "dependencies": {
+    "express": "^4.17.1",
+    "ws": "^7.3.0",
+    "zombie": "^6.1.4"
+  },
+  "devDependencies": {
+    "@types/express": "^4.17.6",
+    "@types/ws": "^7.2.4",
+    "ts-node": "^8.10.2",
+    "typescript": "^3.9.3"
+  }
+}
+\end{lstlisting}

### 2020-06-17 report/sections/evalconcl/eval.tex
Report: feedback

@@
 evaluate the expressiveness
-and performance of our API generation
-toolchain.
+and performance of our generated APIs.
 
@@
 
+\subsubsection{Expressiveness}
 To evaluate the expressiveness
@@
 \item
-\textbf{\tprotocol{Noughts and Crosses} Game:}
+\textbf{\tprotocol{Noughts and Crosses} Game, as introduced
+in \cite{PLACES2020}:}
 
@@
 to show that our work is compatible with multiparty sessions,
-which isn't the case in \cite{MVU2020,Exceptional}.
+which is not the case in \cite{MVU2020,Exceptional}.
 We show how our generated APIs are compatible with the
@@
 \item
-\textbf{\tprotocol{Two Buyers} Protocol, as introduced in \cref{section:twobuyer}:}
+\textbf{\tprotocol{Two Buyers} Protocol, as introduced in 
+\cref{section:twobuyer} and \cite{JACM2016,MPAST}:}
 
@@
 protocol. This \textit{cannot} be implemented in
-\cite{PureScript2019,MVU2020}, which illustrates the novelty
+\cite{PureScript2019,MVU2020,Exceptional}, 
+which illustrates the novelty
 of our \newtheory theory on routed multiparty session types.
@@
 
+\subsubsection{Performance}
 To evaluate the performance of our work,

### 2020-06-17 report/sections/evalconcl/eval/benchmarks.tex
Report: complete chapter 10.1
Report: feedback

@@
 \section{Performance Benchmarks}
+\label{section:benchmarks}
 
@@
 \item We use the production build generated by 
-\texttt{create-react-app} \cite{cra} for all experiments, which performs the
+\textit{create-react-app} \cite{cra} for all experiments, which performs the
 transpilation into JavaScript. We serve the production build using the
-\texttt{serve} package \cite{npmserve} available on \texttt{npm}.
+\textit{serve} package \cite{npmserve} available on \textit{npm}.
 
@@
 using headless browsing functionality from 
-the Zombie.js \cite{zombie} package.
+the \textit{Zombie.js} \cite{zombie} package.
 
@@
 implementations across experiments. Details can be found in their
-corresponding \texttt{package.json} manifests (\cref{appendix:eval}).
+corresponding \texttt{package.json} manifests 
+(\cref{section:evalbenchmarkpackagejson}).
 
-The benchmark compares three implementations of the \tprotocol{Ping Pong} protocol:
+The benchmark compares three implementations of 
+the \tprotocol{Ping Pong} protocol:
 

### 2020-06-17 report/sections/evalconcl/eval/game.tex
Report: feedback

@@
 
-We implement the classic turn-based board game of \textit{Noughts and Crosses}
+We implement the classic turn-based board game of 
+\textit{Noughts and Crosses}
 between two players,
-as introduced in \cref{section:intro}.
+as introduced in \cite{PLACES2020}.
+Both players, identified by either \textit{noughts (O's)}
+or \textit{crosses (X's)} respectively, take turns
+to place a mark on an unoccupied cell of a 3-by-3 grid
+until one player wins (when their markers form one straight
+line on the board) or a stalemate is reached (when all cells
+are occupied and no one wins).
+
 We formalise the game interactions using a Scribble protocol
@@
 \subsection{Game Server}
-We set up the WebSocket server as an Express.js \cite{Express} application
+We set up the WebSocket server as an \textit{Express.js}\footnote{
+Express is a commonly used library for writing lightweight
+web servers in JavaScript.
+} \cite{Express} application
 on top of the Node.js runtime.
@@
 const handleP1Move = new Implementation.S13({
-	Pos: async (move: Point) => {
+	Pos: (*@\hl{async}@*) (move: Point) => {
 		// `board` manages game state;
 		// `board.P1` registers the move and returns the game result
-		const result = await board.P1(move);
+		const result = (*@\hl{await}@*) board.P1(move);
 		switch (result) {
@@
 
-new Svr(wss, handleP1Move, (role, reason) => {
+const cancellation = (role: Roles.All, reason: string) => {
 	console.log(`${role} cancelled because of ${reason}`);
-});
+}
+
+new Svr(
+	wss,						// WebSocket server
+	handleP1Move,		// Game logic
+	cancellation		// Error handler
+);
 \end{lstlisting}
@@
 The main implementation detail for players
-is to make a move. 
+is to make moves. 
 Intuitively, the developer implements a grid and binds a handler
@@
 \cite{PureScript2019,MVU2020,LINKS}.
+
+Code is available at 
+\url{https://github.com/ansonmiu0214/SessionTS-Examples/NoughtsAndCrosses}.

### 2020-06-17 report/sections/evalconcl/eval/twobuyer.tex
Report: feedback

--- 2020-06-12 report/sections/evalconcl/eval/twobuyer.tex
@@
 through implementing novel theory of routed multiparty
-session types formalised in \cref{chap:theory}.
+session types formalised in \cref{chap:theory};
+the implementation of \newtheory in \codegen is 
+explained in \cref{chap:impl}.
 
@@
 \begin{lstlisting}[language=javascript]
-// two buyer seller
+new Implementation.Initial({
+	[Labels.S32.title]: (title) => {
+		return new Implementation.S34([
+			Labels.S34.quote, [100], new Implementation.S35([
+				Labels.S35.quote, [100], new Implementation.S36({
+					[Labels.S36.buy]: () => {
+						return new Implementation.Terminal();
+					},
+					[Labels.S36.cancel]: () => {
+						return new Implementation.Terminal();
+					}
+				})
+			])
+		])
+	}
+})
 \end{lstlisting}
-\captionof{lstlisting}{\dots}
+\captionof{lstlisting}{Two Buyer Seller Implementation}
 \label{lst:evaltwobuyerseller}
@@
 \begin{lstlisting}[language=javascript]
-// two buyer A p2p
+type State = { split: number };
+
+class ProposeSplit extends S11<State> {
+	state = { split: 0 };
+
+	render() {
+		const SendSplit = this.split('onClick', ev => {
+			ev.preventDefault();
+			return [this.state.split];		
+		});
+
+		return <div>
+			<input
+				type='number'
+				value={this.state.split}
+				onChange={ev => this.setState({
+					split: Number(ev.target.value)
+				})}
+				placeholder='Enter split'
+			/>
+			<SendSplit><button>Propose</button></SendSplit>	
+		</div>;	
+	}
+}
 \end{lstlisting}
-\captionof{lstlisting}{\dots}
+\captionof{lstlisting}{Developer Implementation of Peer-to-Peer Interaction
+in Two Buyer}
 \label{lst:evaltwobuyerA}
@@
 The \trole{S}eller endpoint is set up as an Express.js
-application, and both buyers still use the React Context API
-for application state management. We show an example of
-the latter in the developer's implementation for
-buyer \trole{A} receiving and keeping track of the server's quote
-using a React Context in \cref{lst:evaltwobuyercontext}.
+application, and both buyers use the \textit{React Context API}\footnote{
+React Contexts allow components
+to pass data (such as application state)
+through the component tree without
+having to propagate via props at every level.
+}
+for application state management. 
+%We show an example of
+%the latter in the developer's implementation for
+%buyer \trole{A} receiving and keeping track of the server's quote
+%using a React Context in \cref{lst:evaltwobuyercontext}.
+%
+%\begin{figure}[!h]
+%\begin{lstlisting}[language=javascript]
+%// two buyer A context
+%\end{lstlisting}
+%\captionof{lstlisting}{\dots}
+%\label{lst:evaltwobuyercontext}
+%\end{figure}
 
-\begin{figure}[!h]
-\begin{lstlisting}[language=javascript]
-// two buyer A context
-\end{lstlisting}
-\captionof{lstlisting}{\dots}
-\label{lst:evaltwobuyercontext}
-\end{figure}
-
-We include the implementations of the remaining EFSM states
-for buyer \trole{A} in \cref{section:evalcodetwobuyer}
-for completeness;
-interested readers can find the full implementations for the other
+%We include the implementations of the remaining EFSM states
+%for buyer \trole{A} in \cref{section:evalcodetwobuyer}
+%for completeness;
+Interested readers can find the full implementations for the other
 endpoints alongside the generated code on

### 2020-06-17 report/sections/general/implementation.tex
Report: feedback

@@
-\chapter{Implementation}
+\chapter{Implementing \newtheory in \codegen}
 \label{chap:impl}
@@
 as shown in \cref{lst:noderoutereceive}.
-By construction from our API generation strategy,
+By construction of our API generation strategy,
 the incoming message must match the interface defined
@@
 Hence, we need to implement
-the message receiving framework outlined in \cref{subsection:noderuntimereceive}
+the message receiving framework outlined in 
+\cref{subsection:noderuntimereceive}
 instead, which generalises the pair of message/handler queues to two mappings
@@
 and receive handler registration now targets the pair of queues
-specified by the role from which to receive from.
+specified by the role to receive from.
 We illustrate this in \cref{lst:reactrouterreceive}
@@
 Referring to the previous example under this framework,
-$S$ can route $C$'s message to $B$ first, but \lstonelinejs{M2(string)} will be stored
-under the message queue mapped to $C$, so it will not be processed until $B$ registers
-the receive handler for processing $C$'s message. This cannot happen until $B$ processes
+$S$ can route $C$'s message to $B$ first, 
+but \lstonelinejs{M2(string)} will be stored
+under the message queue mapped to $C$, 
+so it will not be processed until $B$ registers
+the receive handler for processing $C$'s message. 
+This cannot happen until $B$ processes
 \lstonelinejs{M1(number)} from $A$, so the order of communication 

### 2020-06-17 report/sections/general/p2p.tex
Report: feedback

@@
 we demonstrate how to support peer-to-peer interactions
+through routing communications
 between web browser endpoints over WebSocket transport,

### 2020-06-17 report/sections/general/p2p/challenges.tex
Report: feedback

@@
 This must be done \textit{very carefully}, as
-naive definitions risk over-serialising the original communication,
+naive definitions of routing risk 
+over-serialising the original communication,
 thereby not preserving all communication traces.

### 2020-06-17 report/sections/general/p2p/twobuyer.tex
Report: feedback

@@
 
-We present the \tprotocol{Two Buyer} protocol in \cref{lst:twobuyer},
+We present the \tprotocol{Two Buyer} protocol in \cref{lst:twobuyer}
+-- the canonical example for peer-to-peer interactions in
+literature \cite{JACM2016,MPAST} --
 which describes a protocol between two buyers \trole{A}, \trole{B}
-and a \trole{S}eller. We assume that the \trole{S}eller runs on
+and a \trole{S}eller.
+We assume that the \trole{S}eller runs on
 the server, and the buyers run on the browser.

### 2020-06-17 report/sections/server/extensions/async.tex
Report: feedback

@@
 
-Asynchronous APIs are commonplace in modern web programming,
-ranging from database operations to third-party API calls.
+Asynchronous APIs are commonplace in modern web programming;
+typical uses range from
+database operations to third-party API calls.
 We motivate the need for supporting asynchronous operations
@@
 \cref{subsection:asyncmotivation},
-demonstrate how we support this through extending
-the generated APIs and runtime in
-\cref{subsection:asyncapi,subsection:asyncruntime}
-respectively, 
+demonstrate how we support asynchronous APIs through extending
+the generated APIs (in \cref{subsection:asyncapi}) 
+and runtime (in \cref{subsection:asyncruntime}),
 and highlight the limitations of our approach in
@@
 \textit{non-blocking} APIs; it is
-recommended that file operations do not block the main execution,
-so a JavaScript database API may require a \textit{callback}
+recommended that IO-bound operations do not block the main execution,
+so a typical JavaScript database API may require a \textit{callback}
 to propagate the return value when it is available.
@@
 
-Callbacks are commonplace in dealing with concurrency in JavaScript
+Callbacks are commonplace in dealing with 
+asynchronous APIs in JavaScript
 \cite{CallbackHell}; we will go over other widely-used
@@
 
-This is also possible for browser-side implementations:
+The same scenario is also possible for browser-side implementations:
 when the user invokes a UI event, the browser-side logic
@@
 This means we should support
-the concurrency primitives built into TypeScript:
+the concurrency patterns built into TypeScript:
 
@@
 as parameters and invoke them when the asynchronous operation is complete.
-The \texttt{setTimeout(fn, delay)} function available on the browser
+For example,
+the \texttt{setTimeout(fn, delay)} function available on the browser
 waits for \texttt{delay} milliseconds before invoking \texttt{fn()} --
@@
 and its error handling construct. The developer passes a ``success''
-handler and ``error'' handler, conventionally named
+callback and ``error'' callback, conventionally named
 ``then'' and ``catch'' respectively, to construct a Promise.
@@
 
-This is rather straightforward.
+To support asynchronous implementations in the
+generated APIs, we use a type signature that allows
+the developer to use \texttt{Promise}s.
 We construct a generic union type, \texttt{MaybePromise<T>},
@@
 it is now permitted to return a \texttt{Promise} that resolves
-to a value of type texttt{Implementation.S42}.
+to a value of type \texttt{Implementation.S42}.
 The same logic applies to other types of EFSM states
@@
 we also support this newer piece of syntactic sugar.
+Moreover, our generated APIs work with both synchronous
+and asynchronous developer implementations transparently,
+as is the convention one would expect from modern JavaScript APIs.
 
@@
 The key observation is that the concrete types for the any
-handlers are the same -- they just might be wrapped in a Promise
+handlers are the same -- the payload may be wrapped in a Promise
 construct.
@@
 return the successor state identifier. Hence, aside from
-refactoring the functions to delay computation using thunks,
+wrapping the functions to delay computation using thunks,
 the functions themselves need to

### 2020-06-17 report/sections/server/extensions/error.tex
Report: feedback

@@
 We motivate the need for handling a variety exceptions in
-modern web programming in \cref{subsection:errormotivation},
+modern web programming in \cref{subsection:errormotivation} and
 demonstrate how we implement a structured error handling
-framework through extending the generated APIs and 
-runtime in \cref{subsection:errorapi,subsection:errorruntime}
-respectively, and explain the limitations of our framework
+framework through extending the generated APIs 
+(in \cref{subsection:errorapi}) and 
+runtime (in \cref{subsection:errorruntime}).
+We highlight the caveats of our error handling framework
+when dealing with asynchronous operations 
+(in \cref{subsection:errorasync}),
+and conclude by explaining the limitations of our framework
 in \cref{subsection:errorlimit}.
@@
 
+// Invoked when socket is closed
 const onClose = ({ target: socket }) => {
@@
 
+// Invoked when server receives connection request.
 const onSubscribe = (event) => {
@@
 	this.setState({
+		// Calls the user-specified cancellation handler,
+		// which returns a React node to render on DOM
 		elem: this.props.cancellation(role, reason);
@@
 \subsection{Caveats with Asynchronous Operations}
+\label{subsection:errorasync}
+
 Implementing session cancellation whilst

### 2020-06-17 report/sections/server/react.tex
Report: feedback

--- 2020-06-05 report/sections/server/react.tex
@@
-\chapter{\fancyname{ReactMPST}: Front-End Session Type Web Development}
+\chapter{\reactcodegen: Front-End Session Type Web Development}
 \label{chap:react}
 
-In this chapter, we present \fancyname{ReactMPST},
+In this chapter, we present \reactcodegen,
 our session type API generation strategy for browser-side endpoints
@@
 
+We highlight the challenges of session-typed GUI programming
+(\cref{section:reactchallenges}) and motivate 
+our approach (\cref{section:reactapproach})
+for adapting the proposal from
+\cite{MVU2020} for the React.js framework, along with a brief
+introduction to the framework.
+We proceed to explain how we leverage React
+to encode the EFSM (\cref{section:reactefsm})
+and implement the session runtime (\cref{section:reactruntime}),
+focusing on our novel approach to preventing
+channel reuse in GUI programming using React.
+We conclude by analysing some alternative designs
+(\cref{section:reactalt})
+and evaluating the limitations of our approach
+(\cref{section:reactlimitations}).
+
 \begin{figure}[!b]
 \centering
-\includegraphics[width=0.55\textwidth]{AdderClientFSM}
+\includegraphics[width=0.6\textwidth]{AdderClientFSM}
 \captionof{figure}{\trole{Client} Endpoint FSM
@@
 \section{Summary}
-\begin{itemize}
-\item static guarantees on avoiding communication mismatch
-\item static guarantees on affine channel usage
-\item leverage the react framework to express the model types approach from Fowler 2020 appropriately, to provide static guarantees, by construction, that the channel actions made available at each state respect the protocol, and by definition, multiparty session type safety
-\end{itemize}
+In this chapter, we have presented
+our session type API generation strategy
+which targets the React framework for browser-side endpoints.
+
+We motivated our approach of
+implementing the concept of model types introduced in \cite{MVU2020}
+using a popular front-end framework in React.
+We define an interpretation of model types
+using the EFSM obtained from a Scribble protocol.
+
+We discussed how to prevent channel reuse in
+GUI programming in React. We build upon the design choices
+implemented in \nodecodegen by concealing channel resources
+in the session runtime and exposing seams for the developer
+to customise what interface to render at each EFSM state,
+what user interface events are allowed to trigger a send action,
+and how to handle messages received from the channel.
+
+% Developers that implement...

### 2020-06-17 report/sections/server/react/approach.tex
Report: feedback

@@
 \section{Approach}
+\label{section:reactapproach}
+
 We motivate our approach from \cite{MVU2020} by extending
@@
 We implement model types for the EFSM on top of the 
-\emph{React.js} framework developed by Facebook \cite{React}.
+\emph{React.js} (React) framework developed by Facebook \cite{React}.
 React is widely used in industry to create scalable single-page
-TypeScript applications, so this makes our workflow beneficial in an
+web applications, so this makes our workflow beneficial in an
 industrial context. 
@@
 \item 
-\textbf{S[40-43].tsx:}
+\textbf{S[40-43].tsx\footnote{
+The \filename{.tsx} file extension allows
+for embedding \textit{JSX} \cite{JSX} elements inside the file.
+JSX is a XML-like syntax extension to JavaScript
+for elements and components.
+}:}
 Developer APIs for implementing EFSM states 

### 2020-06-17 report/sections/server/react/challenges.tex
Report: progress on chapter 5 limitations
Report: feedback

@@
 \section{Challenges}
+\label{section:reactchallenges}
+
 Our goal with \fancyname{ReactMPST} is to
 generate session type implementations for the web browser.
-Session typing the user interface (UI) logic is inherently difficult --
+Integrating session types into user interfaces is inherently difficult --
 assuming channel actions are bound to user interface events 
@@
 
-Existing work (\cref{subsection:sessiontypewebdev})
-do tackle browser-side session typing.
+We recap how existing work 
+(as discussed in \cref{subsection:sessiontypewebdev})
+tackle browser-side session typing.
 Fowler introduced the concept of model types to prevent
@@
 graphical user interface (GUI) programming \cite{MVU2020}, but
-it requires the linear type system available from
-the proprietary Links web programming language \cite{LINKS}.
+it uses the Links web programming language \cite{LINKS} which
+lacks compatibility with the ecosystem of JavaScript
+libraries that one might use in front-end development as well.
 The session type-safe web development framework presented
@@
 a functional target language in PureScript, and relies on
-the Concur UI framework that constructs UIs sequentially.
+the \textit{Concur UI} framework that constructs UIs sequentially.
 
-We find these proposals to come at the cost of
+As motivated in \cref{section:intro},
+we find these proposals to come at the cost of
 limiting developer productivity by adopting unconventional practices
@@
 for web-based GUI programming, and implement them using
-mainstream front-end web development tools (e.g. TypeScript
+mainstream front-end web development tools (i.e. TypeScript
 and React), to provide developers with an intuitive way to implement

### 2020-06-17 report/sections/server/react/efsm.tex
Report: feedback

@@
 as outlined in \cref{table:efsmmodeltypes}.
-\fancyname{ReactMPST} generates an abstract React component class
+\reactcodegen generates an abstract React component class
 for each EFSM state. The developer implements the generated API
@@
 
-Without \fancyname{ReactMPST},
+Without \reactcodegen,
 an implementation of state S40 from \cref{fig:adderclientfsm}
@@
 \begin{lstlisting}[language=javascript,tabsize=2]
+// UI component keeps track of data entered by user
 type State = { num1: number, num2: number, message: string }; (*@\label{line:clientsendstatestate}@*)
+
 class InputWindow extends S40<State> {
@@
 		const { num1, num2, message } = this.state;
+
+		// Generate React component that, when clicked,
+		// sends ADD message with `num1` and `num2`.
 		const (*@\hl{Add}@*) = this.ADD('onClick', (*@\label{line:clientsendfactory1}@*)
 			(ev: React.MouseEvent) => [num1, num2]); (*@\label{line:clientsendpayload1}@*)
+
+		// Generate React component that, when hovered,
+		// sends QUIT message with `message` string.
 		const (*@\hl{Quit}@*) = this.QUIT('onMouseOver', (*@\label{line:clientsendfactory2}@*)
@@
 \begin{lstlisting}[language=javascript,tabsize=2,title=S40.tsx]
+// Expect to receive `factory` function passed in by runtime
 type Props = { factory: SendComponentFactoryFactory };
+
 abstract class S40<State> extends React.Component<Props, State> {
+
+	// Allow developer implementation subclass to use factories
 	protected ADD: SendComponentFactory<[number, number]>;
@@
 		super(props);
+
+		// Generate a factory that generates React components
+		// which sends an [ADD] message along with two numbers,
+		// then transitions to State S42.
 		this.ADD = props.factory<[number, number]>( (*@\label{line:factory1}@*)
 			'ADD', ReceiveState.S42);
+
 		this.QUIT = props.factory<[string]>( (*@\label{line:factory2}@*)
@@
 Referring to \cref{fig:adderclientfsm} again,
-S42 will be rendered when the endpoint is waiting for
-\tmsg{RES}, but once received, S40 will be immediately
+S42 (preceding state) will be rendered when the 
+endpoint is waiting for
+\tmsg{RES}, but once received, S40 (successor state) 
+will be immediately
 rendered. The developer will need to manage the flow
@@
 
-Receive states in \fancyname{ReactMPST}
-work in a similar way as with \fancyname{NodeMPST}.
+Receive states in \reactcodegen
+work in a similar way as with \nodecodegen.
 The runtime will invoke the \tmsg{RES} method
@@
 identifiers and type signatures differ between receive states.
-We adopt the same strategy as with \fancyname{NodeMPST}:
+We adopt the same strategy as with \nodecodegen:
 the receive handler registered by the receive state
@@
 \begin{lstlisting}[language=javascript,tabsize=2]
+// Expect to receive `register` function passed in by runtime
 type Props = {
@@
 };
+
 abstract class S42<S> extends React.Component<Props, S> {
+
+	// Developer must implement receive handlers
+	// for possible messages received at this state.
 	abstract THANKS(payload1: string): void;
@@
 
+	// When component is mounted on DOM,
+	// and the register function is available,
+	// register the message handler with the runtime.
 	componentDidMount() {
@@
 \subsection{Terminal States}
-To be consistent with the conventions defined for \fancyname{NodeMPST},
+To be consistent with the conventions defined for \nodecodegen
 the browser endpoint proactively closes the connection at its terminal state.

### 2020-06-17 report/sections/server/react/limitations.tex
Report: feedback

@@
 \section{Limitations}
+\label{section:reactlimitations}
+
 Our EFSM encoding provides \textbf{affine} channel
@@
 we present examples of developer implementations
-that use separate abstractions (e.g. the \textit{React Context API} or
-the third-party \textit{Redux} state management framework) for propagating 
+that use separate abstractions (e.g. the \textit{React Context API}) 
+for propagating 
 stateful application data in \cref{chap:eval}.

### 2020-06-17 report/sections/server/react/runtime.tex
Report: complete chapter 9
Report: feedback

@@
 for \fancyname{ReactMPST} compared to that of \fancyname{NodeMPST}.
-We focus on the details unique to browser-side endpoints.
+We focus on the details unique to React endpoints.
 
@@
 	endpoint: string,
-	states: {
-		S40: Constructor<S40>,
-		S42: Constructor<S42>,
+	states: {						// mapping of state identifier
+		S40: Constructor<S40>,		// to EFSM state components
+		S42: Constructor<S42>,		// implemented by developer
 		S43: Constructor<S43>,
@@
 	},
-	waiting: React.ReactNode,
-	connectFailed: React.ReactNode,
+	waiting: React.ReactNode,		// render when waiting to start
+	connectFailed: React.ReactNode,	// render if connection failed
 };
+
+export class Session extends React.Component<Props, ...
 \end{lstlisting}
@@
 an error screen if the connection has failed.
+We show an example below.
+
+\begin{lstlisting}[language=javascript,tabsize=2]
+// Developer implementations
+class InputWindow extends S40 { ... };
+class QuitReceived extends S41 { ... };
+class WaitScreen extends S42 { ... };
+class PendingQuit extends S43 { ... };
+
+// Main application component
+class MainApp extends React.Component {
+	render() {
+		return <div>
+			<h1>Adder Client</h1>
+			<Client
+				endpoint='ws://localhost:8080'
+				states={{
+					S40: InputWindow,
+					S41: QuitReceived,
+					S42: WaitScreen,
+					S43: PendingQuit,		
+				}}>
+				waiting={<h1>Pending Connection</h1>}
+				connectFailed={<p>Connection Failed</p>}
+			/>
+		</div>;
+	}
+}
+\end{lstlisting}
 
@@
 For this reason, we create the \texttt{WebSocket} in the
-\texttt{componentDidMount} method as it is guaranteed
+\texttt{componentDidMount}\footnote{
+The \texttt{componentDidMount} method is part
+of the ``commit'' phase in the lifecycle of a React component.
+This is when the component is mounted on the browser DOM and
+methods in the phase are guaranteed to be called exactly once.
+} method as it is guaranteed
 to be only called once.
@@
 WebSocket in its state, and the private API (the React component
-named after the role -- in this case, \trole{Client})after
+named after the role -- in this case, \trole{Client}) after
 as a separate non-exported class
-with the WebSocket available via props. Once the WebSocket
-is instantiated in the \trole{Session} component, it renders the
-\trole{Client} component, passing the non-optional WebSocket value via props.
+with the WebSocket available via props. 
+Once the WebSocket
+is instantiated in the \trole{Session} component,
+it renders the
+\trole{Client} component, 
+passing the non-optional WebSocket value via props.
 
@@
 
-The connection phase is managed in the same way as \fancyname{NodeMPST}:
+The connection phase is managed in the same way as \nodecodegen:
 when the \texttt{Client} component is mounted,
 it sends a connection request to the server and overrides the
-\texttt{onmessage} event handler to listen for the connection confirmation,
-before advancing the EFSM to the initial state to begin executing the protocol.
+\texttt{onmessage} event handler to listen for 
+the connection confirmation,
+before advancing the EFSM to the initial state to 
+begin executing the protocol.
 
 \subsection{Executing the EFSM}
-Unlike \fancyname{NodeMPST}, our EFSM state encoding does not define
+Unlike \nodecodegen, our EFSM state encoding does not define
 continuation states using the actual implementation.
@@
 This enables the UI to be re-rendered, and therefore ``react'',
-to EFSM state transitions, which is precisely the goal of \fancyname{ReactMPST}.
+to EFSM state transitions, which is precisely the goal of 
+\reactcodegen.
 
@@
 Messages are sent through the WebSocket in the same way 
-as \fancyname{NodeMPST} -- objects marked with label and payload
+as \nodecodegen (see \cref{subsection:noderuntimesend}) 
+-- objects marked with label and payload
 are serialised into JSON. The process of sending is always followed
@@
 the successor state as a function argument, so it can make that transition
-immediately after.
+immediately after.\\
 
@@
 \subsection{Receiving Messages}
+\label{subsection:reactruntimereceive}
 

### 2020-06-18 report/sections/general/theory.tex
Report: progress on chapter 8
Report: feedback chapter 8

@@
 that are routed through a participant.
-We introduce the extended syntax for types in 
-\cref{section:syntax}, and give semantics of our
-syntax extensions in \cref{section:lts}.
 
@@
 which uses WebSocket transport.
+We introduce the extended syntax for types in 
+\cref{section:syntax}, and give semantics of our
+syntax extensions in \cref{section:lts}.
 We define a router-parameterised encoding
@@
 
+\input{sections/general/theory/soundcomplete}
+
 \input{sections/general/theory/encoding}
+
+\section{Summary}
+We have presented \newtheory, a variant of
+the canonical MPST theory to express \textit{routed communication}.
+We introduced extensions to syntax and semantics, and proved
+that our extended semantics preserve deadlock-freedom
+for well-formed protocols. We defined an encoding from the
+canonical theory onto \newtheory, and proved the preservation
+of well-formedness and communication.

### 2020-06-18 report/sections/general/theory/encoding.tex
Report: minor fixes to chapter 8
Report: feedback chapter 8

@@
 This encoding is \textit{parameterised} by the router role 
-(conventionally denoted as $\mrole{s}$);
+(conventionally denoted as $\role{s}$);
 the intuition is that we encode all communication interactions
-to involve $\mrole{s}$.
+to involve $\role{s}$.
 For example, as motivated in \cref{chap:p2p},
@@
 We arrive at \cref{lem:enccenter} -- the proof is straightforward
-by induction on global types and left as an exercise to the reader.
+by induction on global types.
 
@@
 \end{lemma}
+
+\begin{proof}
+By induction on the structure of $G$.
+\end{proof}
 
@@
 will carry out the communication through the router
-$\trole{s}$ as a proxy in the encoded local type.
+$\trole{s}$ in the encoded local type.
 This is expressed in both \rulename{Enc-L-Sel}
@@
 as presented in \cref{lem:encprojeq}.
-
-\begin{lemma}
-\[
-\forall G_1, G_2, \mrole{r}, \mrole{s}. ~ \left(
-(\proj{G_1}{r}) = (\proj{G_2}{r}) \wedge \mrole{r} \neq \mrole{s}
-	\Longrightarrow
-\proj{\enc{G_1}{s}}{r} = \proj{\enc{G_2}{s}}{r}
-\right)
-\]
-
-\label{lem:encprojeq}
-\end{lemma}
-
-\begin{proof} By consequence from \cref{lem:enclink}.
-
-Take $G_1, G_2, \mrole{r}, \mrole{s}$ arbitrarily. 
-Assume $(\proj{G_1}{r}) = (\proj{G_2}{r})$ and $\mrole{r} \neq \mrole{s}$.
-
-We need to show $\proj{\enc{G_1}{s}}{r} = \proj{\enc{G_2}{s}}{r}$,
-but by \cref{lem:enclink}, it is sufficient to show
-\[
-\enclocal{\proj{G_1}{r}}{r}{s} = \enclocal{\proj{G_2}{r}}{r}{s}
-\]
-
-Define an ``inner'' function $f(z) = \enclocal{z}{r}{s}$.
-
-By definition, $\forall x, y. (x = y \Longrightarrow f(x) = f(y))$.
-
-We have $(\proj{G_1}{r}) = (\proj{G_2}{r})$ by assumption,
-so we can conclude.
-
-\begin{align*}
-(\proj{G_1}{r}) = (\proj{G_2}{r}) 
-&\Longrightarrow f(\proj{G_1}{r}) = f(\proj{G_2}{r})  \\
-&\Longrightarrow
-\enclocal{\proj{G_1}{r}}{r}{s} = \enclocal{\proj{G_2}{r}}{r}{s}
-\end{align*}
-
-%\[
-%(\proj{G_1}{r}) = (\proj{G_2}{r}) 
-%\Longrightarrow
-%f(\proj{G_1}{r}) = f(\proj{G_2}{r}) 
-%\Longrightarrow
-%\enclocal{\proj{G_1}{r}}{r}{s} = \enclocal{\proj{G_2}{r}}{r}{s}
-%\]
-
-\end{proof}
 
@@
 \end{proof}
+
+We motivated the risk of over-serialising communication
+with a naive routing definition in \cref{section:routerchallenges}.
+We revisit the example and show that our encoding
+preserves valid traces in the original communication.
+
+\begin{example}[Encoding Preserves Semantics]
+Consider the global type
+\[
+G = \gcommone{p}{q}{\tmsg{M1}} . ~
+	\gcommone{s}{q}{\tmsg{M2}} . ~\tend
+\]
+
+We apply our encoding with respect to the router role $\role{s}$
+\[
+\enc{G}{s} = \grouteone{p}{q}{s}{\tmsg{M1}}.~
+	\gcommone{s}{q}{\tmsg{M2}}. ~ \tend
+\]
+
+Recall that $l = \aout{s}{q}{\tmsg{M2}}$ can reduce $G$.
+
+\begin{prooftree}
+\AxiomC{}
+\RightLabel{\rulename{Gr1}}
+\UnaryInfC{$
+\treduce
+	{\left(\gcommone{s}{q}{\tmsg{M2}}.~ \tend\right)}
+	{\tend}
+	{l}
+$}
+\AxiomC{$\subj{l} = \mrole{s} \notin \{ \mrole{p}, \mrole{q} \}$}
+\RightLabel{\rulename{Gr4}}
+\BinaryInfC{$
+\treducelong
+	{\left(\gcommone{p}{q}{\tmsg{M1}} . ~
+		\gcommone{s}{q}{\tmsg{M2}} . ~\tend\right)}
+	{\left(\gcommone{p}{q}{\tmsg{M1}} . ~
+	\gtransone{s}{q}{\tmsg{M2}} . ~\tend\right)}
+	{l}
+$}
+\end{prooftree}
+
+We show that $\enc{l}{s} = l$ can reduce $\enc{G}{s}$.
+
+\begin{prooftree}
+\AxiomC{}
+\RightLabel{\rulename{Gr8}}
+\UnaryInfC{$
+\treduce
+	{\left(\gcommone{s}{q}{\tmsg{M2}}.~ \tend\right)}
+	{\tend}
+	{l}
+$}
+\AxiomC{$\subj{l} = \mrole{s} \notin \{ \mrole{p}, \mrole{q} \}$}
+\RightLabel{\rulename{Gr4}}
+\BinaryInfC{$
+\treducelong
+	{\left(\grouteone{p}{q}{s}{\tmsg{M1}} . ~
+		\gcommone{s}{q}{\tmsg{M2}} . ~\tend\right)}
+	{\left(\grouteone{p}{q}{s}{\tmsg{M1}} . ~
+	\gtransone{s}{q}{\tmsg{M2}} . ~\tend\right)}
+	{l}
+$}
+\end{prooftree}
+
+$\aout{s}{q}{\tmsg{M2}}$ is a prefix of a valid
+execution trace for $G$, given below.
+
+\[
+\treducelong
+	{\treducelong
+		{\treducelong
+			{\treducelong
+				{G}
+				{}
+				{\aout{s}{q}{\tmsg{M2}}}}
+			{}
+			{\aout{p}{q}{\tmsg{M1}}}}
+		{}
+		{\ain{p}{q}{\tmsg{M1}}}}
+	{\tend}
+	{\ain{s}{q}{\tmsg{M2}}}
+\]
+
+Interested readers can verify
+that the encoded trace (given below) 
+is a valid execution trace for
+$\enc{G}{s}$.
+
+\[
+\treducelong
+	{\treducelong
+		{\treducelong
+			{\treducelong
+				{\enc{G}{s}}
+				{}
+				{\aout{s}{q}{\tmsg{M2}}}}
+			{}
+			{\via{s}{\aout{p}{q}{\tmsg{M1}}}}}
+		{}
+		{\via{s}{\ain{p}{q}{\tmsg{M1}}}}}
+	{\tend}
+	{\ain{s}{q}{\tmsg{M2}}}
+\]
+
+\end{example}

### 2020-06-18 report/sections/general/theory/lts.tex
Report: fix notation
Report: feedback chapter 8

@@
 
-We define labelled transition system (LTS) semantics
+We define the labelled transition system (LTS) semantics
 over global types (\cref{subsection:newltsglobal}) 
@@
 within the grammar of global types.
-
-\cite{characterisation} added the construct
+Deni\'elou and Yoshida \cite{characterisation}
+added the construct
 {$\gtrans{p}{q}{j}{l_i: G_i}{i \in I}$}
@@
 but not yet routed to $\role{q}$.
+We extend the projection operator to 
+support this new construct.
+
+\[
+\proj{\gtransroute{p}{q}{s}{j}{l_i: G_i}{i \in I}}{r} = \begin{cases}
+\tbraproxy{p}{s}{l_i: \proj{G_i}{r}}{i \in I} 
+	& \text{if } \mrole{r} = \mrole{q} \\
+\routertrans{p}{q}{j}{l_i: \proj{G_i}{r}}{i \in I}
+	& \text{if } \mrole{r} = \mrole{s} \\
+\proj{G_j}{r}
+	& \text{otherwise} \\
+\end{cases}
+\]
 
 Because routed communication is treated differently
-from normal send and receive actions, the notion
+from direct send and receive actions, the notion
 of asynchrony differs between the two types of communication
 too. This definition allows us to extend
-the LTS semantics defined in \cite{characterisation}
+the LTS semantics from \cite{characterisation}
 more naturally.
@@
 over global types,
-defined by $\treduce{G}{G'}{l}$,
+denoted by $\treduce{G}{G'}{l}$,
 in \cref{fig:newglobal}.
@@
 
+The LTS semantics over local types
+need to be extended for routed communication.
+Because the router holds a ``global perspective''
+on the routed communication, we also need to
+represent the intermediate state (at which the message is 
+being routed by the router) within the grammar
+of local types.
+Similar to \cref{subsection:newltsglobal}, we add 
+the construct 
+$\routertrans{p}{q}{j}{l_i: T_i}{i \in I}$
+to represent that, from the local perspective of
+the router, the message $l_j$
+has been received from $\role{p}$ but not yet routed
+to $\role{q}$.
+
 We define the LTS semantics 
 over local types,
-defined by $\treduce{T}{T'}{l}$,
+denoted by $\treduce{T}{T'}{l}$,
 in \cref{fig:newlocal}.
@@
 The exception is that the new rules pattern-match
-on the proxy role $\role{s}$ on the local type
+on the router role $\role{s}$ on the local type
 and the routed label.
@@
 \begin{align*}
-G_1 &= \gcommone{s}{r}{\tmsg{M1}} \grouteone{r}{q}{s}{\tmsg{M2}} \tend \\
+G_1 &= \gcommone{s}{r}{\tmsg{M1}}.~ \grouteone{r}{q}{s}{\tmsg{M2}}.~ \tend \\
 \proj{G_1}{s} &= \tselone{r}{\tmsg{M1}}. ~ 
 	\routerone{r}{q}{\tmsg{M2}}{\tend} \\~\\
-G_2 &= \gcommone{s}{r}{\tmsg{M1}} \grouteone{p}{r}{s}{\tmsg{M2}} \tend \\
+G_2 &= \gcommone{s}{r}{\tmsg{M1}}.~ \grouteone{p}{r}{s}{\tmsg{M2}}.~ \tend \\
 \proj{G_2}{s} &= \tselone{r}{\tmsg{M1}}. ~ 

### 2020-06-18 report/sections/general/theory/syntax.tex
Report: feedback chapter 8

@@
-\section{Syntax}
+\section{Syntax for Global and Local Types}
 \label{section:syntax}
@@
 in \cref{subsection:newglobal} and
-local types in \cref{subsection:newlocal}.
+local types in \cref{subsection:newlocal}
+to express routed communication.
 The same conventions introduced in \cref{subsection:bgmpst}
@@
 
+Global types range over $G, G', G_i, \dots$
 As motivated by \cref{section:routerchallenges},
@@
 new construct for expressing routed communication.
-Global types range over $G, G', G_i, \dots$
 
@@
 \textit{
-``the communication $\gcomm{p}{q}{l_i: G_i}{i \in I}$
+``the communication interaction 
+from $\role{p}$ to $\role{q}$
 is routed by $\role{s}$''
@@
 $\role{s}$ ranges over the set of roles 
-$\role{p},\role{q},\role{r},\role{s},\dots$,
+$\role{p},\role{q},\role{r},\dots$,
 but we use $\role{s}$ by convention as the server endpoint
@@
 
-We also extend the definition of the participants function,
-$\pt{G}$, in \cref{def:newpt}, to formalise that the router 
+We inherit the definitions of the participants function, $\pt{G}$,
+introduced in \cref{table:pt}; we extend it in \cref{def:newpt}
+to formalise that the router 
 participates in the routed communication.
@@
 
+Local types range over $T, T', T_i, \dots$
 Local types are obtained from global types via
@@
 some arbitrary role $\role{r}$.
-Local types range over $T, T', T_i, \dots$
 
@@
 \mid & \tselproxy{p}{q}{l_i: T_i}{i \in I}
-& \text{\hl{Proxy Selection}} \\
+& \text{\hl{Routed Selection}} \\
 \mid & \tbraproxy{p}{q}{l_i: T_i}{i \in I}
-& \text{\hl{Proxy Branching}} \\
+& \text{\hl{Routed Branching}} \\
 \end{array}
@@
 
-\item \textbf{Proxy selection:}
+\item \textbf{Routed slection:}
 
@@
 \textit{intended recipient} $\role{p}$,
-but the selection is sent to the intermediate proxy $\role{q}$. 
+but the selection is sent to the intermediate router $\role{q}$.,
+instead of to $\role{p}$ directly.
 $\role{r}$ sends their internal choice $l_i$ to $\role{q}$
@@
 
-\item \textbf{Proxy branching:}
+\item \textbf{Routed branching:}
 
@@
 \textit{intended sender} $\role{p}$,
-but the message is received from the intermediate proxy $\role{q}$.
+but the message is received from the intermediate router $\role{q}$,
+instead of from $\role{p}$ directly.
 $\role{r}$ receives the external choice $l_i$ from $\role{q}$
@@
 
-As a general comment, the new local types of proxy selection
-and proxy branching should ``behave'' as normal selection and
+As a general comment, the new local types of routed selection
+and routed branching should ``behave'' as normal selection and
 receive types with respect to their intended recipient and sender
-respectively. We keep track of the proxy role in the syntax
+respectively. We keep track of the router role in the syntax
 to allow us to distinguish between routing communications
@@
 
-We extend the projection operator to be defined
-on routed communication. Given the explanations of the
-new routing constructs for both global and local types,
-\cref{def:newprojection} should be self-explanatory.
+In \cref{def:newprojection},
+we extend the projection operator
+to be defined
+on routed communication.
 
@@
 The projection of $G$ onto $\role{r}$,
-written $\proj{G}{r}$, is defined as:
+written $\proj{G}{r}$, is extended as:
 
@@
 As the merging operator is defined on local
-types, we extend the merging operator to be defined
+types, we extend the merging operator 
+(introduced in \cref{fig:merge}) to be defined
 on the extended syntax in \cref{def:newmerge}.
@@
 The merging operator $\mergeop$ on local types
-is:
+is extended as:
 
@@
 \tmerge{T_k}{T'_k} & \text{if} ~ k \in I \cap J \\
+\text{otherwise} & \text{undefined}
 \end{cases} \\
@@
 
-Recall that proxy selection and proxy
-branching behave in the same way as their ``non-proxified''
+Recall that routed selection and routed
+branching behave in the same way as their ``non-routed''
 counterparts -- the merging operator reflects this
@@
 in canonical MPST theory: a global type $G$ is well-formed
-if a projection exists for all its participants.
+if a projection is defined for all its participants.
 
@@
 We need to define the characteristics that $\role{s}$
-must display in $G$ to prove that it is the \textit{only}
+must display in $G$ to prove that it is a
 router. 
 We formalise this as an \textit{inductively} defined relation,
-$\centroid{G}{s}$, which reads \textit{``$\role{s}$ is the
+$\centroid{G}{s}$, which reads \textit{``$\role{s}$ is a
 centroid in $G$''}.
@@
 \item 
-\rulename{$\centroidop$-End}, 
-\rulename{$\centroidop$-RecVar}, 
-\rulename{$\centroidop$-Rec}:
-Trivial.
-
-\item 
 \rulename{$\centroidop$-Comm}:
-For normal communication, $\role{s}$ must
+For direct communication, $\role{s}$ must
 be a participant and the centroid of all continuations.
@@
 
+We also assume that $G$ has no free type variables
+and is guarded/contractive.
+
 \label{def:newwf}

### 2020-06-18 report/sections/evalconcl/eval.tex
Report: feedback on benchmarks

@@
 evaluate the expressiveness
-and performance of our generated APIs.
-
-%TODO: add when finalised
-%with respect to the main objective
-%of \dots
+and performance of our generated APIs
+with respect to the main objective of
+our project in
+providing
+developers with a development workflow
+that offers communication safety guarantees in modern
+web programming through multiparty session types.
 
@@
 \input{sections/evalconcl/eval/benchmarks}
+
+\section{Summary}
+We have demonstrated
+the expressiveness of our generated APIs
+through implementing two web services describing
+multiparty sessions. We highlighted how our
+approach towards session-typed GUI programming
+prevents channel reuse and 
+makes it intuitive to implement the game board
+of \tprotocol{Noughts and Crosses}.
+We also showed
+that the extensions added to \nodecodegen
+and \reactcodegen
+to support \newtheory
+are transparent to the developer.
+
+Through performance benchmarks against a
+baseline implementation of the \tprotocol{Ping Pong}
+protocol, we analysed the overhead of our implementation
+and reasoned that this overhead will minimise when compared
+with baseline implementations of more complex, multiparty
+protocols involving routed communication.

### 2020-06-18 report/sections/evalconcl/eval/benchmarks.tex
Report: feedback on benchmarks

@@
 
-Whilst web applications implementing our generated APIs enjoy
-communication safety guarantees, the presence of the session runtime acts
-as an additional layer of abstraction between the application logic and the
-WebSocket transport, which is likely to present a performance trade-off.
+Whilst web applications implementing 
+our generated APIs enjoy
+communication safety guarantees, 
+the presence of the session runtime acts
+as an additional layer of abstraction between 
+the application logic and the
+WebSocket transport, which presents
+a performance trade-off; we observe an
+increase in the time taken to perform
+each channel action, as a result of
+the session runtime intercepting channel actions
+and performing additional logic to ensure
+that the endpoint conforms to the protocol.
 
@@
 
-The interested reader may follow the instructions under
-\texttt{benchmarks/README.md} in the project repository \cite{repo}
-to run the benchmarks and visualise the logs using the interactive
-notebook in the same directory.
+Code is available at
+\url{https://github.com/ansonmiu0214/SessionTS-Benchmarks}.
+Interested readers may follow the \texttt{README.md}
+to run the benchmarks and visualise the logs using the 
+interactive notebook in the same directory.
 
@@
 We standardise all packages used in the front- and back-end
-implementations across experiments. Details can be found in their
-corresponding \texttt{package.json} manifests 
-(\cref{section:evalbenchmarkpackagejson}).
+implementations across experiments. Details can be found in
+\cref{section:evalbenchmarkpackagejson}.
 
@@
 \subsection{Overhead}
+The execution time is the total time taken
+for the \trole{Client} to connect to the \trole{Server}
+and complete the parameterised number of round trips.
 We compare the total execution time (\textit{Exec. Time}) 
@@
 		}
-	} else {
-		throw new Error(`Unrecognised label: ${label}`);
-	}
+	} else { throw new Error(`Unrecognised label: ${label}`); }
 }
@@
 \begin{lstlisting}[language=javascript,tabsize=2]
-// NOTE: `Implementation' namespace omitted for brevity.
-
-const logic = new S14({
+const logic = new Implementat.S14({
 	PING: (count) => {
@@
 		if (count === MSGS) {
-			return new S16([Labels.S16.BYE, [count], new S15()]);
+			return new Implementation.S16([
+				Labels.S16.BYE, [count], new Implementation.S15()
+			]);
 		} else {
-			return new S16([Labels.S16.PONG, [count], logic]);
+			return new Implementation.S16([
+				Labels.S16.PONG, [count], logic
+			]);
 		}

### 2020-06-18 report/sections/evalconcl/eval/twobuyer.tex
Report: feedback on benchmarks

@@
 \begin{figure}[!h]
-\begin{lstlisting}[language=javascript]
+\begin{lstlisting}[language=javascript,tabsize=2]
 new Implementation.Initial({
-	[Labels.S32.title]: (title) => {
+	[Labels.S32.title]: async (title) => {
+		const quote = await db.getQuote(title);
 		return new Implementation.S34([
-			Labels.S34.quote, [100], new Implementation.S35([
-				Labels.S35.quote, [100], new Implementation.S36({
+			Labels.S34.quote, [quote], new Implementation.S35([
+				Labels.S35.quote, [quote], new Implementation.S36({
 					[Labels.S36.buy]: () => {

### 2020-06-18 report/sections/background/related.tex
Report: feedback chapter 8
Report: feedback on Ch2

@@
 		\tmsg{Add(int, int)} & :
-			\gcommone{S}{C}{\tmsg{Res(int)}}. \trecvar \\
+			\gcommone{S}{C}{\tmsg{Res(int)}}. ~\trecvar \\
 		\tmsg{Quit()} & : 
-			\gcommone{S}{C}{\tmsg{Terminate()}}. \tend \\		
+			\gcommone{S}{C}{\tmsg{Terminate()}}.~ \tend \\		
 	\end{cases}
@@
 For \cref{fig:bgadderfsm}, this is singleton set
-$\{~\texttt{number}~\}$.
+$\{~\texttt{int}~\}$.
 
@@
 \item \textbf{Channel linearity:}
-Each transition in the EFSM represents a channel resource.
-When the application transitions from some state $S$
-to some successor state $S'$, it must no longer
+Each transition in the EFSM represents one channel resource.
+The application transitions from some state $S$
+to some successor state $S'$
+once \textit{and only once},
+after which it must no longer
 be able to access a reference (e.g. have an alias) to $S$.
@@
 
-\subsubsection{Hybrid Session Verification in Java}
-Hu and Yoshida implemented a workflow in \cite{Hybrid2016}
+\subsubsection{Hybrid Session Verification in Java \cite{Hybrid2016}}
+Hu and Yoshida implemented a workflow
 that performs \textit{hybrid} verification of 
@@
 done at runtime.
-Each state keeps track of its usage in a private boolean flag a
-nd throws an exception when the instance method is called twice.
+Each state keeps track of its usage in a private boolean flag 
+and throws an exception when the instance method is called twice.
 Similarly, the \texttt{SessionEndpoint} class keeps track of 
@@
 
-\subsubsection{Runtime Monitors in Python}
+\subsubsection{Runtime Monitors in Python \cite{Python2017} }
 Neykova targeted the MPST methodology 
-for Python programs in \cite{Python2017} 
+for Python programs
 and proposed to generate \textit{runtime monitors} 
@@
 
-\subsubsection{API Generation in PureScript}
-King et al. presented an approach in \cite{PureScript2019}
+\subsubsection{API Generation in PureScript \cite{PureScript2019}}
+King et al. presented an approach
 for integrating \textit{multiparty} session types 
@@
 PureScript statically resolves these type constraints at
-compile-time.
+compile time.
 
@@
 clicking on it again will reuse the channel resource).
-Whilst \cite{PureScript2019} supports multiparty session types,
+Whilst King et al. \cite{PureScript2019} supports 
+multiparty session types,
 the communication protocol must describe all interactions
@@
 
-\subsubsection{Session-Typed GUI Programming using Links}
+\subsubsection{Session-Typed GUI Programming 
+using Links \cite{MVU2020}}
 Fowler presented the first formal integration
-of session typing and GUI programming in \cite{MVU2020}.
+of session typing and GUI programming .
 His work formalised the \textit{Model-View-Update} (MVU)
@@
 
-\cite{MVU2020} is implemented using
+Fowler's proposal \cite{MVU2020} is implemented using
 the \textit{Links} web programming language
@@
 given side-effects must be made explicit
-in functional programming, \cite{Exceptional} introduces
+in functional programming,
+Fowler et al. \cite{Exceptional} introduces
 a \textit{session cancellation} operator to ensure

### 2020-06-18 report/sections/background/sessiontypes.tex
Report: complete background
Report: review chapter 4
Report: feedback on Ch2

@@
 \section{Session Types}
+\label{section:sessiontypes}
 
@@
 formalises the message passing abstraction in terms of 
-the basic building blocks of sending and receiving processes,
-along with inductively defined continuation processes. 
+the basic building blocks of sending and receiving processes.
 The composition of these primitives allow us to 
@@
 provided by the programming language.
-For example, King et al. leveraged the 
+For example, King et al. \cite{PureScript2019} leveraged the 
 expressive type system of PureScript to perform 
-\textit{static} session type checking in \cite{PureScript2019},
-whilst Neykova and Yoshida introduced dynamic approaches to 
+\textit{static} session type checking,
+whilst Neykova and Yoshida \cite{Python2017} 
+introduced dynamic approaches to 
 check the conformance of Python programs 
-with respect to session types in \cite{Python2017}.
+with respect to session types.
 We discuss these related work, among others, in \cref{section:bgrelated}.
@@
 more specifically, $!P \equiv P \mid {!P}$.
+
+$\equiv$ represents the \textit{structural congruence} 
+equivalence relation
+on processes, meaning that $!P$ is indistinguishable from 
+$P \mid {!P}$. We inherit the definition of structural congruence from
+\cite{C406Lecture}.
+
 \item $(\nu a)~P$ represents a name restriction 
@@
 between parallel processes.
-Whilst \cite{C406Lecture} presents the full operational semantics, 
-we highlight the \rulename{Comm} reduction rule which 
+We inherit the full operational semantics presented by
+Yoshida \cite{C406Lecture}, and
+highlight the \rulename{Comm} reduction rule which 
 specifically models message passing:
@@
 \AxiomC{$
-\forall i \in I \left(
+\forall i \in I.\left(
 \vec x_i = x_1, \dots, x_n
@@
 \mathcal{M} ::= & \ptprocess{p_1}{P_1} ~\mid~
-\ptprocess{p_2}{P_1} ~\mid~\dots~\mid
+\ptprocess{p_2}{P_1} ~\mid~\dots~\mid~
 \ptprocess{p_n}{P_n}
@@
 multiple binary sessions: 
-\cite{C406Lecture} and \cite{MPST} both present 
+Yoshida and Lorenzo \cite{MPST} presents
 counterexamples of well-typed binary sessions that, 
@@
 
-Honda et al. presents {multiparty session types} 
-in \cite{MPAST} to extend the 
+Honda et al. \cite{MPAST} presents {multiparty session types} 
+to extend the 
 binary session typing discipline for 
@@
 describing the interactions between pairs of participants. 
-\cref{fig:mpst} defines the syntax of global types
+\cref{fig:globaltypes} defines the syntax of global types
 inspired by \cite{MPST};
@@
 \caption{Syntax of Global Types}
-\label{fig:mpst}
+\label{fig:globaltypes}
 \end{figure}
@@
 an ill-formed global type; 
-\cite{C406Lecture} presents examples of where this is the case, 
-and \cite{MPST} presents the formal definition of projection.
+Yoshida \cite{C406Lecture} presents examples of where this is the case, 
+and Yoshida and Lorenzo \cite{MPST} 
+presents the formal definition of projection.
 We focus on the definition of projection for the 
@@
 We use the \textit{inductive} definition presented by
-Scalas and Yoshida in \cite{LessIsMore}, shown in \cref{fig:merge}.
+Scalas and Yoshida in \cite{LessIsMore} Definition 3.3,
+shown in \cref{fig:merge}.
 

### 2020-06-18 report/sections/abstract.tex
Report: final proofread

--- 2020-06-16 report/sections/abstract.tex
@@
 \noindent
+Modern web programming
+involves coordinating interactions
+between web browser clients and a web server.
+Typically, the interactions are
+informally described, making
+it difficult to verify communication
+correctness in web-based distributed systems.
+
 Multiparty Session Types (MPST)
-is a typing discipline for concurrent
+are a typing discipline for concurrent
 processes that communicate via
-message-passing. MPST theory can ensure
+message passing. MPST theory can ensure
 communication safety properties and
@@
 is incompatible with modern web
-programming practices and are limited
+programming practices and is limited
 to supporting communication protocols
@@
 We address limitations in the current 
-state-of-the-art by:
+state of the art by:
 \textbf{(1)} implementing \codegen, a code generation
@@
 and are compatible
-with the Node.js runtime and the React.js framework
-for server-side and browser-side endpoints respectively.
+with the Node.js runtime for server-side endpoints
+and the React.js framework
+for browser-side endpoints.
 We evaluate the expressiveness of \codegen for modern
 web programming
-using case studies of protocols found in web-services,
-and analyse the overhead of our implementation through 
-running micro-benchmarks against a baseline implementation.
+using case studies of protocols found in web services,
+and analyse the performance overhead
+through 
+running benchmarks against a baseline implementation.
 

### 2020-06-18 report/sections/background/related.tex
Report: final proofread

@@
 \textit{Scribble toolchain} initiated by 
-Honda et al. in \cite{Scribble}, 
+Honda et al. \cite{Scribble}, 
 through which the toolchain also facilitates 
@@
 to a role, or \textit{endpoint},
-to obtain a local protocol which represents
+to obtain a \textit{local protocol} 
+which represents
 the global protocol viewed from the perspective
@@
 
+We also inherit the characteristics of EFSMs derived
+from well-formed protocols, as presented in \cite{Hybrid2016}:
+
+\begin{itemize}
+\item 
+There is \textit{exactly one} initial state 
+(\texttt{S1} in \cref{fig:bgadderfsm}).
+
+An initial state has no incoming transitions.
+Formally, the state identifier of the initial state
+is not in the range of $\delta$.
+
+\item 
+There is \textit{at most one} terminal state 
+(\texttt{S2} in \cref{fig:bgadderfsm}).
+
+A terminal state has no outgoing transitions.
+Formally, the state identifier of the terminal state
+is not in the domain of $\delta$.
+
+\item
+Every non-terminal state is either a \textit{send}
+state or a \textit{receive} state.
+
+A send state only has send actions in its
+outgoing transitions. \texttt{S3} and \texttt{S4}
+are the send states in \cref{fig:bgadderfsm}.
+
+A receive state only has receive actions in its
+outgoing transitions. \texttt{S1} is the only
+receive state in \cref{fig:bgadderfsm}.
+\end{itemize}
+
 Developers can use the EFSM to verify their implementation
@@
 which the developer
-can use to implement their endpoint application that guarantees
-the following properties:
+can use to implement their endpoint application
+and enjoy the following guarantees:
 
@@
 into web development using 
-the PureScript language \cite{PureScript},
+the PureScript language \cite{PureScript}.
 PureScript is a strongly typed functional language
@@
 The semantics of the state transition function 
-in the EFSM express that a (state, action) tuple
-\textit{uniquely defines} a successor state. 
+in the EFSM express that a \texttt{(state, action)} tuple
+\textit{uniquely defines} a successor \texttt{state}. 
 These semantics can be expressed by 
@@
 Fowler presented the first formal integration
-of session typing and GUI programming .
+of session typing and GUI programming.
 His work formalised the \textit{Model-View-Update} (MVU)
@@
 As it has a linear type system, session types can be
-implemented using native language features, since the
-language can monitor linear channel usage.
+implemented using native language features to
+monitor linear channel usage.
 Links also supports error handling for session types,

### 2020-06-18 report/sections/background/sessiontypes.tex
Report: final proofread

@@
 The process algebra $\pi$-calculus 
-introduced by Milner in \cite{Milner1999} 
+introduced by Milner \cite{Milner1999} 
 formalises the message passing abstraction in terms of 
@@
 \textit{static} session type checking,
-whilst Neykova and Yoshida \cite{Python2017} 
+whilst Neykova \cite{Python2017} 
 introduced dynamic approaches to 
@@
 with respect to session types.
-We discuss these related work, among others, in \cref{section:bgrelated}.
+We discuss these related work, among others, 
+in \cref{section:bgrelated}.
 
@@
 $\pi$-calculus introduced by 
-Honda and Tokoro in \cite{AsyncHonda}.
+Honda and Tokoro \cite{AsyncHonda}.
 Among the many flavours of the calculus 
@@
 
-A \textit{binary session} is a parallel composition 
-of two processes, each representing a distinct participant.
+A \textit{session} represents 
+the sequence of send and receive actions of 
+a single participant. 
+A \textit{binary session} describes a session
+with two distinct participants.
 In the context of web applications, a binary session may
 describe the interactions between client and server.
-Without loss of generality, a \text{session} represents 
-the sequence of send and receive actions of a single participant.
 
@@
 the input process in \cref{fig:async}
-is generalised as a branching process that offers one branch.
+is generalised as a branching process offering one branch.
 \end{itemize}
@@
 \begin{prooftree}
-\AxiomC{$\exists	j \in I. l_j = l$}
-\AxiomC{$\vec x_j = x_1, \dots, x_n$}
-\AxiomC{$e_1 \downarrow v_1 \dots e_n \downarrow v_n$}
+\AxiomC{$
+\exists j \in I. ~ \left( 
+l_j = l
+\qquad
+\vec x_j = x_1, \dots, x_n 
+\qquad
+e_1 \downarrow v_1 \dots e_n \downarrow v_n
+\right)
+$}
 \AxiomC{$\mrole{p} \neq \mrole{q}$}
 \RightLabel{\rulename{Comm}}
-\QuaternaryInfC{$
+\BinaryInfC{$
 \mrole{p} :: \sel{q}{l}{e_1 \dots e_n}.~P \mid 
@@
 process $P$ has session type $S$}. 
-
 The \textit{typing context} records typing assumptions 
@@
 \begin{cases}
-	\tmsg{Succ(int)}: \tselone{Client}{\tmsg{Succ(int)}}. \tend \\
-	\tmsg{Quit()}:  \tend \\
+	\tmsg{Succ(int)} &: \tselone{Client}{\tmsg{Succ(int)}}.~\tend \\
+	\tmsg{Quit()} &:  \tend \\
 \end{cases}
@@
 Regarding payload, the server is expecting \texttt{int},
-but the client sends \texttt{nat}, which is a subtype of \texttt{int}.
+but the client sends \texttt{nat},
+which is a \textit{subtype} of \texttt{int}.
 
-This motivates the concept of \textit{subtypes} in session types,
+This motivates the concept of subtyping in session types,
 which allows a process to be typed by its ``supertype'' when required. 
@@
 	\begin{cases}
-	T_k & \text{if } k \in (I \setminus J) \\
-	T'_k & \text{if } k \in (J \setminus I) \\
-	\tmerge{T_k}{T'_k} & \text{otherwise} \\
+	T_k & \text{if } k \in I \setminus J \\
+	T'_k & \text{if } k \in J \setminus I \\
+	\tmerge{T_k}{T'_k} & \text{if } ~ k \in I \cap J \\
 	\end{cases}
 	& \\
+	& \text{otherwise undefined} &\\
 

### 2020-06-18 report/sections/background/typescript.tex
Report: feedback
Report: final proofread

@@
 to compile a TypeScript program into JavaScript
-source code; since
-JavaScript is a dynamically typed langauge,
-the compilation process features full type erasure.
+source code, with full type erasure.
 
@@
 
-declare function area(sq: ThisSquare): number;
+const area = (sq: ThisSquare) => sq.side * sq.side;
 

### 2020-06-18 report/sections/evalconcl/concl/contributions.tex
Report: feedback on benchmarks
Report: final proofread

@@
 from a Scribble protocol specification.
-Developers write their full-stack applications
-by implementing the generated APIs, and enjoy 
-communication safety guarantees by construction.
+By writing their full-stack applications using
+the generated APIs, developers enjoy 
+communication safety guarantees in their endpoint
+applications by construction.
 APIs generated for server-side endpoints
@@
 statically provide the same level of communication safety
-guarantees from multiparty session type theory, but we do so
+guarantees, but we do so
 using modern web programming practices to increase
@@
 We also do not limit support to protocols that describe
-a server-centric topology: we formalise 
-a theory of routed multiparty session types in 
+a server-centric topology: we formalise \newtheory,
+a theory of routed multiparty session types, in 
 \cref{chap:theory} to prove that it is possible to relax the
@@
 Our work has received positive feedback from academia.
-The intitial part of this project was published as 
+The initial part of this project was published as 
 \emph{Generating Interactive WebSocket 
@@
 are expanded from \cite{PLACES2020}.
+The materials in 
+\cref{chap:ext,chap:p2p,chap:theory,chap:impl,chap:eval}
+are developed exclusively in this project by myself.
 We were also invited to present a part of this work

### 2020-06-18 report/sections/evalconcl/concl/future.tex
Report: review chapter 3
Report: final proofread

@@
 
-\cite{ExplicitConnections} uses a variant of the Scribble protocol
+The work of Hu and Yoshida \cite{ExplicitConnections} 
+uses a variant of the Scribble protocol
 language that supports explicit connection actions. Future work
@@
 The MPST-based web development framework presented by 
-King et al. \cite{PureScript2019} parameterised
+King et al. \cite{PureScript2019} parameterises
 the transport mechanism in the session runtime definition.
@@
 MPST framework: the permitted methods (i.e. channel actions)
-of the EFSM depends on its current state.
+of the EFSM depend on its current state.
 In fact, Kouzapas et al. \cite{StMungo} presents a way to
@@
 an encoding of session types in object-oriented languages
-and discuss their approach with respect to typestates.
+and discusses their approach with respect to typestates.
 

### 2020-06-18 report/sections/evalconcl/eval.tex
Report: final proofread

@@
 we use two case studies
-of protocols found in webservices
+of protocols found in web services
 to demonstrate strengths and weaknesses

### 2020-06-18 report/sections/evalconcl/eval/benchmarks.tex
Report: final proofread

@@
 \includegraphics[width=0.5\textwidth]{pingpongclient}
-\captionof{figure}{User interface of \tprotocol{Ping Pong} \trole{Client}}
+\captionof{figure}{User Interface of \trole{Client} Endpoint in
+\tprotocol{Ping Pong} Protocol}
 \label{fig:pingpongclient}
@@
 \textit{create-react-app} \cite{cra} for all experiments, which performs the
-transpilation into JavaScript. We serve the production build using the
+compilation into JavaScript. We serve the production build using the
 \textit{serve} package \cite{npmserve} available on \textit{npm}.
@@
 
-\item We use the transpiled JavaScript versions of all \trole{Svr}s
+\item We use the compiled JavaScript versions of all \trole{Svr}s
 for the experiments.
@@
 for the \trole{Client} to connect to the \trole{Server}
-and complete the parameterised number of round trips.
+\textit{and} complete the parameterised number of round trips.
 We compare the total execution time (\textit{Exec. Time}) 
@@
 \end{lstlisting}
-\captionof{lstlisting}{Preventing user-triggered linearity violation in
+\captionof{lstlisting}{Preventing Channel Linearity Violation in
 \texttt{bare_safe} \tprotocol{Ping Pong}}
@@
 \end{lstlisting}
-\caption{\texttt{bare} implementation of \tprotocol{Ping Pong} \trole{Svr}}
+\caption{\texttt{bare} Implementation of \tprotocol{Ping Pong} \trole{Svr}}
 \label{lst:nodepingpongbare}
@@
 \begin{lstlisting}[language=javascript,tabsize=2]
-const logic = new Implementat.S14({
+const logic = new Implementation.S14({
 	PING: (count) => {
@@
 \end{lstlisting}
-\caption{\texttt{mpst} implementation of \tprotocol{Ping Pong} \trole{Svr}}
+\caption{\texttt{mpst} Implementation of \tprotocol{Ping Pong} \trole{Svr}}
 \label{lst:nodepingpongmpst}
@@
 \caption{Comparison of \texttt{bare} and \texttt{mpst}
-implementations for \tprotocol{Ping Pong} \trole{Svr}}
+Implementations for \tprotocol{Ping Pong} \trole{Svr}}
 \label{fig:nodepingpong}

### 2020-06-18 report/sections/evalconcl/eval/game.tex
Report: final proofread

@@
 \end{lstlisting}
-\captionof{lstlisting}{Implementing \tprotocol{Noughts and Crosses} Game Server using
-\codegen}
+\captionof{lstlisting}{Implementing \tprotocol{Noughts and Crosses} Game Server}
 \label{lst:gamesvr}
@@
 			const sendPoint = (event: React.MouseEvent) => {
-				// Don't refresh the screen
-				event.preventDefault();
 				return { x, y };

### 2020-06-18 report/sections/general/p2p/approach.tex
Report: final proofread

@@
 such as the WebRTC framework \cite{WebRTC}. 
-However, WebRTC has its own proprietary ``protocol'' 
+However, WebRTC has its own ``protocol'' 
 for establishing connections and exchanging network information,
@@
 We believe that adopting WebRTC deviates from the focus of project in 
-implementing type-safe webservices.
+implementing type-safe web services.
 

### 2020-06-18 report/sections/general/p2p/challenges.tex
Report: final proofread

@@
 By existing LTS semantics over
-global types \cite{characterisation}, 
+global types introduced in 
+Definition 3.2 of \cite{characterisation}, 
 then $\aout{S}{B}{\tmsg{M2}}$
@@
 Our proposal recognises that, in the context of web applications,
-there is a need for server-side endpoints to implement business logic
+\textit{there is a need for server-side endpoints to 
+implement business logic},
 so enforcing the server to be an exclusive orchestrator makes our work
-incompatible with webservice protocols; 
+incompatible with web service protocols; 
 we also argue that the server can still participate in the 
@@
 does not overserialise the permitted interactions. 
-We introduce our extension as \textit{Routed Session Types},
+We introduce our extension as \newtheory,
 and proceed to prove these claims in \cref{chap:theory}.
@@
 Likewise, the implementation of routed session types 
-in \fancyname{SessionTS} needs to accurately reflect the theory
-and be as transparent to the developer as possible to minimise
+in \codegen needs to accurately reflect the theory.
+It also needs to be 
+as transparent to the developer as possible to minimise
 the learning curve overhead demanded 
-by complicating the generated APIs -- we document our extensions
-to \fancyname{SessionTS} in \cref{chap:impl}.
+by complicating the generated APIs.
+We document our extensions
+to \codegen in \cref{chap:impl}.

### 2020-06-18 report/sections/general/theory.tex
Report: final proofread

@@
 syntax extensions in \cref{section:lts}.
-We define a router-parameterised encoding
-for MPST into \newtheory in \cref{section:encoding}
+Regarding the extended semantics on routed session types,
+we prove the soundness and completeness of
+the extended semantics in \cref{section:newtraceeq}.
+In \cref{section:encoding},
+we define a router-parameterised encoding
+for canonical MPST theory into \newtheory
 and prove the preservation of well-formedness
@@
 We introduced extensions to syntax and semantics, and proved
-that our extended semantics preserve deadlock-freedom
-for well-formed protocols. We defined an encoding from the
+that our extended semantics are sound and complete,
+and preserve deadlock-freedom
+for well-formed protocols. 
+We defined an encoding from the
 canonical theory onto \newtheory, and proved the preservation

### 2020-06-18 report/sections/general/theory/lts.tex
Report: final proofread

@@
 for \newtheory,
-building upon the work of Deni\'elou and Yoshida in
+building upon the work of Deni\'elou and Yoshida
 \cite{characterisation}.
@@
 \begin{itemize}
-\item \textbf{Routed send:}
+\item \textbf{Routed Send:}
 
@@
 
-\item \textbf{Routed receive:}
+\item \textbf{Routed Receive:}
 
@@
 been sent from $\role{p}$ to the router $\role{s}$
-but not yet routed to $\role{q}$.
+\textit{but not yet routed to $\role{q}$}.
 We extend the projection operator to 
@@
 
+Because the router $\role{s}$ holds
+Because the router holds a ``global perspective''
+on the routed communication, we also need to
+represent the intermediate state (at which the message is 
+being routed by the router) within the grammar
+of \textit{local types}.
+We extend the grammar of local types with the construct
+$\routertrans{p}{q}{j}{l_i: T_i}{i \in I}$
+to represent that, from the local perspective of
+the router, the message $l_j$
+has been received from $\role{p}$ \textit{but not yet routed
+to $\role{q}$}.
+
 Because routed communication is treated differently
@@
 in \cref{fig:newglobal}.
-We highlight and explain the new rules.
+We highlight and explain the new rules.\\
 
@@
 \label{subsection:newltslocal}
-
-The LTS semantics over local types
-need to be extended for routed communication.
-Because the router holds a ``global perspective''
-on the routed communication, we also need to
-represent the intermediate state (at which the message is 
-being routed by the router) within the grammar
-of local types.
-Similar to \cref{subsection:newltsglobal}, we add 
-the construct 
-$\routertrans{p}{q}{j}{l_i: T_i}{i \in I}$
-to represent that, from the local perspective of
-the router, the message $l_j$
-has been received from $\role{p}$ but not yet routed
-to $\role{q}$.
 

### 2020-06-18 report/sections/general/theory/syntax.tex
Report: final proofread

@@
 we need to handle \textit{routed communication} separately
-from normal send and receive actions.
+from direct send and receive actions.
 We define the syntax of global types for \newtheory 
@@
 
-\item \textbf{Routed communication:}
+\item \textbf{Routed Communication:}
 
@@
 
-\item \textbf{Routed slection:}
+\item \textbf{Routed Selection:}
 
@@
 
-\item \textbf{Routed branching:}
+\item \textbf{Routed Branching:}
 
@@
 \[
-\begin{array}{rcl}
+\begin{array}{rl}
 \tmerge
-{(\tselproxy{p}{q}{l_i: T_i}{i \in I})}
-{(\tselproxy{p}{q}{l_i: T_i}{i \in I})}
-	& = & \tselproxy{p}{q}{l_i: T_i}{i \in I} \\
+{\tselproxy{p}{q}{l_i: T_i}{i \in I}}
+{\tselproxy{p}{q}{l_i: T_i}{i \in I}}
+	& = \tselproxy{p}{q}{l_i: T_i}{i \in I} \\
 
 \tmerge
-{(\tbraproxy{p}{q}{l_i: T_i}{i \in I})}
-{(\tbraproxy{p}{q}{l_j: T'_j}{j \in J})}
-	& = & \tbraproxy{p}{q}{l_k: T''_k}{k \in I \cup J} \\
-\text{where} & & T''_k = \begin{cases}
+{\tbraproxy{p}{q}{l_i: T_i}{i \in I}}
+{\tbraproxy{p}{q}{l_j: T'_j}{j \in J}}
+	& = \tbraproxy{p}{q}{l_k: T''_k}{k \in I \cup J} \\
+\text{where} & T''_k = \begin{cases}
 T_k & \text{if} ~ k \in I \setminus J \\
 T'_k & \text{if} ~ k \in J \setminus I \\
-\tmerge{T_k}{T'_k} & \text{if} ~ k \in I \cap J \\
+\tmerge{T_k}{T'_k} & \text{if } ~ k \in I \cap J \\
+\end{cases} \\
 \text{otherwise} & \text{undefined}
-\end{cases} \\
 \end{array}
@@
 
-In \newtheory, we require that there is \textit{exactly
-one} router, say $\role{s}$, in the global type, so
-we need to express that a global type to be
+In \newtheory, we need to express that a global type is
 well-formed \textit{with respect to the role $\role{s}$
@@
 For direct communication, $\role{s}$ must
-be a participant and the centroid of all continuations.
+be a participant and a centroid of all continuations.
 
@@
 For routed communication, $\role{s}$ must
-be the router and be the centroid of all continuations.
+be the router and be a centroid of all continuations.
 
@@
 
-We also assume that $G$ has no free type variables
-and is guarded/contractive.
+We also assume that the syntax of $G$ is \textit{contractive}, 
+i.e. that type variables are guarded.
 

### 2020-06-18 report/sections/introduction/contributions.tex
Report: final proofread

@@
 over a server-centric WebSocket-based network topology.
+
 Related work \citep{PureScript2019, MVU2020, LINKS, Exceptional} 
@@
 guarantees using familiar industry tools in TypeScript and React,
-and supports peer-to-peer browser interactions over WebSockets.
+and supports peer-to-peer browser interactions over WebSockets. \\
 
@@
 and provide an overview of related work on integrating session types into
-application development. We also introduce TypeScript and highlight the 
+application development, with a focus on the state-of-the-art
+proposals for session-typed web development. 
+We also introduce TypeScript and highlight the 
 defining features of the language.
@@
 we extend \codegen to support advanced
-practices commonly found in full-stack web development in a way that
+practices commonly found in full-stack web development,
+in a way that
 preserves communication safety. 
@@
 protocol to motivate our proposal of providing the server role with 
-routing responsibilities in addition to its existing interactions
+routing responsibilities, in addition to its existing interactions
 specified in the protocol.
@@
 generated APIs against a baseline implementation to evaluate
-the overhead of our work.
+the overhead of our implementation.
 

### 2020-06-18 report/sections/introduction/motivation.tex
Report: final proofread

@@
 the \trole{FlightServer} will \textit{reserve a seat}
-responds with the price,
+and respond with the price,
 to which the \trole{Traveller} confirms or rejects the purchase;
@@
 conventional for modern web programming.
-Developers that use \cite{MVU2020} are required
+Fowler \cite{MVU2020} requires developers
 to implement their web applications in a \textit{functional} 
@@
 web programming language \cite{LINKS}.
-\cite{PureScript2019} also endorses
+King et al. \cite{PureScript2019} also endorses
 functional web programming by requiring developers to
@@
 Whilst communication safety guarantees provided by
-\cite{PureScript2019,MVU2020} are based on
+Fowler \cite{MVU2020} and King et al. \cite{PureScript2019}
+are based on
 canonical MPST theory, they are not contextualised
@@
 
-The current state-of-the-art leverage the WebSocket protocol
+The current state of the art leverage the WebSocket protocol
 as the communication transport between roles, which is
@@
 so the \trole{Traveller} can compare between both options.
-The current state-of-the-art cannot implement this 
+The current state of the art cannot implement this 
 extended flight booking service, as the described
-communication interactions is no longer server-centric.
+communication interactions are no longer server-centric.

### 2020-06-18 report/sections/introduction/objectives.tex
Report: final proofread

@@
 We tackle the limitations of the current
-state-of-the-art with two solutions:
+state of the art with two solutions:
 
@@
 processes that implement \newtheory also enjoy
-guarantees of \textbf{deadlock-freedom}, \textbf{channel linearity},
+guarantees of \textbf{deadlock-freedom}
 and the absence
@@
 We also extend \codegen to implement \newtheory.
+\codegen generates APIs that provide \textbf{channel linearity}
+by construction, and the extended
+implementation to support \newtheory enjoys the same benefits.
 \\
@@
 Our work provides the same communication safety guarantees
-as the current state-of-the-art, but 
+as the current state of the art, but 
 is compatible with modern web programming tools and idioms.
@@
 that the developer may incorporate into their application.
-Whilst some \cite{MVU2020} point out this limits its usability for
+Whilst some \cite{MVU2020} point out that 
+this limits its usability for
 encoding multiparty session types, we believe that the language
@@
 found in modern web development (such as callbacks and
-asynchronous operations), and provides an error handling
+asynchronous operations), and provide an error handling
 mechanism tailored to the dynamic web-based environment

### 2020-06-18 report/sections/server/codegen/testing.tex
Report: final proofread

@@
 \section{Testing}
-
-The usual unit testing practices from software engineering
-apply to our code generator -- the built-in \texttt{unittest} 
-package is more than sufficient for our use case. 
-Instead, we detail our methodology for \textit{system testing} -- 
-executing \fancyname{SessionTS} end-to-end and testing
+The challenge for testing \codegen
+is to verify that the generated code is valid
+TypeScript code.
+Here, we detail our methodology
+for \textit{system testing} -- 
+executing \codegen end-to-end and testing
 the generated code.
@@
 We implement a test suite on top of \texttt{unittest} APIs
-for verifying that \fancyname{SessionTS}
+for verifying that \codegen
 generates \textit{valid TypeScript code}. This is especially useful as
@@
 WebSocket package installed on server-side endpoints), we require a
-sandbox environment to type-check the generated code. 
+\textit{sandbox environment} to type-check the generated code. 
 The sandbox contains the minimal boilerplate required

### 2020-06-18 report/sections/server/extensions/async.tex
Report: final proofread

@@
 even though the callback function has the expected return type.
-The developer will encounter the following:
+The developer will encounter the following type error message:
 
@@
 
-In order to make \fancyname{SessionTS} more relevant and compatible
+In order to make \codegen more relevant and compatible
 with modern web programming practices, it is important to
@@
 \begin{lstlisting}[language=javascript,numbers=none]
-setTimeout(() => console.log('First!', 0);
+setTimeout(() => console.log('First!', 0));
 console.log('Second!');
@@
 \end{lstlisting}
-\captionof{lstlisting}{\tprotocol{2FA} \trole{Svr} Implementation using 
+\captionof{lstlisting}{Implementing \trole{Svr} in \tprotocol{2FA} using 
 Callbacks, \texttt{Promise}s and
@@
 Generally, minimal changes are required for the runtime.
-The key observation is that the concrete types for the any
-handlers are the same -- the payload may be wrapped in a Promise
+The key observation is that the ``concrete'' types for the any
+handlers are the same; the payload may simply be wrapped in a Promise
 construct.
 To handle potential Promises, we wrap existing logic
-as a delayed computation, colloquially known as a \textit{thunk}
+into a delayed computation, colloquially known as a \textit{thunk}
 \cite{Thunk}.
@@
 \captionof{lstlisting}{Generating Runtime Linearity Checks for
-React Send State Components}
+React Send States}
 \label{lst:asynclinearcheck}

### 2020-06-18 report/sections/server/extensions/error.tex
Report: final proofread

@@
 respectively. 
-To respect the convention specified in the WebSocket protocol
-RFC \cite{WebSocketRFC} that the 1xxx close codes 
+To respect the convention specified in the WebSocket Protocol
+\cite{WebSocketRFC} that the 1xxx close codes 
 are reserved for internal use,

### 2020-06-18 report/sections/server/node.tex
Report: final proofread

@@
 the EFSM for server-side endpoints (\cref{section:nodeefsm}),
-and detail the session runtime that is designed
+and present a session runtime designed
 to execute the EFSM on Node.js (\cref{section:noderuntime}).
@@
 \captionof{figure}{\trole{Svr} Endpoint FSM
-in \tprotocol{Adder} protocol}
+in \tprotocol{Adder} Protocol}
 \label{fig:addersvrfsm}

### 2020-06-18 report/sections/server/node/efsm.tex
Report: final proofread

@@
 to define what message to send and how to process a received
-message, without access to the send and receive APIs
+message, without accessing the send and receive APIs
 that are used to execute the EFSM.
@@
 
-We attempt to address this by defining an enum of state identifiers
+We \textit{attempt} to address this by defining an enum of state identifiers
 for each type of state (i.e. an enum for send states, 
@@
 
-Unfortunately, this approach does not work for the simple fact that
+Unfortunately, this approach \textbf{does not work}
+for the simple fact that
 conditional types were not designed to be exploited in this manner.
 The main limitation of conditional types is its 
-\textit{distributivity} when the type ``parameter'' is an
+\textit{distributivity} when the type parameter is an
 union type (which is the case for enums, as
-\texttt{S = SendState.S53 | SendState.S54| ...}), where
+\texttt{S = SendState.S53 | SendState.S54 | ...}), where
 
@@
 (\cref{section:noderuntime}), where we disclose more details regarding
-the role of the \texttt{Implemenation} wrapper API in
+the role of the \texttt{Implementation} wrapper API in
 the runtime.

### 2020-06-18 report/sections/server/node/limitations.tex
Report: final proofread

--- 2020-06-05 report/sections/server/node/limitations.tex
@@
 does not help code readability, as it is neither apparent nor
-self-documenting what \texttt{S51} refers to -- some existing work
-\cite{Hybrid2016} augment the state identifiers with the channel
+self-documenting what \texttt{S51} refers to -- the work of
+Hu and Yoshida
+\cite{Hybrid2016} augment
+the state identifiers with the channel
 action and labels involved (e.g. 
@@
 which does not scale elegantly as the levels of nesting
-increase. For example, the following implementation of
+increase. 
+For example, the following implementation of
 the \tprotocol{One Adder} protocol would be invalid if
@@
 storing values in a database and retrieving them for
-handlers of continuation states), so our generated APIs
+handlers of continuation states), 
+so our generated APIs
 do not \textit{strictly} enforce scoping to be the only way to

### 2020-06-18 report/sections/server/node/runtime.tex
Report: final proofread

@@
 and application logic
-(i.e. their handler implementations).
+(i.e. implementations of the handler APIs).
 It is developer's responsibility to construct the
@@
 
-	// Ignore if role is aleady taken
+	// Ignore if role is already taken
 	if (!waiting.has(role)) { return socket.close(); } (*@\label{line:occupied}@*)
@@
 Finally, we use the parameterised EFSM transition handler
-to notify the rutime which specific state to transition to.
+to notify the runtime which specific state to transition to.
 
@@
 They are decoded using the same interface schema on the receiving end
-using \texttt{JSON.parse()}: whilst the method return type is the
+using \texttt{JSON.parse()}.
+Whilst the method return type is the
 dynamic \lstonelinejs{any} type, 
@@
 	// Optional type, same as `MessageHandler | undefined`
-	private handler?: MessageHandler; | undefined(*@\label{line:noderuntimehandler}@*)
+	private handler?: MessageHandler; (*@\label{line:noderuntimehandler}@*)
 
@@
 \[
-A \to S: \text{\lstonelinejs{M1(string)}}. 
-B \to S: \text{\lstonelinejs{M2(number)}}. \texttt{end} 
+A \to S: \text{\lstonelinejs{M1(string)}}.~ 
+B \to S: \text{\lstonelinejs{M2(number)}}.~\texttt{end} 
 \]
@@
 \[
-A \to S: \text{\lstonelinejs{M1(string)}}. 
-A \to S: \text{\lstonelinejs{M2(number)}}. \texttt{end}
+A \to S: \text{\lstonelinejs{M1(string)}}.~ 
+A \to S: \text{\lstonelinejs{M2(number)}}.~\texttt{end}
 \]
@@
 \end{subfigure}
-\caption{Possible Orderings for Handling Message Event and Preparing
-Receive State}
+\caption{Possible Orderings for Receiving Message and
+Registering Handler}
 \label{fig:nodereceivecompare}

### 2020-06-18 report/sections/server/react.tex
Report: final proofread

@@
 and implement the session runtime (\cref{section:reactruntime}),
-focusing on our novel approach to preventing
+focusing on our novel approach for preventing
 channel reuse in GUI programming using React.
-We conclude by analysing some alternative designs
+We conclude by analysing alternative designs
 (\cref{section:reactalt})
@@
 \captionof{figure}{\trole{Client} Endpoint FSM
-in \tprotocol{Adder} protocol}
+in \tprotocol{Adder} Protocol}
 \label{fig:adderclientfsm}

### 2020-06-18 report/sections/server/react/approach.tex
Report: final proofread

@@
 between UI elements, and empowers the UI to subscribe and
-``react'' to data changes.
+``react'' to data changes;
+we introduce the framework in \cref{subsection:react}.
 We aim to implement similar behaviour with respect to the EFSM:
@@
 are those permitted by the current EFSM state.
-We introduce the framework in \cref{subsection:react}.
 
-When executing \fancyname{SessionTS} to generate code
+When executing \reactcodegen to generate code
 for the \trole{Client} endpoint specifying the

### 2020-06-18 report/sections/server/react/challenges.tex
Report: final proofread

@@
 
-Our goal with \fancyname{ReactMPST} is to
+Our goal with \reactcodegen is to
 generate session type implementations for the web browser.
@@
 tackle browser-side session typing.
-Fowler introduced the concept of model types to prevent
+Fowler \cite{MVU2020} introduced the concept of model types to 
+prevent
 channel linearity violation in his proposal for
 integrating session types with
-graphical user interface (GUI) programming \cite{MVU2020}, but
+GUI programming, but
 it uses the Links web programming language \cite{LINKS} which
@@
 Through our work, we aim to distil the key
-concepts from existing work that provide session type safety
+concepts from \cite{MVU2020,LINKS,PureScript2019} 
+that provide session type safety
 for web-based GUI programming, and implement them using
-mainstream front-end web development tools (i.e. TypeScript
+mainstream front-end web development tools (specifically TypeScript
 and React), to provide developers with an intuitive way to implement

### 2020-06-18 report/sections/server/react/efsm.tex
Report: final proofread

@@
 the developer to customise this through specifying an 
-interface (\cref{line:clientsendstatestate}).
+interface (\cref{line:clientsendstatestate}) as the generic type.
 
@@
 Swapping \cref{line:clientsendpayload1,line:clientsendpayload2}
-will result in a compile-time type error due to invalid 
+will result in a compile time type error due to invalid 
 payload types, which is the intended behaviour.
@@
 of conditional types, and \textbf{(2)} we know the type
-of \texttt{ev} at compile-time from the developer's implementation,
+of \texttt{ev} at compile time from the developer's implementation,
 so we are not resolving index type queries with union types.
@@
 \end{lstlisting}
-\captionof{lstlisting}{Snippet of the \texttt{React.DOMAttributes} Interface}
+\captionof{lstlisting}{Snippet of the 
+\texttt{React.DOMAttributes} Interface}
 \label{lst:reactdomattrs}
@@
 \captionof{lstlisting}{Generated Type for \trole{Client} Send State
-in \tprotocol{Adder} protocol}
+in \tprotocol{Adder} Protocol}
 \label{lst:reactsendcomponent}
@@
 \captionof{lstlisting}{Generated Code for \trole{Client} 
-Receive State S42 in \tprotocol{Adder} protocol}
+Receive State S42 in \tprotocol{Adder} Protocol}
 \label{lst:reactreceivecomponent}

### 2020-06-18 report/sections/server/react/runtime.tex
Report: final proofread

@@
 There are many parallels between the runtime we generate
-for \fancyname{ReactMPST} compared to that of \fancyname{NodeMPST}.
+for \reactcodegen compared to that of \nodecodegen.
 We focus on the details unique to React endpoints.
@@
 type Props = {
-	endpoint: string,
+	endpoint: string,				// WebSocket URL to connect to
 	states: {						// mapping of state identifier
@@
 
-export class Session extends React.Component<Props, ...
+export class Session extends React.Component<Props> { ... }
 \end{lstlisting}
@@
 to concrete implementations. 
-The developer only needs to pass the \textit{constructor} function,
+The developer only needs to pass the \textit{constructor function},
 as the runtime will instantiate the React components separately
@@
 			<h1>Adder Client</h1>
-			<Client
+			<Session
 				endpoint='ws://localhost:8080'
@@
 for performance \cite{ReactState}, so it is possible for the WebSocket message
-event to be queued (and hence, procesesd) before the receive handler
+event to be queued (and hence, processed) before the receive handler
 is registered.
@@
 \end{lstlisting}
-\captionof{lstlisting}{Receive Handler Registration and Message Event
-Listener in \fancyname{ReactMPST}}
+\captionof{lstlisting}{Receive Handler Registration and Message Handler
+in \fancyname{ReactMPST}}
 \label{lst:reactreceiveruntime}

### 2020-06-18 report/sections/acknowledgements.tex
Report: ack

@@
 \begin{abstract}
-These are some acknowledgements.
+
+I would like to thank Prof. Nobuko Yoshida,
+Fangyi Zhou and Dr. Francisco Ferreira for their
+continuous support, encouragement and motivation
+during the project. 
+I extend my gratitude to the members
+of the Mobility Reading Group.
+Their regular meetings have been a source
+of inspiration for me, and helped me spark new ideas
+to incorporate into the project.
+I would also like to thank Prof. Peter Pietzuch
+for supporting me as my personal tutor
+throughout the degree.
+\\
+
+I would like to thank all the friends I have made
+throughout my time at Imperial.
+In particular, to the friends under the namespace
+labelled \textit{Reply 404} at the time of writing:
+I am forever grateful for every bit of shared memory
+over the past four years.
+\\
+
+Finally, I would like to thank my family
+for their unconditional love and support throughout my life,
+and encouraging me to pursue a degree abroad.
+
 \end{abstract}
